---
title: "Adding distribution families"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding distribution families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article describes the workflow for adding a distribution family to
`drmTMB`. It is written for contributors, not end users. A new family is not a
single constructor in `R/family.R`; it is a contract across symbolic equations,
formula grammar, R builders, TMB likelihood code, simulation, tests,
documentation, pkgdown, and after-task notes.

The planning references are `docs/design/02-family-registry.md`,
`docs/design/03-likelihoods.md`, `docs/design/05-testing-strategy.md`, and
`docs/design/19-family-link-contract.md`, plus
`.agents/skills/add-family/SKILL.md`.

## Scope first

Every new family must stay inside the package boundary:

- one response or two responses only;
- one formula per distributional parameter;
- explicit parameter names such as `mu`, `sigma`, `nu`, `rho12`, and future
  second-shape `tau`;
- no higher-dimensional multivariate response models;
- no family is complete without simulation tests and documentation.

In this article, location means a mean-like parameter such as `mu`; scale means
a residual spread or dispersion-like parameter such as `sigma`; shape means a
family-specific parameter such as Student-t `nu`; and coscale means residual
coupling, represented in the bivariate Gaussian model by `rho12`.

Do not add a family only because another package has it. Add one when it serves
a distributional-regression question that `drmTMB` can answer clearly.

## Start with the mathematical contract

Before touching code, write the model as equations. The implemented Student-t
family is the simplest current example of a location-scale-shape family:

```text
y_i | mu_i, sigma_i, nu_i ~ Student-t(mu_i, sigma_i, nu_i)
mu_i = X_mu[i, ] beta_mu
log(sigma_i) = X_sigma[i, ] beta_sigma
nu_i = 2 + exp(X_nu[i, ] beta_nu)
```

The matching R syntax is:

```{r student-family-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(
    y ~ x1,
    sigma ~ x2,
    nu ~ x3
  ),
  family = student(),
  data = dat
)
```

This equation-syntax pair decides almost everything that follows. It tells the
parser which distributional parameters are legal, the builder which model
matrices to create, the TMB template which transformed parameters to use, and
the tests what data-generating process to simulate.

For a two-response family, also define what any residual coupling parameter
means. The implemented all-Gaussian composed family uses:

```text
[y1_i, y2_i]' ~ MVN([mu1_i, mu2_i]', Omega_i)
Omega_i =
  [ sigma1_i^2                         rho12_i sigma1_i sigma2_i ]
  [ rho12_i sigma1_i sigma2_i          sigma2_i^2                ]
rho12_i = 0.99999999 * tanh(eta_rho12_i)
```

with:

```{r bivariate-family-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

For later mixed composed families, such as a Gaussian response plus a count
response, do not assume `rho12` is automatically meaningful. The family design
must say whether `rho12` is an observed residual correlation, a latent residual
correlation, a copula parameter, or unsupported.

## Define the registry entry

Each family needs a small structured contract. The design document lists these
fields:

```text
name
n_response
dpars
links
inverse_links
bounds
density_id
simulate
starting_values
check_data
```

The current lightweight R family constructors expose the pieces that are needed
by the implemented builders:

```{r student-registry, eval = FALSE}
student <- function() {
  structure(
    list(
      name = "student",
      family = "student",
      n_response = 1L,
      dpars = c("mu", "sigma", "nu"),
      links = c(mu = "identity", sigma = "log", nu = "logm2")
    ),
    class = "drm_family"
  )
}
```

Use the canonical GAMLSS-style names unless there is a strong reason not to:

```text
mu     location or mean-like parameter
sigma  scale, residual SD, or dispersion-like parameter
nu     first shape parameter
tau    second shape parameter
```

Aliases such as `skew`, `df`, or `shape` should wait until the canonical grammar
is stable. If an alias is ever added, tests must show that it resolves to the
same internal parameter before model matrices are built.

## Do not assume identity links

The name `mu` does not automatically mean "identity-linked arithmetic mean".
That is true for Gaussian models, but it is not a general rule.

For the implemented lognormal family:

```text
log(y_i) | mu_i, sigma_i ~ Normal(mu_i, sigma_i^2)
E[y_i] = exp(mu_i + sigma_i^2 / 2)
```

`predict(fit, dpar = "mu")` returns the mean of `log(y)`, while `fitted(fit)`
returns the arithmetic mean of `y`. The implemented Gamma mean-CV route uses
`log(mu)` so `mu` is positive and equal to `E[y]`. A future beta family will
need `logit(mu)` because `mu` is a probability.

Before adding a family with non-identity `mu`, update the family link contract
and add tests for all three quantities:

```text
predict(type = "link")
predict(type = "response")
fitted()
```

Those tests protect the API distinction between a distributional parameter and
the expected response.

## Connect the family to the builder

A family constructor is not enough. The high-level fitting function has to route
the family to a model builder. In broad terms, that builder must:

1. validate the formula only uses legal distributional parameters;
2. reject unsupported neighbouring syntax before optimization;
3. build one model matrix for each distributional parameter;
4. create starting values and parameter maps;
5. assemble the data list passed to TMB;
6. record enough metadata for prediction, simulation, summaries, and checks.

For the Student-t family, the builder currently rejects random effects,
`meta_known_V(V = V)`, random-effect scale formulae, and bivariate terms. That is
intentional. A clear unsupported-feature error is better than a plausible fit
whose likelihood has not been implemented.

When adding a family, update formula grammar docs only after the fitting path
and tests exist. If syntax is planned but not fitted, label it as planned.

## Implement the likelihood

The likelihood should be written on stable transformed scales. For the
implemented Student-t family, TMB receives unconstrained linear predictors and
uses:

```text
sigma_i = exp(eta_sigma_i)
nu_i = 2 + exp(eta_nu_i)
z_i = (y_i - mu_i) / sigma_i
```

The log density is:

```text
log f(y_i) =
  lgamma((nu_i + 1) / 2) -
  lgamma(nu_i / 2) -
  0.5 log(nu_i pi) -
  log(sigma_i) -
  0.5 (nu_i + 1) log(1 + z_i^2 / nu_i)
```

The `-log(sigma_i)` term is part of the scaled Student-t density. Likelihood
constants matter because `logLik()`, AIC, BIC, comparator tests, and profile
likelihoods depend on them.

For a new family, document:

- the response-scale parameter definitions;
- the link functions and inverse links;
- the parameter bounds;
- any density constants;
- the starting-value strategy;
- what is reported and predicted on the response scale.

## Add simulation and method support

Every family should be able to simulate from a fitted model. Simulation is not
just a convenience for users; it is how we test whether the likelihood and
extractors agree with the symbolic model.

For a new univariate family, check these methods:

```{r method-checks, eval = FALSE}
coef(fit, "mu")
coef(fit, "sigma")
predict(fit, dpar = "mu")
predict(fit, dpar = "sigma")
simulate(fit, nsim = 2, seed = 1)
residuals(fit)
check_drm(fit)
summary(fit)
```

Add family-specific diagnostics when a parameter has a boundary or biological
interpretation risk. The implemented Student-t path adds `check_drm()` guidance
for `nu`, because very small `nu` is boundary-prone and very large `nu` means
the likelihood is close to Gaussian.

## Write tests before calling the family implemented

A minimal family test set should include four layers.

First, test parameter recovery from a known data-generating process:

```{r recovery-test-pattern, eval = FALSE}
set.seed(1)
dat <- simulate_from_known_parameters()

fit <- drmTMB(
  drm_formula(y ~ x1, sigma ~ x2, nu ~ x3),
  family = student(),
  data = dat
)

expect_equal(fit$opt$convergence, 0)
expect_lt(max(abs(coef(fit, "mu") - beta_mu)), tolerance_mu)
expect_lt(max(abs(coef(fit, "sigma") - beta_sigma)), tolerance_sigma)
expect_lt(max(abs(coef(fit, "nu") - beta_nu)), tolerance_nu)
```

Second, add an independent likelihood check when the density is simple enough:

```{r independent-likelihood-pattern, eval = FALSE}
loglik <- sum(
  stats::dt((y - mu) / sigma, df = nu, log = TRUE) - log(sigma)
)

expect_equal(
  as.numeric(logLik(fit)),
  loglik,
  tolerance = 1e-8
)
```

Third, test methods such as prediction, simulation, residuals, summaries, and
diagnostics.

Fourth, test rejection paths. Unsupported combinations should fail before TMB
optimization with useful messages:

```{r rejection-pattern, eval = FALSE}
expect_error(
  drmTMB(
    drm_formula(y ~ x + meta_known_V(V = V), sigma ~ 1),
    family = student(),
    data = dat
  ),
  "not implemented"
)
```

Use comparator packages only where the parameterization overlaps cleanly. For
example, `metafor` is useful for Gaussian meta-analysis with known sampling
covariance, and `lme4` is useful for overlapping Gaussian mixed models.
Comparator tests should use `testthat::skip_if_not_installed()` so package
checks remain portable.

## Update user and developer documentation

A family task is incomplete until the documentation tells the same story as the
code and tests.

Update these files when relevant:

```text
R/family.R
R/drmTMB.R
R/methods.R
R/check.R
src/drmTMB.cpp
tests/testthat/test-<family>.R
docs/design/02-family-registry.md
docs/design/03-likelihoods.md
docs/design/05-testing-strategy.md
docs/design/06-distribution-roadmap.md
vignettes/distribution-families.Rmd
vignettes/testing-likelihoods.Rmd
README.md
NEWS.md
_pkgdown.yml
```

This list is not automatic ownership. Touch only the files affected by the
feature. The point is to check whether each public claim remains true.

For exported family constructors, add roxygen documentation and examples. If
the family changes the user-facing learning path, add or update a vignette and
make sure `_pkgdown.yml` exposes it in the right section.

## Close with an after-task audit

Before merging or pushing a family change, use the after-task audit. The
minimum closure record should answer:

- What model is now implemented?
- Which equations define the likelihood?
- Which R syntax is supported?
- Which unsupported neighbouring syntax is rejected?
- Which simulation, comparator, or independent likelihood checks passed?
- Did `devtools::document()`, `devtools::test()`, `pkgdown::check_pkgdown()`,
  `pkgdown::build_site()`, and `devtools::check()` pass?
- Which docs, roadmap entries, NEWS bullets, and known-limitations notes were
  synchronized?
- What did not go smoothly?

Write the answer in `docs/dev-log/check-log.md` and in a compact after-task
file under `docs/dev-log/after-task/`.

The goal is not bureaucracy. It is how `drmTMB` avoids the usual modelling
package failure mode: code, equations, examples, and tests drifting apart while
the public API still looks confident.
