---
title: "Animal models and additive relatedness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Animal models and additive relatedness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144,
  dev.args = list(bg = "white")
)
library(drmTMB)

animal_guide_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA),
      legend.position = "bottom"
    )
}

animal_eye_theme <- function() {
  animal_guide_theme() +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_blank(),
      legend.position = "none"
    )
}

relatedness_heatmap_data <- function(x) {
  out <- as.data.frame(as.table(x), stringsAsFactors = FALSE)
  names(out) <- c("row", "column", "relatedness")
  out$row <- factor(out$row, levels = rev(rownames(x)))
  out$column <- factor(out$column, levels = colnames(x))
  out
}

simulate_animal_guide_data <- function(seed = 20260523) {
  set.seed(seed)
  n_individual <- 32
  n_each <- 4
  individual_levels <- paste0("id_", seq_len(n_individual))
  individual <- factor(rep(individual_levels, each = n_each))
  age <- runif(length(individual), -1, 1)
  sex <- factor(rep(rep(c("female", "male"), each = n_each), n_individual / 2))

  A <- outer(seq_len(n_individual), seq_len(n_individual), function(i, j) {
    0.42^abs(i - j)
  })
  diag(A) <- diag(A) + 0.12
  dimnames(A) <- list(individual_levels, individual_levels)
  Ainv <- solve(A)

  animal_effect <- as.vector(t(chol(A)) %*% rnorm(n_individual)) * 0.38
  names(animal_effect) <- individual_levels
  sigma <- exp(-1.10 + 0.12 * (sex == "male"))
  body_size <- 1.6 + 0.45 * age + 0.22 * (sex == "male") +
    animal_effect[individual] + rnorm(length(individual), sd = sigma)

  list(
    data = data.frame(
      body_size = body_size,
      age = age,
      sex = sex,
      individual = individual
    ),
    A = A,
    Ainv = Ainv
  )
}

simulate_animal_q2_guide_data <- function(seed = 20260525) {
  set.seed(seed)
  n_individual <- 32
  n_each <- 4
  individual_levels <- paste0("id_", seq_len(n_individual))
  individual <- factor(rep(individual_levels, each = n_each))
  climate <- rnorm(length(individual))

  A <- outer(seq_len(n_individual), seq_len(n_individual), function(i, j) {
    0.40^abs(i - j)
  })
  diag(A) <- diag(A) + 0.10
  dimnames(A) <- list(individual_levels, individual_levels)
  Ainv <- solve(A)

  z_body <- rnorm(n_individual)
  z_activity <- 0.55 * z_body + sqrt(1 - 0.55^2) * rnorm(n_individual)
  u_body <- as.vector(t(chol(A)) %*% z_body) * 0.40
  u_activity <- as.vector(t(chol(A)) %*% z_activity) * 0.34
  names(u_body) <- individual_levels
  names(u_activity) <- individual_levels

  e_body <- rnorm(length(individual))
  e_activity <- 0.15 * e_body + sqrt(1 - 0.15^2) * rnorm(length(individual))

  list(
    data = data.frame(
      individual = individual,
      climate = climate,
      body_size = 1.2 + 0.30 * climate + u_body[individual] + 0.38 * e_body,
      activity = -0.3 - 0.24 * climate +
        u_activity[individual] + 0.34 * e_activity
    ),
    Ainv = Ainv
  )
}
```

Use `animal()` when individuals have known additive relatedness and the
scientific question is about among-individual location deviations after the
fixed effects have been included. The fitted first slices are Gaussian
location models. They are useful for animal-model and quantitative-genetic
workflows, but they are not yet full parity with every `phylo()` route.

## What is fitted today

| Question | Syntax | Status |
| --- | --- | --- |
| Does one Gaussian response have additive individual deviations? | `animal(1 | individual, pedigree = pedigree)`, `animal(1 | individual, A = A)`, or `animal(1 | individual, Ainv = Ainv)` in `mu` | Fitted first slice. Dense pedigree construction, covariance input, and precision input are supported. |
| Does one numeric fixed-effect slope also have additive individual deviations? | `animal(1 + x | individual, pedigree = pedigree)` in `mu` | Fitted first one-slope slice with independent intercept and slope fields. Multiple structured slopes and intercept-slope correlations remain planned. |
| Does a one-response Gaussian model have additive residual-scale deviations? | `animal(1 | individual, Ainv = Ainv)` in `sigma`, alone or matching the `mu` term | Fitted first residual-scale intercept slice. Matching univariate `mu` and `sigma` animal terms estimate one animal mean-scale correlation; residual-scale structured slopes remain planned. |
| Do two Gaussian response means share an animal-model correlation? | matching labelled `animal(1 | p | individual, pedigree = pedigree)`, `animal(1 | p | individual, A = A)`, or `animal(1 | p | individual, Ainv = Ainv)` terms in `mu1` and `mu2` | Fitted first q=2 location-location slice. `corpairs(level = "animal")` reports the latent animal correlation separately from residual `rho12`. |
| Do animal-model deviations link means and residual scales across two responses? | the same labelled `animal(1 | p | individual, Ainv = Ainv)` term in `mu1`, `mu2`, `sigma1`, and `sigma2` | Fitted first q=4 location-scale slice. `corpairs(level = "animal")` reports six constant latent animal correlations; profile intervals for those derived correlations remain unavailable. |
| Does the animal effect change multiple slopes or a predictor-dependent correlation? | examples such as `animal(1 + x + z | individual, pedigree = pedigree)` or animal `corpair()` formulas | Planned. These routes need separate diagnostics, recovery tests, interval targets, and examples before use. |

The `pedigree` route expects a data frame whose identifiers match the grouping
variable used in the formula. The `A` route uses a precomputed additive
relationship covariance matrix, and the `Ainv` route uses the corresponding
precision matrix. Use exactly one of `pedigree`, `A`, or `Ainv` for a fitted
animal term.

## Start with the smallest useful model

For one response, start with the fitted location-intercept route:

```r
fit_animal <- drmTMB(
  y ~ treatment + animal(1 | individual, pedigree = pedigree),
  data = dat,
  family = gaussian()
)
```

For two responses, use the same animal block label in `mu1` and `mu2`:

```r
fit_animal_q2 <- drmTMB(
  bf(
    mu1 = trait1 ~ treatment +
      animal(1 | p | individual, Ainv = Ainv),
    mu2 = trait2 ~ treatment +
      animal(1 | p | individual, Ainv = Ainv)
  ),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_animal_q2, level = "animal")
rho12(fit_animal_q2)
```

The two extractors answer different questions. `corpairs()` reports the fitted
latent animal-model correlation among individual location deviations. `rho12()`
reports the residual correlation between the two responses after the fixed
effects and random effects have been included.

For reports, do not leave the q=2 animal correlation as a lone dot when interval
evidence is available. The animal q=2 profile interval computes, but it is
recovery-grade and not coverage-checked, unlike the `phylo()` and `relmat()` q=2
mean-mean anchors -- trust the point estimate more than the interval width.
Request interval provenance first, then plot the explicit table:

```r
animal_pairs <- corpairs(fit_animal_q2, level = "animal", conf.int = TRUE)
plot_corpairs(animal_pairs)
```

That plot should be read as an animal-layer correlation display, not as a
residual `rho12` plot. The q=4 rows below are different: current q=4 animal
correlations are derived rows, so their tables may expose point estimates and
`conf.status` without profile Confidence Eyes.

When the biological question is whether additive individual deviations in means
and residual scales covary, use the same labelled animal term in all four
bivariate endpoint formulas:

```r
fit_animal_q4 <- drmTMB(
  bf(
    mu1 = trait1 ~ treatment +
      animal(1 | p | individual, Ainv = Ainv),
    mu2 = trait2 ~ treatment +
      animal(1 | p | individual, Ainv = Ainv),
    sigma1 = ~ treatment +
      animal(1 | p | individual, Ainv = Ainv),
    sigma2 = ~ treatment +
      animal(1 | p | individual, Ainv = Ainv)
  ),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_animal_q4, level = "animal")
```

The q=4 path is constant across individuals and responses: it estimates four
animal endpoint standard deviations and six latent animal correlations. It is
not a direct-SD model and it does not make the residual correlation `rho12`
animal-structured.

## What to inspect

After fitting, inspect the animal layer before interpreting it:

| Output | Use |
| --- | --- |
| `check_drm(fit)` | Confirm the structured layer was recognized and review convergence diagnostics. |
| `fit$sdpars$mu` | Read the fitted animal-model location SD. |
| `ranef(fit, "animal_mu")` | Inspect conditional individual deviations on the fitted animal layer. |
| `summary(fit)$covariance` | Check how the animal SD or correlation is reported beside other covariance layers. |
| `profile_targets(fit)` | See which animal SD or q=2 correlation targets can be profiled directly; q=4 correlations are derived-only in the first slice. |
| `corpairs(fit, level = "animal")` | Read the fitted q=2 animal mean-mean correlation, or the six q=4 animal endpoint correlations when the model has the all-four block. |

If the ordinary repeatability model answers a useful sensitivity question, fit
`(1 | individual)` as a comparison. Report it as an ordinary individual
random-effect model, not as an additive animal model.

## Rendered checks

The small examples below keep three evidence grains separate. The relationship
matrix is raw structure supplied to the model. The SD plot is model-estimated
uncertainty from `confint()`. The q=2 animal-correlation plot is a fitted
latent correlation row from `corpairs()`, so it uses a Confidence Eye and a
dotted zero line.

```{r animal-guide-fit}
animal_example <- simulate_animal_guide_data()
animal_dat <- animal_example$data
A <- animal_example$A
Ainv <- animal_example$Ainv

fit_animal <- drmTMB(
  bf(
    body_size ~ age + sex + animal(1 | individual, Ainv = Ainv),
    sigma ~ 1
  ),
  family = gaussian(),
  data = animal_dat
)
```

```{r animal-relatedness-matrix, fig.width = 5.3, fig.height = 4.8, fig.cap = "Additive relatedness matrix used by the animal-model example; this heatmap is the known input structure, not an uncertainty display.", fig.alt = "Heatmap of the additive relatedness matrix for the animal-model example. Values are highest on the diagonal and fade as individuals are farther apart in the simulated pedigree-like ordering."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  animal_matrix <- relatedness_heatmap_data(A)

  ggplot2::ggplot(
    animal_matrix,
    ggplot2::aes(column, row, fill = relatedness)
  ) +
    ggplot2::geom_tile() +
    ggplot2::coord_equal() +
    ggplot2::scale_fill_gradientn(
      colours = c("#F7FBFF", "#C6DBEF", "#6BAED6", "#08519C"),
      name = "Additive\nrelatedness"
    ) +
    animal_guide_theme() +
    ggplot2::theme(
      axis.text = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank()
    ) +
    ggplot2::labs(
      title = "Animal-model input structure",
      subtitle = "Known additive relatedness among individuals",
      x = "Individual",
      y = "Individual"
    )
}
```

```{r animal-sd-figure, fig.width = 6.6, fig.height = 3.4, fig.cap = "Fitted residual `sigma` and animal-model location SD from a univariate Gaussian model; points and bars are response-scale estimates with 95% Wald intervals from `confint(fit_animal, parm = \"variance_components\")`.", fig.alt = "Horizontal interval display comparing residual sigma and the animal-model location standard deviation. Both rows have hollow point estimates and finite 95 percent Wald intervals."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  animal_sd <- confint(fit_animal, parm = "variance_components")
  animal_sd$estimate <- c(
    unname(exp(coef(fit_animal, "sigma")["(Intercept)"])),
    unname(fit_animal$sdpars$mu["animal(1 | individual)"])
  )
  animal_sd$label <- c("Residual\nsigma", "Animal\nlocation SD")

  ggplot2::ggplot(animal_sd, ggplot2::aes(y = label)) +
    ggplot2::geom_vline(
      xintercept = 0,
      linewidth = 0.45,
      linetype = "dashed",
      colour = "grey55"
    ) +
    ggplot2::geom_errorbar(
      ggplot2::aes(xmin = lower, xmax = upper),
      width = 0,
      linewidth = 1.1,
      colour = "#0072B2"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = estimate),
      shape = 21,
      size = 3.5,
      stroke = 1,
      fill = "white",
      colour = "#0072B2"
    ) +
    ggplot2::scale_x_continuous(
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    animal_eye_theme() +
    ggplot2::labs(
      title = "Animal-model SD is separate from residual sigma",
      subtitle = "Bars are 95% Wald intervals on the response scale",
      x = "Fitted standard deviation",
      y = NULL
    )
}
```

For two response means, the animal-model correlation row is a fitted latent
correlation among individual additive deviations. It is not the residual
`rho12` correlation and it is not a raw sample correlation among observations.

```{r animal-q2-correlation-fit}
animal_q2_example <- simulate_animal_q2_guide_data()
animal_q2_dat <- animal_q2_example$data
Ainv <- animal_q2_example$Ainv

fit_animal_q2_example <- drmTMB(
  bf(
    mu1 = body_size ~ climate +
      animal(1 | p | individual, Ainv = Ainv),
    mu2 = activity ~ climate +
      animal(1 | p | individual, Ainv = Ainv),
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = animal_q2_dat
)

animal_q2_pairs <- corpairs(
  fit_animal_q2_example,
  level = "animal",
  conf.int = TRUE,
  ystep = 0.5,
  ytol = 2
)
animal_q2_pairs
```

```{r animal-q2-confidence-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "Animal-model q=2 location-location correlation from `corpairs()`; the Confidence Eye is a 95% profile interval on the latent animal-correlation scale and the dotted vertical line marks zero correlation. This animal q=2 interval is recovery-grade, not coverage-checked.", fig.alt = "Single-row Confidence Eye for the animal-model mean-mean correlation. The eye spans the 95 percent profile interval and the hollow point estimate is to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  animal_q2_display <- animal_q2_pairs
  animal_q2_display$display_label <- "Animal\nmu1-mu2"

  plot_corpairs(
    animal_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("animal" = "#0072B2")) +
    ggplot2::scale_fill_manual(values = c("animal" = "#0072B2")) +
    animal_eye_theme() +
    ggplot2::labs(
      title = "Animal-model latent mean correlation",
      subtitle = "Eye is a 95% profile interval; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

## Boundaries

The following animal-model routes remain planned:

- sparse large-pedigree precision construction;
- multiple animal-model slopes such as
  `animal(1 + x + z | individual, pedigree = pedigree)`;
- animal-model intercept-slope correlations;
- residual-scale structured slopes;
- predictor-dependent animal `corpair()` regressions;
- generic direct-SD grammar for animal-model standard deviations;
- richer non-Gaussian animal structured effects: labelled covariance blocks,
  multiple structured slopes, structured scale routes beyond the NB2 one-slope
  gate, and q=2 or q=4 non-Gaussian animal covariance blocks. A single q=1
  animal term already fits today for `poisson()`, `nbinom2()`, and `beta()`
  (Poisson and NB2 `mu` intercept plus one-slope, NB2 `sigma` one-slope, and a
  `beta()` animal intercept in `mu` or `sigma`) at recovery grade -- trust the
  point estimate, not the interval.

Use the [structural-dependence overview](structural-dependence.html) when you
are choosing among `animal()`, `phylo()`, `spatial()`, and `relmat()`. Use the
detailed [structural-dependence tutorial](phylogenetic-spatial.html) when you
need the current fitted examples, equations, and broader parity ladder.
