---
title: "Two-tree phylogenetic interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-tree phylogenetic interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

A tale of two phylogenies starts with a row in a paired data set. One partner
comes from one tree, the other partner comes from another tree, and the response
belongs to the pair. In bipartite ecological network data this could be a plant
and a pollinator, a host and a parasite, or any two partner clades where related
partners may have similar pair-level deviations.

Use `phylo_interaction()` when the pair itself should carry a phylogenetic
random effect. The fitted first slice is:

```text
eta_i = X_i beta + z[a_i, b_i]
vec(z) ~ Normal(0, sd_pair^2 (A_partner2 kron A_partner1))
```

Here `a_i` is the first partner level, `b_i` is the second partner level,
`A_partner1` and `A_partner2` are the two phylogenetic covariance matrices, and
`sd_pair` is the standard deviation of the pair-level structured field. For
Poisson and NB2 models, `eta_i` is the log mean. For Gaussian models, `eta_i` is
the location mean.

## What is fitted today

| Question | Syntax | Status |
| --- | --- | --- |
| Do pairs from two partner clades share phylogenetically structured deviations? | `phylo_interaction(1 | plant:pollinator, tree1 = plant_tree, tree2 = pollinator_tree)` | Fitted first q=1 `mu` slice for univariate Gaussian, ordinary Poisson, and ordinary NB2 models. |
| Do pair deviations need to be independent rather than phylogenetic? | precompute `pair_id <- interaction(plant, pollinator)` and use `(1 | pair_id)` | Fitted through ordinary grouped random effects. The direct grammar `(1 | plant:pollinator)` remains future parser sugar. |
| Do you already have a labelled Kronecker precision matrix? | `relmat(1 | plant_pollinator, Q = Q_pair)` | Low-level escape hatch for power users who can build and align the pair precision themselves. |
| Do you need partner main phylogenies plus the pair interaction in one model? | `phylo(1 | plant, ...) + phylo(1 | pollinator, ...) + phylo_interaction(...)` | Planned. Simultaneous structured layers need the next identifiability and reporting checks. |
| Is the response binary incidence? | Bernoulli/binomial pair model | Planned. Current fitted pair route covers Gaussian, Poisson, and NB2 `mu`, not binary incidence. |

## A small count example

The example below uses two small trees and a count response. The same marker can
also enter a Gaussian location model or an ordinary NB2 mean model.

```{r simulate-network}
balanced_partner_tree <- function(n_tip = 4L, prefix = "sp") {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0(prefix, "_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}

set.seed(20260531)
plant_tree <- balanced_partner_tree(4, "plant")
pollinator_tree <- balanced_partner_tree(4, "poll")

pair_grid <- expand.grid(
  plant = plant_tree$tip.label,
  pollinator = pollinator_tree$tip.label,
  KEEP.OUT.ATTRS = FALSE
)

plant_clade <- ifelse(pair_grid$plant %in% plant_tree$tip.label[1:2], -1, 1)
pollinator_clade <- ifelse(
  pair_grid$pollinator %in% pollinator_tree$tip.label[1:2],
  -1,
  1
)
pair_signal <- 0.45 * plant_clade * pollinator_clade
names(pair_signal) <- paste(pair_grid$plant, pair_grid$pollinator, sep = ":")

network_dat <- pair_grid[rep(seq_len(nrow(pair_grid)), each = 5), ]
network_dat$floral_density <- rnorm(nrow(network_dat))
eta <- 0.6 +
  0.25 * network_dat$floral_density +
  pair_signal[paste(network_dat$plant, network_dat$pollinator, sep = ":")]
network_dat$visits <- rpois(nrow(network_dat), lambda = exp(eta))
head(network_dat)
```

Fit the pair-level phylogenetic interaction with one marker:

```{r fit-pair-count}
fit_pair <- drmTMB(
  bf(
    visits ~ floral_density +
      phylo_interaction(
        1 | plant:pollinator,
        tree1 = plant_tree,
        tree2 = pollinator_tree
      )
  ),
  family = poisson(link = "log"),
  data = network_dat
)
fit_pair$sdpars$mu
```

The fitted standard deviation is the scale of the latent pair deviations on the
linear predictor scale. For this Poisson model, that is the log-count scale.
Use `ranef()` when you want conditional pair deviations, and use
`profile_targets()` when you want to see whether the pair SD is registered for
profile-likelihood diagnostics.

```{r inspect-pair-count}
names(ranef(fit_pair))
profile_targets(fit_pair)
```

## Independent pairs are a different question

If the biological question is "do repeated observations of the same pair share
an unexplained deviation?" and no phylogenetic structure is needed, use an
ordinary grouped random effect. Precompute the pair identifier first:

```{r independent-pairs, eval = FALSE}
network_dat$pair_id <- interaction(
  network_dat$plant,
  network_dat$pollinator,
  drop = TRUE
)

fit_independent_pair <- drmTMB(
  visits ~ floral_density + (1 | pair_id),
  family = poisson(link = "log"),
  data = network_dat
)
```

Do not write `(1 | plant:pollinator)` yet. The ordinary random-effect parser
currently expects a simple grouping column, so `pair_id` is the supported route.

## When to reach for `relmat()`

`relmat()` is the low-level route when you already have the pair precision
matrix and have checked that its row names match the pair-level factor in the
data. If the factor is `plant_pollinator`, then the row names and column names of
`Q_pair` must match the levels of `plant_pollinator` exactly.

```{r relmat-pair-route, eval = FALSE}
fit_pair_relmat <- drmTMB(
  visits ~ floral_density +
    relmat(1 | plant_pollinator, Q = Q_pair),
  family = poisson(link = "log"),
  data = network_dat
)
```

For most two-tree pair models, `phylo_interaction()` is safer because it builds
the sparse Kronecker precision from `tree1` and `tree2`, preserves the two
partner names, and keeps the biological meaning visible in the formula.

## What this does not claim

This first slice is a pair-level structured random intercept. It is not yet the
full additive model with a main plant phylogeny, a main pollinator phylogeny,
and a pair interaction:

```{r additive-future, eval = FALSE}
visits ~ floral_density +
  phylo(1 | plant, tree = plant_tree) +
  phylo(1 | pollinator, tree = pollinator_tree) +
  phylo_interaction(
    1 | plant:pollinator,
    tree1 = plant_tree,
    tree2 = pollinator_tree
  )
```

It also does not fit reciprocal evolutionary dynamics, community-only matching
without pair observations, binary incidence, structured pair slopes, or
simultaneous spatial or animal-model interactions. Those are different
statistical questions and need separate model designs.
