---
title: "Bivariate non-Gaussian models: choose a joint or staged association"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bivariate non-Gaussian models: choose a joint or staged association}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

With two non-Gaussian outcomes, first decide what kind of association answers
the scientific question. A direct joint likelihood estimates both outcomes and
their residual association together. A frozen-margin association first fits the
two outcomes separately and then estimates a latent-normal association while
holding the fitted margins fixed. These are different estimands.

This article helps you choose a route for two outcomes measured on the same
rows. The non-Gaussian and mixed-outcome routes are beta development interfaces:
they do not provide interval, coverage, random-effect, or generic-pair claims.
For a released direct Gaussian joint model, use `biv_gaussian()` and
[Changing residual coupling with rho12](bivariate-coscale.html).

## Start by choosing the association scale

| Your paired outcomes | Route | What the association means |
| --- | --- | --- |
| Two Gaussian traits | Direct `biv_gaussian()` model | Gaussian residual `rho12` |
| Two positive traits plausibly lognormal | Post-0.6 direct `biv_lognormal()` model | Log-response residual `rho12` |
| Two heavy-tailed real-valued traits | Post-0.6 direct `biv_student()` model | Shared-Student-t residual/scatter `rho12` |
| A reviewed mixed or discrete pair | `biv_associate()` staged route | Latent-normal copula `eta` after frozen margins; one numeric association slope is beta-supported for Bernoulli x NB2 |

Neither `rho12` nor `eta` is automatically the Pearson correlation of the two
raw response columns. The scale is part of the scientific interpretation, not
just a computational detail.

## Direct joint model: two positive traits

For two positive lognormal outcomes, the model is joint on the log-response
scale:

\[
(\log Y_{1i}, \log Y_{2i}) \sim N_2\left(
(\mu_{1i}, \mu_{2i}),
\begin{bmatrix}
\sigma_1^2 & \rho_{12}\sigma_1\sigma_2 \\
\rho_{12}\sigma_1\sigma_2 & \sigma_2^2
\end{bmatrix}
\right).
\]

Here `rho12` is the correlation of the two log-response residuals. A positive
value means that plants above their fitted log seed-mass expectation also tend
to be above their fitted log root-mass expectation. It is not automatically the
correlation of seed mass and root mass on their original scale.

```{r lognormal-example, eval = FALSE}
fit_log <- drmTMB(
  bf(mu1 = seed_mass ~ treatment, mu2 = root_mass ~ treatment,
     sigma1 = ~ 1, sigma2 = ~ 1, rho12 = ~ 1),
  family = biv_lognormal(), data = plants
)

rho12(fit_log)
fitted(fit_log)
```

`fitted(fit_log)` returns arithmetic marginal means on the original response
scale. The `mu1` and `mu2` predictors remain log-scale locations. These direct
models estimate the two margins and `rho12` together; they are not staged
frozen-margin fits.

## Direct joint model: two heavy-tailed traits

For two real-valued traits with occasional extreme observations, the bivariate
Student-t model uses a shared degrees-of-freedom parameter, `nu`:

```{r student-example, eval = FALSE}
fit_t <- drmTMB(
  bf(mu1 = activity ~ habitat, mu2 = boldness ~ habitat,
     sigma1 = ~ 1, sigma2 = ~ 1, nu = ~ 1, rho12 = ~ 1),
  family = biv_student(), data = behaviour
)

rho12(fit_t)
```

`sigma1` and `sigma2` are Student-t scales, not necessarily marginal standard
deviations. The model has one shared `nu`, which controls tail heaviness. At a
finite `nu`, `rho12 = 0` means zero residual correlation, but not necessarily
complete independence because the two outcomes share the same row-level
heavy-tail mechanism.

## Staged association: different kinds of outcome

If one response is binary and the other is a count, there is no shared direct
Gaussian residual scale. The frozen-margin route instead fits a binary
probability model and a count model, then estimates latent-normal `eta` from
their fitted distributions.

```{r staged-example, eval = FALSE}
paired_data <- na.omit(data.frame(
  bred, offspring, habitat, habitat_score, season
))

assoc <- biv_associate(
  bf(mu = bred ~ habitat),
  bf(mu = offspring ~ habitat, sigma = ~ season),
  family = list(binomial(), nbinom2()),
  data = paired_data,
  association = ~ 1
)

association(assoc)
```

This is one convenient R call, but not one jointly fitted model.
`biv_associate()` fits the two margins, freezes them, and estimates only `eta`.
The reported value is a latent-normal copula correlation conditional on those
fitted margins. It is not `rho12`, a logit coefficient, an odds ratio, or an
observed-scale correlation.

For a binary response, the model observes whether a hidden tendency crossed a
threshold. It does not observe an ordinary numeric residual. Read
[Frozen-margin association for mixed outcome pairs](cross-family.html) for the
binary-threshold explanation, the reviewed pair classes, and the rule for
withheld estimates.

For the literal-Bernoulli x ordinary-NB2 beta route, `association = ~ x` can
also model one numeric association gradient. It estimates a row-specific
latent-normal `eta_i`, after both margins have been fitted and frozen. See
[Association between mixed outcome pairs](cross-family.html) for the exact
syntax, output, and limits.

## What is currently outside these routes

The direct same-family first slices keep `sigma1`, `sigma2`, `rho12`, and `nu`
constant across rows. The staged route usually keeps `eta` constant; its one
beta exception is a single numeric association slope for literal-Bernoulli x
ordinary-NB2. Its reviewed fixed-effect margins may include their explicitly
supported covariates before they are frozen. Neither route currently adds random
effects, general association-predictor grammar, missing-data support, two-stage standard errors, profiles,
confidence intervals, coverage, or a general non-Gaussian bivariate claim.
Choose the route whose stated association scale matches the biological question,
and treat every boundary as part of the model definition.
