---
title: "Changing residual coupling with rho12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing residual coupling with rho12}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

This tutorial introduces bivariate location-scale regression with
predictor-dependent residual correlation. The scientific idea is that two
responses can change in mean, change in residual variance, and change in
residual coupling.

Current fixed-effect syntax uses one formula per distributional parameter:

```r
drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The central modelling question is whether predictors change residual coupling
between two responses, not only their means or variances.

The key parameter is `rho12`. A location-scale model asks whether predictors
change residual SDs. A location-coscale model asks whether predictors also
change the residual covariance structure between two responses; in the current
bivariate Gaussian model, that structure is the residual correlation `rho12`.
In ecology and evolution, this can ask whether the residual association between
traits such as body mass and litter size differs among terrestrial, aquatic,
and aerial lifestyles after modelling response-specific means and dispersion.

## Equations for the implemented model

For observation `i`, the implemented fixed-effect bivariate Gaussian model is:

\[
\begin{bmatrix}
y_{1i} \\
y_{2i}
\end{bmatrix}
\mid
\mu_{1i}, \mu_{2i}, \sigma_{1i}, \sigma_{2i}, \rho_{12i}
\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{1i} \\
\mu_{2i}
\end{bmatrix},
\Omega_i
\right).
\]

The five distributional predictors are:

\[
\begin{aligned}
\mu_{1i} &= X_{\mu 1,i}\beta_{\mu 1}, \\
\mu_{2i} &= X_{\mu 2,i}\beta_{\mu 2}, \\
\log(\sigma_{1i}) &= X_{\sigma 1,i}\beta_{\sigma 1}, \\
\log(\sigma_{2i}) &= X_{\sigma 2,i}\beta_{\sigma 2}, \\
\eta_{\rho12,i} &= X_{\rho12,i}\beta_{\rho12}, \\
\rho_{12i} &= \tanh(\eta_{\rho12,i}).
\end{aligned}
\]

The residual covariance matrix is:

\[
\Omega_i =
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}.
\]

Component meanings:

| Component | Meaning |
|---|---|
| `mu1`, `mu2` | expected values for responses 1 and 2 |
| `sigma1`, `sigma2` | residual SDs for responses 1 and 2 |
| `eta_rho12` | unconstrained linear predictor for residual correlation |
| `rho12` | residual response-response correlation within observation `i` |
| `Omega_i` | residual covariance matrix after means and residual SDs are modelled |

For implementation cross-checking, the same model can also be written in a
compact code-like form:

```text
mu1_i = X_mu1[i, ] beta_mu1
mu2_i = X_mu2[i, ] beta_mu2
log(sigma1_i) = X_sigma1[i, ] beta_sigma1
log(sigma2_i) = X_sigma2[i, ] beta_sigma2
eta_rho12_i = X_rho12[i, ] beta_rho12
rho12_i = tanh(eta_rho12_i)
Omega_i[1, 1] = sigma1_i^2
Omega_i[2, 2] = sigma2_i^2
Omega_i[1, 2] = Omega_i[2, 1] = rho12_i * sigma1_i * sigma2_i
```

Implementation note: `drmTMB` internally evaluates residual correlations as
`0.99999999 * tanh(eta_rho12_i)`. The tiny multiplier keeps covariance
matrices strictly positive definite near the boundaries. It is not a
biological scaling factor, and it should not change how users interpret the
model.

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The mapping is:

| Symbol | Meaning | R syntax source |
|---|---|---|
| `y1_i`, `y2_i` | observed responses | left-hand sides of `mu1` and `mu2` formulas |
| `X_mu1`, `X_mu2` | response-specific mean design matrices | right-hand sides of `mu1` and `mu2` |
| `mu1_i`, `mu2_i` | fitted response-specific means | columns of `fitted(fit)` |
| `X_sigma1`, `X_sigma2` | response-specific residual scale design matrices | right-hand sides of `sigma1` and `sigma2` |
| `X_rho12` | residual correlation design matrix | right-hand side of `rho12` |
| `sigma1_i`, `sigma2_i` | residual SDs for responses 1 and 2 | `predict(fit, dpar = "sigma1")`, `predict(fit, dpar = "sigma2")` |
| `rho12_i` | residual response-response correlation | `rho12(fit)` |

The fixed effects may be the same, overlapping, or different across
parameters. For example, `x1` can appear in all five formulae, while `x2` might
appear only in `mu1`, `sigma1`, and `rho12`. Each parameter still gets its own
coefficient vector.

Read those coefficient vectors as separate biological statements:

| Formula | Example coefficient | Interpretation |
|---|---|---|
| `mu1 = y1 ~ x` | `mu1:x` | predictor effect on the expected value of response 1 |
| `mu2 = y2 ~ x` | `mu2:x` | predictor effect on the expected value of response 2 |
| `sigma1 = ~ x` | `sigma1:x` | multiplicative change in residual SD for response 1 |
| `sigma2 = ~ x` | `sigma2:x` | multiplicative change in residual SD for response 2 |
| `rho12 = ~ x` | `rho12:x` | change in residual response-response coupling after both means and residual SDs are modelled |

The `rho12` slope is not a slope for either response mean. It is a slope on the
correlation-link scale. Use `rho12(fit, newdata = grid)` or a `rho12` curve
when the reader needs the fitted residual correlation on the response scale.

If both responses have the same location predictors, `mvbind()` is a shorthand
for the two location formulas:

```r
drmTMB(
  bf(
    mvbind(y1, y2) ~ x1 + x2,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

This expands internally to `mu1 = y1 ~ x1 + x2` and
`mu2 = y2 ~ x1 + x2`. Use explicit `mu1` and `mu2` formulas whenever the
responses need different location predictors.

## Worked example: behaviour coupling and disturbance

Suppose an ecologist measures two behaviours, activity and boldness, and wants
to know whether an environmental gradient changes their residual coupling after
accounting for response-specific means and residual SDs. The code below uses
simulated data, but the model structure matches that scientific question.

For this example, the fitted biological model is:

\[
\begin{aligned}
\begin{bmatrix}
\text{activity}_i\\
\text{boldness}_i
\end{bmatrix}
&\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{\text{activity},i}\\
\mu_{\text{boldness},i}
\end{bmatrix},
\Omega_i
\right),\\
\mu_{\text{activity},i}
  &= \beta_{a0} + \beta_{a1}\text{food}_i +
     \beta_{a2}\text{temperature}_i,\\
\mu_{\text{boldness},i}
  &= \beta_{b0} + \beta_{b1}\text{food}_i,\\
\log(\sigma_{\text{activity},i})
  &= \gamma_{a0} + \gamma_{a1}\text{food}_i +
     \gamma_{a2}\text{temperature}_i,\\
\log(\sigma_{\text{boldness},i})
  &= \gamma_{b0} + \gamma_{b1}\text{food}_i,\\
\eta_{\rho12,i} &= \delta_0 + \delta_1\text{disturbance}_i,\\
\rho_{12,i} &= \tanh(\eta_{\rho12,i}).
\end{aligned}
\]

Here `rho12` is not the ordinary correlation between raw activity and raw
boldness. It is the residual correlation after the model has already adjusted
for the mean and residual SD predictors in each response.

```{r bivariate-coscale-simulate}
set.seed(1)
n <- 180
dat <- data.frame(
  food = rnorm(n),
  temperature = rnorm(n),
  disturbance = rnorm(n)
)
mu1 <- 0.2 + 0.5 * dat$food + 0.2 * dat$temperature
mu2 <- -0.1 + 0.4 * dat$food
sigma1 <- exp(-0.2 + 0.2 * dat$food - 0.1 * dat$temperature)
sigma2 <- exp(0.1 - 0.2 * dat$food)
eta_rho12 <- -0.1 + 0.4 * dat$disturbance
rho12 <- tanh(eta_rho12)
e1 <- rnorm(n)
e2 <- rho12 * e1 + sqrt(1 - rho12^2) * rnorm(n)
dat$activity <- mu1 + sigma1 * e1
dat$boldness <- mu2 + sigma2 * e2
```

The model lets `food` and `temperature` explain mean activity, lets `food`
explain mean boldness, lets `food` and `temperature` explain residual SD in
activity, lets `food` explain residual SD in boldness, and lets `disturbance`
explain residual correlation:

```{r bivariate-coscale-fit}
fit_biv <- drmTMB(
  drm_formula(
    mu1 = activity ~ food + temperature,
    mu2 = boldness ~ food,
    sigma1 = ~ food + temperature,
    sigma2 = ~ food,
    rho12 = ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Check the fit and inspect the coefficient table:

```{r bivariate-coscale-output}
check_drm(fit_biv)
summary(fit_biv)
```

How to read this output:

1. Rows beginning with `mu1:` describe expected activity, and rows beginning
   with `mu2:` describe expected boldness.
2. Rows beginning with `sigma1:` and `sigma2:` describe log residual SDs for
   activity and boldness.
3. Rows beginning with `rho12:` describe the residual activity-boldness
   correlation on the linear-predictor scale. Use `rho12(fit_biv)` or
   `rho12(fit_biv, newdata = ...)` to read that effect as a correlation.

The `rho12` coefficients are on the linear-predictor scale:

```{r bivariate-coscale-rho12}
coef(fit_biv, "rho12")
head(rho12(fit_biv))
```

`corpairs()` gives the same residual correlation in the long correlation-pair
format. The same table shape now also holds fitted ordinary group-level and
phylogenetic rows; richer spatial and study-level correlation rows remain
planned:

```{r bivariate-coscale-corpairs}
corpairs(fit_biv)[
  c("level", "from_response", "to_response", "class", "parameter",
    "estimate", "min", "max", "modelled")
]
```

The `rho12` coefficients are fitted on the linear-predictor scale
`eta_rho12`; the response transform is
`rho12 = tanh(eta_rho12)` with the tiny internal guard described above. Use
`rho12(fit)` to read them as residual correlations, and use `corpairs(fit)` to
place residual `rho12` in a table that is ready to hold other correlation
levels later. In this example, a positive coefficient for `disturbance` means
that observations with larger disturbance have stronger positive residual
coupling between activity and boldness, after the model has already accounted
for the predictors in `mu1`, `mu2`, `sigma1`, and `sigma2`.

```{r bivariate-coscale-interpretation-table}
newdat <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = c(-1, 0, 1)
)
rho_table <- data.frame(
  disturbance = newdat$disturbance,
  sigma_activity = predict(fit_biv, newdata = newdat, dpar = "sigma1"),
  sigma_boldness = predict(fit_biv, newdata = newdat, dpar = "sigma2"),
  rho12 = rho12(fit_biv, newdata = newdat)
)
rho_table$residual_covariance <- with(
  rho_table,
  rho12 * sigma_activity * sigma_boldness
)
rho_table$residual_variance_activity <- rho_table$sigma_activity^2
rho_table$residual_variance_boldness <- rho_table$sigma_boldness^2
round(rho_table, 3)
```

This table gives three quantities a reader can report directly. `sigma_activity`
and `sigma_boldness` are the fitted residual SDs for each response at the
chosen covariate values. `rho12` is the fitted residual correlation between the
two responses. `residual_covariance` is the off-diagonal element of
`\Omega_i`, computed as `rho12 * sigma_activity * sigma_boldness`. The two
residual variance columns are the diagonal elements of `\Omega_i`.

## What should I report?

For a location-coscale model, a useful result sentence usually needs all three
distributional pieces: means, residual SDs, and residual coupling. A minimal
table for the coscale part is:

```{r bivariate-coscale-report-table}
report_grid <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = c(-1.5, 0, 1.5)
)
report_table <- data.frame(
  disturbance = report_grid$disturbance,
  eta_rho12 = predict(
    fit_biv,
    newdata = report_grid,
    dpar = "rho12",
    type = "link"
  ),
  rho12 = rho12(fit_biv, newdata = report_grid),
  sigma_activity = predict(fit_biv, newdata = report_grid, dpar = "sigma1"),
  sigma_boldness = predict(fit_biv, newdata = report_grid, dpar = "sigma2")
)
report_table$residual_covariance <- with(
  report_table,
  rho12 * sigma_activity * sigma_boldness
)
report_table$residual_variance_activity <- report_table$sigma_activity^2
report_table$residual_variance_boldness <- report_table$sigma_boldness^2
round(report_table, 3)
```

The raw correlation between activity and boldness is not the same estimand as
`rho12`. The raw correlation mixes mean structure, residual SDs, and residual
coupling. The fitted `rho12` is the residual correlation after the model has
accounted for the formulae in `mu1`, `mu2`, `sigma1`, and `sigma2`:

```{r bivariate-coscale-raw-vs-residual}
round(data.frame(
  raw_activity_boldness_correlation = stats::cor(dat$activity, dat$boldness),
  mean_fitted_residual_rho12 = mean(rho12(fit_biv))
), 3)
```

When reporting this model, avoid saying only "activity and boldness were
correlated". A better sentence is: "After modelling response-specific means
and residual SDs, the fitted residual activity-boldness correlation increased
with disturbance." That sentence tells the reader which correlation has been
modelled.

The same response-scale result can be visualised as a `rho12` curve. Here the
ribbon is a 95% Wald interval from `predict_parameters()`; if an interval
column is unavailable, the plot should not silently imply uncertainty.

```{r bivariate-coscale-rho12-curve, fig.width = 6, fig.height = 4, fig.cap = "Fitted residual `rho12` over disturbance after modelling response-specific means and residual SDs; the ribbon is a 95% Wald confidence interval and the dotted line marks zero residual correlation.", fig.alt = "Line plot of fitted residual rho12 correlation over disturbance with a pale 95 percent Wald confidence ribbon and a dotted zero line, increasing from negative to positive values and crossing zero near average disturbance."}
rho_grid <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = seq(-2, 2, length.out = 80)
)
rho_pred <- predict_parameters(
  fit_biv,
  newdata = rho_grid,
  dpar = "rho12",
  conf.int = TRUE
)
rho_grid$rho12 <- rho_pred$estimate
rho_grid$conf.low <- rho_pred$conf.low
rho_grid$conf.high <- rho_pred$conf.high

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(rho_grid, ggplot2::aes(disturbance, rho12)) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
    ggplot2::geom_ribbon(
      ggplot2::aes(ymin = conf.low, ymax = conf.high),
      fill = "#006D77",
      alpha = 0.18,
      colour = NA
    ) +
    ggplot2::geom_line(linewidth = 0.9, colour = "#006D77") +
    ggplot2::coord_cartesian(ylim = c(-1, 1)) +
    ggplot2::labs(
      x = "Disturbance",
      y = "Fitted residual correlation (rho12)",
      title = "Residual coupling changes with disturbance",
      subtitle = "Ribbon is a 95% Wald interval; dotted line marks rho12 = 0"
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold"),
      plot.subtitle = ggplot2::element_text(colour = "grey30"),
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA)
    )
} else {
  old_par <- graphics::par(bg = "white")
  on.exit(graphics::par(old_par), add = TRUE)
  plot(
    rho12 ~ disturbance,
    data = rho_grid,
    type = "l",
    lwd = 2,
    ylim = c(-1, 1),
    xlab = "Disturbance",
    ylab = "Fitted residual correlation (rho12)"
  )
  abline(h = 0, lty = 3, col = "grey60")
}
```

In this example, a positive `disturbance` coefficient for `rho12` means that
activity and boldness become more tightly positively coupled at higher
disturbance, after their means and residual SDs have been modelled. It does
not say that disturbance changes among-individual personality or plasticity
correlations; those require group-level covariance blocks and should appear as
separate rows in `corpairs(fit)`.

A concise reporting sentence would be: after accounting for food,
temperature, and response-specific residual SDs, the fitted residual
correlation between activity and boldness increased along the disturbance
gradient. In biological terms, individuals that were more active than expected
were also more bold than expected under higher disturbance.

## Where residual rho12 stops

The fitted model above has no repeated-individual structure. It estimates
residual coupling: whether two measurements from the same observation are
closer together or farther apart than expected after modelling their means and
residual SDs. That is useful, but it is not the same question as whether some
individuals have consistently high activity and consistently high boldness.

For ecological and evolutionary examples, residual `rho12` can represent
changing residual coupling between traits such as activity and boldness, body
size and fecundity, or leaf area and seed number. A group-level covariance
block asks a different question: do groups, individuals, species, or sites
share latent deviations across fitted components?

## Worked example: individual differences

Suppose the same ecologist repeatedly measures each individual and now asks an
individual-difference question. After accounting for food and disturbance, do
individuals with higher average activity also tend to have higher average
boldness?

The first fitted group-level version adds matching labelled random intercepts
in the mean structure. The shared `p` label creates one covariance block for
the `mu1` and `mu2` random intercepts:

```r
drmTMB(
  formula = bf(
    mu1 = activity ~ food + disturbance + (1 | p | ID),
    mu2 = boldness ~ food              + (1 | p | ID),
    sigma1 = ~1,
    sigma2 = ~1,
    rho12 = ~1,
    corpair(ID, level = "group", block = "p", from = "mu1", to = "mu2") ~ 1
  ),
  family = biv_gaussian(),
  data = dat_group
)
```

Here `sigma1` and `sigma2` are residual scale parameters. The fitted
random-intercept SDs are group-level standard deviations in `sdpars$mu`, not
residual `sigma`.

Symbolically, the implemented random-intercept mean part is:

\[
\begin{aligned}
\mu_{1ij}
  &= X_{\mu 1}[ij, ]\beta_{\mu 1} + b_{0,1j},\\
\mu_{2ij}
  &= X_{\mu 2}[ij, ]\beta_{\mu 2} + b_{0,2j},\\
\begin{bmatrix}
b_{0,1j}\\
b_{0,2j}
\end{bmatrix}
  &\sim \operatorname{MVN}(\mathbf{0}, \Sigma_{\mu,ID}).
\end{aligned}
\]

The covariance matrix `Sigma_mu_ID` contains two group-level standard
deviations and one group-level correlation. That correlation answers whether
individuals with higher average activity also tend to have higher average
boldness. It is not `rho12_i`. The parameter `rho12_i` remains the
observation-level residual correlation in `Omega_i`.

The small simulation below creates repeated observations from that model:

```{r bivariate-coscale-group-simulate}
set.seed(98)
n_ID <- 55
n_each <- 7
ID <- factor(rep(seq_len(n_ID), each = n_each))
id_index <- as.integer(ID)
n <- length(ID)

food <- rnorm(n)
disturbance <- rnorm(n)

sigma_activity <- 0.35
sigma_boldness <- 0.45
sd_activity_ID <- 0.65
sd_boldness_ID <- 0.65
rho_ID <- 0.55
rho_residual <- 0.15

u_activity <- rnorm(n_ID)
u_boldness <- rho_ID * u_activity + sqrt(1 - rho_ID^2) * rnorm(n_ID)

e_activity <- rnorm(n)
e_boldness <- rho_residual * e_activity +
  sqrt(1 - rho_residual^2) * rnorm(n)

dat_group <- data.frame(ID = ID, food = food, disturbance = disturbance)
dat_group$activity <- 0.25 + 0.45 * food + 0.15 * disturbance +
  sd_activity_ID * u_activity[id_index] + sigma_activity * e_activity
dat_group$boldness <- -0.15 + 0.35 * food +
  sd_boldness_ID * u_boldness[id_index] + sigma_boldness * e_boldness
```

Fit the grouped model with matching `(1 | p | ID)` terms in the two location
formulas:

```{r bivariate-coscale-group-fit}
fit_group <- drmTMB(
  bf(
    mu1 = activity ~ food + disturbance + (1 | p | ID),
    mu2 = boldness ~ food + (1 | p | ID),
    sigma1 = ~1,
    sigma2 = ~1,
    rho12 = ~1,
    corpair(ID, level = "group", block = "p", from = "mu1", to = "mu2") ~ 1
  ),
  family = biv_gaussian(),
  data = dat_group,
  control = list(eval.max = 500, iter.max = 500)
)
```

In a real analysis, inspect every row of `check_drm(fit_group)`. The filtered
rows below show the convergence and covariance checks most directly tied to
this individual-difference question:

```{r bivariate-coscale-group-check}
group_checks <- check_drm(fit_group)
group_checks[
  group_checks$check %in% c(
    "optimizer_convergence",
    "random_effect_sd_boundary",
    "rho12_boundary",
    "biv_mu_random_effect_covariance"
  ),
  c("check", "status", "value", "message")
]
```

## Reading group-level covariance

`corpairs()` keeps residual and group-level correlations in the same long table
without giving them the same meaning. Because the group correlation is modelled
with an intercept-only `corpair()` formula, the response-scale interval needs a
one-row `newdata` target:

```{r bivariate-coscale-group-corpairs}
group_cor_dpar <- 'corpair(ID, level = "group", block = "p", from = "mu1", to = "mu2")'
group_cor_newdata <- data.frame(row = 1)
row.names(group_cor_newdata) <- "overall"

pair_table <- corpairs(
  fit_group,
  conf.int = TRUE,
  ystep = 0.5,
  ytol = 2
)
group_cor_ci <- confint(
  fit_group,
  parm = group_cor_dpar,
  method = "profile",
  newdata = group_cor_newdata,
  trace = FALSE,
  ystep = 0.5,
  ytol = 2
)
group_pair_row <- pair_table$level == "group" &
  pair_table$from_dpar == "mu1" &
  pair_table$to_dpar == "mu2"
pair_table$profile_target[group_pair_row] <- group_cor_ci$parm[[1]]
pair_table$conf.low[group_pair_row] <- group_cor_ci$lower[[1]]
pair_table$conf.high[group_pair_row] <- group_cor_ci$upper[[1]]
pair_table$conf.method[group_pair_row] <- group_cor_ci$method[[1]]
pair_table$conf.status[group_pair_row] <- group_cor_ci$conf.status[[1]]
pair_table$interval_source[group_pair_row] <- group_cor_ci$method[[1]]
pair_table[
  ,
  c(
    "level", "group", "block", "from_dpar", "to_dpar", "class",
    "parameter", "estimate", "conf.low", "conf.high", "conf.status",
    "interval_source", "modelled"
  )
]
```

The optional plotting helper consumes that explicit table. Faceting by `level`
keeps the residual `rho12` row visually separate from the group-level
random-intercept correlation:

```{r bivariate-coscale-group-corpairs-plot, fig.width = 6.6, fig.height = 3.4, fig.cap = "`corpairs()` Confidence Eye display separating residual `rho12` from the individual-level mean-mean random-intercept correlation; eyes are 95% profile intervals on the response-correlation scale and the dotted line marks zero correlation.", fig.alt = "Confidence Eye plot comparing residual rho12 correlation with the individual-level mean-mean random-intercept correlation. Each row has a pale horizontal eye spanning a 95 percent profile interval, a hollow point estimate, and a dotted zero line on a correlation scale from minus one to one."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  pair_plot_table <- pair_table
  pair_plot_table$display_label <- ifelse(
    pair_plot_table$level == "residual",
    "Residual\nrho12",
    "Individual\nmean-mean"
  )
  plot_corpairs(pair_plot_table, label = "display_label", facet = NULL) +
    ggplot2::labs(
      title = "Residual and individual-level correlations are separate",
      subtitle = "Eyes are 95% profile intervals; dotted line marks zero correlation",
      colour = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold"),
      plot.subtitle = ggplot2::element_text(colour = "grey30"),
      legend.position = "none"
    )
}
```

The residual row is `rho12`: within-observation activity-boldness coupling.
The group row is the `mu1`/`mu2` random-intercept correlation: among-individual
covariation in average activity and average boldness.

`summary(fit_group)$covariance` gives the same group-level row with the
component SDs and covariance on the interpretation scale:

```{r bivariate-coscale-group-covariance-table}
group_covariance <- summary(fit_group)$covariance
report_group_covariance <- group_covariance[
  ,
  c(
    "level", "group", "block", "from_response", "to_response", "class",
    "correlation", "from_sd", "to_sd", "covariance", "from_scale",
    "to_scale", "covariance_conf.status"
  )
]
numeric_columns <- vapply(report_group_covariance, is.numeric, logical(1))
report_group_covariance[numeric_columns] <- lapply(
  report_group_covariance[numeric_columns],
  round,
  3
)
report_group_covariance
```

For this mean-mean block, `from_sd` is the among-individual SD in average
activity and `to_sd` is the among-individual SD in average boldness. The
`correlation` column is the individual-level activity-boldness correlation,
and `covariance` is `correlation * from_sd * to_sd`. The `identity` scale
columns tell you these quantities are on the response-location scale, not on a
log residual-SD scale.

For this fitted block, `check_drm()` reports whether group levels are
replicated and whether either group-level SD is tiny relative to the matching
residual scale. A note there is not a proof that the model is wrong; it tells
you to inspect whether the group-level SDs and correlation are supported by the
data before treating them as biological conclusions.

Use `profile_targets(fit_group)` before requesting profile intervals for this
block. The SDs are direct random-effect targets, while the modelled
`corpair()` row is a fixed-effect target on the correlation-link scale:

```{r bivariate-coscale-group-targets}
group_targets <- profile_targets(fit_group)
group_target_names <- c(
  "sd:mu:mu1:(1 | p | ID)",
  "sd:mu:mu2:(1 | p | ID)",
  paste0("fixef:", group_cor_dpar, ":(Intercept)")
)
group_targets[
  match(group_target_names, group_targets$parm),
  c("parm", "profile_ready", "profile_note")
]
```

The group-level SD targets are named like `sd:mu:mu1:(1 | p | id)` and
`sd:mu:mu2:(1 | p | id)`. The modelled group-level correlation target begins
with `fixef:corpair(...)`, and `confint(..., newdata = ...)` returns the
response-scale correlation interval used in the plot. These are separate from
the residual-correlation target `rho12`. Returned profile rows include
`profile.boundary` and `profile.message`; a non-`"ok"` message is a prompt to
inspect `check_drm(fit)` and the profile path before treating the interval as a
stable biological estimate.

## Boundaries for this covariance layer

The middle `p` in `(1 | p | ID)` follows grouped covariance-block syntax for
correlated group-level effects. This syntax is distinct from residual `rho12`.
Use `rho12(fit)` for within-observation residual coupling and
`corpairs(fit, level = "group")` for fitted ordinary group-level covariance
rows.

This example teaches only the `mu1`/`mu2` random-intercept covariance row.
Other implemented ordinary bivariate random-intercept rows include
`sigma1`/`sigma2`, one same-response `mu`/`sigma` pair, and the all-four
`mu1`/`mu2`/`sigma1`/`sigma2` block. Read the all-four block as six rows: one
`mu1`-`mu2` row, four mean-scale rows (`mu1`-`sigma1`, `mu1`-`sigma2`,
`mu2`-`sigma1`, and `mu2`-`sigma2`), and one `sigma1`-`sigma2` row.

The first ordinary bivariate slope-slope route is now the matching
`(0 + x | p | id)` block in both `mu1` and `mu2`. Intercept-plus-slope
bivariate blocks, all-four p8/q8 location-scale slope endpoints, random
effects in `rho12`, bivariate `meta_known_V()` plus random effects,
mixed-response families, and ordinary spatial group-level covariance remain
future work unless a later status map says otherwise. For phylogenetic
covariance, use the structural-dependence tutorial and keep those rows
separate from both ordinary group-level covariance and residual `rho12`.
