---
title: "Changing residual coupling with rho12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing residual coupling with rho12}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

This tutorial introduces bivariate location-scale regression with
predictor-dependent residual correlation. The scientific idea is that two
responses can change in mean, change in residual variance, and change in
residual coupling.

Current fixed-effect syntax uses one formula per distributional parameter:

```r
drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The central modelling question is whether predictors change residual coupling
between two responses, not only their means or variances.

The key parameter is `rho12`. A location-scale model asks whether predictors
change residual SDs. A location-coscale model asks whether predictors also
change the residual covariance structure between two responses; in the current
bivariate Gaussian model, that structure is the residual correlation `rho12`.
In ecology and evolution, this can ask whether the residual association between
traits such as body mass and litter size differs among terrestrial, aquatic,
and aerial lifestyles after modelling response-specific means and dispersion.

## Equations for the implemented model

For observation `i`, the implemented fixed-effect bivariate Gaussian model is:

\[
\begin{bmatrix}
y_{1i} \\
y_{2i}
\end{bmatrix}
\mid
\mu_{1i}, \mu_{2i}, \sigma_{1i}, \sigma_{2i}, \rho_{12i}
\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{1i} \\
\mu_{2i}
\end{bmatrix},
\Omega_i
\right).
\]

The five distributional predictors are:

\[
\begin{aligned}
\mu_{1i} &= X_{\mu 1,i}\beta_{\mu 1}, \\
\mu_{2i} &= X_{\mu 2,i}\beta_{\mu 2}, \\
\log(\sigma_{1i}) &= X_{\sigma 1,i}\beta_{\sigma 1}, \\
\log(\sigma_{2i}) &= X_{\sigma 2,i}\beta_{\sigma 2}, \\
\eta_{\rho 12,i} &= X_{\rho 12,i}\beta_{\rho 12}, \\
\rho_{12i} &= \tanh(\eta_{\rho 12,i}).
\end{aligned}
\]

The residual covariance matrix is:

\[
\Omega_i =
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}.
\]

Component meanings:

| Component | Meaning |
|---|---|
| `mu1`, `mu2` | expected values for responses 1 and 2 |
| `sigma1`, `sigma2` | residual SDs for responses 1 and 2 |
| `eta_rho12` | unconstrained linear predictor for residual correlation |
| `rho12` | residual response-response correlation within observation `i` |
| `Omega_i` | residual covariance matrix after means and residual SDs are modelled |

For implementation cross-checking, the same model can also be written in a
compact code-like form:

```text
mu1_i = X_mu1[i, ] beta_mu1
mu2_i = X_mu2[i, ] beta_mu2
log(sigma1_i) = X_sigma1[i, ] beta_sigma1
log(sigma2_i) = X_sigma2[i, ] beta_sigma2
eta_rho12_i = X_rho12[i, ] beta_rho12
rho12_i = tanh(eta_rho12_i)
Omega_i[1, 1] = sigma1_i^2
Omega_i[2, 2] = sigma2_i^2
Omega_i[1, 2] = Omega_i[2, 1] = rho12_i * sigma1_i * sigma2_i
```

Implementation note: `drmTMB` internally evaluates residual correlations as
`0.99999999 * tanh(eta_rho12_i)`. The tiny multiplier keeps covariance
matrices strictly positive definite near the boundaries. It is not a
biological scaling factor, and it should not change how users interpret the
model.

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The mapping is:

| Symbol | Meaning | R syntax source |
|---|---|---|
| `y1_i`, `y2_i` | observed responses | left-hand sides of `mu1` and `mu2` formulas |
| `X_mu1`, `X_mu2` | response-specific mean design matrices | right-hand sides of `mu1` and `mu2` |
| `mu1_i`, `mu2_i` | fitted response-specific means | columns of `fitted(fit)` |
| `X_sigma1`, `X_sigma2` | response-specific residual scale design matrices | right-hand sides of `sigma1` and `sigma2` |
| `X_rho12` | residual correlation design matrix | right-hand side of `rho12` |
| `sigma1_i`, `sigma2_i` | residual SDs for responses 1 and 2 | `predict(fit, dpar = "sigma1")`, `predict(fit, dpar = "sigma2")` |
| `rho12_i` | residual response-response correlation | `rho12(fit)` |

The fixed effects may be the same, overlapping, or different across
parameters. For example, `x1` can appear in all five formulae, while `x2` might
appear only in `mu1`, `sigma1`, and `rho12`. Each parameter still gets its own
coefficient vector.

If both responses have the same location predictors, `mvbind()` is a shorthand
for the two location formulas:

```r
drmTMB(
  bf(
    mvbind(y1, y2) ~ x1 + x2,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

This expands internally to `mu1 = y1 ~ x1 + x2` and
`mu2 = y2 ~ x1 + x2`. Use explicit `mu1` and `mu2` formulas whenever the
responses need different location predictors.

## Worked example: behaviour coupling and disturbance

Suppose an ecologist measures two behaviours, activity and boldness, and wants
to know whether an environmental gradient changes their residual coupling after
accounting for response-specific means and residual SDs. The code below uses
simulated data, but the model structure matches that scientific question.

For this example, the fitted biological model is:

\[
\begin{aligned}
\begin{bmatrix} activity_i \\ boldness_i \end{bmatrix}
&\sim
\operatorname{MVN}
\left(
\begin{bmatrix} \mu_{activity,i} \\ \mu_{boldness,i} \end{bmatrix},
\Omega_i
\right),\\
\mu_{activity,i} &= \beta_{a0} + \beta_{a1} food_i +
                  \beta_{a2} temperature_i,\\
\mu_{boldness,i} &= \beta_{b0} + \beta_{b1} food_i,\\
\log(\sigma_{activity,i}) &= \gamma_{a0} + \gamma_{a1} food_i +
                            \gamma_{a2} temperature_i,\\
\log(\sigma_{boldness,i}) &= \gamma_{b0} + \gamma_{b1} food_i,\\
\eta_{\rho12,i} &= \delta_0 + \delta_1 disturbance_i,\\
\rho_{12,i} &= \tanh(\eta_{\rho12,i}).
\end{aligned}
\]

Here `rho12` is not the ordinary correlation between raw activity and raw
boldness. It is the residual correlation after the model has already adjusted
for the mean and residual SD predictors in each response.

```{r bivariate-coscale-simulate}
set.seed(1)
n <- 180
dat <- data.frame(
  food = rnorm(n),
  temperature = rnorm(n),
  disturbance = rnorm(n)
)
mu1 <- 0.2 + 0.5 * dat$food + 0.2 * dat$temperature
mu2 <- -0.1 + 0.4 * dat$food
sigma1 <- exp(-0.2 + 0.2 * dat$food - 0.1 * dat$temperature)
sigma2 <- exp(0.1 - 0.2 * dat$food)
eta_rho12 <- -0.1 + 0.4 * dat$disturbance
rho12 <- tanh(eta_rho12)
e1 <- rnorm(n)
e2 <- rho12 * e1 + sqrt(1 - rho12^2) * rnorm(n)
dat$activity <- mu1 + sigma1 * e1
dat$boldness <- mu2 + sigma2 * e2
```

The model lets `food` and `temperature` explain mean activity, lets `food`
explain mean boldness, lets `food` and `temperature` explain residual SD in
activity, lets `food` explain residual SD in boldness, and lets `disturbance`
explain residual correlation:

```{r bivariate-coscale-fit}
fit_biv <- drmTMB(
  drm_formula(
    mu1 = activity ~ food + temperature,
    mu2 = boldness ~ food,
    sigma1 = ~ food + temperature,
    sigma2 = ~ food,
    rho12 = ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Check the fit and inspect the coefficient table:

```{r bivariate-coscale-output}
check_drm(fit_biv)
summary(fit_biv)
```

How to read this output:

1. Rows beginning with `mu1:` describe expected activity, and rows beginning
   with `mu2:` describe expected boldness.
2. Rows beginning with `sigma1:` and `sigma2:` describe log residual SDs for
   activity and boldness.
3. Rows beginning with `rho12:` describe the residual activity-boldness
   correlation on the linear-predictor scale. Use `rho12(fit_biv)` or
   `rho12(fit_biv, newdata = ...)` to read that effect as a correlation.

The `rho12` coefficients are on the linear-predictor scale:

```{r bivariate-coscale-rho12}
coef(fit_biv, "rho12")
head(rho12(fit_biv))
```

`corpairs()` gives the same residual correlation in the long correlation-pair
format. The same table shape is the design target for future group-level,
phylogenetic, spatial, and study-level correlations:

```{r bivariate-coscale-corpairs}
corpairs(fit_biv)[
  c("level", "from_response", "to_response", "class", "parameter",
    "estimate", "min", "max", "modelled")
]
```

The `rho12` coefficients are fitted on the linear-predictor scale
`eta_rho12`; the response transform is
`rho12 = tanh(eta_rho12)` with the tiny internal guard described above. Use
`rho12(fit)` to read them as residual correlations, and use `corpairs(fit)` to
place residual `rho12` in a table that is ready to hold other correlation
levels later. In this example, a positive coefficient for `disturbance` means
that observations with larger disturbance have stronger positive residual
coupling between activity and boldness, after the model has already accounted
for the predictors in `mu1`, `mu2`, `sigma1`, and `sigma2`.

```{r bivariate-coscale-interpretation-table}
newdat <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = c(-1, 0, 1)
)
data.frame(
  disturbance = newdat$disturbance,
  sigma_activity = predict(fit_biv, newdata = newdat, dpar = "sigma1"),
  sigma_boldness = predict(fit_biv, newdata = newdat, dpar = "sigma2"),
  rho12 = rho12(fit_biv, newdata = newdat)
)
```

The same response-scale result can be visualised as a `rho12` curve:

```{r bivariate-coscale-rho12-curve, fig.width = 6, fig.height = 4}
rho_grid <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = seq(-2, 2, length.out = 80)
)
rho_grid$rho12 <- rho12(fit_biv, newdata = rho_grid)

plot(
  rho12 ~ disturbance,
  data = rho_grid,
  type = "l",
  lwd = 2,
  ylim = c(-1, 1),
  xlab = "Disturbance",
  ylab = "Fitted residual correlation (rho12)"
)
abline(h = 0, lty = 2, col = "grey60")
```

In this example, a positive `disturbance` coefficient for `rho12` means that
activity and boldness become more tightly positively coupled at higher
disturbance, after their means and residual SDs have been modelled. It does
not say that disturbance changes among-individual personality or plasticity
correlations; those require group-level covariance blocks and should appear as
separate rows in `corpairs(fit)`.

A concise reporting sentence would be: after accounting for food,
temperature, and response-specific residual SDs, the fitted residual
correlation between activity and boldness increased along the disturbance
gradient. In biological terms, individuals that were more active than expected
were also more bold than expected under higher disturbance.

## What this model does not estimate

This fitted model has no bivariate random effects, no group-level
personality/plasticity correlations, no phylogenetic correlation, and no
spatial field correlation. If bivariate random-effect syntax currently errors,
fit the fixed-effect bivariate model shown here or use the implemented
univariate Gaussian random-effect models until bivariate covariance blocks are
added.

For ecological and evolutionary examples, residual `rho12` can represent
changing residual coupling between traits such as activity and boldness, body
size and fecundity, or leaf area and seed number.

## Future group-level equations

The future double-hierarchical version should also allow group-level random
intercepts and random slopes in the mean structure:

```r
drmTMB(
  formula = drm_formula(
    mu1 = activity ~ food + disturbance + (1 + disturbance | p | ID),
    mu2 = boldness ~ food              + (1 + disturbance | p | ID),
    sigma1 = ~ food + disturbance,
    sigma2 = ~ food,
    rho12 = ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

This syntax is planned, not implemented. The two terms in
`(1 + disturbance | p | ID)` would create group-level intercept and slope
variation in the mean part. `sigma1` and `sigma2` are residual scale
parameters; additional random-effect scale components should be named and
extracted as group-level standard deviations, not confused with residual
`sigma`.

Symbolically, the future mean part would look like:

```text
mu1_ij = X_mu1[ij, ] beta_mu1 + b_0_1j + b_disturbance_1j disturbance_ij
mu2_ij = X_mu2[ij, ] beta_mu2 + b_0_2j + b_disturbance_2j disturbance_ij

[b_0_1j, b_disturbance_1j, b_0_2j, b_disturbance_2j]' ~ MVN(0, Sigma_mu_ID)
```

The covariance matrix `Sigma_mu_ID` would contain group-level standard
deviations and group-level correlations. Those correlations answer questions
about between-group association in intercepts and slopes. They are not
`rho12_i`. The parameter `rho12_i` remains the observation-level residual
correlation in `Omega_i`.

The middle `p` in `(1 | p | id)` follows grouped covariance-block syntax for
correlated group-level effects. Labelled one-slope blocks are implemented for
univariate Gaussian `mu`; shared bivariate `mu1`/`mu2` covariance blocks remain
future work. This syntax is distinct from residual `rho12`.

In phylogenetic location-scale models, correlations among phylogenetic mean and
scale effects answer different questions: mean-mean coevolution, scale-scale
co-divergence, and mean-scale evolutionary constraints. Those are group-level
covariance summaries. Residual `rho12` is the observation-level trait coupling
after those structured effects have been accounted for.

The long-term package should therefore support several correlation levels, not
only residual correlation. For example, a comparative model could estimate a
phylogenetic correlation between response-specific evolutionary deviations, a
non-phylogenetic species correlation after shared ancestry is accounted for,
and a residual `rho12` within observations. Those quantities should be named
and extracted separately, for example as `corpars$phylo`,
`corpars$species`, and `rho12(fit)`.
