---
title: "Bivariate non-Gaussian models: choose the association scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bivariate non-Gaussian models: choose the association scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

This article is for an applied researcher with two positive or heavy-tailed
responses. The first decision is not which correlation command to use. It is
which residual scale answers the biological question. `biv_lognormal()` and
`biv_student()` are direct joint models: they retain `rho12` because its scale
is defined by their likelihood. They are different from the post-fit
mixed-family `associate_pairs()` route, which estimates latent-normal `eta`.

## Positive traits: exact bivariate lognormal

For positive lognormal outcomes,
\((\log Y_1,\log Y_2)\) is bivariate normal. `sigma1` and `sigma2` are
log-response SDs and `rho12` is their log-residual correlation; it is not
automatically a raw-scale correlation.

```{r lognormal-example, eval = FALSE}
fit_log <- drmTMB(
  bf(mu1 = seed_mass ~ treatment, mu2 = root_mass ~ treatment,
     sigma1 = ~ 1, sigma2 = ~ 1, rho12 = ~ 1),
  family = biv_lognormal(), data = plants
)
rho12(fit_log)
fitted(fit_log)
```

A positive value means that plants heavier than expected on the log seed-mass
scale also tend to be heavier than expected on the log root-mass scale after
the treatment means are fitted. `fitted(fit_log)` returns arithmetic marginal
means on the original response scale, not the log-response locations shown by
the `mu1` and `mu2` predictors. Replace the example names with two positive
response columns in one complete paired data frame.

## Heavy-tailed traits: exact bivariate Student-t

For paired measurements with heavy tails, use the direct shared-degrees-of-
freedom Student-t model.

```{r student-example, eval = FALSE}
fit_t <- drmTMB(
  bf(mu1 = activity ~ habitat, mu2 = boldness ~ habitat,
     sigma1 = ~ 1, sigma2 = ~ 1, nu = ~ 1, rho12 = ~ 1),
  family = biv_student(), data = behaviour
)
rho12(fit_t)
```

`sigma1` and `sigma2` are Student-t scales, not marginal SDs. The model has
one shared `nu`; at finite `nu`, `rho12 = 0` means uncorrelated but not
independent because both outcomes share the same row-level mixing draw.

## A practical choice

Create the data once, retain rows with both outcomes and all predictors, and
fit both response formulas to that same object. This avoids accidentally
comparing fits with different row identities.

```{r compare-families, eval = FALSE}
# For example: positive_pairs <- na.omit(data.frame(y1, y2, x))
# Positive biomass traits: a log-response correlation is meaningful.
fit_log <- drmTMB(bf(mu1 = y1 ~ x, mu2 = y2 ~ x,
                     sigma1 = ~ 1, sigma2 = ~ 1, rho12 = ~ 1),
                  family = biv_lognormal(), data = positive_pairs)

# Heavy-tailed real-valued traits: use a shared-t residual scale.
fit_t <- drmTMB(bf(mu1 = y1 ~ x, mu2 = y2 ~ x,
                   sigma1 = ~ 1, sigma2 = ~ 1, nu = ~ 1, rho12 = ~ 1),
                family = biv_student(), data = real_pairs)
```

Both first slices require complete pairs and keep `sigma1`, `sigma2`, `rho12`,
and (for Student-t) `nu` intercept-only. They do not currently add random
effects, `rho12` predictors, intervals, coverage, or a general non-Gaussian
bivariate claim. If the responses have different marginal families, read
[Frozen-margin mixed-family association](cross-family.html) instead.
