---
title: "Bivariate non-Gaussian models: choose a joint or staged association"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bivariate non-Gaussian models: choose a joint or staged association}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_penguins <- requireNamespace("palmerpenguins", quietly = TRUE)
```

With two non-Gaussian outcomes, first decide what kind of association answers
the scientific question. A direct joint likelihood estimates both outcomes and
their residual association together. A frozen-margin association first fits the
two outcomes separately and then estimates a latent-normal association while
holding the fitted margins fixed. These are different estimands.

This article helps you choose a route for two outcomes measured on the same
rows. The mixed-outcome route is a beta development interface: it does not
provide interval, coverage, random-effect, or generic-pair claims. The exact
direct lognormal route now has callable uncertainty methods for a constant
`rho12`, with a retained fixed-effect calibration ladder for its tested domain;
do not treat a finite interval as a validated general non-Gaussian claim.
For a released direct Gaussian joint model, use `biv_gaussian()` and
[Changing residual coupling with rho12](bivariate-coscale.html).

## Start by choosing the association scale

| Your paired outcomes | Route | What the association means |
| --- | --- | --- |
| Two Gaussian traits | Direct `biv_gaussian()` model | Gaussian residual `rho12` |
| Two positive traits plausibly lognormal | Post-0.6 direct `biv_lognormal()` model | Log-response residual `rho12` |
| Two heavy-tailed real-valued traits | Post-0.6 direct `biv_student()` model | Shared-Student-t residual/scatter `rho12` |
| A reviewed mixed or discrete pair | `biv_associate()` staged route | Latent-normal copula `eta` after frozen margins; one numeric association slope is beta-supported for Bernoulli x NB2 |

Neither `rho12` nor `eta` is automatically the Pearson correlation of the two
raw response columns. The scale is part of the scientific interpretation, not
just a computational detail.

## Direct joint model: two positive traits

For two positive lognormal outcomes, the model is joint on the log-response
scale:

\[
(\log Y_{1i}, \log Y_{2i}) \sim N_2\left(
(\mu_{1i}, \mu_{2i}),
\begin{bmatrix}
\sigma_1^2 & \rho_{12}\sigma_1\sigma_2 \\
\rho_{12}\sigma_1\sigma_2 & \sigma_2^2
\end{bmatrix}
\right).
\]

Here `rho12` is the correlation of the two log-response residuals. A positive
value means that units above their fitted log-response expectation on one trait
also tend to be above it on the other. It is not automatically the correlation
of the two traits on their original scale.

### Real example: penguin size after measured composition differences

The openly licensed `palmerpenguins` data set records body size in three Palmer
Archipelago penguin species. It is distributed under CC0 and collates data from
the Palmer Station LTER study ([data policy](https://allisonhorst.github.io/palmerpenguins/);
[Gorman et al. 2014](https://doi.org/10.1371/journal.pone.0090081)).

Our question is conditional: after allowing both margins to vary by species,
sex, and centred year, do birds with longer-than-expected flippers also have
greater-than-expected body mass? This is not a causal claim that flippers
determine mass, nor a replacement for a raw scatterplot.

```{r penguin-data, eval = has_penguins}
penguins <- subset(
  palmerpenguins::penguins,
  complete.cases(species, sex, year, flipper_length_mm, body_mass_g) &
    flipper_length_mm > 0 & body_mass_g > 0
)
penguins$year_c <- penguins$year - mean(penguins$year)
nrow(penguins)
```

```{r penguin-raw-scale, eval = has_penguins, fig.width = 6, fig.height = 4, fig.cap = "Raw log-scale flipper length and body mass for analysed complete rows. This descriptive relationship is not the fitted residual association."}
plot(
  log(body_mass_g) ~ log(flipper_length_mm), data = penguins,
  pch = 16, col = grDevices::adjustcolor("#087f8c", alpha.f = 0.45),
  xlab = "log flipper length (mm)", ylab = "log body mass (g)"
)
abline(stats::lm(log(body_mass_g) ~ log(flipper_length_mm), data = penguins),
  col = "#14344a", lwd = 2
)
```

```{r penguin-direct-lognormal, eval = has_penguins}
fit_log <- drmTMB(
  bf(mu1 = flipper_length_mm ~ species + sex + year_c,
     mu2 = body_mass_g ~ species + sex + year_c,
     sigma1 = ~ 1, sigma2 = ~ 1, rho12 = ~ 1),
  family = biv_lognormal(), data = penguins
)

rho12(fit_log)[1]
check_drm(fit_log)

# All three target the log-residual rho12, not a raw-scale correlation.
ci_wald <- confint(fit_log, parm = "rho12", method = "wald")
ci_profile <- confint(
  fit_log, parm = "rho12", method = "profile", profile_engine = "endpoint"
)
ci_bootstrap <- confint(
  fit_log, parm = "rho12", method = "bootstrap", R = 99, seed = 20260724
)
interval_row <- function(x, label) {
  data.frame(
    label = label,
    estimate = unname(rho12(fit_log)[1]),
    lower = x$lower,
    upper = x$upper,
    status = x$conf.status,
    note = x$profile.message,
    stringsAsFactors = FALSE
  )
}
intervals <- rbind(
  interval_row(ci_wald, "Wald"),
  interval_row(ci_profile, "Profile"),
  interval_row(
    ci_bootstrap,
    sprintf(
      "Bootstrap (%d/%d retained)",
      ci_bootstrap$bootstrap.n,
      ci_bootstrap$bootstrap.n + ci_bootstrap$bootstrap.failed
    )
  )
)
intervals
```

```{r penguin-intervals, eval = has_penguins, fig.width = 6, fig.height = 3.5, fig.cap = "Intervals and the fitted direct log-residual association in the penguin model. The plot shows calculation output, not a generic calibration guarantee. The bootstrap label gives retained refits; a low retained fraction is a diagnostic rather than hidden precision."}
at <- rev(seq_len(nrow(intervals)))
plot(
  intervals$lower, at, xlim = c(-1, 1), ylim = c(0.5, 3.5), yaxt = "n",
  pch = 16, xlab = expression(rho[12]), ylab = "", col = "#14344a"
)
segments(intervals$lower, at, intervals$upper, at, lwd = 3, col = "#087f8c")
points(intervals$estimate, at, pch = 16, col = "#14344a")
axis(2, at = at, labels = intervals$label, las = 1)
abline(v = 0, lty = 2, col = "grey50")
```

`fitted(fit_log)` returns arithmetic marginal means on the original response
scale. The `mu1` and `mu2` predictors remain log-scale locations. These direct
models estimate the two margins and `rho12` together; they are not staged
frozen-margin fits. Wald is a fast link-scale approximation; profile likelihood
checks the exact likelihood geometry; the bootstrap refits both margins and
`rho12`. Inspect `check_drm(fit_log)` and bootstrap diagnostics, especially
near \(|\rho_{12}|=1\). In this data set, do not hide failed bootstrap refits:
the printed `bootstrap.n` and `bootstrap.failed` fields are part of the result.
The external data example illustrates interpretation and diagnostics; it does
not itself establish coverage calibration or prove lognormal adequacy. Inspect
residual diagnostics and compare scientifically defensible alternatives before
reporting a substantive association.

## Direct joint model: two heavy-tailed traits

For two real-valued traits with occasional extreme observations, the bivariate
Student-t model uses a shared degrees-of-freedom parameter, `nu`:

```{r student-example, eval = FALSE}
fit_t <- drmTMB(
  bf(mu1 = activity ~ habitat, mu2 = boldness ~ habitat,
     sigma1 = ~ 1, sigma2 = ~ 1, nu = ~ 1, rho12 = ~ 1),
  family = biv_student(), data = behaviour
)

rho12(fit_t)
```

`sigma1` and `sigma2` are Student-t scales, not necessarily marginal standard
deviations. The model has one shared `nu`, which controls tail heaviness. At a
finite `nu`, `rho12 = 0` means zero residual correlation, but not necessarily
complete independence because the two outcomes share the same row-level
heavy-tail mechanism.

## Staged association: different kinds of outcome

If one response is binary and the other is a count, there is no shared direct
Gaussian residual scale. The frozen-margin route instead fits a binary
probability model and a count model, then estimates latent-normal `eta` from
their fitted distributions.

```{r staged-example, eval = FALSE}
paired_data <- na.omit(data.frame(
  bred, offspring, habitat, habitat_score, season
))

assoc <- biv_associate(
  bf(mu = bred ~ habitat),
  bf(mu = offspring ~ habitat, sigma = ~ season),
  family = list(binomial(), nbinom2()),
  data = paired_data,
  association = ~ 1
)

association(assoc)
```

This is one convenient R call, but not one jointly fitted model.
`biv_associate()` fits the two margins, freezes them, and estimates only `eta`.
The reported value is a latent-normal copula correlation conditional on those
fitted margins. It is not `rho12`, a logit coefficient, an odds ratio, or an
observed-scale correlation.

For a binary response, the model observes whether a hidden tendency crossed a
threshold. It does not observe an ordinary numeric residual. Read
[Frozen-margin association for mixed outcome pairs](cross-family.html) for the
binary-threshold explanation, the reviewed pair classes, and the rule for
withheld estimates.

For the literal-Bernoulli x ordinary-NB2 beta route, `association = ~ x` can
also model one numeric association gradient. It estimates a row-specific
latent-normal `eta_i`, after both margins have been fitted and frozen. See
[Association between mixed outcome pairs](cross-family.html) for the exact
syntax, output, and limits.

## What is currently outside these routes

The direct same-family first slices keep `sigma1`, `sigma2`, `rho12`, and `nu`
constant across rows. The staged route usually keeps `eta` constant; its one
beta exception is a single numeric association slope for literal-Bernoulli x
ordinary-NB2. Its reviewed fixed-effect margins may include their explicitly
supported covariates before they are frozen. Neither route currently adds random
effects, general association-predictor grammar, missing-data support, or a
general non-Gaussian bivariate claim. The staged route has no two-stage standard
errors, profiles, confidence intervals, or coverage claim. The direct lognormal
constant-`rho12` profile, Wald, and bootstrap methods have direct calibration
evidence only for the fixed-effect DGP in the Arc 6 coverage artifact; they do
not validate the staged route or a general non-Gaussian association claim.
Choose the route whose stated association scale matches the biological question,
and treat every boundary as part of the model definition.
