---
title: "Changing residual coupling with rho12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing residual coupling with rho12}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

This tutorial introduces bivariate location-scale regression with
predictor-dependent residual correlation. The scientific idea is that two
responses can change in mean, change in residual variance, and change in
residual coupling.

Current fixed-effect syntax uses one formula per distributional parameter:

```r
drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The central modelling question is whether predictors change residual coupling
between two responses, not only their means or variances.

The key parameter is `rho12`. A location-scale model asks whether predictors
change residual SDs. A location-coscale model asks whether predictors also
change the residual covariance structure between two responses; in the current
bivariate Gaussian model, that structure is the residual correlation `rho12`.
In ecology and evolution, this can ask whether the residual association between
traits such as body mass and litter size differs among terrestrial, aquatic,
and aerial lifestyles after modelling response-specific means and dispersion.

## Equations for the implemented model

For observation `i`, the implemented fixed-effect bivariate Gaussian model is:

\[
\begin{bmatrix}
y_{1i} \\
y_{2i}
\end{bmatrix}
\mid
\mu_{1i}, \mu_{2i}, \sigma_{1i}, \sigma_{2i}, \rho_{12i}
\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{1i} \\
\mu_{2i}
\end{bmatrix},
\Omega_i
\right).
\]

The five distributional predictors are:

\[
\begin{aligned}
\mu_{1i} &= X_{\mu 1,i}\beta_{\mu 1}, \\
\mu_{2i} &= X_{\mu 2,i}\beta_{\mu 2}, \\
\log(\sigma_{1i}) &= X_{\sigma 1,i}\beta_{\sigma 1}, \\
\log(\sigma_{2i}) &= X_{\sigma 2,i}\beta_{\sigma 2}, \\
\eta_{\rho12,i} &= X_{\rho12,i}\beta_{\rho12}, \\
\rho_{12i} &= \tanh(\eta_{\rho12,i}).
\end{aligned}
\]

The residual covariance matrix is:

\[
\Omega_i =
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}.
\]

Component meanings:

| Component | Meaning |
|---|---|
| `mu1`, `mu2` | expected values for responses 1 and 2 |
| `sigma1`, `sigma2` | residual SDs for responses 1 and 2 |
| `eta_rho12` | unconstrained linear predictor for residual correlation |
| `rho12` | residual response-response correlation within observation `i` |
| `Omega_i` | residual covariance matrix after means and residual SDs are modelled |

For implementation cross-checking, the same model can also be written in a
compact code-like form:

```text
mu1_i = X_mu1[i, ] beta_mu1
mu2_i = X_mu2[i, ] beta_mu2
log(sigma1_i) = X_sigma1[i, ] beta_sigma1
log(sigma2_i) = X_sigma2[i, ] beta_sigma2
eta_rho12_i = X_rho12[i, ] beta_rho12
rho12_i = tanh(eta_rho12_i)
Omega_i[1, 1] = sigma1_i^2
Omega_i[2, 2] = sigma2_i^2
Omega_i[1, 2] = Omega_i[2, 1] = rho12_i * sigma1_i * sigma2_i
```

Implementation note: `drmTMB` internally evaluates residual correlations as
`0.99999999 * tanh(eta_rho12_i)`. The tiny multiplier keeps covariance
matrices strictly positive definite near the boundaries. It is not a
biological scaling factor, and it should not change how users interpret the
model.

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The mapping is:

| Symbol | Meaning | R syntax source |
|---|---|---|
| `y1_i`, `y2_i` | observed responses | left-hand sides of `mu1` and `mu2` formulas |
| `X_mu1`, `X_mu2` | response-specific mean design matrices | right-hand sides of `mu1` and `mu2` |
| `mu1_i`, `mu2_i` | fitted response-specific means | columns of `fitted(fit)` |
| `X_sigma1`, `X_sigma2` | response-specific residual scale design matrices | right-hand sides of `sigma1` and `sigma2` |
| `X_rho12` | residual correlation design matrix | right-hand side of `rho12` |
| `sigma1_i`, `sigma2_i` | residual SDs for responses 1 and 2 | `predict(fit, dpar = "sigma1")`, `predict(fit, dpar = "sigma2")` |
| `rho12_i` | residual response-response correlation | `rho12(fit)` |

The fixed effects may be the same, overlapping, or different across
parameters. For example, `x1` can appear in all five formulae, while `x2` might
appear only in `mu1`, `sigma1`, and `rho12`. Each parameter still gets its own
coefficient vector.

If both responses have the same location predictors, `mvbind()` is a shorthand
for the two location formulas:

```r
drmTMB(
  bf(
    mvbind(y1, y2) ~ x1 + x2,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

This expands internally to `mu1 = y1 ~ x1 + x2` and
`mu2 = y2 ~ x1 + x2`. Use explicit `mu1` and `mu2` formulas whenever the
responses need different location predictors.

## Worked example: behaviour coupling and disturbance

Suppose an ecologist measures two behaviours, activity and boldness, and wants
to know whether an environmental gradient changes their residual coupling after
accounting for response-specific means and residual SDs. The code below uses
simulated data, but the model structure matches that scientific question.

For this example, the fitted biological model is:

\[
\begin{aligned}
\begin{bmatrix}
\text{activity}_i\\
\text{boldness}_i
\end{bmatrix}
&\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{\text{activity},i}\\
\mu_{\text{boldness},i}
\end{bmatrix},
\Omega_i
\right),\\
\mu_{\text{activity},i}
  &= \beta_{a0} + \beta_{a1}\text{food}_i +
     \beta_{a2}\text{temperature}_i,\\
\mu_{\text{boldness},i}
  &= \beta_{b0} + \beta_{b1}\text{food}_i,\\
\log(\sigma_{\text{activity},i})
  &= \gamma_{a0} + \gamma_{a1}\text{food}_i +
     \gamma_{a2}\text{temperature}_i,\\
\log(\sigma_{\text{boldness},i})
  &= \gamma_{b0} + \gamma_{b1}\text{food}_i,\\
\eta_{\rho12,i} &= \delta_0 + \delta_1\text{disturbance}_i,\\
\rho_{12,i} &= \tanh(\eta_{\rho12,i}).
\end{aligned}
\]

Here `rho12` is not the ordinary correlation between raw activity and raw
boldness. It is the residual correlation after the model has already adjusted
for the mean and residual SD predictors in each response.

```{r bivariate-coscale-simulate}
set.seed(1)
n <- 180
dat <- data.frame(
  food = rnorm(n),
  temperature = rnorm(n),
  disturbance = rnorm(n)
)
mu1 <- 0.2 + 0.5 * dat$food + 0.2 * dat$temperature
mu2 <- -0.1 + 0.4 * dat$food
sigma1 <- exp(-0.2 + 0.2 * dat$food - 0.1 * dat$temperature)
sigma2 <- exp(0.1 - 0.2 * dat$food)
eta_rho12 <- -0.1 + 0.4 * dat$disturbance
rho12 <- tanh(eta_rho12)
e1 <- rnorm(n)
e2 <- rho12 * e1 + sqrt(1 - rho12^2) * rnorm(n)
dat$activity <- mu1 + sigma1 * e1
dat$boldness <- mu2 + sigma2 * e2
```

The model lets `food` and `temperature` explain mean activity, lets `food`
explain mean boldness, lets `food` and `temperature` explain residual SD in
activity, lets `food` explain residual SD in boldness, and lets `disturbance`
explain residual correlation:

```{r bivariate-coscale-fit}
fit_biv <- drmTMB(
  drm_formula(
    mu1 = activity ~ food + temperature,
    mu2 = boldness ~ food,
    sigma1 = ~ food + temperature,
    sigma2 = ~ food,
    rho12 = ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Check the fit and inspect the coefficient table:

```{r bivariate-coscale-output}
check_drm(fit_biv)
summary(fit_biv)
```

How to read this output:

1. Rows beginning with `mu1:` describe expected activity, and rows beginning
   with `mu2:` describe expected boldness.
2. Rows beginning with `sigma1:` and `sigma2:` describe log residual SDs for
   activity and boldness.
3. Rows beginning with `rho12:` describe the residual activity-boldness
   correlation on the linear-predictor scale. Use `rho12(fit_biv)` or
   `rho12(fit_biv, newdata = ...)` to read that effect as a correlation.

The `rho12` coefficients are on the linear-predictor scale:

```{r bivariate-coscale-rho12}
coef(fit_biv, "rho12")
head(rho12(fit_biv))
```

`corpairs()` gives the same residual correlation in the long correlation-pair
format. The same table shape is the design target for future group-level,
phylogenetic, spatial, and study-level correlations:

```{r bivariate-coscale-corpairs}
corpairs(fit_biv)[
  c("level", "from_response", "to_response", "class", "parameter",
    "estimate", "min", "max", "modelled")
]
```

The `rho12` coefficients are fitted on the linear-predictor scale
`eta_rho12`; the response transform is
`rho12 = tanh(eta_rho12)` with the tiny internal guard described above. Use
`rho12(fit)` to read them as residual correlations, and use `corpairs(fit)` to
place residual `rho12` in a table that is ready to hold other correlation
levels later. In this example, a positive coefficient for `disturbance` means
that observations with larger disturbance have stronger positive residual
coupling between activity and boldness, after the model has already accounted
for the predictors in `mu1`, `mu2`, `sigma1`, and `sigma2`.

```{r bivariate-coscale-interpretation-table}
newdat <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = c(-1, 0, 1)
)
rho_table <- data.frame(
  disturbance = newdat$disturbance,
  sigma_activity = predict(fit_biv, newdata = newdat, dpar = "sigma1"),
  sigma_boldness = predict(fit_biv, newdata = newdat, dpar = "sigma2"),
  rho12 = rho12(fit_biv, newdata = newdat)
)
rho_table$residual_covariance <- with(
  rho_table,
  rho12 * sigma_activity * sigma_boldness
)
rho_table$residual_variance_activity <- rho_table$sigma_activity^2
rho_table$residual_variance_boldness <- rho_table$sigma_boldness^2
round(rho_table, 3)
```

This table gives three quantities a reader can report directly. `sigma_activity`
and `sigma_boldness` are the fitted residual SDs for each response at the
chosen covariate values. `rho12` is the fitted residual correlation between the
two responses. `residual_covariance` is the off-diagonal element of
`\Omega_i`, computed as `rho12 * sigma_activity * sigma_boldness`. The two
residual variance columns are the diagonal elements of `\Omega_i`.

## What should I report?

For a location-coscale model, a useful result sentence usually needs all three
distributional pieces: means, residual SDs, and residual coupling. A minimal
table for the coscale part is:

```{r bivariate-coscale-report-table}
report_grid <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = c(-1.5, 0, 1.5)
)
report_table <- data.frame(
  disturbance = report_grid$disturbance,
  eta_rho12 = predict(
    fit_biv,
    newdata = report_grid,
    dpar = "rho12",
    type = "link"
  ),
  rho12 = rho12(fit_biv, newdata = report_grid),
  sigma_activity = predict(fit_biv, newdata = report_grid, dpar = "sigma1"),
  sigma_boldness = predict(fit_biv, newdata = report_grid, dpar = "sigma2")
)
report_table$residual_covariance <- with(
  report_table,
  rho12 * sigma_activity * sigma_boldness
)
report_table$residual_variance_activity <- report_table$sigma_activity^2
report_table$residual_variance_boldness <- report_table$sigma_boldness^2
round(report_table, 3)
```

The raw correlation between activity and boldness is not the same estimand as
`rho12`. The raw correlation mixes mean structure, residual SDs, and residual
coupling. The fitted `rho12` is the residual correlation after the model has
accounted for the formulae in `mu1`, `mu2`, `sigma1`, and `sigma2`:

```{r bivariate-coscale-raw-vs-residual}
round(data.frame(
  raw_activity_boldness_correlation = stats::cor(dat$activity, dat$boldness),
  mean_fitted_residual_rho12 = mean(rho12(fit_biv))
), 3)
```

When reporting this model, avoid saying only "activity and boldness were
correlated". A better sentence is: "After modelling response-specific means
and residual SDs, the fitted residual activity-boldness correlation increased
with disturbance." That sentence tells the reader which correlation has been
modelled.

The same response-scale result can be visualised as a `rho12` curve:

```{r bivariate-coscale-rho12-curve, fig.width = 6, fig.height = 4}
rho_grid <- data.frame(
  food = 0,
  temperature = 0,
  disturbance = seq(-2, 2, length.out = 80)
)
rho_grid$rho12 <- rho12(fit_biv, newdata = rho_grid)

plot(
  rho12 ~ disturbance,
  data = rho_grid,
  type = "l",
  lwd = 2,
  ylim = c(-1, 1),
  xlab = "Disturbance",
  ylab = "Fitted residual correlation (rho12)"
)
abline(h = 0, lty = 2, col = "grey60")
```

In this example, a positive `disturbance` coefficient for `rho12` means that
activity and boldness become more tightly positively coupled at higher
disturbance, after their means and residual SDs have been modelled. It does
not say that disturbance changes among-individual personality or plasticity
correlations; those require group-level covariance blocks and should appear as
separate rows in `corpairs(fit)`.

A concise reporting sentence would be: after accounting for food,
temperature, and response-specific residual SDs, the fitted residual
correlation between activity and boldness increased along the disturbance
gradient. In biological terms, individuals that were more active than expected
were also more bold than expected under higher disturbance.

## What this model does not estimate

This fitted model has no group-level personality or plasticity correlations, no
phylogenetic correlation, and no spatial field correlation. The first
bivariate group-level covariance slices are implemented for matching labelled
random intercepts, such as `(1 | p | ID)` in both `mu1` and `mu2`, or in both
`sigma1` and `sigma2`. If you need random slopes, cross-parameter covariance,
phylogenetic covariance, or spatial covariance, fit the nearest implemented
bivariate or univariate Gaussian random-effect model until those covariance
blocks are added.

For ecological and evolutionary examples, residual `rho12` can represent
changing residual coupling between traits such as activity and boldness, body
size and fecundity, or leaf area and seed number.

## Group-level equations

The first fitted group-level version adds matching labelled random intercepts
in the mean structure:

```r
drmTMB(
  formula = drm_formula(
    mu1 = activity ~ food + disturbance + (1 | p | ID),
    mu2 = boldness ~ food              + (1 | p | ID),
    sigma1 = ~ food + disturbance,
    sigma2 = ~ food,
    rho12 = ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The shared `p` label creates one group-level covariance block for the
`mu1` and `mu2` random intercepts. `sigma1` and `sigma2` are residual scale
parameters; fitted random-intercept SDs are group-level standard deviations in
`sdpars$mu`, not residual `sigma`.

Symbolically, the implemented random-intercept mean part is:

\[
\begin{aligned}
\mu_{1ij}
  &= X_{\mu 1}[ij, ]\beta_{\mu 1} + b_{0,1j},\\
\mu_{2ij}
  &= X_{\mu 2}[ij, ]\beta_{\mu 2} + b_{0,2j},\\
\begin{bmatrix}
b_{0,1j}\\
b_{0,2j}
\end{bmatrix}
  &\sim \operatorname{MVN}(\mathbf{0}, \Sigma_{\mu,ID}).
\end{aligned}
\]

The covariance matrix `Sigma_mu_ID` would contain group-level standard
deviations and one group-level correlation. That correlation answers whether
groups with higher average activity also tend to have higher average boldness.
It is not `rho12_i`. The parameter `rho12_i` remains the observation-level
residual correlation in `Omega_i`.

After fitting this grouped model, `corpairs()` keeps the two correlation levels
separate:

```r
corpairs(fit_group, level = "residual")
corpairs(fit_group, level = "group", group = "ID", block = "p")
```

The first call returns the residual `rho12` row. The second returns the
group-level mean-mean row for the labelled `mu1`/`mu2` random-intercept block.

For this fitted block, `check_drm()` reports whether group levels are
replicated and whether either group-level SD is tiny relative to the matching
residual scale. A note there is not a proof that the model is wrong; it tells
you to inspect whether the group-level SDs and correlation are supported by the
data before treating them as biological conclusions.

Use `profile_targets(fit)` before requesting profile intervals for this block.
The group-level SD targets are named like `sd:mu:mu1:(1 | p | id)` and
`sd:mu:mu2:(1 | p | id)`, while the group-level correlation target is named
like `cor:mu:cor(mu1:(Intercept),mu2:(Intercept) | p | id)`. These are
separate from the residual-correlation target `rho12`. Returned profile rows
include `profile.boundary` and `profile.message`; a non-`"ok"` message is a
prompt to inspect `check_drm(fit)` and the profile path before treating the
interval as a stable biological estimate.

The middle `p` in `(1 | p | id)` follows grouped covariance-block syntax for
correlated group-level effects. Labelled one-slope blocks are implemented for
univariate Gaussian `mu`; bivariate random slopes remain future work. This
syntax is distinct from residual `rho12`.

In phylogenetic location-scale models, correlations among phylogenetic mean and
scale effects answer different questions: mean-mean coevolution, scale-scale
co-divergence, and mean-scale evolutionary constraints. Those are group-level
covariance summaries. Residual `rho12` is the observation-level
trait coupling after fitted structured effects have been accounted for.

The long-term package should therefore support several correlation levels, not
only residual correlation. For example, a comparative model could estimate a
phylogenetic correlation between response-specific evolutionary deviations, a
non-phylogenetic species correlation after shared ancestry is accounted for,
and a residual `rho12` within observations. Those quantities should be named
and extracted separately, for example as `corpars$phylo`,
`corpars$species`, and `rho12(fit)`.
