---
title: "What drmTMB can and can't do"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What drmTMB can and can't do}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

If you are an ecology or evolution graduate student deciding whether
`drmTMB` fits your problem, the first question is not "can I fit this
model?" It is "which numbers from the fitted model can I report with a
straight face?" A model can converge, print coefficients, and still hand you
a confidence/compatibility interval that does not cover the truth at the
rate it claims to. This page separates what `drmTMB` fits from what
`drmTMB` has checked, cell by cell, so you know whether to trust a point
estimate, an interval, or neither before it goes in a manuscript.

Throughout, `mu` is the location term, `sigma` is the scale term,
family-specific parameters such as Student-t `nu` are shape terms, and
coscale means a residual-correlation parameter such as bivariate `rho12` --
the association left between two responses after their means and scales are
modelled. Those four words describe what a formula changes; they say
nothing about whether the resulting interval has been checked. That check
is what the rest of this page reports.

## Four tiers, defined once

Every model/effect combination in `drmTMB` sits in one of four tiers.

1. **Inference-ready.** The confidence/compatibility interval has been
   coverage-checked by simulation and holds, at least at an honest
   small-sample floor (see the caveats below). You can report the point
   estimate and the interval.
2. **Recovery-only.** The point estimate has been checked against a known
   simulated truth and recovers it. The interval has not been
   coverage-checked and should be treated as provisional: report the
   point estimate, and treat any interval on the same row as a plausible
   range rather than a calibrated one.
3. **Diagnostic-only.** A deterministic or single-smoke fit verifies that the
   route is wired to the requested parameter and that its extractors work, but
   no retained-denominator seed ladder supports recovery. Treat the estimate as
   a debugging or feasibility result, not as a recovery-backed scientific
   estimate.
4. **Not available / rejected by design.** `drmTMB` refuses to fit the
   request outright. Some of these routes are planned for a later release;
   some are out of scope for the package's current design. Either way,
   fitting it anyway is not an option -- use the alternative named below
   each entry.

No cell in `drmTMB` is coverage-verified at the strictest bar the project
tracks internally as "supported" (nominal-exact coverage with symmetric
misses on both sides). The census does mark four cells `fit_status =
supported` -- ordinary (unstructured) Gaussian and bivariate-Gaussian random
effects -- but those are point-trustworthy with interval coverage still
planned, not coverage-verified. Read "inference-ready" as an honest floor
for a first fitted structured route, not as parity with a long-mature
fixed-effect interval in a package such as `glmmTMB` or `lme4`.

## At a glance

| Model / effect | Tier | What you can trust |
| --- | --- | --- |
| Ordinary (unstructured) Gaussian / bivariate-Gaussian random effects (intercepts, slopes) | Point-trustworthy (census "supported", coverage planned) | Point estimate -- the everyday random-intercept/slope mixed model; its interval is not yet coverage-verified |
| Gaussian `mu` q1 intercept -- `phylo()`, `spatial()`, `relmat()` | Inference-ready | Point estimate and the default location-axis bias-corrected, small-sample-t Wald interval; coverage is mildly conservative in the tested campaigns |
| Gaussian `sigma` q1 one-slope -- `phylo()`, `animal()`, `relmat()` | Inference-ready | Point estimate and raw uncorrected log-SD Wald-z interval; profile is diagnostic-only at `g = 8` |
| Bivariate Gaussian slope-only `mu1:x`/`mu2:x` q2 mean-mean -- `phylo()`, `relmat()` | Inference-ready (default-corrected channel only) | Point estimate and interval from plain `confint(fit)`; intercept-only and explicitly uncorrected Wald intervals are not promoted by this evidence |
| binomial / Poisson / beta / nbinom2 unstructured `mu` | Inference-ready | Point estimate and Wald interval from `n` &gt;= 50 |
| nbinom2 unstructured location-scale (`mu` and `sigma ~ x`) | Inference-ready | Point estimate and Wald interval on both formulas |
| beta unstructured location-scale, interior proportions | Inference-ready | Point estimate and Wald interval; exact 0 or 1 needs `zero_one_beta()` instead |
| Ordinary independent `mu` random slope `(0 + x \| id)` -- binomial, cumulative-logit, skew-normal, Tweedie, zero-one-beta | Inference-ready with caveats | Profile interval for the natural-scale slope SD, ML only, not `supported`, no Wald or point-bias claim. The tested floor differs by family, so do not generalize one number: skew-normal / Tweedie / zero-one-beta at `M >= 16` (true SD 0.50, ML-Laplace); binomial at `M >= 32` (true SD 0.6, ML-Laplace); cumulative-logit at `M >= 40` (true SD 0.50, adaptive Gauss-Hermite + Cox-Reid). Zero-one-beta is generator-qualified (see the caveat below) |
| Poisson / nbinom2 structured `mu` q1 -- `phylo()`, `spatial()`, `animal()`, `relmat()` | Recovery-only | Point estimate only |
| nbinom2 structured `sigma` -- `phylo()`, `spatial()`, `animal()`, `relmat()` | Recovery-only | Point estimate only; scale-targeting bug fixed in 0.4.0 |
| Row-specific recovery slices: beta `animal()` on `mu`/`sigma`, Student-t `mu ~ spatial(1 + x | ...)`, and Gamma `mu ~ relmat()` | Recovery-only | Point estimate only; no interval or coverage promotion |
| Bivariate Gaussian spatial q2 location-intercept REML -- matching labelled `spatial(1 | p | site, coords = coords)` in `mu1` and `mu2` | Recovery-only | Point estimates for both structured SDs and their latent correlation; requires intercept-only `sigma1`, `sigma2`, and `rho12`, complete pairs, unit weights, no known `meta_V()`, and no additional ordinary random, direct-SD, or `corpair()` layer; no interval or coverage promotion |
| Bivariate Gaussian supplied-`K` relmat q2 location-intercept REML -- matching labelled `relmat(1 | p | id, K = K)` in `mu1` and `mu2` | Recovery-only (`point_fit_recovery`) | Point estimates for both structured SDs and their latent relatedness correlation; requires the same named `K`, group ordering, and label in both formulas, intercept-only `sigma1`, `sigma2`, and `rho12`, complete pairs, unit weights, and no additional random-effect, scale-side, `meta_V()`, direct-SD, or `corpair()` layer; no interval or coverage promotion |
| Row-specific single-smoke slices: ordinal `mu ~ phylo()`, truncated-nbinom2 `hu ~ relmat()`, Student-t `nu ~ phylo()`, Student-t intercept-only `mu ~ spatial(1 | ...)`, Poisson slope-only `mu ~ spatial(0 + x | ...)`, Poisson labelled-scalar `mu ~ spatial()`, Poisson `mu ~ spatial(1 | ...) + (1 | id)`, Poisson `zi ~ spatial()`, fixed-`zi` Poisson `mu ~ spatial()`, and fixed-`zi` NB2 `mu ~ spatial()` | Diagnostic-only | Use only to confirm fit/extractor feasibility; no recovery, interval, or coverage claim |
| Structured effects for lognormal, skew-normal, Tweedie, zero-one-beta | Not available | Use a Gaussian or basic count/proportion structured route, or fixed effects |
| Gaussian pure-`mu` univariate REML -- `spatial()`, `animal()`, `relmat()` | Inference-ready with caveats | Unlabelled intercept or independent intercept plus one numeric slope, with `sigma ~ 1`; report a direct structured-SD profile interval only inside the tested discrete domains below |
| REML for non-Gaussian families; non-phylogenetic mean-side REML outside the pure-`mu` Arc 1a boundary and the exact bivariate spatial and supplied-`K` relmat q2 intercepts above; all other bivariate non-phylogenetic structured REML | Not available (rejected by design) | Use `REML = FALSE` (maximum likelihood) |
| `q4`/`q6`/`q8`/`q12` covariance interval promotion; derived-correlation intervals | Not available (planned) | Use the fitted point estimate; try `profile_targets()` for a direct target |

The rest of this page expands each row with the caveat that changes how you
should read it.

## Tier 1: inference-ready

### Gaussian structured random effects: eleven anchor cells

Eleven conceptual cells across `drmTMB`'s structured Gaussian-family random-effect
routes are inference-ready today. Eight are the established ML model-surface
anchors: six sit on a univariate Gaussian `mu` or `sigma` formula, all at q1,
and two are bivariate-Gaussian slope-only `mu1:x`/`mu2:x` q2 mean-mean blocks.
The other three
are the Arc 1a REML estimator cells for pure-`mu` univariate `spatial()`,
`animal()`, and `relmat()` routes:

- three q1 mean-intercept rows -- `phylo()`, `spatial()`, `relmat()` on
  univariate Gaussian `mu`;
- three q1 sigma one-slope rows -- `phylo()`, `animal()`, `relmat()` on
  univariate Gaussian `sigma`;
- two slope-only q2 mean-mean rows -- `phylo(0 + x | ...)` and
  `relmat(0 + x | ...)` on bivariate Gaussian `mu1`/`mu2`, a joint
  cross-response covariance block. The intercept-only bivariate q2 rows are
  not inference-ready. The univariate
  Gaussian single-response q2 mean-slope route is recovery-only, not
  inference-ready;
- three Arc 1a REML rows -- `spatial()`, `animal()`, and `relmat()` on
  univariate Gaussian `mu`, restricted to an unlabelled intercept or an
  independent intercept plus one numeric slope with `sigma ~ 1`. Each provider
  is one ledger cell that covers both admitted shapes; the cell count is not a
  count of formula shapes.

The machine ledger contains thirteen rows for these eleven conceptual cells
because each of the two bivariate slope-only q2 provider blocks has separate
`mu1` and `mu2` endpoint rows. That storage detail does not create four distinct
scientific covariance blocks.

None of these eleven is free of a caveat, and the caveat is different for
each group. The three q1 `mu` intercept rows are backed by the default
location-axis bias-corrected, small-sample-t Wald channel, with coverage of
0.9705-0.9832 in the retained-denominator campaigns. The three q1
`sigma` one-slope rows are backed by raw, uncorrected log-SD Wald-z intervals:
intercept-SD coverage is 0.9388-0.9633 and slope-SD coverage is
0.9895-0.9957 at `g = 8`, with material miss asymmetry for some intercept
targets. Profile intervals for these q1 `sigma` rows are diagnostic-only at
`g = 8` because their finite-interval rates do not clear the promotion gate.
These are the documented small-sample limitations; do not replace either
evidence channel with a general profile recommendation.

The three Arc 1a REML cells are coverage-backed only over discrete campaign
domains. Here `M` is the number of structured levels (sites, animals, or
relatedness levels, and therefore the structured matrix dimension), while
`n_each` is the number of observations per structured level. Spatial and
`relmat()` use `n_each = 20` and exactly `M = {8, 16, 32}`, while the
`animal(A = A)` campaign uses `n_each = 20` and one fixed `M = 8` matrix.
Coverage clears the pre-specified small-sample floors but is not nominal-exact;
upper-tail miss asymmetry and zero-lower-bound slope profiles remain material.
Do not rewrite these as continuous `M >= ...` claims or treat other pedigrees,
matrices, or sample sizes as coverage-checked.

These are copy-paste forms of the three admitted independent one-slope REML
cells (replace the object names with objects from your analysis):

```r
fit_spatial_reml <- drmTMB(
  bf(y ~ x + spatial(1 + x | site, coords = coords), sigma ~ 1),
  family = gaussian(), data = dat, REML = TRUE
)
fit_animal_reml <- drmTMB(
  bf(y ~ x + animal(1 + x | id, A = A), sigma ~ 1),
  family = gaussian(), data = dat, REML = TRUE
)
fit_relmat_reml <- drmTMB(
  bf(y ~ x + relmat(1 + x | id, K = K), sigma ~ 1),
  family = gaussian(), data = dat, REML = TRUE
)
```

The multi-seed campaigns used exactly the coordinate, `A`, and `K`
representations shown above. Pedigree and `Ainv` animal inputs and relmat `Q`
have deterministic representation-parity evidence only; they do not inherit
the multi-seed campaign claim. Intercept-only versions replace `1 + x` with
`1`; slope-only, labelled, and multiple-slope shapes remain rejected.

For every admitted Arc 1a term, the fitted structured SD scale `s_j` is the
latent-field scale: the covariance is `s_j^2 K_h`, and node `i` has marginal
SD `s_j sqrt(K_h[ii])`. The fitted `s_j` is therefore equal to a node marginal
SD only when the corresponding diagonal entry of `K_h` is one.

The two bivariate-Gaussian slope-only q2 mean-mean rows are inference-ready only
because `confint()`'s defaults already correct for known small-sample bias
on location-axis structured-SD targets: `bias_correct = "location"` shifts the log-scale
point estimate to counter ML shrinkage, and `small_sample_df = "location"`
widens the interval with a `t(df = g - 1)` reference instead of a normal
quantile. Calling plain `confint(fit)` gives you this corrected interval
automatically. If you explicitly turn the correction off
(`bias_correct = "none"`, `small_sample_df = "none"`), the resulting raw
Wald interval under-covers on these two rows -- do not do that for a
bivariate `phylo()`/`relmat()` q2 mean-mean report.

None of these eleven structured cells carries the "supported" `fit_status`
the census reserves for the package's ordinary (unstructured) Gaussian and
bivariate-Gaussian random effects -- and even those supported cells are
point-trustworthy with interval coverage still planned, not
coverage-verified. Report these structured intervals as inference-ready, not
as a fully mature fixed-effect-grade interval.

### Ordinary (unstructured) random effects: point-trustworthy, coverage pending

One group of cells deserves explicit mention even though its interval does
not yet meet the coverage bar for this tier. Ordinary random intercepts and
slopes on Gaussian `mu` -- the everyday `(1 | id)` and `(0 + x | id)` mixed
model -- and their bivariate-Gaussian `mu1`/`mu2` counterpart are the only
cells the capability census marks `fit_status = supported`, the package's
highest fit maturity, so their point estimates are the most trustworthy
random-effect estimates `drmTMB` produces. Their interval coverage, however,
is still planned rather than simulation-verified: until that coverage
campaign lands, read any interval on these rows the way you would a
recovery-only row below, and report the point estimate with confidence.

### Unstructured non-Gaussian fixed effects

A 2026-07-09 coverage campaign
(`docs/dev-log/simulation-artifacts/2026-07-09-nongaussian-unstructured-coverage-pilot/`)
checked the fixed-effect mean coefficients of `binomial()`, `poisson()`,
`beta()`, and `nbinom2()` across `n` in `{50, 150, 500}`, 400 seeds per
cell. Every cell cleared the bar: finite rate at 1.00 and Wald coverage
between 0.922 and 0.973, with no worsening at the smallest sample size and
no systematic under-coverage. That holds even under stress -- a rare-event
binomial design with an approximately 8% base rate, and a low-count
Poisson design with a mean near 1 -- so a scarce or noisy field dataset is
not, by itself, a reason to distrust these intervals.

The same campaign extended to `nbinom2()` location-scale models, where
both the mean coefficients and the `sigma ~ x` dispersion coefficients are
calibrated (coverage 0.93-0.97 across `n`), and to `beta()` location-scale
models on interior proportions, where mean and `sigma ~ x` coefficients are
also calibrated (coverage 0.93-0.95). The `beta()` result carries one hard
requirement: the family strictly needs responses in the open interval
`(0, 1)`. An observation with an exact 0 or exact 1 -- a boundary
proportion that arises naturally from rounding at extreme covariate
values -- produces a non-finite result by design, not a bug to route
around. If your proportions can legitimately sit at the boundary, fit
`zero_one_beta()` instead of `beta()`.

### Ordinary non-Gaussian random slopes: profile intervals with caveats

Five non-Gaussian families now carry an inference-ready-with-caveats interval
for the standard deviation of a single ordinary independent `mu` random slope,
`(0 + x | id)`: binomial, cumulative-logit, skew-normal, Tweedie, and
zero-one-beta. The claim is narrow and the same in shape across them -- a profile interval for
the *natural-scale* slope SD, ML only, no Wald-interval or point-bias claim --
but the tested floor and design differ by family and should not be generalized
to a single number. Skew-normal, Tweedie, and zero-one-beta are certified at
`M >= 16` groups (true slope SD 0.50) with the standard ML-Laplace profile.
Binomial is certified at `M >= 32` (true SD 0.6). Cumulative-logit is certified
at `M >= 40` (true SD 0.50) and reaches this tier through adaptive Gauss-Hermite
quadrature with a Cox-Reid restricted likelihood rather than the Laplace
approximation.

Zero-one-beta carries an extra caveat and is **generator-qualified**. Its
coverage campaign was designed with a fixed 15% structural boundary mass and
interior draws from a beta density, but the interior draws leaked rare
machine-exact ones -- 50, 87, and 193 of them across the `M = 16`, `32`, and
`64` replicate banks -- because the sampler's precision parameter pushes the
beta shape toward a boundary in the upper tail. The reported coverage (0.929,
0.940, 0.952 at `M = 16 / 32 / 64`) therefore describes *the generator as
executed*, not an exactly-15%-boundary design. The leak is not independent of
the estimand: affected replicates carry systematically larger slope-SD
estimates, so the effect of the defect under the intended generator is
unquantified and cannot be recovered from the retained campaign. One of the
three independent promotion reviewers withheld this cell on exactly that ground;
it promotes under the frozen two-withhold rule, and the caveat travels with it.
A strictly-interior-generator rerun is the named next gate before any
exactly-15%-boundary claim.

## Tier 2: recovery-only

A larger set of non-Gaussian structured random-effect routes has verified
point-estimate recovery but no coverage evidence at all. Trust the
coefficient; do not report the interval as calibrated.

This covers Poisson and nbinom2 q1 structured `mu` intercepts and
one-slopes for `phylo()`, `spatial()`, `animal()`, and `relmat()`, and
nbinom2 structured `sigma` for the same four providers. The nbinom2
structured-`sigma` route deserves a specific caution: through 0.3.x, a
`sigma ~ phylo()`/`spatial()`/`animal()`/`relmat()` formula silently
modified the *mean* predictor instead of the scale predictor, so a fitted
model reported a `sigma`-labelled standard deviation that was actually
changing `mu`. That mis-targeting is fixed in 0.4.0, and the route now
correctly modifies scale with verified point-fit recovery -- but intervals
and coverage remain out of scope, so this row stays recovery-only even
after the fix.

A further handful of row-specific slices sit at the same tier: beta `animal()`
structured effects on `mu` and on `sigma`, Student-t
`mu ~ spatial(1 + x | ...)`, and Gamma `mu ~ relmat()` (unlabelled intercept
and one-slope). Each of these fits and has point-recovery evidence; none has a
coverage study behind its interval. If your
scientific question turns on the width of a confidence/compatibility
interval rather than the sign and rough size of an effect, do not build
the argument on one of these rows yet.

## Tier 3: diagnostic-only

Ten exact structured routes currently have deterministic or single-smoke
fit/extractor evidence without a retained-denominator recovery ladder:
ordinal `mu ~ phylo()`, truncated-nbinom2 `hu ~ relmat()`, Poisson
`zi ~ spatial()`, fixed-`zi` Poisson `mu ~ spatial()`, Student-t
`nu ~ phylo(1 | ...)`, Student-t intercept-only `mu ~ spatial(1 | ...)`,
Poisson slope-only `mu ~ spatial(0 + x | ...)`, Poisson labelled-scalar
`mu ~ spatial()`, Poisson `mu ~ spatial(1 | ...) + (1 | id)`, and fixed-`zi`
NB2 `mu ~ spatial(1 | ...)`. These
routes establish that the parser, likelihood target, and extractors connect.
They do **not** establish point-estimate recovery. Use them for feasibility or
debugging only, and do not report an interval or coverage claim.

## Tier 4: not available or rejected by design

**Structured random effects outside the narrow gates above.** `drmTMB`
does not currently accept structured `phylo()`, `spatial()`, `animal()`,
or `relmat()` random effects for lognormal, skew-normal, Tweedie, or
zero-one-beta families. Gamma is the one partial exception: it accepts
`relmat()` on `mu` (recovery-only, listed under Tier 2 above) but rejects
`phylo()`, `spatial()`, and `animal()`. Try instead: model the same
structural dependence through a Gaussian route (for example a log- or
logit-transformed response under `gaussian()`), fall back to the ordinary
Poisson/nbinom2 structured routes described above for count data, or drop to
fixed effects for the family you need while structured support catches up.

**REML outside its Gaussian scope.** REML is implemented for Gaussian
models only, and even inside Gaussian models it accepts a specific
boundary rather than every structured shape. It accepts univariate
phylogenetic mean-side, scale-side, and matched q2 mean-and-scale blocks,
plus univariate spatial, animal, and `relmat()` structured effects on the
scale side, bivariate phylogenetic structured effects in every covariance
layout, and heteroscedastic `sigma ~ x` formulas together with ordinary
(non-phylogenetic) `sigma` random effects. Arc 1a additionally accepts a
pure-`mu`, univariate `spatial()`, `animal()`, or `relmat()` term as an
unlabelled intercept or an independent intercept plus one numeric slope, but
only with constant `sigma ~ 1` and no sigma random effect. Arc 1b-S1 accepts
matching labelled fixed-covariance
`spatial(1 | p | site, coords = coords)` intercepts in bivariate `mu1` and
`mu2`, with intercept-only `sigma1`, `sigma2`, and `rho12`, complete response
pairs, unit weights, no known `meta_V()` covariance, and no additional ordinary
random, direct-SD, or `corpair()` layer, at recovery-only grade. Arc 1b-S2R
admits the analogous location-only supplied-relatedness cell only when both
formulas contain the same labelled `relmat(1 | p | id, K = K)` intercept and
the named `K`, group ordering, and label match exactly. This relmat route has
`point_fit_recovery` evidence: report its two structured SDs and latent
relatedness correlation as point estimates, not calibrated intervals.

The relmat REML exception does not admit `Q = Q`, slopes, q4 or larger blocks,
scale-side terms, extra random effects, incomplete response pairs, non-unit
weights, nonconstant residual formulas, direct-SD models, `meta_V()`, or
`corpair()`. Animal-model bivariate REML and every non-Gaussian family also
remain rejected.
Try instead: fit the same model with `REML = FALSE` (maximum likelihood); ML
is the route to use for any model REML rejects.

**Higher-order covariance promotion.** `q4`, `q6`, `q8`, and `q12`
covariance-interval promotion, plus derived-correlation intervals more
broadly for non-Gaussian structured covariance, remain planned rather than
implemented. Try instead: use the fitted point estimate for the
covariance or correlation summary you need, and check
`profile_targets(fit)` for any direct target that already has an interval
route before assuming none exists.

## Missing data

Missing-data support is a separate axis from the tiers above: it is likelihood
based (missing responses are marginalised, missing predictors are modelled inside
the same likelihood), not multiple imputation and not a posterior. It is
validated against two single-source-of-truth inventories. A positive runtime
test reconciles all 18 fitted response routes with the response ledger;
predictor-family tests still require every family outside
`drm_missing_predictor_families()` to reject. Route-specific tests cover
unsupported response neighbours, so an unsupported request is a clear error,
never a silent wrong likelihood. The full worked walkthrough is in
`vignette("missing-data")`.

The response-missingness board below is generated from the same 18-route ledger
as the capability surface. It distinguishes code admission from completed
validation: a route receives a verified ✓ only at G3 recovery or above.

```{r capability-ledger-missing-response, echo = FALSE, results = "asis"}
cat(
  readLines("includes/capability-ledger-missing-response.md", warn = FALSE),
  sep = "\n"
)
```

Missing-predictor support is a different axis and is not managed by the new
response-missingness ledger:

| Response family | Missing predictor `mi()` (`predictor = "model"`) |
| --- | :---: |
| `gaussian()` | ✓ (broad predictor-model catalogue) |
| `binomial()`, `poisson()`, `nbinom2()`, `beta()` | ✓ (one binary predictor) |
| every other family | — (rejects) |

Two missing-predictor details:

- **Imputation-model catalogue.** When the *response* is a univariate Gaussian
  location model, the *missing predictor* can be modelled with a wide family set —
  Gaussian, Bernoulli/logit, ordered cumulative logit, unordered softmax, beta,
  zero-one beta, beta-binomial (known trials), Poisson, NB2, zero-truncated NB2,
  lognormal, Gamma, and fixed-power Tweedie. The non-Gaussian *responses* above
  currently model **one binary** (Bernoulli/logit) missing predictor each.
- **Structured / multivariate `mi()`.** One intercept-only structured predictor
  model (`phylo()`, `spatial()`, `animal()`, `relmat()`) is available for the
  Gaussian-response `mi()` route only. Multivariate/bivariate missing-predictor
  modelling is not available (that is `gllvmTMB`'s lane).

Not available anywhere yet, and rejected with a family-specific message:
non-binary missing predictors on non-Gaussian responses; multiple missing
predictors; `mi()` with random-effect, structured, or zero-inflated response
terms; response masking combined with `mi()` in the same fit; and EM, profile,
or REML missing-data engines. No fitted response route remains at G0 on the
generated board, but each G3 tick applies only to the route and effect structure
named in its evidence row.

## Full per-family capability map

The table below preserves the original whole-package view alongside the
missing-response board. It shows distributional parameters, fixed and random
effects, structured providers, REML, inference maturity, and both missing-data
axes. Its missing-response column is generated from the same 18-route ledger.

```{r capability-ledger-family-map, echo = FALSE, results = "asis"}
cat(
  readLines("includes/capability-ledger-family-map.md", warn = FALSE),
  sep = "\n"
)
```

## Choosing an interval method

`confint()`'s default method is Wald (`method = "wald"`), and for most
unstructured fixed-effect rows in Tier 1 that default is already calibrated.
Do not collapse the four structured-anchor evidence channels into one method:
q1 `mu` and the exact phylo/relmat slope-only q2 `mu1:x`/`mu2:x` SD rows use
the default location-axis bias-corrected, small-sample-t Wald channel; q1
`sigma` uses raw uncorrected log-SD Wald-z evidence, with profile
diagnostic-only at `g = 8`; and the Arc 1a REML cells use direct structured-SD
profile evidence only inside their tested discrete domains. Near a boundary,
`profile_targets(fit)` can identify a
direct target worth diagnosing, but that does not promote profile intervals
for a row whose ledger calls them diagnostic-only. When neither an admitted
Wald channel nor an admitted profile channel is available for a target,
`method = "bootstrap"` is the last-resort fallback.

For a structured `sigma` on count data where only the point estimate is
verified, prefer a better-tested alternative when the interval matters to
your conclusion: an ordinary `sigma ~ (1 | id)` random intercept, or a
fixed-effect `sigma ~ predictors` model, both of which carry stronger
evidence than the structured-`sigma` recovery-only routes above.

## Where the exhaustive detail lives

This page is a summary. The full, machine-readable capability census --
one row per family, distributional parameter, effect type, structure
provider, estimator, and evidence tier, with a source citation for every
row -- lives under `docs/dev-log/dashboard/capability-census/` in the
package repository. Read it when you need the exact test or simulation
artifact behind a specific cell rather than the tier-level summary given
here.
