---
title: "What drmTMB can and can't do"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What drmTMB can and can't do}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

If you are an ecology or evolution graduate student deciding whether
`drmTMB` fits your problem, the first question is not "can I fit this
model?" It is "which numbers from the fitted model can I report with a
straight face?" A model can converge, print coefficients, and still hand you
a confidence/compatibility interval that does not cover the truth at the
rate it claims to. This page separates what `drmTMB` fits from what
`drmTMB` has checked, cell by cell, so you know whether to trust a point
estimate, an interval, or neither before it goes in a manuscript.

Throughout, `mu` is the location term, `sigma` is the scale term,
family-specific parameters such as Student-t `nu` are shape terms, and
coscale means a residual-correlation parameter such as bivariate `rho12` --
the association left between two responses after their means and scales are
modelled. Those four words describe what a formula changes; they say
nothing about whether the resulting interval has been checked. That check
is what the rest of this page reports.

## Three tiers, defined once

Every model/effect combination in `drmTMB` sits in one of three tiers.

1. **Inference-ready.** The confidence/compatibility interval has been
   coverage-checked by simulation and holds, at least at an honest
   small-sample floor (see the caveats below). You can report the point
   estimate and the interval.
2. **Recovery-only.** The point estimate has been checked against a known
   simulated truth and recovers it. The interval has not been
   coverage-checked and should be treated as provisional: report the
   point estimate, and treat any interval on the same row as a plausible
   range rather than a calibrated one.
3. **Not available / rejected by design.** `drmTMB` refuses to fit the
   request outright. Some of these routes are planned for a later release;
   some are out of scope for the package's current design. Either way,
   fitting it anyway is not an option -- use the alternative named below
   each entry.

No cell in `drmTMB` is coverage-verified at the strictest bar the project
tracks internally as "supported" (nominal-exact coverage with symmetric
misses on both sides). The census does mark four cells `fit_status =
supported` -- ordinary (unstructured) Gaussian and bivariate-Gaussian random
effects -- but those are point-trustworthy with interval coverage still
planned, not coverage-verified. Read "inference-ready" as an honest floor
for a first fitted structured route, not as parity with a long-mature
fixed-effect interval in a package such as `glmmTMB` or `lme4`.

## At a glance

| Model / effect | Tier | What you can trust |
| --- | --- | --- |
| Ordinary (unstructured) Gaussian / bivariate-Gaussian random effects (intercepts, slopes) | Point-trustworthy (census "supported", coverage planned) | Point estimate -- the everyday random-intercept/slope mixed model; its interval is not yet coverage-verified |
| Gaussian `mu` q1 intercept -- `phylo()`, `spatial()`, `relmat()` | Inference-ready | Point estimate and interval; prefer `method = "profile"` |
| Gaussian `sigma` q1 one-slope -- `phylo()`, `animal()`, `relmat()` | Inference-ready | Point estimate and interval; documented mild under-coverage, prefer `method = "profile"` |
| Bivariate Gaussian `mu1`/`mu2` q2 mean-mean -- `phylo()`, `relmat()` | Inference-ready (default-corrected channel only) | Point estimate and interval from plain `confint(fit)`; an explicitly uncorrected Wald interval under-covers |
| binomial / Poisson / beta / nbinom2 unstructured `mu` | Inference-ready | Point estimate and Wald interval from `n` &gt;= 50 |
| nbinom2 unstructured location-scale (`mu` and `sigma ~ x`) | Inference-ready | Point estimate and Wald interval on both formulas |
| beta unstructured location-scale, interior proportions | Inference-ready | Point estimate and Wald interval; exact 0 or 1 needs `zero_one_beta()` instead |
| Poisson / nbinom2 structured `mu` q1 -- `phylo()`, `spatial()`, `animal()`, `relmat()` | Recovery-only | Point estimate only |
| nbinom2 structured `sigma` -- `phylo()`, `spatial()`, `animal()`, `relmat()` | Recovery-only | Point estimate only; scale-targeting bug fixed in 0.4.0 |
| A handful of row-specific slices: Student-t `nu ~ phylo()`, ordinal `mu ~ phylo()`, beta `animal()` on `mu`/`sigma`, Student-t `mu ~ spatial()`, truncated-nbinom2 `hu ~ relmat()`, Gamma `mu ~ relmat()` | Recovery-only | Point estimate only |
| Structured effects for lognormal, skew-normal, Tweedie, zero-one-beta | Not available | Use a Gaussian or basic count/proportion structured route, or fixed effects |
| REML for non-Gaussian families; non-phylogenetic mean-side or any bivariate non-phylogenetic structured REML | Not available (rejected by design) | Use `REML = FALSE` (maximum likelihood) |
| `q4`/`q6`/`q8`/`q12` covariance interval promotion; derived-correlation intervals | Not available (planned) | Use the fitted point estimate; try `profile_targets()` for a direct target |

The rest of this page expands each row with the caveat that changes how you
should read it.

## Tier 1: inference-ready

### Gaussian structured random effects: eight anchor cells

Eight cells across `drmTMB`'s structured Gaussian-family random-effect
routes are inference-ready today. Six sit on a univariate Gaussian `mu` or
`sigma` formula, all at q1; the remaining two are the separate
bivariate-Gaussian family (`mu1`/`mu2` q2 mean-mean blocks):

- three q1 mean-intercept rows -- `phylo()`, `spatial()`, `relmat()` on
  univariate Gaussian `mu`;
- three q1 sigma one-slope rows -- `phylo()`, `animal()`, `relmat()` on
  univariate Gaussian `sigma`;
- two q2 mean-mean rows -- `phylo()`, `relmat()` on bivariate Gaussian
  `mu1`/`mu2`, a joint cross-response covariance block. The univariate
  Gaussian single-response q2 mean-slope route is recovery-only, not
  inference-ready.

None of these eight is free of a caveat, and the caveat is different for
each group. The sigma one-slope rows carry a documented, mild residual
under-coverage: profiled interval coverage on the sigma standard-deviation
member lands around 0.90-0.92 at a typical group count of eight rather than
the nominal 0.95, which matches the expected small-sample floor for this
class of interval, not a defect to work around. Use `method = "profile"`
for these rows, and read a reported interval as honest-but-conservative
rather than nominal-exact.

The two bivariate-Gaussian q2 mean-mean rows are inference-ready only
because `confint()`'s defaults already correct for known small-sample bias
on location-axis structured-SD targets: `bias_correct = "location"` shifts the log-scale
point estimate to counter ML shrinkage, and `small_sample_df = "location"`
widens the interval with a `t(df = g - 1)` reference instead of a normal
quantile. Calling plain `confint(fit)` gives you this corrected interval
automatically. If you explicitly turn the correction off
(`bias_correct = "none"`, `small_sample_df = "none"`), the resulting raw
Wald interval under-covers on these two rows -- do not do that for a
bivariate `phylo()`/`relmat()` q2 mean-mean report.

None of these eight structured cells carries the "supported" `fit_status`
the census reserves for the package's ordinary (unstructured) Gaussian and
bivariate-Gaussian random effects -- and even those supported cells are
point-trustworthy with interval coverage still planned, not
coverage-verified. Report these structured intervals as inference-ready, not
as a fully mature fixed-effect-grade interval.

### Ordinary (unstructured) random effects: point-trustworthy, coverage pending

One group of cells deserves explicit mention even though its interval does
not yet meet the coverage bar for this tier. Ordinary random intercepts and
slopes on Gaussian `mu` -- the everyday `(1 | id)` and `(0 + x | id)` mixed
model -- and their bivariate-Gaussian `mu1`/`mu2` counterpart are the only
cells the capability census marks `fit_status = supported`, the package's
highest fit maturity, so their point estimates are the most trustworthy
random-effect estimates `drmTMB` produces. Their interval coverage, however,
is still planned rather than simulation-verified: until that coverage
campaign lands, read any interval on these rows the way you would a
recovery-only row below, and report the point estimate with confidence.

### Unstructured non-Gaussian fixed effects

A 2026-07-09 coverage campaign
(`docs/dev-log/simulation-artifacts/2026-07-09-nongaussian-unstructured-coverage-pilot/`)
checked the fixed-effect mean coefficients of `binomial()`, `poisson()`,
`beta()`, and `nbinom2()` across `n` in `{50, 150, 500}`, 400 seeds per
cell. Every cell cleared the bar: finite rate at 1.00 and Wald coverage
between 0.922 and 0.973, with no worsening at the smallest sample size and
no systematic under-coverage. That holds even under stress -- a rare-event
binomial design with an approximately 8% base rate, and a low-count
Poisson design with a mean near 1 -- so a scarce or noisy field dataset is
not, by itself, a reason to distrust these intervals.

The same campaign extended to `nbinom2()` location-scale models, where
both the mean coefficients and the `sigma ~ x` dispersion coefficients are
calibrated (coverage 0.93-0.97 across `n`), and to `beta()` location-scale
models on interior proportions, where mean and `sigma ~ x` coefficients are
also calibrated (coverage 0.93-0.95). The `beta()` result carries one hard
requirement: the family strictly needs responses in the open interval
`(0, 1)`. An observation with an exact 0 or exact 1 -- a boundary
proportion that arises naturally from rounding at extreme covariate
values -- produces a non-finite result by design, not a bug to route
around. If your proportions can legitimately sit at the boundary, fit
`zero_one_beta()` instead of `beta()`.

## Tier 2: recovery-only

A larger set of non-Gaussian structured random-effect routes has verified
point-estimate recovery but no coverage evidence at all. Trust the
coefficient; do not report the interval as calibrated.

This covers Poisson and nbinom2 q1 structured `mu` intercepts and
one-slopes for `phylo()`, `spatial()`, `animal()`, and `relmat()`, and
nbinom2 structured `sigma` for the same four providers. The nbinom2
structured-`sigma` route deserves a specific caution: through 0.3.x, a
`sigma ~ phylo()`/`spatial()`/`animal()`/`relmat()` formula silently
modified the *mean* predictor instead of the scale predictor, so a fitted
model reported a `sigma`-labelled standard deviation that was actually
changing `mu`. That mis-targeting is fixed in 0.4.0, and the route now
correctly modifies scale with verified point-fit recovery -- but intervals
and coverage remain out of scope, so this row stays recovery-only even
after the fix.

A further handful of row-specific slices sit at the same tier: Student-t
`nu ~ phylo()`, ordinal `mu ~ phylo()`, beta `animal()` structured effects
on `mu` and on `sigma`, Student-t `mu ~ spatial()`, truncated-nbinom2
`hu ~ relmat()`, and Gamma `mu ~ relmat()` (unlabelled intercept and
one-slope). Each of these fits and recovers its point estimate in a
recovery study; none has a coverage study behind its interval. If your
scientific question turns on the width of a confidence/compatibility
interval rather than the sign and rough size of an effect, do not build
the argument on one of these rows yet.

## Tier 3: not available or rejected by design

**Structured random effects outside the narrow gates above.** `drmTMB`
does not currently accept structured `phylo()`, `spatial()`, `animal()`,
or `relmat()` random effects for lognormal, skew-normal, Tweedie, or
zero-one-beta families. Gamma is the one partial exception: it accepts
`relmat()` on `mu` (recovery-only, listed under Tier 2 above) but rejects
`phylo()`, `spatial()`, and `animal()`. Try instead: model the same
structural dependence through a Gaussian route (for example a log- or
logit-transformed response under `gaussian()`), fall back to the ordinary
Poisson/nbinom2 structured routes described above for count data, or drop to
fixed effects for the family you need while structured support catches up.

**REML outside its Gaussian scope.** REML is implemented for Gaussian
models only, and even inside Gaussian models it accepts a specific
boundary rather than every structured shape. It accepts univariate
phylogenetic mean-side, scale-side, and matched q2 mean-and-scale blocks,
plus univariate spatial, animal, and `relmat()` structured effects on the
scale side, bivariate phylogenetic structured effects in every covariance
layout, and heteroscedastic `sigma ~ x` formulas together with ordinary
(non-phylogenetic) `sigma` random effects. It rejects non-phylogenetic
mean-side structured effects, every bivariate non-phylogenetic structured
route, and every non-Gaussian family outright. Try instead: fit the same
model with `REML = FALSE` (maximum likelihood); ML is the route to use for
any model REML rejects.

**Higher-order covariance promotion.** `q4`, `q6`, `q8`, and `q12`
covariance-interval promotion, plus derived-correlation intervals more
broadly for non-Gaussian structured covariance, remain planned rather than
implemented. Try instead: use the fitted point estimate for the
covariance or correlation summary you need, and check
`profile_targets(fit)` for any direct target that already has an interval
route before assuming none exists.

## Missing data

Missing-data support is a separate axis from the tiers above: it is likelihood
based (missing responses are marginalised, missing predictors are modelled inside
the same likelihood), not multiple imputation and not a posterior. It is
validated against two single-source-of-truth inventories. A positive runtime
test reconciles all 18 fitted response routes with the response ledger;
predictor-family tests still require every family outside
`drm_missing_predictor_families()` to reject. Route-specific tests cover
unsupported response neighbours, so an unsupported request is a clear error,
never a silent wrong likelihood. The full worked walkthrough is in
`vignette("missing-data")`.

The response-missingness board below is generated from the same 18-route ledger
as the capability surface. It distinguishes code admission from completed
validation: a route receives a verified ✓ only at G3 recovery or above.

```{r capability-ledger-missing-response, echo = FALSE, results = "asis"}
cat(
  readLines("includes/capability-ledger-missing-response.md", warn = FALSE),
  sep = "\n"
)
```

Missing-predictor support is a different axis and is not managed by the new
response-missingness ledger:

| Response family | Missing predictor `mi()` (`predictor = "model"`) |
| --- | :---: |
| `gaussian()` | ✓ (broad predictor-model catalogue) |
| `binomial()`, `poisson()`, `nbinom2()`, `beta()` | ✓ (one binary predictor) |
| every other family | — (rejects) |

Two missing-predictor details:

- **Imputation-model catalogue.** When the *response* is a univariate Gaussian
  location model, the *missing predictor* can be modelled with a wide family set —
  Gaussian, Bernoulli/logit, ordered cumulative logit, unordered softmax, beta,
  zero-one beta, beta-binomial (known trials), Poisson, NB2, zero-truncated NB2,
  lognormal, Gamma, and fixed-power Tweedie. The non-Gaussian *responses* above
  currently model **one binary** (Bernoulli/logit) missing predictor each.
- **Structured / multivariate `mi()`.** One intercept-only structured predictor
  model (`phylo()`, `spatial()`, `animal()`, `relmat()`) is available for the
  Gaussian-response `mi()` route only. Multivariate/bivariate missing-predictor
  modelling is not available (that is `gllvmTMB`'s lane).

Not available anywhere yet, and rejected with a family-specific message:
non-binary missing predictors on non-Gaussian responses; multiple missing
predictors; `mi()` with random-effect, structured, or zero-inflated response
terms; response masking combined with `mi()` in the same fit; and EM, profile,
or REML missing-data engines. No fitted response route remains at G0 on the
generated board, but each G3 tick applies only to the route and effect structure
named in its evidence row.

## Full per-family capability map

The table below preserves the original whole-package view alongside the
missing-response board. It shows distributional parameters, fixed and random
effects, structured providers, REML, inference maturity, and both missing-data
axes. Its missing-response column is generated from the same 18-route ledger.

```{r capability-ledger-family-map, echo = FALSE, results = "asis"}
cat(
  readLines("includes/capability-ledger-family-map.md", warn = FALSE),
  sep = "\n"
)
```

## Choosing an interval method

`confint()`'s default method is Wald (`method = "wald"`), and for most
unstructured fixed-effect rows in Tier 1 that default is already
calibrated. Two situations call for a different method. Near a boundary --
a variance component pinned close to zero, a correlation close to `+-1`,
or any row `check_drm()` or `conf.status` flags as boundary-adjacent --
prefer `method = "profile"`. At small group counts, particularly for the
structured-RE anchor cells above, `method = "profile"` is also the
better-behaved choice; `confint()`'s Wald defaults already apply a
small-sample correction (`bias_correct`, `small_sample_df`) for the
location-axis structured targets that need it, but profiling remains the
more robust fallback when a target is listed by `profile_targets(fit)`.
When neither Wald nor profile is available for a target, `method =
"bootstrap"` is the last-resort fallback.

For a structured `sigma` on count data where only the point estimate is
verified, prefer a better-tested alternative when the interval matters to
your conclusion: an ordinary `sigma ~ (1 | id)` random intercept, or a
fixed-effect `sigma ~ predictors` model, both of which carry stronger
evidence than the structured-`sigma` recovery-only routes above.

## Where the exhaustive detail lives

This page is a summary. The full, machine-readable capability census --
one row per family, distributional parameter, effect type, structure
provider, estimator, and evidence tier, with a source citation for every
row -- lives under `docs/dev-log/dashboard/capability-census/` in the
package repository. Read it when you need the exact test or simulation
artifact behind a specific cell rather than the tier-level summary given
here.
