---
title: "Comparing drmTMB with other packages"
description: "Eight fitted comparisons against lme4, metafor, ordinal, and glmmTMB, grouped by how independent the comparator is."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing drmTMB with other packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
has_metafor <- requireNamespace("metafor", quietly = TRUE)
has_metadat <- requireNamespace("metadat", quietly = TRUE)
has_ordinal <- requireNamespace("ordinal", quietly = TRUE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
has_penguins <- requireNamespace("palmerpenguins", quietly = TRUE)
```

```{r}
library(drmTMB)
```

If you already fit mixed models in `lme4`, meta-analyses in `metafor`,
proportional-odds models in `ordinal`, or location-scale models in `glmmTMB`,
this article answers one question in each of those packages' own vocabulary:
on a dataset you may already know, does `drmTMB` land on the same fit?

Eight comparisons follow, each built from a real dataset, a real `drmTMB` fit,
and a real fit from the comparator package alongside it. **These eight models
were chosen because a package this article already depends on fits them too;
that set is small next to what `drmTMB` implements as a whole, and the
package's design documents record the rest of that surface.**

Every comparison here is single-seed and single-dataset. Each one shows that
`drmTMB`'s likelihood and optimizer reach the same optimum as another
implementation of the same model, on the data shown. It is **not** evidence
about interval calibration, coverage, bias, or small-sample behaviour, and no
standard error or confidence interval is compared anywhere in this article.
Agreement licenses the model both packages fit on the dataset shown; for any
other model, `drmTMB`'s evidence is its own recovery and coverage studies,
recorded per capability in the package's design documents and capability
ledger.

## What this article compares

`drmTMB`'s location submodel is compared in every one of the eight fits
below. Its distributional submodel — the `sigma` formula that lets residual
spread depend on covariates — is compared in some of them, and its
random-effect form is not compared here. The table states only what this
article did, comparison by comparison.

| `drmTMB` capability | Comparisons | Compared in this article? | Independence |
| --- | --- | --- | --- |
| Location (`mu`) submodel, fixed effects | 1–8 (all) | yes, eight times | 1–3 STRONG · 4–5 unclassified · 6–8 WEAK |
| Location (`mu`) submodel, ordinary random intercept / slope | 1, 5, 6, 8 | yes, four times | 1 STRONG · 5 unclassified · 6, 8 WEAK |
| Distributional (`sigma`) submodel, fixed effects | 2, 3, 6, 7, 8 | yes, five times | 2, 3 STRONG (a between-study heterogeneity SD, not a residual scale; 2's `sigma` formula is intercept-only, 3's carries a covariate) · 6, 7, 8 WEAK |
| Distributional (`sigma`) submodel, ordinary random effect | none | not compared here | nothing to classify |
| Residual correlation `rho12` with a predictor | none | not compared here | nothing to classify |
| The rest of the implemented model surface | none | not compared here; the design documents record what it contains | nothing to classify |

Comparisons 6, 7, and 8 fit `mu` and `sigma` jointly in one model. The
agreement reported for each is a property of that joint fit; it does not
license either submodel taken on its own.

Three independence labels appear throughout: **STRONG** means the comparator
shares no estimation code with `drmTMB`; **WEAK** means it is built on the
same TMB automatic-differentiation stack and outer optimizer, so agreement is
a consistency check between related implementations rather than a
cross-implementation one; **unclassified** means the comparator is a separate
engine whose independence `drmTMB`'s design documents have not yet
classified.

## Checked against a separate estimation engine

The three comparisons in this section are against `lme4` and `metafor`,
which share no estimation code with `drmTMB`
(`docs/design/242-external-comparator-evidence-class.md`). Agreement here is
a genuine cross-implementation check.

### Comparison 1: contagious bovine pleuropneumonia across periods and herds

If you already fit `lme4::glmer()` on `lme4::cbpp` while learning mixed
models, this is the same model in `drmTMB`: infection risk by observation
period, with a random herd intercept.

```{r comparison1-drmtmb, eval = has_lme4}
data(cbpp, package = "lme4")
fit1 <- drmTMB(
  bf(mu = cbind(incidence, size - incidence) ~ period + (1 | herd)),
  data = cbpp, family = binomial()
)
coef(fit1, "mu")
```

```{r comparison1-comparator, eval = has_lme4}
cmp1 <- lme4::glmer(
  cbind(incidence, size - incidence) ~ period + (1 | herd),
  family = binomial, data = cbpp
)
lme4::fixef(cmp1)
```

Both sides fit the same logit-scale coefficients directly — no conversion —
and the herd random-intercept SD compares to `lme4`'s the same way:

```{r comparison1-re, eval = has_lme4}
summary(fit1)$parameters[, c("term", "estimate")]
attr(lme4::VarCorr(cmp1)$herd, "stddev")
```

```{r comparison1-diff, eval = has_lme4}
coef(fit1, "mu") - lme4::fixef(cmp1)
c(drmTMB = as.numeric(logLik(fit1)), glmer = as.numeric(logLik(cmp1)))
```

The two fits agree to about three decimal places on the coefficients and the
log-likelihood, not to five or six. Both sides use a first-order
(`nAGQ = 1`) Laplace approximation of the marginal likelihood; a gap of this
size between two different Laplace solvers — TMB's automatic-differentiation
inner solve here, `lme4`'s PIRLS solve there — is the expected size for that
comparison, not a sign that either optimizer stopped early.

`binomial()` here has no dispersion parameter, on either side. `cbpp`'s known
extra-binomial variation is therefore unmodelled symmetrically, not a
discrepancy between the packages; `beta_binomial()` adds a dispersion
parameter to this same two-column response if that matters for your own data.
(`nbinom2()` also carries a dispersion parameter, but it is a count family and
takes a single-column count response, so it is not a substitute here.)

**Independence: STRONG** (`lme4`).

### Comparison 2: the pooled BCG vaccine effect

If you run meta-analyses in `metafor`, this is a fixed pooled effect with a
between-trial heterogeneity term, written as `drmTMB` gaussian regression
with the sampling variance supplied through `meta_V()`.

```{r comparison2-data, eval = has_metafor && has_metadat}
data(dat.bcg, package = "metadat")
dat_bcg <- metafor::escalc(
  measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data = dat.bcg
)
```

```{r comparison2-drmtmb, eval = has_metafor && has_metadat}
fit2 <- drmTMB(
  bf(mu = yi ~ 1 + meta_V(V = vi), sigma = ~ 1),
  data = dat_bcg, family = gaussian()
)
coef(fit2, "mu")
```

`metafor::rma.uni()`'s default estimator is REML, and it gives a visibly
different fit:

```{r comparison2-reml, eval = has_metafor && has_metadat}
cmp2_reml <- metafor::rma.uni(yi, vi, data = dat_bcg)
c(coef(cmp2_reml), tau2 = cmp2_reml$tau2)
```

`method = "ML"` is the comparator that matches:

```{r comparison2-ml, eval = has_metafor && has_metadat}
cmp2 <- metafor::rma.uni(yi, vi, data = dat_bcg, method = "ML")
c(coef(cmp2), tau2 = cmp2$tau2)
```

`drmTMB` reports `log(tau)` as the `sigma` intercept; `metafor` reports
`tau^2` directly. Converting with `tau^2 = exp(2 * coef(fit, "sigma"))` and
comparing the two log-likelihoods shows the two packages are maximising the
same function, not merely landing close to each other:

```{r comparison2-convert, eval = has_metafor && has_metadat}
c(drmTMB_tau2 = exp(2 * coef(fit2, "sigma")), metafor_tau2 = cmp2$tau2)
c(drmTMB = as.numeric(logLik(fit2)), metafor = as.numeric(logLik(cmp2)))
```

BCG vaccination reduces tuberculosis risk on average, with real heterogeneity
between the 13 trials, and `drmTMB` reproduces the estimate a `metafor` user
already trusts before being asked to trust anything new.

**Independence: STRONG** (`metafor`,
`docs/design/242-external-comparator-evidence-class.md`).

### Comparison 3: does heterogeneity itself differ by allocation design?

The 13 BCG trials were allocated to treatment in three different ways
(`alternate`, `random`, `systematic`). Rather than splitting the data into
three separate meta-analyses, one `drmTMB` fit gives each allocation type its
own heterogeneity `tau` through `sigma ~ alloc`.

```{r comparison3-drmtmb, eval = has_metafor && has_metadat}
fit3 <- drmTMB(
  bf(mu = yi ~ 1 + meta_V(V = vi), sigma = ~ alloc),
  data = dat_bcg, family = gaussian()
)
coef(fit3, "sigma")
```

```{r comparison3-comparator, eval = has_metafor && has_metadat}
dat_bcg$id <- seq_len(nrow(dat_bcg))
cmp3 <- metafor::rma.mv(
  yi, vi, random = ~ alloc | id, struct = "DIAG", data = dat_bcg, method = "ML"
)
cmp3$tau2
```

`drmTMB`'s `sigma` linear predictor is on the `log(tau)` scale, and each
allocation level's value is the intercept plus that level's contrast, with
`alternate` as the reference on both sides. Exponentiating gives per-level
`tau`, matching `rma.mv()`'s response-scale `tau`:

```{r comparison3-convert, eval = has_metafor && has_metadat}
sigma3 <- coef(fit3, "sigma")
tau_by_level <- exp(c(
  alternate = sigma3[1],
  random = sigma3[1] + sigma3[2],
  systematic = sigma3[1] + sigma3[3]
))
tau_by_level
sqrt(cmp3$tau2)
c(drmTMB = as.numeric(logLik(fit3)), metafor = as.numeric(logLik(cmp3)))
```

The two fits' per-level `tau` values and log-likelihood agree closely, and
that agreement is not a coincidence: with one row per study, `rma.mv()`'s
`~ alloc | id` under `struct = "DIAG"` puts one random effect per study with
variance `tau^2` for that study's allocation level, so the marginal variance
is `v_i + tau^2_{alloc(i)}` on both sides. These are the same model, written
in two packages' syntax, not two models that happen to agree.

Read the three fitted `tau` values as a point-agreement result only, not as a
finding about which allocation design is more heterogeneous: the
`alternate` level carries two studies, `random` seven, and `systematic` four,
and `drmTMB`'s own recorded evidence for this route says heterogeneity
intervals are not usable at this few studies per level. Report the estimates
this comparison licenses — that the two packages reach the same optimum — and
no more.

**Independence: STRONG.** This and Comparison 2 are the two comparisons in
this article that check a `sigma` linear predictor against a separate
estimation engine — Comparison 2's `sigma` formula is intercept-only, and
this one gives it a covariate. In both, `sigma` is a between-study
heterogeneity SD under `meta_V()`, not a residual scale.

## Checked against `ordinal`

`ordinal` is a separate estimation engine, but
`docs/design/242-external-comparator-evidence-class.md` has not yet
classified its independence strength the way it classifies `lme4`,
`metafor`, and `glmmTMB`, so the two comparisons in this section are
reported without one.

### Comparison 4: does serving temperature shift wine bitterness ratings?

If you fit proportional-odds models with `ordinal::clm()`, this is the same
grammar in `cumulative_logit()`: five ordered bitterness ratings, from
Randall (1989)'s wine-tasting data, as a function of serving temperature and
skin contact.

```{r comparison4-drmtmb, eval = has_ordinal}
data(wine, package = "ordinal")
fit4 <- drmTMB(
  bf(mu = rating ~ temp + contact), data = wine, family = cumulative_logit()
)
summary(fit4)$coefficients
summary(fit4)$ordinal$cutpoints
```

```{r comparison4-comparator, eval = has_ordinal}
cmp4 <- ordinal::clm(rating ~ temp + contact, data = wine)
coef(cmp4)
```

```{r comparison4-loglik, eval = has_ordinal}
c(drmTMB = as.numeric(logLik(fit4)), clm = as.numeric(logLik(cmp4)))
```

No conversion is needed: both sides write the same cumulative-logit
convention, `logit P(Y <= j) = alpha_j - x'beta`, and the two log-likelihoods
agree to five decimal places with matching-sign coefficients. Warm serving
and skin contact both push ratings toward "more bitter."

The comparison that makes this pair valuable runs the other way. `ordinal`
can add a scale formula to this same model, and it fits:

```{r comparison4-scale-comparator, eval = has_ordinal}
cmp4_scale <- ordinal::clm(rating ~ temp + contact, scale = ~ temp, data = wine)
as.numeric(logLik(cmp4_scale))
```

`cumulative_logit()` does not accept the equivalent formula in `drmTMB` 0.7.0:

```{r comparison4-drmtmb-scale, eval = has_ordinal, error = TRUE}
drmTMB(
  bf(mu = rating ~ temp + contact, sigma = ~ temp),
  data = wine, family = cumulative_logit()
)
```

`cumulative_logit()` currently fits only a `mu` location formula; a scale or
discrimination formula for this family is not yet implemented. If you need
that, `ordinal::clm(scale = ~ ...)` is the tool that already has it. This
asymmetry is worth sitting with: the two packages' surfaces do not coincide,
and they fail to coincide in the direction that costs `drmTMB` something,
which is the direction worth believing.

**Independence:** `ordinal` is not yet classified in `drmTMB`'s design
documents.

### Comparison 5: do judges differ, and does that change the temperature effect?

Nine judges each rated eight bottles. Adding a judge random intercept asks
whether that matters, and whether `drmTMB` still matches `ordinal::clmm()`
once it is added.

```{r comparison5-drmtmb, eval = has_ordinal}
fit5 <- drmTMB(
  bf(mu = rating ~ temp + contact + (1 | judge)),
  data = wine, family = cumulative_logit()
)
summary(fit5)$coefficients
summary(fit5)$ordinal$cutpoints
```

```{r comparison5-comparator, eval = has_ordinal}
cmp5 <- ordinal::clmm(rating ~ temp + contact + (1 | judge), data = wine)
coef(cmp5)
```

For the random effect, compare SD to SD, not `clmm`'s printed variance:

```{r comparison5-re, eval = has_ordinal}
summary(fit5)$parameters[, c("term", "estimate")]
attr(ordinal::VarCorr(cmp5)$judge, "stddev")
c(drmTMB = as.numeric(logLik(fit5)), clmm = as.numeric(logLik(cmp5)))
```

Every quantity here — slopes, cutpoints, the random-intercept SD, and the
log-likelihood — agrees closely between the two fits. Ignoring judge
identity attenuates the temperature effect: Comparison 4's marginal slope was
smaller than the conditional slope shown here, which is the expected
conditional-versus-marginal shift, and having a working comparator is what
makes it checkable rather than merely plausible.

This licenses point-fit parity only, on this dataset. It does not license any
REML or interval claim for `cumulative_logit()` random effects in general.
The capability-ledger cell behind the model fitted here is `mc-0225`
(`cumulative_logit`, `mu`, ordinary random intercept, ML), recorded at
`interval_feasible`. That tier is narrow: it establishes computational
interval feasibility for that one cell's direct intercept-SD target on a
frozen low-rung fixture, it was not evaluated for coverage or calibration,
and it carries no public interval guidance, no broader family or formula
support, and nothing about correlated or labelled effects. A random *slope*
on this family is a different cell (`mc-0227`, recorded a tier lower at
`point_fit_recovery`) and is not what this comparison fits.

**Independence:** `ordinal` is not yet classified in `drmTMB`'s design
documents.

## Checked against a package built on the same machinery

The three comparisons in this section are against `glmmTMB`, which is built
on the same TMB automatic-differentiation stack and outer optimizer as
`drmTMB` (`docs/design/242-external-comparator-evidence-class.md`).
Agreement here is a consistency check between related implementations, not a
cross-implementation check. It is also, uncomfortably, where `drmTMB`'s
stated differentiator lives: no separate estimation engine checked a
residual-scale submodel anywhere in this article.

```{r block2c-attach, eval = has_glmmTMB}
library(glmmTMB)
```

Attaching `glmmTMB` masks several names it shares with `drmTMB`, including
the family constructors `lognormal()` and `nbinom2()` used below. From this
point on, those two are called as `drmTMB::lognormal()` and
`drmTMB::nbinom2()` for that reason, not for style.

### Comparison 6: does reaction-time variability grow with sleep deprivation?

If you already fit `lme4::lmer(Reaction ~ Days + (Days | Subject))` on
`lme4::sleepstudy`, this asks a further question: does the spread of
reaction times widen with each day of deprivation, on top of the mean?

The scale conversion is checked first on a fixed-effect variant of the
model, where the likelihood is exact and no Laplace step intervenes on
either side:

```{r comparison6-fe-check, eval = has_lme4}
data(sleepstudy, package = "lme4")
fit6_fe <- drmTMB(
  bf(mu = Reaction ~ Days, sigma = ~ Days), data = sleepstudy, family = gaussian()
)
mu_hat <- predict_parameters(fit6_fe, dpar = "mu", type = "response")$estimate
eta_sigma <- predict_parameters(fit6_fe, dpar = "sigma", type = "link")$estimate
c(
  reported = as.numeric(logLik(fit6_fe)),
  `hand, sd = exp(eta)` = sum(dnorm(sleepstudy$Reaction, mu_hat, exp(eta_sigma), log = TRUE)),
  `hand, sd = sqrt(exp(eta))` = sum(dnorm(sleepstudy$Reaction, mu_hat, sqrt(exp(eta_sigma)), log = TRUE))
)
```

`drmTMB`'s `sigma` linear predictor is `log(SD)`; no transform, no squaring.
The wrong alternative misses `drmTMB`'s own reported log-likelihood by
thousands of units, so the link is structural, not something that happens to
hold for one fit.

On the displayed random-effect model, that same no-transform rule is the one
to use, and the two packages' `sigma` coefficients compare directly to
`glmmTMB`'s `dispformula` coefficients:

```{r comparison6-drmtmb, eval = has_lme4}
fit6 <- drmTMB(
  bf(mu = Reaction ~ Days + (1 + Days | Subject), sigma = ~ Days),
  data = sleepstudy, family = gaussian()
)
coef(fit6, "mu")
coef(fit6, "sigma")
```

```{r comparison6-comparator, eval = has_lme4 && has_glmmTMB}
cmp6 <- glmmTMB(
  Reaction ~ Days + (Days | Subject), dispformula = ~ Days, data = sleepstudy
)
fixef(cmp6)$cond
fixef(cmp6)$disp
```

With a random effect on `mu`, `logLik()` is a Laplace marginal on both sides,
so no row-by-row density identity holds the way it did above. What the two
packages share instead is the marginal log-likelihood itself:

```{r comparison6-loglik, eval = has_lme4 && has_glmmTMB}
c(drmTMB = as.numeric(logLik(fit6)), glmmTMB = as.numeric(logLik(cmp6)))
abs(coef(fit6, "sigma") - fixef(cmp6)$disp)
```

The `sigma` slope on `Days` is positive: residual spread grows with each day
of deprivation. A mean-only model cannot show that, because it carries a
single residual SD for every observation; this is an addition to the mean
model you already trust, not a replacement for it.

**Independence: WEAK.** The two fits are a consistency check between related
implementations, not an independent confirmation.

### Comparison 7: are male penguins more variable in body mass than females?

`palmerpenguins::penguins` lets one model ask two things at once: does mean
body mass differ by species, and is one sex more variable than the other?

```{r comparison7-drmtmb, eval = has_penguins && has_glmmTMB}
pen <- palmerpenguins::penguins[
  stats::complete.cases(
    palmerpenguins::penguins[, c("species", "sex", "body_mass_g")]
  ),
]
fit7 <- drmTMB(
  bf(mu = body_mass_g ~ species, sigma = ~ sex),
  data = pen, family = drmTMB::lognormal()
)
coef(fit7, "mu")
coef(fit7, "sigma")
```

```{r comparison7-comparator, eval = has_penguins && has_glmmTMB}
cmp7 <- glmmTMB(
  log(body_mass_g) ~ species, dispformula = ~ sex, data = pen, family = gaussian()
)
fixef(cmp7)$cond
fixef(cmp7)$disp
```

Compare on the log scale, coefficient for coefficient, with `sigma`
unsquared. Two things matter here that a coefficient-by-coefficient eyeball
can miss. First, `drmTMB`'s `mu` is `E[log y]`, not `log E[y]`:

```{r comparison7-mu-check, eval = has_penguins && has_glmmTMB}
mu_hat <- predict_parameters(fit7, dpar = "mu", type = "link")$estimate
eta_sigma <- predict_parameters(fit7, dpar = "sigma", type = "link")$estimate
y <- pen$body_mass_g
c(
  reported = as.numeric(logLik(fit7)),
  `hand, meanlog = mu` = sum(dlnorm(y, mu_hat, exp(eta_sigma), log = TRUE)),
  `hand, meanlog = mu - sigma^2/2` = sum(dlnorm(y, mu_hat - exp(eta_sigma)^2 / 2, exp(eta_sigma), log = TRUE))
)
```

Second, the two packages' raw log-likelihoods are not comparable directly:
`drmTMB` reports it on the original `body_mass_g` scale, including the
log-Jacobian for the log transform, while `glmmTMB` reports it for
`log(body_mass_g)`. The gap between them is exactly that Jacobian term:

```{r comparison7-loglik, eval = has_penguins && has_glmmTMB}
c(drmTMB = as.numeric(logLik(fit7)), glmmTMB = as.numeric(logLik(cmp7)))
sum(-log(y))
```

Coefficient agreement, once the two traps above are accounted for, is close
on every term. Gentoo penguins are heavier than Adelie on the log scale, and
males are noticeably more variable in log body mass than females — the
`sigma ~ sex` submodel is what makes that second finding representable at
all; a model with one residual SD has no parameter for it.

**Independence: WEAK.**

### Comparison 8: are satiated owl broods more variable, not just quieter?

`glmmTMB::Owls` records sibling negotiation calls by nest, food treatment,
and parent sex. Beyond the mean call rate, does food treatment change the
overdispersion too?

The scale conversion is again checked first on a fixed-effect variant, where
the likelihood is exact:

```{r comparison8-fe-check, eval = has_glmmTMB}
fit8_fe <- drmTMB(
  bf(mu = SiblingNegotiation ~ FoodTreatment, sigma = ~ FoodTreatment),
  data = Owls, family = drmTMB::nbinom2()
)
mu_hat <- predict_parameters(fit8_fe, dpar = "mu", type = "response")$estimate
eta_sigma <- predict_parameters(fit8_fe, dpar = "sigma", type = "link")$estimate
y <- Owls$SiblingNegotiation
c(
  reported = as.numeric(logLik(fit8_fe)),
  `hand, size = 1/sigma^2` = sum(dnbinom(y, mu = mu_hat, size = 1 / exp(eta_sigma)^2, log = TRUE)),
  `hand, size = 1/sigma` = sum(dnbinom(y, mu = mu_hat, size = 1 / exp(eta_sigma), log = TRUE)),
  `hand, size = sigma` = sum(dnbinom(y, mu = mu_hat, size = exp(eta_sigma), log = TRUE))
)
```

`drmTMB`'s `sigma` maps to `size = 1 / sigma^2`; `glmmTMB`'s `dispformula`
predictor is `log(size)` directly. The conversion is therefore
`log(theta) = -2 * log(sigma)`, and only the first hand-built likelihood
above reproduces `drmTMB`'s reported value — the other two miss it by a wide
margin, in an order of magnitude that could otherwise look plausible if you
did not check it.

The `drmTMB::` prefix on `nbinom2()` above is mandatory here, not stylistic.
`glmmTMB` also exports a family constructor named `nbinom2()`; once
`glmmTMB` is attached, a bare `family = nbinom2()` silently hands `drmTMB`
`glmmTMB`'s family object instead of its own, and the call is rejected:

```{r comparison8-clash, eval = has_glmmTMB, error = TRUE}
drmTMB(
  bf(mu = SiblingNegotiation ~ FoodTreatment, sigma = ~ FoodTreatment),
  data = Owls, family = nbinom2()
)
```

On the displayed random-effect model, the joint fits agree on the marginal
log-likelihood, and the same `-2 *` conversion carries over to the `sigma`
coefficients:

```{r comparison8-drmtmb, eval = has_glmmTMB}
fit8 <- drmTMB(
  bf(
    mu = SiblingNegotiation ~ FoodTreatment * SexParent + (1 | Nest),
    sigma = ~ FoodTreatment
  ),
  data = Owls, family = drmTMB::nbinom2()
)
coef(fit8, "mu")
coef(fit8, "sigma")
```

```{r comparison8-comparator, eval = has_glmmTMB}
cmp8 <- glmmTMB(
  SiblingNegotiation ~ FoodTreatment * SexParent + (1 | Nest),
  dispformula = ~ FoodTreatment, family = nbinom2, data = Owls
)
fixef(cmp8)$cond
fixef(cmp8)$disp
```

```{r comparison8-convert, eval = has_glmmTMB}
c(drmTMB = as.numeric(logLik(fit8)), glmmTMB = as.numeric(logLik(cmp8)))
sigma8 <- coef(fit8, "sigma")
abs(-2 * sigma8 - fixef(cmp8)$disp)
```

Skip the conversion and the mismatch is large enough to notice — but it is
signed the opposite way for `nbinom2()` than it was for the identity rule in
Comparison 6, which is exactly the kind of error that survives a casual
eyeball because it still lands in a plausible-looking range. Applied
correctly, satiated broods have higher `sigma`, meaning lower `size`,
meaning more overdispersion than food-deprived broods, and the two packages
agree once, and only once, the conversion is applied.

**Independence: WEAK.**

## What these eight comparisons show

Three of the comparisons above — the `lme4` and `metafor` ones — are against
separate estimation engines. Three are against `glmmTMB`, which is built on
the same TMB automatic-differentiation stack and outer optimizer as
`drmTMB`, so they check consistency between related implementations rather
than across independent ones. Two are against `ordinal`, whose independence
`drmTMB`'s design documents have not yet classified. It is worth naming
where the
distributional-parameter comparisons fall among those three groups: every
comparison of a residual-scale submodel in this article — the capability
`drmTMB` exists for — is in the `glmmTMB` group. The separate-engine checks
on a `sigma` linear predictor are the two meta-analysis comparisons, where
`sigma` is a between-study heterogeneity SD, not a residual scale.

These eight models were compared against packages listed in `drmTMB`'s
`DESCRIPTION` under `Suggests`, because those are the packages this article
can call. `drmTMB` implements a great deal more than eight models; what it
implements, and what evidence stands behind each capability, is recorded in
the package's design documents and capability ledger rather than here.
