---
title: "Improving convergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Improving convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Complex distributional models sometimes need more care than the default fit.
This article gives a practical workflow for `drmTMB` users when the optimizer
does not converge, gradients look large, `TMB::sdreport()` reports a
non-positive-definite Hessian, or Wald standard errors are unavailable.

The default optimizer settings are chosen for ordinary models to fit quickly
and reproducibly. They are not meant to be the best possible settings for every
large phylogenetic, spatial, bivariate, location-scale, shape, inflation, or
random-slope model. For complex models, using a larger optimizer budget is a
normal part of the workflow, not a sign that the model has failed.

The short version is: diagnose first, increase optimizer limits when the
diagnostics point to an unfinished optimization, and simplify the model when
the fitted surface is flat because a variance, correlation, shape, dispersion,
or zero-inflation component is weakly identified.

## Start with `check_drm()`

Always inspect the fitted object before interpreting a difficult model:

```{r}
check_drm(fit)
```

The diagnostic table separates several questions that are easy to confuse:

| Diagnostic | What it asks | First response when it warns |
| --- | --- | --- |
| `optimizer_convergence` | Did `nlminb()` return convergence code 0? | Inspect the optimizer message, gradient, and budget rows. |
| `optimizer_budget` | Did the fit hit supplied `iter.max` or `eval.max` limits? | Increase limits if the model is still moving and the structure is defensible. |
| `fixed_gradient` | Is the fixed-parameter gradient small at the selected optimum? | Treat large gradients as an optimization problem before using Wald inference. |
| `hessian_positive_definite` | Did `TMB::sdreport()` report a positive-definite Hessian? | Look for boundary SDs, correlations, shape, dispersion, or inflation parameters. |
| `fixed_se_finite` | Are fixed-effect standard errors finite? | If not, do not use Wald intervals until the Hessian problem is understood. |
| boundary rows | Are fitted SDs, `rho12`, or shape parameters near a boundary? | Consider a simpler model, profile intervals, or a fixed/removal decision. |

A warning is not automatically a failed analysis. It is a prompt to decide
whether the problem is unfinished optimization, weak identifiability, or an
unnecessary model component.

## Use `drm_control()` for optimizer limits

`drmTMB()` uses `stats::nlminb()` by default. The public way to pass optimizer
settings is through `drm_control(optimizer = list(...))`:

```{r}
fit <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  family = gaussian(),
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000)
  )
)
```

This kind of control is often appropriate for complex models:

```{r}
complex_control <- drm_control(
  optimizer = list(iter.max = 5000, eval.max = 5000)
)
```

There is no universal best number. A small Gaussian fixed-effect model should
not need the same budget as a bivariate phylogenetic location-scale model with
random slopes and fitted correlations. Use `check_drm()` to decide whether the
extra budget changed the result or merely gave the optimizer more time to
reach the same boundary.

A plain list is still accepted for backward compatibility:

```{r}
fit <- drmTMB(
  bf(y ~ x, sigma ~ z),
  data = dat,
  control = list(iter.max = 2000, eval.max = 2000)
)
```

Prefer `drm_control()` in new code because it keeps optimizer settings separate
from `drmTMB` controls such as `se`, `keep_data`, `keep_model_frame`,
`keep_tmb_object`, `sparse_fixed`, and `aggregate_gaussian`.

Do not pass `se = FALSE` or storage controls inside `optimizer = list(...)`.
Those names are not `nlminb()` controls:

```{r}
fit_fast <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = FALSE
  )
)
```

Here `se = FALSE` skips `TMB::sdreport()` after optimization. The fitted
coefficients, log likelihood, fitted values, residuals, predictions,
simulations, and retained profile-likelihood paths remain available, but Wald
standard errors, `vcov()`, and Wald confidence intervals are unavailable.

## If the optimizer did not converge

When `optimizer_convergence` warns, read the message and the other diagnostic
rows. A useful escalation ladder is:

1. Check the model formula and data. Confirm that each predictor has variation
   after missing-row removal, each grouping factor has enough levels and
   replication, and each transformed predictor is finite.
2. Center and scale continuous predictors. This is especially important for
   polynomial terms, interactions, phylogenetic or spatial terms, and
   distributional parameters such as `sigma`, `nu`, zero inflation, hurdle
   probability, or `rho12`.
3. Increase `iter.max` and `eval.max` when the optimizer hit a supplied limit
   or stopped before the gradient was small.
4. Fit a simpler version of the model. For example, use `sigma ~ 1` before
   `sigma ~ x`, remove a weak random slope before fitting a full covariance
   block, or fit `mu` random effects before adding scale, shape, inflation, or
   residual-correlation formulas.
5. Compare the estimates from the simpler and larger models. If the main
   scientific effect changes wildly, treat the larger model as unstable until
   simulation or profiling supports it.

Increasing optimizer limits is not a cure for a model that the data cannot
identify. It is useful when the fit was still making progress, the gradient is
not yet small, or a large model simply needed more evaluations.

## If `pdHess` is false

`pdHess = FALSE` means the Hessian used by `TMB::sdreport()` is not positive
definite at the selected optimum. In practice this often means the likelihood
surface is nearly flat in at least one direction.

Common causes are:

- a random-effect SD close to zero;
- a fitted correlation close to -1 or 1, including residual `rho12` or a
  random-effect correlation;
- a zero-inflation or hurdle component estimated near its boundary;
- dispersion or shape parameters that duplicate another source of variation;
- a scale model that is trying to explain the same pattern as a random effect;
- too many random slopes or covariance terms for the available replication;
- badly scaled predictors or responses.

First check whether the gradient is small. If the gradient is large, the model
has not been optimized well enough. If the gradient is small but `pdHess` is
false, the point estimates may still describe the fitted optimum, but Wald
standard errors and Wald confidence intervals should not be treated as reliable.

In that second case, look for the culprit:

```{r}
summary(fit)
check_drm(fit)
fit$sdpars
fit$corpars
```

If one SD is effectively zero, that component may be unnecessary for these
data. If one correlation is close to -1 or 1, the covariance block may be too
rich. If `sigma`, `nu`, zero inflation, hurdle, or one-inflation terms are
extreme, the distributional parameter may be duplicating another part of the
model.

For inference on direct SD or correlation targets, use profile-likelihood
intervals when the target is available:

```{r}
profile_targets(fit)

summary(
  fit,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "sd:mu:(1 | id)"
)
```

Profile intervals are not magic either: a flat profile is evidence that the
parameter is weakly identified. But profiles show the shape of the likelihood
more honestly than a Wald standard error near a boundary.

## Separate optimization from uncertainty

For very large or experimental fits, it can be useful to separate "can the
model optimize?" from "can Wald uncertainty be computed?":

```{r}
fit_no_se <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = FALSE
  )
)

check_drm(fit_no_se)
```

This fit skips `TMB::sdreport()`, so Hessian and finite-SE rows are reported as
notes. Use this mode for exploratory optimization, prediction, simulation, and
large-data stress tests. Refit with `se = TRUE` when Wald standard errors or
Wald confidence intervals are part of the final report:

```{r}
fit_final <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = TRUE
  )
)
```

For memory-heavy objects, keep the TMB object in the final fit if you need
gradient diagnostics or profile intervals:

```{r}
fit_final <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    keep_tmb_object = TRUE
  )
)
```

## Model-specific advice

For ordinary Gaussian location random effects, start with the location model.
Add arbitrary correlated `mu` slopes only when groups have enough replication
to estimate the covariance block. A model such as `(1 + x1 + x2 | id)` is
statistically hungrier than three independent SD terms.

For Gaussian scale models, remember that `sigma` is a residual standard
deviation model. Scale random effects and scale random slopes can be difficult
to separate from location random effects, residual heteroscedasticity, and
outliers. Use enough groups and repeated observations, and check boundary SDs
before interpreting a scale component.

For bivariate models, keep residual `rho12`, group-level correlations,
phylogenetic or spatial correlations, and known sampling covariance `V`
separate. A warning in one correlation layer does not mean the others are the
same problem.

For non-Gaussian models, add complexity in this order: fixed-effect `mu`,
fixed-effect scale or shape if implemented for that family, ordinary `mu`
random effects, then scale, shape, inflation, hurdle, one-inflation, structured
dependence, or cross-parameter covariance only after each simpler layer is
stable. Non-Gaussian shape and inflation formulas can look like extra variance
components when data are sparse.

For `meta_V(V = V)` models, check the scale and conditioning of the known
sampling covariance. A diagonal known variance vector and a dense known
covariance matrix are both accepted inputs, but a poorly conditioned dense `V`
can make the likelihood numerically harder even when the model formula is
simple.

## What is planned but not automatic yet

Several convergence tools are deliberately not automatic in the current
release:

- user-supplied named starts;
- warm-start refits from a simpler fitted model;
- fallback optimizers such as `optim(method = "BFGS")`;
- multi-start jittered searches;
- Hessian eigenvector culprit reporting;
- automatic simplification suggestions.

The current design records these as future controls rather than silently trying
many optimizers behind the user's back. For reproducible science, a fitted
object must record which optimizer, start, objective value, gradient, and
selected optimum produced the reported estimates.

## Further reading

The practical recommendations here follow the same broad logic used by related
mixed-model software:

- the `metafor` note on
  [convergence problems in `rma.mv()`](https://www.metafor-project.org/doku.php/tips:convergence_problems_rma_mv),
  which shows why boundary variance components can make `nlminb()` sensitive
  to iteration and tolerance settings;
- the
  [`glmmTMB` troubleshooting vignette](https://glmmtmb.github.io/glmmTMB/articles/troubleshooting.html),
  especially its discussion of non-positive-definite Hessians, extreme
  parameters, and flat likelihood directions;
- the
  [`lme4` convergence guidance](https://lme4.github.io/lme4/reference/convergence.html),
  which recommends scaling predictors, restarting from the reported optimum,
  and comparing optimizers when warnings are hard to interpret;
- the
  [`TMB::sdreport()` documentation](https://search.r-project.org/CRAN/refmans/TMB/html/sdreport.html),
  which explains why Wald standard errors depend on the fitted-parameter
  Hessian.
