---
title: "Improving convergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Improving convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl=FALSE}
library(drmTMB)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144,
  dev.args = list(bg = "white"),
  purl = FALSE
)

theme_convergence <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey20"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      strip.text = ggplot2::element_text(face = "bold", colour = "grey15"),
      legend.position = "bottom",
      legend.title = ggplot2::element_blank()
    )
}
```

Complex distributional models sometimes need more care than the default fit.
This article gives a practical workflow for `drmTMB` users when the optimizer
does not converge, gradients look large, `TMB::sdreport()` reports a
non-positive-definite Hessian, or Wald standard errors are unavailable. It
picks up where the `check_drm()` step in [Checking and using fitted
models](model-workflow.html) flags a problem.

The default optimizer settings are chosen for ordinary models to fit quickly
and reproducibly. They are not meant to be the best possible settings for every
large phylogenetic, spatial, bivariate, location-scale, shape, inflation, or
random-slope model. For complex models, using a larger optimizer budget is a
normal part of the workflow, not a sign that the model has failed.

The short version is: diagnose first, increase optimizer limits when the
diagnostics point to an unfinished optimization, and simplify the model when
the fitted surface is flat because a variance, correlation, shape, dispersion,
or zero-inflation component is weakly identified.

## Start with `check_drm()`

Always inspect the fitted object before interpreting a difficult model:

```{r, purl=FALSE}
check_drm(fit)
```

The diagnostic table separates several questions that are easy to confuse:

| Diagnostic | What it asks | First response when it warns |
| --- | --- | --- |
| `optimizer_convergence` | Did `nlminb()` return convergence code 0? | Inspect the optimizer message, gradient, and budget rows. |
| `optimizer_budget` | Did the fit hit supplied `iter.max` or `eval.max` limits? | Increase limits if the model is still moving and the structure is defensible. |
| `fixed_gradient` | Is the fixed-parameter gradient small at the selected optimum, and which component is largest? | Treat large gradients as an optimization problem before using Wald inference. |
| `hessian_positive_definite` | Did `TMB::sdreport()` report a positive-definite Hessian? | Look for boundary SDs, correlations, shape, dispersion, or inflation parameters. |
| `standard_errors_finite` | Are fixed-effect standard errors finite? | If not, do not use Wald intervals until the Hessian problem is understood. |
| boundary rows | Are fitted SDs, `rho12`, or shape parameters near a boundary? | Consider a simpler model, profile intervals, or a fixed/removal decision. |

A warning is not automatically a failed analysis. It is a prompt to decide
whether the problem is unfinished optimization, weak identifiability, or an
unnecessary model component.

The same table is often easier to read as a status map. The next figure uses
three tiny fitted examples: a clean Gaussian fit, the same model run with
`drm_control(se = FALSE)`, and a deliberately under-budgeted random-effect fit.
The colours are diagnostic statuses, not uncertainty intervals.

```{r convergence-check-map, eval=has_ggplot2, fig.width=8.2, fig.height=4.8, fig.cap="`check_drm()` status map for three tiny fitted examples. Tiles show diagnostic statuses; they are not confidence intervals or posterior probabilities.", fig.alt="Tile plot of check_drm diagnostic statuses for a clean fit, a no-standard-error stress fit, and a deliberately low-budget random-effect fit. Clean checks are green, skipped Hessian and standard-error checks are blue notes, and low-budget optimizer and gradient checks are orange warnings."}
set.seed(20260523)
diagnostic_data <- data.frame(
  y = 0.5 + 0.7 * stats::rnorm(80) + stats::rnorm(80, sd = 0.35),
  x = stats::rnorm(80),
  z = stats::rnorm(80),
  id = factor(rep(seq_len(20), each = 4))
)

fit_clean <- drmTMB(
  bf(y ~ x, sigma ~ 1),
  data = diagnostic_data,
  family = gaussian()
)
fit_no_se <- drmTMB(
  bf(y ~ x, sigma ~ 1),
  data = diagnostic_data,
  family = gaussian(),
  control = drm_control(se = FALSE)
)
fit_low_budget <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = diagnostic_data,
  family = gaussian(),
  control = drm_control(
    optimizer = list(iter.max = 1, eval.max = 1)
  )
)

diagnostic_fits <- list(
  "Clean fit" = fit_clean,
  "Stress fit\nse = FALSE" = fit_no_se,
  "Budget hit" = fit_low_budget
)
diagnostic_checks <- do.call(
  rbind,
  lapply(names(diagnostic_fits), function(fit_label) {
    checks <- as.data.frame(check_drm(diagnostic_fits[[fit_label]]))
    checks$fit <- fit_label
    checks
  })
)
diagnostic_rows <- c(
  "optimizer_convergence",
  "optimizer_budget",
  "finite_objective",
  "fixed_gradient",
  "sdreport_status",
  "hessian_positive_definite",
  "standard_errors_finite",
  "random_effect_sd_boundary"
)
diagnostic_row_labels <- c(
  "optimizer\nconvergence",
  "optimizer\nbudget",
  "finite\nobjective",
  "fixed\ngradient",
  "sdreport\nstatus",
  "positive-definite\nHessian",
  "finite\nSEs",
  "random-effect\nSD boundary"
)
diagnostic_grid <- expand.grid(
  check = diagnostic_rows,
  fit = names(diagnostic_fits),
  stringsAsFactors = FALSE
)
diagnostic_plot_data <- merge(
  diagnostic_grid,
  diagnostic_checks[, c("check", "fit", "status")],
  by = c("check", "fit"),
  all.x = TRUE,
  sort = FALSE
)
diagnostic_plot_data$status[is.na(diagnostic_plot_data$status)] <- "not applicable"
diagnostic_plot_data$status <- factor(
  diagnostic_plot_data$status,
  levels = c("ok", "note", "warning", "error", "not applicable")
)
diagnostic_plot_data$fit <- factor(
  diagnostic_plot_data$fit,
  levels = names(diagnostic_fits)
)
diagnostic_plot_data$check_label <- factor(
  diagnostic_row_labels[match(diagnostic_plot_data$check, diagnostic_rows)],
  levels = rev(diagnostic_row_labels)
)

ggplot2::ggplot(
  diagnostic_plot_data,
  ggplot2::aes(fit, check_label, fill = status)
) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.8) +
  ggplot2::geom_text(
    ggplot2::aes(label = status),
    size = 3,
    colour = "grey12"
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "ok" = "#66C2A5",
      "note" = "#8DA0CB",
      "warning" = "#FC8D62",
      "error" = "#D53E4F",
      "not applicable" = "grey88"
    ),
    drop = FALSE
  ) +
  ggplot2::labs(
    title = "Read convergence as a diagnostic table, not a single flag",
    subtitle = "Notes and warnings tell different stories: skipped uncertainty is not failed optimization",
    x = NULL,
    y = NULL
  ) +
  theme_convergence() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r convergence-gradient-budget, eval=has_ggplot2, fig.width=7.8, fig.height=3.8, fig.cap="Optimizer-budget fixture from the same tiny fits. Points show recorded optimizer counts and maximum fixed-gradient size; the dotted line marks the `check_drm()` fixed-gradient warning threshold.", fig.alt="Two-panel lollipop plot comparing a clean fit and a deliberately low-budget fit. The low-budget fit has fewer recorded function evaluations and a much larger maximum fixed-gradient value, which crosses the dotted warning threshold."}
extract_check_value <- function(fit, check_name) {
  checks <- as.data.frame(check_drm(fit))
  checks$value[match(check_name, checks$check)]
}
extract_named_number <- function(text, name) {
  pattern <- paste0(".*", name, "=([-+0-9.eE]+).*")
  as.numeric(sub(pattern, "\\1", text))
}

gradient_budget <- data.frame(
  fit = rep(c("Clean fit", "Budget hit"), each = 2),
  metric = rep(c("function evaluations", "maximum fixed gradient"), 2),
  value = c(
    extract_named_number(
      extract_check_value(fit_clean, "optimizer_budget"),
      "function"
    ),
    extract_named_number(
      extract_check_value(fit_clean, "fixed_gradient"),
      "max"
    ),
    extract_named_number(
      extract_check_value(fit_low_budget, "optimizer_budget"),
      "function"
    ),
    extract_named_number(
      extract_check_value(fit_low_budget, "fixed_gradient"),
      "max"
    )
  )
)
gradient_budget$fit <- factor(
  gradient_budget$fit,
  levels = c("Clean fit", "Budget hit")
)
gradient_budget$metric <- factor(
  gradient_budget$metric,
  levels = c("function evaluations", "maximum fixed gradient")
)
gradient_budget$value_plot <- pmax(gradient_budget$value, 1e-8)
gradient_budget$baseline <- ifelse(
  gradient_budget$metric == "function evaluations",
  1,
  1e-8
)

ggplot2::ggplot(
  gradient_budget,
  ggplot2::aes(value_plot, fit, fill = fit)
) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = baseline,
      xend = value_plot,
      y = fit,
      yend = fit,
      colour = fit
    ),
    linewidth = 1.1,
    show.legend = FALSE
  ) +
  ggplot2::geom_point(
    shape = 21,
    size = 4.5,
    colour = "white",
    stroke = 0.8,
    show.legend = FALSE
  ) +
  ggplot2::geom_vline(
    data = data.frame(
      metric = factor(
        "maximum fixed gradient",
        levels = levels(gradient_budget$metric)
      )
    ),
    ggplot2::aes(xintercept = 0.001),
    linetype = "dotted",
    colour = "grey35",
    linewidth = 0.5
  ) +
  ggplot2::facet_wrap(~metric, scales = "free_x") +
  ggplot2::scale_x_log10() +
  ggplot2::scale_fill_manual(
    values = c("Clean fit" = "#0072B2", "Budget hit" = "#D55E00")
  ) +
  ggplot2::scale_colour_manual(
    values = c("Clean fit" = "#0072B2", "Budget hit" = "#D55E00")
  ) +
  ggplot2::labs(
    title = "A low-budget fit looks different from a clean optimum",
    subtitle = "Use the gradient and optimizer-count rows before interpreting Wald output",
    x = "Recorded value on a log scale",
    y = NULL
  ) +
  theme_convergence()
```

## Use `drm_control()` for optimizer limits

`drmTMB()` uses `stats::nlminb()` by default. The public way to pass optimizer
settings is through `drm_control(optimizer = list(...))`:

```{r, purl=FALSE}
fit <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  family = gaussian(),
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000)
  )
)
```

For the common case where you simply want a larger deterministic `nlminb()`
budget, use a named preset:

```{r, purl=FALSE}
complex_control <- drm_control(
  optimizer_preset = "robust"
)
```

`optimizer_preset = "careful"` sets `iter.max = 1000` and `eval.max = 1000`.
`optimizer_preset = "robust"` sets both limits to 5000. The default preset adds
no optimizer controls, so ordinary models still use the ordinary `nlminb()`
defaults.

When the default deterministic `nlminb()` call raises an optimizer error, for
example a non-finite gradient evaluation on a difficult tree topology,
`drmTMB()` retries with `optimizer_preset = "careful"` and then `"robust"`.
The retry is deliberately narrow: it runs only along the existing preset
budgets, warns if a larger preset succeeds, and records the selected preset in
`fit$optimizer_used` plus the full attempt table in `fit$optimizer_attempts`.
If you pass explicit optimizer controls, `drmTMB()` respects those controls
instead of silently replacing them with the preset ladder.

There is no universal best number. A small Gaussian fixed-effect model should
not need the same budget as a bivariate phylogenetic location-scale model with
random slopes and fitted correlations. Use `check_drm()` to decide whether the
extra budget changed the result or merely gave the optimizer more time to
reach the same boundary.

For a model that reports a long iteration history, first check whether the fit
actually hit `iter.max` or `eval.max` in `check_drm()`. If it did, rerun with
`optimizer_preset = "careful"` or `"robust"` and compare estimates, gradients,
and boundary rows with the original fit. If the fitted SDs or correlations stay
near a boundary after a larger budget, simplify the random-effect or
correlation structure before treating the longer run as stronger evidence.

If a preset is almost right, combine it with explicit overrides:

```{r, purl=FALSE}
fit <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer_preset = "robust",
    optimizer = list(eval.max = 8000)
  )
)
```

The fitted object records the expanded optimizer settings in `fit$control`, so
the analysis remains reproducible.

A plain list is still accepted for backward compatibility:

```{r, purl=FALSE}
fit <- drmTMB(
  bf(y ~ x, sigma ~ z),
  data = dat,
  control = list(iter.max = 2000, eval.max = 2000)
)
```

Prefer `drm_control()` in new code because it keeps optimizer settings separate
from `drmTMB` controls such as `se`, `keep_data`, `keep_model_frame`,
`keep_tmb_object`, `sparse_fixed`, and `aggregate_gaussian`.

Do not pass `se = FALSE` or storage controls inside `optimizer = list(...)`.
Those names are not `nlminb()` controls:

```{r, purl=FALSE}
fit_fast <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = FALSE
  )
)
```

Here `se = FALSE` skips `TMB::sdreport()` after optimization. The fitted
coefficients, log likelihood, fitted values, residuals, predictions,
simulations, and retained profile-likelihood paths remain available, but Wald
standard errors, `vcov()`, and Wald confidence intervals are unavailable.

## If the optimizer did not converge

When `optimizer_convergence` warns, read the message and the other diagnostic
rows. A useful escalation ladder is:

1. Check the model formula and data. Confirm that each predictor has variation
   after missing-row removal, each grouping factor has enough levels and
   replication, and each transformed predictor is finite.
2. Center and scale continuous predictors. This is especially important for
   polynomial terms, interactions, phylogenetic or spatial terms, and
   distributional parameters such as `sigma`, `nu`, zero inflation, hurdle
   probability, or `rho12`.
3. Increase `iter.max` and `eval.max` when the optimizer hit a supplied limit
   or stopped before the gradient was small.
4. Fit a simpler version of the model. For example, use `sigma ~ 1` before
   `sigma ~ x`, remove a weak random slope before fitting a full covariance
   block, or fit `mu` random effects before adding scale, shape, inflation, or
   residual-correlation formulas.
5. Compare the estimates from the simpler and larger models. If the main
   scientific effect changes wildly, treat the larger model as unstable until
   simulation or profiling supports it.

Increasing optimizer limits is not a cure for a model that the data cannot
identify. It is useful when the fit was still making progress, the gradient is
not yet small, or a large model simply needed more evaluations.

## If `pdHess` is false

`pdHess = FALSE` means the Hessian used by `TMB::sdreport()` is not positive
definite at the selected optimum. In practice this often means the likelihood
surface is nearly flat in at least one direction.

Common causes are:

- a random-effect SD close to zero;
- a fitted correlation close to -1 or 1, including residual `rho12` or a
  random-effect correlation;
- a zero-inflation or hurdle component estimated near its boundary;
- dispersion or shape parameters that duplicate another source of variation;
- a scale model that is trying to explain the same pattern as a random effect;
- too many random slopes or covariance terms for the available replication;
- badly scaled predictors or responses.

First check whether the gradient is small. If the gradient is large, the model
has not been optimized well enough. If the gradient is small but `pdHess` is
false, the point estimates may still describe the fitted optimum, but Wald
standard errors and Wald confidence intervals should not be treated as reliable.

In that second case, look for the culprit:

```{r, purl=FALSE}
summary(fit)
check_drm(fit)
fit$sdpars
fit$corpars
```

If one SD is effectively zero, that component may be unnecessary for these
data. If one correlation is close to -1 or 1, the covariance block may be too
rich. If `sigma`, `nu`, zero inflation, hurdle, or one-inflation terms are
extreme, the distributional parameter may be duplicating another part of the
model.

For inference on direct SD or correlation targets, use profile-likelihood
intervals when the target is available:

```{r, purl=FALSE}
profile_targets(fit)

summary(
  fit,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "sd:mu:(1 | id)"
)
```

Profile intervals are not magic either: a flat profile is evidence that the
parameter is weakly identified. But profiles show the shape of the likelihood
more honestly than a Wald standard error near a boundary.

## When bivariate phylogenetic correlations are near a boundary

Bivariate phylogenetic models can have more than one correlation layer. Keep
these layers separate when reading diagnostics:

- residual `rho12` is within-row response-response dependence after the fitted
  fixed, random, phylogenetic, or spatial components are accounted for;
- a phylogenetic mean-mean correlation is latent structured covariance between
  the response-specific species effects;
- ordinary species or group correlations are non-phylogenetic random-effect
  covariance layers.

Near-boundary values in any of these layers are warnings, but they are not the
same warning. A residual `rho12` close to `+/-1` says the remaining paired
responses are almost perfectly coupled. A phylogenetic mean-mean correlation
close to `+/-1` says the fitted species effects have an almost one-dimensional
latent covariance. If both layers are present and one moves to the boundary,
the model may be trying to allocate the same biological or measurement signal
to more than one covariance component.

A near-boundary phylogenetic correlation is often *weakly identified* rather
than genuinely close to `+/-1`. If one response carries little phylogenetic
signal, its phylogenetic SD is pulled toward zero and the cross-correlation has
nothing to correlate against, so it drifts to the boundary along a nearly flat
likelihood ridge. The telltale sign is a **large standard error on the
correlation despite a positive-definite Hessian** (`pdHess = TRUE`):
`check_drm()` flags this as a `biv_phylo_mu_covariance` warning and a
`standard_errors_inflated` note, and a likelihood profile of the correlation is
flat with no usable interval. A clean Hessian is necessary, not sufficient. The
fix is to drop the cross-correlation (fit the two phylogenetic effects
independently) or report it as non-identified; the trait means and their
standard errors are largely unaffected.

For these fits, use a staged workflow:

1. Fit a fixed/residual bivariate model and inspect `rho12`.
2. Fit the simpler phylogenetic or ordinary group-level model that contains
   only the covariance layer you need first.
3. Compare `check_drm()` rows for optimizer convergence, fixed-gradient
   culprit, `rho12_boundary`, phylogenetic covariance diagnostics,
   `standard_errors_inflated`, and species replication.
4. Use `profile_targets()` or `corpairs()` only for targets whose interval
   status is reported as available.
5. Treat `se = FALSE` runs as optimizer stress evidence only; they do not
   validate Hessian curvature, Wald standard errors, or Wald confidence
   intervals.

Larger data sets help when they add the right replication. More species can
help estimate a phylogenetic covariance layer, and more independent repeated
observations can help separate residual covariance from species-level
covariance. More rows alone do not guarantee a better fit if the extra rows
carry nearly the same paired response signal and leave residual `rho12` and
structured covariance competing for it.

`drmTMB` does not yet expose public `start_from`, raw starts, raw TMB maps,
fallback optimizers, or stochastic multi-start controls. The reserved control
names are deliberate: future start or optimizer comparison tools must record
which parameters were copied, which targets were skipped or jittered, which
optimizer result was selected, and whether inference is tied to that selected
optimum.

## Penalized / MAP estimation for weakly-identified phylogenetic components

When `pdHess` is false because a phylogenetic SD sits near zero or a
phylogenetic correlation pins at `+/-1`, and a simpler model is not the answer
you want, you can regularize the weakly-identified component with an optional
penalty (a weakly-informative prior). This can turn a boundary-pinned,
non-positive-definite fit into a finite, positive-definite one:

```{r, purl=FALSE}
fit_map <- drmTMB(
  bf(
    mu1    = y1 ~ x + phylo(1 | p | id, tree = tree),
    mu2    = y2 ~ x + phylo(1 | p | id, tree = tree),
    sigma1 = ~ x + phylo(1 | p | id, tree = tree),
    sigma2 = ~ x + phylo(1 | p | id, tree = tree),
    rho12  = ~ 1
  ),
  data = dat,
  family = biv_gaussian(),
  penalty = drm_phylo_penalty(sd_u = 1, sd_alpha = 0.05, cor_sd = 0.5)
)
```

`drm_phylo_penalty()` puts a penalised-complexity prior on each phylogenetic SD
(`sd_u`, `sd_alpha`: a priori `P(sd > sd_u) = sd_alpha`) and, when `cor_sd` is
set, a mean-zero normal `N(0, cor_sd)` on the unconstrained (transformed,
atanh-scale) phylogenetic correlation parameter. The
correlation penalty is the part that pulls a coupled correlation off the `+/-1`
boundary.

This switches the estimator from maximum likelihood to **penalized /
maximum-a-posteriori (MAP)**, and that distinction matters:

- The fit is labeled `estimator = "MAP"` and `check_drm()` adds a
  `penalized_map` note. It is a regularized point estimate, not an ML fit.
- `logLik()` returns the **unpenalized** data log-likelihood (the penalty is
  stored in `fit$phylo_penalty`), so it stays comparable on the data scale --
  but likelihood-ratio tests and AIC *across penalized fits* are not standard.
- The penalty makes a model *fittable*; it does not make a weakly-identified
  component *identified*. Where the data are nearly uninformative, the estimate
  reflects the prior.

Because of that last point, **there is no universal penalty strength**: a
`cor_sd` that is right for one dataset is wrong for another, because it trades
off against the unknown true correlation magnitude. Always run a
prior-sensitivity sweep and interpret only the quantities that are stable
across it:

```{r, purl=FALSE}
sweep <- lapply(c(0.25, 0.5, 1.0), function(s) {
  f <- drmTMB(
    form,
    data = dat, family = biv_gaussian(),
    penalty = drm_phylo_penalty(sd_u = 1, sd_alpha = 0.05, cor_sd = s)
  )
  f$corpars$phylo
})
```

If a correlation barely moves as you tighten the prior, it is data-informed; if
it tracks the prior, it is prior-dominated and should be reported as such (or
not interpreted). For a definitive, prior-free result on a coupled phylogenetic
location-scale model, the clean route is within-group replication (more than
one observation per species), which identifies the full model from the data
itself.

Treat the penalty as a complement to -- not a replacement for -- simplifying
the model, collecting replicated data, or fitting the same model in a Bayesian
framework where the prior is explicit.

## Choosing between maximum likelihood and REML

`drmTMB` fits by maximum likelihood (`REML = FALSE`) by default. Restricted
maximum likelihood (`REML = TRUE`) is available for Gaussian and
bivariate-Gaussian models, including phylogenetic ones. The choice affects only
**variance components** -- phylogenetic and other random-effect standard
deviations -- and mainly at small sample sizes.

Why: maximum likelihood estimates a variance component as if the fixed effects
were known, which biases the variance *downward* by a factor of order `p / n`
(`p` fixed effects, `n` groups; for a phylogenetic effect `n` is the number of
tips). REML accounts for the degrees of freedom spent on the fixed effects and
removes that leading bias. The correction is set by the *ratio* `p / n`, not by
`n` alone: a rich fixed-effect surface -- several mean covariates plus an
`sd_phylo(...) ~ x` surface, which all count toward `p` -- can leave a
noticeable correction even at a few hundred tips. On the standard-deviation
scale the correction is about half the size of the variance correction (roughly
`p / 2n`, by the delta method), which is why the SD movements below are modest.
As `p / n` shrinks toward zero the two estimators converge.

What actually moves: for a Gaussian location-scale model the mean and the
variance are information-orthogonal, so switching ML to REML leaves the
**mean-model coefficients essentially unchanged** and adjusts the
**variance-component estimates** -- the phylogenetic SD and, for an
`sd_phylo(...) ~ x` surface, its coefficients. In a 16-species fit the mean
coefficients moved by under 0.005 while the `sd_phylo` intercept and slope moved
by about 0.04 to 0.06; by a few hundred species even the SD difference is
negligible.

Practical guidance:

- **Small `p / n`** -- hundreds of tips or more against a modest number of fixed
  effects (a large clade, or a dataset in the thousands of tips): use the
  default ML. The correction is a fraction of a percent and does not change the
  conclusions. At Ayumi Mizuno's ~10,440-tip scale it is about 0.05%.
- **Large `p / n`** -- tens of species, or a small subclade carrying a rich
  fixed-effect surface: prefer `REML = TRUE`. This is where ML's downward
  variance bias is largest and REML earns its keep.
- **If you interpret a variance component, fit both and compare.** If the
  phylogenetic SD or the `sd_phylo ~ x` coefficients move materially between ML
  and REML, report that sensitivity; if they barely move, the estimate is
  robust to the choice.

```{r, purl=FALSE}
fit_ml <- drmTMB(form, data = dat, family = gaussian()) # default (ML)
fit_reml <- drmTMB(form, data = dat, family = gaussian(), REML = TRUE)
# Compare the variance-component estimates; the mean coefficients should agree.
```

A REML fit is labeled `estimator = "REML"` and goes through the same
`check_drm()`, profile, and standard-error diagnostics as an ML fit; REML does
not change how you read them.

## Separate optimization from uncertainty

For very large or opt-in-control fits, it can be useful to separate "can the
model optimize?" from "can Wald uncertainty be computed?":

```{r, purl=FALSE}
fit_no_se <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = FALSE
  )
)

check_drm(fit_no_se)
```

This fit skips `TMB::sdreport()`, so Hessian and finite-SE rows are reported as
notes. Use this mode for exploratory optimization, prediction, simulation, and
large-data stress tests. Refit with `se = TRUE` when Wald standard errors or
Wald confidence intervals are part of the final report:

```{r, purl=FALSE}
fit_final <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    se = TRUE
  )
)
```

For memory-heavy objects, keep the TMB object in the final fit if you need
gradient diagnostics or profile intervals:

```{r, purl=FALSE}
fit_final <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ z),
  data = dat,
  control = drm_control(
    optimizer = list(iter.max = 2000, eval.max = 2000),
    keep_tmb_object = TRUE
  )
)
```

## Model-specific advice

For ordinary Gaussian location random effects, start with the location model.
Add arbitrary correlated `mu` slopes only when groups have enough replication
to estimate the covariance block. A model such as `(1 + x1 + x2 | id)` is
statistically hungrier than three independent SD terms.

For Gaussian scale models, remember that `sigma` is a residual standard
deviation model. Scale random effects and scale random slopes can be difficult
to separate from location random effects, residual heteroscedasticity, and
outliers. Use enough groups and repeated observations, and check boundary SDs
before interpreting a scale component.

For bivariate models, keep residual `rho12`, group-level correlations,
phylogenetic or spatial correlations, and known sampling covariance `V`
separate. A warning in one correlation layer does not mean the others are the
same problem.

For non-Gaussian models, add complexity in this order: fixed-effect `mu`,
fixed-effect scale or shape if implemented for that family, ordinary `mu`
random effects, then scale, shape, inflation, hurdle, one-inflation, structured
dependence, or cross-parameter covariance only after each simpler layer is
stable. Non-Gaussian shape and inflation formulas can look like extra variance
components when data are sparse.

For `meta_V(V = V)` models, check the scale and conditioning of the known
sampling covariance. A diagonal known variance vector and a dense known
covariance matrix are both accepted inputs, but a poorly conditioned dense `V`
can make the likelihood numerically harder even when the model formula is
simple.

## What is planned but not automatic yet

Several convergence tools are deliberately not automatic in the current
release:

- user-supplied named starts;
- warm-start refits from a simpler fitted model; names such as `start_from`,
  `warm_start`, and `warm_start_from` are reserved until the source-fit contract
  is implemented;
- fallback optimizers such as `optim(method = "BFGS")` or
  `optim(method = "L-BFGS-B")`; names such as `fallback_optimizer`,
  `fallback_optimizers`, and `optimizer_fallback` are reserved until optimizer
  comparison provenance is implemented;
- multi-start jittered searches;
- Hessian eigenvector culprit reporting;
- automatic simplification suggestions.

The current design records these as future controls rather than silently trying
many optimizers behind the user's back. The automatic `nlminb()` preset retry
above is a narrower error-recovery path, not an alternative-optimizer search.
For reproducible science, a fitted
object must record which optimizer, start, objective value, gradient, and
selected optimum produced the reported estimates.

## Further reading

The practical recommendations here follow the same broad logic used by related
mixed-model software:

- the `metafor` note on
  [convergence problems in `rma.mv()`](https://www.metafor-project.org/doku.php/tips:convergence_problems_rma_mv),
  which shows why boundary variance components can make `nlminb()` sensitive
  to iteration and tolerance settings;
- the
  [`glmmTMB` troubleshooting vignette](https://glmmtmb.github.io/glmmTMB/articles/troubleshooting.html),
  especially its discussion of non-positive-definite Hessians, extreme
  parameters, and flat likelihood directions;
- the
  [`lme4` convergence guidance](https://lme4.github.io/lme4/reference/convergence.html),
  which recommends scaling predictors, restarting from the reported optimum,
  and comparing optimizers when warnings are hard to interpret;
- the
  [`TMB::sdreport()` documentation](https://search.r-project.org/CRAN/refmans/TMB/html/sdreport.html),
  which explains why Wald standard errors depend on the fitted-parameter
  Hessian.
