---
title: "Count abundance and extra zeros"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Count abundance and extra zeros}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Many ecological responses are counts: fledglings per nest, parasites per host,
insects in a trap, or soil invertebrates in a quadrat. A Poisson model is the
natural baseline, but it fixes the variance equal to the mean. The
location-scale count question is sharper: do predictors change the expected
abundance, the extra-Poisson variation around that abundance, or the chance of
a separate structural zero?

This article stays inside the currently implemented `drmTMB` count surface:
fixed-effect univariate NB2 and zero-inflated NB2 models.

The source motivation comes from Nakagawa et al. (2026), who use
location-scale models to discuss heteroscedasticity in continuous, count, and
proportion data. Their count section highlights negative-binomial models for
fledglings, insect colony size, parasites, and soil invertebrates, and it
separates overdispersion from structural-zero processes. `drmTMB` uses the
same scientific split, but reports the NB2 scale as public `sigma` rather than
the native size or precision parameter often written as `theta`.

## Model Equation And Syntax

For an overdispersed count model, `nbinom2()` uses

\[
\begin{aligned}
Y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{NB2}(\mu_i, \text{size}_i),\\
\log(\mu_i)
  &= \log(E_i) + \beta_0
     + \beta_1 \text{restored}_i
     + \beta_2 \text{moisture}_i,\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{restored}_i,\\
\text{size}_i
  &= 1 / \sigma_i^2,\\
E[Y_i]
  &= \mu_i,\\
\operatorname{Var}(Y_i)
  &= \mu_i + \sigma_i^2\mu_i^2.
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Read each parameter before interpreting the fitted model:

| Symbol or syntax | Meaning | In the soil-invertebrate example |
|---|---|---|
| \(Y_i\), `springtails` | observed non-negative integer count | springtails found in trap \(i\) |
| \(E_i\), `trap_nights` | sampling effort exposure | number of nights the trap was active |
| `offset(log(trap_nights))` | forces expected count to scale with effort | estimates abundance rate per trap night |
| \(\mu_i\), `mu` | expected count after accounting for effort | expected springtail abundance in a trap |
| \(\boldsymbol{\beta}_{\mu}\) | location coefficients on the log mean scale | habitat and moisture effects on expected abundance |
| \(\sigma_i\), `sigma` | NB2 extra-Poisson scale | variation beyond the Poisson expectation |
| \(\boldsymbol{\beta}_{\sigma}\) | scale coefficients on the log `sigma` scale | habitat effects on extra-Poisson variation |
| \(\text{size}_i\), `theta` in some papers | native NB2 size or precision | `size_i = 1 / sigma_i^2`, so larger `sigma` means smaller `size` |

The `sigma` slope is a variability slope. If `gamma_1` is the restored-habitat
coefficient, then

\[
\frac{\sigma_\text{restored}}{\sigma_\text{degraded}}
  = \exp(\gamma_1),
\qquad
\frac{\sigma_\text{restored}^2}{\sigma_\text{degraded}^2}
  = \exp(2\gamma_1).
\]

If a paper reports the native NB2 parameter
\(\theta_i = 1 / \sigma_i^2\), the direction reverses:

\[
\frac{\theta_\text{restored}}{\theta_\text{degraded}}
  = \exp(-2\gamma_1).
\]

That reversal is why the tutorial keeps saying `sigma`: in `drmTMB`, larger
`sigma` always means more modelled variation for this family.

## A Soil-Invertebrate Example

Suppose springtails are counted from soil traps in degraded and restored
grassland plots. Restoration may increase average abundance, but restored plots
may also be more patchy while litter and vegetation structure re-establish. A
few bare-soil microsites may be true absences where springtails are not
available to be trapped.

This transparent simulation gives us known structure before fitting:

```{r}
set.seed(106)
n <- 360
soil_counts <- data.frame(
  habitat = factor(
    rep(c("degraded", "restored"), each = n / 2),
    levels = c("degraded", "restored")
  ),
  surface = factor(
    sample(c("litter", "bare"), n, replace = TRUE, prob = c(0.72, 0.28)),
    levels = c("litter", "bare")
  ),
  moisture = as.numeric(scale(runif(n, 0.15, 0.95))),
  trap_nights = sample(2:5, n, replace = TRUE)
)

restored <- as.numeric(soil_counts$habitat == "restored")
bare <- as.numeric(soil_counts$surface == "bare")

rate <- exp(log(2.4) + 0.35 * restored + 0.30 * soil_counts$moisture)
mu <- soil_counts$trap_nights * rate
sigma_nb2 <- exp(-0.75 + 0.40 * restored)
zi <- plogis(-3.2 + 1.6 * bare - 0.25 * restored)

structural_zero <- runif(n) < zi
soil_counts$springtails <- ifelse(
  structural_zero,
  0L,
  rnbinom(n, size = 1 / sigma_nb2^2, mu = mu)
)

head(soil_counts)
```

Start with the NB2 location-scale model before adding a structural-zero
submodel:

```{r}
fit_nb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Run diagnostics before interpreting coefficients:

```{r}
check_drm(fit_nb2)
```

The NB2 fit answers two questions. The `mu` coefficients describe abundance
rates, because the offset has already accounted for trap effort. The `sigma`
coefficients describe extra-Poisson variation in the count component:

```{r}
coef(fit_nb2, "mu")
coef(fit_nb2, "sigma")

sigma_ratio <- exp(coef(fit_nb2, "sigma")["habitatrestored"])
c(
  sigma_ratio_restored_vs_degraded = sigma_ratio,
  variance_multiplier_at_same_mu = sigma_ratio^2
)
```

The second value is the multiplier on the quadratic extra-Poisson part
\(\sigma_i^2\mu_i^2\) when two traps have the same expected count. It is not
the full variance ratio whenever \(\mu_i\) also changes, because NB2 variance
contains both \(\mu_i\) and \(\sigma_i^2\mu_i^2\).

## When Zeros Are A Separate Process

The paper example of soil invertebrates in patchy habitats maps naturally to a
zero-inflated count model. Bare-soil microsites can be true absences, while
litter microsites can still produce ordinary sampling zeros from the NB2 count
component. Add a `zi` formula when the biological question needs that split:

\[
\begin{aligned}
\Pr(Y_i = 0)
  &= \pi_i + (1 - \pi_i)\Pr_{\operatorname{NB2}}(0 \mid \mu_i,\sigma_i),\\
\Pr(Y_i = k > 0)
  &= (1 - \pi_i)\Pr_{\operatorname{NB2}}(k \mid \mu_i,\sigma_i),\\
\operatorname{logit}(\pi_i)
  &= \delta_0 + \delta_1 \text{bare}_i,\\
\operatorname{Var}(Y_i)
  &= (1 - \pi_i)(\mu_i + \sigma_i^2\mu_i^2)
     + \pi_i(1 - \pi_i)\mu_i^2.
\end{aligned}
\]

In this equation, \(\pi_i\) is the structural-zero probability. It is not a
scale parameter and should not be interpreted as overdispersion.

```{r}
fit_zinb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat,
    zi ~ surface
  ),
  family = nbinom2(),
  data = soil_counts
)

check_drm(fit_zinb2)
```

The `zi` coefficients are on the log-odds scale. Prediction gives the
structural-zero probability on the response scale:

```{r}
coef(fit_zinb2, "zi")

zero_grid <- data.frame(
  habitat = factor(c("degraded", "degraded"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "bare"), levels = levels(soil_counts$surface)),
  moisture = 0,
  trap_nights = 3
)

data.frame(
  surface = zero_grid$surface,
  structural_zero_probability = predict(fit_zinb2, newdata = zero_grid, dpar = "zi")
)
```

For the fitted zero-inflated model, `predict(fit_zinb2, dpar = "mu")` returns
the conditional NB2 count mean, `sigma(fit_zinb2)` returns the conditional
NB2 overdispersion scale, and `fitted(fit_zinb2)` returns the unconditional
response mean \((1 - \pi_i)\mu_i\):

```{r}
new_traps <- data.frame(
  habitat = factor(c("degraded", "restored"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "litter"), levels = levels(soil_counts$surface)),
  moisture = c(0, 0),
  trap_nights = c(3, 3)
)

mu_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "mu")
sigma_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "sigma")
zi_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "zi")

data.frame(
  habitat = new_traps$habitat,
  conditional_mean = mu_hat,
  sigma = sigma_hat,
  structural_zero_probability = zi_hat,
  unconditional_mean = (1 - zi_hat) * mu_hat,
  unconditional_variance =
    (1 - zi_hat) * (mu_hat + sigma_hat^2 * mu_hat^2) +
      zi_hat * (1 - zi_hat) * mu_hat^2
)
```

Use AIC only as one check, not as a substitute for the design story. A
zero-inflated model needs a plausible structural-zero process such as bare
soil, unsuitable host tissue, unsurveyable habitat, or true absence.

```{r}
AIC(fit_nb2, fit_zinb2)
```

## Current Boundary

The implemented count path is intentionally fixed-effect and univariate:

```r
drmTMB(
  bf(count ~ habitat + offset(log(effort)), sigma ~ habitat, zi ~ surface),
  family = nbinom2(),
  data = dat
)
```

Do not teach the following as fitted count examples yet:

- random effects in NB2 `mu`, `sigma`, or `zi`;
- `sd(group) ~ ...` random-effect scale models;
- `meta_known_V(V = V)` with counts;
- `phylo()` or `spatial()` count models;
- bivariate or mixed-response count models such as
  `family = c(gaussian(), nbinom2())`;
- COM-Poisson underdispersion models.

Those are useful future routes, but each needs its own likelihood path,
simulation recovery, diagnostics, and source-map update before it becomes
tutorial syntax.
