---
title: "Count abundance and extra zeros"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Count abundance and extra zeros}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

Many ecological responses are counts: fledglings per nest, parasites per host,
insects in a trap, or soil invertebrates in a quadrat. A Poisson model is the
natural baseline, but it fixes the variance equal to the mean. The
location-scale count question is sharper: do predictors change the expected
abundance, the extra-Poisson variation around that abundance, or the chance of
a separate structural zero? This tutorial assumes the location-scale reading
pattern from [When variance carries signal, Part 1](location-scale.html) and extends
it to a count mean and dispersion.

Start with an NB2 model when counts vary more than a Poisson model predicts.
Use `mu` for expected abundance, `sigma` for extra-Poisson variation, and add
`zi` only when a separate zero-producing process makes biological sense.
The worked example fits these three parts with measured predictors. For
repeated sampling within sites, or dependence among species or locations, see
[Adding grouped or structured effects](#adding-grouped-or-structured-effects)
after working through the example.

The source motivation comes from Nakagawa et al. (2026), who use
location-scale models to discuss heteroscedasticity in continuous, count, and
proportion data. Their count section highlights negative-binomial models for
fledglings, insect colony size, parasites, and soil invertebrates, and it
separates overdispersion from structural-zero processes. `drmTMB` uses the
same scientific split, but reports the NB2 scale as public `sigma` rather than
the native size or precision parameter often written as `theta`.

## Model Equation And Syntax

For an overdispersed count model, `nbinom2()` uses

\[
\begin{aligned}
Y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{NB2}(\mu_i, \text{size}_i),\\
\log(\mu_i)
  &= \log(E_i) + \beta_0
     + \beta_1 \text{restored}_i
     + \beta_2 \text{moisture}_i,\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{restored}_i,\\
\text{size}_i
  &= 1 / \sigma_i^2,\\
E[Y_i]
  &= \mu_i,\\
\operatorname{Var}(Y_i)
  &= \mu_i + \sigma_i^2\mu_i^2.
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Read each parameter before interpreting the fitted model:

| Symbol or syntax | Meaning | In the soil-invertebrate example |
|---|---|---|
| \(Y_i\), `springtails` | observed non-negative integer count | springtails found in trap \(i\) |
| \(E_i\), `trap_nights` | sampling effort exposure | number of nights the trap was active |
| `offset(log(trap_nights))` | forces expected count to scale with effort | estimates abundance rate per trap night |
| \(\mu_i\), `mu` | expected count after accounting for effort | expected springtail abundance in a trap |
| \(\boldsymbol{\beta}_{\mu}\) | location coefficients on the log mean scale | habitat and moisture effects on expected abundance |
| \(\sigma_i\), `sigma` | NB2 extra-Poisson scale | variation beyond the Poisson expectation |
| \(\boldsymbol{\beta}_{\sigma}\) | scale coefficients on the log `sigma` scale | habitat effects on extra-Poisson variation |
| \(\text{size}_i\), `theta` in some papers | native NB2 size or precision | `size_i = 1 / sigma_i^2`, so larger `sigma` means smaller `size` |

The `sigma` slope is a variability slope. If `gamma_1` is the restored-habitat
coefficient, then

\[
\frac{\sigma_\text{restored}}{\sigma_\text{degraded}}
  = \exp(\gamma_1),
\qquad
\frac{\sigma_\text{restored}^2}{\sigma_\text{degraded}^2}
  = \exp(2\gamma_1).
\]

If a paper reports the native NB2 parameter
\(\theta_i = 1 / \sigma_i^2\), the direction reverses:

\[
\frac{\theta_\text{restored}}{\theta_\text{degraded}}
  = \exp(-2\gamma_1).
\]

That reversal is why the tutorial keeps saying `sigma`: in `drmTMB`, larger
`sigma` always means more modelled variation for this family.

## A Soil-Invertebrate Example

Suppose springtails are counted from soil traps in degraded and restored
grassland plots. Restoration may increase average abundance, but restored plots
may also be more patchy while litter and vegetation structure re-establish. A
few bare-soil microsites may be true absences where springtails are not
available to be trapped.

This transparent simulation gives us known structure before fitting:

```{r}
set.seed(106)
n <- 360
soil_counts <- data.frame(
  habitat = factor(
    rep(c("degraded", "restored"), each = n / 2),
    levels = c("degraded", "restored")
  ),
  surface = factor(
    sample(c("litter", "bare"), n, replace = TRUE, prob = c(0.72, 0.28)),
    levels = c("litter", "bare")
  ),
  moisture = as.numeric(scale(runif(n, 0.15, 0.95))),
  trap_nights = sample(2:5, n, replace = TRUE)
)

restored <- as.numeric(soil_counts$habitat == "restored")
bare <- as.numeric(soil_counts$surface == "bare")

rate <- exp(log(2.4) + 0.35 * restored + 0.30 * soil_counts$moisture)
mu <- soil_counts$trap_nights * rate
sigma_nb2 <- exp(-0.75 + 0.40 * restored)
zi <- plogis(-3.2 + 1.6 * bare - 0.25 * restored)

structural_zero <- runif(n) < zi
soil_counts$springtails <- ifelse(
  structural_zero,
  0L,
  rnbinom(n, size = 1 / sigma_nb2^2, mu = mu)
)

head(soil_counts)
```

Start with the NB2 location-scale model before adding a structural-zero
submodel:

```{r}
fit_nb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Run diagnostics before interpreting coefficients:

```{r}
check_drm(fit_nb2)
```

The NB2 fit answers two questions. The `mu` coefficients describe abundance
rates, because the offset has already accounted for trap effort. The `sigma`
coefficients describe extra-Poisson variation in the count component:

```{r}
coef(fit_nb2, "mu")
coef(fit_nb2, "sigma")

sigma_ratio <- exp(coef(fit_nb2, "sigma")["habitatrestored"])
c(
  sigma_ratio_restored_vs_degraded = sigma_ratio,
  variance_multiplier_at_same_mu = sigma_ratio^2
)
```

The second value is the multiplier on the quadratic extra-Poisson part
\(\sigma_i^2\mu_i^2\) when two traps have the same expected count. It is not
the full variance ratio whenever \(\mu_i\) also changes, because NB2 variance
contains both \(\mu_i\) and \(\sigma_i^2\mu_i^2\).

## When Zeros Are A Separate Process

The paper example of soil invertebrates in patchy habitats maps naturally to a
zero-inflated count model. Bare-soil microsites can be true absences, while
litter microsites can still produce ordinary sampling zeros from the NB2 count
component. Add a `zi` formula when the biological question needs that split:

\[
\begin{aligned}
\Pr(Y_i = 0)
  &= \pi_i + (1 - \pi_i)\Pr_{\operatorname{NB2}}(0 \mid \mu_i,\sigma_i),\\
\Pr(Y_i = k > 0)
  &= (1 - \pi_i)\Pr_{\operatorname{NB2}}(k \mid \mu_i,\sigma_i),\\
\operatorname{logit}(\pi_i)
  &= \delta_0 + \delta_1 \text{bare}_i,\\
\operatorname{Var}(Y_i)
  &= (1 - \pi_i)(\mu_i + \sigma_i^2\mu_i^2)
     + \pi_i(1 - \pi_i)\mu_i^2.
\end{aligned}
\]

In this equation, \(\pi_i\) is the structural-zero probability. It is not a
scale parameter and should not be interpreted as overdispersion.

```{r}
fit_zinb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat,
    zi ~ surface
  ),
  family = nbinom2(),
  data = soil_counts
)

check_drm(fit_zinb2)
```

The `zi` coefficients are on the log-odds scale. Prediction gives the
structural-zero probability on the response scale:

```{r}
coef(fit_zinb2, "zi")

zero_grid <- data.frame(
  habitat = factor(c("degraded", "degraded"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "bare"), levels = levels(soil_counts$surface)),
  moisture = 0,
  trap_nights = 3
)

data.frame(
  surface = zero_grid$surface,
  structural_zero_probability = predict(fit_zinb2, newdata = zero_grid, dpar = "zi")
)
```

For the fitted zero-inflated model, `predict(fit_zinb2, dpar = "mu")` returns
the conditional NB2 count mean, `sigma(fit_zinb2)` returns the conditional
NB2 overdispersion scale, and `fitted(fit_zinb2)` returns the unconditional
response mean \((1 - \pi_i)\mu_i\):

```{r}
new_traps <- data.frame(
  habitat = factor(c("degraded", "restored"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "litter"), levels = levels(soil_counts$surface)),
  moisture = c(0, 0),
  trap_nights = c(3, 3)
)

mu_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "mu")
sigma_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "sigma")
zi_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "zi")

data.frame(
  habitat = new_traps$habitat,
  conditional_mean = mu_hat,
  sigma = sigma_hat,
  structural_zero_probability = zi_hat,
  unconditional_mean = (1 - zi_hat) * mu_hat,
  unconditional_variance =
    (1 - zi_hat) * (mu_hat + sigma_hat^2 * mu_hat^2) +
      zi_hat * (1 - zi_hat) * mu_hat^2
)
```

```{r count-model-parts-figure, eval=requireNamespace("ggplot2", quietly = TRUE), fig.width=7.4, fig.height=4.8, fig.cap="Response-scale model parts for the zero-inflated NB2 example. Points show fitted conditional counts, unconditional counts, NB2 extra-Poisson scale, and structural-zero probabilities on separate facets with their own x scales. No interval bars are drawn because this figure compares fitted components, not confidence intervals.", fig.alt="Faceted point plot for the zero-inflated NB2 soil-count example. Separate facets show conditional expected counts, unconditional expected counts, NB2 extra-Poisson sigma, and structural-zero probabilities for habitat and surface combinations."}
library(ggplot2)

count_plot_grid <- expand.grid(
  habitat = factor(levels(soil_counts$habitat), levels = levels(soil_counts$habitat)),
  surface = factor(levels(soil_counts$surface), levels = levels(soil_counts$surface))
)
count_plot_grid$moisture <- 0
count_plot_grid$trap_nights <- 3
count_plot_grid$conditional_mean <- predict(
  fit_zinb2,
  newdata = count_plot_grid,
  dpar = "mu"
)
count_plot_grid$sigma <- predict(fit_zinb2, newdata = count_plot_grid, dpar = "sigma")
count_plot_grid$structural_zero_probability <- predict(
  fit_zinb2,
  newdata = count_plot_grid,
  dpar = "zi"
)
count_plot_grid$unconditional_mean <-
  (1 - count_plot_grid$structural_zero_probability) *
  count_plot_grid$conditional_mean
count_plot_grid$row_label <- paste(count_plot_grid$habitat, count_plot_grid$surface)
count_plot_long <- rbind(
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Conditional mean",
    value = count_plot_grid$conditional_mean
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Unconditional mean",
    value = count_plot_grid$unconditional_mean
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "NB2 sigma",
    value = count_plot_grid$sigma
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Structural-zero probability",
    value = count_plot_grid$structural_zero_probability
  )
)
count_plot_long$component <- factor(
  count_plot_long$component,
  levels = c(
    "Conditional mean",
    "Unconditional mean",
    "NB2 sigma",
    "Structural-zero probability"
  )
)
count_plot_long$row_label <- factor(
  count_plot_long$row_label,
  levels = rev(unique(count_plot_grid$row_label))
)

ggplot(count_plot_long, aes(value, row_label, colour = habitat)) +
  geom_point(size = 2.7) +
  facet_wrap(~component, scales = "free_x", ncol = 2) +
  scale_colour_manual(values = c("degraded" = "#D55E00", "restored" = "#009E73")) +
  labs(
    title = "Zero-inflated counts have several fitted pieces",
    subtitle = "Facets keep count means, NB2 sigma, and structural-zero probabilities separate",
    x = "Response-scale fitted value",
    y = NULL,
    colour = "Habitat"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30"),
    strip.text = element_text(face = "bold")
  )
```

Use AIC only as one check, not as a substitute for the design story. A
zero-inflated model needs a plausible structural-zero process such as bare
soil, unsuitable host tissue, unsurveyable habitat, or true absence.

```{r}
AIC(fit_nb2, fit_zinb2)
```

## Adding grouped or structured effects

If you repeatedly sample the same site, add an ordinary random intercept
`(1 | site)` to the count mean. An independent numeric slope such as
`(0 + effort | site)` lets the effect of effort vary among sites. These mean
effects are available for Poisson and NB2 models without zero inflation.
For NB2, `sigma ~ z + (1 | id)` also lets extra-Poisson variation differ among
groups. Keep `sigma` free of random effects in zero-inflated, hurdle, and
zero-truncated models.

If your question concerns structural zeros, start with the fixed-effect model
used above:

```r
drmTMB(
  bf(count ~ habitat + offset(log(effort)), sigma ~ habitat, zi ~ surface),
  family = nbinom2(),
  data = dat
)
```

For dependence described by a tree, coordinates, pedigree, or relatedness
matrix, use `phylo()`, `spatial()`, `animal()`, or `relmat()`, respectively.
Poisson and NB2 models without zero inflation allow one such term with an
intercept and one numeric slope in `mu`. NB2 allows the same terms in `sigma`,
fitted separately from the structured `mu` term. For example, choose the first
formula to model species differences in mean abundance and its response to
`x`, or the second to model species differences in overdispersion:

```r
bf(count ~ habitat + phylo(1 + x | species, tree = tree), sigma ~ z)
bf(count ~ habitat, sigma ~ z + phylo(1 + x | species, tree = tree))
```

These structured models have simulation checks on point estimates; those
checks do not establish reliable confidence intervals. Use
`check_drm(fit)` after fitting, and consult
[Can I fit and report this model?](capability-and-limits.html) before reporting
uncertainty. Structured slope-only terms, multiple slopes, and labelled
covariance blocks are unavailable.

For a hurdle NB2 model, you can explore whether related groups have similar
zero probabilities with one relatedness intercept in `hu`. Supply either a
covariance matrix `K` or its precision matrix `Q`:

```r
bf(count ~ habitat, sigma ~ treatment,
  hu ~ relmat(1 | id, K = K))
```

Zero-inflated Poisson also allows one spatial intercept in the zero
probability, `zi ~ spatial(1 | id, coords = coords)`. Alternatively, keep
zero inflation constant and put the spatial intercept in the mean:

```r
# Poisson
bf(count ~ habitat + spatial(1 | site, coords = coords), zi ~ 1)
# NB2
bf(count ~ habitat + spatial(1 | site, coords = coords), sigma ~ 1, zi ~ 1)
```

The hurdle-relatedness and zero-inflated spatial models above have been checked
for fitting and extracting results only. Their point-estimate accuracy and
confidence intervals have not been validated. Use them for exploratory work;
the fixed-effect example earlier in this tutorial is the starting point for
learning how the zero and count components differ.

One NB2 model without zero inflation can combine a spatial mean intercept
with a relatedness mean intercept:
`mu ~ spatial(1 | site, coords = coords) + relmat(1 | id, Q = Q)`.
Its point estimates have simulation checks, but its intervals do not. This
does not extend to other combinations of structured effects.

For your next fit, keep ordinary count slopes independent, use at most the
structured terms described above, and leave ordinary NB2 `sigma` slopes and
`sd(group)` formulas out. Zero-truncated NB2 allows ordinary mean intercepts
and independent slopes when there is no `hu` formula; an active hurdle formula
does not allow those mean random effects. Known sampling covariance via
`meta_V(V = V)`, bivariate or mixed-response count models, and COM-Poisson
underdispersion models are unavailable. If your design needs one of these,
choose a model that supports it before interpreting results.
