---
title: "Count abundance and extra zeros"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Count abundance and extra zeros}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Many ecological responses are counts: fledglings per nest, parasites per host,
insects in a trap, or soil invertebrates in a quadrat. A Poisson model is the
natural baseline, but it fixes the variance equal to the mean. The
location-scale count question is sharper: do predictors change the expected
abundance, the extra-Poisson variation around that abundance, or the chance of
a separate structural zero?

This article stays inside the currently implemented `drmTMB` count surface:
univariate NB2 models, fixed-effect zero-inflated NB2 models, ordinary Poisson
and NB2 `mu` random-effect paths, and the first ordinary Poisson/NB2 q=1
structured `mu` routes for `phylo()`, `spatial()`, `animal()`, and `relmat()`.
Ordinary NB2 also has a first grouped overdispersion path in `sigma`. The worked
example below remains fixed-effect so the mean, overdispersion, and
structural-zero pieces are easy to read before adding group, structured
log-mean, or grouped-overdispersion effects.

The source motivation comes from Nakagawa et al. (2026), who use
location-scale models to discuss heteroscedasticity in continuous, count, and
proportion data. Their count section highlights negative-binomial models for
fledglings, insect colony size, parasites, and soil invertebrates, and it
separates overdispersion from structural-zero processes. `drmTMB` uses the
same scientific split, but reports the NB2 scale as public `sigma` rather than
the native size or precision parameter often written as `theta`.

## Model Equation And Syntax

For an overdispersed count model, `nbinom2()` uses

\[
\begin{aligned}
Y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{NB2}(\mu_i, \text{size}_i),\\
\log(\mu_i)
  &= \log(E_i) + \beta_0
     + \beta_1 \text{restored}_i
     + \beta_2 \text{moisture}_i,\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{restored}_i,\\
\text{size}_i
  &= 1 / \sigma_i^2,\\
E[Y_i]
  &= \mu_i,\\
\operatorname{Var}(Y_i)
  &= \mu_i + \sigma_i^2\mu_i^2.
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Read each parameter before interpreting the fitted model:

| Symbol or syntax | Meaning | In the soil-invertebrate example |
|---|---|---|
| \(Y_i\), `springtails` | observed non-negative integer count | springtails found in trap \(i\) |
| \(E_i\), `trap_nights` | sampling effort exposure | number of nights the trap was active |
| `offset(log(trap_nights))` | forces expected count to scale with effort | estimates abundance rate per trap night |
| \(\mu_i\), `mu` | expected count after accounting for effort | expected springtail abundance in a trap |
| \(\boldsymbol{\beta}_{\mu}\) | location coefficients on the log mean scale | habitat and moisture effects on expected abundance |
| \(\sigma_i\), `sigma` | NB2 extra-Poisson scale | variation beyond the Poisson expectation |
| \(\boldsymbol{\beta}_{\sigma}\) | scale coefficients on the log `sigma` scale | habitat effects on extra-Poisson variation |
| \(\text{size}_i\), `theta` in some papers | native NB2 size or precision | `size_i = 1 / sigma_i^2`, so larger `sigma` means smaller `size` |

The `sigma` slope is a variability slope. If `gamma_1` is the restored-habitat
coefficient, then

\[
\frac{\sigma_\text{restored}}{\sigma_\text{degraded}}
  = \exp(\gamma_1),
\qquad
\frac{\sigma_\text{restored}^2}{\sigma_\text{degraded}^2}
  = \exp(2\gamma_1).
\]

If a paper reports the native NB2 parameter
\(\theta_i = 1 / \sigma_i^2\), the direction reverses:

\[
\frac{\theta_\text{restored}}{\theta_\text{degraded}}
  = \exp(-2\gamma_1).
\]

That reversal is why the tutorial keeps saying `sigma`: in `drmTMB`, larger
`sigma` always means more modelled variation for this family.

## A Soil-Invertebrate Example

Suppose springtails are counted from soil traps in degraded and restored
grassland plots. Restoration may increase average abundance, but restored plots
may also be more patchy while litter and vegetation structure re-establish. A
few bare-soil microsites may be true absences where springtails are not
available to be trapped.

This transparent simulation gives us known structure before fitting:

```{r}
set.seed(106)
n <- 360
soil_counts <- data.frame(
  habitat = factor(
    rep(c("degraded", "restored"), each = n / 2),
    levels = c("degraded", "restored")
  ),
  surface = factor(
    sample(c("litter", "bare"), n, replace = TRUE, prob = c(0.72, 0.28)),
    levels = c("litter", "bare")
  ),
  moisture = as.numeric(scale(runif(n, 0.15, 0.95))),
  trap_nights = sample(2:5, n, replace = TRUE)
)

restored <- as.numeric(soil_counts$habitat == "restored")
bare <- as.numeric(soil_counts$surface == "bare")

rate <- exp(log(2.4) + 0.35 * restored + 0.30 * soil_counts$moisture)
mu <- soil_counts$trap_nights * rate
sigma_nb2 <- exp(-0.75 + 0.40 * restored)
zi <- plogis(-3.2 + 1.6 * bare - 0.25 * restored)

structural_zero <- runif(n) < zi
soil_counts$springtails <- ifelse(
  structural_zero,
  0L,
  rnbinom(n, size = 1 / sigma_nb2^2, mu = mu)
)

head(soil_counts)
```

Start with the NB2 location-scale model before adding a structural-zero
submodel:

```{r}
fit_nb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat
  ),
  family = nbinom2(),
  data = soil_counts
)
```

Run diagnostics before interpreting coefficients:

```{r}
check_drm(fit_nb2)
```

The NB2 fit answers two questions. The `mu` coefficients describe abundance
rates, because the offset has already accounted for trap effort. The `sigma`
coefficients describe extra-Poisson variation in the count component:

```{r}
coef(fit_nb2, "mu")
coef(fit_nb2, "sigma")

sigma_ratio <- exp(coef(fit_nb2, "sigma")["habitatrestored"])
c(
  sigma_ratio_restored_vs_degraded = sigma_ratio,
  variance_multiplier_at_same_mu = sigma_ratio^2
)
```

The second value is the multiplier on the quadratic extra-Poisson part
\(\sigma_i^2\mu_i^2\) when two traps have the same expected count. It is not
the full variance ratio whenever \(\mu_i\) also changes, because NB2 variance
contains both \(\mu_i\) and \(\sigma_i^2\mu_i^2\).

## When Zeros Are A Separate Process

The paper example of soil invertebrates in patchy habitats maps naturally to a
zero-inflated count model. Bare-soil microsites can be true absences, while
litter microsites can still produce ordinary sampling zeros from the NB2 count
component. Add a `zi` formula when the biological question needs that split:

\[
\begin{aligned}
\Pr(Y_i = 0)
  &= \pi_i + (1 - \pi_i)\Pr_{\operatorname{NB2}}(0 \mid \mu_i,\sigma_i),\\
\Pr(Y_i = k > 0)
  &= (1 - \pi_i)\Pr_{\operatorname{NB2}}(k \mid \mu_i,\sigma_i),\\
\operatorname{logit}(\pi_i)
  &= \delta_0 + \delta_1 \text{bare}_i,\\
\operatorname{Var}(Y_i)
  &= (1 - \pi_i)(\mu_i + \sigma_i^2\mu_i^2)
     + \pi_i(1 - \pi_i)\mu_i^2.
\end{aligned}
\]

In this equation, \(\pi_i\) is the structural-zero probability. It is not a
scale parameter and should not be interpreted as overdispersion.

```{r}
fit_zinb2 <- drmTMB(
  bf(
    springtails ~ habitat + moisture + offset(log(trap_nights)),
    sigma ~ habitat,
    zi ~ surface
  ),
  family = nbinom2(),
  data = soil_counts
)

check_drm(fit_zinb2)
```

The `zi` coefficients are on the log-odds scale. Prediction gives the
structural-zero probability on the response scale:

```{r}
coef(fit_zinb2, "zi")

zero_grid <- data.frame(
  habitat = factor(c("degraded", "degraded"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "bare"), levels = levels(soil_counts$surface)),
  moisture = 0,
  trap_nights = 3
)

data.frame(
  surface = zero_grid$surface,
  structural_zero_probability = predict(fit_zinb2, newdata = zero_grid, dpar = "zi")
)
```

For the fitted zero-inflated model, `predict(fit_zinb2, dpar = "mu")` returns
the conditional NB2 count mean, `sigma(fit_zinb2)` returns the conditional
NB2 overdispersion scale, and `fitted(fit_zinb2)` returns the unconditional
response mean \((1 - \pi_i)\mu_i\):

```{r}
new_traps <- data.frame(
  habitat = factor(c("degraded", "restored"), levels = levels(soil_counts$habitat)),
  surface = factor(c("litter", "litter"), levels = levels(soil_counts$surface)),
  moisture = c(0, 0),
  trap_nights = c(3, 3)
)

mu_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "mu")
sigma_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "sigma")
zi_hat <- predict(fit_zinb2, newdata = new_traps, dpar = "zi")

data.frame(
  habitat = new_traps$habitat,
  conditional_mean = mu_hat,
  sigma = sigma_hat,
  structural_zero_probability = zi_hat,
  unconditional_mean = (1 - zi_hat) * mu_hat,
  unconditional_variance =
    (1 - zi_hat) * (mu_hat + sigma_hat^2 * mu_hat^2) +
      zi_hat * (1 - zi_hat) * mu_hat^2
)
```

```{r count-model-parts-figure, eval=requireNamespace("ggplot2", quietly = TRUE), fig.width=7.4, fig.height=4.8, fig.cap="Response-scale model parts for the zero-inflated NB2 example. Points show fitted conditional counts, unconditional counts, NB2 extra-Poisson scale, and structural-zero probabilities on separate facets with their own x scales. No interval bars are drawn because this figure compares fitted components, not confidence intervals.", fig.alt="Faceted point plot for the zero-inflated NB2 soil-count example. Separate facets show conditional expected counts, unconditional expected counts, NB2 extra-Poisson sigma, and structural-zero probabilities for habitat and surface combinations."}
library(ggplot2)

count_plot_grid <- expand.grid(
  habitat = factor(levels(soil_counts$habitat), levels = levels(soil_counts$habitat)),
  surface = factor(levels(soil_counts$surface), levels = levels(soil_counts$surface))
)
count_plot_grid$moisture <- 0
count_plot_grid$trap_nights <- 3
count_plot_grid$conditional_mean <- predict(
  fit_zinb2,
  newdata = count_plot_grid,
  dpar = "mu"
)
count_plot_grid$sigma <- predict(fit_zinb2, newdata = count_plot_grid, dpar = "sigma")
count_plot_grid$structural_zero_probability <- predict(
  fit_zinb2,
  newdata = count_plot_grid,
  dpar = "zi"
)
count_plot_grid$unconditional_mean <-
  (1 - count_plot_grid$structural_zero_probability) *
  count_plot_grid$conditional_mean
count_plot_grid$row_label <- paste(count_plot_grid$habitat, count_plot_grid$surface)
count_plot_long <- rbind(
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Conditional mean",
    value = count_plot_grid$conditional_mean
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Unconditional mean",
    value = count_plot_grid$unconditional_mean
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "NB2 sigma",
    value = count_plot_grid$sigma
  ),
  data.frame(
    count_plot_grid[c("habitat", "row_label")],
    component = "Structural-zero probability",
    value = count_plot_grid$structural_zero_probability
  )
)
count_plot_long$component <- factor(
  count_plot_long$component,
  levels = c(
    "Conditional mean",
    "Unconditional mean",
    "NB2 sigma",
    "Structural-zero probability"
  )
)
count_plot_long$row_label <- factor(
  count_plot_long$row_label,
  levels = rev(unique(count_plot_grid$row_label))
)

ggplot(count_plot_long, aes(value, row_label, colour = habitat)) +
  geom_point(size = 2.7) +
  facet_wrap(~component, scales = "free_x", ncol = 2) +
  scale_colour_manual(values = c("degraded" = "#D55E00", "restored" = "#009E73")) +
  labs(
    title = "Zero-inflated counts have several fitted pieces",
    subtitle = "Facets keep count means, NB2 sigma, and structural-zero probabilities separate",
    x = "Response-scale fitted value",
    y = NULL,
    colour = "Habitat"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30"),
    strip.text = element_text(face = "bold")
  )
```

Use AIC only as one check, not as a substitute for the design story. A
zero-inflated model needs a plausible structural-zero process such as bare
soil, unsuitable host tissue, unsurveyable habitat, or true absence.

```{r}
AIC(fit_nb2, fit_zinb2)
```

## Current Boundary

The implemented count path is still intentionally narrow. Fixed-effect
Poisson, NB2, zero-inflated Poisson, zero-inflated NB2, zero-truncated NB2,
and hurdle NB2 models are fitted. Ordinary non-zero-inflated Poisson and NB2
`mu` models can also fit unlabelled random intercepts and independent numeric
random slopes such as `(1 | site)` and `(0 + effort | site)`. The
first structured count routes are ordinary Poisson and ordinary NB2 q=1
phylogenetic `mu` intercepts, `phylo(1 | species, tree = tree)`, on the
log-mean scale. Ordinary NB2 also fits `sigma ~ z + (1 | id)` as a grouped
overdispersion random intercept on the log-`sigma` scale. The zero-inflated,
hurdle, truncated, and other scale-side count routes remain fixed-effect only.

```r
drmTMB(
  bf(count ~ habitat + offset(log(effort)), sigma ~ habitat, zi ~ surface),
  family = nbinom2(),
  data = dat
)
```

Do not teach the following as fitted count examples yet:

- correlated or labelled NB2 `mu` random-slope blocks;
- NB2 `sigma` random slopes, labelled blocks, structured effects, or `zi`
  random effects;
- random effects in zero-inflated Poisson, zero-inflated NB2, truncated NB2,
  or hurdle NB2 count components;
- `sd(group) ~ ...` random-effect scale models;
- `meta_V(V = V)` or deprecated `meta_known_V(V = V)` with counts;
- zero-inflated or hurdle structured count models, structured count slopes,
  simultaneous structured count types, and labelled structured count covariance;
- NB2 phylogenetic slopes, NB2 `sigma` phylogeny, and zero-inflated NB2
  phylogeny;
- phylogenetic count slopes or labelled q2/q4 phylogenetic count covariance;
- bivariate or mixed-response count models such as
  `family = c(gaussian(), nbinom2())`;
- COM-Poisson underdispersion models.

Those are useful future routes, but each needs its own likelihood path,
simulation recovery, diagnostics, and source-map update before it becomes
tutorial syntax.
