---
title: "Cross-family bivariate models: development note"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-family bivariate models: development note}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

Cross-family bivariate models are a development route for a later release, not
a `drmTMB` 0.6.0 analysis workflow. The native TMB engine supports the released
one-response families and bivariate Gaussian models; use it for analysis and
reporting.

The implemented post-0.6 development route is a post-fit association object,
not a new joint TMB family and not a Julia bridge. It takes two already-fitted
margins on the same externally constructed complete paired rows, freezes their
stage-1 estimates, and then estimates a latent-normal association. The bounded
pair classes are fixed-effect Gaussian × literal-Bernoulli, Gaussian ×
ordinary-NB2, literal-Bernoulli × ordinary-NB2, and literal Bernoulli ×
Bernoulli, with an intercept-only
association predictor and a point estimate plus diagnostics only. The NB2 margin may model both `mu` and `sigma` with
fixed covariates; both resulting rowwise vectors are frozen. This
association is `eta`, not the native bivariate-Gaussian residual correlation
`rho12`, an observed-scale correlation, or the `corpairs()` extractor (which
is distinct from the `corpair()` formula marker).

The Julia bridge remains halted/deferred and does not make Julia, JuliaCall, or
DRM.jl a package dependency. The post-fit route is not a released 0.6.0
analysis surface. Arc 6.1 regression and Arc 6.2 new-pair local smokes are
recorded separately; neither supplies recovery, interval-inference, coverage,
or capability evidence. The deferred bridge is also not available for 0.6.0
analysis.

For the released bivariate workflow, see
[Changing residual coupling with `rho12`](bivariate-coscale.html). For the
deferred Julia boundary, see [Julia engine: future support](julia-engine.html);
keep `engine = "tmb"` as the default.

## Development boundary

The following syntax shows the implemented development interface. It is not a
supported 0.6.0 fitting or reporting recipe. Targeted tests exist, but a smoke
and all stronger claims remain separately gated.

```r
# Post-0.6 development interface: do not use for 0.6.0 release analysis.
# paired_complete_data must be the same complete rows for both margins;
# input order may be reversed.
# fit_gaussian <- drmTMB(y_continuous ~ x, family = gaussian(),
#                        data = paired_complete_data)
# fit_nbinom2 <- drmTMB(bf(mu = y_count ~ x, sigma = ~ z),
#                       family = nbinom2(), data = paired_complete_data)
# assoc <- associate_pairs(
#   fit_gaussian, fit_nbinom2,
#   kernel = latent_normal(), association = ~ 1
# )
# Or pair a literal 0/1 Bernoulli fit with an ordinary NB2 fit. The binary ×
# count route uses a fail-closed latent-normal rectangle calculation; it has no
# recovery, interval, coverage, or capability claim.
```

The object rejects binomial trial counts, non-literal binary values,
zero-inflated/hurdle/truncated NB2, mismatched or partial rows, weights,
offsets, `mi()`, `meta_V()`, REML, random/phylogenetic/structured effects, and
association slopes. It must not refit either margin. Standard errors, intervals,
profiles, `vcov()`, residuals, quantiles, and `emmeans` are also outside these
slices. Recovery, interval, coverage, and capability promotion remain separate
work; none is implied by this development implementation.

## Three frozen-margin examples

The association object answers a narrower question than a joint bivariate
family: after two margins have been fitted separately, is there residual
co-occurrence on the latent-normal scale? `eta` is not `rho12` and is not an
observed-scale correlation.

```{r gaussian-binary-example, eval = FALSE}
fit_continuous <- drmTMB(y_continuous ~ habitat, gaussian(), paired_data)
fit_binary <- drmTMB(bf(mu = y_present ~ habitat), binomial(), paired_data)
assoc_gb <- associate_pairs(fit_continuous, fit_binary,
  kernel = latent_normal(), association = ~ 1)
association(assoc_gb)
```

```{r gaussian-count-example, eval = FALSE}
fit_count <- drmTMB(bf(mu = offspring ~ habitat, sigma = ~ season),
  nbinom2(), paired_data)
assoc_gn <- associate_pairs(fit_continuous, fit_count,
  kernel = latent_normal(), association = ~ 1)
association(assoc_gn)
```

```{r binary-count-example, eval = FALSE}
assoc_bn <- associate_pairs(fit_binary, fit_count,
  kernel = latent_normal(), association = ~ 1)
association(assoc_bn)
```

The last example uses a bivariate-normal rectangle because both outcomes are
discrete. `drmTMB` retains endpoint and integration diagnostics; if a tail
rectangle is unresolved, `eta` is withheld rather than repaired by probability
clipping. These are fixed-effect, complete-pair, intercept-only development
routes, not standard-error, interval, coverage, random-effect, missing-data,
or generic discrete-pair support.

The 0.6.0 capability ledger and release scope manifest are the source of truth
for any later promotion decision.
