---
title: "Cross-family bivariate models with the Julia engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-family bivariate models with the Julia engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article documents the *cross-family* bivariate route: two responses from
**different** distribution families, coupled through a single shared
latent-scale correlation, fitted through the DRM.jl engine via
`engine = "julia"`. It is written for analysts who want to ask whether two
traits measured on incompatible response scales -- for example a continuous
Gaussian trait and a Poisson count -- move together once each trait's own mean
and scale model is in place.

The route is **experimental** (recovery grade -- trust the point estimate of the
latent correlation, not the interval). Its R dispatch ships in `drmTMB` and
routes admitted mixed-family pairs through the Julia bridge with no R-level
decline gate; what is developed separately, and not bundled, is the DRM.jl
cross-family latent engine it calls, so a working fit needs a DRM.jl checkout
that provides it. On the default `drmTMB` install the Julia bridge admits the
Gaussian one-response and Gaussian two-response cells (see
`vignette("julia-engine")`); the cross-family pair uses the same
`engine = "julia"` mechanism and the same `family = c(famA(), famB())`
two-response syntax. Every live code chunk below is
marked `eval = FALSE` so the article builds without Julia, JuliaCall, or DRM.jl
present. Each chunk is followed by a narration of the output you should expect on
a working cross-family checkout.

## Why couple different families

`drmTMB`'s native bivariate Gaussian model (`biv_gaussian()`, or equivalently
`family = c(gaussian(), gaussian())`) answers a familiar question: after
modelling each response's mean and residual scale, do the two residuals
co-vary? The residual correlation `rho12` is the answer, and it lives on the
shared Gaussian scale where "residual" has an unambiguous meaning.

That construction breaks the moment the two responses are not both Gaussian. A
count has no residual that is interchangeable with a Gaussian residual; a
strictly positive trait, a proportion, and a Gaussian measurement each carry
their own mean--variance relationship and their own natural scale. You cannot
correlate a raw Poisson residual with a raw Gaussian residual and expect the
number to mean anything portable.

The cross-family model solves this by moving the dependence off the response
scale and onto a **shared latent scale**. Each response keeps its own family,
its own link, and its own distributional formulas for mean and scale. The
coupling is a single correlation `rho12` between the two latent linear
predictors -- the standardized, Gaussian-scale quantities that sit *behind* each
family's link function. This is the same device that underlies copula and latent
multivariate generalized-linear models: the marginals are whatever each trait
needs, and a Gaussian dependence structure ties them together.

Concrete questions this answers:

- Is a continuous performance trait (Gaussian) associated with a count of
  offspring or seeds (Poisson) at the individual level, after each is regressed
  on the same covariates?
- Do two non-Gaussian traits -- say a positive-continuous size and a count --
  share latent variation that no single-family model can express?

The scientific payoff is a correlation you can interpret across mismatched
scales, with each trait still described by the family that fits it.

## What the latent-scale correlation means

Write the two responses as `y1` from family A and `y2` from family B. The
cross-family model gives each its own mean and scale formulas, exactly as a
univariate `drmTMB` fit would:

```text
y1_i ~ familyA(mu1_i, sigma1_i)
y2_i ~ familyB(mu2_i, sigma2_i)

g1(mu1_i) = X_mu1[i, ] beta_mu1      (familyA link g1, e.g. log for Poisson)
g2(mu2_i) = X_mu2[i, ] beta_mu2      (familyB link g2, e.g. identity for Gaussian)
```

The dependence enters through a pair of correlated latent Gaussian variables,
one per response, with unit variance and correlation `rho12`:

```text
[z1_i, z2_i]' ~ Normal(0, R),   R = [ 1      rho12 ]
                                    [ rho12  1     ]
rho12 = tanh(eta_rho12)
```

Each family draws its response conditional on its own latent variable. For a
Gaussian response the latent variable *is* the standardized residual, so the
construction reduces to the ordinary residual correlation when both families are
Gaussian -- the cross-family model is a strict generalization of
`biv_gaussian()`. For a non-Gaussian response, `z` is the latent driver that the
family's link and variance function transform into the observed scale.

Interpretation of `rho12`:

- `rho12` is a **latent-scale** correlation, not a correlation of the raw
  responses and not a group-level, phylogenetic, or spatial correlation. It is
  the residual coupling after each trait's mean and scale model is removed,
  expressed on the common Gaussian latent scale.
- A positive `rho12` means individuals whose latent value for trait 1 is above
  what its covariates predict also tend to have a latent value for trait 2 above
  what *its* covariates predict. With a Gaussian trait and a Poisson count, that
  reads as: individuals larger than expected on the continuous trait tend to be
  the same individuals that count higher than their covariate-predicted rate.
- It is bounded in `(-1, 1)` by the `tanh` parameterization of `eta_rho12`,
  which is the same scale convention native `drmTMB` uses for residual `rho12`.

Because the marginals stay in their own families, the correlation is comparable
across the scale mismatch: it is not contaminated by the fact that a count has a
mean-linked variance while a Gaussian trait does not.

## Set up the bridge

The cross-family route uses the identical JuliaCall bridge as the rest of the
Julia engine, so the setup is the same as in `vignette("julia-engine")`. The R
package does not install Julia, JuliaCall, or DRM.jl for you; the bridge expects
that R can load `JuliaCall`, that Julia can load DRM.jl, and that `drmTMB` knows
which DRM.jl environment to activate.

```{r install-juliacall, eval = FALSE}
install.packages("JuliaCall")
```

Point `drmTMB` at a DRM.jl checkout with an R option during local development:

```{r set-drmjl-option, eval = FALSE}
options(drmTMB.DRM.jl.path = "/path/to/DRM.jl")
```

or with an environment variable for shell-driven checks:

```sh
export DRM_JL_PATH="/path/to/DRM.jl"
```

If neither is set, `drmTMB` looks for a sibling checkout at `../DRM.jl` from the
current working directory.

## The cross-family usage pattern

The syntax is the two-response `bf()` front end plus a **two-element family
vector** whose elements differ. `family = c(famA(), famB())` is the established
way to declare a two-response model; supplying two *different* families is what
selects the cross-family route. The first family governs response 1
(`mu1`/`sigma1`), the second governs response 2 (`mu2`/`sigma2`), and `rho12`
couples their latent scales.

The example below simulates an individual-level data set where a continuous
performance trait (Gaussian) and a reproductive count (Poisson) both respond to
a body-condition covariate, and the two share positive latent variation.

```{r simulate-crossfamily, eval = FALSE}
set.seed(20260610)
n <- 200
dat <- data.frame(condition = runif(n, -1.5, 1.5))

## shared latent draws with correlation rho12_true
rho12_true <- 0.45
z1 <- rnorm(n)
z2 <- rho12_true * z1 + sqrt(1 - rho12_true^2) * rnorm(n)

## response 1: Gaussian continuous trait on the identity scale
mu1 <- 2.0 + 1.1 * dat$condition
sigma1 <- 0.8
dat$performance <- mu1 + sigma1 * z1

## response 2: Poisson count on the log scale, driven by the second latent
log_mu2 <- 0.4 + 0.6 * dat$condition + 0.5 * z2
dat$offspring <- rpois(n, lambda = exp(log_mu2))
```

Declare one formula per distributional parameter with `bf()`, then pass the two
differing families as a vector and request the Julia engine:

```{r fit-crossfamily, eval = FALSE}
bform <- bf(
  mu1 = performance ~ condition,
  mu2 = offspring   ~ condition,
  sigma1 = ~ 1
)

fit_xf <- drmTMB(
  bform,
  family = c(gaussian(), poisson(link = "log")),  # DIFFERENT families
  data   = dat,
  engine = "julia"
)

fit_xf
```

A successful cross-family fit prints as a lightweight `<drmTMB Julia-engine
cross-family fit>`: the bridge returns plain pieces from DRM.jl and `drmTMB`
reconstructs the coefficients, likelihood summary, and the latent correlation
payload, rather than pretending to be a full native fit. (In-sample fitted
values are *not* wired for this object yet -- see the extractor table below and
`vignette("capability-and-limits")`.)

Note the parameter set. The Gaussian response keeps a `sigma1` scale formula;
the Poisson response has no separate dispersion parameter, so there is no
`sigma2` entry. The latent correlation is not requested through a formula --
DRM.jl estimates it from the coupling of the two latent scales, and you read it
back with `rho_latent(fit_xf)` (see below).

## Reading the latent-scale correlation

The point estimate of the latent correlation comes back through a dedicated
accessor, `rho_latent()`. This is deliberately *not* `rho12()`: `rho12()` is
reserved for the native bivariate-Gaussian residual correlation and has no
cross-family method, so `rho12(fit_xf)` errors. Use `rho_latent(fit_xf)`, which
returns the latent-scale correlation that couples the two responses here.

```{r extract-rho, eval = FALSE}
## the latent-scale correlation between the two responses
rho_latent(fit_xf)

## coefficients split by distributional parameter
coef(fit_xf)
```

Expected shape of the output: `rho_latent(fit_xf)` returns the fitted latent
correlation -- for the simulation above, a single positive value near the
`rho12_true = 0.45` used to generate the data (recovery is approximate at
`n = 200` and depends on the count's information content). `coef(fit_xf)` is a
named list with only the two location blocks: a `mu1` block on the Gaussian
identity scale and a `mu2` block on the Poisson log scale. The Gaussian
dispersion block is not in this default output; reach it with
`coef(fit_xf, dpar = "sigma1")` (log residual SD of the Gaussian trait). There is
no `rho12`/`eta_rho12` block in the coefficients -- the latent correlation is a
separate scalar returned by `rho_latent(fit_xf)`.

Read each block on its own family's scale: `mu1` effects are changes in expected
performance per unit condition; `mu2` effects are changes in log expected
offspring count per unit condition; `rho12` is the latent coupling that ties the
two together.

## Confidence intervals for rho12

The Julia engine exposes interval estimation through `confint()` with a
`method` argument. For the latent correlation, the **profile-likelihood**
interval is the right default: it respects the `(-1, 1)` boundary of a
correlation far better than a symmetric Wald interval, which is especially
important when `rho12` is moderate-to-large or the sample is small.

```{r confint-profile, eval = FALSE}
## profile-likelihood interval for the latent-scale correlation
confint(fit_xf, parm = "rho_latent", method = "profile", level = 0.95)
```

Expected output: a one-row matrix (or data frame) giving the profile lower and
upper limits for `rho12` at the 95% level, with both limits inside `(-1, 1)`.
For the simulated data the interval should comfortably exclude zero, supporting
the conclusion that the continuous trait and the count share positive latent
variation after each is adjusted for body condition.

A Wald interval is available as the faster, symmetric alternative when the
latent-scale Hessian block is returned and `rho12` sits away from the boundary:

```{r confint-wald, eval = FALSE}
## symmetric Wald interval, when the latent-scale Hessian block is available
confint(fit_xf, parm = "rho_latent", method = "wald", level = 0.95)
```

Prefer the profile interval for reporting. The Wald interval can stray outside
`(-1, 1)` and understate asymmetry near the boundary; treat it as a quick check,
not the headline number, exactly as the rest of the Julia-engine inference
surface advises.

## What the cross-family bridge returns

When the route is live, most of the familiar R extractors work on the returned
`drmTMB_julia` object, each reading the cross-family payload that DRM.jl sends
back (the exceptions are noted in the table and in
`vignette("capability-and-limits")`):

| R method | Cross-family behaviour |
| --- | --- |
| `coef()` / `fixef()` | fixed-effect coefficients split by distributional parameter, each on its own family's link scale |
| `rho_latent()` | the latent-scale correlation payload |
| `confint(method = "profile")` | profile-likelihood interval for `rho_latent` (preferred) |
| `confint(method = "wald")` | symmetric Wald interval for `rho_latent` when its Hessian block is available |
| `logLik()`, `nobs()`, `df.residual()` | likelihood and dimension summaries |
| `fitted()` / `predict(dpar = ...)` | **not wired yet** for the cross-family (`drmTMB_julia_xfam`) object -- these currently return `NULL` rather than in-sample means; use the native `engine = "tmb"` path for fitted values (see `vignette("capability-and-limits")`) |
| `is_converged()` | convergence flag returned by DRM.jl |

This is enough to fit a cross-family pair and read the latent correlation with
its profile interval -- the same latent-correlation workflow the Gaussian bridge
already supports, extended to the case where the two responses do not share a
scale. (In-sample fitted means are not yet wired for this object; use the native
`engine = "tmb"` path if you need fitted values for plotting or reporting.)

## Current status and boundaries

To keep the article honest about what ships today:

- The Gaussian one-response and Gaussian two-response cells are the admitted
  Julia-engine routes on the default branch. They have R-scale parity tests
  against the native TMB backend (see `vignette("julia-engine")`).
- The cross-family route -- `family = c(famA(), famB())` with **different**
  families through `engine = "julia"` -- is experimental (recovery grade). Its R
  dispatch ships in `drmTMB` and actively routes admitted mixed-family pairs
  (gaussian / poisson / binomial / nbinom2 / beta / gamma axes) through the
  bridge to DRM.jl -- there is no R-level decline gate. What is developed
  separately is the DRM.jl cross-family latent engine: without a DRM.jl checkout
  that provides it, the request errors at the DRM.jl setup boundary, not by an R
  decline. Its coefficient-scale and latent-correlation parity tests ship as
  skip-gated tests that run against such a checkout.
- The underlying two-response *syntax* is not new: `c(gaussian(), gaussian())`
  already routes to the native bivariate Gaussian engine. The cross-family work
  is about admitting **unequal** family pairs and returning their latent `rho12`
  with profile and Wald intervals through the bridge.
- For any analysis you need to run now on the default branch, use the native
  `engine = "tmb"` path with the families it supports, and use `engine =
  "julia"` only when deliberately exercising the bridge.

The design intent is a narrow, well-tested extension: the same `bf()` front end,
the same `family = c(...)` two-response declaration, and the `rho_latent()` and
`confint(method = "profile")` accessors, now spanning two responses that live on
different scales and are coupled by a single interpretable latent correlation.

## See also

- `vignette("julia-engine")` -- the Julia bridge mechanism, setup, admitted
  cells, and native-vs-Julia parity evidence.
- `vignette("bivariate-coscale")` -- residual `rho12` and co-scale modelling in
  the native Gaussian bivariate engine.
- `vignette("distribution-families")` -- the response families available in
  `drmTMB`.
- `vignette("capability-and-limits")` -- known 0.6.0 limitations, including the
  not-yet-wired Julia cross-family extractors (`fitted()`, `predict()`,
  `vcov()`, `residuals()`, and the `summary()` coefficient table return `NULL`
  on a `drmTMB_julia_xfam` fit).
