---
title: "Association between mixed outcome pairs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association between mixed outcome pairs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<style>
h1.title {
  hyphens: none;
  overflow-wrap: normal;
  word-break: normal;
}
</style>

Use this page when you have two outcomes measured on the same individual, site,
or sampling occasion, but the outcomes need different models. For example, you
might record whether an individual bred (`0` or `1`) and how many offspring it
produced (a count). The question is: after accounting for predictors such as
habitat, do rows that do better than expected for one outcome also tend to do
better than expected for the other?

The one-call tool on this page returns that association and its numerical
diagnostics. It is a beta tool: it gives a point estimate, not a standard error,
confidence interval, profile, coverage claim, or a general bivariate-family
interface. If both outcomes are Gaussian and you need a jointly fitted model,
use `biv_gaussian()` and its residual correlation `rho12` instead.

Behind the scenes, the tool fits each outcome model first, holds those fitted
margins fixed, and estimates a latent-normal association called `eta`. You do
not need to understand that machinery before deciding whether the question and
data requirements below match your study.

## The question in an ecological example

Suppose each individual has a breeding outcome (`0` or `1`) and an offspring
count. Habitat may explain both outcomes. The question is not whether the raw
zeroes and counts have a large Pearson correlation. It is:

> After habitat has been accounted for in each outcome, do individuals that are
> more likely than expected to breed also tend to have more offspring than
> expected?

The answer is expressed on an unobserved common scale. A positive `eta` means
that the same rows tend to be unexpectedly high for both fitted outcomes on that
unobserved Gaussian scale. It does not mean that `eta` is an odds ratio, a logit
coefficient, an observed-scale correlation, or native residual `rho12`.

## One call, two stages

`biv_associate()` is a convenience wrapper. It makes the workflow one R call,
but it is still a two-stage estimator: it fits the two margins, freezes their
estimated parameters, and estimates only `eta` in stage 2. Changing the stage-2
association never refits the mean or scale parameters from either margin.

```{r staged-diagram, fig.width = 8, fig.height = 3.1, fig.cap = "The frozen-margin workflow. The two response models are fitted first on the same complete rows. Their fitted probabilities or distributions are held fixed while stage 2 estimates a latent-normal association, eta. This is not a jointly fitted model and has no association interval in the current beta."}
old_par <- par(mar = c(0, 0, 1, 0))
plot.new()
plot.window(xlim = c(0, 12), ylim = c(0, 4))
boxes <- list(
  c(0.3, 1.3, 2.7, 2.7), c(3.5, 2.2, 6.2, 3.5),
  c(3.5, 0.5, 6.2, 1.8), c(8.2, 1.3, 11.6, 2.7)
)
cols <- c("#E7F3F5", "#F2E8C9", "#F2E8C9", "#DCEAD7")
for (i in seq_along(boxes)) {
  rect(boxes[[i]][1], boxes[[i]][2], boxes[[i]][3], boxes[[i]][4],
       col = cols[i], border = "#112638", lwd = 1.2)
}
text(1.5, 2.35, "One complete\npaired data set", cex = 0.9)
text(4.85, 3.05, "Stage 1\nBinary margin\nP(breeding = 1)", cex = 0.84)
text(4.85, 1.15, "Stage 1\nCount margin\nE(offspring)", cex = 0.84)
text(9.9, 2.35, "Stage 2\nFreeze both margins\nEstimate eta (or eta_i)", cex = 0.88)
arrows(2.7, 2.35, 3.45, 2.85, length = 0.08, lwd = 1.2)
arrows(2.7, 1.65, 3.45, 1.15, length = 0.08, lwd = 1.2)
arrows(6.2, 2.85, 8.15, 2.45, length = 0.08, lwd = 1.2)
arrows(6.2, 1.15, 8.15, 2.05, length = 0.08, lwd = 1.2)
par(old_par)
```

In symbols, stage 1 produces two fitted marginal parameter sets,
\(\widehat\theta_1\) and \(\widehat\theta_2\). Stage 2 estimates

\[
\widehat\beta_A = \arg\max_{\beta_A} \sum_i
\log f_{\eta_i}(y_{1i}, y_{2i} \mid \widehat\theta_1, \widehat\theta_2),
\qquad a_i=X_{A,i}\beta_A,\quad \eta_i = \tanh(a_i).
\]

For the usual constant association, \(X_A=1\), so every row has the same
`eta`. The beta Bernoulli x ordinary-NB2 slope route uses
\(X_{A,i}=(1,x_i)\). In computation, `drmTMB` keeps `eta` infinitesimally
inside -1 and 1, because exact endpoints make the latent-normal probability
calculation unstable. This is a plug-in association criterion, not a joint
maximum-likelihood fit.
The stage-2 Hessian therefore cannot give an ordinary valid standard error for
`eta`, because it treats stage-1 uncertainty as fixed.

## What a binary outcome contributes

A binary observation has no ordinary numeric residual. If the breeding margin
predicts a probability \(p_i = 0.80\), we observe only whether breeding happened,
not how far above or below its predicted tendency the individual lay.

The association calculation uses an auxiliary latent variable \(U_{Bi}\):

\[
U_{Bi} \sim N(0, 1), \qquad
B_i = 1\{U_{Bi} > \Phi^{-1}(1 - p_i)\}.
\]

This threshold gives \(P(B_i = 1) = p_i\). The latent variable is not an
individual measurement and should not be reported as an individual binary
residual. For a Gaussian outcome, the corresponding latent quantity is its
standardized residual \((Y_i - \mu_i)/\sigma_i\). `eta` describes the
correlation between these latent quantities after the margins are frozen.

## Fit a reviewed pair in one call

Create one data frame containing both responses and every predictor, then remove
incomplete rows once. A binary response must contain exactly one literal `0` or
`1` per row. It cannot be a proportion, a `cbind(successes, failures)` response,
or a binomial count with multiple trials.

```{r one-call-example, eval = FALSE}
paired_data <- na.omit(data.frame(
  bred, offspring, habitat, habitat_score, season
))

assoc <- biv_associate(
  bf(mu = bred ~ habitat),
  bf(mu = offspring ~ habitat, sigma = ~ season),
  family = list(binomial(), nbinom2()),
  data = paired_data,
  association = ~ 1
)

association(assoc)
```

The current reviewed classes are Gaussian x literal-Bernoulli, Gaussian x
ordinary-NB2, literal-Bernoulli x literal-Bernoulli, literal-Bernoulli x
ordinary-NB2, and ordinary-NB2 x ordinary-NB2. The last two show why
"cross-family" is only a historical label for this page: the same
frozen-margin machinery also has two reviewed same-family discrete classes.

`association(assoc)` returns `eta` unless the numerical diagnostic is
`boundary_unresolved`; a `near_boundary` status remains flagged. For the
example, a positive value says that a greater latent tendency to breed is
associated with a greater latent tendency for offspring after habitat and season
have been accounted for. It is not a raw-data correlation.

## Can the association itself be modelled?

Usually start with `association = ~ 1`: it estimates one common `eta` after the
predictors in the two margins have been accounted for. In the beta
literal-Bernoulli x ordinary-NB2 route only, you can also ask whether the
association changes along one numeric environmental gradient. For example,
suppose `habitat_score` is a numeric measure rather than a habitat category:

```{r association-slope-example, eval = FALSE}
assoc_by_habitat <- biv_associate(
  bf(mu = bred ~ habitat_score),
  bf(mu = offspring ~ habitat_score, sigma = ~ season),
  family = list(binomial(), nbinom2()),
  data = paired_data,
  association = ~ habitat_score
)

association(assoc_by_habitat)
association(assoc_by_habitat, type = "fitted")
```

This fits \(a_i=\beta_0+\beta_1\,\texttt{habitat_score}_i\) and converts it
to a row-specific latent association \(\eta_i=\tanh(a_i)\).
`association()` first returns the two coefficients on the association-link
scale; `type = "fitted"` returns the corresponding `eta_i` values for the
rows used to fit the margins. These are point estimates only: no standard
errors, intervals, profiles, or coverage claim is available.

This is similar in spirit to allowing a predictor for a residual `rho12`, but
it is not the same model or parameter. `rho12` belongs to a direct joint
likelihood, whereas this beta route freezes two separately fitted margins before
estimating its association link. A categorical habitat, multiple association
predictors, interactions, random effects, new-data association prediction, and
association regressions for the other pair classes remain unsupported.

| Question | Current beta answer |
| --- | --- |
| Does habitat predict breeding? | Put `habitat` in the binary margin. |
| Does season predict offspring variation? | Put `season` in the count `sigma` margin. |
| Is there one remaining association after those effects? | Estimate `association = ~ 1`. |
| Does association change along one numeric habitat score? | For beta Bernoulli x ordinary-NB2 only, use `association = ~ habitat_score`. |
| Does association change among habitat categories or with several predictors? | Not yet supported. |

## When the estimate is withheld

For pairs with two discrete outcomes, the calculation evaluates a bivariate
normal probability region that corresponds to the observed outcome pair. Some
extreme tail regions can be numerically unresolved. In that case `drmTMB`
withholds `eta` rather than clips a probability or reports a repaired estimate.

```{r unresolved-example, eval = FALSE}
if (identical(assoc$status, "boundary_unresolved")) {
  assoc$diagnostics
  # Report that no association estimate was returned. Do not clip or repair it.
}
```

This is the correct development-boundary result. Inspect the diagnostics, report
that no association estimate was available, and do not convert an unresolved
tail calculation into a scientific conclusion.

## How this differs from rho12

`rho12` belongs to a direct joint likelihood. In `biv_gaussian()` it is a
Gaussian residual correlation; in `biv_lognormal()` it is a log-response
residual correlation. `eta` is neither of those. It is a latent-normal
association obtained after separate margins have been fitted and frozen. The
new beta association slope changes the latent `eta_i`, not a direct `rho12`.

Use [Changing residual coupling with rho12](bivariate-coscale.html) for the
released Gaussian joint model. Use [Bivariate non-Gaussian models](bivariate-nongaussian.html)
to choose between an exact same-family joint model and this staged association
route.
