---
title: "Distributional outputs and adequacy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional outputs and adequacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.4,
  fig.height = 4.2,
  dpi = 130
)
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

A fitted `mu` formula answers "what is the expected response, given the
covariates?" It does not answer two other questions an applied paper usually
needs: *"did I model the whole conditional distribution, not just its
average?"* and *"how extreme, or how likely to exceed a threshold, is this
observation for its covariates?"* A model can have a perfectly reasonable
mean structure and still misrepresent the spread, tails, or boundary
behaviour of the response -- and coefficients alone never reveal that. This
article follows the `check_drm()` step in the post-fit workflow described in
[Checking and using fitted models](model-workflow.html).

This article covers `drmTMB`'s distributional-output and adequacy layer,
built on one shared foundation: every fitted family exposes its own density,
CDF, and quantile function at the fitted parameters
(`fitted_distribution()`). Two families of outputs are built on it:

- **Adequacy diagnostics** -- `residuals(fit, type = "quantile")`,
  `worm_plot()`, `qq_plot()` -- randomized quantile residuals (Dunn & Smyth
  1996) that should look like a standard normal sample under a correctly
  specified fixed-effect distributional form.
- **Distributional outputs** -- `predict(fit, type = "quantile")`,
  `exceedance()`, `centile_chart()` -- conditional quantiles, exceedance
  probabilities, and centile curves at the fitted model.

The worked contrast below follows Fig. 4c of the GAMLSS Primer (Merder,
Rigby, Stasinopoulos et al., *Nature Reviews Methods Primers*, 2026, 6:49):
a location-only fit to heteroscedastic data looks fine on a coefficient
table, but its quantile residuals bend a worm plot and fail a normality
test; the matching location-scale fit is flat and passes.

## Setup

```{r setup-data}
set.seed(20260712)
n <- 300
x <- stats::rnorm(n)
sigma_true <- exp(0.1 + 0.9 * x)
y <- 0.3 + 0.6 * x + stats::rnorm(n, sd = sigma_true)
dat <- data.frame(y = y, x = x)
```

`y` is generated with a residual SD that grows with `x`: a Gaussian
location-scale problem, not a location-only one.

## Quantile residuals and worm/QQ plots on a correctly specified fit

Fit the model that matches the data-generating process -- `mu` and `sigma`
both depend on `x` -- and compute randomized quantile residuals.

```{r fit-true}
fit_true <- drmTMB(bf(y ~ x, sigma ~ x), family = gaussian(), data = dat)
r_true <- residuals(fit_true, type = "quantile")
stats::ks.test(r_true, "pnorm")
```

The residuals are indistinguishable from a standard normal sample: a
Kolmogorov-Smirnov test does not reject. `worm_plot()` (a detrended QQ plot)
and `qq_plot()` show the same thing visually -- a flat scatter around the
reference line is "no detectable departure from N(0,1)," never proof the
model is correct.

```{r worm-true, fig.cap="Worm plot of quantile residuals for the correctly specified location-scale fit; the smooth stays near the zero reference across the bulk, with only sparse-tail drift (the KS test does not reject).", fig.alt="Worm plot for the correctly specified fit: points scatter around the zero reference and the smooth holds near zero across the bulk, drifting only in the sparse tails -- no systematic misfit."}
worm_plot(fit_true) +
  ggplot2::labs(
    subtitle = "Fixed-effect adequacy: no detectable departure from N(0,1)"
  )
```

```{r qq-true, fig.cap="Normal QQ plot of quantile residuals for the correctly specified fit; points on the diagonal are no detectable departure from N(0,1).", fig.alt="Normal QQ plot with points tracking the diagonal reference line for the correctly specified fit."}
qq_plot(fit_true)
```

## Reproducing the Fig-4c contrast: a location-only fit to heteroscedastic data

Now fit a location-only model to the same data -- `sigma` held constant, so
the growing spread in `y` has nowhere to go but into the residuals.

```{r fit-mis}
fit_mis <- drmTMB(bf(y ~ x, sigma ~ 1), family = gaussian(), data = dat)
r_mis <- residuals(fit_mis, type = "quantile")
stats::ks.test(r_mis, "pnorm")
```

The KS test now rejects sharply. The worm plot bends into the same
systematic curve the GAMLSS Primer's Fig-4c shows: at extreme covariate
values the observations are consistently too close to (or too far from)
`mu`, because the constant `sigma` cannot track the true variance growth.

```{r worm-mis, fig.cap="Worm plot of quantile residuals for the location-only fit to heteroscedastic data; the systematic bend flags a mis-specified distributional shape.", fig.alt="Worm plot showing a clear curved departure from the zero reference line for the location-only fit, in contrast to the flat scatter under the correctly specified fit."}
worm_plot(fit_mis) +
  ggplot2::labs(
    subtitle = "Fixed-effect adequacy: bend flags a detectable departure from N(0,1)"
  )
```

Both `worm_plot()` and `qq_plot()` apply this package's Confidence-Eye
figure contract: hollow points, no filled dots, no interval bars around the
reference line -- the reference line is a null-hypothesis benchmark, not an
uncertainty interval.

The two fits have the same `mu` formula and produce similar coefficient
tables; only the adequacy check separates them. That is the point of a
distributional diagnostic: it catches a mis-specification a mean-only
residual plot would miss.

## Distributional outputs: quantiles, exceedance, and centiles

Beyond checking the fit, the same `fitted_distribution()` foundation
answers conditional-quantile and threshold questions directly, using the
correctly specified `fit_true`.

### Conditional quantiles

```{r predict-quantile}
newdata <- data.frame(x = c(-1, 0, 1))
predict(fit_true, newdata = newdata, type = "quantile", prob = c(0.025, 0.5, 0.975))
```

The output widens at extreme `x` because `sigma` grows with `x` in this
model -- exactly the heteroscedasticity the location-only fit above could
not represent.

### Exceedance probability

`exceedance()` returns `Pr(Y > threshold | x)` at the fitted model, a thin
wrapper over the shared CDF.

```{r exceedance}
exceedance(fit_true, threshold = 2, newdata = newdata)
```

### Centile chart

`centile_chart()` draws fitted response centiles against one covariate,
holding every other predictor at a reference value.

```{r centile-chart, fig.cap="Model-conditional centile chart for the correctly specified location-scale fit; centiles fan out with x because sigma grows with x.", fig.alt="Centile chart with five fitted centile curves fanning out as x increases, reflecting the fitted heteroscedasticity."}
centile_chart(fit_true, covariate = "x") +
  ggplot2::scale_colour_viridis_d()
```

The centiles fan out as `x` increases -- a direct visual reading of the
fitted `sigma ~ x` relationship that a mean-only prediction plot would not
show.

## What this layer detects -- and what it does not

**Detects.** The adequacy diagnostic reliably catches distributional
**shape or atom** mis-specification that a family cannot reabsorb through
its own free parameters: heavy tails fit as Gaussian, ignored
overdispersion or zero-inflation in a family with no free dispersion
parameter, ignored truncation, a missing zero/one atom, and the
heteroscedasticity example reproduced above. A 400-seed gated campaign
across all 18 fitted families (tweedie: 99 of 400 seeds locally, 66/99
dispersion-arm non-convergence, full run deferred to Totoro)
(`docs/dev-log/simulation-artifacts/2026-07-12-dg3-power-arm-gated/`)
confirms type-I error at or below the nominal rate under a correctly
specified fit, with power >= 0.8 (commonly 0.9-1.0) for these detectable
mis-specifications.

**Does not detect -- a structural blind spot, not evidence of adequacy.**
When a fitted family's own free nuisance, dispersion, or inflation
parameter can absorb a mis-specification, the marginal quantile residual
stays close to N(0,1) even though the mean or variance structure is wrong.
Examples from the same campaign: heteroscedasticity absorbed by Student-t
`nu` (power 0.035, versus 1.0 for the same heteroscedasticity under
Gaussian, which has no absorbing parameter); missing zero-inflation
absorbed by `nbinom2`'s `sigma` (power 0.035, versus 0.96 under Poisson);
and a constant-vs-covariate zero-inflation/hurdle/zero-one-inflation
mechanism, which splits in two: for `hurdle_nbinom2`/`zero_one_beta` power
stays flat near the type-I rate regardless of sample size (a genuine
structural blind spot); for `zi_poisson`/`zi_nbinom2` power rises with
sample size but stays far below 0.8 even at n = 3000, so it is impractical
to detect at realistic sample sizes rather than permanently undetectable.
**A mean-structure diagnostic, not this one, is what catches an absorbed
mis-specification.**

**Wording and scope, throughout this package:**

- Passing this diagnostic means "no detectable departure from N(0,1)," and
  is worded that way everywhere -- never "the model is adequate," "valid,"
  or "correct."
- `residuals(type = "quantile")`, `worm_plot()`, and `qq_plot()` are
  **fixed-effect** adequacy: for a random-effect or structured fit, they are
  conditional on the fixed-effect prediction, not a marginal check of the
  full model.
- `predict(type = "quantile")` and `exceedance()` results carry
  `attr(., "calibrated") <- FALSE`: they are distributional (plug-in)
  outputs at the fitted parameter point estimate, and do not propagate
  parameter uncertainty. `centile_chart()` is a model-conditional curve at
  that same point estimate, not a WHO-style or other population reference
  standard.
- A distributional-output or adequacy tick never changes, or implies
  anything about, a family's separate inference-tier status (see
  `vignettes/capability-and-limits.Rmd`). Fitting quantiles and residuals
  for a family does not certify its confidence intervals.

See `docs/dev-log/known-limitations.md` for the full scope note and
`docs/dev-log/simulation-artifacts/2026-07-12-dg3-power-arm-gated/` for the
underlying campaign this section summarizes.
