---
title: "Distributional regression with drmTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional regression with drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

`drmTMB` fits distributional regression models for one or two responses. Start
here when you want to fit a first model, check that the fitted object matches
your scientific question, and find the next article without reading the whole
roadmap.

The central workflow is simple: write one formula for each parameter you want
to estimate, then check that the fitted model matches the symbolic model you had
in mind. The broader implemented map lives in
[What can I fit today?](model-map.html), and family choice lives in
[Choosing response families](distribution-families.html).

For example, an applied user might ask: do mean trait values and residual
variability change with an environmental predictor, after accounting for
repeated measures from the same site or species? In `drmTMB`, that question is
written as one formula for the mean and one formula for the residual scale.

## Install the preview

`drmTMB` is not on CRAN yet. Install the current development version (the
`0.6.0` line) from GitHub with `pak`:

```r
install.packages("pak")
pak::pak("itchyshin/drmTMB")
```

The earlier `v0.5.0` tag predates the current line and is not a supported
install target; a `0.6.0` release will be tagged when it reaches CRAN.

You need R 4.1.0 or newer and a working compiler toolchain because TMB models
are compiled during installation. If installation fails while compiling C++,
install the usual R build tools for your platform: Rtools on Windows, Xcode
Command Line Tools on macOS, or the R development toolchain on Linux.

The core runtime dependencies are installed automatically by `pak`: `cli`,
`Matrix`, `TMB`, and the compiled headers from `RcppEigen` and `TMB`. The
articles and development checks also use optional packages such as `glmmTMB`,
`lme4`, `MASS`, `metafor`, `knitr`, `rmarkdown`, `testthat`, `withr`, and
`pkgdown`.

## Fit your first model

Start with a Gaussian location-scale model when the response is continuous and
the scientific question is about both the expected value and predictability.
In the small example below, habitat and temperature affect mean growth, while
habitat also changes residual variation:

```{r first-model-data}
set.seed(13)
n <- 120
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)
mu <- 1 + 0.6 * (dat$habitat == "grassland") + 0.4 * dat$temperature
sigma <- exp(-0.5 + 0.45 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma)
```

The fitted model uses one formula for `mu` and one formula for `sigma`:

```{r first-model-fit}
fit <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)

check_drm(fit)
```

Read the `sigma` coefficient as a log residual-SD contrast. Exponentiating it
gives an SD ratio; exponentiating twice the coefficient gives a residual
variance ratio:

```{r first-model-interpret}
sigma_habitat <- coef(fit, "sigma")["habitatgrassland"]
data.frame(
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

For a fuller walkthrough of fitted means, residual SDs, and residual
variances, read [When variance carries signal, Part
1](location-scale.html). Continue to [Part
2](location-scale-scale.html) when predictors model a grouped or
phylogenetic random-effect SD through `sd()`.

## Learning path

Start with the question you want to answer, not with the full list of features.
The tutorials are arranged so that each article pairs symbolic equations, R
syntax, fitted output, and interpretation.

The table below follows the same order as the rest of the site: check the
first fit, choose a family, work through the matching interpretation
tutorial, add structured dependence if you need it, then handle uncertainty,
diagnostics, and honest limits.

| If your question is... | Read this first | Main parameter |
| --- | --- | --- |
| After fitting, which checks and interval-status columns should I read? | [Checking and using fitted models](model-workflow.html) | `check_drm()`, `profile_targets()`, `conf.status` |
| Am I modelling residual variation, group-level variation, or likelihood weights? | [Which scale are you modelling?](which-scale.html) | `sigma`, `sd(group)`, `weights` |
| Which family fits my response — continuous, count, proportion, or robust? | [Choosing response families](distribution-families.html) | family-specific `mu`, `sigma`, `nu`, `zi`, `hu` |
| Do group or phylogenetic covariance rows answer a bivariate question beyond residual coupling? | [What can I fit today?](model-map.html) | `corpairs()`, `corpair()`, `sd1(group, level = "phylogenetic")`, `sd2(group, level = "phylogenetic")` |
| Do predictors change the mean and residual variability of one response? | [When variance carries signal, Part 1](location-scale.html) | `mu`, `sigma` |
| Do predictors change residual variability and the SD of a grouped or phylogenetic random effect? | [When variance carries signal, Part 2](location-scale-scale.html) | `mu`, `sigma`, `sd(group)` |
| Do counts show extra-Poisson variation or structural zeros? | [Count abundance and extra zeros](count-nbinom2.html) | NB2 `mu`, `sigma`, `zi` |
| Are proportions successes out of trials, continuous rates inside `(0, 1)`, or continuous rates with structural exact boundaries? | [Proportions and success rates](proportion-beta-binomial.html) | beta-binomial, beta, and zero-one beta `mu`, `sigma`, `zoi`, `coi` |
| Does a predictor change residual coupling between two responses? | [Changing residual coupling with rho12](bivariate-coscale.html) | `rho12` |
| Do repeated individuals have correlated average responses after accounting for residual coupling? | [Changing residual coupling with rho12](bivariate-coscale.html) | `corpairs(level = "group")` |
| Do effect sizes have known sampling variances or covariance? (a specialist route, not a family choice for raw response data) | [Mean effects and residual heterogeneity in meta-analysis](meta-analysis.html) | `meta_V(V = V)`, `sigma` |
| Do species trait means remain similar after accounting for shared ancestry, do sites share coordinate-structured deviations, or do individuals have known relatedness? | [Structural dependence overview](structural-dependence.html) | `animal()`, `phylo()`, `spatial()`, `relmat()` |
| I have more than one candidate fit — which one should I report? | [Model selection with AIC and BIC](model-selection.html) | `AIC()`, `BIC()` |
| Is the fitted distribution itself adequate, not just the mean structure? | [Distributional outputs and adequacy](distributional-outputs-and-adequacy.html) | `worm_plot()`, `qq_plot()`, `centile_chart()` |
| The optimizer did not converge cleanly — what do I check? | [Improving convergence](convergence.html) | `optimizer_convergence`, `fixed_gradient` |
| My fit is slow or memory-heavy — how do I keep fitting at this scale? | [Working with large data](large-data.html) | `keep_data`, `keep_model_frame`, `se = FALSE` |
| What does `drmTMB` not support yet, and where is the line between fitted, first-slice, and planned syntax? | [What drmTMB can and can't do](capability-and-limits.html) | fitted, first-slice, and planned status |

Here `nu` is a shape parameter, `zi` and `hu` are zero-process parameters, and
`sd(group)` refers to a group-level random-effect standard deviation rather than
the residual scale `sigma`. Use the model guide "What can I fit today?" when you
need the longer status map before choosing syntax.

For a first applied analysis, fit the simplest model that answers the question,
run `check_drm()`, and then read the coefficient table on the parameter scale
used by the model. For example, `sigma` coefficients are on a log scale in
Gaussian location-scale models, while `rho12(fit)` returns residual
correlations on the response scale.

For slope and variance-component questions, name the estimand before reporting
the number. A `mu` slope is an expected-response effect, a `sigma` slope is a
log residual-SD effect, a random-slope SD is among-group variation in a
reaction norm, and `sd(group) ~ x_group` is a model for the SD of a group-level
mean effect. Those four quantities can all involve a predictor, but they answer
different biological questions.

## What can I fit today?

The detailed implementation map now lives in the model guide "What can I fit
today?". This getting-started article keeps the first fitted example close to
the top, then sends you to the right guide or worked tutorial.

At a high level, the current public surface is:

| Model family or component | Current status | Main article |
| --- | --- | --- |
| Gaussian location-scale fixed effects | stable | "When variance carries signal, Part 1" |
| Gaussian `mu` random intercepts and one-slope blocks | stable | "When variance carries signal, Part 1" |
| residual-scale random intercepts and independent slopes | stable or first slice, depending on term | "Which scale are you modelling?" |
| unlabelled Gaussian `mu` random-intercept SD formulas | first slice | "Which scale are you modelling?" |
| fixed-effect non-Gaussian families | stable for named fixed-effect paths | "Choosing response families" |
| fixed-effect NB2 and zero-inflated NB2 count examples | stable for fixed effects | "Count abundance and extra zeros" |
| bounded-response proportion examples | beta, beta-binomial, and zero-one beta fixed effects plus ordinary `mu` random intercepts and independent numeric slopes; its exact slope cell is inference-ready with caveats for true SD 0.50 and M>=16; zero-one beta handles structural exact boundaries | "Proportions and success rates" |
| Gaussian meta-analysis with known sampling covariance | stable | "Mean effects and residual heterogeneity in meta-analysis" |
| fixed-effect bivariate Gaussian `rho12` | stable | "Changing residual coupling with rho12" |
| ordinary bivariate covariance and `corpairs()` rows | first slice | "What can I fit today?" |
| phylogenetic Gaussian `mu`, q=2, q=4, and direct-SD covariance slices | first slices | "Structural dependence" |
| coordinate spatial Gaussian `mu` intercept, one numeric slope, and bivariate q=2 location covariance | first slices | "Structural dependence" |
| fitted interval status and derived-summary flags | first slice | "Checking and using fitted models" |

After fitting any model, run `check_drm()` before interpreting the estimates:

```r
check_drm(fit)
```

The diagnostic table flags convergence, finite objective values, Hessian
status, fixed-effect standard errors, dropped rows, positive scale values,
near-zero random-effect SDs, residual `rho12` boundaries, Student-t `nu`
boundary behaviour, known sampling covariance summaries, random-effect
replication, and weak random-slope designs. A `note` marks something to
inspect; a `warning` or `error` should be resolved before treating estimates as
stable.

The key rule is to keep correlation layers separate. A bivariate residual
`rho12` is within-observation coupling after the two means and residual SDs are
modelled. A random-effect correlation is a group-level quantity. Phylogenetic
rows from `corpairs(..., level = "phylogenetic")` are structured-effect
quantities, and the first coordinate-spatial mean-mean row is reported by
`corpairs(..., level = "spatial")` when matching labelled spatial terms are
fitted in `mu1` and `mu2`. The constant coordinate-spatial q=4 location-scale
block is also fitted when matching labelled spatial terms appear in all four
`mu1`/`mu2`/`sigma1`/`sigma2` endpoints. Predictor-dependent spatial
`corpair()` rows remain planned. Those layers should not be reported as if they
were the same estimand.

Once this first fit is checked, follow the Learning path table above in the
order given, starting with [Checking and using fitted models](model-workflow.html).
