---
title: "Distributional regression with drmTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional regression with drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

`drmTMB` is being developed as a model-first package: the documentation should
pair symbolic equations with the R syntax that fits them. The current
implementation includes Gaussian location-scale models with fixed effects,
ordinary location and scale random effects, group-level random-effect scale
formulae such as `sd(id) ~ x_group`, fixed-effect non-Gaussian families,
known-covariance Gaussian meta-analysis, bivariate Gaussian residual coupling
with `rho12 ~ predictors`, first ordinary bivariate covariance blocks,
phylogenetic Gaussian covariance slices, and coordinate-based spatial `mu`
intercept and one-slope fields. The articles "What can I fit today?" and
"Choosing response families" give the broader implemented map.

The current package is already useful for one central workflow: write one
formula for each parameter you want to estimate, then check that the fitted
model matches the symbolic model you had in mind.

## Install the preview

`drmTMB` is not on CRAN yet. Install the tagged preview from GitHub with
`pak`:

```r
install.packages("pak")
pak::pak("itchyshin/drmTMB@v0.1.1")
```

For the newest development build from `main`, use:

```r
pak::pak("itchyshin/drmTMB")
```

You need R 4.1.0 or newer and a working compiler toolchain because TMB models
are compiled during installation. If installation fails while compiling C++,
install the usual R build tools for your platform: Rtools on Windows, Xcode
Command Line Tools on macOS, or the R development toolchain on Linux.

The core runtime dependencies are installed automatically by `pak`: `cli`,
`Matrix`, `TMB`, and the compiled headers from `RcppEigen` and `TMB`. The
articles and development checks also use optional packages such as `glmmTMB`,
`lme4`, `MASS`, `metafor`, `knitr`, `rmarkdown`, `testthat`, `withr`, and
`pkgdown`.

For example, an applied user might ask: do mean trait values and residual
variability change with an environmental predictor, after accounting for
repeated measures from the same site or species? In `drmTMB`, that question is
written as one formula for the mean and one formula for the residual scale.

## Fit your first model

Start with a Gaussian location-scale model when the response is continuous and
the scientific question is about both the expected value and predictability.
In the small example below, habitat and temperature affect mean growth, while
habitat also changes residual variation:

```{r first-model-data}
set.seed(13)
n <- 120
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)
mu <- 1 + 0.6 * (dat$habitat == "grassland") + 0.4 * dat$temperature
sigma <- exp(-0.5 + 0.45 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma)
```

The fitted model uses one formula for `mu` and one formula for `sigma`:

```{r first-model-fit}
fit <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)

check_drm(fit)
```

Read the `sigma` coefficient as a log residual-SD contrast. Exponentiating it
gives an SD ratio; exponentiating twice the coefficient gives a residual
variance ratio:

```{r first-model-interpret}
sigma_habitat <- coef(fit, "sigma")["habitatgrassland"]
data.frame(
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

For a fuller walkthrough of fitted means, residual SDs, and residual
variances, read "When variance carries signal".

## Learning path

Start with the question you want to answer, not with the full list of features.
The tutorials are arranged so that each article pairs symbolic equations, R
syntax, fitted output, and interpretation.

| If your question is... | Read this first | Main parameter |
| --- | --- | --- |
| Do predictors change the mean and residual variability of one response? | [When variance carries signal](location-scale.html) | `mu`, `sigma` |
| Am I modelling residual variation, group-level variation, or likelihood weights? | [Which scale are you modelling?](which-scale.html) | `sigma`, `sd(group)`, `weights` |
| Which continuous, count, proportion, or robust family should I use? | [Choosing response families](distribution-families.html) | family-specific `mu`, `sigma`, `nu`, `zi`, `hu` |
| Does a predictor change residual coupling between two responses? | [Changing residual coupling with rho12](bivariate-coscale.html) | `rho12` |
| Do effect sizes have known sampling variances or covariance? | [Mean effects and residual heterogeneity in meta-analysis](meta-analysis.html) | `meta_known_V(V = V)`, `sigma` |
| Do group or phylogenetic covariance rows answer a bivariate question beyond residual coupling? | [What can I fit today?](model-map.html) | `corpairs()`, `corpair()`, `sd_phylo*()` |
| Do species trait means remain similar after accounting for shared ancestry, or do sites share coordinate-structured deviations? | [Structural dependence](phylogenetic-spatial.html) | `phylo()`, `spatial()` |
| After fitting, which checks and interval-status columns should I read? | [Checking and using fitted models](model-workflow.html) | `check_drm()`, `profile_targets()`, `conf.status` |

Here `nu` is a shape parameter, `zi` and `hu` are zero-process parameters, and
`sd(group)` refers to a group-level random-effect standard deviation rather than
the residual scale `sigma`. Use the model guide "What can I fit today?" when you
need the longer status map before choosing syntax.

For a first applied analysis, fit the simplest model that answers the question,
run `check_drm()`, and then read the coefficient table on the parameter scale
used by the model. For example, `sigma` coefficients are on a log scale in
Gaussian location-scale models, while `rho12(fit)` returns residual
correlations on the response scale.

For slope and variance-component questions, name the estimand before reporting
the number. A `mu` slope is an expected-response effect, a `sigma` slope is a
log residual-SD effect, a random-slope SD is among-group variation in a
reaction norm, and `sd(group) ~ x_group` is a model for the SD of a group-level
mean effect. Those four quantities can all involve a predictor, but they answer
different biological questions.

## What can I fit today?

The detailed implementation map now lives in the model guide "What can I fit
today?". This getting-started article keeps the first fitted example close to
the top, then sends you to the right guide or worked tutorial.

At a high level, the current public surface is:

| Model family or component | Current status | Main article |
| --- | --- | --- |
| Gaussian location-scale fixed effects | implemented | "When variance carries signal" |
| Gaussian `mu` random intercepts and one-slope blocks | implemented | "When variance carries signal" |
| residual-scale random intercepts and independent slopes | implemented | "Which scale are you modelling?" |
| unlabelled Gaussian `mu` random-intercept SD formulas | implemented | "Which scale are you modelling?" |
| fixed-effect non-Gaussian families | implemented | "Choosing response families" |
| Gaussian meta-analysis with known sampling covariance | implemented | "Mean effects and residual heterogeneity in meta-analysis" |
| fixed-effect bivariate Gaussian `rho12` | implemented | "Changing residual coupling with rho12" |
| ordinary bivariate covariance and `corpairs()` rows | implemented first slice | "What can I fit today?" |
| phylogenetic Gaussian `mu`, q=2, q=4, and direct-SD covariance slices | implemented first slices | "Structural dependence" |
| coordinate spatial Gaussian `mu` intercept and one numeric slope | implemented first slice | "Structural dependence" |
| fitted interval status and derived-summary flags | implemented first slice | "Checking and using fitted models" |

After fitting any model, run `check_drm()` before interpreting the estimates:

```r
check_drm(fit)
```

The diagnostic table flags convergence, finite objective values, Hessian
status, fixed-effect standard errors, dropped rows, positive scale values,
near-zero random-effect SDs, residual `rho12` boundaries, Student-t `nu`
boundary behaviour, known sampling covariance summaries, random-effect
replication, and weak random-slope designs. A `note` marks something to
inspect; a `warning` or `error` should be resolved before treating estimates as
stable.

The key rule is to keep correlation layers separate. A bivariate residual
`rho12` is within-observation coupling after the two means and residual SDs are
modelled. A random-effect correlation is a group-level quantity. Phylogenetic
rows from `corpairs(..., level = "phylogenetic")` are structured-effect
quantities. Coordinate spatial fits currently report field SDs and conditional
modes; richer spatial correlation rows remain planned. Those layers should not
be reported as if they were the same estimand.
