---
title: "Distributional regression with drmTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional regression with drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

`drmTMB` fits distributional regression models for one or two responses. Start
here if you have one response and want to ask whether predictors change its
average, its residual variability, or both. This article fits a small Gaussian
model, explains the output, and shows what to check before interpretation.

The central workflow is simple: write one formula for each parameter you want
to estimate, fit the simplest model that answers the question, and check the
fit before interpreting it. If you instead have a phylogenetic tree, begin
with [Phylogenetic mixed models](phylogenetic-models.html). If your response is
an effect size with known sampling variance or covariance, begin with
[Mean effects and residual heterogeneity](meta-analysis.html). The broader
map, including other response types, is [What can I fit today?](model-map.html).

The parameter names stay consistent across the site. Location is a
family-specific centre or location parameter; for Gaussian models it is the
expected response, but it need not be the unconditional response mean. Scale
describes residual variability; shape describes distribution features beyond
location and scale; and coscale means residual correlation such as `rho12`
between two responses. A family may expose only a subset of these components.

For example, an applied user might ask: do mean trait values and residual
variability change with an environmental predictor, after accounting for
repeated measures from the same site or species? In `drmTMB`, that question is
written as one formula for the mean and one formula for the residual scale.

## Install

This vignette describes `drmTMB` 0.7.1, the current development version.
`drmTMB` 0.7.0 is the first CRAN release. Install it from CRAN:

```r
install.packages("drmTMB")
```

Or install the current development source from GitHub with `pak`:

```r
install.packages("pak")
pak::pak("itchyshin/drmTMB")
```

You need R 4.1.0 or newer and a working compiler toolchain because TMB models
are compiled during installation. If installation fails while compiling C++,
install the usual R build tools for your platform: Rtools on Windows, Xcode
Command Line Tools on macOS, or the R development toolchain on Linux.

The core runtime dependencies are installed automatically by `pak`: `cli`,
`Matrix`, `TMB`, and the compiled headers from `RcppEigen` and `TMB`. The
articles and development checks also use optional packages such as `glmmTMB`,
`lme4`, `MASS`, `metafor`, `knitr`, `rmarkdown`, `testthat`, `withr`, and
`pkgdown`.

## Fit your first model

Start with a Gaussian location-scale model when the response is continuous and
the scientific question is about both the expected value and predictability.
In the small example below, habitat and temperature affect mean growth, while
habitat also changes residual variation:

```{r first-model-data}
set.seed(13)
n <- 120
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)
mu <- 1 + 0.6 * (dat$habitat == "grassland") + 0.4 * dat$temperature
sigma <- exp(-0.5 + 0.45 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma)
```

The fitted model uses one formula for `mu` and one formula for `sigma`:

```{r first-model-fit}
fit <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)

check_drm(fit)
```

Read the `sigma` coefficient as a log residual-SD contrast. Exponentiating it
gives an SD ratio; exponentiating twice the coefficient gives a residual
variance ratio:

```{r first-model-interpret}
sigma_habitat <- coef(fit, "sigma")["habitatgrassland"]
data.frame(
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

For a fuller walkthrough of fitted means, residual SDs, and residual
variances, read [When variance carries signal, Part
1](location-scale.html). Continue to [Part
2](location-scale-scale.html) when predictors model a grouped or
phylogenetic random-effect SD through `sd()`.

## Choose your next guide

The first fit above is the right next step when your response is continuous and
you want to model its average and residual variation. Choose another guide only
when your data require it.

| If your question is... | Continue with | Key distinction |
| --- | --- | --- |
| Which response family fits continuous, count, proportion, robust, or zero-heavy data? | [Choosing response families](distribution-families.html) | A family determines what `mu`, `sigma`, and any extra parameters mean. |
| Do related species, sites, animals, or a supplied relationship matrix share deviations? | [Structural dependence overview](structural-dependence.html) | A structured deviation is not the same as residual variation or residual correlation. |
| Do two responses remain associated after their means and residual SDs are modelled? | [Changing residual coupling with rho12](bivariate-coscale.html) | `rho12` is residual coupling, not a correlation among groups or species. |
| Do effect sizes come with known sampling variances or covariance? | [Mean effects and residual heterogeneity](meta-analysis.html) | This is a Gaussian known-variance analysis, not a family for raw observations. |
| Is an estimate or interval safe to report? | [Can I fit and report this model?](capability-and-limits.html) | A model that fits can still have weak diagnostics or unsupported uncertainty. |

Use [What can I fit today?](model-map.html) only when you need a compact
route-and-limitation map. For any model, run `check_drm()` before interpreting
coefficients or intervals.

For a first applied analysis, fit the simplest model that answers the question,
run `check_drm()`, and then read the coefficient table on the parameter scale
used by the model. For example, `sigma` coefficients are on a log scale in
Gaussian location-scale models, while `rho12(fit)` returns residual
correlations on the response scale.

For slope and variance-component questions, name the estimand before reporting
the number. A `mu` slope is an expected-response effect, a `sigma` slope is a
log residual-SD effect, a random-slope SD is among-group variation in a
reaction norm, and `sd(group) ~ x_group` is a model for the SD of a group-level
mean effect. Those four quantities can all involve a predictor, but they answer
different biological questions.

## Check before interpreting

After fitting any model, run `check_drm()` before interpreting the estimates:

```r
check_drm(fit)
```

The diagnostic table checks convergence, gradients, Hessian status, standard
errors, dropped rows, scale values, random-effect replication, and relevant
parameter boundaries. Inspect a `note`; resolve a `warning` or `error` before
treating estimates as stable. The [errors, warnings, and convergence
guide](convergence.html) explains what to try next.

The key rule is to keep correlation layers separate. A bivariate residual
`rho12` is within-observation coupling after the two means and residual SDs are
modelled. A random-effect correlation is a group-level quantity. Phylogenetic
rows from `corpairs(..., level = "phylogenetic")` are structured-effect
quantities, and the first coordinate-spatial mean-mean row is reported by
`corpairs(..., level = "spatial")` when matching labelled spatial terms are
fitted in `mu1` and `mu2`. The constant coordinate-spatial q=4 location-scale
block is also fitted when matching labelled spatial terms appear in all four
`mu1`/`mu2`/`sigma1`/`sigma2` endpoints. Predictor-dependent spatial
`corpair()` rows remain planned. Those layers should not be reported as if they
were the same estimand.

Use [Can I fit and report this model?](capability-and-limits.html) for the
current reporting boundary and named fallback; use the [model map](model-map.html)
when you need syntax detail. Once this first fit is checked, continue with
[Checking and using fitted models](https://itchyshin.github.io/drmTMB/articles/model-workflow.html) or choose a scientific
question from the Learning path table above.
