---
title: "Distributional regression with drmTMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional regression with drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

`drmTMB` is being developed as a model-first package: the documentation should
pair symbolic equations with the R syntax that fits them. The current
implementation includes Gaussian location-scale models with fixed effects,
location random effects, residual-scale random intercepts and slopes, and
group-level random-effect scale formulae such as `sd(id) ~ x_group`. It also includes
fixed-effect Student-t, lognormal, Gamma, beta, Poisson, zero-inflated Poisson,
NB2, and zero-inflated NB2 models, known-covariance Gaussian meta-analysis,
intercept-only phylogenetic Gaussian location effects, and fixed-effect
bivariate Gaussian location-coscale models with `rho12 ~ predictors`. The
article "Choosing response families" gives the broader implemented family map.

The current package is already useful for one central workflow: write one
formula for each parameter you want to estimate, then check that the fitted
model matches the symbolic model you had in mind.

## Install the preview

`drmTMB` is not on CRAN yet. Install the tagged preview from GitHub with
`pak`:

```r
install.packages("pak")
pak::pak("itchyshin/drmTMB@v0.1.1")
```

For the newest development build from `main`, use:

```r
pak::pak("itchyshin/drmTMB")
```

You need R 4.1.0 or newer and a working compiler toolchain because TMB models
are compiled during installation. If installation fails while compiling C++,
install the usual R build tools for your platform: Rtools on Windows, Xcode
Command Line Tools on macOS, or the R development toolchain on Linux.

The core runtime dependencies are installed automatically by `pak`: `cli`,
`Matrix`, `TMB`, and the compiled headers from `RcppEigen` and `TMB`. The
articles and development checks also use optional packages such as `glmmTMB`,
`lme4`, `MASS`, `metafor`, `knitr`, `rmarkdown`, `testthat`, `withr`, and
`pkgdown`.

For example, an applied user might ask: do mean trait values and residual
variability change with an environmental predictor, after accounting for
repeated measures from the same site or species? In `drmTMB`, that question is
written as one formula for the mean and one formula for the residual scale.

## Fit your first model

Start with a Gaussian location-scale model when the response is continuous and
the scientific question is about both the expected value and predictability.
In the small example below, habitat and temperature affect mean growth, while
habitat also changes residual variation:

```{r first-model-data}
set.seed(13)
n <- 120
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)
mu <- 1 + 0.6 * (dat$habitat == "grassland") + 0.4 * dat$temperature
sigma <- exp(-0.5 + 0.45 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma)
```

The fitted model uses one formula for `mu` and one formula for `sigma`:

```{r first-model-fit}
fit <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)

check_drm(fit)
```

Read the `sigma` coefficient as a log residual-SD contrast. Exponentiating it
gives an SD ratio; exponentiating twice the coefficient gives a residual
variance ratio:

```{r first-model-interpret}
sigma_habitat <- coef(fit, "sigma")["habitatgrassland"]
data.frame(
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

For a fuller walkthrough of fitted means, residual SDs, and residual
variances, read "When variance carries signal".

## Learning path

Start with the question you want to answer, not with the full list of features.
The tutorials are arranged so that each article pairs symbolic equations, R
syntax, fitted output, and interpretation. The first route is usually a model
class. Biological case studies should become separate articles when they need
their own data preparation, figures, or multi-model narrative.

| Route | Start when your question is... | Read this first | Main parameters |
| --- | --- | --- | --- |
| One-response Gaussian | Do predictors change the mean and residual variability of one response? | "When variance carries signal" | `mu`, `sigma` |
| Scale audit | Am I modelling residual variation, group-level variation, or likelihood weights? | "Which scale are you modelling?" | `sigma`, `sd(group)`, `weights` |
| Response family choice | Which continuous, count, proportion, or robust family should I use? | "Choosing response families" | family-specific `mu`, `sigma`, `nu`, `zi`, `hu` |
| Two-response residual coupling | Does a predictor change residual coupling between two responses? | "Changing residual coupling with rho12" | `rho12` |
| Evidence synthesis | Do effect sizes have known sampling variances or covariance? | "Mean effects and residual heterogeneity in meta-analysis" | `meta_known_V(V = V)`, `sigma` |
| Structured dependence | Do species trait means remain similar after accounting for shared ancestry? | "Structured dependence: implemented phylogeny and planned spatial models" | `phylo()`, `sigma_phylo` |
| Post-fit workflow | How should I check, predict, simulate, and report a fitted model? | "Checking and using fitted models" | `check_drm()`, `predict()`, `simulate()` |

For a first applied analysis, fit the simplest model that answers the question,
run `check_drm()`, and then read the coefficient table on the parameter scale
used by the model. For example, `sigma` coefficients are on a log scale in
Gaussian location-scale models, while `rho12(fit)` returns residual
correlations on the response scale.

## Implemented now

For a fixed-effect Gaussian location-scale model:

\[
y_i \mid \mu_i, \sigma_i \sim
\operatorname{Normal}(\mu_i, \sigma_i^2)
\]

\[
\mu_i = \beta_0 + \beta_1 x1_i
\]

\[
\log(\sigma_i) = \gamma_0 + \gamma_1 x1_i
\]

the matching syntax is:

```r
drmTMB(
  drm_formula(y ~ x1, sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

Here `sigma_i` is the residual or within-observation standard deviation. The
location formula may also include ordinary grouped random effects. Adding an
independent random intercept and random slope for `id` gives:

\[
y_{ij} \mid \mu_{ij}, \sigma_{ij} \sim
\operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2)
\]

\[
\mu_{ij} =
\beta_0 + \beta_1 x1_{ij} + b_{0j} + b_{1j}x1_{ij}
\]

\[
\log(\sigma_{ij}) = \gamma_0 + \gamma_1 x1_{ij}
\]

\[
b_{0j} \sim
\operatorname{Normal}(0, sd_{\mu,\mathrm{id},0}^2),
\qquad
b_{1j} \sim
\operatorname{Normal}(0, sd_{\mu,\mathrm{id},1}^2)
\]

with matching syntax:

```r
drmTMB(
  drm_formula(y ~ x1 + (1 | id) + (0 + x1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

The separate `(1 | id)` and `(0 + x1 | id)` terms are independent in the
current implementation.

The ordinary correlated random intercept-slope form is also implemented for
univariate Gaussian `mu`:

\[
\mu_{ij} =
\beta_0 + \beta_1 x1_{ij} + b_{0j} + b_{1j}x1_{ij}
\]

\[
\begin{bmatrix}
b_{0j} \\
b_{1j}
\end{bmatrix}
\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
0 \\
0
\end{bmatrix},
\Sigma_{\mu,\mathrm{id}}
\right)
\]

```r
drmTMB(
  drm_formula(y ~ x1 + (1 + x1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

The intercept-slope correlation from this block is a group-level correlation
reported under `corpars$mu`; it is not residual bivariate `rho12`.

A labelled version is implemented for univariate Gaussian `mu`:

```r
drmTMB(
  drm_formula(y ~ x1 + (1 + x1 | p | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

The middle label `p` is retained in group-level SD and correlation names. In
one-response Gaussian models, the same label can span matching `mu` and
`sigma` intercept or one-slope blocks. In the bivariate Gaussian path, matching
`(1 | p | id)` terms in `mu1` and `mu2` now fit one cross-response
random-intercept covariance block; bivariate sharing across `sigma` formulas
remains planned.

Residual-scale random intercepts and slopes in `sigma` are also implemented:

```r
drmTMB(
  drm_formula(y ~ x1 + (1 | id), sigma ~ x1 + (1 | id)),
  family = gaussian(),
  data = dat
)
```

```r
drmTMB(
  drm_formula(y ~ x1, sigma ~ x1 + (1 + x1 | id)),
  family = gaussian(),
  data = dat
)
```

This is a model for residual or within-observation scale. It is distinct from
`sd(id) ~ x_group`, which models the standard deviation of an unlabelled
group-level `mu` random intercept.

Symbolically, those two scale models target different quantities:

\[
\log(\sigma_{ij}) =
X_{\sigma}[ij,]\beta_{\sigma} + a_j,
\qquad
a_j \sim \operatorname{Normal}(0, sd_{\sigma,\mathrm{id}}^2)
\]

\[
b_j = sd_{\mu,\mathrm{id},j}u_j,
\qquad
u_j \sim \operatorname{Normal}(0, 1),
\qquad
\log(sd_{\mu,\mathrm{id},j}) = W_{\mathrm{id}}[j,]\alpha_{\mathrm{id}}
\]

So `sigma ~ x + (1 | id)` models residual-scale heterogeneity, whereas
`sigma ~ x + (1 + x | id)` additionally models group-to-group differences in
how `x` changes residual scale. `sd(id) ~ x_group` models among-`id`
heterogeneity in the mean.

The first fitted phylogenetic path attaches a tree-structured random intercept
to the location model:

\[
y_i \mid \mu_i, \sigma_i \sim
\operatorname{Normal}(\mu_i, \sigma_i^2)
\]

\[
\mu_i = X_{\mu}[i,]\beta_{\mu} + a_{\mathrm{species}[i]},
\qquad
\log(\sigma_i) = X_{\sigma}[i,]\beta_{\sigma}
\]

\[
a \sim \operatorname{MVN}(0, \sigma_{\mathrm{phylo}}^2 A)
\]

with matching syntax:

```r
drmTMB(
  drm_formula(
    y ~ x1 + phylo(1 | species, tree = tree),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

The public `phylo()` path takes an ultrametric tree with branch lengths. Internally,
`drmTMB` uses the sparse augmented A-inverse route so the fitted model is a
phylogenetic random intercept, not a dense covariance matrix supplied by the
user.

The first implemented bivariate seed covers fixed-effect location-coscale
regression with predictor-dependent residual correlation. Here "coscale" means
the residual covariance structure; in the current bivariate Gaussian model,
that structure is the residual correlation `rho12`:

```r
drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The matching symbolic model is:

\[
\begin{bmatrix}
y_{1i} \\
y_{2i}
\end{bmatrix}
\sim
\operatorname{MVN}
\left(
\begin{bmatrix}
\mu_{1i} \\
\mu_{2i}
\end{bmatrix},
\Omega_i
\right)
\]

\[
\begin{aligned}
\mu_{1i} &= X_{\mu 1}[i,]\beta_{\mu 1}, \\
\mu_{2i} &= X_{\mu 2}[i,]\beta_{\mu 2}, \\
\log(\sigma_{1i}) &= X_{\sigma 1}[i,]\beta_{\sigma 1}, \\
\log(\sigma_{2i}) &= X_{\sigma 2}[i,]\beta_{\sigma 2}, \\
\eta_{\rho 12,i} &= X_{\rho 12}[i,]\beta_{\rho 12}, \\
\rho_{12i} &= \tanh(\eta_{\rho 12,i})
\end{aligned}
\]

\[
\Omega_i =
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}
\]

The fitted response-scale residual correlations are available with
`rho12(fit)`. Internally, the TMB likelihood applies a tiny numerical guard to
the `tanh()` transform so covariance matrices stay strictly positive definite;
that guard is an implementation detail, not part of the biological
interpretation.

After fitting any model, run `check_drm()` before interpreting the estimates:

```r
check_drm(fit)
```

The diagnostic table reports optimizer convergence, optimizer evaluation counts,
finite objective values, fixed-parameter gradients, Hessian status, finite
fixed-effect standard errors, dropped rows, positive scale values, random-effect
standard deviations near zero, residual `rho12` boundaries when relevant,
Student-t `nu` boundary behaviour, known sampling covariance summaries,
random-effect replication, and weak random-slope design checks. A `note`
records something to inspect, such as dropped rows, an optimizer budget that was
fully used, a singly observed random-effect level, a very large Student-t `nu`,
or a random-slope block with little within-group variation; a `warning` or
`error` should be resolved before treating estimates as stable.

Meta-analysis uses the ordinary Gaussian family plus a known sampling
covariance marker:

```r
drmTMB(
  bf(
    yi ~ x1 + x2 + meta_known_V(V = V),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

Use this when each observation has known sampling uncertainty, such as a set of
effect sizes from studies. The marker supplies known sampling variance or
covariance; the fitted `sigma` remains the extra heterogeneity SD.

For diagonal known sampling variances `vi`, this is:

\[
yi_i \mid \mu_i, \sigma_i, vi_i \sim
\operatorname{Normal}(\mu_i, vi_i + \sigma_i^2)
\]

For a dense or block-diagonal known covariance matrix `V`, it is:

\[
y \mid \mu, \sigma, V \sim
\operatorname{MVN}\left(\mu, V + \operatorname{diag}(\sigma_i^2)\right)
\]

Bivariate Gaussian meta-analysis uses the same idea, but the known sampling
covariance is row-paired so each study contributes both responses:

\[
y_{\mathrm{stack}} =
(y_{11}, y_{21}, \ldots, y_{1n}, y_{2n})'
\]

\[
y_{\mathrm{stack}} \sim
\operatorname{MVN}(\mu_{\mathrm{stack}}, V + \Omega_{\mathrm{stack}})
\]

\[
\Omega_i =
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}
\]

The matching syntax is:

```r
V <- meta_vcov_bivariate(v1 = v1, v2 = v2, cor12 = sampling_cor12)

drmTMB(
  bf(
    mu1 = y1 ~ x1 + meta_known_V(V = V),
    mu2 = y2 ~ x1,
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ x1
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

The known `V` describes within-study sampling uncertainty. The fitted `rho12`
describes the residual covariance component after that known uncertainty has
been included. It should only be called a study-level correlation when a
separate study-level random effect is part of the model.

For one Gaussian response, the fitted double-hierarchical covariance block uses
the same label in the `mu` and `sigma` formulas. The block can be intercept-only
or can include one numeric slope:

```r
drmTMB(
  formula = drm_formula(
    y ~ x + (1 + x | p | ID),
    sigma ~ x + (1 + x | p | ID)
  ),
  family = gaussian(),
  data = dat
)
```

This asks whether groups with higher expected response, or steeper response
slopes, also tend to have higher residual SD or steeper residual-SD slopes. The
fitted pairwise correlations are group-level mean-scale, slope-scale, and
scale-slope quantities in `corpars$mu_sigma` and `corpairs(fit)`, not residual
`rho12`.

## Next covariance steps

The first bivariate group-level covariance slice adds matching labelled random
intercepts to each mean formula. The bivariate double-hierarchical version
remains planned: it will add random slopes and residual-scale random effects,
following the
individual-differences literature on personality, plasticity, predictability,
and malleability:

```r
drmTMB(
  formula = drm_formula(
    mu1 = y1 ~ x1 + x2 + (1 | p | ID),
    mu2 = y2 ~ x1      + (1 | p | ID),
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Here `sigma1` and `sigma2` are residual scale parameters. The random-effect
standard deviations and correlation from `(1 | p | ID)` are separate
group-level quantities.

For the mathematical details, start with the location-scale and bivariate
coscale tutorials. They spell out which part of the R formula creates each
linear predictor, which scale each parameter uses, and which quantities are
residual scales versus group-level standard deviations.
