---
title: "Figure gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Figure gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This gallery shows the kinds of figures `drmTMB` users can build for model
interpretation, uncertainty display, fitted interactions, fitted correlation
summaries, and later simulation diagnostics. It is different from the internal
count-pilot diagnostics used while developing the first Poisson/NB2 simulation
plumbing.

The examples here use small simulated data so the page can build quickly. They
are meant to show figure patterns, not biological conclusions. The interaction
figures deliberately keep three pieces visible: the observed data, the fitted
surface or cell means, and the conditioning values used to make the display.

```{r setup}
library(drmTMB)
library(ggplot2)

theme_drmtmb_gallery <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey30"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

habitat_palette <- c("reef" = "#0072B2", "kelp" = "#D55E00")
scale_drmtmb_colour <- scale_colour_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
scale_drmtmb_fill <- scale_fill_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
```

## Fitted mean with a confidence band

Start with the figure most readers expect: raw observations, fitted mean, and a
95% confidence band on the response scale. This example is also a
categorical-by-continuous interaction: the temperature slope differs by
habitat.

```{r location-scale-fit, fig.alt="Scatterplot of simulated growth against temperature for two habitats, with fitted mean lines and 95 percent confidence bands for each habitat."}
set.seed(2591)
n <- 160
habitat_levels <- c("reef", "kelp")
fish <- data.frame(
  temperature = runif(n, -1.6, 1.6),
  habitat = factor(
    sample(habitat_levels, n, replace = TRUE),
    levels = habitat_levels
  )
)
eta_mu <- 0.4 +
  0.7 * fish$temperature +
  0.35 * (fish$habitat == "kelp") +
  0.18 * fish$temperature * (fish$habitat == "kelp")
eta_sigma <- -0.55 + 0.35 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit_growth <- drmTMB(
  bf(growth ~ temperature * habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)

grid_growth <- expand.grid(
  temperature = seq(-1.6, 1.6, length.out = 80),
  habitat = levels(fish$habitat)
)
pred_growth <- predict_parameters(
  fit_growth,
  newdata = grid_growth,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)
pred_growth$panel <- factor(
  ifelse(
    pred_growth$dpar == "mu",
    "Expected growth (mu)",
    "Residual SD (sigma)"
  ),
  levels = c("Expected growth (mu)", "Residual SD (sigma)")
)

pred_mu <- pred_growth[pred_growth$dpar == "mu", , drop = FALSE]

ggplot(fish, aes(temperature, growth, colour = habitat)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_ribbon(
    data = pred_mu,
    aes(
      x = temperature,
      y = estimate,
      ymin = conf.low,
      ymax = conf.high,
      fill = habitat
    ),
    inherit.aes = FALSE,
    alpha = 0.16,
    colour = NA
  ) +
  geom_line(data = pred_mu, aes(y = estimate), linewidth = 0.8) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Expected growth along a temperature gradient",
    subtitle = "Raw observations, fitted mean, and 95% confidence band",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Habitat",
    fill = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

## Distributional-parameter surfaces

The same fit can show more than the mean. `predict_parameters()` returns a long
table with one row per distributional parameter, and
`plot_parameter_surface()` draws estimate lines plus finite interval bands when
they are present.

```{r parameter-surface, fig.alt="Two-panel plot showing fitted expected growth and residual standard deviation surfaces along temperature for reef and kelp habitats."}
plot_parameter_surface(
  pred_growth,
  x = "temperature",
  colour = "habitat",
  dpar = c("mu", "sigma"),
  facet = "panel"
) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Location and residual-scale surfaces",
    subtitle = "Both panels use the same prediction grid",
    x = "Temperature anomaly",
    y = "Estimate on response scale",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

The `mu` panel is the expected response. The `sigma` panel is the fitted
residual standard deviation, so an upward line means lower predictability or
larger residual heterogeneity at warmer values.

Distributional parameters are not all residual standard deviations. Shape,
inflation, hurdle, and residual-correlation parameters answer different
questions and often live on bounded response scales. The next panel set uses
three fitted models and the same plotting helper, but each panel names the
estimand explicitly.

```{r shape-inflation-rho12-panels, fig.alt="Three-panel plot showing a Student-t degrees-of-freedom shape parameter, a zero-inflation probability, and a residual rho12 correlation along a predictor gradient with 95 percent confidence bands."}
set.seed(2594)
n_tail <- 2500
tail_dat <- data.frame(tail_pressure = runif(n_tail, -1.5, 1.5))
tail_nu <- 2 + exp(log(4) - tail_dat$tail_pressure)
tail_dat$robust_trait <- 0.35 +
  0.45 * tail_dat$tail_pressure +
  0.55 * stats::rt(n_tail, df = tail_nu)
fit_tail <- drmTMB(
  bf(robust_trait ~ tail_pressure, sigma ~ 1, nu ~ tail_pressure),
  family = student(),
  data = tail_dat
)
tail_grid <- data.frame(tail_pressure = seq(-1.5, 1.5, length.out = 80))
pred_tail <- predict_parameters(
  fit_tail,
  newdata = tail_grid,
  dpar = "nu",
  conf.int = TRUE
)
pred_tail$gradient <- pred_tail$tail_pressure
pred_tail$panel <- "Tail weight (Student-t nu)"
pred_tail$model <- "Tail weight"

set.seed(2595)
n_count <- 180
count_dat <- data.frame(detection_difficulty = runif(n_count, -1.2, 1.2))
mu_count <- exp(0.9 + 0.35 * count_dat$detection_difficulty)
zi_count <- plogis(-1.1 + 1.25 * count_dat$detection_difficulty)
count_dat$count <- ifelse(
  runif(n_count) < zi_count,
  0,
  stats::rpois(n_count, mu_count)
)
fit_zi <- drmTMB(
  bf(count ~ detection_difficulty, zi ~ detection_difficulty),
  family = poisson(link = "log"),
  data = count_dat
)
zi_grid <- data.frame(detection_difficulty = seq(-1.2, 1.2, length.out = 80))
pred_zi <- predict_parameters(
  fit_zi,
  newdata = zi_grid,
  dpar = "zi",
  conf.int = TRUE
)
pred_zi$gradient <- pred_zi$detection_difficulty
pred_zi$panel <- "Structural-zero probability (zi)"
pred_zi$model <- "Structural-zero probability"

set.seed(2596)
n_pair <- 160
pair_dat <- data.frame(disturbance = runif(n_pair, -1.4, 1.4))
rho_true <- tanh(-0.15 + 0.75 * pair_dat$disturbance)
z1 <- stats::rnorm(n_pair)
z2 <- stats::rnorm(n_pair)
pair_dat$activity <- 0.25 + 0.35 * pair_dat$disturbance + z1
pair_dat$boldness <- -0.1 +
  0.2 * pair_dat$disturbance +
  rho_true * z1 +
  sqrt(1 - rho_true^2) * z2
fit_pair <- drmTMB(
  bf(
    mu1 = activity ~ disturbance,
    mu2 = boldness ~ disturbance,
    sigma1 ~ 1,
    sigma2 ~ 1,
    rho12 ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = pair_dat
)
rho_grid <- data.frame(disturbance = seq(-1.4, 1.4, length.out = 80))
pred_rho <- predict_parameters(
  fit_pair,
  newdata = rho_grid,
  dpar = "rho12",
  conf.int = TRUE
)
pred_rho$gradient <- pred_rho$disturbance
pred_rho$panel <- "Residual correlation (rho12)"
pred_rho$model <- "Residual correlation"

pred_extra <- rbind(
  pred_tail[intersect(names(pred_tail), names(pred_zi))],
  pred_zi[intersect(names(pred_tail), names(pred_zi))],
  pred_rho[intersect(names(pred_tail), names(pred_zi))]
)
pred_extra$panel <- factor(
  pred_extra$panel,
  levels = c(
    "Tail weight (Student-t nu)",
    "Structural-zero probability (zi)",
    "Residual correlation (rho12)"
  )
)

plot_parameter_surface(
  pred_extra,
  x = "gradient",
  colour = "model",
  facet = "panel",
  dpar = c("nu", "zi", "rho12"),
  point = FALSE
) +
  geom_hline(
    data = data.frame(
      .drmTMB_plot_facet = factor(
        "Residual correlation (rho12)",
        levels = levels(pred_extra$panel)
      )
    ),
    aes(yintercept = 0),
    inherit.aes = FALSE,
    linetype = "dashed",
    colour = "grey55"
  ) +
  scale_colour_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  labs(
    title = "Shape, inflation, and residual-correlation panels",
    subtitle = "Each panel is a different distributional parameter on its response scale",
    x = "Predictor gradient",
    y = "Estimate on response scale",
    colour = "Parameter",
    fill = "Parameter"
  ) +
  theme_drmtmb_gallery() +
  theme(legend.position = "none")
```

The `nu` panel is a Student-t degrees-of-freedom parameter: lower values mean
heavier tails, and here the fitted model estimates a changing tail shape
through `nu ~ tail_pressure`. The `zi` panel is a structural-zero probability,
not a mean count. The `rho12` panel is residual coupling between two responses
after their own means and scales have been modelled.

## Discrete comparisons

For a factor predictor, the same plotting helper uses interval bars rather than
ribbons. This is useful for compact tutorial figures and estimated marginal
comparisons.

```{r discrete-comparison, fig.alt="Point interval plot comparing expected growth between reef and kelp habitats at average temperature."}
habitat_grid <- data.frame(
  temperature = 0,
  habitat = factor(c("reef", "kelp"), levels = levels(fish$habitat))
)
pred_habitat <- predict_parameters(
  fit_growth,
  newdata = habitat_grid,
  dpar = "mu",
  conf.int = TRUE
)

plot_parameter_surface(
  pred_habitat,
  x = "habitat",
  line = FALSE,
  point = TRUE
) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated habitat contrast at average temperature",
    subtitle = "Intervals are Wald intervals from predict_parameters()",
    x = NULL,
    y = "Expected growth"
  ) +
  theme_drmtmb_gallery()
```

## Estimated marginal means

For the first supported `emmeans` path, a fixed-effect univariate `mu` model can
produce estimated marginal means that are easy to turn into publication-style
figures. The same rule applies: keep the conditioning values visible.

```{r emmeans-display, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.alt="Point interval plot of estimated marginal mean growth for reef and kelp habitats conditioned at temperature zero."}
emm_habitat <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat,
      at = list(temperature = 0),
      type = "response"
    )
  )
)

ggplot(emm_habitat, aes(habitat, emmean, colour = habitat)) +
  geom_pointrange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    linewidth = 0.7,
    fatten = 2.4
  ) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated marginal mean growth",
    subtitle = "Conditioned at temperature = 0",
    x = NULL,
    y = "Expected growth",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

## Categorical by categorical interaction

For two categorical predictors, show every fitted cell and its interval. Use
position dodging so the uncertainty intervals do not sit on top of each other.
Raw observations can sit behind the intervals when they help readers see the
data support for each cell.

```{r cat-cat-interaction, fig.alt="Categorical by categorical interaction plot showing raw growth observations and fitted growth intervals for each habitat and season combination."}
set.seed(2592)
n_cell <- 30
reef <- expand.grid(
  habitat = factor(habitat_levels, levels = habitat_levels),
  season = factor(c("dry", "wet"), levels = c("dry", "wet")),
  replicate = seq_len(n_cell)
)
reef$mu <- with(
  reef,
  0.6 +
    0.35 * (habitat == "kelp") -
    0.15 * (season == "wet") +
    0.45 * (habitat == "kelp") * (season == "wet")
)
reef$growth <- rnorm(nrow(reef), mean = reef$mu, sd = 0.45)

fit_cat_cat <- drmTMB(
  bf(growth ~ habitat * season, sigma ~ 1),
  family = gaussian(),
  data = reef
)

grid_cat_cat <- expand.grid(
  habitat = levels(reef$habitat),
  season = levels(reef$season)
)
pred_cat_cat <- predict_parameters(
  fit_cat_cat,
  newdata = grid_cat_cat,
  dpar = "mu",
  conf.int = TRUE
)

raw_cat_cat <- reef

ggplot(pred_cat_cat, aes(habitat, estimate, colour = season)) +
  geom_jitter(
    data = raw_cat_cat,
    aes(y = growth, colour = season),
    width = 0.11,
    height = 0,
    alpha = 0.22,
    size = 1.1,
    show.legend = FALSE
  ) +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.35),
    linewidth = 0.65,
    fatten = 2.3
  ) +
  scale_colour_manual(values = c("dry" = "#0072B2", "wet" = "#E69F00")) +
  labs(
    title = "Categorical by categorical interaction",
    subtitle = "Small points are observations; intervals are fitted cell means",
    x = "Habitat",
    y = "Growth",
    colour = "Season"
  ) +
  theme_drmtmb_gallery()
```

## Continuous by continuous interaction

For two continuous predictors, choose scientifically meaningful slices of one
predictor and show fitted slopes across the other. The values should be named
in the legend so the reader can interpret the conditioning.

```{r cont-cont-interaction, fig.alt="Continuous by continuous interaction plot showing raw observations in grey and fitted temperature slopes at low, average, and high moisture with confidence bands."}
set.seed(2593)
n_cc <- 180
soil <- data.frame(
  temperature = runif(n_cc, -1.5, 1.5),
  moisture = runif(n_cc, -1.4, 1.4)
)
soil$growth <- rnorm(
  n_cc,
  mean = 0.3 +
    0.55 * soil$temperature +
    0.45 * soil$moisture -
    0.35 * soil$temperature * soil$moisture,
  sd = 0.45
)

fit_cont_cont <- drmTMB(
  bf(growth ~ temperature * moisture, sigma ~ 1),
  family = gaussian(),
  data = soil
)

grid_cont_cont <- expand.grid(
  temperature = seq(-1.5, 1.5, length.out = 80),
  moisture = c(-1, 0, 1)
)
grid_cont_cont$moisture_level <- factor(
  grid_cont_cont$moisture,
  levels = c(-1, 0, 1),
  labels = c("low moisture", "average moisture", "high moisture")
)

pred_cont_cont <- predict_parameters(
  fit_cont_cont,
  newdata = grid_cont_cont,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(
  pred_cont_cont,
  aes(temperature, estimate, colour = moisture_level, fill = moisture_level)
) +
  geom_point(
    data = soil,
    aes(temperature, growth),
    inherit.aes = FALSE,
    alpha = 0.18,
    size = 1,
    colour = "grey35"
  ) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.14, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  labs(
    title = "Continuous by continuous interaction",
    subtitle = "Moisture is held at three named slices",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Moisture slice",
    fill = "Moisture slice"
  ) +
  theme_drmtmb_gallery()
```

The raw point cloud is not coloured by the sliced predictor because the fitted
lines are conditional displays, not clusters in the data. This keeps the
visual message honest: the model is being read at three moisture values, while
the observations span the whole moisture gradient.

## Fitted correlation summaries

Correlation displays should keep layers separate. Residual `rho12`,
ordinary-group correlations, and phylogenetic or spatial correlations are not
the same parameter even when all are correlations.

```{r correlation-display, fig.alt="Point interval plot showing separate residual, group, and phylogenetic correlation layers with short row labels."}
pair_table <- data.frame(
  level = c("Residual", "Group", "Phylogenetic"),
  class = c("rho12", "random slope", "trait"),
  parameter = c("Trait coupling", "Intercept-slope", "Species covariance"),
  estimate = c(-0.32, 0.44, 0.61),
  modelled = c(TRUE, FALSE, FALSE),
  conf.low = c(-0.54, 0.12, 0.29),
  conf.high = c(-0.05, 0.68, 0.79),
  conf.status = c("wald", "profile", "profile")
)
pair_table$display <- factor(
  c(
    "Residual\nrho12",
    "Group\nintercept-slope",
    "Phylogenetic\ntrait covariance"
  ),
  levels = c(
    "Group\nintercept-slope",
    "Phylogenetic\ntrait covariance",
    "Residual\nrho12"
  )
)

plot_corpairs(pair_table, label = "display") +
  labs(
    title = "Correlation layers stay separate",
    subtitle = "Colour identifies the correlation layer; intervals are drawn only where available"
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.y = element_text(size = 10, lineheight = 0.95),
    legend.position = "none"
  )
```

## Simulation operating characteristics

Simulation figures answer different questions from model-interpretation
figures. They should show bias, error, coverage, runtime, and failures beside
the model surface being tested.

```{r simulation-operating-characteristics, fig.alt="Simulation summary plots showing illustrative bias and interval coverage by estimand across Gaussian location-scale, proportion, and count surfaces."}
sim_summary <- data.frame(
  surface = rep(
    c("Gaussian location-scale", "Proportion", "Count"),
    each = 4
  ),
  estimand = rep(c("beta_x", "sigma", "sd_intercept", "rho12"), 3),
  bias = c(
    0.01, -0.02, 0.04, 0.00,
    -0.01, 0.03, NA, NA,
    0.02, -0.01, -0.04, NA
  ),
  rmse = c(
    0.07, 0.09, 0.13, 0.11,
    0.08, 0.10, NA, NA,
    0.10, 0.09, 0.16, NA
  ),
  coverage = c(
    0.95, 0.94, 0.91, 0.93,
    0.94, 0.92, NA, NA,
    0.93, 0.95, 0.89, NA
  )
)

ggplot(sim_summary, aes(estimand, bias, colour = surface, group = surface)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 2.2, na.rm = TRUE) +
  geom_line(na.rm = TRUE) +
  coord_flip() +
  scale_colour_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Bias by estimand",
    x = NULL,
    y = "Mean estimate minus truth",
    colour = "Surface"
  ) +
  theme_drmtmb_gallery()

ggplot(sim_summary, aes(estimand, coverage, fill = surface)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey60") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, na.rm = TRUE) +
  coord_flip(ylim = c(0.75, 1)) +
  scale_fill_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Interval coverage",
    x = NULL,
    y = "Empirical coverage",
    fill = "Surface"
  ) +
  theme_drmtmb_gallery()
```

These figures use a small illustrative table with the same columns that
simulation summaries should carry. The comprehensive simulation programme
should not be count-only. It should add continuous, meta-analysis, proportion,
count, ordinal, bivariate, random-slope, and structured-dependence surfaces as
each surface becomes ready. Real simulation reports should also show the run
manifest and warning/error ledger so missing or failed fits are not hidden
behind clean-looking panels.

## Future simulation result articles

Specialised simulation result articles should come later, after the
comprehensive simulation surfaces are ready. Those articles can be organised by
scientific question: power, bias, coverage, runtime, convergence, and failure
patterns. They should also be organised across data types, not only counts:
continuous location-scale models, meta-analysis with known `V`, proportions,
counts, ordinal responses, bivariate `rho12`, random-slope surfaces, and
structured dependence.

This figure gallery is broader and lighter. It shows the visual language users
should see across tutorials, examples, model interpretation, and later
simulation reports.
