---
title: "Figure gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Figure gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This gallery shows the kinds of figures `drmTMB` users can build for model
interpretation, uncertainty display, fitted interactions, fitted correlation
summaries, and later simulation diagnostics. It is different from the internal
count-pilot diagnostics used while developing the first Poisson/NB2 simulation
plumbing. The panels apply the same location-scale reading pattern taught in
the family tutorials, such as [When variance carries
signal](location-scale.html); read one of those first for the modelling
context behind a given figure.

The examples here use small simulated data so the page can build quickly. They
are meant to show figure patterns, not biological conclusions. The interaction
figures deliberately keep three pieces visible: the observed data, the fitted
surface or cell means, and the conditioning values used to make the display.
Across the gallery, the right display depends on the figure's job. Raw data are
often warranted and sometimes preferred, especially when the reader needs to see
sample grain, spread, outliers, or repeated-measures structure. When the figure
makes an inferential claim, pair the estimate with uncertainty if the fitted
object or simulation output provides it: a slope estimate can be a line with a
confidence band, a cell mean can be a point with an interval, and selected
row-wise summaries can use Confidence Eyes. When uncertainty is unavailable or
outside the current support boundary, the figure should say so rather than
drawing a decorative interval.

The examples keep consistency within a figure family rather than forcing one
geometry everywhere. Raw-data displays, model-surface displays, row-wise
interval summaries, simulation summaries, and support-boundary strips each use
their own grammar because they answer different reader questions.

```{r setup}
library(drmTMB)
library(ggplot2)

theme_drmtmb_gallery <- function() {
  theme_minimal(base_size = 10.5) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(size = 10.5, colour = "grey15"),
      axis.text = element_text(size = 9.5, colour = "grey30"),
      plot.title.position = "plot",
      plot.title = element_text(size = 12.5, face = "bold"),
      plot.subtitle = element_text(size = 10.2, colour = "grey30"),
      strip.text = element_text(size = 10, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(colour = "grey20"),
      legend.text = element_text(size = 9.5),
      legend.key.width = grid::unit(1.0, "lines"),
      plot.margin = margin(5.5, 9, 5.5, 5.5)
    )
}
theme_confidence_eye <- function() {
  theme_drmtmb_gallery() +
    theme(
      axis.line.x = element_line(colour = "grey35", linewidth = 0.35),
      axis.ticks.x = element_line(colour = "grey35", linewidth = 0.35),
      axis.ticks.length.x = grid::unit(3, "pt"),
      panel.grid.major.x = element_line(colour = "grey88", linewidth = 0.35),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
}
theme_point_interval_gallery <- function() {
  theme_drmtmb_gallery() +
    theme(
      panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.35),
      panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
      axis.title = element_text(size = 10, colour = "grey15"),
      axis.text = element_text(size = 10, colour = "grey30"),
      plot.title = element_text(size = 12, face = "bold"),
      plot.subtitle = element_text(size = 10, colour = "grey30"),
      plot.margin = margin(5.5, 10, 5.5, 5.5)
    )
}

gallery_colours <- c(
  blue = "#0072B2",
  green = "#009E73",
  orange = "#D55E00",
  yellow = "#E69F00",
  pink = "#CC79A7",
  sky = "#56B4E9",
  charcoal = "#4D4D4D"
)
gallery_colour <- function(name) unname(gallery_colours[name])

habitat_palette <- c(
  "reef" = gallery_colour("blue"),
  "kelp" = gallery_colour("orange")
)
season_palette <- c(
  "dry" = gallery_colour("blue"),
  "wet" = gallery_colour("yellow")
)
surface_palette <- c(
  "Gaussian location-scale" = gallery_colour("blue"),
  "Proportion" = gallery_colour("green"),
  "Count" = gallery_colour("pink")
)
make_confidence_eye_distribution <- function(
  parameter,
  estimate,
  std_error,
  transform = c("identity", "exp", "tanh"),
  null_value = 0,
  level = 0.95,
  n = 500
) {
  transform <- match.arg(transform)
  cutoff <- 0.5 * qchisq(level, df = 1)
  theta <- seq(
    estimate - sqrt(2 * cutoff) * std_error,
    estimate + sqrt(2 * cutoff) * std_error,
    length.out = n
  )
  value <- switch(
    transform,
    identity = theta,
    exp = exp(theta),
    tanh = tanh(theta)
  )
  height <- pmax(cutoff - 0.5 * ((theta - estimate) / std_error)^2, 0) /
    cutoff
  data.frame(
    parameter = parameter,
    theta = theta,
    value = value,
    cdf = pnorm(theta, mean = estimate, sd = std_error),
    height = height,
    null_value = switch(
      transform,
      identity = null_value,
      exp = exp(null_value),
      tanh = tanh(null_value)
    ),
    estimate_value = switch(
      transform,
      identity = estimate,
      exp = exp(estimate),
      tanh = tanh(estimate)
    )
  )
}
guard_rho <- function(x) pmin(pmax(x, -0.999999), 0.999999)
response_scale_se_to_log_se <- function(estimate, std_error) {
  ifelse(is.finite(estimate) & estimate > 0, std_error / estimate, NA_real_)
}
response_scale_se_to_z_se <- function(estimate, std_error) {
  rho <- guard_rho(estimate)
  std_error / pmax(1 - rho^2, .Machine$double.eps)
}

scale_drmtmb_colour <- scale_colour_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
scale_drmtmb_fill <- scale_fill_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
```

## Fitted mean with a confidence band

Start with the figure most readers expect: raw observations, fitted mean, and a
95% confidence band on the response scale. This example is also a
categorical-by-continuous interaction: the temperature slope differs by
habitat.

```{r location-scale-fit, fig.cap="Raw observations and fitted `mu` surfaces for a Gaussian location-scale model; ribbons are 95% Wald confidence bands from `predict_parameters()`.", fig.alt="Scatterplot of simulated growth against temperature for two habitats, with fitted mean lines and 95 percent confidence bands for each habitat."}
set.seed(2591)
n <- 160
habitat_levels <- c("reef", "kelp")
fish <- data.frame(
  temperature = runif(n, -1.6, 1.6),
  habitat = factor(
    sample(habitat_levels, n, replace = TRUE),
    levels = habitat_levels
  )
)
eta_mu <- 0.4 +
  0.7 * fish$temperature +
  0.35 * (fish$habitat == "kelp") +
  0.18 * fish$temperature * (fish$habitat == "kelp")
eta_sigma <- -0.55 + 0.35 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit_growth <- drmTMB(
  bf(growth ~ temperature * habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)

grid_growth <- expand.grid(
  temperature = seq(-1.6, 1.6, length.out = 80),
  habitat = levels(fish$habitat)
)
pred_growth <- predict_parameters(
  fit_growth,
  newdata = grid_growth,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)
pred_growth$panel <- factor(
  ifelse(
    pred_growth$dpar == "mu",
    "Expected growth (mu)",
    "Residual SD (sigma)"
  ),
  levels = c("Expected growth (mu)", "Residual SD (sigma)")
)

pred_mu <- pred_growth[pred_growth$dpar == "mu", , drop = FALSE]
pred_sigma <- pred_growth[pred_growth$dpar == "sigma", , drop = FALSE]

ggplot(fish, aes(temperature, growth, colour = habitat)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_ribbon(
    data = pred_mu,
    aes(
      x = temperature,
      y = estimate,
      ymin = conf.low,
      ymax = conf.high,
      fill = habitat
    ),
    inherit.aes = FALSE,
    alpha = 0.16,
    colour = NA
  ) +
  geom_line(data = pred_mu, aes(y = estimate), linewidth = 0.8) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Expected growth along a temperature gradient",
    subtitle = "Raw observations, fitted mean, and 95% confidence band",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Habitat",
    fill = "Habitat"
  ) +
  theme_drmtmb_gallery() +
  guides(fill = "none")
```

## Distributional-parameter surfaces

The same fit can show more than the mean. `predict_parameters()` returns a long
table with one row per distributional parameter, and
`plot_parameter_surface()` draws estimate lines plus finite interval bands when
they are present.

```{r parameter-surface, fig.cap="Fitted response-scale `mu` and `sigma` surfaces; ribbons are 95% Wald confidence bands from `predict_parameters()`, and `sigma` is shared across habitats because the scale formula contains temperature but not habitat.", fig.alt="Two-panel line and ribbon plot showing fitted expected growth surfaces by habitat and a shared residual standard deviation surface along temperature, with Wald confidence bands."}
pred_surface <- rbind(
  pred_mu,
  pred_sigma[pred_sigma$habitat == levels(fish$habitat)[1], , drop = FALSE]
)
pred_surface$surface_group <- as.character(pred_surface$habitat)
pred_surface$surface_group[pred_surface$dpar == "sigma"] <- "shared sigma"
surface_group_palette <- c(
  habitat_palette,
  "shared sigma" = gallery_colour("green")
)

plot_parameter_surface(
  pred_surface,
  x = "temperature",
  colour = "surface_group",
  dpar = c("mu", "sigma"),
  facet = "panel",
  point = FALSE
) +
  scale_colour_manual(
    values = surface_group_palette,
    breaks = c("reef", "kelp", "shared sigma")
  ) +
  scale_fill_manual(
    values = surface_group_palette,
    breaks = c("reef", "kelp", "shared sigma")
  ) +
  labs(
    title = "Mean and residual-scale surfaces",
    subtitle = "Wald bands from predict_parameters(); sigma is shared",
    x = "Temperature anomaly",
    y = "Estimate on response scale",
    colour = NULL
  ) +
  theme_drmtmb_gallery()
```

The `mu` panel is the expected response. The `sigma` panel is the fitted
residual standard deviation, so an upward line means lower predictability or
larger residual heterogeneity at warmer values.

For scale models, the observed-data analogue is not the raw response itself.
Raw `growth` lives on the location scale. A useful scale check is the residual
magnitude after subtracting the fitted `mu`: under a Gaussian model,
`E(|residual|) = sigma * sqrt(2 / pi)`. This keeps the observed scatter visible
without putting raw response points on a `sigma` axis.

```{r residual-scale-observed-check, fig.cap="Observed absolute residual magnitudes compared with the fitted expectation implied by the response-scale `sigma` surface.", fig.alt="Scatterplot of absolute residual magnitudes against temperature, with a fitted expected absolute residual curve derived from the residual sigma surface."}
pred_row_growth <- predict_parameters(
  fit_growth,
  newdata = fish,
  dpar = c("mu", "sigma")
)
row_mu <- pred_row_growth[pred_row_growth$dpar == "mu", c("row", "estimate")]
row_sigma <- pred_row_growth[
  pred_row_growth$dpar == "sigma",
  c("row", "estimate")
]
names(row_mu)[2] <- "fitted_mu"
names(row_sigma)[2] <- "fitted_sigma"
scale_check <- merge(row_mu, row_sigma, by = "row")
scale_check <- cbind(fish[scale_check$row, ], scale_check)
scale_check$abs_residual <- abs(scale_check$growth - scale_check$fitted_mu)

pred_sigma$expected_abs_residual <- pred_sigma$estimate * sqrt(2 / pi)
pred_sigma$quantity <- "Expected absolute residual"

ggplot(scale_check, aes(temperature, abs_residual, colour = habitat)) +
  geom_point(alpha = 0.22, size = 1.2) +
  geom_line(
    data = pred_sigma,
    aes(
      y = expected_abs_residual,
      colour = quantity,
      group = quantity
    ),
    linewidth = 0.9
  ) +
  scale_colour_manual(values = c(
    habitat_palette,
    "Expected absolute residual" = gallery_colour("green")
  )) +
  labs(
    title = "Residual magnitudes carry the scale information",
    subtitle = "Points are absolute residuals; the green curve is sigma * sqrt(2 / pi)",
    x = "Temperature anomaly",
    y = "Absolute residual",
    colour = NULL
  ) +
  coord_cartesian(ylim = c(0, NA)) +
  theme_drmtmb_gallery()
```

Confidence intervals are only one slice through the uncertainty surface. This
gallery calls the fuller display a Confidence Eye: a pale confidence region
with a hollow point-estimate circle. The idea is close to
[Raindrop plots](https://doi.org/10.1198/0003130032369), but the name and use
are deliberately distinct. A Confidence Eye is not a posterior density; it is a
frequentist compatibility display that gives confidence intervals a visible
shape without adding interval bars, filled estimate dots, or outlines by
default. The code below uses a normal likelihood approximation from the estimate
and standard error, draws the finite 95% confidence region, and keeps the no-effect
line and hollow estimate circle visible together. Both rows are shown on the
fitted coefficient
scale so the no-effect line has the same meaning; the `sigma` row is a log-SD
effect, so positive values mean one unit of temperature increases residual
scale. This shared-axis display is only appropriate when the predictor scale is
common or standardized; coefficients for different predictors or different
units should be faceted, standardized, or converted to a named contrast before
visual comparison.

```{r confidence-distribution-slopes, fig.width=7.2, fig.height=3.1, fig.cap="Confidence Eye display for fixed-effect slope estimates; pale regions are finite 95% Wald confidence regions and hollow circles are point estimates.", fig.alt="Confidence Eye rows for the fitted mu temperature slope and log residual SD temperature effect, with a shared no-effect reference line, pale finite confidence regions, and hollow point-estimate circles."}
coef_growth <- summary(fit_growth)$coefficients
confidence_curves <- rbind(
  make_confidence_eye_distribution(
    parameter = "mu temperature slope",
    estimate = coef_growth["mu:temperature", "estimate"],
    std_error = coef_growth["mu:temperature", "std_error"],
    transform = "identity",
    null_value = 0
  ),
  make_confidence_eye_distribution(
    parameter = "sigma temperature log-SD effect",
    estimate = coef_growth["sigma:temperature", "estimate"],
    std_error = coef_growth["sigma:temperature", "std_error"],
    transform = "identity",
    null_value = 0
  )
)
confidence_curves$parameter <- factor(
  confidence_curves$parameter,
  levels = rev(c("mu temperature slope", "sigma temperature log-SD effect"))
)
confidence_curves$y <- as.numeric(confidence_curves$parameter)
confidence_curves$drop_height <- 0.17 * confidence_curves$height
confidence_reference <- unique(
  confidence_curves[c("parameter", "null_value", "estimate_value")]
)
confidence_reference$y <- as.numeric(confidence_reference$parameter)
confidence_palette <- c(
  "mu temperature slope" = gallery_colour("blue"),
  "sigma temperature log-SD effect" = gallery_colour("green")
)

ggplot(confidence_curves, aes(value, ymin = y - drop_height, ymax = y + drop_height)) +
  geom_ribbon(aes(fill = parameter), alpha = 0.24, colour = NA) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = "grey45",
    linewidth = 0.45
  ) +
  geom_point(
    data = confidence_reference,
    aes(x = estimate_value, y = y, colour = parameter),
    inherit.aes = FALSE,
    shape = 21,
    fill = "white",
    size = 3.1,
    stroke = 1.1
  ) +
  scale_colour_manual(values = confidence_palette) +
  scale_fill_manual(values = confidence_palette) +
  scale_y_continuous(
    breaks = seq_along(levels(confidence_curves$parameter)),
    labels = levels(confidence_curves$parameter),
    expand = expansion(mult = c(0.28, 0.32))
  ) +
  scale_x_continuous(
    breaks = c(0, 0.25, 0.50, 0.75),
    expand = expansion(mult = 0.04)
  ) +
  coord_cartesian(xlim = c(-0.05, 0.82)) +
  labs(
    title = NULL,
    subtitle = NULL,
    x = "Coefficient",
    y = NULL,
    colour = "Parameter"
  ) +
  theme_confidence_eye() +
  theme(
    legend.position = "none"
  )
```

Distributional parameters are not all residual standard deviations. Shape,
inflation, hurdle, and residual-correlation parameters answer different
questions and often live on bounded response scales. The next panel set uses
three fitted models and the same plotting helper, but each panel names the
estimand explicitly.

```{r shape-inflation-rho12-panels, fig.cap="Response-scale distributional-parameter panels for `nu`, `zi`, and residual `rho12`; ribbons are 95% Wald confidence bands.", fig.alt="Three-panel plot showing a Student-t degrees-of-freedom shape parameter, a zero-inflation probability, and a residual rho12 correlation along a predictor gradient with 95 percent confidence bands."}
set.seed(2594)
n_tail <- 2500
tail_dat <- data.frame(tail_pressure = runif(n_tail, -1.5, 1.5))
tail_nu <- 2 + exp(log(4) - tail_dat$tail_pressure)
tail_dat$robust_trait <- 0.35 +
  0.45 * tail_dat$tail_pressure +
  0.55 * stats::rt(n_tail, df = tail_nu)
fit_tail <- drmTMB(
  bf(robust_trait ~ tail_pressure, sigma ~ 1, nu ~ tail_pressure),
  family = student(),
  data = tail_dat
)
tail_grid <- data.frame(tail_pressure = seq(-1.5, 1.5, length.out = 80))
pred_tail <- predict_parameters(
  fit_tail,
  newdata = tail_grid,
  dpar = "nu",
  conf.int = TRUE
)
pred_tail$gradient <- pred_tail$tail_pressure
pred_tail$panel <- "Tail weight (nu)"
pred_tail$model <- "Tail weight"

set.seed(2595)
n_count <- 180
count_dat <- data.frame(detection_difficulty = runif(n_count, -1.2, 1.2))
mu_count <- exp(0.9 + 0.35 * count_dat$detection_difficulty)
zi_count <- plogis(-1.1 + 1.25 * count_dat$detection_difficulty)
count_dat$count <- ifelse(
  runif(n_count) < zi_count,
  0,
  stats::rpois(n_count, mu_count)
)
fit_zi <- drmTMB(
  bf(count ~ detection_difficulty, zi ~ detection_difficulty),
  family = poisson(link = "log"),
  data = count_dat
)
zi_grid <- data.frame(detection_difficulty = seq(-1.2, 1.2, length.out = 80))
pred_zi <- predict_parameters(
  fit_zi,
  newdata = zi_grid,
  dpar = "zi",
  conf.int = TRUE
)
pred_zi$gradient <- pred_zi$detection_difficulty
pred_zi$panel <- "Structural zero (zi)"
pred_zi$model <- "Structural-zero probability"

set.seed(2596)
n_pair <- 160
pair_dat <- data.frame(disturbance = runif(n_pair, -1.4, 1.4))
rho_true <- tanh(-0.15 + 0.75 * pair_dat$disturbance)
z1 <- stats::rnorm(n_pair)
z2 <- stats::rnorm(n_pair)
pair_dat$activity <- 0.25 + 0.35 * pair_dat$disturbance + z1
pair_dat$boldness <- -0.1 +
  0.2 * pair_dat$disturbance +
  rho_true * z1 +
  sqrt(1 - rho_true^2) * z2
fit_pair <- drmTMB(
  bf(
    mu1 = activity ~ disturbance,
    mu2 = boldness ~ disturbance,
    sigma1 ~ 1,
    sigma2 ~ 1,
    rho12 ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = pair_dat
)
rho_grid <- data.frame(disturbance = seq(-1.4, 1.4, length.out = 80))
pred_rho <- predict_parameters(
  fit_pair,
  newdata = rho_grid,
  dpar = "rho12",
  conf.int = TRUE
)
pred_rho$gradient <- pred_rho$disturbance
pred_rho$panel <- "Residual rho12"
pred_rho$model <- "Residual correlation"

pred_extra <- rbind(
  pred_tail[intersect(names(pred_tail), names(pred_zi))],
  pred_zi[intersect(names(pred_tail), names(pred_zi))],
  pred_rho[intersect(names(pred_tail), names(pred_zi))]
)
pred_extra$panel <- factor(
  pred_extra$panel,
  levels = c(
    "Tail weight (nu)",
    "Structural zero (zi)",
    "Residual rho12"
  )
)

plot_parameter_surface(
  pred_extra,
  x = "gradient",
  colour = "model",
  facet = "panel",
  dpar = c("nu", "zi", "rho12"),
  point = FALSE
) +
  geom_hline(
    data = data.frame(
      .drmTMB_plot_facet = factor(
        "Residual rho12",
        levels = levels(pred_extra$panel)
      )
    ),
    aes(yintercept = 0),
    inherit.aes = FALSE,
    linetype = "dashed",
    colour = "grey55"
  ) +
  scale_colour_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  labs(
    title = "Shape, inflation, and residual-correlation panels",
    subtitle = "Each panel is a different distributional parameter on its response scale",
    x = "Predictor gradient",
    y = "Estimate on response scale",
    colour = "Parameter",
    fill = "Parameter"
  ) +
  theme_drmtmb_gallery() +
  theme(legend.position = "none")
```

The `nu` panel is a Student-t degrees-of-freedom parameter: lower values mean
heavier tails, and here the fitted model estimates a changing tail shape
through `nu ~ tail_pressure`. The `zi` panel is a structural-zero probability,
not a mean count. The `rho12` panel is residual coupling between two responses
after their own means and scales have been modelled.

## Random-effect and variance-component figures

Random-effect figures need a different visual grammar from residual-scale
figures. Residual `sigma` is the observation-level standard deviation. A
group-level SD, such as the SD of site intercepts or temperature slopes, is
among-group variation in a fitted random effect.

```{r random-effect-variance-components, fig.height=3.05, fig.cap="Confidence Eye display for residual and among-site standard deviations; SD intervals are shaped on the fitted log-SD scale and displayed as SDs.", fig.alt="Confidence Eye plot comparing residual sigma, among-site intercept SD, and among-site temperature-slope SD, with pale finite Wald confidence regions and hollow point-estimate circles."}
set.seed(2621)
n_site <- 24
n_site_obs <- 6
site_levels <- seq_len(n_site)
site_random <- data.frame(
  site = factor(site_levels),
  intercept = rnorm(n_site, mean = 0, sd = 0.42),
  slope = rnorm(n_site, mean = 0, sd = 0.24)
)
reef_re <- data.frame(
  site = factor(rep(site_levels, each = n_site_obs), levels = site_levels)
)
reef_re$temperature <- rep(seq(-1, 1, length.out = n_site_obs), n_site)
site_lookup <- as.integer(reef_re$site)
reef_re$growth <- 0.25 +
  0.62 * reef_re$temperature +
  site_random$intercept[site_lookup] +
  site_random$slope[site_lookup] * reef_re$temperature +
  rnorm(nrow(reef_re), mean = 0, sd = 0.34)

fit_re_slopes <- drmTMB(
  bf(growth ~ temperature + (1 + temperature | site), sigma ~ 1),
  family = gaussian(),
  data = reef_re,
  control = drm_control(optimizer = list(eval.max = 160L, iter.max = 160L))
)

component_rows <- summary(fit_re_slopes)$parameters
variance_components <- data.frame(
  component = c(
    "Residual sigma",
    "Site intercept SD",
    "Site temperature-slope SD"
  ),
  component_type = c(
    "Residual scale",
    "Group intercept SD",
    "Group slope SD"
  ),
  estimate = c(
    component_rows["sigma", "estimate"],
    component_rows[
      "sd:mu:(1 + temperature | site):(Intercept)",
      "estimate"
    ],
    component_rows[
      "sd:mu:(1 + temperature | site):temperature",
      "estimate"
    ]
  ),
  response_std_error = c(
    component_rows["sigma", "std_error"],
    component_rows[
      "sd:mu:(1 + temperature | site):(Intercept)",
      "std_error"
    ],
    component_rows[
      "sd:mu:(1 + temperature | site):temperature",
      "std_error"
    ]
  )
)
variance_components$link_estimate <- log(variance_components$estimate)
variance_components$link_std_error <- response_scale_se_to_log_se(
  variance_components$estimate,
  variance_components$response_std_error
)
variance_components <- variance_components[
  is.finite(variance_components$link_std_error),
]
variance_components$component <- factor(
  variance_components$component,
  levels = rev(variance_components$component)
)
variance_components$component_type <- factor(
  variance_components$component_type,
  levels = c("Residual scale", "Group intercept SD", "Group slope SD")
)

component_palette <- c(
  "Residual scale" = gallery_colour("blue"),
  "Group intercept SD" = gallery_colour("green"),
  "Group slope SD" = gallery_colour("orange")
)

component_cloud <- do.call(
  rbind,
  lapply(seq_len(nrow(variance_components)), function(i) {
    row <- variance_components[i, ]
    make_confidence_eye_distribution(
      parameter = row$component,
      estimate = row$link_estimate,
      std_error = row$link_std_error,
      transform = "exp"
    )
  })
)
component_cloud <- merge(
  component_cloud,
  variance_components[c("component", "component_type")],
  by.x = "parameter",
  by.y = "component"
)
component_cloud$parameter <- factor(
  component_cloud$parameter,
  levels = levels(variance_components$component)
)
component_cloud$y_index <- as.numeric(component_cloud$parameter)
component_cloud$ymin <- component_cloud$y_index -
  0.17 * component_cloud$height
component_cloud$ymax <- component_cloud$y_index +
  0.17 * component_cloud$height
variance_components$y_index <- as.numeric(variance_components$component)

ggplot() +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = "grey45",
    linewidth = 0.45
  ) +
  geom_ribbon(
    data = component_cloud,
    aes(
      value,
      ymin = ymin,
      ymax = ymax,
      fill = component_type,
      group = parameter
    ),
    alpha = 0.24,
    colour = NA
  ) +
  geom_point(
    data = variance_components,
    aes(estimate, y_index, colour = component_type),
    shape = 21,
    fill = "white",
    size = 3.1,
    stroke = 1.1
  ) +
  scale_colour_manual(values = component_palette) +
  scale_fill_manual(values = component_palette) +
  scale_y_continuous(
    breaks = seq_along(levels(variance_components$component)),
    labels = levels(variance_components$component),
    expand = expansion(add = 0.35)
  ) +
  scale_x_continuous(
    breaks = c(0, 0.1, 0.2, 0.3, 0.4),
    expand = expansion(mult = c(0, 0.04))
  ) +
  coord_cartesian(xlim = c(-0.02, 0.48)) +
  labs(
    title = NULL,
    subtitle = NULL,
    x = "Standard deviation",
    y = NULL,
    colour = "Component"
  ) +
  theme_confidence_eye() +
  guides(colour = "none", fill = "none")
```

The plot compares three standard deviations, but only one is residual `sigma`.
The other two are random-effect SDs: how much sites differ in baseline growth,
and how much sites differ in the temperature slope. The eyes are fast Wald
summaries on the log-SD scale. For slower likelihood-profile intervals on these
SDs, use `profile_targets(fit_re_slopes)` to choose direct profile-ready
targets rather than profiling every fixed effect by default.

Readers used to Bayesian model summaries often expect one compact display that
keeps fixed effects, standard deviations, and correlations together. For this
frequentist fit, the Confidence Eye keeps the confidence region visually
subordinate to the estimate: the pale eye is the finite 95% Wald confidence
region and the hollow circle is the point estimate. This compact panel is a
row-by-row reading aid rather than a magnitude comparison across parameter
classes. Fixed effects use their fitted coefficient scale, SDs are shaped on the
log-SD scale and shown back as SDs, and the correlation is shaped on Fisher's
`z` scale and shown back as a correlation.

```{r coefficient-intervals, fig.height=4.4, fig.cap="Mixed-parameter Confidence Eye display; fixed effects use fitted-scale Wald intervals, SD rows use log-SD shaping, and the correlation row uses Fisher-z shaping.", fig.alt="Compact Confidence Eye display showing fixed effects, residual and group-level standard deviations, and the random intercept-slope correlation as pale finite confidence regions with hollow point-estimate circles."}
fixed_rows <- summary(fit_re_slopes)$coefficients
parameter_rows <- summary(fit_re_slopes)$parameters
coefficient_specs <- data.frame(
  group = c(
    "Fixed effects",
    "Fixed effects",
    "SDs (log-scale Wald)",
    "SDs (log-scale Wald)",
    "SDs (log-scale Wald)",
    "Correlation (Fisher-z Wald)"
  ),
  parameter = c(
    "mu intercept",
    "mu temperature slope",
    "residual sigma",
    "site intercept SD",
    "site temperature-slope SD",
    "intercept-slope correlation"
  ),
  display = c(
    "mu\nintercept",
    "mu\ntemperature slope",
    "residual\nsigma",
    "site\nintercept SD",
    "site temperature-\nslope SD",
    "intercept-slope\ncorrelation"
  ),
  estimate = c(
    fixed_rows["mu:(Intercept)", "estimate"],
    fixed_rows["mu:temperature", "estimate"],
    parameter_rows["sigma", "estimate"],
    parameter_rows[
      "sd:mu:(1 + temperature | site):(Intercept)",
      "estimate"
    ],
    parameter_rows[
      "sd:mu:(1 + temperature | site):temperature",
      "estimate"
    ],
    parameter_rows[
      "cor:mu:cor((Intercept),temperature | site)",
      "estimate"
    ]
  ),
  response_std_error = c(
    fixed_rows["mu:(Intercept)", "std_error"],
    fixed_rows["mu:temperature", "std_error"],
    parameter_rows["sigma", "std_error"],
    parameter_rows[
      "sd:mu:(1 + temperature | site):(Intercept)",
      "std_error"
    ],
    parameter_rows[
      "sd:mu:(1 + temperature | site):temperature",
      "std_error"
    ],
    parameter_rows[
      "cor:mu:cor((Intercept),temperature | site)",
      "std_error"
    ]
  ),
  transform = c("identity", "identity", "exp", "exp", "exp", "tanh"),
  reference = c(0, 0, 0, 0, 0, 0),
  stringsAsFactors = FALSE
)
coefficient_specs$link_estimate <- coefficient_specs$estimate
coefficient_specs$link_std_error <- coefficient_specs$response_std_error
sd_rows <- coefficient_specs$transform == "exp"
coefficient_specs$link_estimate[sd_rows] <- log(coefficient_specs$estimate[sd_rows])
coefficient_specs$link_std_error[sd_rows] <- response_scale_se_to_log_se(
  coefficient_specs$estimate[sd_rows],
  coefficient_specs$response_std_error[sd_rows]
)
cor_rows <- coefficient_specs$transform == "tanh"
coefficient_specs$link_estimate[cor_rows] <- atanh(
  guard_rho(coefficient_specs$estimate[cor_rows])
)
coefficient_specs$link_std_error[cor_rows] <- response_scale_se_to_z_se(
  coefficient_specs$estimate[cor_rows],
  coefficient_specs$response_std_error[cor_rows]
)
coefficient_specs <- coefficient_specs[is.finite(coefficient_specs$link_std_error), ]
coefficient_specs$row_id <- rev(seq_len(nrow(coefficient_specs)))
coefficient_specs$group <- factor(
  coefficient_specs$group,
  levels = c(
    "Fixed effects",
    "SDs (log-scale Wald)",
    "Correlation (Fisher-z Wald)"
  )
)
coefficient_cloud <- do.call(
  rbind,
  lapply(seq_len(nrow(coefficient_specs)), function(i) {
    row <- coefficient_specs[i, ]
    make_confidence_eye_distribution(
      parameter = row$parameter,
      estimate = row$link_estimate,
      std_error = row$link_std_error,
      transform = row$transform
    )
  })
)
coefficient_cloud <- merge(
  coefficient_cloud,
  coefficient_specs[c("parameter", "group", "display", "row_id")],
  by = "parameter",
  sort = FALSE
)
coefficient_cloud$group <- factor(
  coefficient_cloud$group,
  levels = levels(coefficient_specs$group)
)
coefficient_cloud$ymin <- coefficient_cloud$row_id -
  0.17 * coefficient_cloud$height
coefficient_cloud$ymax <- coefficient_cloud$row_id +
  0.17 * coefficient_cloud$height

interval_group_palette <- c(
  "Fixed effects" = gallery_colour("blue"),
  "SDs (log-scale Wald)" = gallery_colour("green"),
  "Correlation (Fisher-z Wald)" = gallery_colour("orange")
)

ggplot() +
  geom_vline(
    xintercept = seq(-0.5, 0.75, by = 0.25),
    colour = "grey90",
    linewidth = 0.35
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    colour = "grey55",
    linewidth = 0.55
  ) +
  geom_ribbon(
    data = coefficient_cloud,
    aes(x = value, ymin = ymin, ymax = ymax, fill = group, group = parameter),
    alpha = 0.24,
    colour = NA
  ) +
  geom_point(
    data = coefficient_specs,
    aes(x = estimate, y = row_id, colour = group),
    shape = 21,
    fill = "white",
    size = 3.1,
    stroke = 1.1
  ) +
  scale_colour_manual(values = interval_group_palette) +
  scale_fill_manual(values = interval_group_palette) +
  scale_x_continuous(
    limits = c(-0.5, 0.75),
    breaks = c(-0.5, 0, 0.5),
    expand = expansion(mult = 0.04)
  ) +
  scale_y_continuous(
    breaks = coefficient_specs$row_id,
    labels = coefficient_specs$display,
    expand = expansion(add = 0.42)
  ) +
  labs(
    title = NULL,
    subtitle = NULL,
    x = "Estimate",
    y = NULL,
    colour = NULL
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.y = element_text(size = 10.5, colour = "grey30"),
    axis.ticks.y = element_blank(),
    axis.line.x = element_line(colour = "grey35", linewidth = 0.35),
    axis.ticks.x = element_line(colour = "grey35", linewidth = 0.35),
    axis.ticks.length.x = grid::unit(3, "pt"),
    panel.grid = element_blank()
  ) +
  guides(colour = "none", fill = "none")
```

A random-slope figure should show repeated site behaviour. The translucent
lines below are conditional site-level trajectories: each site borrows strength
from the population fit, but sites can still differ in both intercept and
temperature slope.

```{r random-slope-modes, fig.cap="Repeated-measures raw observations with conditional site trajectories and the population-average fitted trend; the figure shows fitted modes, not interval uncertainty.", fig.alt="Repeated-measures random-slope plot with grey observed site data, translucent site-level fitted trajectories, and a darker population-average fitted line."}
site_intercept <- ranef(fit_re_slopes, "mu")$terms[[
  "(1 + temperature | site):(Intercept)"
]]
site_slope_deviation <- ranef(fit_re_slopes, "mu")$terms[[
  "(1 + temperature | site):temperature"
]]
site_effects <- data.frame(
  site = names(site_slope_deviation),
  intercept_deviation = unname(site_intercept),
  slope_deviation = unname(site_slope_deviation)
)
site_effects$slope_direction <- factor(
  ifelse(
    site_effects$slope_deviation < 0,
    "Lower than average",
    "Higher than average"
  ),
  levels = c("Lower than average", "Higher than average")
)
fixed_re <- summary(fit_re_slopes)$coefficients
site_lines <- expand.grid(
  site = names(site_slope_deviation),
  temperature = seq(-1, 1, length.out = 50),
  KEEP.OUT.ATTRS = FALSE
)
site_lines <- merge(site_lines, site_effects, by = "site")
site_lines$fitted_growth <- (
  fixed_re["mu:(Intercept)", "estimate"] +
    site_lines$intercept_deviation
) + (
  fixed_re["mu:temperature", "estimate"] +
    site_lines$slope_deviation
) * site_lines$temperature
population_line <- data.frame(
  temperature = seq(-1, 1, length.out = 80)
)
population_line$fitted_growth <- fixed_re["mu:(Intercept)", "estimate"] +
  fixed_re["mu:temperature", "estimate"] * population_line$temperature

ggplot() +
  geom_point(
    data = reef_re,
    aes(temperature, growth),
    colour = "grey55",
    alpha = 0.38,
    size = 1.15
  ) +
  geom_line(
    data = site_lines,
    aes(
      temperature,
      fitted_growth,
      group = site,
      colour = slope_direction
    ),
    alpha = 0.48,
    linewidth = 0.55
  ) +
  geom_line(
    data = population_line,
    aes(temperature, fitted_growth),
    colour = gallery_colour("charcoal"),
    linewidth = 1.05
  ) +
  scale_colour_manual(values = c(
    "Lower than average" = gallery_colour("blue"),
    "Higher than average" = gallery_colour("orange")
  )) +
  labs(
    title = "Site trajectories show random slopes",
    subtitle = "Lines are conditional modes, not interval uncertainty",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Site slope"
  ) +
  theme_drmtmb_gallery()
```

A direct `sd(site) ~ reef_cover` formula asks a third question: does a
site-level predictor change the among-site SD in expected growth? The
prediction table keeps this target as `sd(site)`, not residual `sigma`.

```{r random-effect-sd-surface, fig.cap="Fitted among-site random-intercept SD surface along reef cover; the rug shows site-level support, and no band is drawn because the table reports `conf.status = \"wald_unavailable\"` for this derived surface.", fig.alt="Line plot showing a fitted among-site random-intercept SD surface along reef cover, with rug marks for site support and no confidence band because profile or bootstrap intervals are needed for this derived surface."}
set.seed(2622)
n_sd_site <- 28
n_sd_each <- 5
site_sd_data <- data.frame(
  site = factor(seq_len(n_sd_site)),
  reef_cover = runif(n_sd_site, -1.2, 1.2)
)
site_sd <- exp(-0.6 + 0.45 * site_sd_data$reef_cover)
site_deviation <- rnorm(n_sd_site, mean = 0, sd = site_sd)
reef_sd <- site_sd_data[rep(seq_len(n_sd_site), each = n_sd_each), ]
reef_sd$temperature <- rep(seq(-1.2, 1.2, length.out = n_sd_each), n_sd_site)
reef_sd$growth <- 0.4 +
  0.5 * reef_sd$temperature +
  site_deviation[as.integer(reef_sd$site)] +
  rnorm(nrow(reef_sd), mean = 0, sd = 0.35)

fit_site_sd <- drmTMB(
  bf(growth ~ temperature + (1 | site), sigma ~ 1, sd(site) ~ reef_cover),
  family = gaussian(),
  data = reef_sd,
  control = drm_control(optimizer = list(eval.max = 180L, iter.max = 180L))
)

grid_site_sd <- prediction_grid(
  fit_site_sd,
  focal = "reef_cover",
  at = list(reef_cover = seq(-1.2, 1.2, length.out = 80))
)
pred_site_sd <- predict_parameters(
  fit_site_sd,
  newdata = grid_site_sd,
  dpar = "sd(site)",
  conf.int = TRUE
)
pred_site_sd$quantity <- "Among-site SD"

plot_parameter_surface(
  pred_site_sd,
  x = "reef_cover",
  colour = "quantity",
  facet = NULL,
  point = FALSE
) +
  scale_colour_manual(values = c(
    "Among-site SD" = gallery_colour("green")
  )) +
  geom_rug(
    data = site_sd_data,
    aes(x = reef_cover),
    inherit.aes = FALSE,
    sides = "b",
    colour = "grey55",
    alpha = 0.50,
    linewidth = 0.45
  ) +
  geom_point(
    data = pred_site_sd[seq(1, nrow(pred_site_sd), length.out = 9), ],
    aes(reef_cover, estimate),
    inherit.aes = FALSE,
    colour = gallery_colour("green"),
    fill = "white",
    shape = 21,
    size = 1.7,
    stroke = 0.55
  ) +
  labs(
    title = "Fitted among-site SD surface",
    subtitle = "Rug marks support; profile/bootstrap intervals needed",
    x = "Reef cover",
    y = "Estimated among-site SD",
    colour = NULL
  ) +
  theme_drmtmb_gallery() +
  guides(colour = "none")
```

This line is a fitted random-effect SD surface. It should not be read as a raw
growth curve, and it should not be overlaid with raw `growth` points. The grey
rug marks show where the site-level predictor was observed. The table behind
the plot reports `conf.status = "wald_unavailable"`, which is the honest
interval boundary until a profile or bootstrap route is attached to these
derived surfaces.

## Discrete comparisons

For a factor predictor, the same plotting helper uses interval bars rather than
ribbons. This is useful for compact tutorial figures and estimated marginal
comparisons.

```{r discrete-comparison, fig.height=2.05, fig.cap="Fitted habitat means at average temperature; points and horizontal intervals are response-scale Wald summaries from `predict_parameters()`.", fig.alt="Compact horizontal point interval plot comparing expected growth between reef and kelp habitats at average temperature."}
habitat_grid <- data.frame(
  temperature = 0,
  habitat = factor(c("reef", "kelp"), levels = levels(fish$habitat))
)
pred_habitat <- predict_parameters(
  fit_growth,
  newdata = habitat_grid,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(pred_habitat, aes(estimate, habitat, colour = habitat)
) +
  geom_segment(
    aes(x = conf.low, xend = conf.high, yend = habitat),
    linewidth = 0.75,
    lineend = "round"
  ) +
  geom_point(size = 2.7) +
  scale_drmtmb_colour +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.08))) +
  labs(
    title = "Fitted habitat means at average temperature",
    subtitle = "Intervals are Wald intervals from predict_parameters()",
    x = "Expected growth",
    y = NULL,
    colour = NULL
  ) +
  theme_point_interval_gallery() +
  guides(colour = "none")
```

## Categorical by categorical interaction

For two categorical predictors, show every fitted cell and its interval. Use
position dodging so the uncertainty intervals do not sit on top of each other.
Raw observations can sit behind the intervals when they help readers see the
data support for each cell.

```{r cat-cat-interaction, fig.cap="Habitat-by-season fitted mean summaries over raw observations; large points are fitted `mu` cell means and bars are 95% Wald intervals.", fig.alt="Categorical by categorical interaction plot showing raw growth observations, fitted mu cell means, and Wald intervals for each habitat and season combination."}
set.seed(2592)
n_cell <- 30
reef <- expand.grid(
  habitat = factor(habitat_levels, levels = habitat_levels),
  season = factor(c("dry", "wet"), levels = c("dry", "wet")),
  replicate = seq_len(n_cell)
)
reef$mu <- with(
  reef,
  0.6 +
    0.35 * (habitat == "kelp") -
    0.15 * (season == "wet") +
    0.45 * (habitat == "kelp") * (season == "wet")
)
reef$growth <- rnorm(nrow(reef), mean = reef$mu, sd = 0.45)

fit_cat_cat <- drmTMB(
  bf(growth ~ habitat * season, sigma ~ 1),
  family = gaussian(),
  data = reef
)

grid_cat_cat <- expand.grid(
  habitat = levels(reef$habitat),
  season = levels(reef$season)
)
pred_cat_cat <- predict_parameters(
  fit_cat_cat,
  newdata = grid_cat_cat,
  dpar = "mu",
  conf.int = TRUE
)

raw_cat_cat <- reef

cat_cat_position <- position_dodge(width = 0.45)

ggplot(pred_cat_cat, aes(habitat, estimate, colour = season)) +
  geom_point(
    data = raw_cat_cat,
    aes(y = growth, colour = season),
    position = position_jitterdodge(
      jitter.width = 0.08,
      jitter.height = 0,
      dodge.width = 0.45
    ),
    alpha = 0.22,
    size = 1.1,
    show.legend = FALSE
  ) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    position = cat_cat_position,
    width = 0.08,
    linewidth = 0.65
  ) +
  geom_point(
    position = cat_cat_position,
    size = 2.3
  ) +
  scale_colour_manual(values = season_palette) +
  labs(
    title = "Habitat-season fitted means",
    subtitle = "Small points are observations; large points and bars are fitted mu summaries",
    x = "Habitat",
    y = "Growth",
    colour = "Season"
  ) +
  theme_drmtmb_gallery()
```

## Estimated marginal means and marginal summaries

For the first supported `emmeans` path, a fixed-effect univariate `mu` model can
produce estimated marginal means that are easy to turn into publication-style
figures. The same rule applies: keep the target and conditioning values
visible. These displays are `mu` summaries, not `sigma`, fitted-response,
random-effect, or bivariate summaries.

```{r emmeans-display, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.height=2.05, fig.cap="Estimated marginal mean `mu` summaries for habitat at temperature zero, using the `emmeans` bridge for fixed-effect univariate means.", fig.alt="Compact horizontal point interval plot of estimated marginal mean growth for reef and kelp habitats conditioned at temperature zero."}
emm_habitat <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat,
      at = list(temperature = 0),
      type = "response"
    )
  )
)

ggplot(emm_habitat, aes(emmean, habitat, colour = habitat)) +
  geom_segment(
    aes(x = asymp.LCL, xend = asymp.UCL, yend = habitat),
    linewidth = 0.75,
    lineend = "round"
  ) +
  geom_point(size = 2.7) +
  scale_drmtmb_colour +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.08))) +
  labs(
    title = "Estimated marginal means from emmeans",
    subtitle = "Supported fixed-effect mu target, conditioned at temperature = 0",
    x = "Expected growth",
    y = NULL,
    colour = NULL
  ) +
  theme_point_interval_gallery() +
  guides(colour = "none")
```

`emmeans` reference grids can also condition on a factor. In the next display,
the `mu` cell means are grouped by season, so the reader sees the fitted
habitat contrast inside each season rather than a single averaged contrast.

```{r emmeans-factor-grid, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.height=2.45, fig.cap="Season-conditioned estimated marginal mean `mu` summaries; intervals come from the `emmeans` fixed-effect reference grid.", fig.alt="Compact horizontal point interval plot of estimated marginal mean growth for reef and kelp habitats shown separately for dry and wet seasons."}
emm_season <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_cat_cat,
      ~ habitat | season,
      type = "response"
    )
  )
)
emm_season$habitat_index <- as.numeric(emm_season$habitat)
emm_season$season_offset <- ifelse(emm_season$season == "dry", -0.08, 0.08)
emm_season$habitat_y <- emm_season$habitat_index + emm_season$season_offset

ggplot(emm_season, aes(emmean, habitat_y, colour = season)) +
  geom_segment(
    aes(x = asymp.LCL, xend = asymp.UCL, yend = habitat_y),
    linewidth = 0.7,
    lineend = "round"
  ) +
  geom_point(size = 2.6) +
  scale_colour_manual(values = season_palette) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.08))) +
  scale_y_continuous(
    breaks = seq_along(levels(emm_season$habitat)),
    labels = levels(emm_season$habitat),
    limits = c(0.65, length(levels(emm_season$habitat)) + 0.35)
  ) +
  labs(
    title = "Season-conditioned marginal means",
    subtitle = "Habitat mu summaries are shown separately for each season",
    x = "Expected growth",
    y = "Habitat",
    colour = "Season"
  ) +
  theme_point_interval_gallery() +
  theme(legend.position = "bottom")
```

For a numeric interaction, use an explicit reference grid. Here `temperature`
is held at three named values so the habitat contrast does not hide the
conditioning rule.

```{r emmeans-interaction-grid, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.height=3.35, fig.cap="Estimated marginal mean `mu` trends for habitat across three named temperature values; intervals are fixed-effect `emmeans` intervals.", fig.alt="Point interval plot of estimated marginal mean growth for reef and kelp habitats at three temperature values."}
emm_temperature <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat | temperature,
      at = list(temperature = c(-1, 0, 1)),
      type = "response"
    )
  )
)
emm_temperature$temperature_slice <- factor(
  emm_temperature$temperature,
  levels = c(-1, 0, 1),
  labels = c("cooler", "average", "warmer")
)

ggplot(emm_temperature, aes(temperature_slice, emmean, colour = habitat)) +
  geom_line(
    aes(group = habitat),
    linewidth = 0.65,
    alpha = 0.85
  ) +
  geom_errorbar(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    width = 0.08,
    linewidth = 0.65
  ) +
  geom_point(size = 2.5) +
  scale_drmtmb_colour +
  labs(
    title = "Marginal means across temperature slices",
    subtitle = "The same fixed-effect mu bridge, evaluated at named temperature slices",
    x = "Temperature slice",
    y = "Expected growth",
    colour = "Habitat"
  ) +
  theme_point_interval_gallery()
```

Empirical marginal summaries use a different rule. `prediction_grid(...,
margin = "empirical")` crosses a focal value with fitted rows, and
`marginal_parameters()` averages the resulting parameter predictions. The helper
itself is still a plug-in point summary, so the interval bars below are a
display approximation: they average row-wise Wald prediction limits from
`predict_parameters(conf.int = TRUE)` within each habitat.

```{r empirical-marginal-summary, fig.cap="Empirical marginal `mu` summaries by habitat; faint marks are fitted-row predictions, large points are plug-in marginal means, and bars average row-wise Wald limits as a display approximation.", fig.alt="Point-and-interval plot of empirical marginal expected growth for reef and kelp habitats, where faint marks are fitted-row predictions, large points are plug-in marginal means, and bars average row-wise Wald prediction limits."}
empirical_habitat_grid <- prediction_grid(
  fit_growth,
  focal = "habitat",
  at = list(habitat = levels(fish$habitat)),
  margin = "empirical"
)
empirical_habitat <- marginal_parameters(
  fit_growth,
  newdata = empirical_habitat_grid,
  dpar = "mu",
  by = "habitat"
)
empirical_habitat_rowwise <- predict_parameters(
  fit_growth,
  newdata = empirical_habitat_grid,
  dpar = "mu",
  conf.int = TRUE
)
empirical_habitat_wald <- aggregate(
  cbind(conf.low, conf.high) ~ habitat,
  data = empirical_habitat_rowwise,
  FUN = mean
)
empirical_habitat_plot <- empirical_habitat
empirical_habitat_plot$conf.low <- empirical_habitat_wald$conf.low[
  match(empirical_habitat_plot$habitat, empirical_habitat_wald$habitat)
]
empirical_habitat_plot$conf.high <- empirical_habitat_wald$conf.high[
  match(empirical_habitat_plot$habitat, empirical_habitat_wald$habitat)
]
empirical_habitat_plot$conf.status <- "wald"
empirical_habitat_plot$interval_source <- "averaged_rowwise_wald"

ggplot(empirical_habitat_rowwise, aes(habitat, estimate, fill = habitat)) +
  geom_violin(
    width = 0.55,
    alpha = 0.16,
    colour = NA,
    trim = FALSE
  ) +
  geom_point(
    aes(colour = habitat),
    position = position_jitter(width = 0.08, height = 0),
    alpha = 0.22,
    size = 0.65,
    show.legend = FALSE
  ) +
  geom_errorbar(
    data = empirical_habitat_plot,
    aes(
      y = estimate,
      ymin = conf.low,
      ymax = conf.high,
      colour = habitat
    ),
    width = 0.10,
    linewidth = 0.7
  ) +
  geom_point(
    data = empirical_habitat_plot,
    aes(y = estimate, colour = habitat),
    size = 2.4
  ) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Empirical marginal mu summary",
    subtitle = "Large points are plug-in means; bars average row-wise Wald limits",
    x = NULL,
    y = "Expected growth",
    colour = "Habitat",
    fill = "Habitat"
  ) +
  guides(fill = "none") +
  theme_drmtmb_gallery()
```

Unsupported targets should be visible before a reader tries to plot them. The
current `emmeans` bridge is for fixed-effect univariate `mu`; other targets
should use parameter prediction, `corpairs()`, profile intervals, or a future
feature-specific design.

```{r emmeans-boundary-strip, fig.height=3.6, fig.cap="Support-boundary strip for `emmeans`; fixed-effect univariate `mu` is supported, while other targets show the fitted route to use instead.", fig.alt="Marker plot showing fixed-effect univariate mu as the supported emmeans target and sigma, bivariate, zero-inflated, hurdle, ordinal, and random-effect targets as boundary routes with alternatives named beside each marker."}
emmeans_boundary <- data.frame(
  target = c(
    "Fixed-effect univariate mu",
    "sigma",
    "Bivariate responses",
    "Zero-inflated or hurdle means",
    "Ordinal expected scores",
    "Random-effect targets"
  ),
  status = c(
    "Supported emmeans route",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary"
  )
)
emmeans_boundary$route <- c(
  "supported EMMs",
  "predict_parameters()",
  "separate bivariate design",
  "family-specific marginal route",
  "ordinal support later",
  "ranef(), profile_targets()"
)
emmeans_boundary$target <- factor(
  emmeans_boundary$target,
  levels = rev(emmeans_boundary$target)
)
emmeans_boundary$status <- factor(
  emmeans_boundary$status,
  levels = c("Supported emmeans route", "Unsupported boundary")
)
emmeans_boundary$status_label <- ifelse(
  emmeans_boundary$status == "Supported emmeans route",
  "supported",
  "boundary"
)

ggplot(emmeans_boundary, aes(y = target)) +
  geom_segment(
    aes(x = 0.92, xend = 2.35, yend = target),
    colour = "grey86",
    linewidth = 0.35
  ) +
  geom_point(
    aes(x = 1, fill = status),
    shape = 21,
    size = 3.4,
    colour = "white",
    stroke = 0.8
  ) +
  geom_text(
    aes(x = 1.14, label = status_label, colour = status),
    hjust = 0,
    size = 2.85,
    fontface = "bold"
  ) +
  geom_text(
    aes(x = 1.65, label = route),
    hjust = 0,
    size = 2.75,
    colour = "grey25"
  ) +
  scale_fill_manual(values = c(
    "Supported emmeans route" = "#0072B2",
    "Unsupported boundary" = "#E69F00"
  )) +
  scale_colour_manual(values = c(
    "Supported emmeans route" = "#0072B2",
    "Unsupported boundary" = "#8A5700"
  )) +
  scale_x_continuous(
    limits = c(0.86, 2.55),
    breaks = c(1.14, 1.65),
    labels = c("Status", "Current route")
  ) +
  labs(
    title = "Current emmeans support boundary",
    subtitle = "Only fixed-effect univariate mu is advertised as an emmeans target",
    x = NULL,
    y = NULL,
    fill = "Status"
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.x = element_text(size = 8.8, face = "bold", colour = "grey35"),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

## Continuous by continuous interaction

For two continuous predictors, choose scientifically meaningful slices of one
predictor and show fitted slopes across the other. The values should be named
in the legend so the reader can interpret the conditioning.

```{r cont-cont-interaction, fig.cap="Continuous-by-continuous fitted `mu` interaction over raw observations; ribbons are 95% Wald confidence bands at low, average, and high moisture.", fig.alt="Continuous by continuous interaction plot showing raw observations in grey and fitted temperature slopes at low, average, and high moisture with confidence bands."}
set.seed(2593)
n_cc <- 180
soil <- data.frame(
  temperature = runif(n_cc, -1.5, 1.5),
  moisture = runif(n_cc, -1.4, 1.4)
)
soil$growth <- rnorm(
  n_cc,
  mean = 0.3 +
    0.55 * soil$temperature +
    0.45 * soil$moisture -
    0.35 * soil$temperature * soil$moisture,
  sd = 0.45
)

fit_cont_cont <- drmTMB(
  bf(growth ~ temperature * moisture, sigma ~ 1),
  family = gaussian(),
  data = soil
)

grid_cont_cont <- expand.grid(
  temperature = seq(-1.5, 1.5, length.out = 80),
  moisture = c(-1, 0, 1)
)
grid_cont_cont$moisture_level <- factor(
  grid_cont_cont$moisture,
  levels = c(-1, 0, 1),
  labels = c("low moisture", "average moisture", "high moisture")
)

pred_cont_cont <- predict_parameters(
  fit_cont_cont,
  newdata = grid_cont_cont,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(
  pred_cont_cont,
  aes(temperature, estimate, colour = moisture_level, fill = moisture_level)
) +
  geom_point(
    data = soil,
    aes(temperature, growth),
    inherit.aes = FALSE,
    alpha = 0.18,
    size = 1,
    colour = "grey35"
  ) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.14, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c(
    "low moisture" = gallery_colour("blue"),
    "average moisture" = gallery_colour("green"),
    "high moisture" = gallery_colour("orange")
  )) +
  scale_fill_manual(values = c(
    "low moisture" = gallery_colour("blue"),
    "average moisture" = gallery_colour("green"),
    "high moisture" = gallery_colour("orange")
  )) +
  labs(
    title = "Continuous by continuous interaction",
    subtitle = "Ribbons are 95% Wald bands at three moisture slices",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Moisture slice",
    fill = "Moisture slice"
  ) +
  theme_drmtmb_gallery()
```

The raw point cloud is not coloured by the sliced predictor because the fitted
lines are conditional displays, not clusters in the data. This keeps the
visual message honest: the model is being read at three moisture values, while
the observations span the whole moisture gradient.

## Fitted correlation summaries

Correlation displays should keep layers separate. Residual `rho12`,
ordinary-group correlations, phylogenetic correlations, coordinate-spatial
correlations, animal-model correlations, and lower-level `relmat()`
correlations are not the same parameter even when all are correlations. A
fitted workflow should start from `corpairs(fit)` and keep the interval
provenance columns.

A Confidence Eye should appear only when finite interval bounds exist. The
compact example below mimics a `corpairs()`-style result for fitted correlation
layers. The pale eye is the finite 95% confidence region, transformed on
Fisher's `z` scale for correlations. The hollow circle marks the estimate.
Rows without finite bounds should remain point-only diagnostics; the plot
should not invent an eye.

```{r correlation-display, fig.width=7.6, fig.height=5.4, fig.cap="Correlation-row Confidence Eye display; finite 95% regions are shaped on Fisher's z scale and shown on the correlation scale.", fig.alt="Correlation-row Confidence Eye plot for residual rho12, group intercept-slope, phylogenetic, spatial, animal, and relmat location-location rows, with pale finite confidence regions and hollow point-estimate circles."}
pair_table <- data.frame(
  level = c("residual", "group", "phylogenetic", "spatial", "animal", "relmat"),
  class = c(
    "residual",
    "intercept-slope",
    "location-location",
    "location-location",
    "location-location",
    "location-location"
  ),
  parameter = c(
    "rho12",
    "cor((Intercept), x | id)",
    "cor(mu1, mu2 | phylo)",
    "cor(mu1, mu2 | spatial)",
    "cor(mu1, mu2 | animal)",
    "cor(mu1, mu2 | relmat)"
  ),
  estimate = c(-0.32, 0.44, 0.61, 0.38, 0.52, -0.27),
  modelled = rep(TRUE, 6),
  conf.low = c(-0.54, 0.12, 0.32, 0.05, 0.18, -0.58),
  conf.high = c(-0.05, 0.68, 0.80, 0.63, 0.75, 0.10),
  conf.status = rep("profile", 6),
  interval_source = rep("illustrative_profile", 6),
  display = c(
    "Residual\nrho12",
    "Group\nintercept-slope",
    "Phylogenetic\nlocation-location",
    "Spatial\nlocation-location",
    "Animal\nlocation-location",
    "relmat\nlocation-location"
  ),
  stringsAsFactors = FALSE
)
pair_table$row_id <- rev(seq_len(nrow(pair_table)))
pair_table$has_interval <- is.finite(pair_table$conf.low) &
  is.finite(pair_table$conf.high)
correlation_palette <- c(
  "residual" = gallery_colour("blue"),
  "group" = gallery_colour("green"),
  "phylogenetic" = gallery_colour("orange"),
  "spatial" = gallery_colour("sky"),
  "animal" = gallery_colour("pink"),
  "relmat" = gallery_colour("charcoal")
)
make_correlation_confidence_eye <- function(row, level = 0.95, n = 240) {
  if (!all(is.finite(c(row$estimate, row$conf.low, row$conf.high)))) {
    return(data.frame())
  }
  z_estimate <- atanh(guard_rho(row$estimate))
  z_low <- atanh(guard_rho(row$conf.low))
  z_high <- atanh(guard_rho(row$conf.high))
  z_cutoff <- qnorm(1 - (1 - level) / 2)
  z_se <- max((z_high - z_low) / (2 * z_cutoff), .Machine$double.eps)
  cutoff <- 0.5 * qchisq(level, df = 1)
  z <- seq(
    z_estimate - sqrt(2 * cutoff) * z_se,
    z_estimate + sqrt(2 * cutoff) * z_se,
    length.out = n
  )
  rho <- tanh(z)
  height <- pmax(cutoff - 0.5 * ((z - z_estimate) / z_se)^2, 0) /
    cutoff * 0.17
  data.frame(
    rho = rho,
    height = height,
    ymin = row$row_id - height,
    ymax = row$row_id + height,
    row_id = row$row_id,
    display = row$display,
    level = row$level
  )
}
correlation_cloud <- do.call(
  rbind,
  lapply(seq_len(nrow(pair_table)), function(i) make_correlation_confidence_eye(pair_table[i, ]))
)

ggplot() +
  geom_vline(
    xintercept = seq(-1, 1, by = 0.5),
    colour = "grey90",
    linewidth = 0.35
  ) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.55, linetype = "dotted") +
  geom_ribbon(
    data = correlation_cloud,
    aes(
      rho,
      ymin = ymin,
      ymax = ymax,
      fill = level,
      group = interaction(level, display)
    ),
    alpha = 0.24,
    colour = NA
  ) +
  geom_point(
    data = pair_table,
    aes(estimate, row_id, colour = level),
    shape = 21,
    fill = "white",
    size = 3.1,
    stroke = 1.1,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values = correlation_palette) +
  scale_fill_manual(values = correlation_palette, guide = "none") +
  scale_x_continuous(
    limits = c(-1, 1),
    breaks = seq(-1, 1, by = 0.5),
    expand = expansion(mult = 0.04)
  ) +
  scale_y_continuous(
    breaks = pair_table$row_id,
    labels = pair_table$display,
    expand = expansion(add = 0.42)
  ) +
  labs(
    title = NULL,
    subtitle = NULL,
    x = "Correlation estimate",
    y = NULL,
    colour = NULL
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.y = element_text(size = 10.5, colour = "grey30"),
    axis.ticks.y = element_blank(),
    axis.line.x = element_line(colour = "grey35", linewidth = 0.35),
    axis.ticks.x = element_line(colour = "grey35", linewidth = 0.35),
    axis.ticks.length.x = grid::unit(3, "pt"),
    panel.grid = element_blank()
  ) +
  guides(colour = "none", fill = "none")
```

The figure should be read as an uncertainty display, not as a new model class.
Rows with finite bounds get default Confidence Eyes. Rows without finite bounds
should be shown as point-only diagnostics. This prevents the plot from implying
uncertainty evidence that the fitted object or extractor did not provide.

```{r correlation-layer-boundaries, fig.height=3.7, fig.cap="Support-boundary strip for correlation layers; fitted q=2 rows, partial richer q=4 or regression rows, and planned residual-`rho12` routes are kept separate.", fig.alt="Marker plot showing fitted q2 support for residual, ordinary group, phylogenetic, spatial, animal, and relmat correlation layers, with partial richer q4 or regression routes separated from planned residual-rho12 boundaries."}
layer_boundary <- expand.grid(
  layer = c(
    "Residual rho12",
    "Ordinary group",
    "Phylogenetic",
    "Spatial",
    "Animal",
    "relmat"
  ),
  target = c("Constant q2 row", "Regression or q4/scale extension"),
  stringsAsFactors = FALSE
)
layer_boundary$support <- "Planned"
layer_boundary$support[
  layer_boundary$target == "Constant q2 row" &
    layer_boundary$layer %in% c(
      "Residual rho12",
      "Ordinary group",
      "Phylogenetic",
      "Spatial",
      "Animal",
      "relmat"
    )
] <- "Fitted q2"
layer_boundary$support[
  layer_boundary$target == "Regression or q4/scale extension" &
    layer_boundary$layer %in% c("Spatial", "Animal", "relmat")
] <- "Partial q4"
layer_boundary$support <- factor(
  layer_boundary$support,
  levels = c("Fitted q2", "Partial q4", "Planned")
)
layer_boundary$layer <- factor(
  layer_boundary$layer,
  levels = c(
    "Residual rho12",
    "Ordinary group",
    "Phylogenetic",
    "Spatial",
    "Animal",
    "relmat"
  )
)
layer_boundary$target <- factor(
  layer_boundary$target,
  levels = c("Constant q2 row", "Regression or q4/scale extension")
)
ggplot(layer_boundary, aes(target, layer, fill = support)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = support), size = 2.8, colour = "white", fontface = "bold") +
  scale_fill_manual(
    values = c(
      "Fitted q2" = gallery_colour("green"),
      "Partial q4" = gallery_colour("orange"),
      "Planned" = gallery_colour("charcoal")
    )
  ) +
  labs(
    title = "Correlation layers have different support boundaries",
    subtitle = "q2 constants are separate from richer regression and q4 extensions",
    x = NULL,
    y = NULL,
    fill = "Status"
  ) +
  theme_drmtmb_gallery() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

The grid prevents a common mistake: seeing one fitted correlation route and
assuming all correlation layers now share the same regression and q4 support.

## Simulation operating characteristics

Simulation figures answer different questions from model-interpretation
figures. They should show bias, error, coverage, runtime, and failures beside
the model surface being tested.

```{r simulation-operating-characteristics, fig.width=8.6, fig.height=4.8, fig.cap="Simulation grammar fixture showing replicate-error bias clouds with MCSE intervals and coverage block proportions with binomial MCSE whiskers.", fig.alt="Simulation summary plots showing fixture replicate-error bias clouds with mean and MCSE intervals, plus interval coverage points with binomial MCSE whiskers, by estimand across Gaussian location-scale, proportion, and count surfaces."}
sim_targets <- data.frame(
  surface = rep(
    c("Gaussian location-scale", "Proportion", "Count"),
    each = 4
  ),
  estimand = rep(c("beta_x", "sigma", "sd_intercept", "rho12"), 3),
  bias = c(
    0.01, -0.02, 0.04, 0.00,
    -0.01, 0.03, NA, NA,
    0.02, -0.01, -0.04, NA
  ),
  rmse = c(
    0.07, 0.09, 0.13, 0.11,
    0.08, 0.10, NA, NA,
    0.10, 0.09, 0.16, NA
  ),
  coverage = c(
    0.95, 0.94, 0.91, 0.93,
    0.94, 0.92, NA, NA,
    0.93, 0.95, 0.89, NA
  )
)
sim_targets$n_rep <- 500L
sim_targets$error_sd <- sqrt(
  pmax(sim_targets$rmse^2 - sim_targets$bias^2, 0)
)
sim_bias_rows <- do.call(
  rbind,
  lapply(seq_len(nrow(sim_targets)), function(i) {
    row <- sim_targets[i, ]
    if (is.na(row$bias) || is.na(row$rmse)) {
      return(NULL)
    }
    data.frame(
      surface = row$surface,
      estimand = row$estimand,
      replicate = seq_len(row$n_rep),
      bias_error = row$bias +
        qnorm(ppoints(row$n_rep)) * row$error_sd
    )
  })
)
summarise_bias_rows <- function(rows) {
  data.frame(
    surface = rows$surface[1],
    estimand = rows$estimand[1],
    bias = mean(rows$bias_error),
    rmse = sqrt(mean(rows$bias_error^2)),
    error_sd = stats::sd(rows$bias_error),
    n_rep = nrow(rows)
  )
}
sim_summary <- do.call(
  rbind,
  lapply(
    split(
      sim_bias_rows,
      interaction(sim_bias_rows$surface, sim_bias_rows$estimand, drop = TRUE)
    ),
    summarise_bias_rows
  )
)
sim_summary <- merge(
  sim_targets[c("surface", "estimand", "coverage")],
  sim_summary,
  by = c("surface", "estimand"),
  all.x = TRUE,
  sort = FALSE
)
sim_summary$n_rep <- 500L
sim_summary$bias_mcse <- sim_summary$error_sd / sqrt(sim_summary$n_rep)
sim_summary$bias_low <- sim_summary$bias - 1.96 * sim_summary$bias_mcse
sim_summary$bias_high <- sim_summary$bias + 1.96 * sim_summary$bias_mcse
sim_summary$coverage_mcse <- sqrt(
  sim_summary$coverage * (1 - sim_summary$coverage) / sim_summary$n_rep
)
sim_summary$coverage_low <- pmax(
  0,
  sim_summary$coverage - 1.96 * sim_summary$coverage_mcse
)
sim_summary$coverage_high <- pmin(
  1,
  sim_summary$coverage + 1.96 * sim_summary$coverage_mcse
)
sim_summary$surface <- factor(
  sim_summary$surface,
  levels = names(surface_palette)
)
estimand_labels <- c(
  "beta_x" = "beta_x fixed slope",
  "sigma" = "sigma residual SD",
  "sd_intercept" = "sd_intercept group SD",
  "rho12" = "rho12 residual correlation"
)
sim_summary$estimand_label <- factor(
  estimand_labels[sim_summary$estimand],
  levels = rev(unname(estimand_labels))
)
surface_facet_labels <- c(
  "Gaussian location-scale" = "Gaussian\nlocation-scale",
  "Proportion" = "Proportion",
  "Count" = "Count"
)
display_replicates <- 55
sim_rain <- do.call(
  rbind,
  lapply(split(sim_bias_rows, sim_bias_rows[c("surface", "estimand")]), function(rows) {
    if (nrow(rows) == 0L) {
      return(NULL)
    }
    keep <- round(seq(1, nrow(rows), length.out = display_replicates))
    rows[keep, , drop = FALSE]
  })
)
sim_rain$surface <- factor(sim_rain$surface, levels = levels(sim_summary$surface))
sim_rain$estimand_label <- factor(
  estimand_labels[sim_rain$estimand],
  levels = levels(sim_summary$estimand_label)
)
sim_bias_missing <- sim_summary[is.na(sim_summary$bias), , drop = FALSE]

ggplot(sim_rain, aes(bias_error, estimand_label)) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey60") +
  geom_violin(
    aes(
      fill = surface
    ),
    width = 0.62,
    alpha = 0.16,
    colour = NA,
    scale = "width",
    trim = FALSE
  ) +
  geom_point(
    aes(colour = surface),
    position = position_jitter(height = 0.05, width = 0),
    alpha = 0.22,
    size = 0.75,
    show.legend = FALSE
  ) +
  geom_errorbar(
    data = sim_summary,
    aes(
      xmin = bias_low,
      xmax = bias_high,
      y = estimand_label,
      colour = surface
    ),
    inherit.aes = FALSE,
    orientation = "y",
    width = 0.16,
    linewidth = 0.65,
    na.rm = TRUE
  ) +
  geom_point(
    data = sim_summary,
    aes(bias, estimand_label, colour = surface),
    size = 2.6,
    na.rm = TRUE
  ) +
  geom_text(
    data = sim_bias_missing,
    aes(x = -0.47, y = estimand_label, label = "not\ntargeted"),
    colour = "grey55",
    size = 2.25,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~surface,
    nrow = 1,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  scale_fill_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "Bias by estimand",
    subtitle = "Faint dots are fixture replicate errors; large points and bars show mean bias with 95% MCSE intervals",
    x = NULL,
    y = "Estimate minus truth",
    colour = "Surface"
  ) +
  guides(fill = "none") +
  theme_drmtmb_gallery()

sim_coverage_present <- sim_summary[!is.na(sim_summary$coverage), , drop = FALSE]
sim_coverage_present$y_id <- as.numeric(sim_coverage_present$estimand_label)
make_coverage_blocks <- function(row, n_block = 18) {
  block_size <- max(8L, floor(row$n_rep / n_block))
  block_coverage <- qbinom(
    ppoints(n_block),
    size = block_size,
    prob = row$coverage
  ) / block_size
  data.frame(
    surface = row$surface,
    estimand_label = row$estimand_label,
    block_coverage = block_coverage
  )
}
sim_coverage_blocks <- do.call(
  rbind,
  lapply(seq_len(nrow(sim_coverage_present)), function(i) {
    make_coverage_blocks(sim_coverage_present[i, ])
  })
)
sim_coverage_missing <- sim_summary[is.na(sim_summary$coverage), , drop = FALSE]

ggplot(sim_coverage_present, aes(coverage, estimand_label, colour = surface)) +
  geom_vline(xintercept = 0.95, linetype = "dotted", colour = "grey60") +
  geom_point(
    data = sim_coverage_blocks,
    aes(block_coverage, estimand_label, colour = surface),
    position = position_jitter(height = 0.07, width = 0),
    alpha = 0.20,
    size = 0.75,
    inherit.aes = FALSE
  ) +
  geom_errorbar(
    aes(xmin = coverage_low, xmax = coverage_high, colour = surface),
    orientation = "y",
    width = 0.18,
    linewidth = 0.45,
    na.rm = TRUE
  ) +
  geom_point(size = 2.3, na.rm = TRUE) +
  geom_text(
    data = sim_coverage_missing,
    aes(x = 0.755, y = estimand_label, label = "not\ntargeted"),
    colour = "grey55",
    size = 2.25,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~surface,
    nrow = 1,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  coord_cartesian(xlim = c(0.75, 1)) +
  scale_x_continuous(
    breaks = c(0.75, 0.85, 0.95),
    labels = c("0.75", "0.85", "0.95")
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "Interval coverage",
    subtitle = "Faint block dots sample replicate-block coverage; large points and bars show 95% binomial MCSE",
    x = "Empirical coverage",
    y = NULL,
    colour = "Surface"
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  theme_drmtmb_gallery()
```

These figures use a small illustrative table with the same columns that
simulation summaries should carry. The bias panel shows the intended
raincloud-style grammar: real reports should draw the replicate-level
errors, then overplot the mean and MCSE interval for each defined
surface-estimand cell. The panel still leaves estimands disconnected because
`beta_x`, `sigma`, `sd_intercept`, and `rho12` are categorical targets, not
steps along one trajectory. The coverage panel uses the more utilitarian
simulation grammar: replicate or replicate-block proportions first, then
aggregate coverage and a 95% binomial MCSE interval. This small fixture
illustrates block points; formal simulation reports should use the stored
replicate rows or stored block summaries rather than reconstructing raw
uncertainty from aggregate CSVs alone. The comprehensive simulation programme
should not be count-only. It should add continuous, meta-analysis, proportion,
count, ordinal, bivariate, random-slope, and structured-dependence surfaces as
each surface becomes ready. Real simulation reports should also show the run
manifest and warning/error ledger so missing or failed fits are not hidden
behind clean-looking panels.

## Gallery source map

The gallery is easier to audit when every figure names its source object,
extractor, interval source, and support boundary. Use this table as a quick
check before adapting a plot to a manuscript, tutorial, or simulation report.

```{r gallery-source-map, eval=TRUE}
gallery_source_map <- data.frame(
  figure = c(
    "Mean with confidence band",
    "mu and sigma surfaces",
    "nu, zi, and rho12 panels",
    "Variance-component Confidence Eye",
    "Conditional random-slope modes",
    "sd(site) surface",
    "Discrete habitat comparison",
    "Categorical interaction",
    "Simple emmeans display",
    "Factor-conditioned emmeans grid",
    "Interaction emmeans grid",
    "Empirical marginal mu summary",
    "emmeans support boundary",
    "Continuous interaction",
    "Correlation estimate rows",
    "Correlation support boundary",
    "Simulation bias and coverage"
  ),
  source = c(
    "fit_growth",
    "fit_growth",
    "fit_tail, fit_zi, fit_pair",
    "fit_re_slopes",
    "fit_re_slopes",
    "fit_site_sd",
    "fit_growth",
    "fit_cat_cat",
    "fit_growth",
    "fit_cat_cat",
    "fit_growth",
    "fit_growth",
    "emmeans_boundary",
    "fit_cont_cont",
    "pair_table",
    "layer_boundary",
    "sim_summary"
  ),
  extractor_or_plotter = c(
    "predict_parameters(dpar = \"mu\") + ggplot2",
    "predict_parameters() + plot_parameter_surface()",
    "predict_parameters() + plot_parameter_surface()",
    "summary(fit)$parameters + ggplot2",
    "ranef() + ggplot2",
    "prediction_grid() + predict_parameters() + plot_parameter_surface()",
    "predict_parameters(dpar = \"mu\") + plot_parameter_surface()",
    "predict_parameters(dpar = \"mu\") + ggplot2",
    "emmeans::emmeans(type = \"response\") + ggplot2",
    "emmeans::emmeans(~ habitat | season) + ggplot2",
    "emmeans::emmeans(~ habitat | temperature) + ggplot2",
    "prediction_grid(margin = \"empirical\") + marginal_parameters() + predict_parameters()",
    "ggplot2 status strip",
    "predict_parameters(dpar = \"mu\") + ggplot2",
    "corpairs-compatible table + ggplot2",
    "ggplot2 status strip",
    "ggplot2 fixture display"
  ),
  interval_source = c(
    "Wald intervals from predict_parameters()",
    "Wald bands from predict_parameters()",
    "Wald intervals from each fitted parameter table",
    "Wald regions shaped on log-SD scale",
    "No intervals; conditional modes only",
    "conf.status = \"wald_unavailable\"; no band",
    "Wald intervals from predict_parameters()",
    "Wald intervals from predict_parameters()",
    "asymptotic emmeans intervals",
    "asymptotic emmeans intervals",
    "asymptotic emmeans intervals",
    "Plug-in mean; bars average row-wise Wald limits",
    "No intervals; status only",
    "Wald intervals from predict_parameters()",
    "Finite bounds supplied in compatible pair table",
    "No intervals; status only",
    "Illustrative summaries, not final evidence"
  ),
  support_boundary = c(
    "fixed-effect univariate mu",
    "fixed-effect mu and sigma",
    "fixed-effect shape, inflation, and residual rho12",
    "ordinary Gaussian random-slope fit",
    "shrunken deviations, not variance intervals",
    "modelled random-effect SD surface without Wald bands",
    "fixed-effect univariate mu",
    "fixed-effect univariate mu",
    "fixed-effect univariate mu only",
    "fixed-effect univariate mu only",
    "fixed-effect univariate mu only",
    "empirical marginalisation, not emmeans",
    "unsupported targets remain unsupported",
    "fixed-effect univariate mu",
    "implemented residual, group, phylogenetic, spatial, animal, and relmat q2 rows",
    "partial q4 location-scale blocks for spatial, animal, and relmat (recovery-grade); regression and remaining richer paths planned",
    "display grammar for later result articles"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(gallery_source_map)
```

## Future simulation result articles

Specialised simulation result articles should come later, after the
comprehensive simulation surfaces are ready. Those articles can be organised by
scientific question: power, bias, coverage, runtime, convergence, and failure
patterns. They should also be organised across data types, not only counts:
continuous location-scale models, meta-analysis with known `V`, proportions,
counts, ordinal responses, bivariate `rho12`, random-slope surfaces, and
structured dependence.

This figure gallery is broader and lighter. It shows the visual language users
should see across tutorials, examples, model interpretation, and later
simulation reports.
