---
title: "Figure gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Figure gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This gallery shows the kinds of figures `drmTMB` users can build for model
interpretation, uncertainty display, fitted interactions, fitted correlation
summaries, and simulation diagnostics. It is different from the internal
count-pilot diagnostics used while developing the first Poisson/NB2 simulation
plumbing.

The examples here use small simulated data so the page can build quickly. They
are meant to show figure patterns, not biological conclusions.

```{r setup}
library(drmTMB)
library(ggplot2)

theme_drmtmb_gallery <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

scale_drmtmb_colour <- scale_colour_manual(
  values = c("reef" = "#1b9e77", "kelp" = "#d95f02")
)
scale_drmtmb_fill <- scale_fill_manual(
  values = c("reef" = "#1b9e77", "kelp" = "#d95f02")
)
```

## Fitted mean with a confidence band

Start with the figure most readers expect: raw observations, fitted mean, and a
95% confidence band on the response scale. This example is also a
categorical-by-continuous interaction: the temperature slope differs by
habitat.

```{r location-scale-fit}
set.seed(2591)
n <- 160
fish <- data.frame(
  temperature = runif(n, -1.6, 1.6),
  habitat = factor(sample(c("reef", "kelp"), n, replace = TRUE))
)
eta_mu <- 0.4 +
  0.7 * fish$temperature +
  0.35 * (fish$habitat == "kelp") +
  0.18 * fish$temperature * (fish$habitat == "kelp")
eta_sigma <- -0.55 + 0.35 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit_growth <- drmTMB(
  bf(growth ~ temperature * habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)

grid_growth <- expand.grid(
  temperature = seq(-1.6, 1.6, length.out = 80),
  habitat = levels(fish$habitat)
)
pred_growth <- predict_parameters(
  fit_growth,
  newdata = grid_growth,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)

pred_mu <- pred_growth[pred_growth$dpar == "mu", , drop = FALSE]

ggplot(fish, aes(temperature, growth, colour = habitat)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_ribbon(
    data = pred_mu,
    aes(
      x = temperature,
      y = estimate,
      ymin = conf.low,
      ymax = conf.high,
      fill = habitat
    ),
    inherit.aes = FALSE,
    alpha = 0.16,
    colour = NA
  ) +
  geom_line(data = pred_mu, aes(y = estimate), linewidth = 0.8) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Expected growth along a temperature gradient",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Habitat",
    fill = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

## Distributional-parameter surfaces

The same fit can show more than the mean. `predict_parameters()` returns a long
table with one row per distributional parameter, and
`plot_parameter_surface()` draws estimate lines plus finite interval bands when
they are present.

```{r parameter-surface}
plot_parameter_surface(
  pred_growth,
  x = "temperature",
  colour = "habitat",
  dpar = c("mu", "sigma")
) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Location and residual-scale surfaces",
    x = "Temperature anomaly",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

The `mu` panel is the expected response. The `sigma` panel is the fitted
residual standard deviation, so an upward line means lower predictability or
larger residual heterogeneity at warmer values.

## Discrete comparisons

For a factor predictor, the same plotting helper uses interval bars rather than
ribbons. This is useful for compact tutorial figures and estimated marginal
comparisons.

```{r discrete-comparison}
habitat_grid <- data.frame(
  temperature = 0,
  habitat = factor(c("reef", "kelp"), levels = levels(fish$habitat))
)
pred_habitat <- predict_parameters(
  fit_growth,
  newdata = habitat_grid,
  dpar = "mu",
  conf.int = TRUE
)

plot_parameter_surface(
  pred_habitat,
  x = "habitat",
  line = FALSE,
  point = TRUE
) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated habitat contrast at average temperature",
    x = NULL,
    y = "Expected growth"
  ) +
  theme_drmtmb_gallery()
```

## Estimated marginal means

For the first supported `emmeans` path, a fixed-effect univariate `mu` model can
produce estimated marginal means that are easy to turn into publication-style
figures. The same rule applies: keep the conditioning values visible.

```{r emmeans-display, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE)}
emm_habitat <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat,
      at = list(temperature = 0),
      type = "response"
    )
  )
)

ggplot(emm_habitat, aes(habitat, emmean, colour = habitat)) +
  geom_pointrange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    linewidth = 0.7,
    fatten = 2.4
  ) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated marginal mean growth",
    subtitle = "Conditioned at temperature = 0",
    x = NULL,
    y = "Expected growth",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

## Categorical by categorical interaction

For two categorical predictors, show every fitted cell and its interval. Use
position dodging so the uncertainty intervals do not sit on top of each other.

```{r cat-cat-interaction}
set.seed(2592)
n_cell <- 30
reef <- expand.grid(
  habitat = factor(c("reef", "kelp")),
  season = factor(c("dry", "wet")),
  replicate = seq_len(n_cell)
)
reef$mu <- with(
  reef,
  0.6 +
    0.35 * (habitat == "kelp") -
    0.15 * (season == "wet") +
    0.45 * (habitat == "kelp") * (season == "wet")
)
reef$growth <- rnorm(nrow(reef), mean = reef$mu, sd = 0.45)

fit_cat_cat <- drmTMB(
  bf(growth ~ habitat * season, sigma ~ 1),
  family = gaussian(),
  data = reef
)

grid_cat_cat <- expand.grid(
  habitat = levels(reef$habitat),
  season = levels(reef$season)
)
pred_cat_cat <- predict_parameters(
  fit_cat_cat,
  newdata = grid_cat_cat,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(pred_cat_cat, aes(habitat, estimate, colour = season)) +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.35),
    linewidth = 0.65,
    fatten = 2.3
  ) +
  scale_colour_manual(values = c("dry" = "#0072B2", "wet" = "#E69F00")) +
  labs(
    title = "Categorical by categorical interaction",
    x = "Habitat",
    y = "Expected growth",
    colour = "Season"
  ) +
  theme_drmtmb_gallery()
```

## Continuous by continuous interaction

For two continuous predictors, choose scientifically meaningful slices of one
predictor and show fitted slopes across the other. The values should be named
in the legend so the reader can interpret the conditioning.

```{r cont-cont-interaction}
set.seed(2593)
n_cc <- 180
soil <- data.frame(
  temperature = runif(n_cc, -1.5, 1.5),
  moisture = runif(n_cc, -1.4, 1.4)
)
soil$growth <- rnorm(
  n_cc,
  mean = 0.3 +
    0.55 * soil$temperature +
    0.45 * soil$moisture -
    0.35 * soil$temperature * soil$moisture,
  sd = 0.45
)

fit_cont_cont <- drmTMB(
  bf(growth ~ temperature * moisture, sigma ~ 1),
  family = gaussian(),
  data = soil
)

grid_cont_cont <- expand.grid(
  temperature = seq(-1.5, 1.5, length.out = 80),
  moisture = c(-1, 0, 1)
)
grid_cont_cont$moisture_level <- factor(
  grid_cont_cont$moisture,
  levels = c(-1, 0, 1),
  labels = c("low moisture", "average moisture", "high moisture")
)

pred_cont_cont <- predict_parameters(
  fit_cont_cont,
  newdata = grid_cont_cont,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(
  pred_cont_cont,
  aes(temperature, estimate, colour = moisture_level, fill = moisture_level)
) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.14, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  labs(
    title = "Continuous by continuous interaction",
    x = "Temperature anomaly",
    y = "Expected growth",
    colour = "Moisture slice",
    fill = "Moisture slice"
  ) +
  theme_drmtmb_gallery()
```

## Fitted correlation summaries

Correlation displays should keep layers separate. Residual `rho12`,
ordinary-group correlations, and phylogenetic or spatial correlations are not
the same parameter even when all are correlations.

```{r correlation-display}
pair_table <- data.frame(
  level = c("residual", "group", "phylogenetic"),
  class = c("rho12", "mean-slope", "trait"),
  parameter = c("trait coupling", "intercept-slope", "species covariance"),
  estimate = c(-0.32, 0.44, 0.61),
  modelled = c(TRUE, FALSE, FALSE),
  conf.low = c(-0.54, 0.12, 0.29),
  conf.high = c(-0.05, 0.68, 0.79),
  conf.status = c("wald", "profile", "profile")
)

plot_corpairs(pair_table, facet = "level") +
  labs(title = "Correlation layers stay separate") +
  theme_drmtmb_gallery()
```

## Simulation operating characteristics

Simulation figures answer different questions from model-interpretation
figures. They should show bias, error, coverage, runtime, and failures beside
the model surface being tested.

```{r simulation-operating-characteristics}
sim_summary <- data.frame(
  surface = rep(
    c("Gaussian location-scale", "Proportion", "Count"),
    each = 4
  ),
  estimand = rep(c("beta_x", "sigma", "sd_intercept", "rho12"), 3),
  bias = c(
    0.01, -0.02, 0.04, 0.00,
    -0.01, 0.03, NA, NA,
    0.02, -0.01, -0.04, NA
  ),
  rmse = c(
    0.07, 0.09, 0.13, 0.11,
    0.08, 0.10, NA, NA,
    0.10, 0.09, 0.16, NA
  ),
  coverage = c(
    0.95, 0.94, 0.91, 0.93,
    0.94, 0.92, NA, NA,
    0.93, 0.95, 0.89, NA
  )
)

ggplot(sim_summary, aes(estimand, bias, colour = surface, group = surface)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 2.2, na.rm = TRUE) +
  geom_line(na.rm = TRUE) +
  coord_flip() +
  scale_colour_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Bias by estimand",
    x = NULL,
    y = "Mean estimate minus truth",
    colour = "Surface"
  ) +
  theme_drmtmb_gallery()

ggplot(sim_summary, aes(estimand, coverage, fill = surface)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey60") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, na.rm = TRUE) +
  coord_flip(ylim = c(0.75, 1)) +
  scale_fill_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Interval coverage",
    x = NULL,
    y = "Empirical coverage",
    fill = "Surface"
  ) +
  theme_drmtmb_gallery()
```

These figures use a small illustrative table with the same columns that
simulation summaries should carry. The comprehensive simulation programme
should not be count-only. It should add continuous, meta-analysis, proportion,
count, ordinal, bivariate, random-slope, and structured-dependence surfaces as
each surface becomes ready. Real simulation reports should also show the run
manifest and warning/error ledger so missing or failed fits are not hidden
behind clean-looking panels.

## Future simulation result articles

Specialised simulation result articles should come later, after the
comprehensive simulation surfaces are ready. Those articles can be organised by
scientific question: power, bias, coverage, runtime, convergence, and failure
patterns. They should also be organised across data types, not only counts:
continuous location-scale models, meta-analysis with known `V`, proportions,
counts, ordinal responses, bivariate `rho12`, random-slope surfaces, and
structured dependence.

This figure gallery is broader and lighter. It shows the visual language users
should see across tutorials, examples, model interpretation, and later
simulation reports.
