---
title: "Figure gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Figure gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This gallery shows the kinds of figures `drmTMB` users can build for model
interpretation, uncertainty display, fitted interactions, fitted correlation
summaries, and later simulation diagnostics. It is different from the internal
count-pilot diagnostics used while developing the first Poisson/NB2 simulation
plumbing.

The examples here use small simulated data so the page can build quickly. They
are meant to show figure patterns, not biological conclusions. The interaction
figures deliberately keep three pieces visible: the observed data, the fitted
surface or cell means, and the conditioning values used to make the display.

```{r setup}
library(drmTMB)
library(ggplot2)

theme_drmtmb_gallery <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey30"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

habitat_palette <- c("reef" = "#0072B2", "kelp" = "#D55E00")
scale_drmtmb_colour <- scale_colour_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
scale_drmtmb_fill <- scale_fill_manual(
  values = habitat_palette,
  limits = names(habitat_palette)
)
```

## Fitted mean with a confidence band

Start with the figure most readers expect: raw observations, fitted mean, and a
95% confidence band on the response scale. This example is also a
categorical-by-continuous interaction: the temperature slope differs by
habitat.

```{r location-scale-fit, fig.alt="Scatterplot of simulated growth against temperature for two habitats, with fitted mean lines and 95 percent confidence bands for each habitat."}
set.seed(2591)
n <- 160
habitat_levels <- c("reef", "kelp")
fish <- data.frame(
  temperature = runif(n, -1.6, 1.6),
  habitat = factor(
    sample(habitat_levels, n, replace = TRUE),
    levels = habitat_levels
  )
)
eta_mu <- 0.4 +
  0.7 * fish$temperature +
  0.35 * (fish$habitat == "kelp") +
  0.18 * fish$temperature * (fish$habitat == "kelp")
eta_sigma <- -0.55 + 0.35 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit_growth <- drmTMB(
  bf(growth ~ temperature * habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)

grid_growth <- expand.grid(
  temperature = seq(-1.6, 1.6, length.out = 80),
  habitat = levels(fish$habitat)
)
pred_growth <- predict_parameters(
  fit_growth,
  newdata = grid_growth,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)
pred_growth$panel <- factor(
  ifelse(
    pred_growth$dpar == "mu",
    "Expected growth (mu)",
    "Residual SD (sigma)"
  ),
  levels = c("Expected growth (mu)", "Residual SD (sigma)")
)

pred_mu <- pred_growth[pred_growth$dpar == "mu", , drop = FALSE]

ggplot(fish, aes(temperature, growth, colour = habitat)) +
  geom_point(alpha = 0.35, size = 1.4) +
  geom_ribbon(
    data = pred_mu,
    aes(
      x = temperature,
      y = estimate,
      ymin = conf.low,
      ymax = conf.high,
      fill = habitat
    ),
    inherit.aes = FALSE,
    alpha = 0.16,
    colour = NA
  ) +
  geom_line(data = pred_mu, aes(y = estimate), linewidth = 0.8) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Expected growth along a temperature gradient",
    subtitle = "Raw observations, fitted mean, and 95% confidence band",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Habitat",
    fill = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

## Distributional-parameter surfaces

The same fit can show more than the mean. `predict_parameters()` returns a long
table with one row per distributional parameter, and
`plot_parameter_surface()` draws estimate lines plus finite interval bands when
they are present.

```{r parameter-surface, fig.alt="Two-panel plot showing fitted expected growth and residual standard deviation surfaces along temperature for reef and kelp habitats."}
plot_parameter_surface(
  pred_growth,
  x = "temperature",
  colour = "habitat",
  dpar = c("mu", "sigma"),
  facet = "panel"
) +
  scale_drmtmb_colour +
  scale_drmtmb_fill +
  labs(
    title = "Location and residual-scale surfaces",
    subtitle = "Both panels use the same prediction grid",
    x = "Temperature anomaly",
    y = "Estimate on response scale",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

The `mu` panel is the expected response. The `sigma` panel is the fitted
residual standard deviation, so an upward line means lower predictability or
larger residual heterogeneity at warmer values.

Distributional parameters are not all residual standard deviations. Shape,
inflation, hurdle, and residual-correlation parameters answer different
questions and often live on bounded response scales. The next panel set uses
three fitted models and the same plotting helper, but each panel names the
estimand explicitly.

```{r shape-inflation-rho12-panels, fig.alt="Three-panel plot showing a Student-t degrees-of-freedom shape parameter, a zero-inflation probability, and a residual rho12 correlation along a predictor gradient with 95 percent confidence bands."}
set.seed(2594)
n_tail <- 2500
tail_dat <- data.frame(tail_pressure = runif(n_tail, -1.5, 1.5))
tail_nu <- 2 + exp(log(4) - tail_dat$tail_pressure)
tail_dat$robust_trait <- 0.35 +
  0.45 * tail_dat$tail_pressure +
  0.55 * stats::rt(n_tail, df = tail_nu)
fit_tail <- drmTMB(
  bf(robust_trait ~ tail_pressure, sigma ~ 1, nu ~ tail_pressure),
  family = student(),
  data = tail_dat
)
tail_grid <- data.frame(tail_pressure = seq(-1.5, 1.5, length.out = 80))
pred_tail <- predict_parameters(
  fit_tail,
  newdata = tail_grid,
  dpar = "nu",
  conf.int = TRUE
)
pred_tail$gradient <- pred_tail$tail_pressure
pred_tail$panel <- "Tail weight (Student-t nu)"
pred_tail$model <- "Tail weight"

set.seed(2595)
n_count <- 180
count_dat <- data.frame(detection_difficulty = runif(n_count, -1.2, 1.2))
mu_count <- exp(0.9 + 0.35 * count_dat$detection_difficulty)
zi_count <- plogis(-1.1 + 1.25 * count_dat$detection_difficulty)
count_dat$count <- ifelse(
  runif(n_count) < zi_count,
  0,
  stats::rpois(n_count, mu_count)
)
fit_zi <- drmTMB(
  bf(count ~ detection_difficulty, zi ~ detection_difficulty),
  family = poisson(link = "log"),
  data = count_dat
)
zi_grid <- data.frame(detection_difficulty = seq(-1.2, 1.2, length.out = 80))
pred_zi <- predict_parameters(
  fit_zi,
  newdata = zi_grid,
  dpar = "zi",
  conf.int = TRUE
)
pred_zi$gradient <- pred_zi$detection_difficulty
pred_zi$panel <- "Structural-zero probability (zi)"
pred_zi$model <- "Structural-zero probability"

set.seed(2596)
n_pair <- 160
pair_dat <- data.frame(disturbance = runif(n_pair, -1.4, 1.4))
rho_true <- tanh(-0.15 + 0.75 * pair_dat$disturbance)
z1 <- stats::rnorm(n_pair)
z2 <- stats::rnorm(n_pair)
pair_dat$activity <- 0.25 + 0.35 * pair_dat$disturbance + z1
pair_dat$boldness <- -0.1 +
  0.2 * pair_dat$disturbance +
  rho_true * z1 +
  sqrt(1 - rho_true^2) * z2
fit_pair <- drmTMB(
  bf(
    mu1 = activity ~ disturbance,
    mu2 = boldness ~ disturbance,
    sigma1 ~ 1,
    sigma2 ~ 1,
    rho12 ~ disturbance
  ),
  family = c(gaussian(), gaussian()),
  data = pair_dat
)
rho_grid <- data.frame(disturbance = seq(-1.4, 1.4, length.out = 80))
pred_rho <- predict_parameters(
  fit_pair,
  newdata = rho_grid,
  dpar = "rho12",
  conf.int = TRUE
)
pred_rho$gradient <- pred_rho$disturbance
pred_rho$panel <- "Residual correlation (rho12)"
pred_rho$model <- "Residual correlation"

pred_extra <- rbind(
  pred_tail[intersect(names(pred_tail), names(pred_zi))],
  pred_zi[intersect(names(pred_tail), names(pred_zi))],
  pred_rho[intersect(names(pred_tail), names(pred_zi))]
)
pred_extra$panel <- factor(
  pred_extra$panel,
  levels = c(
    "Tail weight (Student-t nu)",
    "Structural-zero probability (zi)",
    "Residual correlation (rho12)"
  )
)

plot_parameter_surface(
  pred_extra,
  x = "gradient",
  colour = "model",
  facet = "panel",
  dpar = c("nu", "zi", "rho12"),
  point = FALSE
) +
  geom_hline(
    data = data.frame(
      .drmTMB_plot_facet = factor(
        "Residual correlation (rho12)",
        levels = levels(pred_extra$panel)
      )
    ),
    aes(yintercept = 0),
    inherit.aes = FALSE,
    linetype = "dashed",
    colour = "grey55"
  ) +
  scale_colour_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "Tail weight" = "#0072B2",
    "Structural-zero probability" = "#009E73",
    "Residual correlation" = "#D55E00"
  )) +
  labs(
    title = "Shape, inflation, and residual-correlation panels",
    subtitle = "Each panel is a different distributional parameter on its response scale",
    x = "Predictor gradient",
    y = "Estimate on response scale",
    colour = "Parameter",
    fill = "Parameter"
  ) +
  theme_drmtmb_gallery() +
  theme(legend.position = "none")
```

The `nu` panel is a Student-t degrees-of-freedom parameter: lower values mean
heavier tails, and here the fitted model estimates a changing tail shape
through `nu ~ tail_pressure`. The `zi` panel is a structural-zero probability,
not a mean count. The `rho12` panel is residual coupling between two responses
after their own means and scales have been modelled.

## Random-effect and variance-component figures

Random-effect figures need a different visual grammar from residual-scale
figures. Residual `sigma` is the observation-level standard deviation. A
group-level SD, such as the SD of site intercepts or temperature slopes, is
among-group variation in a fitted random effect.

```{r random-effect-variance-components, fig.alt="Dot plot comparing residual sigma, among-site intercept SD, and among-site temperature-slope SD on their fitted response-scale units."}
set.seed(2621)
n_site <- 24
n_site_obs <- 6
site_levels <- seq_len(n_site)
site_random <- data.frame(
  site = factor(site_levels),
  intercept = rnorm(n_site, mean = 0, sd = 0.42),
  slope = rnorm(n_site, mean = 0, sd = 0.24)
)
reef_re <- data.frame(
  site = factor(rep(site_levels, each = n_site_obs), levels = site_levels)
)
reef_re$temperature <- rep(seq(-1, 1, length.out = n_site_obs), n_site)
site_lookup <- as.integer(reef_re$site)
reef_re$growth <- 0.25 +
  0.62 * reef_re$temperature +
  site_random$intercept[site_lookup] +
  site_random$slope[site_lookup] * reef_re$temperature +
  rnorm(nrow(reef_re), mean = 0, sd = 0.34)

fit_re_slopes <- drmTMB(
  bf(growth ~ temperature + (1 + temperature | site), sigma ~ 1),
  family = gaussian(),
  data = reef_re,
  control = drm_control(optimizer = list(eval.max = 160L, iter.max = 160L))
)

component_rows <- summary(fit_re_slopes)$parameters
variance_components <- data.frame(
  component = c(
    "Residual sigma",
    "Site intercept SD",
    "Site temperature-slope SD"
  ),
  estimate = c(
    component_rows["sigma", "estimate"],
    component_rows[
      "sd:mu:(1 + temperature | site):(Intercept)",
      "estimate"
    ],
    component_rows[
      "sd:mu:(1 + temperature | site):temperature",
      "estimate"
    ]
  )
)
variance_components$component <- factor(
  variance_components$component,
  levels = rev(variance_components$component)
)

ggplot(variance_components, aes(estimate, component)) +
  geom_point(size = 2.8, colour = "#0072B2") +
  geom_segment(
    aes(x = 0, xend = estimate, yend = component),
    linewidth = 0.7,
    colour = "#0072B2",
    alpha = 0.55
  ) +
  labs(
    title = "Residual scale and group-level SDs are different quantities",
    subtitle = "The slope SD measures among-site variation in temperature responses",
    x = "Fitted standard deviation",
    y = NULL
  ) +
  theme_drmtmb_gallery()
```

The dot plot compares three standard deviations, but only one is residual
`sigma`. The other two are random-effect SDs: how much sites differ in baseline
growth, and how much sites differ in the temperature slope. For intervals on
these SDs, use `profile_targets(fit_re_slopes)` to choose direct profile-ready
targets rather than treating the quick point plot as an uncertainty analysis.

Conditional random effects answer a related but different question. They are
shrunken site-level deviations from the population intercept or slope, not
variance components.

```{r random-slope-modes, fig.alt="Ordered dot plot of conditional site-level temperature-slope deviations from a fitted random-slope model, centered around zero."}
site_slope <- ranef(fit_re_slopes, "mu")$terms[[
  "(1 + temperature | site):temperature"
]]
site_slope <- data.frame(
  site = names(site_slope),
  slope_deviation = unname(site_slope)
)
site_slope$site <- factor(
  site_slope$site,
  levels = site_slope$site[order(site_slope$slope_deviation)]
)

ggplot(site_slope, aes(slope_deviation, site)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 1.7, colour = "#D55E00") +
  labs(
    title = "Conditional random-slope deviations",
    subtitle = "Each point is a site-specific deviation from the average temperature slope",
    x = "Deviation in temperature slope",
    y = "Site"
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

A direct `sd(site) ~ reef_cover` formula asks a third question: does a
site-level predictor change the among-site SD in expected growth? The
prediction table keeps this target as `sd(site)`, not residual `sigma`.

```{r random-effect-sd-surface, fig.alt="Line plot showing a fitted among-site random-intercept SD surface along reef cover, with no confidence band because direct random-effect SD surfaces do not yet supply Wald intervals."}
set.seed(2622)
n_sd_site <- 28
n_sd_each <- 5
site_sd_data <- data.frame(
  site = factor(seq_len(n_sd_site)),
  reef_cover = runif(n_sd_site, -1.2, 1.2)
)
site_sd <- exp(-0.6 + 0.45 * site_sd_data$reef_cover)
site_deviation <- rnorm(n_sd_site, mean = 0, sd = site_sd)
reef_sd <- site_sd_data[rep(seq_len(n_sd_site), each = n_sd_each), ]
reef_sd$temperature <- rep(seq(-1.2, 1.2, length.out = n_sd_each), n_sd_site)
reef_sd$growth <- 0.4 +
  0.5 * reef_sd$temperature +
  site_deviation[as.integer(reef_sd$site)] +
  rnorm(nrow(reef_sd), mean = 0, sd = 0.35)

fit_site_sd <- drmTMB(
  bf(growth ~ temperature + (1 | site), sigma ~ 1, sd(site) ~ reef_cover),
  family = gaussian(),
  data = reef_sd,
  control = drm_control(optimizer = list(eval.max = 180L, iter.max = 180L))
)

grid_site_sd <- prediction_grid(
  fit_site_sd,
  focal = "reef_cover",
  at = list(reef_cover = seq(-1.2, 1.2, length.out = 80))
)
pred_site_sd <- predict_parameters(
  fit_site_sd,
  newdata = grid_site_sd,
  dpar = "sd(site)",
  conf.int = TRUE
)

plot_parameter_surface(
  pred_site_sd,
  x = "reef_cover",
  facet = NULL,
  point = FALSE
) +
  labs(
    title = "Among-site SD surface",
    subtitle = "Requested Wald intervals remain unavailable for direct random-effect SD surfaces",
    x = "Reef cover",
    y = "Among-site SD in expected growth"
  ) +
  theme_drmtmb_gallery()
```

This line is a fitted random-effect SD surface. It should not be read as a raw
growth curve, and it should not be overlaid with raw `growth` points. The
table behind the plot reports `conf.status = "wald_unavailable"`, which is the
honest interval boundary until a profile or bootstrap route is attached to
these derived surfaces.

## Discrete comparisons

For a factor predictor, the same plotting helper uses interval bars rather than
ribbons. This is useful for compact tutorial figures and estimated marginal
comparisons.

```{r discrete-comparison, fig.alt="Point interval plot comparing expected growth between reef and kelp habitats at average temperature."}
habitat_grid <- data.frame(
  temperature = 0,
  habitat = factor(c("reef", "kelp"), levels = levels(fish$habitat))
)
pred_habitat <- predict_parameters(
  fit_growth,
  newdata = habitat_grid,
  dpar = "mu",
  conf.int = TRUE
)

plot_parameter_surface(
  pred_habitat,
  x = "habitat",
  line = FALSE,
  point = TRUE
) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated habitat contrast at average temperature",
    subtitle = "Intervals are Wald intervals from predict_parameters()",
    x = NULL,
    y = "Expected growth"
  ) +
  theme_drmtmb_gallery()
```

## Categorical by categorical interaction

For two categorical predictors, show every fitted cell and its interval. Use
position dodging so the uncertainty intervals do not sit on top of each other.
Raw observations can sit behind the intervals when they help readers see the
data support for each cell.

```{r cat-cat-interaction, fig.alt="Categorical by categorical interaction plot showing raw growth observations and fitted growth intervals for each habitat and season combination."}
set.seed(2592)
n_cell <- 30
reef <- expand.grid(
  habitat = factor(habitat_levels, levels = habitat_levels),
  season = factor(c("dry", "wet"), levels = c("dry", "wet")),
  replicate = seq_len(n_cell)
)
reef$mu <- with(
  reef,
  0.6 +
    0.35 * (habitat == "kelp") -
    0.15 * (season == "wet") +
    0.45 * (habitat == "kelp") * (season == "wet")
)
reef$growth <- rnorm(nrow(reef), mean = reef$mu, sd = 0.45)

fit_cat_cat <- drmTMB(
  bf(growth ~ habitat * season, sigma ~ 1),
  family = gaussian(),
  data = reef
)

grid_cat_cat <- expand.grid(
  habitat = levels(reef$habitat),
  season = levels(reef$season)
)
pred_cat_cat <- predict_parameters(
  fit_cat_cat,
  newdata = grid_cat_cat,
  dpar = "mu",
  conf.int = TRUE
)

raw_cat_cat <- reef

ggplot(pred_cat_cat, aes(habitat, estimate, colour = season)) +
  geom_jitter(
    data = raw_cat_cat,
    aes(y = growth, colour = season),
    width = 0.11,
    height = 0,
    alpha = 0.22,
    size = 1.1,
    show.legend = FALSE
  ) +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.35),
    linewidth = 0.65,
    fatten = 2.3
  ) +
  scale_colour_manual(values = c("dry" = "#0072B2", "wet" = "#E69F00")) +
  labs(
    title = "Categorical by categorical interaction",
    subtitle = "Small points are observations; intervals are fitted cell means",
    x = "Habitat",
    y = "Growth",
    colour = "Season"
  ) +
  theme_drmtmb_gallery()
```

## Estimated marginal means and marginal summaries

For the first supported `emmeans` path, a fixed-effect univariate `mu` model can
produce estimated marginal means that are easy to turn into publication-style
figures. The same rule applies: keep the target and conditioning values
visible. These displays are `mu` summaries, not `sigma`, fitted-response,
random-effect, or bivariate summaries.

```{r emmeans-display, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.alt="Point interval plot of estimated marginal mean growth for reef and kelp habitats conditioned at temperature zero."}
emm_habitat <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat,
      at = list(temperature = 0),
      type = "response"
    )
  )
)

ggplot(emm_habitat, aes(habitat, emmean, colour = habitat)) +
  geom_pointrange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    linewidth = 0.7,
    fatten = 2.4
  ) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated marginal mean growth",
    subtitle = "Supported fixed-effect mu target, conditioned at temperature = 0",
    x = NULL,
    y = "Expected growth",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

`emmeans` reference grids can also condition on a factor. In the next display,
the `mu` cell means are grouped by season, so the reader sees the fitted
habitat contrast inside each season rather than a single averaged contrast.

```{r emmeans-factor-grid, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.alt="Point interval plot of estimated marginal mean growth for reef and kelp habitats shown separately for dry and wet seasons."}
emm_season <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_cat_cat,
      ~ habitat | season,
      type = "response"
    )
  )
)

ggplot(emm_season, aes(habitat, emmean, colour = season)) +
  geom_pointrange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    position = position_dodge(width = 0.35),
    linewidth = 0.65,
    fatten = 2.3
  ) +
  scale_colour_manual(values = c("dry" = "#0072B2", "wet" = "#E69F00")) +
  labs(
    title = "Factor-conditioned estimated marginal means",
    subtitle = "Habitat mu summaries are shown separately for each season",
    x = "Habitat",
    y = "Expected growth",
    colour = "Season"
  ) +
  theme_drmtmb_gallery()
```

For a numeric interaction, use an explicit reference grid. Here `temperature`
is held at three named values so the habitat contrast does not hide the
conditioning rule.

```{r emmeans-interaction-grid, eval = has_ggplot2 && requireNamespace("emmeans", quietly = TRUE), fig.alt="Point interval plot of estimated marginal mean growth for reef and kelp habitats at three temperature values."}
emm_temperature <- suppressMessages(
  as.data.frame(
    emmeans::emmeans(
      fit_growth,
      ~ habitat | temperature,
      at = list(temperature = c(-1, 0, 1)),
      type = "response"
    )
  )
)
emm_temperature$temperature_slice <- factor(
  emm_temperature$temperature,
  levels = c(-1, 0, 1),
  labels = c("cooler", "average", "warmer")
)

ggplot(emm_temperature, aes(temperature_slice, emmean, colour = habitat)) +
  geom_pointrange(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    position = position_dodge(width = 0.42),
    linewidth = 0.65,
    fatten = 2.3
  ) +
  scale_drmtmb_colour +
  labs(
    title = "Estimated marginal means across an interaction grid",
    subtitle = "The same fixed-effect mu bridge, evaluated at named temperature slices",
    x = "Temperature slice",
    y = "Expected growth",
    colour = "Habitat"
  ) +
  theme_drmtmb_gallery()
```

Empirical marginal summaries use a different rule. `prediction_grid(...,
margin = "empirical")` crosses a focal value with fitted rows, and
`marginal_parameters()` averages the resulting parameter predictions. This
point display deliberately has no interval bars because the current empirical
summary is a plug-in average.

```{r empirical-marginal-summary, fig.alt="Point plot of empirical marginal expected growth for reef and kelp habitats without interval bars."}
empirical_habitat_grid <- prediction_grid(
  fit_growth,
  focal = "habitat",
  at = list(habitat = levels(fish$habitat)),
  margin = "empirical"
)
empirical_habitat <- marginal_parameters(
  fit_growth,
  newdata = empirical_habitat_grid,
  dpar = "mu",
  by = "habitat"
)

plot_parameter_surface(
  empirical_habitat,
  x = "habitat",
  facet = NULL,
  line = FALSE,
  point = TRUE
) +
  scale_drmtmb_colour +
  labs(
    title = "Empirical marginal mu summary",
    subtitle = "Averaged over the fitted-row covariate distribution",
    x = NULL,
    y = "Average expected growth"
  ) +
  theme_drmtmb_gallery()
```

Unsupported targets should be visible before a reader tries to plot them. The
current `emmeans` bridge is for fixed-effect univariate `mu`; other targets
should use parameter prediction, `corpairs()`, profile intervals, or a future
feature-specific design.

```{r emmeans-boundary-strip, fig.alt="Tile plot showing fixed-effect univariate mu as the supported emmeans target and sigma, bivariate, zero-inflated, hurdle, ordinal, and random-effect targets as unsupported boundaries."}
emmeans_boundary <- data.frame(
  target = c(
    "Fixed-effect univariate mu",
    "sigma",
    "Bivariate responses",
    "Zero-inflated or hurdle means",
    "Ordinal expected scores",
    "Random-effect targets"
  ),
  status = c(
    "Supported emmeans route",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary",
    "Unsupported boundary"
  )
)
emmeans_boundary$target <- factor(
  emmeans_boundary$target,
  levels = rev(emmeans_boundary$target)
)
emmeans_boundary$status <- factor(
  emmeans_boundary$status,
  levels = c("Supported emmeans route", "Unsupported boundary")
)
emmeans_boundary$status_label <- ifelse(
  emmeans_boundary$status == "Supported emmeans route",
  "supported\nmu EMM",
  "unsupported\nboundary"
)

ggplot(emmeans_boundary, aes(status, target, fill = status)) +
  geom_tile(width = 0.84, height = 0.72, colour = "white", linewidth = 0.6) +
  geom_text(aes(label = status_label), size = 3.1, lineheight = 0.9) +
  scale_fill_manual(values = c(
    "Supported emmeans route" = "#0072B2",
    "Unsupported boundary" = "#E69F00"
  )) +
  labs(
    title = "Current emmeans support boundary",
    subtitle = "Only fixed-effect univariate mu is advertised as an emmeans target",
    x = "Current status",
    y = NULL,
    fill = "Status"
  ) +
  theme_drmtmb_gallery() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

## Continuous by continuous interaction

For two continuous predictors, choose scientifically meaningful slices of one
predictor and show fitted slopes across the other. The values should be named
in the legend so the reader can interpret the conditioning.

```{r cont-cont-interaction, fig.alt="Continuous by continuous interaction plot showing raw observations in grey and fitted temperature slopes at low, average, and high moisture with confidence bands."}
set.seed(2593)
n_cc <- 180
soil <- data.frame(
  temperature = runif(n_cc, -1.5, 1.5),
  moisture = runif(n_cc, -1.4, 1.4)
)
soil$growth <- rnorm(
  n_cc,
  mean = 0.3 +
    0.55 * soil$temperature +
    0.45 * soil$moisture -
    0.35 * soil$temperature * soil$moisture,
  sd = 0.45
)

fit_cont_cont <- drmTMB(
  bf(growth ~ temperature * moisture, sigma ~ 1),
  family = gaussian(),
  data = soil
)

grid_cont_cont <- expand.grid(
  temperature = seq(-1.5, 1.5, length.out = 80),
  moisture = c(-1, 0, 1)
)
grid_cont_cont$moisture_level <- factor(
  grid_cont_cont$moisture,
  levels = c(-1, 0, 1),
  labels = c("low moisture", "average moisture", "high moisture")
)

pred_cont_cont <- predict_parameters(
  fit_cont_cont,
  newdata = grid_cont_cont,
  dpar = "mu",
  conf.int = TRUE
)

ggplot(
  pred_cont_cont,
  aes(temperature, estimate, colour = moisture_level, fill = moisture_level)
) +
  geom_point(
    data = soil,
    aes(temperature, growth),
    inherit.aes = FALSE,
    alpha = 0.18,
    size = 1,
    colour = "grey35"
  ) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.14, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  scale_fill_manual(values = c(
    "low moisture" = "#0072B2",
    "average moisture" = "#555555",
    "high moisture" = "#D55E00"
  )) +
  labs(
    title = "Continuous by continuous interaction",
    subtitle = "Moisture is held at three named slices",
    x = "Temperature anomaly",
    y = "Growth",
    colour = "Moisture slice",
    fill = "Moisture slice"
  ) +
  theme_drmtmb_gallery()
```

The raw point cloud is not coloured by the sliced predictor because the fitted
lines are conditional displays, not clusters in the data. This keeps the
visual message honest: the model is being read at three moisture values, while
the observations span the whole moisture gradient.

## Fitted correlation summaries

Correlation displays should keep layers separate. Residual `rho12`,
ordinary-group correlations, phylogenetic correlations, spatial correlations,
and future animal or `relmat()` layers are not the same parameter even when all
are correlations. A fitted workflow should start from `corpairs(fit)`; this
gallery uses a compact compatible table so the page builds quickly while
preserving the same required columns.

```{r correlation-display, fig.alt="Faceted point interval plot showing separate residual, group, and phylogenetic correlation layers with short row labels."}
pair_table <- data.frame(
  level = c("residual", "group", "phylogenetic"),
  class = c("residual", "mean-slope", "mean-mean"),
  parameter = c(
    "rho12",
    "cor((Intercept),temperature | site)",
    "cor(mu1,mu2 | phylo | species)"
  ),
  estimate = c(-0.32, 0.44, 0.61),
  modelled = c(TRUE, FALSE, FALSE),
  conf.low = c(-0.54, 0.12, 0.29),
  conf.high = c(-0.05, 0.68, 0.79),
  conf.status = c("wald", "profile", "profile"),
  layer = c("Residual coscale", "Ordinary group", "Phylogenetic")
)
pair_table$layer <- factor(
  pair_table$layer,
  levels = c("Residual coscale", "Ordinary group", "Phylogenetic")
)
pair_table$display <- factor(
  c(
    "Residual\nrho12",
    "Group\nintercept-slope",
    "Phylogenetic\nlocation-location"
  ),
  levels = c(
    "Residual\nrho12",
    "Group\nintercept-slope",
    "Phylogenetic\nlocation-location"
  )
)

correlation_palette <- c(
  "Residual coscale" = "#0072B2",
  "Ordinary group" = "#009E73",
  "Phylogenetic" = "#D55E00"
)

plot_corpairs(pair_table, colour = "layer", facet = "layer", label = "display") +
  scale_colour_manual(values = correlation_palette) +
  labs(
    title = "Implemented correlation rows stay in their own layers",
    subtitle = "Intervals are drawn only when the pair table has finite bounds",
    colour = "Layer"
  ) +
  theme_drmtmb_gallery() +
  theme(
    axis.text.y = element_text(size = 10, lineheight = 0.95),
    legend.position = "none"
  )
```

The residual row is the fitted within-observation coscale `rho12`. The group
row is a latent random-effect correlation, here an intercept-slope row from an
ordinary grouped model. The phylogenetic row is a species-level structured
location-location correlation. These rows can live in the same `corpairs()`
table, but the figure keeps them apart because they answer different
biological questions.

```{r correlation-layer-boundaries, fig.alt="Tile plot showing implemented residual, group, and phylogenetic correlation-pair layers separately from planned spatial, animal, and relmat boundaries."}
layer_boundary <- data.frame(
  layer = c(
    "Residual rho12",
    "Ordinary group",
    "Phylogenetic",
    "Spatial",
    "Animal",
    "relmat"
  ),
  status = c(
    "Fitted corpairs row",
    "Fitted corpairs row",
    "Fitted corpairs row",
    "Planned boundary",
    "Planned boundary",
    "Planned boundary"
  )
)
layer_boundary$layer <- factor(
  layer_boundary$layer,
  levels = rev(layer_boundary$layer)
)
layer_boundary$status <- factor(
  layer_boundary$status,
  levels = c("Fitted corpairs row", "Planned boundary")
)
layer_boundary$status_label <- ifelse(
  layer_boundary$status == "Fitted corpairs row",
  "fitted\nrow",
  "planned\nboundary"
)

ggplot(layer_boundary, aes(status, layer, fill = status)) +
  geom_tile(width = 0.82, height = 0.72, colour = "white", linewidth = 0.6) +
  geom_text(aes(label = status_label), size = 3.1, lineheight = 0.9) +
  scale_fill_manual(values = c(
    "Fitted corpairs row" = "#0072B2",
    "Planned boundary" = "#E69F00"
  )) +
  labs(
    title = "Correlation-layer support boundary",
    subtitle = "Spatial, animal, and relmat rows are reserved until fitted correlation-pair evidence exists",
    x = "Current status",
    y = NULL,
    fill = "Status"
  ) +
  theme_drmtmb_gallery() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

The status strip is not an estimate plot. Coordinate-spatial Gaussian `mu`
slopes are fitted elsewhere, but spatial `corpairs()` rows remain planned.
`animal()` and `relmat()` are also reserved structured-effect markers until the
likelihood, diagnostics, profile-target, recovery-test, and example evidence
exist.

## Simulation operating characteristics

Simulation figures answer different questions from model-interpretation
figures. They should show bias, error, coverage, runtime, and failures beside
the model surface being tested.

```{r simulation-operating-characteristics, fig.alt="Simulation summary plots showing illustrative bias and interval coverage by estimand across Gaussian location-scale, proportion, and count surfaces."}
sim_summary <- data.frame(
  surface = rep(
    c("Gaussian location-scale", "Proportion", "Count"),
    each = 4
  ),
  estimand = rep(c("beta_x", "sigma", "sd_intercept", "rho12"), 3),
  bias = c(
    0.01, -0.02, 0.04, 0.00,
    -0.01, 0.03, NA, NA,
    0.02, -0.01, -0.04, NA
  ),
  rmse = c(
    0.07, 0.09, 0.13, 0.11,
    0.08, 0.10, NA, NA,
    0.10, 0.09, 0.16, NA
  ),
  coverage = c(
    0.95, 0.94, 0.91, 0.93,
    0.94, 0.92, NA, NA,
    0.93, 0.95, 0.89, NA
  )
)

ggplot(sim_summary, aes(estimand, bias, colour = surface, group = surface)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 2.2, na.rm = TRUE) +
  geom_line(na.rm = TRUE) +
  coord_flip() +
  scale_colour_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Bias by estimand",
    x = NULL,
    y = "Mean estimate minus truth",
    colour = "Surface"
  ) +
  theme_drmtmb_gallery()

ggplot(sim_summary, aes(estimand, coverage, fill = surface)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey60") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, na.rm = TRUE) +
  coord_flip(ylim = c(0.75, 1)) +
  scale_fill_manual(values = c(
    "Gaussian location-scale" = "#0072B2",
    "Proportion" = "#009E73",
    "Count" = "#CC79A7"
  )) +
  labs(
    title = "Interval coverage",
    x = NULL,
    y = "Empirical coverage",
    fill = "Surface"
  ) +
  theme_drmtmb_gallery()
```

These figures use a small illustrative table with the same columns that
simulation summaries should carry. The comprehensive simulation programme
should not be count-only. It should add continuous, meta-analysis, proportion,
count, ordinal, bivariate, random-slope, and structured-dependence surfaces as
each surface becomes ready. Real simulation reports should also show the run
manifest and warning/error ledger so missing or failed fits are not hidden
behind clean-looking panels.

## Future simulation result articles

Specialised simulation result articles should come later, after the
comprehensive simulation surfaces are ready. Those articles can be organised by
scientific question: power, bias, coverage, runtime, convergence, and failure
patterns. They should also be organised across data types, not only counts:
continuous location-scale models, meta-analysis with known `V`, proportions,
counts, ordinal responses, bivariate `rho12`, random-slope surfaces, and
structured dependence.

This figure gallery is broader and lighter. It shows the visual language users
should see across tutorials, examples, model interpretation, and later
simulation reports.
