---
title: "How drmTMB fits: the integral and optimiser layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How drmTMB fits: the integral and optimiser layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Fitting a model with random effects in `drmTMB` is **two nested problems**, and
almost all of the machinery you hear about — the Laplace approximation,
automatic differentiation, `nlminb`, REML, the Julia engine — is really an
answer to one of the two. Separating them is the quickest way to understand what
happens when you call `drmTMB()`, and which control to reach for when a fit
struggles.

> **The one-line version.** The **integral layer** turns the intractable
> marginal likelihood into something the computer can evaluate; the **optimiser
> layer** climbs it. They are *orthogonal* — the optimiser does not care how the
> marginal was approximated, and the approximation does not care which optimiser
> climbs it.

## The two questions

A model has fixed parameters `θ` (regression coefficients, variance components,
dispersion, shape) and latent random effects `u` (study/group effects,
phylogenetic deviations, residual correlations). `drmTMB` does not maximise the
joint likelihood of the data *and* `u`; it integrates `u` out, leaving a
**marginal** likelihood in `θ` alone:

```
L(θ) = ∫ p(y | u, θ) · p(u | θ) du.
```

Fitting then splits cleanly:

1. **The integral (marginal) layer** — make `L(θ)` computable. The integral has
   no closed form for non-Gaussian families, so it must be approximated.
2. **The optimiser layer** — maximise the resulting `L̂(θ)` over `θ`.

```
+-- OUTER . optimise theta -------------------------------------+
|   climb L^(theta):  nlminb (default) / optim . ML or REML     |   <- optimiser layer
+---------------------------------------------------------------+
        v  every evaluation of L^(theta) first builds the marginal ...
+-- MIDDLE . the marginal  L(theta) = int p(y|u,theta) p(u|theta) du --+
|   make it tractable:  Laplace approximation (via TMB)                |   <- integral layer
+----------------------------------------------------------------------+
        v  which in turn must handle the latent u ...
+-- INNER . the latent u ---------------------------------------+
|   Laplace -> inner Newton finds the mode u^(theta)            |
+---------------------------------------------------------------+
```

## Layer 1 — the integral, handled by TMB

`drmTMB` delegates the marginal to **Template Model Builder (TMB)**, which does
two things:

- **The Laplace approximation.** TMB replaces the random-effect integrand with a
  Gaussian centred at its mode `û(θ)`, matched in curvature (the Hessian) there.
  Finding that mode is an *inner* optimisation that TMB runs for you on every
  outer step. For a Gaussian random effect on the mean the approximation is
  *exact*; it is an approximation precisely when the integrand is non-Gaussian
  (binary, count, skewed), which is where dispersion and shape parameters pick up
  any error first.
- **Exact gradients by automatic differentiation.** TMB differentiates the
  Laplace-approximated marginal *through* the inner mode (the implicit function
  theorem), so the optimiser receives an exact `∇L̂(θ)` rather than a
  finite-difference estimate. This is what makes the outer optimisation fast and
  stable even with many variance parameters.

## Layer 2 — the optimiser

`drmTMB` hands TMB's `(L̂, ∇L̂)` to a standard R optimiser:

- **`nlminb` is the default** deterministic outer optimiser. When a fit returns
  an optimiser error, `drmTMB` automatically retries with the built-in
  `"careful"` and `"robust"` presets and records which one succeeded — see
  [Improving convergence](convergence.html).
- **`optim` (BFGS / L-BFGS-B)** is available as an alternative. L-BFGS is a
  *quasi-Newton* method — it learns curvature from successive gradients instead
  of forming a Hessian, so it needs only `∇L̂` and scales to many parameters. (It
  is the default optimiser in the related `gllvm` package, and in `drmTMB`'s
  Julia engine.)
- **ML by default; REML optional.** Setting `REML = TRUE` changes *which*
  likelihood you climb — TMB integrates the fixed effects out as well, the same
  restricted likelihood `lme4` and `metafor` target — not the optimiser that
  climbs it.

```{r, eval = FALSE}
# Same model, same integral layer (TMB Laplace) — different optimiser-layer choices:
fit    <- drmTMB(bf(y ~ x + (1 | study)), data = d)                 # nlminb, ML
fit_re <- drmTMB(bf(y ~ x + (1 | study)), data = d, REML = TRUE)    # restricted likelihood
```

## The Julia engine swaps the integral layer

`drmTMB(..., engine = "julia")` keeps the grammar, the data marshalling, and the
post-fit methods exactly as they are, but routes the **integral layer** to the
companion `DRM.jl` engine. That engine adds marginal approximations beyond
Laplace — a variational/ELBO option and Gauss–Hermite quadrature for the
cross-family bivariate models — and contributes an `O(p)` sparse phylogenetic
path. The optimiser layer is again L-BFGS on the Julia side. See
[Running models with the Julia engine](julia-engine.html).

## Why "orthogonal" matters

Because the optimiser only ever consumes a value and a gradient, the two layers
move independently: you can switch `engine = "tmb"` ↔ `"julia"` (changing the
integral layer) without changing how you read the fit, and you can change
optimiser presets without touching the likelihood. It also explains where other
methods sit — *EM*, for instance, is an alternative to the **optimiser** layer
(it ascends by alternating an E-step and an M-step instead of climbing a
gradient), while the *variational* approximation is an alternative in the
**integral** layer.

A practical consequence: **profile-likelihood confidence intervals are the
costliest thing the engine does**, because the layers nest. Every outer `θ`-step
triggers an inner mode-find, and a profile CI adds a *third* level — fix one
component of `θ`, re-optimise the rest, and repeat that constrained fit along a
search for the likelihood-ratio threshold.

## Where each piece lives

| Layer | Question it answers | In `drmTMB` |
|---|---|---|
| **Integral** | how to handle the latent `u` | Laplace via **TMB** (`engine = "tmb"`); Laplace / variational / Gauss–Hermite via **DRM.jl** (`engine = "julia"`) |
| **Optimiser** | how to maximise `L̂(θ)` | **`nlminb`** (default) with `careful`/`robust` retries; `optim` (BFGS / L-BFGS-B); **ML** or **REML** |

## See also

- [Improving convergence](convergence.html) — the optimiser-layer controls in practice.
- [When variance carries signal](location-scale.html) — what the marginal is estimating.
- [Running models with the Julia engine](julia-engine.html) — swapping the integral layer.
