---
title: "Using the Julia engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Julia engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

The Julia engine is available now as an **optional backend** for `drmTMB`.
You keep the R formula and data interface and select `engine = "julia"` when
fitting a supported model. The default, `engine = "tmb"`, does not require Julia.

Julia is useful for repeated fitting and for models that can use DRM.jl's sparse
structured solvers. It also fits ordinary regression models: a large phylogeny
is not a prerequisite. Which backend is faster depends on the model, requested
inference, data size, and whether Julia has already started and compiled the code.

This guide starts with an ordinary regression, then shows group-level and
phylogenetic scale regression. The examples are not run during the standard R
package build, so that installing or reading `drmTMB` does not require Julia.

## Setup and prerequisites

Install **Julia 1.10 or later** and the optional R package `JuliaCall`. DRM.jl
currently needs a local source checkout; installing a registered package named
`DRM` is not the setup route used by this bridge.

Run once in a terminal, choosing a directory where you keep source packages:

```sh
git clone https://github.com/itchyshin/DRM.jl.git
julia --project=DRM.jl -e 'using Pkg; Pkg.instantiate()'
```

Then, in R, set the checkout path **before the first Julia call**:

```{r julia-install, eval=FALSE}
install.packages("JuliaCall")
Sys.setenv(DRM_JL_PATH = "/absolute/path/to/DRM.jl")
library(drmTMB)
```

The same setup works in ordinary noninteractive scripts run with `Rscript`;
no test opt-in is required. Julia startup remains disabled in marked R package
check processes unless explicitly enabled for repository testing. Do not enable
live Julia tests on CRAN or win-builder.

Replace the path with the directory containing DRM.jl's `Project.toml` and
`src/`. The bridge initializes Julia when needed. The first call includes
startup and compilation; time later calls separately when comparing repeated
workflows. Keep the checkout revision with your analysis so results can be
reproduced; a package version alone may describe more than one source build.

## Start with an ordinary regression

These data are generated in R and use the same formula for both backends:

```{r julia-ordinary, eval=FALSE}
set.seed(42)
dat <- data.frame(x = seq(-1, 1, length.out = 120))
dat$y <- 1 + 0.8 * dat$x + rnorm(nrow(dat), sd = exp(-0.4 + 0.2 * dat$x))
form <- bf(y ~ x, sigma ~ x)

fit_r <- drmTMB(form, data = dat, family = gaussian(), engine = "tmb")
fit_julia <- drmTMB(form, data = dat, family = gaussian(), engine = "julia")

summary(fit_julia)
coef(fit_julia)
predict(fit_julia, dpar = "mu")
predict(fit_julia, dpar = "sigma", type = "response")
logLik(fit_r)
logLik(fit_julia)
```

The `mu` coefficients describe the conditional mean. The `sigma` coefficients
are on the log standard-deviation scale: exponentiating a coefficient gives a
multiplicative change in residual standard deviation, not residual variance.
Compare the same estimator and requested outputs when checking the two fits.

## Model variation between groups

In a location-scale-scale model, predictors can enter the mean, residual
standard deviation, and group-level standard deviation. Here sex is constant
within each individual, and each individual has repeated observations:

```{r julia-group-lss, eval=FALSE}
set.seed(43)
individuals <- data.frame(
  individual = factor(seq_len(40)),
  sex = factor(rep(c("female", "male"), each = 20))
)
individuals$b <- rnorm(40, sd = exp(-0.3 + 0.25 * (individuals$sex == "male")))
personality_data <- individuals[rep(seq_len(40), each = 4), ]
personality_data$exploration_score <-
  1 + 0.5 * (personality_data$sex == "male") + personality_data$b +
  rnorm(nrow(personality_data), sd = 0.5)

formula_lss <- bf(
  exploration_score ~ sex + (1 | individual),
  sigma ~ sex,
  sd(individual) ~ sex
)
fit_lss <- drmTMB(
  formula_lss, family = gaussian(), data = personality_data,
  REML = TRUE, engine = "julia"
)
summary(fit_lss)
names(coef(fit_lss))
```

Use the returned coefficient-block names when selecting a block with `coef()`;
formula text such as `sd(individual)` need not be the stored block name.
Gaussian REML is available on supported routes. It accounts for estimating
specified fixed effects, but does not guarantee unbiased estimates in every
model or sample. Keep ML/REML choices explicit when comparing fits.

## Add a phylogeny

The Julia bridge accepts rooted polytomies with positive branch lengths.
Tip labels remain literal: spaces, apostrophes and Unicode do not need to be
renamed. The bridge quotes labels for transport and restores observation order.
Zero-length branches and unary nodes remain unsupported; this bridge uses
ultrametric trees on the correlation scale.

The same scale-regression idea applies to a phylogenetic group effect. This
small simulated example creates an ultrametric tree and matches data rows to
its tip labels explicitly:

```{r julia-phylo-lss, eval=FALSE}
install.packages("ape") # once, if not already installed
set.seed(44)
bird_tree <- ape::rcoal(40)
bird_data <- data.frame(
  species = bird_tree$tip.label,
  habitat = rnorm(40),
  latitude = runif(40, -1, 1)
)
A <- ape::vcv(bird_tree, corr = TRUE)[bird_data$species, bird_data$species]
sd_phylo <- exp(-0.3 + 0.2 * bird_data$habitat)
u <- sd_phylo * drop(t(chol(A)) %*% rnorm(40))
bird_data$log_body_mass <- 1 + 0.4 * bird_data$habitat + u +
  rnorm(40, sd = exp(-0.7 + 0.1 * bird_data$latitude))

formula_phylo_lss <- bf(
  log_body_mass ~ habitat + phylo(1 | species, tree = bird_tree),
  sigma ~ latitude,
  sd(species, level = "phylogenetic") ~ habitat
)
fit_phylo_lss <- drmTMB(
  formula_phylo_lss, family = gaussian(), data = bird_data,
  REML = TRUE, engine = "julia"
)
summary(fit_phylo_lss)
```

Here `habitat` predicts both the mean and the phylogenetic standard deviation;
`latitude` predicts residual standard deviation. The dense matrix above is only
for generating this small example, not a recipe for simulating very large trees.

## Missing responses

For supported Gaussian routes, `response = "include"` retains the full tree and
group design while conditioning the likelihood on observed responses. This is
not predictor imputation. Proper tree pruning can preserve covariances between
observed tips; dropping a response does not inherently change those covariances.

```{r julia-missing-response, eval=FALSE}
bird_data_with_missing <- bird_data
bird_data_with_missing$log_body_mass[c(3, 9)] <- NA_real_
fit_missing <- drmTMB(
  formula_phylo_lss, family = gaussian(), data = bird_data_with_missing,
  missing = miss_control(response = "include"),
  REML = TRUE, engine = "julia"
)
summary(fit_missing)
```

Check each prediction method's documented target and row behavior. Retaining
unobserved tips in the model does not by itself establish support for every
prediction target, such as `predict(..., dpar = "sd(species)")`.

## Inference and post-fit methods

The bridge has methods for `summary()`, `coef()`, `vcov()`, `logLik()`,
`predict()` and `confint()`. Availability depends on the fitted route and target;
a `drmTMB_julia` object does not automatically inherit every native-TMB diagnostic.

```{r julia-intervals, eval=FALSE}
vcov(fit_julia)
confint(fit_julia, method = "wald")
confint(fit_julia, method = "profile", parm = "fixef:mu:x")
```

With matching development versions of drmTMB and DRM.jl, transformed terms
retain their public formula names. For a model containing `I(x^2)`, select
`parm = "fixef:mu:I(x^2)"`; factor and interaction selectors likewise use the
names returned by `coef(fit_julia)`. Do not substitute generated Julia column
numbers. Supported fixed-effect predictions on new data retain the training
centering, scaling and polynomial basis. Older fitted objects without this
label metadata keep their original coefficient names.

These interval examples use the ordinary **ML** fit above. Check the documented
objective and target before comparing ML and REML inference. Agreement between
engines does not establish nominal interval coverage.

Inspect `conf.status` and `profile.message`, not just the limits. A
`profile_failed` result has an invalid endpoint; signed infinity can be a
failure placeholder, and a transformed SD lower limit of zero does not clear
that failure. A profile that did not cross the threshold within the searched
range is a separate diagnostic. Generic Julia profiles now report failed
nuisance solves, but successful optimizer termination does not by itself prove
a global optimum or a practical large-tree runtime.

### Which routes have bridge-side profile and bootstrap intervals

`r_bridge_status` in `inst/extdata/julia-capabilities.tsv` marks whether
`engine = "julia"` profile and bootstrap intervals have been checked against
`engine = "tmb"` on the same fitted target, not just whether the family fits.
Two routes carry `r_bridge_status` `"supported"` today:
`base_gaussian_location_scale` (the fixed-effect Gaussian location-scale
model) and `plain_binomial_nonphylo` (a `stats::binomial()` fit without
`phylo()`). For each, Wald and profile confidence intervals on a fixed-effect
mean coefficient agree with `engine = "tmb"` within `1e-4`, and both engines
converge. Bootstrap intervals (`R = 99`) overlap between engines, but this is
a weaker claim than profile or Wald agreement: `engine = "tmb"` and
`engine = "julia"` draw bootstrap replicates from independent random-number
streams, so the same `seed` does not reproduce the same replicates across
engines, and there is no same-seed design to compare against. Treat a
bootstrap-interval comparison as overlap evidence, not agreement within
tolerance, even where both engines report `0/99` failed replicates.

`biv_gaussian_residual` (the residual-only bivariate Gaussian route, fitting
`rho12` without a structured or phylogenetic covariance term) is expected to
join those two once drmTMB#1187 lands; as of this vignette it still carries
`r_bridge_status` `"partial"` because the bridge has no profile- or
bootstrap-ready target for that route on any parameter. `#1187` adds one.
Until it merges, use `engine = "tmb"` for residual bivariate Gaussian
intervals.

### REML standard errors are not directly comparable across engines

Where REML is admitted on the bridge (see `docs/design/261-reml-by-route.md`),
a fixed-effect mean-block Wald standard error is not expected to match
between engines by construction, not error: `engine = "tmb"`'s `sdreport()`
moves the mean coefficients into TMB's random-effect block under REML and
propagates variance-parameter uncertainty into their reported standard
errors, while DRM.jl profiles the mean coefficients out exactly at each
variance-parameter value and reports the canonical GLS covariance
`(Xmu' Vhat^-1 Xmu)^-1` with no such propagation (drmTMB#1201). Two more REML
routes are landing after this vignette: the mean-only phylogenetic Gaussian
cell (drmTMB#1199) and the residual-only bivariate Gaussian cell
(drmTMB#1197).

## One modelled missing predictor (development route)

The Julia engine also has a narrow, **development** route for one
modelled missing predictor. It is limited to a Gaussian identity-link response,
one bare additive `mi(x)` term, complete fixed-effect exogenous designs, and a
Gaussian or Bernoulli fixed-effect predictor model. Missing `x` values are
integrated in the joint likelihood; they are not filled before fitting.

```{r julia-missing-predictor, eval=FALSE}
set.seed(45)
n <- 80
joint_data <- data.frame(z = seq(-1, 1, length.out = n))
joint_data$x <- 0.2 + 0.7 * joint_data$z + rnorm(n, sd = 0.25)
joint_data$y <- 0.4 + 0.5 * joint_data$z + 0.8 * joint_data$x + rnorm(n, sd = 0.3)
joint_data$x[c(8, 21, 53)] <- NA_real_
joint_data$y[c(15, 53)] <- NA_real_

fit_joint <- drmTMB(
  bf(y ~ z + mi(x), sigma ~ 1), family = gaussian(), data = joint_data,
  impute = list(x = x ~ z),
  missing = miss_control(response = "include", predictor = "model"),
  engine = "julia"
)
coef(fit_joint)
imputed(fit_joint, rows = "all")
```

For a binary predictor, use
`impute = list(x = impute_model(x ~ z, family = binomial()))`. The R development
route also accepts `response = "drop"`; it drops missing-response rows during
preparation. That preprocessing differs from native-TMB behaviour, so it is not
a response-policy parity claim.

This route excludes other response families, additional or interacted `mi()`
terms, random or structured effects, offsets, weights, non-default controls,
and REML. `summary()` and Wald `confint()` require a usable returned covariance;
profile and bootstrap intervals are unsupported. For a Gaussian predictor, the
predictor-SD Wald interval is transformed to the natural SD scale, can cross
zero, and is not a native-interval-parity or coverage result. Two public bridge
adapter cases pass. Native training prediction and binary `newdata` handling
have been repaired and checked independently. Full numerical parity remains
open: small differences in optimizer stopping affect coefficients and predictions
beyond the declared tolerance. This development route makes no speed or
interval-coverage claim.

## Species counts, sparse solvers and parallel work

**There is no universal 5,000-species limit.** The limit belongs to particular
dense Gaussian location-scale-scale (LSS) routes, and is a limit on observations,
not a general Julia limit.

For a single phylogenetic LSS component, DRM.jl automatically chooses its sparse
tree engine once there are more than 500 species. In direct Julia use, you can
also request that route explicitly with `sparse = true` or
`algorithm = :sparse_lbfgs`. The sparse route avoids constructing the dense
species-by-species covariance matrix. A deliberately forced dense fallback, and
the current multi-component LSS route, still stop at 5,000 observations. Repeated
observations per species can therefore reach the dense limit before a dataset has
5,000 species.

Other sparse phylogenetic routes have different practical limits. Species count,
observation count, number of latent effects, and requested uncertainty all affect
cost.

A dense double-precision 5,000 by 5,000 matrix alone occupies about 200 MB, before
factorizations and temporary storage. Its storage grows quadratically and a
general dense factorization grows cubically. Tree-based augmented-state solvers
can avoid the dense covariance, but the precision is **tree-sparse**, not generally
tridiagonal. Crossed effects and other structures can introduce additional fill.

Sparse likelihood calculations, model construction, standard errors, profiles,
and bootstrap refits are different workloads. A fast likelihood evaluation is
not evidence that a complete fit or interval calculation takes the same time.
There is no single speedup factor for all Julia-backed models.

To make multiple Julia threads available, set `JULIA_NUM_THREADS` before Julia
starts (for example in the shell launching R). A multi-threaded runtime does not
make every fit parallel automatically. Use the documented options for the
specific workflow, and avoid combining many Julia workers with many BLAS threads.

## Current boundaries

The Julia backend is implemented, but it is not yet a replacement for every
native-TMB workflow. Check the [capability guide](capability-and-limits.html) for the
model you intend to fit. In particular:

- The one-modelled-predictor development route above is available through
  `engine = "julia"`; other missing-predictor models remain native-TMB workflows.
- Missing-response inclusion and REML depend on the model route. Most
  non-Gaussian bridge routes fit ML only, but large-p Poisson `phylo()` is a
  genuine exception: the bridge fits `REML = TRUE` there by Cox-Reid Laplace,
  a route native `engine = "tmb"` does not have. See the REML column in the
  table below and `docs/design/261-reml-by-route.md` before assuming ML-only.
- Accepted families do not imply that every combination of structured effects,
  weights, controls and post-fit methods is admitted. `phylo()`, `relmat()`,
  `animal()` and `spatial()` terms on the fixed-effect-only families below are
  already refused by existing gates before Julia starts. An ordinary
  random-effect term (`(1 | g)`, a random slope) or an `sd()`/`sd_phylo()`
  scale submodel on one of these families lands with drmTMB#1196: for
  `student()`, `lognormal()`, `truncated_nbinom2()`, `zero_one_beta()`,
  `tweedie()`, `beta_binomial()`, `cumulative_logit()`, and `skew_normal()`
  once #1176 lands, such a term is refused before Julia starts, naming the
  family and pointing to `engine = "tmb"`. Until #1196 merges, that
  combination is not refused by `drmTMB` before Julia starts; the call
  round-trips to DRM.jl and DRM.jl refuses it in its own words instead (for
  example, `TruncatedNegBinomial2() currently supports fixed effects only`,
  pinned in `tests/testthat/test-julia-family-truncated_nbinom2.R`). Treat any
  such combination as unsupported regardless of which side reports the
  refusal.
- `predict()` on new data currently has narrower Julia support than in-sample
  prediction; do not assume that every distributional parameter is available.

If the bridge rejects a combination, use the native engine for that analysis
and retain the warning or error when reporting the limitation.

## Family and route reference

Every family below has a same-target parity receipt against `engine = "tmb"`
on a committed fixture (coefficients and log-likelihood within 1e-4, Wald SEs
within 1e-3 relative), except the one row marked "not yet admitted". The
"Receipt" column names the capability-ledger row
(`inst/extdata/julia-capabilities.tsv`, `r_bridge_status` column) where one
exists; three fixed-effect-only families were admitted without a ledger row
yet, so their receipt lives only in the family's own test file.

| Family | `engine = "julia"` route | Receipt |
|---|---|---|
| `gaussian()` | fixed effect; mean-only `phylo()`; coupled mean+`sigma` phylo; ordinary random intercept/slope on `mu` and `sigma`; `relmat()`/`spatial()`/`animal()` | `base_gaussian_location_scale`, `gaussian_phylo_mean`, `gaussian_random_intercept_mu`, `gaussian_random_slope_mu`, `gaussian_sigma_random_intercept`, `general_covariance_structured`, `location_scale_scale` (partial/experimental; see the TSV for which) |
| `biv_gaussian()` | fixed effect (residual `rho12`); coupled q = 4 phylo | `biv_gaussian_residual`, `biv_q4_phylo_reml` |
| `student()` | fixed effect only | `fe_student` |
| `lognormal()` | fixed effect only | `fe_lognormal` |
| `poisson()` | fixed effect; large-p `phylo()`; `zi` dpar on the fixed-effect route | `fe_poisson`, `phylo_count_large_p`, `zi_poisson` |
| `nbinom2()` | fixed effect; large-p `phylo()`; coupled mean+`sigma` phylo; `zi` dpar; `hu` dpar (bridge spelling only -- see note below) | `fe_nbinom2`, `phylo_count_large_p`, `zi_nbinom2`, `hurdle_nbinom2` |
| `gamma()` | fixed effect; large-p `phylo()`; coupled mean+`sigma` phylo | `fe_gamma`, `phylo_gamma_beta_binomial` |
| `beta()` | fixed effect; large-p `phylo()`; coupled mean+`sigma` phylo | `fe_beta`, `phylo_gamma_beta_binomial` |
| `stats::binomial(link = "logit")` | fixed effect; large-p mean-only `phylo()` | `plain_binomial_nonphylo`, `phylo_gamma_beta_binomial` |
| `truncated_nbinom2()` | fixed effect only | no ledger row yet; `tests/testthat/test-julia-family-truncated_nbinom2.R` |
| `zero_one_beta()` | fixed effect only | no ledger row yet; `tests/testthat/test-julia-family-zero_one_beta.R` |
| `tweedie()` | fixed effect only | no ledger row yet; `tests/testthat/test-julia-family-tweedie.R` |
| `beta_binomial()` | fixed effect only (`phylo()` refuses: DRM.jl's phylo `BetaBinomial` route is constant-`sigma` only, no receipt) | `fe_beta_binomial` |
| `cumulative_logit()` | fixed effect only | no ledger row yet; `tests/testthat/test-julia-family-cumulative_logit.R` |
| `skew_normal()` | not yet admitted (open PR) | none |
| any other family (`zi_poisson()`/`zi_nbinom2()`/`hurdle_nbinom2()` as their own family constructors, mixture families, etc.) | refused | none |

The `zi` and `hu` dpars are not separate families: they are formula terms
(`bf(y ~ x, zi ~ x)` or `bf(y ~ x, hu ~ x)`) added to `poisson()` or
`nbinom2()`, and the bridge admits them through those families' fixed-effect
route rather than through a family-specific ledger row. `hu` has a spelling
mismatch between engines today: native `engine = "tmb"` fits the hurdle model
as `truncated_nbinom2()` with `hu ~ ...`, while `engine = "julia"` fits it as
`nbinom2()` with `hu ~ ...` (DRM.jl reads `hu` on `nbinom2()` as the hurdle
model). The same call does not fit on both engines yet; see `hurdle_nbinom2`
in the capability table for the measured cross-spelling receipt and the
tracked defect.
### Julia optimizer controls

For the base Julia bridge, use the Julia-native controls through the existing
`drm_control()` interface:

```r
control = drm_control(optimizer = list(g_tol = 1e-6, algorithm = "lbfgs"))
```

These settings control DRM.jl's gradient tolerance and solver choice. They do
not make its optimizer equivalent to TMB's `nlminb()` presets or iteration
budgets. TMB-only controls, including `iter.max`, `eval.max`, storage flags and
`optimizer_preset`, are rejected before Julia starts; use `engine = "tmb"` for
those controls.
