---
title: "Running models with the Julia engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running models with the Julia engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article is for R users, contributors, and early testers who want to run a
`drmTMB` formula through the experimental DRM.jl engine. The purpose is not to
replace the default TMB backend. The purpose is to make the R-to-Julia boundary
small, testable, and honest before wider family and post-fit parity work begins.
It assumes a first default-engine fit already succeeded; this article is about
adding the Julia bridge as a second engine, not about fitting a first model.

The current bridge is a first slice. It supports Gaussian one-response and
two-response models that use fixed-effect distributional formulas. It also
marshals one Gaussian phylogenetic mean model,
`phylo(1 | species, tree = tree)` with `sigma ~ 1`, through DRM.jl's default
all-node sparse route. That phylogenetic cell now has narrow
`confint()` support for the residual phylogenetic SD in the `mu` model with
`method = "profile"` or `method = "bootstrap"`.

It now also marshals the Gaussian **σ-phylo location-scale** cell — a `phylo()`
random effect on *both* `mu` and `sigma` — as a **separate / uncorrelated** block,
so the mean-scale phylogenetic correlation is not forced into a single block (the
artefact that pins it at ±1 when a scale-axis variance collapses). It accepts
`REML = TRUE` (restricted maximum likelihood, better conditioned near the variance
boundary) when the installed DRM.jl build supports that REML cell. The formula is
`sigma ~ phylo(1 | species, tree = tree)` alongside
`mu ~ ... + phylo(1 | species, tree = tree)`. The R-side profile/bootstrap
inventory is narrower than the fitted-model route: univariate profile/bootstrap
is currently exposed for the `mu` phylogenetic SD target, while the bivariate
q = 4 route exposes one among-axis SD target for each of `mu1`, `mu2`,
`sigma1`, and `sigma2`.

It additionally routes the bivariate **q = 4 phylogenetic location-scale model**
(the PLSM): intercept `phylo(1 | p | species)` on all four axes (`mu1`, `mu2`,
`sigma1`, `sigma2`), sharing one tree and grouping factor, fit through DRM.jl's
exact-gradient sparse q = 4 engine. The bridge translates drmTMB's labelled
covariance-block grammar `phylo(1 | p | species)` to DRM.jl's `phylo(1 | species)`
(DRM.jl implies the 4x4 between-species covariance from the four shared axes).
Round-trip parity against `engine = "tmb"` matches the log-likelihood and fixed
effects on a seeded q = 4 fixture (verified locally; CI does not run Julia).
Use the top-level `REML = TRUE` / `REML = FALSE` argument, as in native
mixed-model workflows; the bridge converts that public switch to DRM.jl's
internal `method = :REML` only for supported REML cells.

Beyond Gaussian, the bridge also routes a set of non-Gaussian **phylogenetic**
cells: a `phylo(1 | group)` random intercept on Poisson, NB2, Gamma, Beta, or
Binomial responses, plus `relmat()`, `animal()`, and `spatial()` structured
routes for Gaussian, Poisson, NB2, and Gamma. These are net-new capabilities
with no native TMB equivalent, so treat them at recovery grade — the parity
tests check a finite-and-sane floor, not full coefficient parity — and keep
`engine = "tmb"` when you need the validated native surface.

The bridge does not yet support weights,
imputation, non-default control settings, most missing-data routes,
general fixed-effect non-Gaussian models, random-effect scale formulas beyond
the σ-phylo location-scale block, `corpair()` formula entries, simulation, or
persistent Julia fit handles. `missing = miss_control(response = "include")` is
admitted only for Gaussian and bivariate Gaussian observed-response masks.

For ordinary analyses, keep using the default native engine:

```{r native-engine, eval = FALSE}
fit <- drmTMB(
  bf(y ~ x, sigma ~ x),
  family = gaussian(),
  data = dat
)
```

Use `engine = "julia"` only when you are deliberately testing the bridge:

```{r julia-engine-shape, eval = FALSE}
fit_julia <- drmTMB(
  bf(y ~ x, sigma ~ x),
  family = gaussian(),
  data = dat,
  engine = "julia"
)
```

The bivariate q = 4 phylogenetic location-scale route uses matching
`phylo(1 | p | species, tree = tree)` terms on all four axes:

```{r julia-engine-q4, eval = FALSE}
fit_q4 <- drmTMB(
  bf(
    mu1 = y1 ~ x + phylo(1 | p | species, tree = tree),
    mu2 = y2 ~ x + phylo(1 | p | species, tree = tree),
    sigma1 = ~ 1 + phylo(1 | p | species, tree = tree),
    sigma2 = ~ 1 + phylo(1 | p | species, tree = tree),
    rho12 = ~ 1
  ),
  family = biv_gaussian(),
  data = dat,
  engine = "julia"
)
```

## Set up the bridge

The R package does not install Julia, JuliaCall, or DRM.jl for you. The bridge
expects three pieces:

- R can load `JuliaCall`;
- Julia can load DRM.jl;
- `drmTMB` knows which DRM.jl environment to activate.

Install the R-side optional package with:

```{r install-juliacall, eval = FALSE}
install.packages("JuliaCall")
```

Then choose one DRM.jl discovery route. During local development, the most
explicit route is an R option:

```{r set-drmjl-option, eval = FALSE}
options(drmTMB.DRM.jl.path = "/path/to/DRM.jl")
```

For shell-driven checks, use an environment variable:

```sh
export DRM_JL_PATH="/path/to/DRM.jl"
```

If neither is set, `drmTMB` looks for a sibling checkout at `../DRM.jl` from the
current working directory. That default is useful when the two repositories sit
next to each other under the same parent folder.

## Fit the first supported model

For one response, the Gaussian bridge keeps the same public parameter contract
as native `drmTMB`:

```text
y_i ~ Normal(mu_i, sigma_i^2)
mu_i = X_mu[i, ] beta_mu
log(sigma_i) = X_sigma[i, ] beta_sigma
```

The matching R syntax is:

```{r gaussian-one-response, eval = FALSE}
set.seed(20260608)
n <- 120
dat <- data.frame(
  temperature = runif(n, -1.5, 1.5),
  canopy_open = rbinom(n, 1, 0.5)
)
mu <- 8 + 1.2 * dat$temperature + 0.7 * dat$canopy_open
sigma <- exp(-0.4 + 0.35 * dat$canopy_open)
dat$growth <- rnorm(n, mean = mu, sd = sigma)

fit_julia <- drmTMB(
  bf(growth ~ temperature + canopy_open, sigma ~ canopy_open),
  family = gaussian(),
  data = dat,
  engine = "julia"
)

coef(fit_julia)
vcov(fit_julia)
logLik(fit_julia)
sigma(fit_julia)
is_converged(fit_julia)
```

This example asks whether warmer plots have higher expected plant growth and
whether open-canopy plots are less predictable after modelling the mean.
`canopy_open` is an explicit 0/1 indicator in this bridge example so the R and
Julia formula engines receive the same model matrix. Read the `mu`
coefficients as expected-growth effects and the `sigma` coefficients as log
residual-SD effects. The fitted object has class `drmTMB_julia`, so the bridge
can grow without pretending it is already a full native `drmTMB` fit.

## Compare native TMB and Julia-engine output

For bridge development, fit the same model with both engines and compare the
returned quantities directly:

```{r compare-engines-code, eval = FALSE}
fit_tmb <- drmTMB(
  bf(growth ~ temperature + canopy_open, sigma ~ canopy_open),
  family = gaussian(),
  data = dat
)

fit_julia <- drmTMB(
  bf(growth ~ temperature + canopy_open, sigma ~ canopy_open),
  family = gaussian(),
  data = dat,
  engine = "julia"
)

coef(fit_tmb)
coef(fit_julia)
logLik(fit_tmb)
logLik(fit_julia)
```

On a local run of this article's plant-growth data, the two engines returned
the same fitted surface to numerical tolerance:

```text
Gaussian plant growth coefficients
term                    tmb        julia        diff
mu_(Intercept)          7.974736   7.974737    1.06e-06
mu_temperature          1.223783   1.223783    7.05e-07
mu_canopy_open          0.440543   0.440541   -1.67e-06
sigma_(Intercept)      -0.332546  -0.332546    6.45e-08
sigma_canopy_open       0.358766   0.358768    2.57e-06

fit summaries
quantity        tmb        julia       diff
logLik       -153.3282  -153.3282   5.77e-10
nobs          120        120        0
df              5          5        0

max_abs_coef_diff   = 2.57e-06
max_abs_fitted_diff = 2.09e-06
max_abs_sigma_diff  = 2.70e-06
```

Using the two-response plant-trait code in the next section, the same check
works for the bivariate Gaussian bridge:

```text
Bivariate plant traits coefficients
term                    tmb        julia        diff
mu1_(Intercept)        14.85555   14.85555     0
mu1_soil_moisture       2.04480    2.04480     0
mu2_(Intercept)         5.85979    5.85979     0
mu2_soil_moisture       1.28606    1.28606     0
sigma1_(Intercept)      0.20735    0.20734    -3.31e-06
sigma2_(Intercept)     -0.39242   -0.39242    -3.31e-06
rho12_(Intercept)       0.39271    0.39270    -2.17e-06

fit summaries
quantity        tmb        julia       diff
logLik       -257.7595  -257.7595   1.93e-09
nobs          100        100        0
df              7          7        0

max_abs_coef_diff     = 3.31e-06
max_abs_mu1_fit_diff  = 0
max_abs_mu2_fit_diff  = 0
max_abs_sigma1_diff   = 4.07e-06
max_abs_sigma2_diff   = 2.24e-06
max_abs_rho12_diff    = 1.87e-06
```

Those numbers are a bridge check, not a broad statistical validation claim. The
test says this particular Gaussian formula and data route arrives at the same
answer through the native TMB engine and the Julia engine.

## Fit a two-response Gaussian model

For two responses, the bridge uses the same residual-correlation vocabulary as
native `drmTMB`. The model is:

```text
[y1_i, y2_i]' ~ Normal([mu1_i, mu2_i]', Omega_i)
Omega_i =
  [ sigma1_i^2                    rho12_i sigma1_i sigma2_i ]
  [ rho12_i sigma1_i sigma2_i     sigma2_i^2                ]
rho12_i = tanh(eta_rho12_i)
```

Use separate response formulas when the responses have separate location
models. The code below simulates a small two-trait data set where soil moisture
changes both trait means and the two residual traits remain positively
correlated:

```{r gaussian-two-response, eval = FALSE}
set.seed(20260609)
n2 <- 100
plant_traits <- data.frame(soil_moisture = runif(n2, -1, 1))
mu1 <- 15 + 2.0 * plant_traits$soil_moisture
mu2 <- 6 + 1.1 * plant_traits$soil_moisture
sigma1 <- 1.1
sigma2 <- 0.7
rho12_true <- 0.35
z1 <- rnorm(n2)
z2 <- rho12_true * z1 + sqrt(1 - rho12_true^2) * rnorm(n2)
plant_traits$leaf_area <- mu1 + sigma1 * z1
plant_traits$root_mass <- mu2 + sigma2 * z2

bform <- bf(
  mu1 = leaf_area ~ soil_moisture,
  mu2 = root_mass ~ soil_moisture,
  sigma1 = ~ 1,
  sigma2 = ~ 1,
  rho12 = ~ 1
)

fit_biv_tmb <- drmTMB(
  bform,
  family = biv_gaussian(),
  data = plant_traits
)

fit_biv_julia <- drmTMB(
  bform,
  family = biv_gaussian(),
  data = plant_traits,
  engine = "julia"
)

coef(fit_biv_tmb)
coef(fit_biv_julia)
rho12(fit_biv_tmb)
rho12(fit_biv_julia)
```

The Julia-engine-only form is shorter once the check is no longer comparing
backends:

```{r gaussian-two-response-julia-only, eval = FALSE}
fit_biv_julia <- drmTMB(
  bf(
    mu1 = leaf_area ~ soil_moisture,
    mu2 = root_mass ~ soil_moisture,
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ 1
  ),
  family = biv_gaussian(),
  data = plant_traits,
  engine = "julia"
)

coef(fit_biv_julia)
fitted(fit_biv_julia)
sigma(fit_biv_julia)
rho12(fit_biv_julia)
```

Here `rho12` is the residual correlation after the two means and residual
scales are modelled. It is not a group-level, phylogenetic, or spatial
correlation. In this plant-trait example, a positive `rho12` would mean that
individuals with larger leaf area than expected for their soil-moisture value
also tend to have larger root mass than expected for that same covariate.

## What the bridge returns

After fitting, these familiar R methods work on the lightweight
`drmTMB_julia` object. The R side calls `DRM.drm_bridge()` through JuliaCall,
DRM.jl returns plain pieces, and `drmTMB` reconstructs coefficient values,
likelihood summaries, fitted values, residuals, scale summaries, and residual
correlations when present.

| R method | Current bridge behaviour |
| --- | --- |
| `coef()` and `fixef()` | fixed-effect coefficients split by distributional parameter |
| `vcov()` | fixed-effect covariance matrix when DRM.jl returns one; partially available routes keep finite blocks and `NaN` for unavailable blocks |
| `fit$sdpars` | structured-effect scale summaries that are not fixed effects, such as the Julia phylogenetic SD |
| `logLik()`, `nobs()`, `df.residual()`, `deviance()` | likelihood and dimension summaries |
| `fitted()` and `residuals()` | stored in-sample response-scale summaries |
| `sigma()` | stored fitted scale summaries |
| `corpairs()` and `rho12()` | residual-correlation payload for two-response Gaussian fits |
| `predict()` | stored in-sample `mu`, `sigma`, or `rho12`; no `newdata` yet |
| `is_converged()` | convergence flag returned by DRM.jl |

This is enough for bridge smoke tests and simple fit inspection. It is not yet
enough for the full native post-fit surface. In particular, the default Julia
Gaussian phylogenetic route currently returns point estimates, fitted values,
and a mean fixed-effect covariance block. Scale and variance-component
covariance entries remain unavailable in this first slice, so `vcov()` contains
`NaN` in those blocks and the fitted object records uncertainty status as
`partial`. For the Gaussian phylogenetic SD target, use the narrow
`confint()` routes described below: profile intervals are now admitted for that
target, and bootstrap intervals are available when repeated simulation refits
are the right uncertainty check.

## What works now

For applied users, the bridge should feel like a narrow extension of
`drmTMB`: the same formula, the same public parameter scales, faster fitting
for admitted large-tree models, and early errors for unsupported neighbouring
models. DRM.jl can have more native fitting routes than `drmTMB` exposes, but
`engine = "julia"` should choose a default by distribution family and model
structure, then admit only cells with R-scale parity tests.

| Model cell | Julia route | R bridge status |
| --- | --- | --- |
| Fixed-effect Gaussian location-scale | direct Gaussian ML route | admitted |
| Fixed-effect bivariate Gaussian | bivariate Gaussian ML route with `rho12` | admitted |
| Gaussian `phylo(1 | species)` on `mu`, `sigma ~ 1` | sparse all-node Gaussian route | admitted for point estimates, fitted values, mean fixed-effect covariance, and profile/bootstrap intervals for the phylogenetic SD target |
| Gaussian σ-phylo location-scale: `phylo()` on `mu` AND `sigma` (separate / uncorrelated block), optional `REML = TRUE` | sparse location-scale Gaussian route | admitted for fitting and REML forwarding when the DRM.jl build supports the cell; univariate profile/bootstrap extraction still exposes the `mu` phylogenetic SD target, not a separate `sigma`-axis target |
| Bivariate q = 4 phylogenetic location-scale: labelled `phylo()` on `mu1`, `mu2`, `sigma1`, and `sigma2`, optional `REML = TRUE` | exact-gradient sparse q = 4 route | admitted; `REML = TRUE` is forwarded when the DRM.jl build supports the q = 4 REML cell, with profile/bootstrap diagnostics for the four among-axis SDs |
| Gaussian phylogenetic slopes or multiple distinct structured random-effect blocks | sparse all-node likelihood route | blocked in the R bridge |
| Non-Gaussian fixed-effect models | family-specific ML route after coefficient-scale parity tests | blocked in the R bridge |
| Non-Gaussian phylogenetic models: `phylo(1 | group)` intercept on Poisson, NB2, Gamma, Beta, or Binomial (plus `relmat()`, `animal()`, `spatial()` structured routes for Gaussian, Poisson, NB2, Gamma) | sparse all-node Laplace / general-covariance route | admitted by the R gate; net-new capability with no native TMB equivalent, so parity coverage is narrower than Gaussian (finite-and-sane floor, not coefficient parity) |

That split matters. Julia can keep several internal algorithms, but the R
bridge should not ask applied users to choose among them before the parity
contract is settled. The public story is simpler: the R bridge chooses the
current DRM.jl default for each admitted model cell, reports the result on
`drmTMB` scales, and rejects neighbouring syntax until it has tests.

Once one of these admitted cells runs, see
[Cross-family bivariate models with the Julia engine](cross-family.html) for
the mixed-family route the bridge also supports: two responses from different
distribution families sharing one latent-scale correlation.

## Plot Julia-engine fits with ggplot2

The practical benefit of the bridge is that the fitted values come back to R.
Even when DRM.jl does the fitting work, the plot can still use ordinary
`ggplot2`, the original R data frame, and familiar `drmTMB` extractors.

```{r ggplot-julia-fit, eval = FALSE}
library(ggplot2)

plot_dat <- data.frame(
  dat,
  mu_hat = as.numeric(fitted(fit_julia)),
  sigma_hat = as.numeric(sigma(fit_julia))
)
plot_dat <- plot_dat[order(plot_dat$canopy_open, plot_dat$temperature), ]

ggplot(plot_dat, aes(temperature, growth, colour = factor(canopy_open))) +
  geom_point(alpha = 0.55) +
  geom_line(aes(y = mu_hat), linewidth = 1) +
  labs(
    x = "Temperature anomaly",
    y = "Plant growth",
    colour = "Open canopy",
    title = "Mean growth fitted by DRM.jl, plotted in R"
  )

ggplot(plot_dat, aes(factor(canopy_open), sigma_hat, fill = factor(canopy_open))) +
  geom_boxplot(width = 0.5, alpha = 0.7) +
  labs(
    x = "Open canopy",
    y = "Fitted residual SD",
    fill = "Open canopy",
    title = "Residual predictability from a Julia-engine fit"
  )
```

This is the practical value of the bridge for applied work: Julia can do the
fitting, while the analysis, plotting, and reporting stay in the R workflow.
Julia also has strong graphics packages, including Makie, AlgebraOfGraphics,
and Plots, but the first R-bridge plotting story should stay simple: bring the
Julia-engine results back to R and keep using `ggplot2`.

## Time the two engines

Speed comparisons need to separate Julia startup from model-fitting time. The
first JuliaCall fit in a fresh R session pays setup costs for Julia and the
DRM.jl environment. After that setup, a warm-session `system.time()` check can
compare repeated fits of the same model.

They should also record the parallel backend, not just the number of cores. In
`drmTMB`, `parallel = "multicore"` means process-based parallelism through
`parallel::mclapply()`, which forks multiple R worker processes on Unix-like
systems. In DRM.jl, `threads = true` means within-process shared-memory
parallelism using the Julia threads available to the Julia process. R can still
use lower-level threading through BLAS/LAPACK, OpenMP, and TMB templates, so
benchmark metadata should record BLAS and OpenMP thread settings as well as the
high-level profile or bootstrap backend.

BLAS is the low-level linear-algebra layer used for dense matrix operations
such as matrix multiplication, Cholesky factorization, and least-squares
solves. Depending on the installed BLAS library, those operations can use their
own threads. OpenMP is a compiled-code threading system used by C, C++, and
Fortran libraries; a TMB template can use OpenMP if the template and compiler
were built for it. These threads are different from the outer profile or
bootstrap workers. If a benchmark starts 20 outer Julia workers and BLAS also
starts 20 inner workers, the machine may try to schedule far more runnable
threads than it has cores. That is oversubscription, and it can hide the real
engine comparison.

| label | meaning | process model |
|---|---|---|
| R `none` | one R process, no explicit profile/bootstrap parallel backend | single process |
| R `multicore` | `parallel::mclapply()` profile/bootstrap workers | forked R processes |
| R TMB OpenMP | threaded C++ code inside a TMB template | one R process calling OpenMP code |
| Julia `threads = false` | one Julia process, one Julia thread | single process |
| Julia `threads = true` | one Julia process, N Julia threads | shared-memory threads |

For fair comparisons, keep hidden low-level threads explicit. A good timing row
records the language, backend, worker or thread count, process model, BLAS
threads, OpenMP threads, platform, and whether the row is cold-start or warm.

```{r timing-code-helpers, eval = FALSE}
make_gaussian_data <- function(n, seed = 20260608L) {
  set.seed(seed + n)
  dat <- data.frame(
    temperature = runif(n, -1.5, 1.5),
    canopy_open = rbinom(n, 1, 0.5)
  )
  mu <- 8 + 1.2 * dat$temperature + 0.7 * dat$canopy_open
  sigma <- exp(-0.4 + 0.35 * dat$canopy_open)
  dat$growth <- rnorm(n, mean = mu, sd = sigma)
  dat
}

time_reps <- function(expr, reps = 3L) {
  elapsed <- numeric(reps)
  for (i in seq_len(reps)) {
    gc()
    elapsed[[i]] <- system.time(force(expr()))[["elapsed"]]
  }
  elapsed
}
```

For the supported fixed-effect Gaussian bridge, time each data size after both
engines have been warmed:

```{r timing-code-gaussian, eval = FALSE}
bench_gaussian_bridge <- function(n, reps = 3L) {
  dat <- make_gaussian_data(n)
  form <- bf(growth ~ temperature + canopy_open, sigma ~ canopy_open)

  invisible(drmTMB(form, family = gaussian(), data = dat))
  invisible(drmTMB(form, family = gaussian(), data = dat, engine = "julia"))

  tmb_time <- time_reps(function() {
    drmTMB(form, family = gaussian(), data = dat)
  }, reps)
  julia_time <- time_reps(function() {
    drmTMB(form, family = gaussian(), data = dat, engine = "julia")
  }, reps)

  data.frame(
    n = n,
    tmb_median_s = median(tmb_time),
    julia_median_s = median(julia_time),
    tmb_min_s = min(tmb_time),
    julia_min_s = min(julia_time),
    reps = reps,
    speedup = median(tmb_time) / median(julia_time)
  )
}

do.call(rbind, lapply(c(100L, 1000L, 10000L), bench_gaussian_bridge))
```

On the same local checkout, `tools/benchmark-julia-engines.R` gave the following
fixed-effect Gaussian bridge rows from R:

```text
n      tmb_first_s   julia_first_s   tmb_warm_s   julia_warm_s   reps   speedup   logLik_diff
100    0.123         15.886          0.016        0.002          3      8.00      1.41e-09
1000   0.034          0.053          0.036        0.003          3     12.00      7.21e-09
10000  0.245          0.024          0.240        0.024          3     10.00      1.87e-08
```

Here `speedup` is the median native TMB warm time divided by the median
Julia-engine warm time. `julia_first_s` for the first row includes JuliaCall
startup; later first-call columns are first calls in the same R process after
clearing the R-side bridge state. These rows are local bridge evidence, not a
profile/bootstrap threading result. The benchmark script records package
versions, BLAS/OpenMP thread settings, convergence flags, and parity checks
beside the timing rows.

For a phylogenetic model, the important pressure axis is the number of species
or tree tips, not only the number of observation rows. The Julia bridge now
marshals a single Gaussian phylogenetic mean model:

```{r phylo-tmb-benchmark-shape, eval = FALSE}
fit_phylo_tmb <- drmTMB(
  bf(growth ~ temperature + phylo(1 | species, tree = tree), sigma ~ 1),
  family = gaussian(),
  data = dat_phylo
)
```

The Julia comparison uses the same formula and data:

```{r phylo-timing-target, eval = FALSE}
fit_phylo_julia <- drmTMB(
  bf(growth ~ temperature + phylo(1 | species, tree = tree), sigma ~ 1),
  family = gaussian(),
  data = dat_phylo,
  engine = "julia"
)
```

Here the scientific question is whether broad differences in avian body mass
are associated with wing morphology and beak length after accounting for shared
ancestry. The term `phylo(1 | species, tree = tree)` adds a phylogenetic
random intercept: species that are close on the Hackett tree are allowed to
have more similar log body masses than unrelated species. The phylogenetic SD
is the residual ancestry-linked variation in the location model on the response
scale.

The bridge keeps fitted values aligned with the original R data and returns the
phylogenetic SD on the same response-scale convention as native `drmTMB`. The
lower-level tree serialization and sparse all-node representation are bridge
details; the applied check is whether the same formula gives comparable
likelihoods, fitted values, and phylogenetic SD estimates.

This route returns a finite mean-coefficient covariance block. Residual-scale
and phylogenetic-SD Wald covariance entries remain unavailable in this first
slice, so the fitted object reports `uncertainty$status = "partial"` rather
than pretending full Wald inference is available. Use profile likelihood or
bootstrap for the admitted phylogenetic SD interval.

For the article benchmark, use the real AVONET trait table and Hackett bird
phylogeny from a local `pigauto` or `BACE` checkout when those sibling
repositories are available. The full AVONET/Hackett pair has 9,993 complete
species for this large comparative morphology example, so the large row is
labelled `9993` rather than rounded up to `10000`.

<details>
<summary>Show the benchmark helper code used to generate the table</summary>

```{r timing-code-avonet-phylo, eval = FALSE}
find_avonet_files <- function() {
  candidates <- list(
    list(
      data = file.path("..", "pigauto", "avonet", "AVONET3_BirdTree.csv"),
      tree = file.path("..", "pigauto", "avonet", "Stage2_Hackett_MCC_no_neg.tre")
    ),
    list(
      data = file.path("..", "BACE", "dev", "testing_data", "AVONET.csv"),
      tree = file.path("..", "BACE", "dev", "testing_data", "Hackett_tree.tre")
    )
  )
  for (x in candidates) {
    if (file.exists(x$data) && file.exists(x$tree)) {
      return(x)
    }
  }
  stop("AVONET and Hackett tree files were not found in sibling checkouts.")
}

make_positive_branch_lengths <- function(tree, eps = 1e-8) {
  tree$edge.length <- pmax(tree$edge.length, eps)
  tree
}

make_avonet_phylo_data <- function(n_species = NULL) {
  paths <- find_avonet_files()
  avonet <- read.csv(paths$data, check.names = FALSE)
  tree0 <- ape::read.tree(paths$tree)

  dat0 <- data.frame(
    species = gsub(" ", "_", avonet$Species3, fixed = TRUE),
    mass = avonet$Mass,
    hand_wing = avonet[["Hand-Wing.Index"]],
    beak = avonet[["Beak.Length_Culmen"]]
  )
  dat0 <- dat0[complete.cases(dat0), , drop = FALSE]
  available <- tree0$tip.label[tree0$tip.label %in% dat0$species]

  if (!is.null(n_species) && n_species < length(available)) {
    available <- available[seq_len(n_species)]
  }

  tree <- make_positive_branch_lengths(ape::keep.tip(tree0, available))
  dat <- dat0[match(tree$tip.label, dat0$species), , drop = FALSE]
  dat$species <- factor(dat$species, levels = tree$tip.label)
  dat$log_mass <- log(dat$mass)
  dat$hand_wing_z <- as.numeric(scale(dat$hand_wing))
  dat$beak_z <- as.numeric(scale(dat$beak))

  list(data = dat, tree = tree)
}

phylo_sd_value <- function(fit) {
  vals <- unlist(fit$sdpars$mu, use.names = FALSE)
  if (length(vals) == 0L) {
    return(NA_real_)
  }
  as.numeric(vals[[1L]])
}

convergence_code <- function(fit) {
  if (inherits(fit, "drmTMB_julia")) {
    return(as.integer(!is_converged(fit)))
  }
  fit$fit$convergence
}

bench_phylo_bridge <- function(n, reps = 1L) {
  av <- make_avonet_phylo_data(n_species = n)
  tree <- av$tree
  form <- bf(
    log_mass ~ hand_wing_z + beak_z + phylo(1 | species, tree = tree),
    sigma ~ 1
  )

  invisible(drmTMB(form, family = gaussian(), data = av$data))
  invisible(drmTMB(form, family = gaussian(), data = av$data, engine = "julia"))

  tmb_time <- time_reps(function() {
    drmTMB(form, family = gaussian(), data = av$data)
  }, reps)
  julia_time <- time_reps(function() {
    drmTMB(form, family = gaussian(), data = av$data, engine = "julia")
  }, reps)

  fit_tmb <- drmTMB(form, family = gaussian(), data = av$data)
  fit_julia <- drmTMB(form, family = gaussian(), data = av$data, engine = "julia")
  coef_tmb <- unlist(coef(fit_tmb), use.names = TRUE)
  coef_julia <- unlist(coef(fit_julia), use.names = TRUE)

  data.frame(
    model = "AVONET phylogenetic Gaussian",
    n = n,
    species = nrow(av$data),
    rows = nrow(av$data),
    tmb_s = median(tmb_time),
    julia_s = median(julia_time),
    reps = reps,
    speedup = median(tmb_time) / median(julia_time),
    logLik_diff = abs(as.numeric(logLik(fit_tmb)) - as.numeric(logLik(fit_julia))),
    max_coef_diff = max(abs(coef_tmb[names(coef_julia)] - coef_julia), na.rm = TRUE),
    max_fitted_diff = max(abs(fitted(fit_tmb) - fitted(fit_julia)), na.rm = TRUE),
    sd_phylo_diff = abs(phylo_sd_value(fit_tmb) - phylo_sd_value(fit_julia)),
    tmb_convergence = convergence_code(fit_tmb),
    julia_convergence = convergence_code(fit_julia),
    julia_uncertainty = fit_julia$uncertainty$status,
    julia_finite_dpars = paste(fit_julia$uncertainty$finite_dpars, collapse = ";")
  )
}

do.call(rbind, lapply(c(100L, 1000L, 9993L), bench_phylo_bridge))
```

</details>

The same script also measured the AVONET/Hackett bridge after the current
Gaussian phylogenetic default and partial-covariance patch. Regenerate this
comparison with
`DRMTMB_ENGINE_BENCH_MODE=phylo Rscript --vanilla tools/benchmark-julia-engines.R`
when refreshing the article numbers:

| species | TMB warm s | Julia warm s | TMB / Julia | logLik diff | phylo SD diff | note |
|---:|---:|---:|---:|---:|---:|---|
| 100 | 0.147 | 0.057 | 2.58 | 5.07e-06 | 1.10e-05 | small-data warning row |
| 1,000 | 2.667 | 0.257 | 10.38 | 1.01e-08 | 1.88e-07 | clean bridge smoke row |
| 9,993 | 61.113 | 3.784 | 16.15 | 1.16e-03 | 1.91e-04 | post-cache warm bridge row; native TMB convergence code 1 |

The 100-species row is still a small-data warning row: the likelihood and
fitted values agree, but one coefficient differs because the variance split is
weakly identified at that size. The 1,000-species row is the first clean bridge
smoke row: both engines report convergence and the likelihoods agree closely.
The 9,993-species row is the important scaling row: after caching the repeated
same-tree bridge payload, the Julia bridge is about 16.1 times faster in this
local median warm R-side run, while native TMB returned convergence code 1
under the default optimizer budget. In the same overhead-reduction run, the
direct DRM.jl sparse kernel took `2.623` seconds; that direct Julia number is
useful contributor context, but the public article table stays at the R bridge
surface.

The native R row is already using `drmTMB`'s sparse phylogenetic precision, so
the headline is not "dense R versus sparse Julia." The headline is that the
Julia engine can fit the same R formula through the bridge faster on this
large real tree, while keeping fitted values, likelihood, and phylogenetic SD
on comparable scales. Algorithm comparisons between Julia routes are useful
for contributors, but they are not the main reader-facing story for this
article.

Contributor benchmark reports in DRM.jl keep the lower-level algorithm and
thread-scaling details. This R article keeps the main comparison at the bridge
surface: what a `drmTMB` user can fit, which intervals are admitted, and how
the R and Julia-engine results compare on the same public scale.

## Intentional errors

The first bridge slice fails early when a request would look supported but is
not yet parity-tested. These errors protect users from silent scale drift or
missing feature parity.

| Request | Why it errors now | What to use |
| --- | --- | --- |
| JuliaCall or DRM.jl is not found | the bridge is optional and does not install Julia-side dependencies | follow the setup steps above, or use native `engine = "tmb"` |
| `weights = w` | likelihood weights are not marshalled through the bridge | native `engine = "tmb"` |
| `impute = ...` or `missing = miss_control(predictor = "model")` | missing-predictor and imputation routes stay TMB-native | native `engine = "tmb"` |
| `missing = miss_control(response = "include")` outside Gaussian or bivariate Gaussian | observed-response masks are admitted only for the Gaussian bridge cells | native `engine = "tmb"` |
| non-default `control` | optimizer and storage controls do not yet map to DRM.jl | native `engine = "tmb"` |
| fixed-effect (non-phylogenetic) non-Gaussian families | count and proportion families route through Julia only with a phylogenetic random intercept; a fixed-effect non-Gaussian model stays TMB-native | native `engine = "tmb"`, or add a `phylo(1 | species)` intercept for Poisson, NB2, Gamma, Beta, or Binomial |
| partial or mismatched bivariate `phylo()` axes | the q = 4 route requires matching labelled intercepts on `mu1`, `mu2`, `sigma1`, and `sigma2` | native `engine = "tmb"` |
| Scale or variance-component Wald intervals from the Gaussian phylo route | the Julia route returns only the mean fixed-effect covariance block in this slice | Julia profile/bootstrap for the admitted phylogenetic SD target below, or native fit for other targets |
| `predict(fit, newdata = ..., dpar = "sigma")` | fresh-data Julia prediction is currently limited to location parameters | native fit or direct DRM.jl call |

For example, this should fail before JuliaCall setup:

```{r nongaussian-guard, eval = FALSE}
drmTMB(
  bf(count ~ x, sigma ~ 1),
  family = nbinom2(),
  data = counts,
  engine = "julia"
)
```

That is the right behaviour. Count families such as NB2 route through the
Julia engine only with a phylogenetic random intercept (the large-p sparse
lane); a fixed-effect `nbinom2` model stays TMB-native. Add a
`phylo(1 | species)` intercept to use the Julia route, or keep
`engine = "tmb"` for non-phylogenetic count models.

## Profile and bootstrap direction

The bridge is also a natural foundation for profile-like and bootstrap
workloads, because those tasks refit the same model many times. The right
division of labour is:

| Task | R should own | Julia should own |
| --- | --- | --- |
| Profile likelihood | target selection, R-scale constraints, plots, and interval summaries | fast constrained refits for admitted model cells |
| Parametric bootstrap | simulation design, replicate bookkeeping, seeds, summaries, and plots | repeated model fits after each simulated response is generated |
| Simulation recovery | scenario grid, tidy summaries, diagnostics, and reporting | high-throughput fitting for each scenario-replicate cell |

The first R-side repeated-refit benchmark lives in
`tools/benchmark-r-julia-bootstrap-refits.R`. The current version compares
public native `confint()` with public Julia-engine `confint()` on the same
AVONET/Hackett 1,000-species Gaussian phylogenetic model. This run used
`B = 10`, `JULIA_NUM_THREADS = 4`, and pinned BLAS/OpenMP-style low-level
threads to one.

For the admitted phylogenetic SD target, start with `method = "profile"` when
you want a likelihood-based confidence interval for one fitted model. In this
AVONET model, the profiled target is the residual phylogenetic SD in the
location model for log body mass. Biologically, it measures the amount of
remaining among-lineage variation after hand-wing index and beak length are
included. A narrow interval well above zero supports residual phylogenetic
structure; a wide interval or one near zero would warn that the phylogenetic
component is weakly identified.

The fresh parity run gives the same endpoints on the R response scale:

| route | backend | workers | elapsed s | lower | upper | status |
|---|---|---:|---:|---:|---:|---|
| R native profile | serial R process | 1 | 13.357 | 1.162186 | 1.350848 | profile |
| R native profile | `multicore` endpoint workers | 2 | 7.611 | 1.162186 | 1.350848 | profile |
| Julia-engine profile | serial Julia worker | 1 | 2.548 | 1.162188 | 1.350846 | profile |
| Julia-engine profile | Julia endpoint threads | 2 | 1.126 | 1.162188 | 1.350846 | profile |

The Julia profile endpoints differ from native R by about `1e-6` to `3e-6` on
this row. This is endpoint-parity evidence for one admitted target, not a
coverage or near-boundary calibration study. These are warm-session elapsed
timings after Julia has been initialized; they exclude the one-time
JuliaCall/DRM.jl setup recorded in the benchmark artifact's `first_call_s`
column. In elapsed time, Julia profile was about `5.24x` faster than R serial
without endpoint threading and about `6.76x` faster than R multicore when Julia
used endpoint threads.

Bootstrap remains useful when the scientific question is about
simulation-based uncertainty in a reported biological quantity, such as a
transformed phylogenetic SD or a downstream prediction. Use a larger `R`, such
as `999`, for an analysis. On the same row with `B = 10`:

| route | backend | workers | elapsed s | used / failed | lower | upper |
|---|---|---:|---:|---:|---:|---:|
| R native bootstrap | serial R process | 1 | 28.806 | 9 / 1 | 1.141876 | 1.231067 |
| R native bootstrap | `multicore` workers | 4 | 8.077 | 9 / 1 | 1.141876 | 1.231067 |
| Julia-engine bootstrap | serial Julia worker | 1 | 4.898 | 10 / 0 | 1.146677 | 1.237243 |
| Julia-engine bootstrap | Julia threads | 4 | 1.104 | 10 / 0 | 1.146677 | 1.237243 |

These bootstrap rows are a small-`B` timing smoke, not a coverage study or an
endpoint-parity claim. Because `B = 10` is tiny and the native R and Julia rows
used different successful replicate sets, compare these rows for timing and
failure handling only, not interval agreement. On this 1,000-species smoke row,
repeated Julia refits are fast, and threading helps.

The first admitted target is:

```{r julia-phylo-confint-example, eval = FALSE}
profile_targets(fit_phylo_julia)

confint(
  fit_phylo_julia,
  parm = "sd:mu:phylo(1 | species)",
  method = "profile",
  threads = TRUE
)

confint(
  fit_phylo_julia,
  parm = "sd:mu:phylo(1 | species)",
  method = "bootstrap",
  R = 999,
  seed = 20260609,
  threads = TRUE
)
```

This is still deliberately narrow. For univariate Gaussian phylogenetic fits,
profile/bootstrap extraction supports the `mu` random-effect SD target shown
above. For the bivariate q = 4 phylogenetic location-scale fit below, it
supports the four among-axis SD targets `mu1`, `mu2`, `sigma1`, and `sigma2`.
It does not yet expose profile/bootstrap intervals for fixed effects,
non-Gaussian families, a separate univariate `sigma`-phylo SD target, multiple
structured terms, or neighbouring syntax. When `threads = TRUE`, Julia can use
only the threads available in the already-started Julia runtime, so benchmark
output should record the returned `julia.threads`, `julia.workers`, and
`julia.blas_threads` columns.
Near-zero phylogenetic SD targets still need profile-shape and simulation
checks before this route can claim calibrated boundary behaviour.

The R bridge should not expose broader profile or bootstrap acceleration until
each worker returns the same coefficient scales, likelihood convention,
convergence status, and failure messages that a native `drmTMB` user expects.

## Bivariate boundary-honest among-axis SD intervals

The phylogenetic targets above are all univariate: one residual phylogenetic SD
in one `mu` model. The bridge now extends the same boundary-honest interval
discipline to the bivariate Gaussian phylogenetic cell, where the structured
effect is a 4x4 among-axis covariance over the two means and the two log scales.
Here the targets are four among-axis standard deviations, `sd_mu1`, `sd_mu2`,
`sd_sigma1`, and `sd_sigma2`: the among-lineage SD of each axis after shared
ancestry is modelled.

This is the cell where Wald intervals are not just imprecise but unavailable.
An among-axis SD is a non-negative variance component, so its true value can sit
on the boundary at zero. Two distinct things are at work here. First, the
among-axis SDs are *derived* quantities — functions of the fitted covariance
blocks — and the native engine returns `derived_interval_unavailable` for them
rather than a Wald band; that is the reported status for these RE-SD targets in
its own right, not a consequence of any single fit being degenerate. Second, when
one axis collapses the fitted covariance is near-singular and the Hessian at the
optimum stops being positive-definite (`pdHess = FALSE`), so even a symmetric Wald
band would be untrustworthy at the boundary. Either way the Wald route gives no
usable interval here; native `engine = "tmb"` fits the bivariate covariance blocks
and the Julia engine sits on top of that same fit to supply profile and bootstrap
intervals it cannot.

The division of labour is the same as for the univariate cell. Native
`engine = "tmb"` fits the model. The Julia engine adds the boundary-honest
intervals: `confint()` on an `engine = "julia"` bivariate fit returns one row per
admitted among-axis SD, with a collapsed axis reported as `lower = 0` rather than
as a spuriously narrow Wald band.

```{r biv-phylo-fit-shape, eval = FALSE}
bform_biv_phylo <- bf(
  mu1 = leaf_area ~ soil_moisture + phylo(1 | p | species, tree = tree),
  mu2 = root_mass ~ soil_moisture + phylo(1 | p | species, tree = tree),
  sigma1 = ~ 1 + phylo(1 | p | species, tree = tree),
  sigma2 = ~ 1 + phylo(1 | p | species, tree = tree),
  rho12 = ~ 1
)

fit_biv_phylo_julia <- drmTMB(
  bform_biv_phylo,
  family = biv_gaussian(),
  data = plant_traits,
  engine = "julia",
  REML = TRUE
)
```

`REML = TRUE` is the public restricted-likelihood switch for this cell when the
installed DRM.jl build includes q = 4 REML support. The bridge forwards the
REML request to Julia; it does not by itself validate the wall-time or coverage
properties of a full 10k-tip, multi-tree workflow. For variance-component
uncertainty, inspect the profile or bootstrap diagnostics returned below rather
than treating Wald intervals as calibrated at the boundary.

With no `parm`, `confint()` returns all four admitted among-axis SD targets as a
multi-row table. `confint()` defaults to `method = "wald"`; for these variance
components pass `method = "profile"` (recommended) or `method = "bootstrap"`:

```{r biv-phylo-confint-example, eval = FALSE}
confint(
  fit_biv_phylo_julia,
  method = "profile",
  threads = TRUE
)

confint(
  fit_biv_phylo_julia,
  parm = "sd:mu1:phylo(1 | species)",
  method = "bootstrap",
  R = 999,
  seed = 20260609,
  threads = TRUE
)
```

The profile rows are real profile likelihood, not a bootstrap fallback: the
table reports `profile.engine = "julia_profile_result"` and a per-row
`profile.boundary` flag, and each row is reported according to *its own* axis —
not all four axes collapse. An **identified** axis (positive among-axis signal)
returns a strictly-positive lower bound and `profile.boundary = FALSE`: an
ordinary two-sided profile interval. A **collapsed** axis returns `lower = 0`,
with the upper end carried out as far as the profile remains flat, and
`profile.boundary = TRUE`. The collapsed case is the boundary-honest behaviour:
the interval admits that the data place no positive lower bound on that axis,
instead of reporting a symmetric Wald band that the non-invertible Hessian could
not have produced.

The four targets carry the `parm` labels `sd:mu1:phylo(1 | species)`,
`sd:mu2:phylo(1 | species)`, `sd:sigma1:phylo(1 | species)`, and
`sd:sigma2:phylo(1 | species)` (the grouping variable is taken from your fitted
formula). Each is reported by its own axis state:

| Axis state | `lower` | `profile.boundary` | interpretation |
| --- | --- | --- | --- |
| identified (positive signal) | `> 0` | `FALSE` | ordinary two-sided profile interval |
| collapsed to the boundary | `0` | `TRUE` | profile carried to flatness; no positive lower bound |

### With missing responses

The bivariate among-axis interval path also runs when part of one response is
missing. Because the bivariate Julia route uses a per-cell observed mask, a
`mu1` value can be `NA` for a tip while that tip's `mu2` value is still used, and
the tree is kept whole. The admitted route is `missing = miss_control(response =
"include")`, which fits the observed data rather than dropping or imputing rows:

```{r biv-phylo-missing-shape, eval = FALSE}
plant_traits$leaf_area[sample(nrow(plant_traits), 0.25 * nrow(plant_traits))] <- NA

fit_biv_phylo_miss <- drmTMB(
  bform_biv_phylo,
  family = biv_gaussian(),
  data = plant_traits,
  missing = miss_control(response = "include"),
  engine = "julia"
)

confint(fit_biv_phylo_miss, method = "profile", threads = TRUE)
```

This is the only admitted missing-data route through the bridge. `response =
"include"` is admitted for Gaussian and bivariate Gaussian responses; predictor
missingness, imputation, and other `miss_control()` settings stay TMB-native and
still error early.

### Status and honest limits

| Model cell | Native `engine = "tmb"` | Julia engine adds |
| --- | --- | --- |
| Bivariate Gaussian `phylo(1 | species)` among-axis SDs | fits the 4x4 covariance blocks; the derived among-axis SD Wald intervals come back `derived_interval_unavailable` (and at a collapsed axis `pdHess = FALSE` makes a symmetric band untrustworthy regardless) | profile (recommended) and bootstrap intervals for the four among-axis SDs, with `lower = 0` at a collapsed axis |
| Same cell with `missing = miss_control(response = "include")` | observed-data fit | same four boundary-honest among-axis SD intervals on the observed-data fit |

The principle is the same one used for fixed effects elsewhere in this article.
Wald is enough for fixed effects unless `pdHess = FALSE`; for random-effect
variance components such as these among-axis SDs, use profile (recommended) or
bootstrap, and refuse Wald rather than report an interval the Hessian could not
support. The bridge gives Wald for the fixed effects, profile or bootstrap for
the SDs, and correctly declines Wald for the SDs.

Two limits stay explicit. The boundary call itself, `lower = 0` at a collapsed
axis, is the robust, verified part. But profile coverage for a small number of
tips near the boundary is irreducibly uncertain, so this route does not claim
calibrated small-sample coverage. Bootstrap undercoverage on the scale axes is
measured rather than assumed, at roughly 0.52 in the runs to date, so prefer the
profile default for these targets and read a bootstrap interval as a
simulation-based cross-check, not a calibrated band. As elsewhere, this is
endpoint-honesty for an admitted cell, not a coverage study.
```


## Contributor workflow

Treat bridge work as a two-repository contract. The R half lives in `drmTMB`;
the Julia primitive lives in `DRM.jl`.

| Layer | Main files | Responsibility |
| --- | --- | --- |
| R entry point | `R/drmTMB.R` | preserve `engine = "tmb"` as the default and dispatch only explicit `engine = "julia"` calls |
| R bridge glue | `R/julia-bridge.R` | validate support, serialize formulas and trees, call JuliaCall, and reconstruct `drmTMB_julia` |
| R tests | `tests/testthat/test-julia-bridge.R` | test formula/tree marshalling, object methods, target inventory, response-scale interval conversion, and guardrails without requiring JuliaCall |
| Julia primitive | `DRM.jl/src/bridge.jl` | build DRM.jl formulas from strings, fit the model, run the narrow profile/bootstrap primitive, and flatten results to R-friendly pieces |
| Julia default routes | `DRM.jl/src/gaussian_core.jl` and `DRM.jl/src/location_only.jl` | choose the default Gaussian phylogenetic route and keep algorithm alternatives out of the R-facing contract until tested |
| Julia tests | `DRM.jl/test/test_bridge.jl` and `DRM.jl/test/test_conjugate_em.jl` | compare bridge output against native DRM.jl fits and pin the default sparse phylo route |
| Public docs | this article, `NEWS.md`, and the DRM.jl bridge page | keep support wording narrower than the code can accidentally dispatch |
| Project memory | `docs/dev-log/check-log.md` and after-task notes | record exact checks, limitations, and issue-maintenance status |

Run these focused gates first:

```sh
julia --project=. test/test_bridge.jl
```

```sh
julia --project=. test/test_conjugate_em.jl
```

```sh
Rscript -e 'devtools::load_all(); testthat::test_file("tests/testthat/test-julia-bridge.R")'
```

When JuliaCall is not installed in the main R library, use a temporary library
for the integration smoke:

```sh
tmp_lib=$(mktemp -d /tmp/drmtmb-juliacall-lib-XXXXXX)
R_LIBS_USER="$tmp_lib" Rscript -e '
.libPaths(c(Sys.getenv("R_LIBS_USER"), .libPaths()))
install.packages("JuliaCall", lib = Sys.getenv("R_LIBS_USER"),
  repos = "https://cloud.r-project.org", quiet = TRUE)
devtools::load_all("/path/to/drmTMB")
options(drmTMB.DRM.jl.path = "/path/to/DRM.jl")
dat <- data.frame(y = rnorm(24), x = seq(-1, 1, length.out = 24))
fit <- drmTMB(bf(y ~ x, sigma ~ 1), data = dat, engine = "julia")
print(fit)
print(coef(fit))
'
```

After those checks pass, record the result in the check log and after-task
report. Full `devtools::test()`, `devtools::check()`, and
`pkgdown::check_pkgdown()` are still needed before a release or broad PR merge.

## How to widen the bridge

Do not widen the R bridge by only adding a family name to a dispatch table.
DRM.jl can keep extra native routes, experimental algorithms, and Julia-first
features. The R bridge should admit a route only when it matches the R-facing
`drmTMB` contract or fails early with a specific message. A new admitted family
or structure needs all of the following:

1. A DRM.jl bridge test that fits the family through `drm_bridge()` and compares
   coefficients, covariance, log likelihood, fitted values, and scale summaries
   against a native DRM.jl fit.
2. A drmTMB-side test that the R object methods expose the returned fields on
   the intended public scale.
3. A coefficient-scale decision. If DRM.jl and drmTMB use different working
   scales, document the transform and test the matching covariance Jacobian.
4. A guardrail test showing unsupported neighbouring syntax still fails early.
5. Updated public docs that say exactly which formula, family, and post-fit
   methods are admitted.

This checklist is why the R bridge admits only cells that have R-scale parity
tests: Gaussian fixed-effect and bivariate models, the Gaussian phylogenetic
and σ-phylo location-scale cells, the bivariate q = 4 phylogenetic
location-scale route, and gated non-Gaussian phylogenetic cells (a
`phylo(1 | group)` intercept on Poisson, NB2, Gamma, Beta, or Binomial, plus
`relmat()`, `animal()`, and `spatial()` structured routes for Gaussian,
Poisson, NB2, and Gamma). It is better for a bridge to be narrow and dependable
than broad and ambiguous.
