---
title: "Working with large data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with large data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144,
  dev.args = list(bg = "white"),
  purl = FALSE
)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

theme_large_data <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey20"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA),
      legend.position = "bottom",
      legend.title = ggplot2::element_blank()
    )
}
```

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    min-width: 720px;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

Large data changes the workflow. A model that is statistically reasonable can
still fail because R builds large model frames, dense model matrices, fitted
objects, predictions, and residual vectors before the scientific question is
answered. This article is for users fitting ecological, evolutionary, or
environmental-science models with many observation rows, many species, or
both, once the standard post-fit workflow in [Checking and using fitted
models](model-workflow.html) starts choking on size.

The short version is: fit the smallest model that answers the question, keep
only the fitted-object pieces you need, and benchmark on a smaller version of
the real data before running the full analysis.

## What is implemented now

`drmTMB` has a first memory-light fitted-object path through `drm_control()`.
For a large univariate Gaussian phylogenetic location model, use:

```{r, purl=FALSE}
library(drmTMB)

fit <- drmTMB(
  bf(
    y ~ x1 + x2 + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat,
  control = drm_control(
    optimizer = list(eval.max = 1000, iter.max = 1000),
    se = FALSE,
    keep_data = FALSE,
    keep_model_frame = FALSE,
    keep_tmb_object = FALSE
  )
)
```

Here `keep_data = FALSE` removes the stored complete-case data frame from the
fitted object after optimization. `keep_model_frame = FALSE` removes stored
model frames after TMB data are built, including nested model-frame caches for
direct random-effect SD models such as `sd(species, level = "phylogenetic") ~ z_species` and
fitted q=2 latent-correlation models such as `corpair(...) ~ ecology`.
`se = FALSE` skips the `TMB::sdreport()` step after optimization. The fitted
coefficients, fitted values, residuals, prediction, simulation, `sigma`,
`rho12`, and `corpairs()` extraction still use the stored model matrices,
terms, offsets, response vectors, response-name metadata, random-effect
metadata, and optimized parameters. Standard errors, `vcov()`, and Wald
confidence intervals are unavailable in this mode, and `summary()` marks the
standard-error state explicitly. `keep_tmb_object = FALSE` separately removes
the TMB automatic-differentiation object after optimization.

This control is useful when the fitted model is kept for interpretation,
prediction, or plotting, but the original data frame and TMB object do not need
to travel with it.

For final interval estimates, keep the pieces required by the interval method.
Wald intervals need `se = TRUE`, profile intervals need
`keep_tmb_object = TRUE`, and bootstrap intervals need `keep_data = TRUE`
because the simulated responses are refit with the original model data.
Exploratory memory-light fits can therefore use `se = FALSE` and
`keep_tmb_object = FALSE`, while a final inference refit should keep the
standard-error covariance, TMB object, or model data that the chosen interval
method needs.

For factor-heavy fixed-effect Gaussian location models, the first sparse
fixed-effect path is also available:

```{r, purl=FALSE}
fit_sparse <- drmTMB(
  bf(y ~ habitat + x1, sigma ~ 1),
  family = gaussian(),
  data = dat,
  control = drm_control(sparse_fixed = TRUE)
)
```

This path stores the `mu` fixed-effect design as a sparse `Matrix` object and
uses a sparse TMB matrix multiply. It is intentionally limited to fixed-effect
univariate Gaussian `mu` models with intercept-only `sigma`.

For repeated-row Gaussian fixed-effect models, the first aggregation path is
also available:

```{r, purl=FALSE}
fit_agg <- drmTMB(
  bf(y ~ habitat + season, sigma ~ effort_class),
  family = gaussian(),
  data = dat,
  control = drm_control(aggregate_gaussian = TRUE)
)
```

This path collapses rows that have the same processed `mu` and `sigma` design
state after model-row filtering. TMB evaluates the Gaussian likelihood with
cell summaries `n`, `sum(y)`, and `sum(y^2)`, while the fitted object still
keeps original-row model matrices and response vectors so `predict(fit_agg)`,
`fitted(fit_agg)`, and `residuals(fit_agg)` return one value per original
model row.

## What this does not solve yet

The current memory-light path does not prevent R from building model frames and
model matrices before optimization. It can drop model frames after fitting, but
those objects still need to exist during model construction. That means the
control helps fitted-object size, but it is not a complete solution for
millions of rows.

These pieces remain planned:

- sparse fixed-effect matrices for `sigma`, random-effect, phylogenetic,
  spatial, bivariate, and non-Gaussian models;
- Gaussian sufficient-statistic aggregation beyond the first fixed-effect
  univariate path. Random effects, direct-SD formulas, structured effects,
  known sampling covariance, bivariate Gaussian models, non-Gaussian families,
  non-unit likelihood weights, and combined sparse fixed-effect matrices are
  still rejected;
- sparse or block-sparse storage for full-matrix
  `meta_V(V = V)`. The current full known-covariance route keeps the
  retained `V` as a dense R matrix and is for small to moderate correlated
  sampling-error fits;
- memory-light modes that avoid constructing dense model frames or dense model
  matrices in the first place;
- repeated benchmark runs for million-row, bivariate, factor-heavy,
  non-Gaussian, and 10,000-species analyses.

If a model has a high-cardinality factor in a fixed-effect formula, the dense
`model.matrix()` step may dominate memory before TMB sees the data. Treat that
as a sparse-design problem, not as a phylogenetic precision problem.

If a model has many repeated Gaussian rows with the same `mu` and `sigma`
design rows, aggregation is a separate tool from sparse matrices. In the first
implemented path, the likelihood uses cell summaries such as `n`, `sum(y)`,
and `sum(y^2)`, but fitted-row summaries remain original-row summaries.

## Choose the interval route deliberately

For large phylogenetic Gaussian fits, start with Wald intervals for routine
reporting. They are fast and now include direct response-scale targets such as
residual `sigma` and `sd:mu:phylo(1 | species)` when `TMB::sdreport()`
succeeds:

```{r, purl=FALSE}
ci_wald <- confint(
  fit_final,
  parm = c("sigma", "sd:mu:phylo(1 | species)"),
  method = "wald"
)
```

Use profile intervals when you want a likelihood-based check for a small number
of fitted targets. With the default `profile_engine = "auto"`, direct scalar
scale, SD, and correlation targets use the endpoint solver instead of building a
full adaptive profile curve:

```{r, purl=FALSE}
ci_profile <- confint(
  fit_final,
  parm = c("sigma", "sd:mu:phylo(1 | species)"),
  method = "profile"
)
```

If you need the previous full-profile route for comparison or debugging, ask
for it explicitly:

```{r, purl=FALSE}
ci_profile_full <- confint(
  fit_final,
  parm = "sd:mu:phylo(1 | species)",
  method = "profile",
  profile_engine = "tmbprofile"
)
```

On Unix-like systems, `parallel = "multicore"` can split independent profile
targets across workers. For a single direct endpoint target, it can split the
lower and upper endpoint searches. When `workers = NULL`, `drmTMB` uses about
half the detected physical cores, capped at the number of jobs and at 10
workers:

```{r, purl=FALSE}
ci_profile_parallel <- confint(
  fit_final,
  parm = c("sigma", "sd:mu:phylo(1 | species)"),
  method = "profile",
  parallel = "multicore",
  workers = NULL
)
```

Bootstrap intervals are a heavier simulation-refit audit. They require the
fitted model data, so do not combine them with
`drm_control(keep_data = FALSE)` unless you refit the final model with
`keep_data = TRUE` first. Use a small `R` as a smoke test, then increase `R`
only for the final check:

```{r, purl=FALSE}
ci_boot <- confint(
  fit_final,
  parm = c("sigma", "sd:mu:phylo(1 | species)"),
  method = "bootstrap",
  R = 99,
  seed = 1,
  parallel = "multicore",
  workers = NULL
)
```

Treat these routes as different evidence, not as interchangeable defaults. Wald
intervals are the routine first pass, endpoint profiles are likelihood-based
checks for direct scalar targets, full `TMB::tmbprofile()` is the debugging and
comparison route, and bootstrap is a slower refit-based audit.

The current endpoint benchmark artifact gives a scale for those choices. The
plot below uses the synthetic Gaussian phylogenetic benchmark rows recorded in
`docs/dev-log/benchmarks/profile-scalar-endpoint-v2.csv`. It intentionally uses
points only: elapsed seconds are local timing measurements from one machine, not
confidence intervals.

```{r large-data-profile-benchmark-timing, eval=has_ggplot2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.4, fig.height=4.4, fig.cap="Development benchmark timings for direct scalar profile intervals. Points are elapsed seconds from one local synthetic Gaussian phylogenetic benchmark table; they are performance measurements, not confidence intervals.", fig.alt="Horizontal dot plot comparing tmbprofile, endpoint, and endpoint multicore elapsed seconds for residual sigma and phylogenetic SD benchmark scenarios. Endpoint and endpoint multicore are faster than tmbprofile in every shown successful benchmark row."}
profile_bench <- data.frame(
  scenario = c(
    rep("sigma\n10k rows / 1k species", 3),
    rep("phylo SD\n10k rows / 1k species", 3),
    rep("phylo SD\n100k rows / 1k species", 3),
    rep("phylo SD\n100k rows / 5k species", 3)
  ),
  engine = rep(
    c("Full tmbprofile", "Endpoint", "Endpoint + 2 workers"),
    times = 4
  ),
  elapsed_sec = c(
    32.690, 6.993, 3.777,
    21.774, 7.610, 4.734,
    235.344, 51.193, 27.652,
    249.426, 43.857, 24.527
  )
)

profile_bench$scenario <- factor(
  profile_bench$scenario,
  levels = c(
    "sigma\n10k rows / 1k species",
    "phylo SD\n10k rows / 1k species",
    "phylo SD\n100k rows / 1k species",
    "phylo SD\n100k rows / 5k species"
  )
)
profile_bench$engine <- factor(
  profile_bench$engine,
  levels = c("Full tmbprofile", "Endpoint", "Endpoint + 2 workers")
)

ggplot2::ggplot(
  profile_bench,
  ggplot2::aes(elapsed_sec, scenario, colour = engine, shape = engine)
) +
  ggplot2::geom_line(
    ggplot2::aes(group = scenario),
    colour = "grey78",
    linewidth = 0.45
  ) +
  ggplot2::geom_point(size = 3.2) +
  ggplot2::scale_x_log10(
    breaks = c(3, 5, 10, 30, 100, 300),
    labels = c("3", "5", "10", "30", "100", "300"),
    limits = c(3, 320)
  ) +
  ggplot2::scale_colour_manual(
    values = c(
      "Full tmbprofile" = "#4D4D4D",
      "Endpoint" = "#0072B2",
      "Endpoint + 2 workers" = "#009E73"
    )
  ) +
  ggplot2::scale_shape_manual(
    values = c(
      "Full tmbprofile" = 16,
      "Endpoint" = 17,
      "Endpoint + 2 workers" = 15
    )
  ) +
  ggplot2::labs(
    title = "Endpoint profiles reduce direct scalar interval time",
    subtitle = "Synthetic local benchmarks; points are elapsed seconds, not uncertainty",
    x = "Elapsed seconds (log scale)",
    y = NULL
  ) +
  theme_large_data()
```

## Check the fit with the storage choice in mind

Run `check_drm()` before interpreting a large fit:

```{r, purl=FALSE}
check_drm(fit)
```

If `keep_tmb_object = FALSE`, the fixed-gradient row is a note rather than an
`ok` result. That note means the gradient cannot be re-evaluated because
`fit$obj` was intentionally dropped. Refit with
`drm_control(keep_tmb_object = TRUE)` when you need the gradient diagnostic for
a final report.

If `se = FALSE`, the `sdreport_status`, Hessian, and finite-standard-error
rows are notes rather than `ok` results. That note means standard errors were
intentionally skipped; refit with `drm_control(se = TRUE)` when Wald standard
errors, `vcov()`, or Wald confidence intervals are part of the final inference.

The `optimizer_budget` row compares `nlminb()` iterations and function
evaluations with supplied `iter.max` and `eval.max`. If it is a note or warning,
inspect convergence, gradients, and model structure before increasing the
limits.

Also inspect the `fixed_effect_design_size` row. It reports the retained
fixed-effect design size, the widest design block, the matrix class, and the
density of the largest retained block. A dense-fit note with a low density
means high-cardinality factors or sparse interactions may be driving memory use
before TMB optimization. That is a sparse fixed-effect design problem;
simplifying the formula or using `drm_control(sparse_fixed = TRUE)` when the
model is inside the implemented Gaussian `mu` scope is more relevant than
changing the phylogenetic tree.

For full-matrix `meta_V(V = V)` fits, also inspect the
`known_sampling_covariance` row. It is a note for dense-matrix storage and
reports dimension, density, size, rank, and conditioning. A low-density dense
`V` often means the statistical structure is block-sparse, but the current fit
has still paid dense R-matrix storage costs.

For aggregated Gaussian fits, `check_drm()` also includes a
`gaussian_aggregation` row. That row reports the number of original rows, the
number of aggregation cells, the compression ratio, and the largest cell size.
If the compression ratio is close to 1, aggregation was valid but did not buy
much row compression for that model.

## Avoid accidental full-row output

Functions such as `predict(fit)`, `fitted(fit)`, `residuals(fit)`, and
`simulate(fit)` can create vectors or data frames with one value per
observation. On a five-million-row dataset, that output is itself large.

For interpretation plots, prefer smaller `newdata` grids:

```{r, purl=FALSE}
new_dat <- data.frame(
  x1 = seq(-2, 2, length.out = 100),
  x2 = 0,
  species = dat$species[[1]]
)

predict(fit, newdata = new_dat, dpar = "mu")
```

For residual checks, decide whether you need every row or a targeted sample.
For simulations, start with small `nsim` and store only the summaries needed
for the biological question.

## Benchmark before the full run

The repository includes a non-CRAN benchmark harness:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --memory-light true --output bench/results.csv
```

The script generates a synthetic Gaussian phylogenetic location dataset, fits:

```r
y ~ x1 + x2 + phylo(1 | species, tree = tree)
sigma ~ 1
```

and records elapsed time, fitted-object size, model-matrix size, TMB-data size,
the largest retained fixed-effect design block and density, prediction time,
residual time, convergence code, optimizer message, optimizer evaluation
counts, and fitted scale summaries. Use it to learn how your machine behaves
before running the full analysis.

For factor-heavy designs, add:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --factor-heavy true --memory-light true --output bench/results.csv
```

To benchmark the first sparse fixed-effect path, remove the phylogenetic
structured effect explicitly:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --structured none --factor-heavy true --sparse-fixed true \
  --memory-light true --output bench/results.csv
```

That sparse smoke scenario fits a fixed-effect Gaussian location model, not a
phylogenetic model, because `sparse_fixed = TRUE` is currently implemented only
for univariate Gaussian `mu` fixed effects with intercept-only `sigma`.

To benchmark the first Gaussian aggregation path, use a non-phylogenetic
repeated-cell scenario:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --structured none --aggregate-gaussian true --aggregation-cells 100 \
  --memory-light true --output bench/results.csv
```

That scenario creates repeated fixed-effect cells deliberately, so the
benchmark can record the fitted aggregation-cell count and compression ratio.

For a model with a scale predictor, add:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --sigma-x true --memory-light true --output bench/results.csv
```

The benchmark is intentionally outside the package test suite. It can be slow,
and it measures local hardware, BLAS, compiler, and operating-system behaviour
as much as package code.

## What the current development benchmarks say

The development benchmark table in `docs/dev-log/benchmark-results.md` records
selected local runs. Treat the numbers as planning evidence, not as speed
promises for your computer.

| Scenario | Local result | What to learn |
| --- | --- | --- |
| 100k rows, 1k species, ordinary location model | converged; about 25 seconds; about 1.4 GB macOS max RSS | This is the first useful local rung before trying a real large analysis. |
| 100k rows, 5k species | converged; about 32 seconds; about 1.7 GB macOS max RSS | Species-index pressure is separate from row-count pressure. |
| 500k rows, 1k species | converged; about 131 seconds; about 5.1 GB macOS max RSS | Row pressure is feasible on the test machine, but peak memory is already substantial. |
| 100k rows, 1k species, 40-level fixed factor | did not converge under the benchmark optimizer budget | Dense fixed-effect matrices remain a real scaling limitation. |

These results support cautious use of the current sparse phylogenetic location
path. They do not prove million-row readiness, 10,000-species readiness,
bivariate large-data readiness, or non-Gaussian large-data readiness.

## Practical checklist

- Start with a fixed-effect or smaller phylogenetic model that answers a
  simpler version of the question.
- Fit with `keep_data = FALSE`, `keep_model_frame = FALSE`, `se = FALSE`, and
  `keep_tmb_object = FALSE` when you do not need to carry the original data,
  model frames, standard-error covariance, or TMB object inside the saved fit.
- Refit the final model with `se = TRUE` for Wald intervals,
  `keep_tmb_object = TRUE` for profile intervals, and `keep_data = TRUE` for
  bootstrap intervals.
- Use Wald intervals first for direct `sigma`, structured SD, and correlation
  targets; use endpoint profile intervals as a likelihood-based check on the
  few targets that matter most.
- Keep `keep_tmb_object = TRUE` for final diagnostic refits where the fixed
  gradient check matters.
- Inspect `check_drm(fit)` for fixed-effect design-size notes before assuming
  the phylogenetic part caused memory pressure.
- Use `newdata` grids for interpretation instead of creating full-row
  predictions by default.
- Run the benchmark script at 100k rows before trying million-row analyses.
- Treat a non-zero optimizer code as a diagnostic result, not as timing
  evidence.
- Treat sparse fixed-effect matrices outside fixed-effect Gaussian `mu` models,
  and data aggregation in all models, as future scaling features rather than
  current guarantees.
