---
title: "Working with large data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with large data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Large data changes the workflow. A model that is statistically reasonable can
still fail because R builds large model frames, dense model matrices, fitted
objects, predictions, and residual vectors before the scientific question is
answered. This article is for users fitting ecological, evolutionary, or
environmental-science models with many observation rows, many species, or both.

The short version is: fit the smallest model that answers the question, keep
only the fitted-object pieces you need, and benchmark on a smaller version of
the real data before running the full analysis.

## What is implemented now

`drmTMB` has a first memory-light fitted-object path through `drm_control()`.
For a large univariate Gaussian phylogenetic location model, use:

```{r}
library(drmTMB)

fit <- drmTMB(
  bf(
    y ~ x1 + x2 + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat,
  control = drm_control(
    optimizer = list(eval.max = 1000, iter.max = 1000),
    keep_data = FALSE,
    keep_model_frame = FALSE,
    keep_tmb_object = FALSE
  )
)
```

Here `keep_data = FALSE` removes the stored complete-case data frame from the
fitted object after optimization. `keep_model_frame = FALSE` removes stored
model frames after TMB data are built. `keep_tmb_object = FALSE` removes the
TMB automatic-differentiation object after optimization. The fitted
coefficients, standard errors, fitted values, residuals, simulation, and
`sigma` extraction still use the stored model matrices, terms, offsets,
response vectors, response-name metadata, optimized parameters, and `sdreport`
object.

This control is useful when the fitted model is kept for interpretation,
prediction, or plotting, but the original data frame and TMB object do not need
to travel with it.

## What this does not solve yet

The current memory-light path does not prevent R from building model frames and
model matrices before optimization. It can drop model frames after fitting, but
those objects still need to exist during model construction. That means the
control helps fitted-object size, but it is not a complete solution for
millions of rows.

These pieces remain planned:

- sparse fixed-effect matrices for factor-heavy models;
- Gaussian sufficient-statistic aggregation when repeated rows can be collapsed
  without changing the likelihood;
- explicit benchmark runs for 100k, 500k, 1M, and 5M rows with 1k to 10k
  species.

If a model has a high-cardinality factor in a fixed-effect formula, the dense
`model.matrix()` step may dominate memory before TMB sees the data. Treat that
as a sparse-design problem, not as a phylogenetic precision problem.

## Check the fit with the storage choice in mind

Run `check_drm()` before interpreting a large fit:

```{r}
check_drm(fit)
```

If `keep_tmb_object = FALSE`, the fixed-gradient row is a note rather than an
`ok` result. That note means the gradient cannot be re-evaluated because
`fit$obj` was intentionally dropped. Refit with
`drm_control(keep_tmb_object = TRUE)` when you need the gradient diagnostic for
a final report.

Also inspect the `fixed_effect_design_size` row. It reports the retained dense
fixed-effect design size and the widest design block. A note there means
high-cardinality factors or interactions may be driving memory use before TMB
optimization. That is a sparse fixed-effect design problem; simplifying the
formula or waiting for the future sparse path is more relevant than changing
the phylogenetic tree.

## Avoid accidental full-row output

Functions such as `predict(fit)`, `fitted(fit)`, `residuals(fit)`, and
`simulate(fit)` can create vectors or data frames with one value per
observation. On a five-million-row dataset, that output is itself large.

For interpretation plots, prefer smaller `newdata` grids:

```{r}
new_dat <- data.frame(
  x1 = seq(-2, 2, length.out = 100),
  x2 = 0,
  species = dat$species[[1]]
)

predict(fit, newdata = new_dat, dpar = "mu")
```

For residual checks, decide whether you need every row or a targeted sample.
For simulations, start with small `nsim` and store only the summaries needed
for the biological question.

## Benchmark before the full run

The repository includes a non-CRAN benchmark harness:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --memory-light true --output bench/results.csv
```

The script generates a synthetic Gaussian phylogenetic location dataset, fits:

```r
y ~ x1 + x2 + phylo(1 | species, tree = tree)
sigma ~ 1
```

and records elapsed time, fitted-object size, model-matrix size, TMB-data size,
prediction time, residual time, convergence code, optimizer message, optimizer
evaluation counts, and fitted scale summaries. Use it to learn how your
machine behaves before running the full analysis.

For factor-heavy designs, add:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --factor-heavy true --memory-light true --output bench/results.csv
```

For a model with a scale predictor, add:

```sh
Rscript bench/large-phylo-location.R --rows 100000 --species 1000 \
  --sigma-x true --memory-light true --output bench/results.csv
```

The benchmark is intentionally outside the package test suite. It can be slow,
and it measures local hardware, BLAS, compiler, and operating-system behaviour
as much as package code.

## Practical checklist

- Start with a fixed-effect or smaller phylogenetic model that answers a
  simpler version of the question.
- Fit with `keep_data = FALSE`, `keep_model_frame = FALSE`, and
  `keep_tmb_object = FALSE` when you do not need to carry the original data,
  model frames, or TMB object inside the saved fit.
- Keep `keep_tmb_object = TRUE` for final diagnostic refits where the fixed
  gradient check matters.
- Inspect `check_drm(fit)` for fixed-effect design-size notes before assuming
  the phylogenetic part caused memory pressure.
- Use `newdata` grids for interpretation instead of creating full-row
  predictions by default.
- Run the benchmark script at 100k rows before trying million-row analyses.
- Treat sparse fixed-effect matrices and data aggregation as future scaling
  features, not current guarantees.
