---
title: "When variance carries signal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When variance carries signal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}

location_scale_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_line(
        colour = "grey90",
        linewidth = 0.3
      ),
      panel.grid.major.y = ggplot2::element_line(
        colour = "grey87",
        linewidth = 0.35
      ),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(colour = "grey20")
    )
}

location_scale_habitat_scales <- function() {
  list(
    ggplot2::scale_colour_manual(
      values = c("forest" = "#0072B2", "grassland" = "#009E73")
    ),
    ggplot2::scale_fill_manual(
      values = c("forest" = "#0072B2", "grassland" = "#009E73")
    )
  )
}
```

This tutorial starts with a biological question: do two habitats differ only in
mean growth, or does one habitat also make growth less predictable? In a
Gaussian location-scale model, the location part models the expected response
`mu`, while the scale part models the residual standard deviation `sigma`.
That residual scale is not just a nuisance parameter; it can be the scientific
answer when the question is about individual variability, predictability, or
remaining heterogeneity after the mean model is accounted for.

If you are fitting your first model, start with the
[worked growth example](#worked-example-growth-mean-and-predictability), then
return to the syntax overview below when you want to adapt the formula.

The core Gaussian location-scale route supports fixed effects, optional random
effects in the location formula, residual-scale random intercepts and
independent random slopes in the `sigma` formula, and random-effect scale
formulas such as `sd(id) ~ x_group`. The same pattern is used throughout the
documentation: write the model symbolically first, show the matching R syntax,
fit the model, and interpret the fitted output.

Current syntax:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Random-intercept location syntax:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

Simple random-slope location syntax:

```r
drmTMB(
  bf(y ~ x1 + (0 + x1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

Residual-scale random-effect syntax:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2 + (1 | id) + (0 + w | id)),
  family = gaussian(),
  data = dat
)
```

Residual-scale random effects live on `log(sigma)`. The independent
residual-scale slope route `sigma ~ z + (0 + w | id)` fits a random slope SD
for `w`, and the unlabelled correlated block `sigma ~ z + (1 + w | id)` is
also implemented (recovery grade: trust the point estimate, not the
interval). Only labelled residual-scale slope covariance such as
`sigma ~ z + (1 + w | p | id)` or `sigma ~ z + (0 + w | p | id)` remains
planned.

The companion article "Which scale are you modelling?" gives a side-by-side
guide to `sigma ~ x`, `sigma ~ x + (1 | id)`,
`sigma ~ x + (0 + w | id)`, `sd(id) ~ x_group`,
random-slope correlations, and bivariate residual `rho12`.

## Model equations and matching R syntax

The basic discipline is simple: each estimated parameter gets its own linear
predictor. For Gaussian location-scale regression those parameters are the
mean, `mu`, and the residual standard deviation, `sigma`.

Throughout this article, `Normal(a, b)` uses variance as the second argument.
Thus `Normal(mu_i, sigma_i^2)` means a Gaussian distribution with mean `mu_i`
and standard deviation `sigma_i`.

For the fixed-effect Gaussian location-scale model, the symbolic model is:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= \beta_0 + \beta_1 x_{1i},\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 x_{2i}.
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The location formula `y ~ x1` defines the design matrix for the `beta`
coefficients. The scale formula `sigma ~ x2` defines the design matrix for the
`gamma` coefficients. The log link makes `sigma_i` positive:

\[
\sigma_i = \exp(\gamma_0 + \gamma_1 x_{2i}).
\]

The same model can be written more compactly as:

\[
\begin{aligned}
\eta_{\mu i} &= X_{\mu}[i, ]\beta_{\mu},\\
\eta_{\sigma i} &= X_{\sigma}[i, ]\beta_{\sigma},\\
\mu_i &= \eta_{\mu i},\\
\sigma_i &= \exp(\eta_{\sigma i}).
\end{aligned}
\]

The mapping is:

| Symbol | Meaning | R syntax source |
|---|---|---|
| `y_i` | observed response | left-hand side of `y ~ x1` |
| `X_mu` | design matrix for the mean | right-hand side of `y ~ x1` |
| `beta_mu` | mean coefficients | `coef(fit, "mu")` |
| `mu_i` | fitted mean | `fitted(fit)` or `predict(fit, dpar = "mu")` |
| `X_sigma` | design matrix for the residual scale | right-hand side of `sigma ~ x2` |
| `beta_sigma` | log-scale coefficients | `coef(fit, "sigma")` |
| `sigma_i` | residual standard deviation | `sigma(fit)` or `predict(fit, dpar = "sigma")` |

With a random intercept in the location part, the model becomes:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,id}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Here `sigma_ij` is the residual standard deviation. The random-intercept scale
`sd_mu_id` is a separate group-level standard deviation. In implemented
double-hierarchical scale models, `sd(id) ~ x_group` exposes this group-level
scale component by name; it should not be confused with residual `sigma`.

Residual-scale random intercepts enter a different part of the model:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= X_{\mu}[ij, ]\beta_{\mu} + b_j,\\
\log(\sigma_{ij})
  &= X_{\sigma}[ij, ]\beta_{\sigma} + a_j,\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,id}^2),\\
a_j
  &\sim \operatorname{Normal}(0, sd_{\sigma,id}^2).
\end{aligned}
\]

The matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2 + (1 | id) + (0 + w | id)),
  family = gaussian(),
  data = dat
)
```

Here `a_j` is residual-scale heterogeneity, or group-to-group variation in
within-group variability. It is still not the same quantity as `sd_mu_id`, the
standard deviation of the `mu` random intercepts.

A double-hierarchical scale formula targets the random-effect standard
deviation itself. For example:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}, b_j
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &= sd_{\mu,id,j} u_j,\\
u_j
  &\sim \operatorname{Normal}(0, 1),\\
\log(sd_{\mu,id,j})
  &= \alpha_0 + \alpha_1 x_{\text{group},j}.
\end{aligned}
\]

The matching syntax is:

```r
drmTMB(
  bf(
    y ~ x1 + (1 | id),
    sigma ~ x2,
    sd(id) ~ x_group
  ),
  family = gaussian(),
  data = dat
)
```

This model has one residual scale, `sigma_ij`, and one group-level scale,
`sd_mu_id,j`. They answer different questions. The `sigma` formula asks whether
observations are more or less variable around their fitted mean. The `sd(id)`
formula asks whether among-group differences in the mean are larger for some
types of groups.

With two additive random intercepts, the equation has two group-level scale
components:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= X_{\mu}[i, ]\beta_{\mu}
     + b_{\text{site}[i]}
     + b_{\text{observer}[i]},\\
\log(\sigma_i)
  &= X_{\sigma}[i, ]\beta_{\sigma},\\
b_{\text{site},k}
  &\sim \operatorname{Normal}(0, sd_{\mu,site}^2),\\
b_{\text{observer},l}
  &\sim \operatorname{Normal}(0, sd_{\mu,observer}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | site) + (1 | observer), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

With a simple random slope, the random effect multiplies the predictor value:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + x_{1ij} b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,x,id}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (0 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The current independent random-intercept and random-slope syntax writes the two
group-level components separately:

```r
drmTMB(
  bf(y ~ x1 + (1 | id) + (0 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

This is intentionally not the same as a correlated random intercept/slope block.
The ordinary correlated block is also implemented:

```r
drmTMB(
  bf(y ~ x1 + (1 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The intercept-slope correlation from `(1 + x1 | id)` is a group-level
correlation reported under `corpars$mu`. A labelled version is also
implemented for univariate Gaussian `mu`:

```r
drmTMB(
  bf(y ~ x1 + (1 + x1 | p | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Here `p` is retained as a group-level covariance-block label. Matching
intercept-only `mu` and `sigma` terms can now use the same label for the first
mean-scale covariance block; larger cross-formula and bivariate blocks remain
future work.

Ordinary Gaussian `mu` can also carry larger numeric grouped blocks, for
example `(1 + x1 + moisture | id)`. These q > 2 blocks estimate one SD per
coefficient and the constant correlations among those coefficients. Their SDs
are direct profile targets, but the q > 2 correlations are derived rows rather
than direct profile-interval targets. Treat them as advanced fits: they need
enough groups, repeated observations, and covariate spread to avoid weak-slope
or boundary diagnostics.

This is where ordinary language can become slippery: the model may contain
several scale quantities, but only one of them is the residual `sigma`. The
others should be named by what they scale:

- residual SD: `sigma_i`, from `sigma ~ x2`;
- site random-intercept SD in `mu`: `sd_mu_site`, from the site random
  intercept term in the location formula;
- observer random-intercept SD in `mu`: `sd_mu_observer`, from the observer
  random intercept term in the location formula;
- ID random-slope SD in `mu`: `sd_mu_x_id`, from a random-slope term such as
  `(0 + x1 | id)`;
- ID random intercept-slope correlation in `mu`: `corpars$mu`, from a
  correlated block such as `(1 + x1 | id)`;
- random-effect scale model: `sd(site)_k`, from syntax such as
  `sd(site) ~ x_group` when the target is an unlabelled Gaussian `mu` random
  intercept.

When the model includes a correlated block such as `(1 + x1 | p | id)`,
`corpairs(fit, level = "group")` returns the ordinary group-level correlation
row. For this one-slope block the row has class `mean-slope`: it asks whether
groups with higher baseline `mu` also tend to have steeper or shallower
mean-model slopes. It is not `rho12`, because `rho12` belongs only to
bivariate residual response-response coupling.

For a biological reaction-norm question, let `x1` be centred temperature and
`id` be population. The fixed slope `beta_1` is the average temperature effect.
The random-intercept SD asks how much populations differ at the centred
temperature. The random-slope SD asks how much populations differ in thermal
plasticity. The intercept-slope correlation asks whether populations with high
baseline response tend to have steeper or shallower thermal slopes:

\[
\begin{aligned}
\mu_{ij}
  &= \beta_0 + \beta_1 temperature_{ij}
     + b_{0j} + temperature_{ij} b_{1j},\\
\begin{bmatrix}b_{0j}\\ b_{1j}\end{bmatrix}
  &\sim \operatorname{Normal}\left(
    \begin{bmatrix}0\\0\end{bmatrix},
    \Sigma_{\text{population}}
  \right).
\end{aligned}
\]

The matching fitted ordinary syntax is:

```r
drmTMB(
  bf(y ~ temperature + (1 + temperature | p | population), sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

Read the fitted object as `sdpars$mu` for the baseline and slope SDs,
`corpairs(fit, class = "mean-slope")` for the group-level intercept-slope
correlation, and `profile_targets(fit)` for the direct interval target names.

The location formula models the conditional mean, `mu`, on the identity scale.
The scale formula models the residual standard deviation, `sigma`, on the log
scale:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= X_{\mu}\beta_{\mu} + b_{group[i]},\\
\log(\sigma_i)
  &= X_{\sigma}\beta_{\sigma},\\
b_{\text{group}}
  &= sd_{\text{group}} u_{\text{group}},\\
u_{\text{group}}
  &\sim \operatorname{Normal}(0, 1).
\end{aligned}
\]

This means coefficients returned by `coef(fit, "sigma")` are log-standard
deviation effects. Exponentiating a `sigma` coefficient gives the multiplicative
change in residual standard deviation for a one-unit change in that predictor.
For example, a `sigma` coefficient of `0.3` means `exp(0.3) = 1.35`, or about a
35% higher residual standard deviation.

For biological interpretation, keep the slope-like quantities separate and
translate each one to the scale on which it answers the biological question:

| Model component | Example syntax | Response-scale reading | Biological sentence |
|---|---|---|---|
| fixed mean slope | `growth ~ temperature` | `beta_temperature` is an additive change in expected growth per temperature unit | average growth changes with temperature |
| fixed residual-scale slope | `sigma ~ temperature` | `exp(gamma_temperature)` is the residual-SD ratio per temperature unit; `exp(2 * gamma_temperature)` is the residual-variance ratio | growth becomes more or less predictable along the temperature gradient |
| mean random-slope SD | `(0 + temperature | population)` or `(1 + temperature | population)` | `sd_mu_temperature_population` is the SD of population-specific temperature slopes, in growth-per-temperature units | populations differ in thermal reaction norms |
| residual-scale random-slope SD | `sigma ~ (0 + temperature | population)` | the SD is on the log-residual-SD slope scale; `exp(sd)` and `exp(-sd)` give a rough one-SD spread in residual-SD ratios per temperature unit | populations differ in how predictability changes with temperature |
| random-effect scale slope | `sd(population) ~ habitat` | `exp(alpha_habitat)` is the ratio of among-population SDs; `exp(2 * alpha_habitat)` is the corresponding among-population variance ratio | among-population differences in expected growth are larger in one habitat |

The first two rows are fixed-effect slopes. The next two rows are variance
components for group-specific slopes. The last row is a direct model for a
group-level SD. A single ecological phrase such as "temperature-dependent
variability" can point to any of these, so name the model component before
stating the biological conclusion. Current `sd(population) ~ habitat` syntax
targets an unlabelled random intercept; coefficient-specific syntax such as
`sd(population, dpar = "mu", coef = "temperature") ~ habitat` is reserved until
random-slope SD regression has likelihood code and simulation tests.
For residual-scale random slopes, keep the same boundary in mind here:
independent `sigma` slopes fit on `log(sigma)` and the unlabelled correlated
block `sigma ~ z + (1 + w | id)` is also fitted (recovery grade), while
labelled univariate residual-scale slope covariance remains planned.

### Name the biological response before the symbol

When adapting the syntax, replace placeholders such as `y`, `x1`, and `x2`
with trait names before writing the interpretation. A useful tutorial sentence
defines the measured response, the covariates, the distributional parameters,
and the biological meaning of each slope.

For example, the parrot example in the phylogenetic location-scale paper asks
whether forest habitat changes the average beak length of parrot species, the
among-species heterogeneity in beak length, or both. Written in the
non-phylogenetic Gaussian syntax used in this tutorial, the teaching version is:

\[
\begin{aligned}
\text{beak\_length}_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= \beta_0^{(\mu)}
     + \beta_{\text{mass}}^{(\mu)}\text{body\_mass}_i
     + \beta_{\text{forest}}^{(\mu)}\text{forest}_i,\\
\log(\sigma_i)
  &= \beta_0^{(\sigma)}
     + \beta_{\text{mass}}^{(\sigma)}\text{body\_mass}_i
     + \beta_{\text{forest}}^{(\sigma)}\text{forest}_i .
\end{aligned}
\]

```r
drmTMB(
  bf(beak_length ~ body_mass + forest, sigma ~ body_mass + forest),
  family = gaussian(),
  data = parrots
)
```

| Quantity | Definition in the parrot example | Biological reading |
|---|---|---|
| `beak_length_i` | measured, centred, or log-centred beak length for species `i` | the focal morphological trait |
| `body_mass_i` | centred log body mass for species `i` | a size covariate, so the beak-length comparison is not just body-size allometry |
| `forest_i` | habitat indicator for species `i`, commonly 1 for forest-living and 0 for non-forest species | the ecological contrast |
| `mu_i` | expected beak length for species `i` after conditioning on body mass and habitat | the location, or mean trait value |
| `sigma_i` | residual standard deviation of beak length for species `i` after conditioning on body mass and habitat | the scale, or remaining among-species heterogeneity |
| `beta_forest^(mu)` | forest versus non-forest contrast in `mu` | whether forest species have longer or shorter expected beaks |
| `beta_forest^(sigma)` | forest versus non-forest contrast in `log(sigma)` | whether forest species are more or less heterogeneous in beak length |

Report the scale coefficient on the scale that matches the claim:
`exp(beta_forest^(sigma))` is the residual-SD ratio for forest versus
non-forest species, while `exp(2 * beta_forest^(sigma))` is the residual-variance
ratio. A mean effect and a scale effect are different biological claims. The
first says forest habitat shifts average beak length; the second says forest
habitat changes beak-length predictability or evolvability after the mean model
has been accounted for.

The same naming rule works for behavioural or life-history examples. If the
response is `aggressiveness` and the predictor is `life_history_pace`, then
`beta_life_history_pace^(mu)` is the expected change in mean aggressiveness,
whereas `beta_life_history_pace^(sigma)` is the change in residual
predictability of aggressiveness. If the location formula also contains
`(0 + life_history_pace | population)`, the random-slope SD is a third quantity:
it measures how much populations differ in the life-history slope of
aggressiveness.

For effect-size synthesis, use the same habit of defining every symbol, but keep
the teaching path in the separate
[meta-analysis article](meta-analysis.html). There the known sampling variance
or covariance is `V`, and the fitted `sigma` model describes extra
heterogeneity after `V` has been included.

## Worked example: growth mean and predictability

Suppose an ecologist measures juvenile growth in two habitats and records the
temperature at each observation. The location question is whether mean growth
differs between habitats and changes with temperature. The scale question is
whether residual growth variability differs between habitats after accounting
for those mean effects.

For this example, the fitted model is:

\[
\begin{aligned}
\text{growth}_i &\sim \operatorname{Normal}(\mu_i, \sigma_i^2), \\
\mu_i &= \beta_0 +
        \beta_1 \mathbb{1}(\text{habitat}_i = \text{grassland}) +
        \beta_2 \text{temperature}_i, \\
\log(\sigma_i) &= \gamma_0 +
        \gamma_1 \mathbb{1}(\text{habitat}_i = \text{grassland}).
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

The code below simulates one dataset with the same structure. Simulated data
keep the vignette small and reproducible; a real analysis would replace this
block with field or laboratory measurements.

```{r location-scale-simulate}
set.seed(42)
n <- 240
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)

mu <- 8 + 1.2 * (dat$habitat == "grassland") + 0.6 * dat$temperature
sigma_true <- exp(log(0.6) + 0.55 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma_true)
```

Fit the location-scale model:

```{r location-scale-fit}
fit_growth <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

Before interpreting coefficients, check the fitted object:

```{r location-scale-check}
check_drm(fit_growth)
```

The interval target inventory is part of the same interpretation gate. For this
fixed-effect example, every coefficient is a direct profile target:

```{r location-scale-profile-targets}
profile_targets(fit_growth)[
  ,
  c("parm", "estimate", "profile_ready", "profile_note")
]
```

Now print the coefficient table:

```{r location-scale-summary}
summary(fit_growth)
```

How to read this output:

1. Rows beginning with `mu:` are mean-growth effects on the response scale.
   For example, `mu:temperature` is the expected change in mean growth for a
   one-unit increase in temperature.
2. Rows beginning with `sigma:` are log-residual-standard-deviation effects.
   For example, `sigma:habitatgrassland` is not an additive change in growth;
   it is a log-scale change in residual variability.
3. Exponentiating a `sigma` coefficient gives the multiplicative change in
   residual standard deviation.

```{r location-scale-interpret-sigma}
sigma_habitat <- coef(fit_growth, "sigma")["habitatgrassland"]
data.frame(
  coefficient = sigma_habitat,
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

In this fitted example, `exp(sigma:habitatgrassland)` is the estimated ratio of
grassland residual SD to forest residual SD. A value near 2 means that
grassland has about twice the residual SD of forest after accounting for mean
habitat and temperature effects. Because residual variance is SD squared, the
same fitted coefficient implies about four times the residual variance.

It is often clearer to report the fitted residual SDs and variances directly:

```{r location-scale-response-table}
newdat <- data.frame(
  habitat = factor(c("forest", "grassland"), levels = levels(dat$habitat)),
  temperature = 0
)

growth_report <- data.frame(
  habitat = newdat$habitat,
  fitted_mean_growth = predict(fit_growth, newdata = newdat, dpar = "mu"),
  fitted_residual_sd = predict(fit_growth, newdata = newdat, dpar = "sigma")
)
growth_report$fitted_residual_variance <- growth_report$fitted_residual_sd^2
growth_report
```

This table maps the model back to the scientific question. The mean column
summarises the location model. The residual SD column summarises the scale
model in the units of growth. The residual variance column is the
variance-facing version of the same fitted Gaussian scale model.

The fitted mean can be checked against the raw response pattern. Here the
points are observed growth values; the fitted lines and ribbons come from
`predict_parameters()` on an explicit temperature-by-habitat grid:

```{r location-scale-growth-figure, fig.cap = "Raw growth observations and fitted response-scale `mu` surfaces for the Gaussian location-scale example; ribbons are 95% Wald confidence bands from `predict_parameters()`.", fig.alt = "Scatterplot of simulated growth over temperature for forest and grassland habitats, with fitted mean growth lines and 95 percent Wald confidence bands for each habitat."}
growth_mu_grid <- prediction_grid(
  fit_growth,
  focal = c("temperature", "habitat"),
  at = list(
    temperature = seq(
      min(dat$temperature),
      max(dat$temperature),
      length.out = 80
    )
  )
)

growth_mu_surface <- predict_parameters(
  fit_growth,
  newdata = growth_mu_grid,
  dpar = "mu",
  conf.int = TRUE
)

unique(growth_mu_surface[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    dat,
    ggplot2::aes(x = temperature, y = growth, colour = habitat)
  ) +
    ggplot2::geom_point(alpha = 0.38, size = 1.25) +
    ggplot2::geom_ribbon(
      data = growth_mu_surface,
      ggplot2::aes(
        x = temperature,
        ymin = conf.low,
        ymax = conf.high,
        fill = habitat
      ),
      inherit.aes = FALSE,
      alpha = 0.18,
      colour = NA
    ) +
    ggplot2::geom_line(
      data = growth_mu_surface,
      ggplot2::aes(y = estimate),
      linewidth = 0.85
    ) +
    location_scale_habitat_scales() +
    ggplot2::labs(
      title = "Mean growth and observed scatter",
      subtitle = "Points are raw growth; ribbons are 95% Wald bands for mu",
      x = "Temperature",
      y = "Growth",
      colour = "Habitat",
      fill = "Habitat"
    ) +
    location_scale_theme() +
    ggplot2::guides(fill = "none")
}
```

The scale model is a different display. Do not put raw `growth` points on the
`sigma` axis; plot the fitted residual SD and name the interval source:

```{r location-scale-sigma-contrast-figure, fig.width = 5.8, fig.height = 2.6, fig.cap = "Fitted residual standard deviations by habitat at average temperature; horizontal intervals are 95% Wald confidence intervals requested from `predict_parameters()`.", fig.alt = "Horizontal point interval plot comparing fitted residual standard deviation for forest and grassland habitats at temperature zero, with 95 percent Wald confidence intervals."}
growth_sigma_grid <- prediction_grid(
  fit_growth,
  focal = "habitat",
  condition = list(temperature = 0)
)

growth_sigma_surface <- predict_parameters(
  fit_growth,
  newdata = growth_sigma_grid,
  dpar = "sigma",
  conf.int = TRUE
)

unique(growth_sigma_surface[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])

if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    growth_sigma_surface,
    ggplot2::aes(x = estimate, y = habitat, colour = habitat)
  ) +
    ggplot2::geom_segment(
      ggplot2::aes(x = conf.low, xend = conf.high, yend = habitat),
      linewidth = 0.8
    ) +
    ggplot2::geom_point(size = 2.8) +
    location_scale_habitat_scales() +
    ggplot2::labs(
      title = "Residual SD differs by habitat",
      subtitle = "Points are fitted sigma values; bars are 95% Wald intervals",
      x = "Fitted residual SD (sigma)",
      y = NULL,
      colour = "Habitat"
    ) +
    location_scale_theme() +
    ggplot2::guides(colour = "none")
}
```

The compact translation table below is the reporting layer Pat should be able
to read without returning to the equations:

```{r location-scale-translation-table}
growth_translation <- data.frame(
  model_piece = c(
    "fixed mean slope",
    "fixed residual-SD contrast",
    "fixed residual-variance contrast"
  ),
  fitted_term = c(
    "mu:temperature",
    "exp(sigma:habitatgrassland)",
    "exp(2 * sigma:habitatgrassland)"
  ),
  response_scale_value = c(
    unname(coef(fit_growth, "mu")["temperature"]),
    unname(exp(sigma_habitat)),
    unname(exp(2 * sigma_habitat))
  ),
  interpretation = c(
    "mean growth change per one-unit temperature increase",
    "grassland residual SD divided by forest residual SD",
    "grassland residual variance divided by forest residual variance"
  )
)
growth_translation$response_scale_value <-
  round(growth_translation$response_scale_value, 3)
growth_translation
```

A report can now say three concrete things. Mean growth increases by the fitted
`mu:temperature` slope for each one-unit temperature increase. Grassland has
the fitted residual-SD ratio in the second row after accounting for mean habitat
and temperature effects. Because Gaussian variance is `sigma^2`, the residual
variance contrast is the third row. If that third row is larger than one,
individual growth is less predictable in grassland.

The same reporting discipline carries over to hierarchical versions of the
question:

| If the biological question is... | Fit this kind of term | Report this quantity | Check before reporting |
|---|---|---|---|
| Do populations differ in thermal plasticity? | `(0 + temperature | population)` or `(1 + temperature | population)` in the `mu` formula | the random-slope SD, in growth-per-temperature units | `check_drm(fit)` and the matching `sd:mu:...temperature...` row in `profile_targets(fit)` |
| Do high-baseline populations also have steeper thermal reaction norms? | `(1 + temperature | population)` in the `mu` formula | the random-slope SD, and when the correlated block is fitted, the group-level intercept-slope correlation from `corpairs(fit, class = "mean-slope")` | `check_drm(fit)`, the matching `sd:mu:...temperature...` row, and the fitted correlation row in `profile_targets(fit)` |
| Does predictability change with temperature? | `sigma ~ temperature` | `exp(gamma_temperature)` as the residual-SD ratio, and `exp(2 * gamma_temperature)` when the paper talks about variance | `check_drm(fit)` and the `fixef:sigma:temperature` row in `profile_targets(fit)` |
| Do habitats differ in among-population variation? | `(1 | population)` plus `sd(population) ~ habitat` | `exp(alpha_habitat)` as the ratio of among-population SDs | `check_drm(fit)` and the `fixef:sd(population):...` row in `profile_targets(fit)` |

These three rows answer different biological questions. `sigma ~ temperature`
models residual variation among observations. `(0 + temperature | population)`
models population-to-population differences in the mean slope. `sd(population)
~ habitat` models the size of among-population mean differences.

## Curved responses and interactions

Ecological responses are often curved rather than straight-line. Growth may
increase from cold to moderate temperatures and then decline at high
temperatures. A second-order term is usually enough for this kind of question;
third-order polynomials are harder to explain and should be used only when the
question and data justify them.

The symbolic model below asks whether mean growth has a quadratic temperature
response and whether the temperature slope differs between habitats. It also
asks whether residual variability changes along the same temperature gradient:

\[
\begin{aligned}
\text{growth}_i &\sim \operatorname{Normal}(\mu_i, \sigma_i^2), \\
\mu_i &= \beta_0 + \beta_1 \text{temperature}_i +
        \beta_2 \text{temperature}_i^2 +
        \beta_3 \mathbb{1}(\text{habitat}_i = \text{grassland}) +
        \beta_4 \text{temperature}_i
          \mathbb{1}(\text{habitat}_i = \text{grassland}), \\
\log(\sigma_i) &= \gamma_0 + \gamma_1 \text{temperature}_i^2.
\end{aligned}
\]

The matching R formula uses ordinary R formula syntax:

```r
drmTMB(
  drm_formula(
    growth ~ habitat * temperature + I(temperature^2),
    sigma ~ I(temperature^2)
  ),
  family = gaussian(),
  data = dat_curve
)
```

The term `habitat * temperature` expands to the two main effects plus their
interaction. The term `I(temperature^2)` tells R to square `temperature`
before building the design matrix. The alternative `poly(temperature, 2)` is
also supported and can be numerically useful, but its coefficients are basis
coefficients; for applied interpretation, predicted means and residual SDs are
usually clearer than the raw polynomial coefficients.

```{r location-scale-curved-simulate}
set.seed(43)
n_curve <- 260
dat_curve <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n_curve / 2)),
  temperature = runif(n_curve, -2, 2)
)

mu_curve <- 8 +
  0.6 * dat_curve$temperature -
  0.45 * dat_curve$temperature^2 +
  0.8 * (dat_curve$habitat == "grassland") +
  0.35 * dat_curve$temperature * (dat_curve$habitat == "grassland")
sigma_curve <- exp(log(0.55) + 0.18 * dat_curve$temperature^2)
dat_curve$growth <- rnorm(n_curve, mean = mu_curve, sd = sigma_curve)
```

```{r location-scale-curved-fit}
fit_curve <- drmTMB(
  drm_formula(
    growth ~ habitat * temperature + I(temperature^2),
    sigma ~ I(temperature^2)
  ),
  family = gaussian(),
  data = dat_curve
)

summary(fit_curve)
```

For a curved model, the most useful first output is often a prediction table.
Here we compare cold, moderate, and warm temperatures in the two habitats:

```{r location-scale-curved-predictions}
curve_grid <- expand.grid(
  habitat = levels(dat_curve$habitat),
  temperature = c(-1.5, 0, 1.5)
)

data.frame(
  curve_grid,
  fitted_mean_growth = predict(fit_curve, newdata = curve_grid, dpar = "mu"),
  fitted_residual_sd = predict(fit_curve, newdata = curve_grid, dpar = "sigma")
)
```

This table is easier to explain than the individual polynomial coefficients.
The fitted mean column describes the location curve: expected growth is highest
near the middle of the temperature range in this simulated example. The fitted
residual SD column describes the scale curve: growth is less predictable at
temperature extremes when the fitted residual SD is larger.

For several predictors, base R interaction expansions work in the same way.
For example, `(temperature + moisture + canopy)^2` expands to all three main
effects and all pairwise interactions. That formula is often easier to read
than writing every interaction by hand, but it should still be tied to a clear
biological question.

If the `sigma` formula is omitted, `drmTMB` fits an intercept-only residual
standard deviation:

```{r location-scale-intercept-only}
fit0 <- drmTMB(
  drm_formula(growth ~ habitat + temperature),
  family = gaussian(),
  data = dat
)

coef(fit0, "sigma")
```

That simpler model can still estimate habitat and temperature effects on mean
growth, but it assumes one residual SD for all observations.

## Meta-analysis as Gaussian regression

Meta-analysis is the same Gaussian regression idea with known sampling
covariance added to the likelihood. For diagonal known sampling variance:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i, v_i
  &\sim \operatorname{Normal}(\mu_i, v_i + \sigma_i^2),\\
\mu_i
  &= X_{\mu}[i, ]\beta_{\mu},\\
\sigma_i
  &= \exp(X_{\sigma}[i, ]\beta_{\sigma}).
\end{aligned}
\]

with matching syntax:

```r
drmTMB(
  bf(yi ~ x1 + meta_V(V = vi), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

The preferred marker is `meta_V(V = V)`; deprecated `meta_known_V(V = V)`
remains a compatibility alias. In concrete code, `V` can be a vector such as
`vi`. Here `vi` is known sampling variance. The estimated `sigma_i` is the
additional heterogeneity standard deviation after known sampling error has been
included. In meta-analysis prose this quantity is often called `tau`, but the
`drmTMB` API keeps the name `sigma` because it is still the residual scale
parameter of the Gaussian regression.

When `V` is a full known covariance matrix, the same model is:

\[
y \mid \mu, \sigma, V
  \sim \operatorname{MVN}\left(\mu, V + \operatorname{diag}(\sigma_i^2)\right).
\]

## Current implementation caveats

- The current fitter supports `family = gaussian()` with fixed effects for
  `mu` and `sigma`, plus random intercepts, independent numeric random slopes,
  and labelled or unlabelled ordinary correlated intercept-slope blocks in
  `mu`.
- Residual-scale random intercepts and independent random slopes in `sigma`,
  such as `sigma ~ x2 + (1 | id)` and `sigma ~ x2 + (0 + w | id)`, are
  implemented for univariate Gaussian models, along with the unlabelled
  correlated intercept-slope block `sigma ~ x2 + (1 + w | id)` (recovery
  grade).
- The `sigma` formula is the observation-level residual standard deviation. It
  is not a random-effect scale component.
- Missing rows are dropped when they are incomplete for variables used by
  either the `mu` or `sigma` formula.
- Diagonal and dense full known sampling covariance via `meta_V(V = V)` is
  supported for Gaussian meta-analysis; deprecated `meta_known_V(V = V)`
  remains a compatibility alias.
- Fixed-effect bivariate Gaussian `rho12` models are implemented in the
  bivariate coscale article.
- `sd(group)` random-effect scale formulae are implemented for one or more
  distinct unlabelled Gaussian `mu` random intercepts, with group-level
  predictors that are constant within the named grouping variable.
- Labelled univariate residual-scale slope covariance blocks, slope-specific
  `sd()` targets, labelled-block `sd()` targets, same-response labelled
  location-scale slope covariance beyond the first random-intercept slice,
  multiple phylogenetic slopes,
  phylogenetic slope correlations, standalone or partial phylogenetic scale
  terms, mesh/SPDE spatial fields, multiple spatial slopes, spatial slope
  correlations, spatial `corpair()` regression, and non-Gaussian random
  effects are planned but not implemented in this Gaussian tutorial.
- Phylogenetic random effects are implemented in univariate Gaussian `mu`
  formulas as `phylo(1 | species, tree = tree)` and one numeric
  `phylo(1 + x | species, tree = tree)` slope.
- Coordinate-spatial random effects are implemented in univariate Gaussian `mu`
  formulas as `spatial(1 | site, coords = coords)` and one numeric
  `spatial(1 + x | site, coords = coords)` slope. Matching labelled
  coordinate-spatial q=2 `mu1`/`mu2` and constant q=4 location-scale blocks are
  fitted in bivariate Gaussian models; read the structural-dependence articles
  before treating those first slices as routine coverage evidence.

Multiple random-effect scale components use explicit `sd()` formulas, not
extra residual `sigma` formulas:

```r
drmTMB(
  bf(
    y ~ x1 + (1 | study) + (1 | species),
    sigma ~ x1,
    sd(study) ~ x1,
    sd(species) ~ 1
  ),
  family = gaussian(),
  data = dat
)
```
