---
title: "When variance carries signal, Part 2: location-scale-scale models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When variance carries signal, Part 2: location-scale-scale models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

A location-scale model asks whether predictors change the expected response
`mu` and the residual standard deviation `sigma`. A location-scale-scale model
adds a third submodel: predictors can also change the standard deviation of a
latent random effect. `drmTMB` writes this third submodel as
`sd(group) ~ predictors`.

Read [Part 1: location-scale models](location-scale.html) first if `mu` and
`sigma` are new to you. Read [Which scale are you
modelling?](which-scale.html) when you need to distinguish residual SD,
group-level SD, known sampling variance, and likelihood weights.

## Personality, predictability, and repeatability

Suppose an exploration score is recorded repeatedly for each individual. We
want to ask whether sex predicts three different features of behaviour:

1. the mean exploration score;
2. between-individual variation in expected scores; and
3. within-individual variation around each expected score.

For observation $j$ from individual $i$, write

\[
\begin{aligned}
y_{ij} &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{e,i}^2),\\
\mu_{ij} &= \beta_0 + \beta_1\operatorname{sex}_i + b_i,\\
b_i &\sim \operatorname{Normal}(0, \sigma_{b,i}^2),\\
\log(\sigma_{e,i}) &= \gamma_0 + \gamma_1\operatorname{sex}_i,\\
\log(\sigma_{b,i}) &= \alpha_0 + \alpha_1\operatorname{sex}_i.
\end{aligned}
\]

Here, $\operatorname{sex}_i=0$ denotes female and
$\operatorname{sex}_i=1$ denotes male. $\sigma_{b,i}$ is the
between-individual SD and $\sigma_{e,i}$ is the within-individual residual SD.
The matching `drmTMB` formula is

```r
bf(
  exploration_score ~ sex + (1 | individual),
  sigma ~ sex,
  sd(individual) ~ sex
)
```

The same predictor appears in three formulas, but its coefficients answer
three separate questions. Sex must be constant within each individual because
it is used to model `sd(individual)`.

## Simulate and fit the three submodels

The example below gives females and males different means, between-individual
SDs, and within-individual SDs. It is intentionally simple enough that the
three sources of signal remain visible.

```{r personality-simulate}
set.seed(20260721)
n_individual <- 80L
n_each <- 6L

individual_info <- data.frame(
  individual = factor(seq_len(n_individual)),
  sex = factor(
    rep(c("female", "male"), each = n_individual / 2),
    levels = c("female", "male")
  )
)

mean_by_sex <- c(female = 0.35, male = 0.70)
between_sd_by_sex <- c(female = 0.65, male = 0.40)
within_sd_by_sex <- c(female = 0.35, male = 0.60)

individual_effect <- stats::rnorm(
  n_individual,
  sd = between_sd_by_sex[individual_info$sex]
)

personality <- individual_info[rep(seq_len(n_individual), each = n_each), ]
personality$exploration_score <-
  mean_by_sex[personality$sex] +
  individual_effect[as.integer(personality$individual)] +
  stats::rnorm(
    nrow(personality),
    sd = within_sd_by_sex[personality$sex]
  )
```

```{r personality-fit}
fit_personality <- drmTMB(
  bf(
    exploration_score ~ sex + (1 | individual),
    sigma ~ sex,
    sd(individual) ~ sex
  ),
  family = gaussian(),
  data = personality
)

round(coef(fit_personality, "mu"), 3)
round(coef(fit_personality, "sigma"), 3)
round(coef(fit_personality, "sd(individual)"), 3)
```

The `sigma` and `sd(individual)` coefficients use log-SD links. For example,
exponentiating the male coefficient from `sigma` gives the male-to-female
ratio of within-individual SDs. Exponentiating the corresponding
`sd(individual)` coefficient gives the male-to-female ratio of
between-individual SDs.

```{r personality-data-figure, echo = FALSE, fig.width = 7, fig.height = 4.1, fig.cap = "Repeated exploration scores for females and males. Grey points are observations, blue ticks are individual means, and the vermillion line is the fitted sex-specific mean. The spread of individual means represents between-individual variation; scatter around each individual mean represents within-individual variation. Both panels use the same vertical scale.", fig.alt = "Two panels show repeated exploration scores for female and male individuals. Grey observations cluster around blue individual-mean ticks, while a vermillion horizontal line marks the fitted sex-specific mean. The female panel has wider spread among individual means, while the male panel has more scatter within individuals."}
individual_summary <- stats::aggregate(
  exploration_score ~ individual + sex,
  data = personality,
  FUN = mean
)
individual_summary <- individual_summary[
  order(individual_summary$sex, individual_summary$exploration_score),
]
individual_summary$display_id <- ave(
  individual_summary$exploration_score,
  individual_summary$sex,
  FUN = seq_along
)

plot_data <- merge(
  personality,
  individual_summary[c("individual", "sex", "display_id")],
  by = c("individual", "sex"),
  sort = FALSE
)

sex_grid <- data.frame(
  sex = factor(c("female", "male"), levels = levels(personality$sex))
)
sex_grid$fitted_mean <- predict(
  fit_personality,
  newdata = sex_grid,
  dpar = "mu"
)

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = display_id, y = exploration_score)
) +
  ggplot2::geom_point(
    position = ggplot2::position_jitter(width = 0.08, height = 0, seed = 1),
    colour = "grey55",
    alpha = 0.42,
    size = 0.8
  ) +
  ggplot2::geom_segment(
    data = individual_summary,
    ggplot2::aes(
      x = display_id - 0.26,
      xend = display_id + 0.26,
      y = exploration_score,
      yend = exploration_score
    ),
    inherit.aes = FALSE,
    colour = "#0072B2",
    linewidth = 0.65
  ) +
  ggplot2::geom_hline(
    data = sex_grid,
    ggplot2::aes(yintercept = fitted_mean),
    inherit.aes = FALSE,
    colour = "#D55E00",
    linewidth = 0.8
  ) +
  ggplot2::geom_text(
    data = sex_grid,
    ggplot2::aes(
      x = Inf,
      y = fitted_mean,
      label = "fitted sex mean"
    ),
    inherit.aes = FALSE,
    colour = "#D55E00",
    hjust = 1.05,
    vjust = -0.6,
    size = 3
  ) +
  ggplot2::facet_wrap(
    ~sex,
    nrow = 1,
    labeller = ggplot2::as_labeller(c(female = "Female", male = "Male"))
  ) +
  ggplot2::labs(
    x = "Individuals, ordered within sex by their observed mean",
    y = "Exploration score"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank()
  )
```

## Read the fitted scales and calculate repeatability

Predictions with `newdata` are population-level values. They give the three
fitted quantities for each sex directly on their natural scales.

```{r personality-components}
sex_grid$mean_score <- predict(
  fit_personality, newdata = sex_grid, dpar = "mu"
)
sex_grid$between_individual_sd <- predict(
  fit_personality, newdata = sex_grid, dpar = "sd(individual)"
)
sex_grid$within_individual_sd <- predict(
  fit_personality, newdata = sex_grid, dpar = "sigma"
)

sex_grid$repeatability <- with(
  sex_grid,
  between_individual_sd^2 /
    (between_individual_sd^2 + within_individual_sd^2)
)

repeatability_table <- sex_grid[c(
  "sex",
  "mean_score",
  "between_individual_sd",
  "within_individual_sd",
  "repeatability"
)]
repeatability_table[-1] <- lapply(
  repeatability_table[-1],
  round,
  digits = 3
)
repeatability_table
```

In behavioural ecology, this intraclass correlation is usually called
**repeatability**. For sex $s$,

\[
R_s = \frac{\sigma_{b,s}^2}
{\sigma_{b,s}^2 + \sigma_{e,s}^2}.
\]

Repeatability is derived from the two fitted scale submodels; it is not a
fourth formula. The calculation above is a point estimate. It does not supply
an uncertainty interval for this nonlinear ratio.

```{r personality-component-figure, echo = FALSE, fig.width = 7, fig.height = 2.9, fig.cap = "Model-implied values for females and males. Panels show the expected exploration score, between-individual SD from `sd(individual)`, and within-individual residual SD from `sigma`. Lines aid comparison and are not uncertainty intervals.", fig.alt = "Three small panels compare fitted female and male values. The expected score and within-individual standard deviation are higher for males, while the between-individual standard deviation is higher for females. No uncertainty intervals are shown."}
component_surface <- rbind(
  data.frame(
    sex = sex_grid$sex,
    component = "Expected score",
    estimate = sex_grid$mean_score
  ),
  data.frame(
    sex = sex_grid$sex,
    component = "Between-individual SD",
    estimate = sex_grid$between_individual_sd
  ),
  data.frame(
    sex = sex_grid$sex,
    component = "Within-individual SD",
    estimate = sex_grid$within_individual_sd
  )
)
component_surface$component <- factor(
  component_surface$component,
  levels = c(
    "Expected score",
    "Between-individual SD",
    "Within-individual SD"
  )
)

ggplot2::ggplot(
  component_surface,
  ggplot2::aes(x = sex, y = estimate, group = 1)
) +
  ggplot2::geom_line(colour = "grey65", linewidth = 0.55) +
  ggplot2::geom_point(colour = "#0072B2", size = 2.4) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f", estimate)),
    vjust = -0.8,
    colour = "grey20",
    size = 3.1
  ) +
  ggplot2::facet_wrap(~component, scales = "free_y", nrow = 1) +
  ggplot2::expand_limits(y = 0) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.02, 0.18))
  ) +
  ggplot2::scale_x_discrete(labels = c(female = "Female", male = "Male")) +
  ggplot2::labs(x = NULL, y = "Fitted value") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

## A short phylogenetic extension

The same three-part logic applies when the latent location effect follows a
phylogeny. Start from the familiar constant-scale model

\[
\begin{aligned}
y_i &= \mu_i + a_i + e_i,\\
\mu_i &= \beta_0 + \beta_1 T_i,\\
\mathbf a &\sim \operatorname{MVN}(\mathbf 0, \sigma_a^2 A),\\
e_i &\sim \operatorname{Normal}(0, \sigma_e^2),
\end{aligned}
\]

where $A$ is the phylogenetic correlation matrix and $T_i$ is temperature. A
location-scale-scale extension lets both SDs change linearly with temperature:

\[
\log(\sigma_{e,i}) = \gamma_0 + \gamma_1 T_i,
\qquad
\log(\sigma_{a,i}) = \alpha_0 + \alpha_1 T_i.
\]

The matching syntax is deliberately simple:

```r
fit_phylo_lss <- drmTMB(
  bf(
    trait ~ temperature + phylo(1 | species, tree = tree),
    sigma ~ temperature,
    sd(species, level = "phylogenetic") ~ temperature
  ),
  family = gaussian(),
  data = dat
)
```

| Formula | Model quantity | Interpretation |
|---|---|---|
| `trait ~ temperature + phylo(...)` | $\mu_i$ and $a_i$ | expected trait and phylogenetically correlated location deviation |
| `sigma ~ temperature` | $\sigma_{e,i}$ | independent SD |
| `sd(species, level = "phylogenetic") ~ temperature` | $\sigma_{a,i}$ | SD of the phylogenetic location deviation |

The scalar covariance $\sigma_a^2 A$ is the constant-SD starting point. The
third formula generalizes it by allowing the phylogenetic-effect SD to vary
among species with temperature. The older formula spelling
`sd_phylo(species) ~ temperature` is soft-deprecated; fitted objects retain
the output label `sd_phylo(species)` for extractor compatibility.

With repeated observations within species, `sigma` is the within-species
residual SD. With one response row per species, it is the independent
non-phylogenetic species-level deviation. It should not automatically be
described as measurement error.

## Current boundaries

- Ordinary `sd(group) ~ predictors` is implemented for distinct unlabelled
  Gaussian `mu` random intercepts. The matching `(1 | group)` term must appear
  in the location formula, and scale predictors must be constant within group.
- `sd(group, level = "phylogenetic") ~ predictors` targets the location
  phylogenetic effect introduced by `phylo()` in `mu`; predictors must be
  constant within species.
- This article teaches the Gaussian route. Check narrow non-Gaussian cells in
  [What can I fit today?](model-map.html) rather than generalizing from it.
- Random effects on the right-hand side of an `sd()` formula and generic
  direct-SD levels for `spatial()`, `animal()`, and `relmat()` remain separate
  implementation and validation questions.
- Use `check_drm()`, inspect the terminal gradient and Hessian diagnostics,
  and retain failed fits. A clean optimizer code alone does not establish that
  every SD surface is well identified.

For broader phylogenetic syntax, continue to [Phylogenetic mixed
models](phylogenetic-models.html). For prediction tables and the distinction
between `sigma` and `sd(group)`, return to [Which scale are you
modelling?](which-scale.html).

## Reference

Nakagawa, S., Mizuno, A., Williams, C., Lagisz, M., Yang, Y., and Drobniak,
S. M. (2025). Quantifying macro-evolutionary patterns of trait mean and
variance with phylogenetic location-scale models. *Methods in Ecology and
Evolution*. [doi:10.1111/2041-210X.70160](https://doi.org/10.1111/2041-210X.70160).
