---
title: "When variance carries signal, Part 2: location-scale-scale models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When variance carries signal, Part 2: location-scale-scale models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
has_ape <- requireNamespace("ape", quietly = TRUE)
```

A location-scale model asks whether predictors change the expected response
`mu` and the residual standard deviation `sigma`. A location-scale-scale model
adds a second scale submodel: predictors can also change the standard
deviation of a latent random effect. `drmTMB` writes that third submodel as
`sd(group) ~ predictors`.

Read [Part 1: location-scale models](location-scale.html) first if `mu` and
`sigma` are new to you. Read [Which scale are you
modelling?](which-scale.html) when you need to distinguish residual SD,
group-level SD, known sampling variance, and likelihood weights.

## Three submodels, three questions

For an ordinary grouped Gaussian model,

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}, b_j
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \mathbf{x}_{ij}^{\mathsf T}\boldsymbol\beta + b_j,\\
\log(\sigma_{ij})
  &= \mathbf{z}_{ij}^{\mathsf T}\boldsymbol\gamma,\\
b_j
  &= sd_{\mu,j} u_j,\qquad u_j \sim \operatorname{Normal}(0,1),\\
\log(sd_{\mu,j})
  &= \mathbf{w}_{j}^{\mathsf T}\boldsymbol\alpha.
\end{aligned}
\]

The matching syntax is:

```r
drmTMB(
  bf(
    y ~ x + (1 | group),
    sigma ~ z,
    sd(group) ~ w_group
  ),
  family = gaussian(),
  data = dat
)
```

| Submodel | Quantity | Scientific question |
|---|---|---|
| `y ~ x + (1 | group)` | `mu` | How does `x` change the expected response, and how do groups deviate from it? |
| `sigma ~ z` | residual SD | Does `z` change variation of observations around their fitted mean? |
| `sd(group) ~ w_group` | group random-effect SD | Does `w_group` change the spread of group-level mean deviations? |

`sigma` and `sd(group)` are not interchangeable names. They scale different
parts of the model. The right-hand-side variables in `sd(group) ~ w_group`
must be constant within each level of `group` after incomplete rows have been
removed.

## A grouped example

Suppose growth is measured repeatedly in populations. Temperature may change
mean growth and residual predictability, while habitat may change how widely
population means differ.

```{r ordinary-sd-simulate}
set.seed(20260721)
n_population <- 28L
n_each <- 8L
population_info <- data.frame(
  population = factor(seq_len(n_population)),
  habitat = rep(c("forest", "grassland"), length.out = n_population)
)
fish <- population_info[rep(seq_len(n_population), each = n_each), ]
fish$temperature <- stats::rnorm(nrow(fish))

sd_population <- exp(
  -0.45 + 0.55 * (population_info$habitat == "grassland")
)
population_effect <- stats::rnorm(n_population, sd = sd_population)
residual_sd <- exp(-0.7 + 0.20 * fish$temperature)
fish$growth <- 0.8 + 0.6 * fish$temperature +
  population_effect[as.integer(fish$population)] +
  stats::rnorm(nrow(fish), sd = residual_sd)
```

```{r ordinary-sd-fit}
fit_group_sd <- drmTMB(
  bf(
    growth ~ temperature + (1 | population),
    sigma ~ temperature,
    sd(population) ~ habitat
  ),
  family = gaussian(),
  data = fish
)

round(coef(fit_group_sd, "sigma"), 3)
round(coef(fit_group_sd, "sd(population)"), 3)
```

Both coefficient vectors use log-SD links. For example,
`exp(alpha_habitat)` is the ratio of among-population SDs, and
`exp(2 * alpha_habitat)` is the corresponding variance ratio. Those ratios do
not describe residual variability among observations; that remains the job of
`sigma`.

Use the same grammar when the location model has several additive random
intercepts:

```r
bf(
  y ~ x + (1 | study) + (1 | species),
  sigma ~ z,
  sd(study) ~ study_type,
  sd(species) ~ habitat
)
```

Each `sd()` formula targets its named random intercept. It does not create a
new random effect by itself: the matching `(1 | group)` term must be present
in the location formula.

## Phylogenetic location-scale-scale models

Species-level traits often contain two kinds of variation: a phylogenetic
component shared according to ancestry and an independent component that does
not follow the tree. Let $A$ be the phylogenetic covariance matrix, let
$D_a = \operatorname{diag}(\sigma_{a,1},\ldots,\sigma_{a,n})$, and let
$D_e = \operatorname{diag}(\sigma_{e,1},\ldots,\sigma_{e,n})$. A Gaussian
phylogenetic location-scale-scale model is

\[
\begin{aligned}
\mathbf y
  &= \boldsymbol\mu + \mathbf a + \mathbf e,\\
\boldsymbol\mu
  &= X_{\mu}\boldsymbol\beta,\\
\log(\boldsymbol\sigma_e)
  &= X_e\boldsymbol\gamma,\\
\log(\boldsymbol\sigma_a)
  &= X_a\boldsymbol\alpha,\\
\mathbf a
  &\sim \operatorname{MVN}(\mathbf 0,D_a A D_a),\\
\mathbf e
  &\sim \operatorname{MVN}(\mathbf 0,D_e I D_e).
\end{aligned}
\]

After integrating over both deviations,

\[
\mathbf y \sim \operatorname{MVN}
\left(
  \boldsymbol\mu,
  D_a A D_a + D_e I D_e
\right).
\]

The three formula components map directly to these equations:

```r
drmTMB(
  bf(
    trait ~ climate + phylo(1 | species, tree = tree),
    sigma ~ climate,
    sd(species, level = "phylogenetic") ~ climate
  ),
  family = gaussian(),
  data = dat
)
```

| Formula | Model quantity | Interpretation |
|---|---|---|
| `trait ~ climate + phylo(...)` | $\mu_i$ and $a_i$ | expected trait and phylogenetically correlated location deviation |
| `sigma ~ climate` | $\sigma_{e,i}$ | independent, non-phylogenetic SD |
| `sd(species, level = "phylogenetic") ~ climate` | $\sigma_{a,i}$ | SD of the phylogenetic location deviation |

With repeated observations within species, `sigma` is the within-species
residual SD. With one response row per species, as in many comparative
analyses, the same statistical component is the independent
non-phylogenetic species-level deviation. It should not automatically be
described as measurement error.

The generic formula spelling is `sd(species, level = "phylogenetic")`. Fitted
objects currently retain the internal output label `sd_phylo(species)` for
extractor compatibility. The older formula spelling
`sd_phylo(species) ~ climate` is soft-deprecated and should not be used in new
analyses.

## A small phylogenetic example

This small build-time example uses repeated observations within species so the
two SD surfaces are easier to distinguish. We simulate climate-dependent
within-species and phylogenetic SDs, then fit the three submodels together. The
next section gives the corresponding one-row-per-species analysis syntax.

```{r phylogenetic-sd-simulate, eval = has_ape}
set.seed(20260722)
n_species <- 32L
n_each <- 10L
tip <- paste0("sp", seq_len(n_species))
tree <- ape::rcoal(n_species, tip.label = tip)
A <- drmTMB:::drm_phylo_tip_covariance(tree)

climate_species <- as.numeric(scale(seq(-1.5, 1.5, length.out = n_species)))
sigma_a_species <- exp(-0.35 + 0.55 * climate_species)
base_phylo <- as.vector(t(chol(A)) %*% stats::rnorm(n_species))
phylo_effect <- stats::setNames(sigma_a_species * base_phylo, tip)

species <- rep(tip, each = n_each)
climate <- climate_species[match(species, tip)]
within_species_x <- rep(seq(-1, 1, length.out = n_each), n_species)
sigma_e <- exp(-0.90 - 0.20 * climate)
mu <- 0.3 - 0.25 * climate + 0.40 * within_species_x
trait <- mu + phylo_effect[species] + stats::rnorm(length(species), sd = sigma_e)
bird <- data.frame(trait, climate, within_species_x, species)
```

```{r phylogenetic-sd-fit, eval = has_ape}
fit_phylo_sd <- drmTMB(
  bf(
    trait ~ climate + within_species_x + phylo(1 | species, tree = tree),
    sigma ~ climate,
    sd(species, level = "phylogenetic") ~ climate
  ),
  family = gaussian(),
  data = bird,
  control = drm_control(
    optimizer = list(eval.max = 1000L, iter.max = 1000L)
  )
)

round(coef(fit_phylo_sd, "sigma"), 3)
round(coef(fit_phylo_sd, "sd_phylo(species)"), 3)
```

The `sigma` coefficients describe changes in $\log(\sigma_e)$. The
`sd_phylo(species)` output coefficients describe changes in
$\log(\sigma_a)$. A predictor can therefore increase one source of variation
while decreasing the other.

## The same predictors in all three submodels

An ecological analysis may use the same climate basis in all three submodels.
For example, a linear-plus-quadratic temperature and precipitation model is:

```r
fit_climate <- drmTMB(
  bf(
    trait ~
      1 + temp_z + I(temp_z^2) + precip_z + I(precip_z^2) +
      phylo(1 | spp_id, tree = tree),

    sigma ~
      1 + temp_z + I(temp_z^2) + precip_z + I(precip_z^2),

    sd(spp_id, level = "phylogenetic") ~
      1 + temp_z + I(temp_z^2) + precip_z + I(precip_z^2)
  ),
  family = gaussian(),
  data = dat
)
```

The repeated predictor names create separate coefficient vectors; they do not
force the three effects to be equal. A theory-motivated covariate can also
belong to only one submodel. For example, body mass or a temperature-by-body-
mass interaction may enter the location formula without being added to either
scale formula.

## Deriving the phylogenetic variance proportion

At species $i$, the phylogenetic share of the modelled random-plus-independent
variance is

\[
H^2_{\mathrm{phylo},i}
  = \frac{\sigma_{a,i}^2}
  {\sigma_{a,i}^2 + \sigma_{e,i}^2}.
\]

This is a derived quantity, not a fourth formula. The label describes a
phylogenetic variance proportion; it should not automatically be interpreted
as narrow-sense heritability.

```{r phylogenetic-variance-proportion, eval = has_ape}
sigma_e_hat <- predict(fit_phylo_sd, dpar = "sigma")
sigma_a_hat <- predict(fit_phylo_sd, dpar = "sd_phylo(species)")
species_order <- names(sigma_a_hat)
sigma_e_by_species <- tapply(sigma_e_hat, bird$species, mean)
climate_by_species <- tapply(bird$climate, bird$species, unique)

variance_partition <- data.frame(
  species = species_order,
  climate = unname(climate_by_species[species_order]),
  sigma_e = unname(sigma_e_by_species[species_order]),
  sigma_a = unname(sigma_a_hat)
)
variance_partition$H2_phylo <- with(
  variance_partition,
  sigma_a^2 / (sigma_a^2 + sigma_e^2)
)
head(variance_partition[order(variance_partition$climate), ], 4)
```

Pointwise intervals for this ratio must propagate the joint covariance of the
`sigma` and phylogenetic-`sd()` coefficients. Do not combine two separate
coefficient intervals as though the scale surfaces were independent. When an
analysis is repeated over posterior trees, tree-to-tree uncertainty adds
another pooling layer beyond the single-tree model shown here.

## Current boundaries

- Ordinary `sd(group) ~ predictors` is implemented for distinct unlabelled
  Gaussian `mu` random intercepts. It does not yet target a named random slope
  or labelled covariance block.
- `sd(group, level = "phylogenetic") ~ predictors` targets the location
  phylogenetic effect introduced by `phylo()` in `mu`. Predictors must be
  constant within species.
- This article teaches the Gaussian route. Narrow non-Gaussian evidence cells
  should be checked in [What can I fit today?](model-map.html) rather than
  generalized from this example.
- A direct phylogenetic `sd()` surface combined with an additional
  phylogenetic random effect inside `sigma` is a different model and is not
  supported by this route.
- Random effects on the right-hand side of an `sd()` formula and generic
  direct-SD levels for `spatial()`, `animal()`, and `relmat()` require separate
  implementation and validation.
- Use `check_drm()`, inspect the terminal gradient and Hessian diagnostics,
  and retain failed fits. A clean optimizer code alone does not establish that
  every SD surface is well identified.

For the broader phylogenetic syntax, continue to
[Phylogenetic mixed models](phylogenetic-models.html). For prediction tables
and the difference between `sigma` and ordinary `sd(group)`, return to
[Which scale are you modelling?](which-scale.html).

## Reference

Nakagawa, S., Mizuno, A., Williams, C., Lagisz, M., Yang, Y., and Drobniak,
S. M. (2025). Quantifying macro-evolutionary patterns of trait mean and
variance with phylogenetic location-scale models. *Methods in Ecology and
Evolution*. [doi:10.1111/2041-210X.70160](https://doi.org/10.1111/2041-210X.70160).
