---
title: "When variance carries signal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When variance carries signal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

This tutorial starts with a biological question: do two habitats differ only in
mean growth, or does one habitat also make growth less predictable? In a
Gaussian location-scale model, the location part models the expected response
`mu`, while the scale part models the residual standard deviation `sigma`.
That residual scale is not just a nuisance parameter; it can be the scientific
answer when the question is about individual variability, predictability, or
heterogeneous heterogeneity.

If you are fitting your first model, start with the
[worked growth example](#worked-example-growth-mean-and-predictability), then
return to the syntax overview below when you want to adapt the formula.

The first implemented `drmTMB` model class is Gaussian location-scale
regression with fixed effects, optional random effects in the location formula,
residual-scale random intercepts and independent random slopes in the `sigma`
formula, and random-effect scale formulas such as `sd(id) ~ x_group`. The same
pattern will be used throughout the documentation: write the model symbolically
first, then show the matching R syntax, fit the model, and interpret the fitted
output.

Current syntax:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Random-intercept location syntax:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

Simple random-slope location syntax:

```r
drmTMB(
  bf(y ~ x1 + (0 + x1 | id), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

Residual-scale random-effect syntax:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2 + (1 | id) + (0 + w | id)),
  family = gaussian(),
  data = dat
)
```

The companion article "Which scale are you modelling?" gives a side-by-side
guide to `sigma ~ x`, `sigma ~ x + (1 | id)`,
`sigma ~ x + (0 + w | id)`, `sd(id) ~ x_group`,
random-slope correlations, and bivariate residual `rho12`.

## Model equations and matching R syntax

The basic discipline is simple: each estimated parameter gets its own linear
predictor. For Gaussian location-scale regression those parameters are the
mean, `mu`, and the residual standard deviation, `sigma`.

Throughout this article, `Normal(a, b)` uses variance as the second argument.
Thus `Normal(mu_i, sigma_i^2)` means a Gaussian distribution with mean `mu_i`
and standard deviation `sigma_i`.

For the fixed-effect Gaussian location-scale model, the symbolic model is:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= \beta_0 + \beta_1 x_{1i},\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 x_{2i}.
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The location formula `y ~ x1` defines the design matrix for the `beta`
coefficients. The scale formula `sigma ~ x2` defines the design matrix for the
`gamma` coefficients. The log link makes `sigma_i` positive:

\[
\sigma_i = \exp(\gamma_0 + \gamma_1 x_{2i}).
\]

The same model can be written more compactly as:

\[
\begin{aligned}
\eta_{\mu i} &= X_{\mu}[i, ]\beta_{\mu},\\
\eta_{\sigma i} &= X_{\sigma}[i, ]\beta_{\sigma},\\
\mu_i &= \eta_{\mu i},\\
\sigma_i &= \exp(\eta_{\sigma i}).
\end{aligned}
\]

The mapping is:

| Symbol | Meaning | R syntax source |
|---|---|---|
| `y_i` | observed response | left-hand side of `y ~ x1` |
| `X_mu` | design matrix for the mean | right-hand side of `y ~ x1` |
| `beta_mu` | mean coefficients | `coef(fit, "mu")` |
| `mu_i` | fitted mean | `fitted(fit)` or `predict(fit, dpar = "mu")` |
| `X_sigma` | design matrix for the residual scale | right-hand side of `sigma ~ x2` |
| `beta_sigma` | log-scale coefficients | `coef(fit, "sigma")` |
| `sigma_i` | residual standard deviation | `sigma(fit)` or `predict(fit, dpar = "sigma")` |

With a random intercept in the location part, the model becomes:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,id}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Here `sigma_ij` is the residual standard deviation. The random-intercept scale
`sd_mu_id` is a separate group-level standard deviation. In implemented
double-hierarchical scale models, `sd(id) ~ x_group` exposes this group-level
scale component by name; it should not be confused with residual `sigma`.

Residual-scale random intercepts enter a different part of the model:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= X_{\mu}[ij, ]\beta_{\mu} + b_j,\\
\log(\sigma_{ij})
  &= X_{\sigma}[ij, ]\beta_{\sigma} + a_j,\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,id}^2),\\
a_j
  &\sim \operatorname{Normal}(0, sd_{\sigma,id}^2).
\end{aligned}
\]

The matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | id), sigma ~ x2 + (1 | id) + (0 + w | id)),
  family = gaussian(),
  data = dat
)
```

Here `a_j` is residual-scale heterogeneity, or group-to-group variation in
within-group variability. It is still not the same quantity as `sd_mu_id`, the
standard deviation of the `mu` random intercepts.

A double-hierarchical scale formula targets the random-effect standard
deviation itself. For example:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}, b_j
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &= sd_{\mu,id,j} u_j,\\
u_j
  &\sim \operatorname{Normal}(0, 1),\\
\log(sd_{\mu,id,j})
  &= \alpha_0 + \alpha_1 x_{\text{group},j}.
\end{aligned}
\]

The matching syntax is:

```r
drmTMB(
  bf(
    y ~ x1 + (1 | id),
    sigma ~ x2,
    sd(id) ~ x_group
  ),
  family = gaussian(),
  data = dat
)
```

This model has one residual scale, `sigma_ij`, and one group-level scale,
`sd_mu_id,j`. They answer different questions. The `sigma` formula asks whether
observations are more or less variable around their fitted mean. The `sd(id)`
formula asks whether among-group differences in the mean are larger for some
types of groups.

With two additive random intercepts, the equation has two group-level scale
components:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= X_{\mu}[i, ]\beta_{\mu}
     + b_{\text{site}[i]}
     + b_{\text{observer}[i]},\\
\log(\sigma_i)
  &= X_{\sigma}[i, ]\beta_{\sigma},\\
b_{\text{site},k}
  &\sim \operatorname{Normal}(0, sd_{\mu,site}^2),\\
b_{\text{observer},l}
  &\sim \operatorname{Normal}(0, sd_{\mu,observer}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (1 | site) + (1 | observer), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

With a simple random slope, the random effect multiplies the predictor value:

\[
\begin{aligned}
y_{ij} \mid \mu_{ij}, \sigma_{ij}
  &\sim \operatorname{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij}
  &= \beta_0 + \beta_1 x_{1ij} + x_{1ij} b_j,\\
\log(\sigma_{ij})
  &= \gamma_0 + \gamma_1 x_{2ij},\\
b_j
  &\sim \operatorname{Normal}(0, sd_{\mu,x,id}^2).
\end{aligned}
\]

and the matching syntax is:

```r
drmTMB(
  bf(y ~ x1 + (0 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The current independent random-intercept and random-slope syntax writes the two
group-level components separately:

```r
drmTMB(
  bf(y ~ x1 + (1 | id) + (0 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

This is intentionally not the same as a correlated random intercept/slope block.
The ordinary correlated block is also implemented:

```r
drmTMB(
  bf(y ~ x1 + (1 + x1 | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

The intercept-slope correlation from `(1 + x1 | id)` is a group-level
correlation reported under `corpars$mu`. A labelled version is also
implemented for univariate Gaussian `mu`:

```r
drmTMB(
  bf(y ~ x1 + (1 + x1 | p | id), sigma ~ x2),
  family = gaussian(),
  data = dat
)
```

Here `p` is retained as a group-level covariance-block label. Matching
intercept-only `mu` and `sigma` terms can now use the same label for the first
mean-scale covariance block; larger cross-formula and bivariate blocks remain
future work.

This is where ordinary language can become slippery: the model may contain
several scale quantities, but only one of them is the residual `sigma`. The
others should be named by what they scale:

- residual SD: `sigma_i`, from `sigma ~ x2`;
- site random-intercept SD in `mu`: `sd_mu_site`, from the site random
  intercept term in the location formula;
- observer random-intercept SD in `mu`: `sd_mu_observer`, from the observer
  random intercept term in the location formula;
- ID random-slope SD in `mu`: `sd_mu_x_id`, from a random-slope term such as
  `(0 + x1 | id)`;
- ID random intercept-slope correlation in `mu`: `corpars$mu`, from a
  correlated block such as `(1 + x1 | id)`;
- random-effect scale model: `sd(site)_k`, from syntax such as
  `sd(site) ~ x_group` when the target is an unlabelled Gaussian `mu` random
  intercept.

When the model includes a correlated block such as `(1 + x1 | p | id)`,
`corpairs(fit, level = "group")` returns the ordinary group-level correlation
row. For this one-slope block the row has class `mean-slope`: it asks whether
groups with higher baseline `mu` also tend to have steeper or shallower
mean-model slopes. It is not `rho12`, because `rho12` belongs only to
bivariate residual response-response coupling.

For a biological reaction-norm question, let `x1` be centred temperature and
`id` be population. The fixed slope `beta_1` is the average temperature effect.
The random-intercept SD asks how much populations differ at the centred
temperature. The random-slope SD asks how much populations differ in thermal
plasticity. The intercept-slope correlation asks whether populations with high
baseline response tend to have steeper or shallower thermal slopes:

\[
\begin{aligned}
\mu_{ij}
  &= \beta_0 + \beta_1 temperature_{ij}
     + b_{0j} + temperature_{ij} b_{1j},\\
\begin{bmatrix}b_{0j}\\ b_{1j}\end{bmatrix}
  &\sim \operatorname{Normal}\left(
    \begin{bmatrix}0\\0\end{bmatrix},
    \Sigma_{\text{population}}
  \right).
\end{aligned}
\]

The matching fitted ordinary syntax is:

```r
drmTMB(
  bf(y ~ temperature + (1 + temperature | p | population), sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

Read the fitted object as `sdpars$mu` for the baseline and slope SDs,
`corpairs(fit, class = "mean-slope")` for the group-level intercept-slope
correlation, and `profile_targets(fit)` for the direct interval target names.

The location formula models the conditional mean, `mu`, on the identity scale.
The scale formula models the residual standard deviation, `sigma`, on the log
scale:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= X_{\mu}\beta_{\mu} + b_{group[i]},\\
\log(\sigma_i)
  &= X_{\sigma}\beta_{\sigma},\\
b_{\text{group}}
  &= sd_{\text{group}} u_{\text{group}},\\
u_{\text{group}}
  &\sim \operatorname{Normal}(0, 1).
\end{aligned}
\]

This means coefficients returned by `coef(fit, "sigma")` are log-standard
deviation effects. Exponentiating a `sigma` coefficient gives the multiplicative
change in residual standard deviation for a one-unit change in that predictor.
For example, a `sigma` coefficient of `0.3` means `exp(0.3) = 1.35`, or about a
35% higher residual standard deviation.

For biological interpretation, keep the slope-like quantities separate:

| Quantity | Model term | Biological claim |
|---|---|---|
| fixed mean slope | `y ~ temperature` | average growth changes with temperature |
| fixed residual-scale slope | `sigma ~ temperature` | residual predictability changes with temperature |
| mean random-slope SD | `(0 + temperature | population)` or `(1 + temperature | population)` | populations differ in thermal reaction norms |
| residual-scale random-slope SD | `sigma ~ (0 + temperature | population)` | populations differ in how residual variability changes with temperature |
| random-effect scale slope | `sd(population) ~ habitat` | the among-population SD in expected growth differs by habitat |

The first two are fixed-effect slopes. The next two are variance components for
group-specific slopes. The last is a direct model for a group-level SD. A
single ecological phrase such as "temperature-dependent variability" can point
to any of these, so the tutorial should name the model component before naming
the biological conclusion.

## Worked example: growth mean and predictability

Suppose an ecologist measures juvenile growth in two habitats and records the
temperature at each observation. The location question is whether mean growth
differs between habitats and changes with temperature. The scale question is
whether residual growth variability differs between habitats after accounting
for those mean effects.

For this example, the fitted model is:

\[
\begin{aligned}
\text{growth}_i &\sim \operatorname{Normal}(\mu_i, \sigma_i^2), \\
\mu_i &= \beta_0 +
        \beta_1 \mathbb{1}(\text{habitat}_i = \text{grassland}) +
        \beta_2 \text{temperature}_i, \\
\log(\sigma_i) &= \gamma_0 +
        \gamma_1 \mathbb{1}(\text{habitat}_i = \text{grassland}).
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

The code below simulates one dataset with the same structure. Simulated data
keep the vignette small and reproducible; a real analysis would replace this
block with field or laboratory measurements.

```{r location-scale-simulate}
set.seed(42)
n <- 240
dat <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n / 2)),
  temperature = rnorm(n)
)

mu <- 8 + 1.2 * (dat$habitat == "grassland") + 0.6 * dat$temperature
sigma_true <- exp(log(0.6) + 0.55 * (dat$habitat == "grassland"))
dat$growth <- rnorm(n, mean = mu, sd = sigma_true)
```

Fit the location-scale model:

```{r location-scale-fit}
fit_growth <- drmTMB(
  drm_formula(growth ~ habitat + temperature, sigma ~ habitat),
  family = gaussian(),
  data = dat
)
```

Before interpreting coefficients, check the fitted object:

```{r location-scale-check}
check_drm(fit_growth)
```

Now print the coefficient table:

```{r location-scale-summary}
summary(fit_growth)
```

How to read this output:

1. Rows beginning with `mu:` are mean-growth effects on the response scale.
   For example, `mu:temperature` is the expected change in mean growth for a
   one-unit increase in temperature.
2. Rows beginning with `sigma:` are log-residual-standard-deviation effects.
   For example, `sigma:habitatgrassland` is not an additive change in growth;
   it is a log-scale change in residual variability.
3. Exponentiating a `sigma` coefficient gives the multiplicative change in
   residual standard deviation.

```{r location-scale-interpret-sigma}
sigma_habitat <- coef(fit_growth, "sigma")["habitatgrassland"]
data.frame(
  coefficient = sigma_habitat,
  residual_sd_ratio = exp(sigma_habitat),
  residual_variance_ratio = exp(2 * sigma_habitat)
)
```

In this fitted example, `exp(sigma:habitatgrassland)` is the estimated ratio of
grassland residual SD to forest residual SD. A value near 2 means that
grassland has about twice the residual SD of forest after accounting for mean
habitat and temperature effects. Because residual variance is SD squared, the
same fitted coefficient implies about four times the residual variance.

It is often clearer to report the fitted residual SDs and variances directly:

```{r location-scale-response-table}
newdat <- data.frame(
  habitat = factor(c("forest", "grassland"), levels = levels(dat$habitat)),
  temperature = 0
)

growth_report <- data.frame(
  habitat = newdat$habitat,
  fitted_mean_growth = predict(fit_growth, newdata = newdat, dpar = "mu"),
  fitted_residual_sd = predict(fit_growth, newdata = newdat, dpar = "sigma")
)
growth_report$fitted_residual_variance <- growth_report$fitted_residual_sd^2
growth_report
```

This table maps the model back to the scientific question. The mean column
summarises the location model. The residual SD column summarises the scale
model in the units of growth. The residual variance column is the
variance-facing version of the same fitted Gaussian scale model. If grassland
has a larger fitted residual SD and residual variance, the model says that
individual growth is less predictable in grassland even after the mean
difference and temperature effect have been modelled.

## Curved responses and interactions

Ecological responses are often curved rather than straight-line. Growth may
increase from cold to moderate temperatures and then decline at high
temperatures. A second-order term is usually enough for this kind of question;
third-order polynomials are harder to explain and should be used only when the
question and data justify them.

The symbolic model below asks whether mean growth has a quadratic temperature
response and whether the temperature slope differs between habitats. It also
asks whether residual variability changes along the same temperature gradient:

\[
\begin{aligned}
\text{growth}_i &\sim \operatorname{Normal}(\mu_i, \sigma_i^2), \\
\mu_i &= \beta_0 + \beta_1 \text{temperature}_i +
        \beta_2 \text{temperature}_i^2 +
        \beta_3 \mathbb{1}(\text{habitat}_i = \text{grassland}) +
        \beta_4 \text{temperature}_i
          \mathbb{1}(\text{habitat}_i = \text{grassland}), \\
\log(\sigma_i) &= \gamma_0 + \gamma_1 \text{temperature}_i^2.
\end{aligned}
\]

The matching R formula uses ordinary R formula syntax:

```r
drmTMB(
  drm_formula(
    growth ~ habitat * temperature + I(temperature^2),
    sigma ~ I(temperature^2)
  ),
  family = gaussian(),
  data = dat_curve
)
```

The term `habitat * temperature` expands to the two main effects plus their
interaction. The term `I(temperature^2)` tells R to square `temperature`
before building the design matrix. The alternative `poly(temperature, 2)` is
also supported and can be numerically useful, but its coefficients are basis
coefficients; for applied interpretation, predicted means and residual SDs are
usually clearer than the raw polynomial coefficients.

```{r location-scale-curved-simulate}
set.seed(43)
n_curve <- 260
dat_curve <- data.frame(
  habitat = factor(rep(c("forest", "grassland"), each = n_curve / 2)),
  temperature = runif(n_curve, -2, 2)
)

mu_curve <- 8 +
  0.6 * dat_curve$temperature -
  0.45 * dat_curve$temperature^2 +
  0.8 * (dat_curve$habitat == "grassland") +
  0.35 * dat_curve$temperature * (dat_curve$habitat == "grassland")
sigma_curve <- exp(log(0.55) + 0.18 * dat_curve$temperature^2)
dat_curve$growth <- rnorm(n_curve, mean = mu_curve, sd = sigma_curve)
```

```{r location-scale-curved-fit}
fit_curve <- drmTMB(
  drm_formula(
    growth ~ habitat * temperature + I(temperature^2),
    sigma ~ I(temperature^2)
  ),
  family = gaussian(),
  data = dat_curve
)

summary(fit_curve)
```

For a curved model, the most useful first output is often a prediction table.
Here we compare cold, moderate, and warm temperatures in the two habitats:

```{r location-scale-curved-predictions}
curve_grid <- expand.grid(
  habitat = levels(dat_curve$habitat),
  temperature = c(-1.5, 0, 1.5)
)

data.frame(
  curve_grid,
  fitted_mean_growth = predict(fit_curve, newdata = curve_grid, dpar = "mu"),
  fitted_residual_sd = predict(fit_curve, newdata = curve_grid, dpar = "sigma")
)
```

This table is easier to explain than the individual polynomial coefficients.
The fitted mean column describes the location curve: expected growth is highest
near the middle of the temperature range in this simulated example. The fitted
residual SD column describes the scale curve: growth is less predictable at
temperature extremes when the fitted residual SD is larger.

For several predictors, base R interaction expansions work in the same way.
For example, `(temperature + moisture + canopy)^2` expands to all three main
effects and all pairwise interactions. That formula is often easier to read
than writing every interaction by hand, but it should still be tied to a clear
biological question.

If the `sigma` formula is omitted, `drmTMB` fits an intercept-only residual
standard deviation:

```{r location-scale-intercept-only}
fit0 <- drmTMB(
  drm_formula(growth ~ habitat + temperature),
  family = gaussian(),
  data = dat
)

coef(fit0, "sigma")
```

That simpler model can still estimate habitat and temperature effects on mean
growth, but it assumes one residual SD for all observations.

## Meta-analysis as Gaussian regression

Meta-analysis is the same Gaussian regression idea with known sampling
covariance added to the likelihood. For diagonal known sampling variance:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma_i, v_i
  &\sim \operatorname{Normal}(\mu_i, v_i + \sigma_i^2),\\
\mu_i
  &= X_{\mu}[i, ]\beta_{\mu},\\
\sigma_i
  &= \exp(X_{\sigma}[i, ]\beta_{\sigma}).
\end{aligned}
\]

with matching syntax:

```r
drmTMB(
  bf(yi ~ x1 + meta_known_V(V = vi), sigma ~ x1),
  family = gaussian(),
  data = dat
)
```

The canonical marker is `meta_known_V(V = V)`; in concrete code, `V` can be a
vector such as `vi`. Here `vi` is known sampling variance. The estimated
`sigma_i` is the additional heterogeneity standard deviation after known
sampling error has been included. In meta-analysis prose this quantity is often
called `tau`, but the `drmTMB` API keeps the name `sigma` because it is still
the residual scale parameter of the Gaussian regression.

When `V` is a full known covariance matrix, the same model is:

\[
y \mid \mu, \sigma, V
  \sim \operatorname{MVN}\left(\mu, V + \operatorname{diag}(\sigma_i^2)\right).
\]

## Current implementation caveats

- The current fitter supports `family = gaussian()` with fixed effects for
  `mu` and `sigma`, plus random intercepts, independent numeric random slopes,
  and labelled or unlabelled ordinary correlated intercept-slope blocks in
  `mu`.
- Residual-scale random intercepts and independent random slopes in `sigma`,
  such as `sigma ~ x2 + (1 | id)` and `sigma ~ x2 + (0 + w | id)`, are
  implemented for univariate Gaussian models.
- The `sigma` formula is the observation-level residual standard deviation. It
  is not a random-effect scale component.
- Missing rows are dropped when they are incomplete for variables used by
  either the `mu` or `sigma` formula.
- Diagonal and dense full known sampling covariance via `meta_known_V(V = V)` is
  supported for Gaussian meta-analysis.
- Fixed-effect bivariate Gaussian `rho12` models are implemented in the
  bivariate coscale article.
- `sd(group)` random-effect scale formulae are implemented for one or more
  distinct unlabelled Gaussian `mu` random intercepts, with group-level
  predictors that are constant within the named grouping variable.
- Correlated residual-scale slope blocks, slope-specific `sd()` targets,
  labelled-block `sd()` targets, cross-formula labelled covariance sharing
  beyond the first random-intercept slice, phylogenetic slopes, standalone or
  partial phylogenetic scale terms, spatial `sigma`, bivariate spatial
  covariance, mesh/SPDE spatial fields, multiple spatial slopes, and
  non-Gaussian random effects are planned but not implemented in this Gaussian
  tutorial.
- Intercept-only phylogenetic random effects are implemented in univariate
  Gaussian `mu` formulas as `phylo(1 | species, tree = tree)`.
- Coordinate-spatial random effects are implemented in univariate Gaussian `mu`
  formulas as `spatial(1 | site, coords = coords)` and one numeric
  `spatial(1 + x | site, coords = coords)` slope.

Multiple random-effect scale components use explicit `sd()` formulas, not
extra residual `sigma` formulas:

```r
drmTMB(
  bf(
    y ~ x1 + (1 | study) + (1 | species),
    sigma ~ x1,
    sd(study) ~ x1,
    sd(species) ~ 1
  ),
  family = gaussian(),
  data = dat
)
```
