---
title: "Random-effects meta-analysis with known sampling variances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random-effects meta-analysis with known sampling variances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
```

<style>
@media (max-width: 575.98px) {
  main h1 {
    hyphens: none;
    overflow-wrap: normal;
    word-break: normal;
  }

  main table {
    display: block;
    max-width: 100%;
    overflow-x: auto;
  }
}
</style>

A random-effects meta-analysis pools effect sizes from several studies while
acknowledging that the studies differ. Each study reports an effect size
together with its sampling variance, which is treated as **known**. The model
estimates two things on top of that known uncertainty: a pooled mean effect and
the between-study heterogeneity. This is a specialist route for data that
already are effect sizes with known sampling variances; for raw observations,
start with [Distributional regression with drmTMB](drmTMB.html) instead.

In `drmTMB` this is ordinary Gaussian distributional regression with a known
sampling covariance. There is no separate meta-analysis family. You fit
`family = gaussian()`, you supply the known per-study variances with
`meta_V()` inside the location formula, and the between-study heterogeneity is
the residual scale `sigma`.

Throughout, `Normal(a, b)` uses the variance (not the standard deviation) as
its second argument.

## The model

Write \(y_i\) for the observed effect size from study \(i\) and \(v_i\) for its
**known** sampling variance. The random-effects model is

\[
y_i \mid \mu, \tau, v_i
  \;\sim\;
  \operatorname{Normal}\!\left(\mu,\; v_i + \tau^2\right),
\qquad i = 1, \ldots, K.
\]

Each study sees the same pooled mean \(\mu\) but its own total variance
\(v_i + \tau^2\): the known sampling variance \(v_i\) that the primary study
already quantified, plus a shared between-study variance \(\tau^2\) that the
meta-analysis estimates. When \(\tau = 0\) every study is just a noisy
measurement of one common effect (a fixed-effect, or common-effect,
meta-analysis); when \(\tau > 0\) the true study effects themselves scatter
around \(\mu\).

The two unknowns map onto the two `drmTMB` distributional parameters:

| Meta-analysis quantity | Symbol | `drmTMB` parameter | How to read it |
|---|---|---|---|
| pooled effect | \(\mu\) | `mu` intercept | the average effect across studies |
| known sampling variance | \(v_i\) | `meta_V(V = vi)` | supplied, not estimated |
| between-study SD | \(\tau\) | `sigma` | how much true effects differ across studies |
| between-study variance | \(\tau^2\) | `sigma^2` | the heterogeneity variance \(\tau^2\) |

The matching R syntax is

```r
drmTMB(
  bf(yi ~ 1 + meta_V(V = vi), sigma ~ 1),
  family = gaussian(),
  data = dat
)
```

Two points are worth stating plainly.

First, `vi` must be a **variance**. If your dataset stores standard errors,
square them (`vi <- se^2`) before fitting.

Second, the residual scale `sigma` *is* the between-study heterogeneity \(\tau\).
Meta-analysts usually call this quantity `tau` and report `tau^2`. `drmTMB`
keeps the public parameter name `sigma` so that meta-analysis uses exactly the
same distributional grammar as every other Gaussian model in the package; you
recover the familiar `tau^2` simply by squaring `sigma`. The marker
`meta_V()` puts the *known* variance into the likelihood; `sigma` estimates the
*unknown* variance that is left over.

## A simulated dataset

We simulate `K = 30` studies. Each study has a true effect drawn around a
common mean, and we observe that true effect with known sampling error.

```{r}
set.seed(101)

K <- 30
mu_true <- 0.40 # pooled effect
tau_true <- 0.30 # between-study SD

# Known sampling variances: larger studies (smaller vi) and smaller studies.
vi <- runif(K, 0.02, 0.10)

# True study effects scatter around mu_true with SD tau_true.
theta_i <- rnorm(K, mean = mu_true, sd = tau_true)

# Observed effect sizes: each true effect seen with its known sampling error.
yi <- rnorm(K, mean = theta_i, sd = sqrt(vi))

dat <- data.frame(study = factor(seq_len(K)), yi = yi, vi = vi)
head(dat)
```

The data frame has one row per study: the effect size `yi` and its known
sampling variance `vi`.

## Fitting the model

```{r}
fit <- drmTMB(
  bf(yi ~ 1 + meta_V(V = vi), sigma ~ 1),
  family = gaussian(),
  data = dat
)

summary(fit)
```

Before reading the coefficients, confirm the fit is sound. A clean optimisation
and a positive-definite Hessian are the gate for trusting the Wald standard
errors and intervals.

```{r}
fit$opt$convergence # 0 means the optimiser converged
fit$sdr$pdHess # TRUE means the Hessian is positive definite
```

```{r}
check_drm(fit)
```

## The pooled effect

The pooled effect is the `mu` intercept. Its Wald confidence interval comes from
`confint()`.

```{r}
mu_hat <- coef(fit, "mu")[["(Intercept)"]]
mu_hat

confint(fit, parm = "mu:(Intercept)")[, c("parm", "lower", "upper")]
```

In this run the pooled estimate is about `r round(mu_hat, 3)`, which sits near
the simulated `mu_true = 0.40`. The interval reflects uncertainty in the mean
*after* both the known sampling variances and the estimated between-study
heterogeneity have been accounted for.

## Between-study heterogeneity

The between-study SD \(\tau\) is the residual scale `sigma`. Because the `sigma`
formula here is intercept-only, every study shares the same value, so we take
the first element. Squaring it gives the heterogeneity variance \(\tau^2\) that
meta-analysis reports.

```{r}
tau_hat <- sigma(fit)[1]
c(tau = unname(tau_hat), tau_squared = unname(tau_hat^2))
```

Heterogeneity is easier to communicate as a proportion. \(I^2\) is the share of
the total variation that is between-study rather than sampling noise. With the
usual "typical" within-study variance \(\tilde v\) of Higgins and Thompson
(2002),

\[
\tilde v =
  \frac{(K-1)\sum_i w_i}{\left(\sum_i w_i\right)^2 - \sum_i w_i^2},
\qquad w_i = 1 / v_i,
\qquad
I^2 = \frac{\tau^2}{\tau^2 + \tilde v}.
\]

```{r}
w <- 1 / dat$vi
v_typical <- ((K - 1) * sum(w)) / (sum(w)^2 - sum(w^2))
I2 <- tau_hat^2 / (tau_hat^2 + v_typical)

c(
  tau_squared = unname(tau_hat^2),
  typical_v = v_typical,
  I2_percent = unname(100 * I2)
)
```

An \(I^2\) of this size means a substantial fraction of the variation among the
observed effect sizes reflects genuine differences between studies, not just
within-study sampling error. The pooled mean is still meaningful, but it is a
mean of effects that really do differ.

## Cross-check against metafor

The same model can be fitted with `metafor::rma()` using maximum likelihood. It
should agree with `drmTMB`, because both fit the identical random-effects
likelihood \(y_i \sim \operatorname{Normal}(\mu, v_i + \tau^2)\). This is a
useful sanity check when you first adopt the `drmTMB` spelling.

```{r}
if (requireNamespace("metafor", quietly = TRUE)) {
  rma_fit <- metafor::rma(yi = yi, vi = vi, method = "ML", data = dat)

  comparison <- data.frame(
    quantity = c("pooled mu", "tau^2", "I^2 (%)"),
    drmTMB = c(mu_hat, tau_hat^2, 100 * I2),
    metafor = c(as.numeric(rma_fit$beta), rma_fit$tau2, rma_fit$I2)
  )
  print(comparison, row.names = FALSE, digits = 4)
}
```

The two engines return the same pooled effect and the same heterogeneity
variance. `drmTMB` is doing ML random-effects meta-analysis; it simply spells
the known sampling variance as `meta_V()` and the heterogeneity as `sigma`.

## REML for the heterogeneity

The ML estimate of \(\tau^2\) is known to be biased downward, because it does
not account for the degrees of freedom spent estimating \(\mu\). When the mean
model is fixed and you only want a better heterogeneity estimate, restricted
maximum likelihood (`REML = TRUE`) is the standard remedy. Keep ML
(`REML = FALSE`, the default) whenever you intend to compare different
fixed-effect mean models with AIC or BIC, since restricted likelihoods are not
comparable across different mean structures.

```{r}
fit_reml <- drmTMB(
  bf(yi ~ 1 + meta_V(V = vi), sigma ~ 1),
  family = gaussian(),
  data = dat,
  REML = TRUE
)

data.frame(
  estimator = c("ML", "REML"),
  pooled_mu = c(coef(fit, "mu")[[1]], coef(fit_reml, "mu")[[1]]),
  tau = c(sigma(fit)[1], sigma(fit_reml)[1]),
  tau_squared = c(sigma(fit)[1]^2, sigma(fit_reml)[1]^2)
)
```

The REML between-study variance is slightly larger than the ML one, as expected.

## Meta-regression: moderators on the mean

If a study-level covariate might explain part of the variation in effect sizes,
add it to the `mu` formula. This is a random-effects meta-regression: the known
sampling variances stay in `meta_V()`, `sigma` becomes the *residual* (after
moderators) between-study SD, and the new coefficient measures how the effect
size changes with the moderator.

```{r}
set.seed(202)
dat$dose <- scale(runif(K, 1, 10))[, 1] # a study-level moderator
# Give the effect size a genuine dependence on the moderator.
dat$yi <- dat$yi + 0.25 * dat$dose

fit_mr <- drmTMB(
  bf(yi ~ 1 + dose + meta_V(V = vi), sigma ~ 1),
  family = gaussian(),
  data = dat
)

coef(fit_mr, "mu")
```

The `dose` coefficient is the change in the pooled effect per one-SD change in
the moderator. After fitting a moderator, the residual `sigma` is the
between-study heterogeneity that the moderator did *not* explain; comparing it
with the no-moderator `sigma` shows how much heterogeneity the moderator
absorbed.

```{r}
c(
  residual_tau_no_moderator = unname(sigma(fit)[1]),
  residual_tau_with_moderator = unname(sigma(fit_mr)[1])
)
```

## Multiple effect sizes per study

The worked example above has one effect size per study, so `sigma` carries the
whole between-study story. When a study contributes several effect sizes, two
levels of variation appear: a study-level random effect for the studies, and a
residual for the effect sizes within a study. Those are different questions, and
`drmTMB` keeps them in different places:

```r
# Schematic: several effect sizes per study (not evaluated here).
drmTMB(
  bf(yi ~ 1 + moderator + (1 | study) + meta_V(V = vi), sigma ~ 1),
  family = gaussian(),
  data = dat_repeated
)
```

Here `(1 | study)` is the between-study random effect and `sigma` is the
within-study residual heterogeneity, while `meta_V(V = vi)` still supplies the
known sampling variances. A grouping factor used in `(1 | study)` must have at
least one study with repeated rows; a data set with exactly one row per study is
the single-level model shown above, where `sigma` alone represents
between-study heterogeneity.

## Known sampling variance is not a weight

Inverse-variance weights and known sampling variances answer different
questions, and `meta_V()` is not the same as the top-level `weights` argument.

A likelihood weight multiplies a study's contribution to the log-likelihood:

\[
\ell(\theta) = \sum_i w_i \, \ell_i(\theta).
\]

A known sampling variance enters the *covariance* of the response:

\[
y_i \sim \operatorname{Normal}(\mu,\; v_i + \tau^2).
\]

So `weights = 1 / vi` is **not** the random-effects meta-analysis model. It
rescales how much each row counts toward the likelihood; it does not put `vi`
into the modelled sampling variance, and it does not let `tau^2` be estimated on
top of the known variances. For meta-analysis with known sampling variances,
use `meta_V(V = vi)`. Reserve `weights =` for genuine likelihood weights such as
externally defined case weights.

## Notes on the function names

* `meta_V(V = V)` is the current marker for known sampling variance or
  covariance. The argument may be a column of variances (as above), a vector, a
  diagonal matrix, or a dense covariance matrix when the effect sizes are
  correlated.
* `meta_known_V(V = V)` is a deprecated alias kept only for backward
  compatibility. It routes to the same additive known-variance likelihood but
  emits a deprecation warning; prefer `meta_V()` in new code.
* There is intentionally no `meta_gaussian()` family and no `tau ~` syntax.
  Meta-analysis reuses `family = gaussian()` and `sigma ~ ...` so that it shares
  the distributional-regression grammar with the rest of `drmTMB`.

## Reference

Higgins, J. P. T., and Thompson, S. G. (2002). Quantifying heterogeneity in a
meta-analysis. *Statistics in Medicine*, 21(11), 1539-1558.
