---
title: "Mean effects and residual heterogeneity in meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mean effects and residual heterogeneity in meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Meta-analysis in `drmTMB` is ordinary Gaussian distributional regression with
known sampling covariance. The sampling covariance is supplied with
`meta_V(V = V)` inside the location formula; it is not a separate
`meta_gaussian()` family. Deprecated `meta_known_V(V = V)` remains supported
only as a compatibility alias for older examples.

The distinction matters. A meta-analysis usually has two sources of variation:
known sampling error from the original studies, and unknown heterogeneity that
the meta-analytic model estimates. In `drmTMB`, known sampling covariance is
`V`, and unknown residual heterogeneity is `sigma`.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

The source logic mirrors recent location-scale meta-analysis papers and notes:
Nakagawa et al. (2025) frame heterogeneity as a biological and methodological
response, Yang and Nakagawa's distributional-regression manuscript treats
location, scale, and shape as modelled distributional parameters, Rodriguez et
al. (2023) argue that categorical moderators should usually allow unequal
between-study heterogeneity, and Nakagawa's 2022 unifying note separates
ordinary row weights from proportional sampling-variance components. The fitted
syntax below stays inside the currently implemented `drmTMB` surface.

## The model

For a univariate meta-analysis with independent sampling errors, the model is:

\[
y_i \mid \mu_i, \sigma_i, v_i
  \sim \operatorname{Normal}(\mu_i, v_i + \sigma_i^2)
\]

\[
\mu_i = \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma}.
\]

The matching R syntax is:

```r
drmTMB(
  drm_formula(
    yi ~ x1 + x2 + meta_V(V = vi),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

Here `vi` contains known sampling variances. If a dataset contains standard
errors, square them before fitting. The `sigma` formula models the extra
heterogeneity left after the known sampling variance has been included. In
meta-analysis papers this extra heterogeneity SD is often written as `tau`;
`drmTMB` keeps the public parameter name `sigma` so that meta-analysis uses the
same distributional grammar as other Gaussian models.

Read each symbol before interpreting coefficients:

| Symbol or syntax | Meaning in a meta-analysis | In the restoration example |
|---|---|---|
| \(y_i\), `yi` | observed effect size for effect-size row \(i\) | standardized biodiversity response to restoration |
| \(v_i\), `vi` | known sampling variance for \(y_i\), treated as known before fitting | uncertainty from the primary restoration study |
| \(\mathbf{V}\), `V` | known sampling covariance matrix when effect sizes are correlated | a dense or block-diagonal covariance matrix for correlated effect sizes |
| \(\mu_i\), `mu` | expected effect size after location moderators | mean restoration effect for a habitat and duration value |
| \(\boldsymbol{\beta}_{\mu}\) | location coefficients in the mean model | average forest-grassland and duration contrasts |
| \(\sigma_i\), `sigma` | fitted extra residual heterogeneity SD after known sampling variance is included | remaining among-effect-size unpredictability |
| \(\boldsymbol{\beta}_{\sigma}\) | log-SD coefficients for extra heterogeneity | habitat effects on residual heterogeneity, reported as `exp(coef)` SD ratios |
| `sd(study)` | study-level random-effect SD, when repeated effect sizes are grouped within studies | among-study heterogeneity distinct from residual `sigma` |
| `weights = w` | ordinary row log-likelihood multiplier | not a substitute for known sampling variance |

For correlated effect sizes, the same idea is written with a full known
sampling covariance matrix:

\[
\mathbf{y} \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu},
    \mathbf{V} + \operatorname{diag}(\sigma_1^2,\ldots,\sigma_n^2)
  \right).
\]

The right-hand side of `meta_V(V = V)` may be a variance column, a vector, a
diagonal matrix, a dense block-diagonal matrix, or a dense full covariance
matrix.

Dense means dense. If `V` is supplied as a full matrix, `drmTMB` stores the
retained covariance as a dense R matrix. This is useful for small to moderate
correlated-effect examples and for likelihood-comparator checks, but it is not
the large-meta-analysis storage route. Large block-diagonal or sparse known
covariance structures need a future sparse or block-sparse path before they
should be treated as scalable.

## Worked example: restoration effects

Suppose we are synthesising effect sizes from studies that compare ecological
restoration with a control condition. The response `yi` is a standardised
effect size for biodiversity. The known sampling variance `vi` comes from each
primary study. We ask two questions:

1. Is the mean restoration effect different in forests and grasslands?
2. Is the unexplained among-effect-size heterogeneity larger in one habitat?

This simulated dataset stands in for the structure of a real ecological
meta-analysis:

```{r}
set.seed(20)
n <- 80
dat <- data.frame(
  study = factor(seq_len(n)),
  habitat = factor(
    rep(c("grassland", "forest"), length.out = n),
    levels = c("grassland", "forest")
  ),
  duration = scale(runif(n, 1, 8))[, 1]
)

dat$vi <- runif(n, 0.015, 0.08)

mu <- 0.12 + 0.22 * (dat$habitat == "forest") + 0.08 * dat$duration
sigma_extra <- exp(-1.65 + 0.55 * (dat$habitat == "forest"))

dat$yi <- rnorm(n, mean = mu, sd = sqrt(dat$vi + sigma_extra^2))
head(dat)
```

The fitted model mirrors the equations above:

```{r}
fit_meta <- drmTMB(
  drm_formula(
    yi ~ habitat + duration + meta_V(V = vi),
    sigma ~ habitat
  ),
  family = gaussian(),
  data = dat
)

summary(fit_meta)
```

Before trusting the coefficient table in your own analysis, run
`check_drm(fit_meta)`. If a `meta_V(V = V)` model with a moderator in `sigma`
reports `pdHess = FALSE`, treat the fitted heterogeneity pattern as a diagnostic
result rather than settled inference. First try the same mean model with
`sigma ~ 1`; then add the `sigma` moderator back only if the Hessian, gradients,
and fitted extra-heterogeneity values look stable.

The `mu` rows describe the average effect-size model. In this simulation the
forest coefficient is positive, so the model estimates a larger mean
restoration effect for forests than for grasslands, after accounting for study
duration and known sampling variance.

The `sigma` rows are on the log-SD scale. They describe extra heterogeneity,
not sampling error. The forest contrast can be read as a multiplicative change
in the residual heterogeneity SD:

```{r}
sigma_coef <- coef(fit_meta, "sigma")
sigma_coef
exp(sigma_coef["habitatforest"])
```

In this run, forest studies have about `r round(exp(sigma_coef["habitatforest"]), 2)`
times the extra heterogeneity SD of grassland studies. That multiplier applies
to `sigma_i`; it does not alter the known sampling variances `vi`.

## Categorical moderators and heterogeneous heterogeneity

Rodriguez et al. (2023) recommend treating unequal between-study heterogeneity
as the default when testing categorical moderators. In `drmTMB`, the current
Gaussian meta-analysis route expresses that idea with a `sigma` formula rather
than a new `tau ~` syntax. For the two-habitat example above,

\[
\log(\sigma_i) =
  \gamma_0 + \gamma_1 \operatorname{forest}_i,
\]

so the fitted extra heterogeneity SDs are

\[
\sigma_{\mathrm{grassland}} = \exp(\gamma_0),
\qquad
\sigma_{\mathrm{forest}} = \exp(\gamma_0 + \gamma_1).
\]

The scale coefficient \(\gamma_1\) is therefore not another mean effect. It is
the log ratio of forest to grassland residual heterogeneity SD:

\[
\frac{\sigma_{\mathrm{forest}}}
     {\sigma_{\mathrm{grassland}}}
  = \exp(\gamma_1).
\]

If the report uses variance-facing meta-analysis language, the corresponding
extra heterogeneity variance ratio is:

\[
\frac{\sigma_{\mathrm{forest}}^2}
     {\sigma_{\mathrm{grassland}}^2}
  = \exp(2\gamma_1).
\]

Biologically, the mean forest coefficient asks whether restoration has a larger
average biodiversity effect in forests. The `sigma` forest coefficient asks a
different question: are forest restoration effects less predictable across
studies after known sampling error and duration are accounted for? Those two
answers can point in the same direction, opposite directions, or only one can be
important.

It is often helpful to translate fitted values back to the response scale and
to the variance scale used in many meta-analysis summaries:

```{r}
meta_report <- data.frame(
  habitat = dat$habitat,
  known_sampling_variance = dat$vi,
  fitted_extra_heterogeneity_sd = sigma(fit_meta)
)
meta_report$fitted_extra_heterogeneity_variance <-
  meta_report$fitted_extra_heterogeneity_sd^2
meta_report$total_observation_variance <-
  meta_report$known_sampling_variance +
  meta_report$fitted_extra_heterogeneity_variance

meta_by_habitat <- aggregate(
  meta_report[c(
    "fitted_extra_heterogeneity_sd",
    "fitted_extra_heterogeneity_variance",
    "total_observation_variance"
  )],
  list(habitat = meta_report$habitat),
  mean
)
meta_by_habitat
```

For this model, the fitted residual heterogeneity SD is larger in forest
studies. The heterogeneity variance column is the same fitted `sigma` model
reported as `sigma^2`. The total observation variance column adds the known
sampling variance `vi`, so it is larger than the estimated heterogeneity
variance alone. A biological interpretation would be: restoration effects are
not only larger on average in forests, they are also less predictable across
forest studies after known sampling uncertainty is accounted for.

Choose the report scale deliberately:

| Quantity | Computation | Interpretation |
|---|---|---|
| known sampling variance | `vi` or the supplied `V` | uncertainty known before fitting |
| extra heterogeneity SD | `sigma(fit_meta)` | residual between-effect-size SD after known sampling variance is included |
| extra heterogeneity variance | `sigma(fit_meta)^2` | variance-scale version of the fitted `sigma` model |
| total observation variance | `vi + sigma(fit_meta)^2` for diagonal `V` | sampling variance plus estimated extra heterogeneity |

This is why `drmTMB` keeps the public parameter name `sigma`: it is the fitted
extra heterogeneity scale in the Gaussian likelihood. If a meta-analysis report
needs the usual variance-facing language, square `sigma` and say that the
reported column is `sigma^2`.

```{r meta-variance-components-figure, eval=requireNamespace("ggplot2", quietly = TRUE), fig.width=7.2, fig.height=4.4, fig.cap="Variance-scale summary for the restoration meta-analysis example. Stacked bars separate mean known sampling variance from fitted extra heterogeneity variance by habitat; points mark the mean total observation variance. These are variance components, not confidence intervals.", fig.alt="Stacked bar plot by habitat showing known sampling variance and fitted extra heterogeneity variance as separate components, with a point marking total observation variance for grassland and forest studies."}
library(ggplot2)

meta_components <- rbind(
  data.frame(
    habitat = meta_by_habitat$habitat,
    component = "Known sampling variance",
    variance = aggregate(
      known_sampling_variance ~ habitat,
      data = meta_report,
      mean
    )$known_sampling_variance
  ),
  data.frame(
    habitat = meta_by_habitat$habitat,
    component = "Fitted extra heterogeneity",
    variance = meta_by_habitat$fitted_extra_heterogeneity_variance
  )
)
meta_components$component <- factor(
  meta_components$component,
  levels = c("Fitted extra heterogeneity", "Known sampling variance")
)

ggplot(meta_components, aes(habitat, variance, fill = component)) +
  geom_col(width = 0.58, colour = "white", linewidth = 0.3) +
  geom_point(
    data = meta_by_habitat,
    aes(
      x = habitat,
      y = total_observation_variance
    ),
    inherit.aes = FALSE,
    size = 2.8,
    shape = 21,
    fill = "white",
    colour = "grey20",
    stroke = 0.9
  ) +
  scale_fill_manual(values = c(
    "Known sampling variance" = "#56B4E9",
    "Fitted extra heterogeneity" = "#CC79A7"
  )) +
  labs(
    title = "Known sampling variance is not residual heterogeneity",
    subtitle = "Bars are variance components; white points mark their mean total",
    x = "Habitat",
    y = "Variance",
    fill = "Component"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30")
  )
```

A quick diagnostic check flags optimizer, Hessian, row-filtering, scale, and
known sampling covariance problems before interpretation:

```{r}
check_drm(fit_meta)
```

For full-matrix known covariance fits, the `known_sampling_covariance` row is a
note rather than a problem by itself. It reports the retained matrix dimension,
dense storage, density, size, rank, and conditioning so you can see whether the
current dense route still matches the scale of the analysis.

For location-scale meta-analysis fits such as
`bf(y ~ moderator + meta_V(V = v), sigma ~ moderator)`, a clean Hessian is the
gate for Hessian-based SEs and Wald intervals. If `check_drm()` reports
`hessian_positive_definite` as a warning, the known sampling variance still
enters the covariance additively through `meta_V(V = v)`, but the estimated
extra-heterogeneity coefficients should be treated as point estimates until a
profile, bootstrap, or simpler `sigma` model supports them.

## Weights are not sampling variances

The top-level `weights =` argument and `meta_V(V = V)` answer different
questions. Ordinary likelihood weights multiply each observation's
log-likelihood contribution:

\[
\ell(\theta) =
  \sum_i w_i \ell_i(\theta).
\]

Known sampling covariance changes the covariance model:

\[
\mathbf{y} \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu},
    \mathbf{V} + \boldsymbol{\Omega}_{\mathrm{estimated}}
  \right).
\]

Thus `weights = 1 / vi` is not the same model as
`meta_V(V = vi)`. Use `meta_V(V = vi)` when `vi` is known sampling variance.
Use `weights =` only when the intended analysis is a weighted likelihood, such
as case weights or externally defined row weights. The practical consequence is
that `weights = 1 / vi` changes how much each row contributes to the
likelihood; it does not put `vi` into the modelled sampling variance. The
current dense full-`V` path rejects non-unit weights until joint-block
weighting has a separate design.

A proportional sampling-variance model, for example a term such as
\(\pi_i \sim \operatorname{Normal}(0, \phi_\pi / w_i)\), is a different model
again. It is not implemented in `drmTMB` yet and should not be mimicked with
the top-level `weights =` argument. The preferred additive known-`V` spelling
is `meta_V(V = V)`; the proportional branch remains design-only:

```r
# Preferred additive known-V spelling.
meta_V(V = V)

# Future design only; this call is not implemented.
meta_V(w = w, scale = "proportional")
```

Deprecated `meta_known_V(V = V)` remains a compatibility alias for the additive
known-`V` form, not a second likelihood path. The response is already on the
left-hand side of the model formula, so `meta_V()` should not repeat it as a
positional argument. Current fitted examples should use `meta_V(V = V)` for
known sampling covariance and `weights =` only for likelihood weights.

## Known sampling variance plus group-level heterogeneity

Some meta-analyses contain multiple effect sizes per study, species, lab, or
site. Then `sigma` and `sd(group)` answer different questions. The next code
block is schematic; it assumes a dataset such as `dat_repeated`, with multiple
rows per study and a study-level `habitat` value that is constant within each
study after missing rows are removed:

```r
fit_study <- drmTMB(
  drm_formula(
    yi ~ treatment + (1 | study) + meta_V(V = vi),
    sigma ~ 1,
    sd(study) ~ habitat
  ),
  family = gaussian(),
  data = dat_repeated
)
```

The symbolic model is:

\[
y_i = \mu_i + b_{j[i]} + r_i + s_i,
\qquad
\mu_i = \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu},
\]

\[
b_j \sim \operatorname{Normal}(0, \omega_j^2),
\qquad
\log(\omega_j) =
  \mathbf{x}_{\omega j}^{\top}\boldsymbol{\alpha},
\]

\[
r_i \sim \operatorname{Normal}(0, \sigma_i^2),
\qquad
\mathbf{s} \sim \operatorname{MVN}(\mathbf{0}, \mathbf{V}).
\]

Here `s_i` is the known sampling-error component supplied by
`meta_V(V = vi)` or `meta_V(V = V)`. The `sigma` formula models remaining
observation-level heterogeneity. The `sd(study)` formula models `\omega_j`,
the study-level random-effect standard deviation. Predictors in
`sd(study) ~ ...` must be constant within each study after missing-data
filtering; otherwise a single study-level random-effect SD would not be well
defined.

After integrating over the random effects, the marginal covariance is:

\[
\boldsymbol{\Omega} =
  \mathbf{V}
  + \operatorname{diag}(\sigma_1^2,\ldots,\sigma_n^2)
  + \mathbf{Z}_{study}
    \operatorname{diag}(\omega_1^2,\ldots,\omega_J^2)
    \mathbf{Z}_{study}^{\top}.
\]

This is the distinction to keep in mind: `sigma ~ habitat` asks whether
effect-size-level residual heterogeneity changes with habitat, while
`sd(study) ~ habitat` asks whether the study-level random-effect SD changes
with habitat.

## Bivariate meta-analysis

Bivariate meta-analysis uses the same Gaussian bivariate route as other
location-coscale models, where coscale means the residual correlation
component `rho12`. The known sampling covariance should include the within-study
covariance between the two effect-size estimates when it is available.

For study \(i\), write:

\[
\mathbf{y}_i =
  \begin{bmatrix} y_{1i} \\ y_{2i} \end{bmatrix},
\qquad
\mathbf{S}_i =
  \begin{bmatrix}
    v_{1i} & c_{12i} \\
    c_{12i} & v_{2i}
  \end{bmatrix}.
\]

The fitted residual covariance is:

\[
\boldsymbol{\Omega}_i =
  \begin{bmatrix}
    \sigma_{1i}^2
      & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
    \rho_{12i}\sigma_{1i}\sigma_{2i}
      & \sigma_{2i}^2
  \end{bmatrix},
\qquad
\rho_{12i} = \tanh(\eta_{\rho12 i}).
\]

The full observation model is:

\[
\mathbf{y}_i \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu}_i,
    \mathbf{S}_i + \boldsymbol{\Omega}_i
  \right).
\]

The helper `meta_vcov_bivariate()` builds the common row-paired matrix:

```r
V <- meta_vcov_bivariate(
  v1 = c(0.04, 0.03),
  v2 = c(0.05, 0.02),
  cor12 = c(0.4, 0.2)
)
V
```

The stacking is `[y1_1, y2_1, y1_2, y2_2, ...]`. In a fitted model, this known
sampling covariance is added to the estimated residual covariance:

```r
fit_biv <- drmTMB(
  drm_formula(
    mu1 = y1 ~ habitat + meta_V(V = V),
    mu2 = y2 ~ habitat,
    sigma1 = ~ habitat,
    sigma2 = ~ habitat,
    rho12 = ~ habitat
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Here `rho12` is the estimated residual correlation after the known within-study
sampling covariance has already been included. In a model where the estimated
residual component represents between-study heterogeneity, this is the
between-study residual correlation. In other words, `rho12` should not be asked
to explain sampling correlation that belongs in `V`. The current implementation
expects complete bivariate rows and a dense `2n` by `2n` row-paired `V` matrix.

For reporting, translate the fitted residual scales and residual correlation
into marginal residual variances and residual covariance:

```r
s <- sigma(fit_biv)
residual_variance_1 <- s$sigma1^2
residual_variance_2 <- s$sigma2^2
residual_covariance_12 <- rho12(fit_biv) * s$sigma1 * s$sigma2
```

These are residual heterogeneity summaries after the known sampling covariance
has been accounted for. They are not replacements for the supplied `V`.

## Implementation caveats

Dense full covariance is useful for correlated effect sizes and can represent
block-diagonal covariance. Sparse covariance storage is planned for larger
phylogenetic and spatial workloads.

Bivariate known-`V` fitting currently uses a dense row-paired matrix and
complete bivariate rows. Missing single outcomes and sparse matrix storage need
separate likelihood and user-interface decisions.

Negative and missing known variances are rejected or removed through the same
row-retention rules as the model data; full matrices must be symmetric and
positive semidefinite after row subsetting.

Current Gaussian meta-analysis supports `meta_V(V = V)` with ordinary `sigma`
predictors, but `meta_V(V = V) + sigma ~ moderator` can be weakly identified in
some datasets and may give `pdHess = FALSE`. In that case, treat Wald SEs and
Wald intervals for the extra-heterogeneity model as unreliable until a profile,
bootstrap, or simpler `sigma` model supports the target.

Do not put `offset(0.5 * log(v))` in the `sigma` formula to represent known
sampling error. In `drmTMB`, known sampling variance belongs in
`meta_V(V = v)`, and residual heterogeneity belongs in `sigma ~ ...`. Formula
offsets are currently count-family `mu` exposure syntax, so offsets in Gaussian
or meta-analytic `sigma` formulas are intentionally rejected.

The API should keep `family = gaussian()` and `sigma ~ ...`; it should not
introduce `meta_gaussian()` or `tau ~` syntax without a separate design
decision.

For large sparse covariance matrices, incomplete bivariate rows, or dense
full-`V` models that also need non-unit likelihood weights, the current best
practice is to simplify the fitted example, run a sensitivity analysis with
the supported covariance structure, and treat the unsupported path as a design
request rather than forcing it through `weights =`.
