---
title: "Mean effects and residual heterogeneity in meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mean effects and residual heterogeneity in meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Meta-analysis in `drmTMB` is ordinary Gaussian distributional regression with
known sampling covariance. The sampling covariance is supplied with
`meta_known_V(V = V)` inside the location formula; it is not a separate
`meta_gaussian()` family.

The distinction matters. A meta-analysis usually has two sources of variation:
known sampling error from the original studies, and unknown heterogeneity that
the meta-analytic model estimates. In `drmTMB`, known sampling covariance is
`V`, and unknown residual heterogeneity is `sigma`.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

## The model

For a univariate meta-analysis with independent sampling errors, the model is:

\[
y_i \mid \mu_i, \sigma_i, v_i
  \sim \operatorname{Normal}(\mu_i, v_i + \sigma_i^2)
\]

\[
\mu_i = \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma}.
\]

The matching R syntax is:

```r
drmTMB(
  drm_formula(
    yi ~ x1 + x2 + meta_known_V(V = vi),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

Here `vi` contains known sampling variances. If a dataset contains standard
errors, square them before fitting. The `sigma` formula models the extra
heterogeneity left after the known sampling variance has been included. In
meta-analysis papers this extra heterogeneity SD is often written as `tau`;
`drmTMB` keeps the public parameter name `sigma` so that meta-analysis uses the
same distributional grammar as other Gaussian models.

For correlated effect sizes, the same idea is written with a full known
sampling covariance matrix:

\[
\mathbf{y} \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu},
    \mathbf{V} + \operatorname{diag}(\sigma_1^2,\ldots,\sigma_n^2)
  \right).
\]

The right-hand side of `meta_known_V(V = V)` may be a variance column, a vector,
a diagonal matrix, a dense block-diagonal matrix, or a dense full covariance
matrix.

Dense means dense. If `V` is supplied as a full matrix, `drmTMB` stores the
retained covariance as a dense R matrix. This is useful for small to moderate
correlated-effect examples and for likelihood-comparator checks, but it is not
the large-meta-analysis storage route. Large block-diagonal or sparse known
covariance structures need a future sparse or block-sparse path before they
should be treated as scalable.

## Worked example: restoration effects

Suppose we are synthesising effect sizes from studies that compare ecological
restoration with a control condition. The response `yi` is a standardised
effect size for biodiversity. The known sampling variance `vi` comes from each
primary study. We ask two questions:

1. Is the mean restoration effect different in forests and grasslands?
2. Is the unexplained among-effect-size heterogeneity larger in one habitat?

This simulated dataset stands in for the structure of a real ecological
meta-analysis:

```{r}
set.seed(20)
n <- 80
dat <- data.frame(
  study = factor(seq_len(n)),
  habitat = factor(
    rep(c("grassland", "forest"), length.out = n),
    levels = c("grassland", "forest")
  ),
  duration = scale(runif(n, 1, 8))[, 1]
)

dat$vi <- runif(n, 0.015, 0.08)

mu <- 0.12 + 0.22 * (dat$habitat == "forest") + 0.08 * dat$duration
sigma_extra <- exp(-1.65 + 0.55 * (dat$habitat == "forest"))

dat$yi <- rnorm(n, mean = mu, sd = sqrt(dat$vi + sigma_extra^2))
head(dat)
```

The fitted model mirrors the equations above:

```{r}
fit_meta <- drmTMB(
  drm_formula(
    yi ~ habitat + duration + meta_known_V(V = vi),
    sigma ~ habitat
  ),
  family = gaussian(),
  data = dat
)

summary(fit_meta)
```

The `mu` rows describe the average effect-size model. In this simulation the
forest coefficient is positive, so the model estimates a larger mean
restoration effect for forests than for grasslands, after accounting for study
duration and known sampling variance.

The `sigma` rows are on the log-SD scale. They describe extra heterogeneity,
not sampling error. The forest contrast can be read as a multiplicative change
in the residual heterogeneity SD:

```{r}
sigma_coef <- coef(fit_meta, "sigma")
sigma_coef
exp(sigma_coef["habitatforest"])
```

In this run, forest studies have about `r round(exp(sigma_coef["habitatforest"]), 2)`
times the extra heterogeneity SD of grassland studies. That multiplier applies
to `sigma_i`; it does not alter the known sampling variances `vi`.

It is often helpful to translate fitted values back to the response scale and
to the variance scale used in many meta-analysis summaries:

```{r}
meta_report <- data.frame(
  habitat = dat$habitat,
  known_sampling_variance = dat$vi,
  fitted_extra_heterogeneity_sd = sigma(fit_meta)
)
meta_report$fitted_extra_heterogeneity_variance <-
  meta_report$fitted_extra_heterogeneity_sd^2
meta_report$total_observation_variance <-
  meta_report$known_sampling_variance +
  meta_report$fitted_extra_heterogeneity_variance

meta_by_habitat <- aggregate(
  meta_report[c(
    "fitted_extra_heterogeneity_sd",
    "fitted_extra_heterogeneity_variance",
    "total_observation_variance"
  )],
  list(habitat = meta_report$habitat),
  mean
)
meta_by_habitat
```

For this model, the fitted residual heterogeneity SD is larger in forest
studies. The heterogeneity variance column is the same fitted `sigma` model
reported as `sigma^2`. The total observation variance column adds the known
sampling variance `vi`, so it is larger than the estimated heterogeneity
variance alone. A biological interpretation would be: restoration effects are
not only larger on average in forests, they are also less predictable across
forest studies after known sampling uncertainty is accounted for.

Choose the report scale deliberately:

| Quantity | Computation | Interpretation |
|---|---|---|
| known sampling variance | `vi` or the supplied `V` | uncertainty known before fitting |
| extra heterogeneity SD | `sigma(fit_meta)` | residual between-effect-size SD after known sampling variance is included |
| extra heterogeneity variance | `sigma(fit_meta)^2` | variance-scale version of the fitted `sigma` model |
| total observation variance | `vi + sigma(fit_meta)^2` for diagonal `V` | sampling variance plus estimated extra heterogeneity |

This is why `drmTMB` keeps the public parameter name `sigma`: it is the fitted
extra heterogeneity scale in the Gaussian likelihood. If a meta-analysis report
needs the usual variance-facing language, square `sigma` and say that the
reported column is `sigma^2`.

A quick diagnostic check flags optimizer, Hessian, row-filtering, scale, and
known sampling covariance problems before interpretation:

```{r}
check_drm(fit_meta)
```

For full-matrix known covariance fits, the `known_sampling_covariance` row is a
note rather than a problem by itself. It reports the retained matrix dimension,
dense storage, density, size, rank, and conditioning so you can see whether the
current dense route still matches the scale of the analysis.

## Weights are not sampling variances

The top-level `weights =` argument and `meta_known_V(V = V)` answer different
questions. Ordinary likelihood weights multiply each observation's
log-likelihood contribution:

\[
\ell(\theta) =
  \sum_i w_i \ell_i(\theta).
\]

Known sampling covariance changes the covariance model:

\[
\mathbf{y} \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu},
    \mathbf{V} + \boldsymbol{\Omega}_{\mathrm{estimated}}
  \right).
\]

Thus `weights = 1 / vi` is not the same model as
`meta_known_V(V = vi)`. Use `meta_known_V(V = vi)` when `vi` is known sampling
variance. Use `weights =` only when the intended analysis is a weighted
likelihood, such as case weights or externally defined row weights. The
practical consequence is that `weights = 1 / vi` changes how much each row
contributes to the likelihood; it does not put `vi` into the modelled sampling
variance. The current dense full-`V` path rejects non-unit weights until joint
block weighting has a separate design.

## Known sampling variance plus group-level heterogeneity

Some meta-analyses contain multiple effect sizes per study, species, lab, or
site. Then `sigma` and `sd(group)` answer different questions. The next code
block is schematic; it assumes a dataset such as `dat_repeated`, with multiple
rows per study and a study-level `habitat` value that is constant within each
study after missing rows are removed:

```r
fit_study <- drmTMB(
  drm_formula(
    yi ~ treatment + (1 | study) + meta_known_V(V = vi),
    sigma ~ 1,
    sd(study) ~ habitat
  ),
  family = gaussian(),
  data = dat_repeated
)
```

The symbolic model is:

\[
y_i = \mu_i + b_{j[i]} + r_i + s_i,
\qquad
\mu_i = \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu},
\]

\[
b_j \sim \operatorname{Normal}(0, \omega_j^2),
\qquad
\log(\omega_j) =
  \mathbf{x}_{\omega j}^{\top}\boldsymbol{\alpha},
\]

\[
r_i \sim \operatorname{Normal}(0, \sigma_i^2),
\qquad
\mathbf{s} \sim \operatorname{MVN}(\mathbf{0}, \mathbf{V}).
\]

Here `s_i` is the known sampling-error component supplied by
`meta_known_V(V = vi)` or `meta_known_V(V = V)`. The `sigma` formula models
remaining observation-level heterogeneity. The `sd(study)` formula models
`\omega_j`, the study-level random-effect standard deviation. Predictors in
`sd(study) ~ ...` must be constant within each study after missing-data
filtering; otherwise a single study-level random-effect SD would not be well
defined.

After integrating over the random effects, the marginal covariance is:

\[
\boldsymbol{\Omega} =
  \mathbf{V}
  + \operatorname{diag}(\sigma_1^2,\ldots,\sigma_n^2)
  + \mathbf{Z}_{study}
    \operatorname{diag}(\omega_1^2,\ldots,\omega_J^2)
    \mathbf{Z}_{study}^{\top}.
\]

This is the distinction to keep in mind: `sigma ~ habitat` asks whether
effect-size-level residual heterogeneity changes with habitat, while
`sd(study) ~ habitat` asks whether the study-level random-effect SD changes
with habitat.

## Bivariate meta-analysis

Bivariate meta-analysis uses the same Gaussian bivariate route as other
location-coscale models, where coscale means the residual correlation
component `rho12`. The known sampling covariance should include the within-study
covariance between the two effect-size estimates when it is available.

For study \(i\), write:

\[
\mathbf{y}_i =
  \begin{bmatrix} y_{1i} \\ y_{2i} \end{bmatrix},
\qquad
\mathbf{S}_i =
  \begin{bmatrix}
    v_{1i} & c_{12i} \\
    c_{12i} & v_{2i}
  \end{bmatrix}.
\]

The fitted residual covariance is:

\[
\boldsymbol{\Omega}_i =
  \begin{bmatrix}
    \sigma_{1i}^2
      & \rho_{12i}\sigma_{1i}\sigma_{2i} \\
    \rho_{12i}\sigma_{1i}\sigma_{2i}
      & \sigma_{2i}^2
  \end{bmatrix},
\qquad
\rho_{12i} = \tanh(\eta_{\rho12 i}).
\]

The full observation model is:

\[
\mathbf{y}_i \sim
  \operatorname{MVN}
  \left(
    \boldsymbol{\mu}_i,
    \mathbf{S}_i + \boldsymbol{\Omega}_i
  \right).
\]

The helper `meta_vcov_bivariate()` builds the common row-paired matrix:

```r
V <- meta_vcov_bivariate(
  v1 = c(0.04, 0.03),
  v2 = c(0.05, 0.02),
  cor12 = c(0.4, 0.2)
)
V
```

The stacking is `[y1_1, y2_1, y1_2, y2_2, ...]`. In a fitted model, this known
sampling covariance is added to the estimated residual covariance:

```r
fit_biv <- drmTMB(
  drm_formula(
    mu1 = y1 ~ habitat + meta_known_V(V = V),
    mu2 = y2 ~ habitat,
    sigma1 = ~ habitat,
    sigma2 = ~ habitat,
    rho12 = ~ habitat
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Here `rho12` is the estimated residual correlation after the known within-study
sampling covariance has already been included. In a model where the estimated
residual component represents between-study heterogeneity, this is the
between-study residual correlation. In other words, `rho12` should not be asked
to explain sampling correlation that belongs in `V`. The current implementation
expects complete bivariate rows and a dense `2n` by `2n` row-paired `V` matrix.

For reporting, translate the fitted residual scales and residual correlation
into marginal residual variances and residual covariance:

```r
s <- sigma(fit_biv)
residual_variance_1 <- s$sigma1^2
residual_variance_2 <- s$sigma2^2
residual_covariance_12 <- rho12(fit_biv) * s$sigma1 * s$sigma2
```

These are residual heterogeneity summaries after the known sampling covariance
has been accounted for. They are not replacements for the supplied `V`.

## Implementation caveats

Dense full covariance is useful for correlated effect sizes and can represent
block-diagonal covariance. Sparse covariance storage is planned for larger
phylogenetic and spatial workloads.

Bivariate known-`V` fitting currently uses a dense row-paired matrix and
complete bivariate rows. Missing single outcomes and sparse matrix storage need
separate likelihood and user-interface decisions.

Negative and missing known variances are rejected or removed through the same
row-retention rules as the model data; full matrices must be symmetric and
positive semidefinite after row subsetting.

The API should keep `family = gaussian()` and `sigma ~ ...`; it should not
introduce `meta_gaussian()` or `tau ~` syntax without a separate design
decision.

For large sparse covariance matrices, incomplete bivariate rows, or dense
full-`V` models that also need non-unit likelihood weights, the current best
practice is to simplify the fitted example, run a sensitivity analysis with
the supported covariance structure, and treat the unsupported path as a design
request rather than forcing it through `weights =`.
