---
title: "Handling missing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
library(drmTMB)
```

Missing data in `drmTMB` has two different meanings. A response can be missing,
so the observation has no measured outcome for one response cell. A predictor
can be missing, so the model needs a value of a covariate that was not observed.
The current implementation treats those cases differently.

For missing responses, `drmTMB` can keep the original rows and evaluate the
observed-response likelihood. Missing response cells contribute no direct
response likelihood, but the fitted object still knows where those rows were.
For missing predictors, `drmTMB` requires an explicit predictor model through
`mi()` and `impute`. The predictor model is part of the same likelihood as the
response model; it is not a preprocessing fill-in step.

The broadest missing-predictor module covers one missing predictor at a time in a
univariate Gaussian location model. The supported predictor families are
family-aware: Gaussian for continuous predictors, Bernoulli/logit for binary
predictors, cumulative logit for ordered categories, baseline softmax for
unordered categories, beta and zero-one beta for proportions, beta-binomial for
successes out of known trials, Poisson/NB2/zero-truncated NB2 for counts,
lognormal/Gamma for positive continuous predictors, and Tweedie for
semi-continuous predictors with exact zeros. The non-Gaussian response routes
support a `poisson()`, `binomial()`, `nbinom2()`, or `beta()` response with one
binary missing predictor. Complete covariates used to explain the missing
predictor can still be ordinary model-matrix inputs. Transformed `mi()` terms
and multiple missing predictors are later work.

On the response-masking side, `missing = miss_control(response = "include")`
retains and marginalises missing responses for `gaussian()`, bivariate Gaussian,
`student()`, `skew_normal()`, `lognormal()`, `Gamma(link = "log")`,
`tweedie()`, `zero_one_beta()`, `binomial()`, `poisson()`, `nbinom2()`, and
`beta()` fits, plus whole-row `beta_binomial()` and ordered-factor
`cumulative_logit()` fits. Student-t,
lognormal, and Gamma include ordinary random-intercept recovery evidence;
skew-normal, Tweedie, and zero-one beta are fixed-effect routes. These ticks do not automatically extend to
every structured modifier supported by a family.

The response model and the imputation model use two different formula
containers. The main `drmTMB()` model uses `drm_formula()` or its short alias
`bf()`. Entries inside `impute = list(...)` are ordinary two-sided predictor
formulas, such as `body_mass ~ moisture + canopy`, or `impute_model(...)`
objects when a predictor-model family must be stated explicitly. Do not wrap
the `impute` entry itself in `drm_formula()`.

## Capability matrix

The two missing-data modes are validated per response family. A checkmark means
the mode is implemented and recovery-tested; a dash means it is not implemented
and the fit rejects loudly (a family-specific error), so an unsupported request
never silently degrades to a wrong likelihood.

| Response family | `response = "include"` (mask/marginalise) | `predictor = "model"` (`mi()`) |
| --- | :---: | :---: |
| `gaussian()` | ✓ | ✓ (broad predictor-family catalogue) |
| bivariate Gaussian | ✓ (partial `y1`/`y2` pairs) | — |
| `binomial()` | ✓ | ✓ (one binary predictor) |
| `poisson()` | ✓ | ✓ (one binary predictor) |
| `nbinom2()` | ✓ | ✓ (one binary predictor) |
| `beta()` | ✓ | ✓ (one binary predictor) |
| `student()` | ✓ (ordinary random-intercept evidence) | — |
| `lognormal()` | ✓ (ordinary random-intercept evidence) | — |
| `Gamma(link = "log")` | ✓ (ordinary random-intercept evidence) | — |
| `skew_normal()` | ✓ (fixed effects only) | — |
| `tweedie()` | ✓ (fixed effects; zero atom plus continuous component) | — |
| `zero_one_beta()` | ✓ (fixed effects; zero/one atoms plus interior) | — |
| `beta_binomial()` | ✓ (whole success/failure row; ordinary random-intercept evidence) | — |
| `cumulative_logit()` | ✓ (ordered factors; all declared categories observed) | — |
| all other families (`truncated_nbinom2()`, zero-inflated and hurdle routes) | — | — |

The allow-lists are single sources of truth in the package
(`drm_missing_response_families()` and `drm_missing_predictor_families()`), and
an anti-drift test asserts that every family outside them still rejects. Only the
Gaussian response carries the broad predictor-family catalogue tabulated below;
the non-Gaussian responses currently model one binary (Bernoulli/logit) missing
predictor each.

## What is fitted now

The default is still conservative complete-case fitting:

```{r}
miss_control()
```

Use `missing = miss_control(response = "include")` when you want supported
missing responses retained in the fitted object. Use
`missing = miss_control(predictor = "model")` only when the response formula
contains one supported `mi()` term and `impute` supplies the corresponding
predictor model.

| Missing-data case | Current route | Main boundary |
| --- | --- | --- |
| Missing univariate Gaussian response | `missing = miss_control(response = "include")` | Predictors, grouping variables, offsets, weights, and structured-effect inputs must be complete. |
| Missing non-Gaussian response (`student()`, `skew_normal()`, `lognormal()`, `Gamma()`, `tweedie()`, `zero_one_beta()`, `binomial()`, `poisson()`, `nbinom2()`, `beta()`) | `missing = miss_control(response = "include")` | Missing response cells are marginalised out; predictors and other inputs must be complete. Lognormal/Gamma observed responses must be positive; Tweedie allows zero and positive values; zero-one beta needs observed interior data plus optional 0/1 atoms; strict `beta()` needs interior `(0, 1)` responses. |
| Missing bivariate Gaussian response cell | `missing = miss_control(response = "include")` | Partial `y1`/`y2` rows are supported without dense known sampling covariance. |
| Missing beta-binomial response row | `missing = miss_control(response = "include")` | Missing either success or failure masks the whole encoded row; a partial pair never contributes. |
| Missing cumulative-logit response | `missing = miss_control(response = "include")` | Use an ordered factor with explicit levels; every declared category must remain represented among observed rows. |
| One missing numeric predictor | `bf(y ~ z + mi(x), sigma ~ 1)` with `impute = list(x = x ~ z)` | The missing `mi()` variable must be numeric and modelled with one Gaussian predictor model. |
| One missing binary predictor | `bf(y ~ z + mi(treatment), sigma ~ 1)` with `impute = list(treatment = impute_model(treatment ~ z, family = binomial()))` | The missing `mi()` variable must be binary and modelled with one fixed-effect Bernoulli/logit predictor model. |
| Non-Gaussian response with one missing binary predictor | `bf(y ~ z + mi(treatment))`, `family = poisson()`/`binomial()`/`nbinom2()`/`beta()`, with `impute = list(treatment = impute_model(treatment ~ z, family = binomial()))` | The response must be complete; the missing predictor must be binary; zero inflation, random effects, and structured response terms are not implemented for these routes. |
| One missing ordered predictor | `bf(y ~ z + mi(score), sigma ~ 1)` with `impute = list(score = impute_model(score ~ z, family = cumulative_logit()))` | The missing `mi()` variable must be ordered categorical and modelled with one fixed-effect cumulative-logit predictor model. |
| One missing unordered predictor | `bf(y ~ z + mi(habitat), sigma ~ 1)` with `impute = list(habitat = impute_model(habitat ~ z, family = categorical()))` | The missing `mi()` variable must be unordered categorical with at least three levels and modelled with one fixed-effect baseline-category softmax predictor model. |
| One missing strict proportion predictor | `bf(y ~ z + mi(cover), sigma ~ 1)` with `impute = list(cover = impute_model(cover ~ z, family = beta()))` | The missing `mi()` variable must be numeric and strictly inside `(0, 1)`, then modelled with one fixed-effect beta predictor model. |
| One missing boundary proportion predictor | `bf(y ~ z + mi(cover), sigma ~ 1)` with `impute = list(cover = impute_model(cover ~ z, family = zero_one_beta()))` | The missing `mi()` variable must be numeric in `[0, 1]`, may include exact zeros or ones, and is modelled with one fixed-effect zero-one beta predictor model. |
| One missing denominator-aware proportion predictor | `bf(y ~ z + mi(cover), sigma ~ 1)` with `impute = list(cover = impute_model(success ~ z, family = beta_binomial(), trials = trials))` | The `mi()` variable is the proportion used by the response model; the predictor model uses success counts and complete known trial denominators. |
| One missing Poisson count predictor | `bf(y ~ z + mi(abundance), sigma ~ 1)` with `impute = list(abundance = impute_model(abundance ~ z, family = poisson()))` | The missing `mi()` variable must be a non-negative integer count and modelled with one fixed-effect Poisson predictor model. |
| One missing negative-binomial count predictor | `bf(y ~ z + mi(abundance), sigma ~ 1)` with `impute = list(abundance = impute_model(abundance ~ z, family = nbinom2()))` | The missing `mi()` variable must be a non-negative integer count and modelled with one fixed-effect NB2 predictor model. |
| One missing zero-truncated count predictor | `bf(y ~ z + mi(abundance), sigma ~ 1)` with `impute = list(abundance = impute_model(abundance ~ z, family = truncated_nbinom2()))` | The missing `mi()` variable must be a positive integer count and modelled with one fixed-effect zero-truncated NB2 predictor model. |
| One missing lognormal positive predictor | `bf(y ~ z + mi(biomass), sigma ~ 1)` with `impute = list(biomass = impute_model(biomass ~ z, family = lognormal()))` | The missing `mi()` variable must be positive and modelled with one fixed-effect lognormal predictor model. |
| One missing Gamma positive predictor | `bf(y ~ z + mi(biomass), sigma ~ 1)` with `impute = list(biomass = impute_model(biomass ~ z, family = Gamma(link = "log")))` | The missing `mi()` variable must be positive and modelled with one fixed-effect Gamma mean-CV predictor model. |
| One missing Tweedie semi-continuous predictor | `bf(y ~ z + mi(biomass), sigma ~ 1)` with `impute = list(biomass = impute_model(biomass ~ z, family = tweedie()))` | The missing `mi()` variable must be non-negative, may include exact zeros, and is modelled with one fixed-effect Tweedie predictor model. |
| Grouped or structured predictor model | `impute = list(x = x ~ z + (1 | group))` or `impute = list(x = x ~ z + relmat(1 | line, Q = Q))` | One grouped random intercept or one intercept-only structured covariate field. |
| Missing-predictor summaries | `imputed(fit)` | Conditional modes for Gaussian predictors, conditional probabilities for binary predictors, conditional expected scores for ordered predictors, conditional modal categories for unordered predictors, conditional means for strict proportion, boundary proportion, denominator-aware beta-binomial, lognormal, Gamma, and Tweedie predictors, and conditional expected counts for count predictors. |

This is likelihood-based missing-data handling. It is not Bayesian posterior
imputation, not multiple imputation, and not an EM interface.

## Choosing the predictor family

The predictor family should describe the missing covariate, not the response.
This table is the practical starting point:

| Missing predictor type | Use in `impute_model()` | What is integrated or summed |
| --- | --- | --- |
| Continuous on the real line | bare `x ~ z` or `impute_model(x ~ z, family = gaussian())` | a Gaussian latent predictor by Laplace approximation |
| Binary, logical, or two-level factor | `impute_model(x ~ z, family = binomial())` | the two possible states |
| Ordered categories | `impute_model(x ~ z, family = cumulative_logit())` | all ordered levels |
| Unordered categories | `impute_model(x ~ z, family = categorical())` | all unordered levels |
| Continuous proportion in `(0, 1)` | `impute_model(x ~ z, family = beta())` | deterministic beta quadrature nodes |
| Continuous proportion in `[0, 1]` with real zeros/ones | `impute_model(x ~ z, family = zero_one_beta())` | exact zero/one mass plus beta quadrature |
| Successes out of known trials | `impute_model(success ~ z, family = beta_binomial(), trials = trials)` | all integer successes from `0` to `trials` |
| Non-negative counts | `impute_model(x ~ z, family = poisson())` or `impute_model(x ~ z, family = nbinom2())` | deterministic count states |
| Positive counts | `impute_model(x ~ z, family = truncated_nbinom2())` | deterministic positive count states |
| Positive continuous values | `impute_model(x ~ z, family = lognormal())` or `impute_model(x ~ z, family = Gamma(link = "log"))` | deterministic positive quadrature nodes |
| Non-negative semi-continuous values with exact zeros | `impute_model(x ~ z, family = tweedie())` | exact zero mass plus positive quadrature nodes |

Rendering this article is also a smoke test: the worked examples below fit one
model for each supported predictor-family route.

## Relation to gllvmTMB and glmmTMB

The sibling `gllvmTMB` missing-data layer uses the same vocabulary:
`miss_control()`, `mi()`, `impute`, `fit$missing_data`, and `imputed()` for
predictors. `gllvmTMB` adds `predict_missing()` for masked responses; in
`drmTMB` that information comes from `fitted()`, `residuals()`, and
`fit$missing_data`. The `gllvmTMB` multivariate trait engine needs
per-unit products across trait rows, so its missing-predictor implementation is
not a drop-in copy of the scalar `drmTMB` route. The shared contract is the
important part: missing responses are masked or marginalised, missing
predictors are modelled inside the likelihood, and fitted summaries are not
posterior draws.

`glmmTMB` has a different missing-data surface. Its documented `na.action`
argument follows ordinary R model-frame handling: the default drops
observations containing `NA` in model variables, while `na.exclude` still drops
those observations for fitting but returns `NA` predictions and residuals for
excluded cases
([glmmTMB reference](https://glmmtmb.github.io/glmmTMB/reference/glmmTMB.html)).
The `drmTMB` route described here is therefore an in-model missing-data layer,
not a wrapper around `na.action`.

## Missing responses

For a univariate Gaussian response, the included-row likelihood is:

\[
\ell_i =
  I_i \log p(y_i \mid \mu_i, \sigma_i),
\]

where \(I_i = 1\) when the response is observed and \(I_i = 0\) when the
response is missing. A missing response row is retained for row accounting, but
it adds zero direct response likelihood.

This example creates a simple growth data set with three missing response
values:

```{r}
set.seed(20260531)
n <- 40
growth_data <- data.frame(
  temperature = seq(-1.6, 1.6, length.out = n)
)
growth_data$growth <- 0.5 + 0.8 * growth_data$temperature +
  rnorm(n, sd = 0.25)
growth_data$growth[c(6, 17, 32)] <- NA_real_

fit_response <- drmTMB(
  bf(growth ~ temperature, sigma ~ 1),
  family = gaussian(),
  data = growth_data,
  missing = miss_control(response = "include"),
  control = drm_control(se = FALSE)
)
```

`nobs()` reports the number of likelihood-contributing responses, not the
number of retained rows:

```{r}
nobs(fit_response)
fit_response$missing_data$counts
```

The fitted object preserves original-row accounting:

```{r}
head(data.frame(
  original_row = fit_response$missing_data$original_row,
  observed_y = fit_response$missing_data$observed_y,
  fitted = fitted(fit_response),
  residual = residuals(fit_response)
), 10)
```

Rows with missing responses have fitted values because their predictors are
known. Their residuals are `NA` because no response was observed.

## Partial bivariate response pairs

For bivariate Gaussian models, one response can be observed while the other is
missing. A complete pair contributes the bivariate Gaussian density with
`rho12`. A row with only `y1` observed contributes the marginal Gaussian density
for `y1`; a row with only `y2` observed contributes the marginal density for
`y2`. A row with both responses missing contributes zero direct response
likelihood but keeps its row identity.

Symbolically, the row contribution is:

```text
both observed     -> log p(y1_i, y2_i | mu1_i, mu2_i, sigma1_i, sigma2_i, rho12_i)
only y1 observed  -> log p(y1_i | mu1_i, sigma1_i)
only y2 observed  -> log p(y2_i | mu2_i, sigma2_i)
both missing      -> 0
```

The residual correlation `rho12` is learned directly from complete response
pairs. Rows with one observed response still help the marginal location and
scale models, but they do not directly identify `rho12`.

```{r}
trait_data <- data.frame(
  x = seq(-1.4, 1.4, length.out = 36)
)
trait_data$activity <- 0.3 + 0.7 * trait_data$x + rnorm(36, sd = 0.25)
trait_data$boldness <- -0.1 - 0.4 * trait_data$x + rnorm(36, sd = 0.30)
trait_data$activity[c(5, 16, 29)] <- NA_real_
trait_data$boldness[c(8, 16, 31)] <- NA_real_

fit_pair <- drmTMB(
  bf(
    mu1 = activity ~ x,
    mu2 = boldness ~ x,
    sigma1 = ~1,
    sigma2 = ~1,
    rho12 = ~1
  ),
  family = c(gaussian(), gaussian()),
  data = trait_data,
  missing = miss_control(response = "include"),
  control = drm_control(se = FALSE)
)

fit_pair$missing_data$counts
table(fit_pair$missing_data$response_pattern)
```

Use the response-pattern counts before interpreting `rho12`. If complete pairs
are scarce, `drmTMB` warns that the residual correlation is weakly identified.

## Missing predictors

Missing predictors are not handled by dropping a row into the response
likelihood with an arbitrary filled-in covariate. The response model needs the
predictor value, so `drmTMB` asks for an explicit model for that predictor.

For the first implemented route, the response model is a univariate Gaussian
location model:

\[
\begin{aligned}
y_i \mid x_i, \mu_i, \sigma
  &\sim \operatorname{Normal}(\mu_i, \sigma^2),\\
\mu_i
  &= \beta_0 + \beta_1 z_i + \beta_2 x_i.
\end{aligned}
\]

The `impute` formula defines a Gaussian model for the missing predictor:

\[
x_i \sim \operatorname{Normal}(W_i \alpha, \sigma_x^2).
\]

Observed `x` values contribute their predictor-model density and enter the
response model directly. Missing `x` values are latent TMB random effects.
TMB's Laplace approximation integrates them from the joint likelihood.

```{r}
set.seed(20260532)
n <- 48
predictor_data <- data.frame(
  moisture = seq(-1.5, 1.5, length.out = n),
  canopy = cos(seq_len(n) / 5)
)
predictor_data$body_mass_full <- 0.2 + 0.7 * predictor_data$moisture -
  0.2 * predictor_data$canopy + rnorm(n, sd = 0.08)
predictor_data$growth <- 0.6 + 1.1 * predictor_data$body_mass_full -
  0.3 * predictor_data$moisture + rnorm(n, sd = 0.20)
predictor_data$body_mass <- predictor_data$body_mass_full
predictor_data$body_mass[c(7, 19, 34, 43)] <- NA_real_

fit_predictor <- drmTMB(
  bf(growth ~ moisture + mi(body_mass), sigma ~ 1),
  family = gaussian(),
  data = predictor_data,
  impute = list(body_mass = body_mass ~ moisture + canopy),
  missing = miss_control(predictor = "model")
)

coef(fit_predictor, "mu")
imputed(fit_predictor)
```

Read `imputed(fit_predictor)` as a fitted missing-predictor summary. The
`estimate` column is the conditional mode used by the fitted likelihood. The
`std_error` column is a likelihood-based conditional standard error when
`TMB::sdreport()` is available. It is not a posterior interval and it is not a
set of multiple-imputation draws.

## Binary missing predictors

Binary predictors are a separate likelihood problem. They should not be forced
through a Gaussian predictor model just because some values are missing. For a
binary missing predictor, put the predictor formula in the `impute` list and
state the predictor-model family with `impute_model()`:

```{r}
set.seed(20260533)
n <- 50
binary_data <- data.frame(
  z = seq(-1.6, 1.6, length.out = n)
)
binary_data$treatment_full <- as.numeric(
  sin(seq_len(n) * 1.7) + 0.35 * binary_data$z > 0
)
binary_data$growth <- 0.45 + 0.55 * binary_data$z +
  1.25 * binary_data$treatment_full + rnorm(n, sd = 0.08)
binary_data$treatment <- factor(binary_data$treatment_full, levels = c(0, 1))
binary_data$treatment[c(6, 15, 28, 43)] <- NA

fit_binary <- drmTMB(
  bf(growth ~ z + mi(treatment), sigma ~ 1),
  family = gaussian(),
  data = binary_data,
  impute = list(
    treatment = impute_model(treatment ~ z, family = binomial())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_binary, "mi_treatment")
imputed(fit_binary)
```

Here `treatment ~ z` is the formula for the missing predictor itself. It is not
wrapped in `drm_formula()`. Internally, `drmTMB` sums over the two possible
states for each missing treatment value. `imputed(fit_binary)` reports the
fitted conditional probability that the missing predictor is in the second
factor level.

## Non-Gaussian responses with binary missing predictors

The non-Gaussian response routes use the same binary predictor model, but the
response likelihood changes. This is useful when the outcome is a count,
proportion, or Bernoulli/binomial outcome and a two-level covariate is partly
missing. The first example uses a Poisson count response:

```{r}
set.seed(20260543)
n <- 64
poisson_binary_data <- data.frame(
  z = seq(-1.5, 1.8, length.out = n)
)
poisson_binary_data$treatment_full <- as.numeric(
  sin(seq_len(n) * 1.3) + 0.3 * poisson_binary_data$z > 0
)
lambda_count <- exp(
  0.25 + 0.45 * poisson_binary_data$z +
    0.75 * poisson_binary_data$treatment_full
)
poisson_binary_data$count <- pmax(
  0,
  round(lambda_count + sqrt(lambda_count) * cos(seq_len(n) / 4))
)
poisson_binary_data$treatment <- factor(
  poisson_binary_data$treatment_full,
  levels = c(0, 1)
)
poisson_binary_data$treatment[c(8, 19, 31, 46, 57)] <- NA

fit_poisson_binary <- drmTMB(
  bf(count ~ z + mi(treatment)),
  family = poisson(),
  data = poisson_binary_data,
  impute = list(
    treatment = impute_model(treatment ~ z, family = binomial())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_poisson_binary, "mu")
coef(fit_poisson_binary, "mi_treatment")
imputed(fit_poisson_binary)
```

For a missing treatment value, `drmTMB` sums over treatment 0 and treatment 1,
combining the Bernoulli predictor-model probability with the Poisson response
likelihood for the observed count. This route is intentionally narrow: the count
response must be observed, and zero-inflated, grouped, or structured Poisson
response models with `mi()` remain later work.

The overdispersed-count (`nbinom2()`) and proportion (`beta()`) responses reuse
the identical 2-point sum; only the response density and its dispersion change.
The negative-binomial response carries `size = exp(-2 * log_sigma)`:

```{r}
fit_nbinom2_binary <- drmTMB(
  bf(count ~ z + mi(treatment), sigma ~ 1),
  family = nbinom2(),
  data = poisson_binary_data,
  impute = list(
    treatment = impute_model(treatment ~ z, family = binomial())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_nbinom2_binary, "mu")
coef(fit_nbinom2_binary, "mi_treatment")
```

The beta response models an interior proportion with a logit mean and precision
`phi = exp(-2 * log_sigma)`:

```{r}
beta_binary_data <- poisson_binary_data
beta_binary_data$cover <- plogis(
  -0.2 + 0.5 * beta_binary_data$z +
    0.7 * beta_binary_data$treatment_full + 0.15 * cos(seq_len(nrow(beta_binary_data)))
)

fit_beta_binary <- drmTMB(
  bf(cover ~ z + mi(treatment), sigma ~ 1),
  family = beta(),
  data = beta_binary_data,
  impute = list(
    treatment = impute_model(treatment ~ z, family = binomial())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_beta_binary, "mu")
coef(fit_beta_binary, "mi_treatment")
```

A `binomial()` response works the same way with a `cbind(successes, failures)`
or 0/1 outcome. Each route is intentionally narrow in the same way as the
Poisson one: the response must be observed, and zero inflation, random effects,
and structured response terms with `mi()` remain later work.

## Ordered missing predictors

Ordered predictors are also finite-state missing predictors. They need their own
ordinal predictor model rather than a Gaussian model on arbitrary category
codes:

```{r}
set.seed(20260535)
n <- 60
ordered_data <- data.frame(
  z = seq(-1.5, 1.5, length.out = n)
)
ordered_levels <- c("low", "medium", "high")
ordered_data$score_full <- cut(
  ordered_data$z + sin(seq_len(n) * 1.4),
  breaks = c(-Inf, -0.45, 0.55, Inf),
  labels = ordered_levels,
  ordered_result = TRUE
)
ordered_data$growth <- 0.25 + 0.5 * ordered_data$z +
  0.45 * as.numeric(ordered_data$score_full) + rnorm(n, sd = 0.18)
ordered_data$score <- ordered_data$score_full
ordered_data$score[c(8, 18, 36, 51)] <- NA

fit_ordered <- drmTMB(
  bf(growth ~ z + mi(score), sigma ~ 1),
  family = gaussian(),
  data = ordered_data,
  impute = list(
    score = impute_model(score ~ z, family = cumulative_logit())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_ordered, "mi_score")
imputed(fit_ordered)
```

Here the predictor model is a cumulative-logit model for the ordered score.
For a missing score, `drmTMB` sums over all ordered categories. The reported
`estimate` is the fitted conditional expected category score, and the
missing-data metadata stores the conditional probability for each ordered
level.

## Unordered missing predictors

Unordered categorical predictors use a baseline-category softmax model. This is
for nominal categories where the order of the levels is not meaningful:

```{r}
set.seed(20260536)
n <- 66
categorical_data <- data.frame(
  z = seq(-1.7, 1.7, length.out = n)
)
categorical_score <- sin(seq_len(n) / 4) + 0.45 * categorical_data$z
categorical_data$habitat_full <- factor(
  ifelse(
    categorical_score < -0.35,
    "forest",
    ifelse(categorical_score < 0.55, "grass", "wetland")
  ),
  levels = c("forest", "grass", "wetland")
)
habitat_effect <- c(forest = -0.35, grass = 0.2, wetland = 0.75)
categorical_data$growth <- 0.15 + 0.5 * categorical_data$z +
  unname(habitat_effect[as.character(categorical_data$habitat_full)]) +
  rnorm(n, sd = 0.12)
categorical_data$habitat <- categorical_data$habitat_full
categorical_data$habitat[c(7, 16, 31, 48, 60)] <- NA

fit_categorical <- drmTMB(
  bf(growth ~ z + mi(habitat), sigma ~ 1),
  family = gaussian(),
  data = categorical_data,
  impute = list(
    habitat = impute_model(habitat ~ z, family = categorical())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_categorical, "mi_habitat")
imputed(fit_categorical)
```

Here the first factor level is the baseline category in the predictor model.
For a missing habitat value, `drmTMB` sums over all habitat levels.
`imputed(fit_categorical)` reports the fitted modal category code, and the
missing-data metadata stores the fitted conditional probability for each
unordered level.

This is the pattern for broader non-Gaussian missing predictors too: put the
predictor model in the `impute` list, and make the family explicit. The next
sections use the same idea for strict proportion, boundary proportion,
denominator-aware proportion, count, and positive continuous predictors.

## Strict proportion missing predictors

Strict proportion predictors use a beta predictor model. This route is for
continuous proportions inside `(0, 1)`, not exact zeros, exact ones, or counted
successes out of a known denominator:

```{r}
set.seed(20260537)
n <- 72
proportion_data <- data.frame(
  z = seq(-1.8, 1.8, length.out = n)
)
proportion_data$cover_full <- plogis(
  -0.25 + 0.9 * proportion_data$z + 0.18 * sin(seq_len(n) / 5)
)
proportion_data$growth <- 0.35 + 1.25 * proportion_data$cover_full -
  0.30 * proportion_data$z + rnorm(n, sd = 0.10)
proportion_data$cover <- proportion_data$cover_full
proportion_data$cover[c(8, 19, 34, 51, 67)] <- NA_real_

fit_proportion <- drmTMB(
  bf(growth ~ z + mi(cover), sigma ~ 1),
  family = gaussian(),
  data = proportion_data,
  impute = list(
    cover = impute_model(cover ~ z, family = beta())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_proportion, "mi_cover")
coef(fit_proportion, "sigma_mi_cover")
imputed(fit_proportion)
```

For a missing cover value, `drmTMB` integrates over the possible proportion
values using deterministic quadrature under the fitted beta predictor model and
the Gaussian response likelihood. `imputed(fit_proportion)` reports the fitted
conditional quadrature mean for each missing cover value.

## Boundary proportion missing predictors

Use a zero-one beta predictor model when a proportion covariate is continuous
on `[0, 1]` and exact zeros or ones are real observations, not coding mistakes.
The first route uses `cover ~ z` for the predictor-model mean and estimates
constant interior scale, exact-boundary probability `zoi`, and conditional
exact-one probability `coi`:

```{r}
set.seed(20260541)
n <- 78
boundary_prop_data <- data.frame(
  z = seq(-1.8, 1.8, length.out = n)
)
boundary_prop_data$cover_full <- plogis(
  -0.15 + 0.8 * boundary_prop_data$z + 0.12 * sin(seq_len(n) / 4)
)
boundary_prop_data$cover_full[seq(6, n, by = 18)] <- 0
boundary_prop_data$cover_full[seq(13, n, by = 19)] <- 1
boundary_prop_data$growth <- 0.30 + 1.10 * boundary_prop_data$cover_full -
  0.25 * boundary_prop_data$z + rnorm(n, sd = 0.10)
boundary_prop_data$cover <- boundary_prop_data$cover_full
boundary_prop_data$cover[c(8, 21, 39, 58, 73)] <- NA_real_

fit_boundary_proportion <- drmTMB(
  bf(growth ~ z + mi(cover), sigma ~ 1),
  family = gaussian(),
  data = boundary_prop_data,
  impute = list(
    cover = impute_model(cover ~ z, family = zero_one_beta())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_boundary_proportion, "mi_cover")
coef(fit_boundary_proportion, "sigma_mi_cover")
coef(fit_boundary_proportion, "zoi_mi_cover")
coef(fit_boundary_proportion, "coi_mi_cover")
imputed(fit_boundary_proportion)
```

For a missing boundary-proportion value, `drmTMB` integrates over exact zero
mass, exact one mass, and interior beta quadrature under the fitted predictor
model and the Gaussian response likelihood. `imputed(fit_boundary_proportion)`
reports the fitted conditional quadrature mean for each missing cover value.

## Denominator-aware proportion missing predictors

Use a beta-binomial predictor model when the covariate used in the response
model is a proportion, but the observed data are success counts out of known
trial totals. The main formula still uses the proportion through `mi(cover)`.
The `impute_model()` formula uses the success-count column on the left-hand
side and `trials` for the known denominator:

```{r}
set.seed(20260542)
n <- 70
denom_prop_data <- data.frame(
  z = seq(-1.7, 1.7, length.out = n),
  trials = rep(8:16, length.out = n)
)
p_cover <- plogis(-0.20 + 0.80 * denom_prop_data$z)
denom_prop_data$success_full <- qbinom(
  ppoints(n),
  size = denom_prop_data$trials,
  prob = p_cover
)
denom_prop_data$cover_full <-
  denom_prop_data$success_full / denom_prop_data$trials
denom_prop_data$growth <- 0.25 + 1.10 * denom_prop_data$cover_full -
  0.22 * denom_prop_data$z + rnorm(n, sd = 0.10)
denom_prop_data$success <- denom_prop_data$success_full
denom_prop_data$cover <- denom_prop_data$cover_full
denom_prop_data$success[c(8, 21, 39, 58)] <- NA_real_
denom_prop_data$cover[c(8, 21, 39, 58)] <- NA_real_

fit_denominator_proportion <- drmTMB(
  bf(growth ~ z + mi(cover), sigma ~ 1),
  family = gaussian(),
  data = denom_prop_data,
  impute = list(
    cover = impute_model(
      success ~ z,
      family = beta_binomial(),
      trials = trials
    )
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_denominator_proportion, "mi_cover")
coef(fit_denominator_proportion, "sigma_mi_cover")
imputed(fit_denominator_proportion)
```

For a missing success count, `drmTMB` sums over all possible successes from
zero to the known trial total. `imputed(fit_denominator_proportion)` reports
the fitted conditional proportion mean, not a filled-in integer count.

## Count missing predictors

Count predictors use a Poisson predictor model when the mean-variance
relationship is close to Poisson. This route is for non-negative integer
counts, not continuous biomass:

```{r}
set.seed(20260538)
n <- 76
count_data <- data.frame(
  z = seq(-1.6, 1.7, length.out = n)
)
lambda_count <- exp(0.35 + 0.55 * count_data$z)
count_data$abundance_full <- pmax(
  0,
  round(lambda_count + sqrt(lambda_count) * sin(seq_len(n) / 3))
)
count_data$growth <- 0.20 + 0.42 * count_data$abundance_full -
  0.25 * count_data$z + rnorm(n, sd = 0.10)
count_data$abundance <- count_data$abundance_full
count_data$abundance[c(7, 18, 33, 49, 68)] <- NA_real_

fit_count <- drmTMB(
  bf(growth ~ z + mi(abundance), sigma ~ 1),
  family = gaussian(),
  data = count_data,
  impute = list(
    abundance = impute_model(abundance ~ z, family = poisson())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_count, "mi_abundance")
imputed(fit_count)
```

For a missing abundance value, `drmTMB` sums over a deterministic count support
under the fitted Poisson predictor model and the Gaussian response likelihood.
`imputed(fit_count)` reports the fitted conditional expected count for each
missing abundance value.

Use `nbinom2()` instead when the missing count predictor is overdispersed:

```{r}
fit_count_nb <- drmTMB(
  bf(growth ~ z + mi(abundance), sigma ~ 1),
  family = gaussian(),
  data = count_data,
  impute = list(
    abundance = impute_model(abundance ~ z, family = nbinom2())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_count_nb, "mi_abundance")
coef(fit_count_nb, "sigma_mi_abundance")
imputed(fit_count_nb)
```

The negative-binomial route uses the same NB2 parameterization as a count
response model, \(Var(x_i) = \mu_i + \sigma_x^2 \mu_i^2\). Missing counts are
summed over a deterministic count support under the fitted NB2 predictor model.

Use `truncated_nbinom2()` when the predictor is a positive count by design,
such as days until first event among records where the event occurred:

```{r}
positive_count_data <- count_data
positive_count_data$abundance_full <- count_data$abundance_full + 1
positive_count_data$abundance <- positive_count_data$abundance_full
positive_count_data$abundance[c(7, 18, 33, 49, 68)] <- NA_real_
positive_count_data$growth <- 0.20 + 0.42 * positive_count_data$abundance_full -
  0.25 * positive_count_data$z + rnorm(nrow(positive_count_data), sd = 0.10)

fit_count_trunc_nb <- drmTMB(
  bf(growth ~ z + mi(abundance), sigma ~ 1),
  family = gaussian(),
  data = positive_count_data,
  impute = list(
    abundance = impute_model(abundance ~ z, family = truncated_nbinom2())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_count_trunc_nb, "mi_abundance")
coef(fit_count_trunc_nb, "sigma_mi_abundance")
imputed(fit_count_trunc_nb)
```

The zero-truncated route excludes zero from the missing-count support. Use
`nbinom2()` instead when zero is a possible observed predictor value.

## Positive continuous missing predictors

Positive continuous predictors can use a lognormal or Gamma predictor model.
These routes are for strictly positive biomass, concentration, duration, or
size predictors, not exact zeros or semi-continuous variables:

```{r}
set.seed(20260539)
n <- 76
positive_data <- data.frame(
  z = seq(-1.6, 1.6, length.out = n)
)
positive_data$biomass_full <- exp(
  0.15 + 0.55 * positive_data$z + 0.10 * sin(seq_len(n) / 4)
)
positive_data$growth <- 0.40 + 0.85 * positive_data$biomass_full -
  0.25 * positive_data$z + rnorm(n, sd = 0.10)
positive_data$biomass <- positive_data$biomass_full
positive_data$biomass[c(7, 18, 33, 49, 65)] <- NA_real_

fit_positive <- drmTMB(
  bf(growth ~ z + mi(biomass), sigma ~ 1),
  family = gaussian(),
  data = positive_data,
  impute = list(
    biomass = impute_model(biomass ~ z, family = lognormal())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_positive, "mi_biomass")
coef(fit_positive, "sigma_mi_biomass")
imputed(fit_positive)
```

For a missing biomass value, `drmTMB` integrates over log-scale predictor states
under the fitted lognormal predictor model and the Gaussian response likelihood.
`imputed(fit_positive)` reports the fitted conditional quadrature mean for each
missing biomass value.

Use a Gamma predictor model when the positive covariate is better described by
a mean-CV model on the original scale:

```{r}
fit_positive_gamma <- drmTMB(
  bf(growth ~ z + mi(biomass), sigma ~ 1),
  family = gaussian(),
  data = positive_data,
  impute = list(
    biomass = impute_model(biomass ~ z, family = Gamma(link = "log"))
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_positive_gamma, "mi_biomass")
coef(fit_positive_gamma, "sigma_mi_biomass")
imputed(fit_positive_gamma)
```

For a missing biomass value under the Gamma route, `drmTMB` integrates over
positive predictor states under the fitted Gamma mean-CV predictor model and the
Gaussian response likelihood.

## Semi-continuous missing predictors with exact zeros

Use a Tweedie predictor model when the missing covariate is non-negative and can
have exact zeros as part of the same semi-continuous process. The first route
fits the predictor mean and scale, fixes the predictor-model Tweedie power at
1.5, and integrates missing predictor values over the exact zero mass plus a
deterministic positive quadrature grid:

```{r}
set.seed(20260540)
n <- 82
semi_data <- data.frame(
  z = seq(-1.7, 1.8, length.out = n)
)
semi_data$biomass_full <- ifelse(
  seq_len(n) %% 6 %in% c(0, 1),
  0,
  exp(0.10 + 0.45 * semi_data$z) *
    (1 + 0.16 * sin(seq_len(n) / 4))
)
semi_data$growth <- 0.25 + 0.62 * semi_data$biomass_full -
  0.18 * semi_data$z + rnorm(n, sd = 0.10)
semi_data$biomass <- semi_data$biomass_full
semi_data$biomass[c(7, 18, 31, 44, 58, 76)] <- NA_real_

fit_tweedie_predictor <- drmTMB(
  bf(growth ~ z + mi(biomass), sigma ~ 1),
  family = gaussian(),
  data = semi_data,
  impute = list(
    biomass = impute_model(biomass ~ z, family = tweedie())
  ),
  missing = miss_control(predictor = "model"),
  control = drm_control(se = FALSE)
)

coef(fit_tweedie_predictor, "mi_biomass")
coef(fit_tweedie_predictor, "sigma_mi_biomass")
imputed(fit_tweedie_predictor)
```

`imputed(fit_tweedie_predictor)` reports the fitted conditional quadrature mean
for each missing biomass value. This is still a likelihood-based summary, not a
draw from a posterior distribution or a multiple-imputation data set.

## Grouped and structured predictor models

The same one-predictor route can include one grouped random intercept in the
predictor model:

```r
drmTMB(
  bf(growth ~ moisture + mi(body_mass), sigma ~ 1),
  family = gaussian(),
  data = predictor_data,
  impute = list(body_mass = body_mass ~ moisture + (1 | site)),
  missing = miss_control(predictor = "model")
)
```

Use this when missingness in the predictor is tied to a grouping level and a
separate covariate model is scientifically defensible. The grouped effect is a
covariate-model random intercept, not a response-model random effect.

Structured covariate models are also explicit. For example, a missing
species-level trait can use a known relatedness matrix:

```r
drmTMB(
  bf(growth ~ moisture + mi(body_mass), sigma ~ 1),
  family = gaussian(),
  data = predictor_data,
  impute = list(body_mass = body_mass ~ moisture + relmat(1 | line, Q = Q)),
  missing = miss_control(predictor = "model")
)
```

The current structured route supports one intercept-only `phylo()`,
coordinate `spatial()`, `animal()`, or `relmat()` covariate field. It does not
automatically inherit structure from the response model. That explicitness is
intentional: a model for a missing trait is a scientific model for the trait,
not just a technical patch for a response regression.

## Inspect the missing-data metadata

Every fitted route records a `fit$missing_data` object. The most useful fields
are:

| Field | Meaning |
| --- | --- |
| `version` | The fitted implementation slice, such as `"MD1"`, `"MD2"`, `"MD3a"`, `"MD3b"`, `"MD4"`, `"MD6a"`, `"MD6b"`, `"MD6c"`, `"MD7a"`, `"MD7b"`, `"MD7c"`, `"MD7d"`, `"MD7e"`, `"MD7f"`, `"MD8a"`, `"MD8b"`, `"MD8c"`, `"MD9a"` (Poisson-response `mi()`), `"MD-binomial-mi"`, `"MD-nbinom2-mi"`, or `"MD-beta-mi"`. |
| `response_policy` | Whether response rows were dropped or included. |
| `predictor_policy` | Whether missing predictors fail or are modelled. |
| `original_row` | Row indices in the user's original data. |
| `model_row` | Row indices in the retained model frame. |
| `observed_y` | Logical response mask for univariate fits, or a two-column mask for bivariate fits. |
| `counts` | Retained rows, observed responses, missing responses, and likelihood-contributing rows. |
| `predictors` | Metadata and fitted values for explicitly modelled `mi()` predictors. |

For ordinary user work, prefer `nobs()`, `fitted()`, `residuals()`, and
`imputed()` over hand-parsing internals. The metadata is there so row
accounting stays auditable.

## Current boundaries

The current implementation deliberately keeps several features out of scope:

- EM, profile, and REML missing-data engines are not implemented.
- Multiple missing predictors are not implemented.
- Missing predictors in non-Gaussian response models are implemented for the
  `poisson()`, `binomial()`, `nbinom2()`, and `beta()` responses, each with one
  binary (Bernoulli/logit) missing predictor. Other non-Gaussian responses, and
  non-binary missing predictors on non-Gaussian responses, remain later work.
- Hurdle count predictor models, transformed `mi()` terms, interactions inside
  `mi()`, splines inside `mi()`, and predictor-model random slopes are not
  implemented.
- Grouped and structured binary, ordered, unordered, beta/proportion,
  zero-one beta, Poisson/NB2, lognormal, Gamma, or Tweedie predictor models
  are not implemented.
- Dense known sampling-covariance matrices with bivariate partial-response rows
  are not implemented.
- `imputed()` does not summarize missing responses.
- Simulated imputations, posterior means, credible intervals, and
  multiple-imputation pooling are not implemented.
- Measurement-error models are a later modelling lane, not part of this
  missing-data interface.

When you need one of those routes, treat the current error as a design boundary
rather than a failed fit. For now, fit the supported likelihood-based route, use
complete-case fitting, or handle the unsupported missingness outside `drmTMB`
with a method whose assumptions you can state.
