---
title: "Checking and using fitted models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking and using fitted models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

After fitting a distributional model, the next question is not only "what are
the coefficients?" A useful workflow checks the fit, extracts the fitted
distributional parameters, predicts to meaningful covariate values, examines
residuals, and simulates from the fitted model.

This article uses a small growth example, but the same workflow applies to
other Gaussian location-scale models in `drmTMB`.

## Fit the model

Suppose growth is measured along a temperature gradient in two habitat types.
The model asks whether temperature and habitat change expected growth, and
whether temperature also changes residual variability.

Symbolically:

```text
growth_i | mu_i, sigma_i ~ Normal(mu_i, sigma_i^2)
mu_i = beta_0 + beta_1 temperature_i + beta_2 habitat_i
log(sigma_i) = gamma_0 + gamma_1 temperature_i
```

Matching R syntax:

```{r}
library(drmTMB)

set.seed(12)
n <- 120
fish <- data.frame(
  temperature = runif(n, -1.5, 1.5),
  habitat = factor(sample(c("reef", "kelp"), n, replace = TRUE))
)
eta_mu <- 1 + 0.7 * fish$temperature + 0.35 * (fish$habitat == "kelp")
eta_sigma <- -0.5 + 0.25 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit <- drmTMB(
  bf(growth ~ temperature + habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Here `growth ~ temperature + habitat` defines the location model, and
`sigma ~ temperature` defines the residual standard deviation model on the
log scale.

## Check before interpreting

Run `check_drm()` before interpreting estimates:

```{r}
check_drm(fit)
```

A clean diagnostic table does not prove that the model is scientifically
correct, but it catches common problems early: optimizer convergence, gradients,
optimizer evaluation counts, finite objective values, Hessian status, dropped
rows, finite fixed-effect standard errors, positive scale values, residual
`rho12` values near the boundary, Student-t `nu` values near the boundary or
close to Gaussian behaviour, known sampling covariance summaries, random-effect
standard deviations near zero, and random-effect design issues when random
effects are present.

## Read coefficients by parameter

Each distributional parameter has its own coefficient vector:

```{r}
coef(fit, "mu")
coef(fit, "sigma")
```

The `sigma` coefficients are on the log-standard-deviation scale. A positive
temperature coefficient in `sigma` means that residual variability increases
with temperature. Exponentiating the coefficient gives a multiplicative change
in residual standard deviation for a one-unit temperature increase.

`summary()` puts the fixed-effect estimates, response-scale scale, shape, and
correlation summaries, log likelihood, and convergence code in one place:

```{r}
summary(fit)
```

For confidence intervals, keep the fast default summary separate from the
expensive profile step. Wald intervals are available for fixed effects:

```{r}
summary(fit, conf.int = TRUE)
```

Profile intervals are available for selected direct targets such as constant
`sigma`, random-effect SDs, random-effect correlations, and constant residual
`rho12`:

```{r, eval = FALSE}
summary(fit, conf.int = TRUE, method = "profile", ci_parm = "sigma")
```

Use `profile_targets(fit)` before asking for a profile interval. A row is
profile-ready only when it maps to a direct TMB target and the fitted object
kept its TMB object. Interval-aware output uses `conf.status` to explain rows
without intervals: `newdata_required` means profile a supplied row with
`confint(..., newdata = ...)`, while `derived_interval_unavailable` means the
point estimate combines several quantities and needs a later derived-profile
method. Successful profile rows also include `profile.boundary` and
`profile.message`; values other than `"ok"` tell you to check boundary or
near-correlation-limit behaviour before interpreting the interval.

## Predict distributional parameters

`predict()` returns one distributional parameter at a time. For fitted rows,
`fitted(fit)` and `predict(fit, dpar = "mu")` return the same location values
for this fixed-effect model:

```{r}
head(fitted(fit))
head(predict(fit, dpar = "mu"))
head(sigma(fit))
```

For new covariate values, supply `newdata`. These are fixed-effect,
population-level predictions:

```{r}
new_fish <- data.frame(
  temperature = c(-1, 0, 1),
  habitat = factor("reef", levels = levels(fish$habitat))
)

predict(fit, newdata = new_fish, dpar = "mu")
predict(fit, newdata = new_fish, dpar = "sigma")
predict(fit, newdata = new_fish, dpar = "sigma", type = "link")
```

Use `predict_parameters()` when the interpretation task needs several
distributional parameters on the same grid:

```{r}
predict_parameters(fit, newdata = new_fish, dpar = c("mu", "sigma"))
```

Use `marginal_parameters()` when a grid should be reduced to group-level
interpretation summaries:

```{r}
marginal_parameters(fit, newdata = new_fish, dpar = c("mu", "sigma"))
```

For this Gaussian model, the response-scale `sigma` predictions are residual
standard deviations. The link-scale predictions are the corresponding
log-standard deviations. Report residual variance as `sigma(fit)^2` or
`predict(fit, newdata = new_fish, dpar = "sigma")^2`; do not square the
`sigma` coefficient, because coefficients are fitted on the log-SD scale.

In meta-analytic Gaussian models with `meta_known_V(V = V)`, `sigma(fit)`
still returns the modelled residual heterogeneity scale. Simulation and
Pearson residuals combine the known sampling covariance with that residual
scale internally.

## Inspect residuals

Response residuals are observed minus fitted location:

```{r}
head(residuals(fit))
```

Pearson residuals divide by the fitted observation scale:

```{r}
head(residuals(fit, type = "pearson"))
```

For bivariate Gaussian fits, `residuals(fit, type = "pearson")` returns a
two-column matrix whitened by the fitted `sigma1`, `sigma2`, and residual
correlation `rho12`.

## Simulate from the fitted model

Simulation is a compact way to ask whether the fitted model generates data on
the same scale as the observed response:

```{r}
sims <- simulate(fit, nsim = 3, seed = 1)
head(sims)
```

For this univariate model, the output has one column per simulated data set.
For bivariate Gaussian models, simulations are returned in paired columns such
as `sim_1_y1` and `sim_1_y2`.

## A compact post-fit checklist

For routine fitted models, the post-fit loop is:

1. `check_drm(fit)` before interpreting estimates.
2. `coef(fit, dpar)` or `summary(fit)` for parameter-specific coefficients.
3. `predict(fit, dpar = ...)` for fitted or new-data distributional
   parameters.
4. `residuals(fit, type = "pearson")` to inspect standardized lack of fit.
5. `simulate(fit)` to compare fitted-model behaviour with the observed data.

The same loop will extend as more families arrive. The parameter names change,
but the discipline stays the same: check the fit, then interpret each
distributional parameter on its own scale.
