---
title: "Checking and using fitted models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking and using fitted models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

After fitting a distributional model, the next question is not only "what are
the coefficients?" A useful workflow checks the fit, extracts the fitted
distributional parameters, predicts to meaningful covariate values, examines
residuals, and simulates from the fitted model.

This article uses a small growth example, but the same workflow applies to
other Gaussian location-scale models in `drmTMB`.

## Fit the model

Suppose growth is measured along a temperature gradient in two habitat types.
The model asks whether temperature and habitat change expected growth, and
whether temperature also changes residual variability.

Symbolically:

\[
\begin{aligned}
\text{growth}_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= \beta_0 + \beta_1 \text{temperature}_i
     + \beta_2 \mathbb{1}(\text{habitat}_i = \text{kelp}),\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{temperature}_i .
\end{aligned}
\]

Matching R syntax:

```{r}
library(drmTMB)

set.seed(12)
n <- 120
fish <- data.frame(
  temperature = runif(n, -1.5, 1.5),
  habitat = factor(sample(c("reef", "kelp"), n, replace = TRUE))
)
eta_mu <- 1 + 0.7 * fish$temperature + 0.35 * (fish$habitat == "kelp")
eta_sigma <- -0.5 + 0.25 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit <- drmTMB(
  bf(growth ~ temperature + habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Here `growth ~ temperature + habitat` defines the location model, and
`sigma ~ temperature` defines the residual standard deviation model on the
log scale.

## Check before interpreting

Run `check_drm()` before interpreting estimates:

```{r}
check_drm(fit)
```

A clean diagnostic table does not prove that the model is scientifically
correct, but it catches common problems early: optimizer convergence, gradients,
optimizer evaluation counts, finite objective values, Hessian status, dropped
rows, finite fixed-effect standard errors, positive scale values, residual
`rho12` values near the boundary, Student-t `nu` values near the boundary or
close to Gaussian behaviour, known sampling covariance summaries, random-effect
standard deviations near zero, and random-effect design issues when random
effects are present.

## Read the ordinary summary first

The ordinary R habit is still the right first habit in `drmTMB`:

```{r}
summary(fit)
```

`summary()` puts the fixed-effect estimates, response-scale scale, shape,
random-effect SD, correlation summaries, fitted covariance summaries, derived
variance ratios, log likelihood, and convergence code in one place when those
quantities are present. The fixed-effect table includes standard errors when
`TMB::sdreport()` was computed successfully.

Read the printed summary as a map, then use a more specific extractor only
when you need more detail:

| Summary component | What it reports | Use it for |
| --- | --- | --- |
| `coefficients` | Fixed-effect coefficients for each distributional parameter, usually on the model's link scale. | Direction, sign, and approximate uncertainty of terms in `mu`, `sigma`, `rho12`, shape, or zero-inflation formulas. |
| `parameters` | Direct response-scale parameters, fitted parameter ranges, random-effect SDs, and fitted correlations. | Constant `sigma`, residual `rho12`, `sd:mu:(1 | group)`, and quick response-scale uncertainty when `std_error` is available. |
| `covariance` | Fitted random-effect covariance and correlation rows for supported covariance blocks. | Latent correlation layers such as group-level, phylogenetic, or structured-effect covariance summaries. |
| `derived` | Simple variance-ratio summaries whose ingredients are unambiguous. | Repeatability or phylogenetic signal point estimates, including the random-effect and residual variances used in the ratio. |
| `confint` | Confidence-interval rows only when `conf.int = TRUE`. | Checking which requested fixed-effect or profile-likelihood intervals were actually computed. |

The printed summary is meant to answer "what did this fitted model estimate?"
For workflows that need the full object behind a row, use `fixef()`, `sigma()`,
`rho12()`, `ranef()`, `corpairs()`, or `profile_targets()` after reading the
summary.

For models with random effects, `summary()` is also the first place to look for
variance components. This compact example adds a site random intercept to the
location model:

```{r}
set.seed(13)
n_site <- 18
n_per_site <- 6
site <- factor(rep(seq_len(n_site), each = n_per_site))
site_effect <- rnorm(n_site, sd = 0.45)

fish_site <- data.frame(
  site = site,
  temperature = runif(n_site * n_per_site, -1.5, 1.5),
  habitat = factor(
    sample(c("reef", "kelp"), n_site * n_per_site, replace = TRUE)
  )
)

site_mu <- site_effect[as.integer(fish_site$site)]
fish_site$growth <- rnorm(
  nrow(fish_site),
  mean = 1 + 0.7 * fish_site$temperature +
    0.35 * (fish_site$habitat == "kelp") + site_mu,
  sd = 0.35
)

fit_site <- drmTMB(
  bf(growth ~ temperature + habitat + (1 | site), sigma ~ 1),
  family = gaussian(),
  data = fish_site
)

summary(fit_site)
```

In that output, the `sd:mu:(1 | site)` row is the fitted random-intercept
standard deviation on the response scale. Squaring it gives the corresponding
random-effect variance component. When the ingredients are unambiguous, the
`derived` component also reports repeatability and shows both the
random-effect variance and the residual variance used in that ratio.

When `TMB::sdreport()` succeeds, direct response-scale parameter rows such as
constant `sigma` and `sd:mu:(1 | site)` also include delta-method standard
errors. Treat those as quick local approximations. For fitted direct SD or
correlation targets, profile-likelihood confidence intervals remain the safer
interval route. `drmTMB` is a frequentist package, so these are confidence
intervals, not Bayesian credible intervals:

```{r}
profile_targets(fit_site)
```

```{r, eval = FALSE}
summary(
  fit_site,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "sd:mu:(1 | site)"
)
```

Each distributional parameter has its own coefficient vector:

```{r}
coef(fit, "mu")
coef(fit, "sigma")
```

The `sigma` coefficients are on the log-standard-deviation scale. A positive
temperature coefficient in `sigma` means that residual variability increases
with temperature. Exponentiating the coefficient gives a multiplicative change
in residual standard deviation for a one-unit temperature increase.

Use the scale of the fitted parameter when deciding what to report:

| Output | Fitted scale | Report when the question is... |
| --- | --- | --- |
| `coef(fit, "mu")` | response units | expected growth, treatment effects, or mean trait differences |
| `coef(fit, "sigma")` | log residual SD | multiplicative residual-SD or residual-variance contrasts |
| `sigma(fit)` | residual SD | fitted predictability or extra heterogeneity on the response scale |
| `rho12(fit)` | residual correlation | residual coupling between two responses after means and residual SDs are modelled |
| `corpairs(fit)` | fitted correlation rows | group-level, phylogenetic, or residual correlation layers, depending on `level` |
| `profile_targets(fit)` | interval target inventory | whether a fitted SD or correlation target can receive a profile interval |

For confidence intervals, keep the fast default summary separate from the
expensive profile step. Wald 95% confidence intervals are available for fixed
effects:

```{r}
summary(fit, conf.int = TRUE)
```

Profile-likelihood 95% confidence intervals are available for selected direct
targets such as constant `sigma`, random-effect SDs, random-effect
correlations, and constant residual `rho12`. Fixed effects keep Wald
confidence intervals in the same profile summary unless fixed-effect profile
targets are selected:

```{r, eval = FALSE}
summary(fit, conf.int = TRUE, method = "profile", ci_parm = "sigma")
```

Use the profile call when a report needs fixed-effect confidence intervals and
profile-likelihood intervals for variance components, SDs, or correlations in
one summary.

Use `profile_targets(fit)` before asking for a profile interval. A row is
profile-ready only when it maps to a direct TMB target and the fitted object
kept its TMB object. Interval-aware output uses `conf.status` to explain rows
without intervals: `newdata_required` means profile a supplied row with
`confint(..., newdata = ...)`, while `derived_interval_unavailable` means the
point estimate combines several quantities and needs a later derived-profile
method. Successful profile rows also include `profile.boundary` and
`profile.message`; values other than `"ok"` tell you to check boundary or
near-correlation-limit behaviour before interpreting the interval.

Read `conf.status` as an action column:

| `conf.status` | What it means | What to do next |
| --- | --- | --- |
| `wald` | A Wald interval was returned for a fixed-effect row. | Use it for routine fixed-effect summaries, while remembering it is asymptotic. |
| `profile` | A profile-likelihood interval was returned. | Inspect `profile.boundary` and `profile.message` before treating the interval as stable. |
| `profile_ready` | The row is a direct profile target, but the current summary call did not request it. | Use the row's `parm` value in `confint(..., method = "profile")` or in `summary(..., method = "profile", ci_parm = ...)`. |
| `newdata_required` | The fitted surface needs a concrete row before it can be profiled. | Build a biologically meaningful `newdata` row and call `confint(..., newdata = ...)`. |
| `derived_interval_unavailable` | The point estimate combines multiple fitted quantities and has no validated interval method yet. | Report the point estimate as exploratory, or simplify to a direct target if interval support is essential. |
| `wald_unavailable` | Wald uncertainty is not available for this row, or `sdreport()` was skipped or failed. | Use a profile-ready target when one exists; otherwise report the missing interval explicitly. |
| `target_unavailable` or `profile_unavailable` | The row is descriptive or has no current direct interval target. | Treat the interval as unsupported and check `profile_targets(fit)` before trying another route. |
| `not_requested` | The table is reporting a point estimate without an interval request. | Refit is not needed; request intervals only for the rows you plan to interpret. |

## Predict distributional parameters

`predict()` returns one distributional parameter at a time. For fitted rows,
`fitted(fit)` and `predict(fit, dpar = "mu")` return the same location values
for this fixed-effect model:

```{r}
head(fitted(fit))
head(predict(fit, dpar = "mu"))
head(sigma(fit))
```

For new covariate values, supply `newdata`. These are fixed-effect,
population-level predictions. You can write the grid directly:

```{r}
new_fish <- data.frame(
  temperature = c(-1, 0, 1),
  habitat = factor("reef", levels = levels(fish$habitat))
)

predict(fit, newdata = new_fish, dpar = "mu")
predict(fit, newdata = new_fish, dpar = "sigma")
predict(fit, newdata = new_fish, dpar = "sigma", type = "link")
```

Or build an explicit prediction grid. This keeps the focal term, conditioned
nuisance predictors, and grid rule attached to the data:

```{r}
temperature_grid <- prediction_grid(
  fit,
  focal = "temperature",
  at = list(temperature = c(-1, 0, 1)),
  condition = list(habitat = "reef")
)

attr(temperature_grid, "prediction_grid")
```

Use `predict_parameters()` when the interpretation task needs several
distributional parameters on the same grid:

```{r}
predict_parameters(fit, newdata = temperature_grid, dpar = c("mu", "sigma"))
```

Use an empirical grid when the question is about the fitted response at focal
temperatures while averaging over the fitted-row covariate distribution. Here
each requested temperature is crossed with the fitted model rows, so the
habitat distribution is the one in the fitted data:

```{r}
temperature_empirical <- prediction_grid(
  fit,
  focal = "temperature",
  at = list(temperature = c(-1, 0, 1)),
  margin = "empirical"
)

attr(temperature_empirical, "prediction_grid")
```

Then reduce the empirical grid by the focal term:

```{r}
marginal_parameters(
  fit,
  newdata = temperature_empirical,
  dpar = c("mu", "sigma"),
  by = "temperature"
)
```

These helpers are tables, not plotting functions. They are the current Phase 17
surface for future visualization work: each row names the distributional
parameter, fitted component, prediction scale, estimate, interval status, and
grid values. These first tables are point-estimate summaries, so
`conf.status = "not_requested"` and
`interval_source = "not_available"` tell you that no interval has been
computed.

The first optional plotting helper consumes that table directly:

```{r}
pred_temperature <- predict_parameters(
  fit,
  newdata = temperature_grid,
  dpar = c("mu", "sigma")
)

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_parameter_surface(pred_temperature, x = "temperature", colour = "habitat")
}
```

The figure is a convenience layer, not a new estimand. Keep reference grids,
marginalization choices, and interval sources visible instead of hiding those
decisions inside a plot.

For reports, pair model surfaces with the raw response pattern. The raw-data
panel uses the observed response scale:

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    fish,
    ggplot2::aes(x = temperature, y = growth, colour = habitat)
  ) +
    ggplot2::geom_point(alpha = 0.65) +
    ggplot2::labs(
      x = "Temperature",
      y = "Observed growth",
      colour = "Habitat"
    )
}
```

The fitted panels then use the prediction table. Keep `mu` and `sigma` on their
own y-axes because they are different biological quantities. When a plot keeps
one distributional parameter, `plot_parameter_surface()` labels the y-axis with
the parameter and prediction scale:

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_parameter_surface(
    pred_temperature,
    x = "temperature",
    colour = "habitat",
    dpar = "mu",
    facet = NULL
  )
}
```

```{r}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_parameter_surface(
    pred_temperature,
    x = "temperature",
    colour = "habitat",
    dpar = "sigma",
    facet = NULL
  )
}
```

Do not put raw `growth` points on the `sigma` panel. Raw observations show the
response; fitted `sigma` shows the residual standard deviation implied by the
model. If you add interval ribbons later, check the table first and draw them
only when `interval_source` names a real source such as `"wald"` or
`"profile"`.

For this Gaussian model, the response-scale `sigma` predictions are residual
standard deviations. The link-scale predictions are the corresponding
log-standard deviations. Report residual variance as `sigma(fit)^2` or
`predict(fit, newdata = new_fish, dpar = "sigma")^2`; do not square the
`sigma` coefficient, because coefficients are fitted on the log-SD scale.

In meta-analytic Gaussian models with `meta_known_V(V = V)`, `sigma(fit)`
still returns the modelled residual heterogeneity scale. Simulation and
Pearson residuals combine the known sampling covariance with that residual
scale internally.

## Inspect residuals

Response residuals are observed minus fitted location:

```{r}
head(residuals(fit))
```

Pearson residuals divide by the fitted observation scale:

```{r}
head(residuals(fit, type = "pearson"))
```

For bivariate Gaussian fits, `residuals(fit, type = "pearson")` returns a
two-column matrix whitened by the fitted `sigma1`, `sigma2`, and residual
correlation `rho12`.

## Simulate from the fitted model

Simulation is a compact way to ask whether the fitted model generates data on
the same scale as the observed response:

```{r}
sims <- simulate(fit, nsim = 3, seed = 1)
head(sims)
```

For this univariate model, the output has one column per simulated data set.
For bivariate Gaussian models, simulations are returned in paired columns such
as `sim_1_y1` and `sim_1_y2`.

## A compact post-fit checklist

For routine fitted models, the post-fit loop is:

1. `check_drm(fit)` before interpreting estimates.
2. `summary(fit)` for fixed effects, response-scale parameters, variance
   components, derived summaries, interval status, and the extractor to use
   next.
3. `profile_targets(fit)` before asking for profile-likelihood intervals on
   fitted SD or correlation targets.
4. `predict(fit, dpar = ...)` for fitted or new-data distributional
   parameters.
5. `residuals(fit, type = "pearson")` to inspect standardized lack of fit.
6. `simulate(fit)` to compare fitted-model behaviour with the observed data.

The same loop will extend as more families arrive. The parameter names change,
but the discipline stays the same: check the fit, then interpret each
distributional parameter on its own scale.
