---
title: "Model selection with AIC and BIC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model selection with AIC and BIC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Model selection comes after candidate models have been chosen, fitted, and
checked. In `drmTMB`, AIC and BIC can compare fitted distributional-regression
models, but they do not decide whether a candidate set is biologically sensible
or whether a weak fit should be interpreted.

AIC and BIC both use the fitted log likelihood and the number of estimated
parameters:

```text
AIC = -2 * logLik + 2 * k
BIC = -2 * logLik + log(n) * k
```

`k` is the model degrees of freedom and `n` is the number of observations used
by the fitted model. BIC penalizes extra parameters more strongly than AIC once
`n > exp(2)`. Compare models only when they were fitted to the same response
scale and the same analysis rows. Do not compare a model fit to raw counts with
one fit to transformed counts, or two fits that silently dropped different
missing rows.

For Gaussian mixed models, use ML (`REML = FALSE`) when AIC or BIC compares
different fixed-effect formulas. REML is useful for Gaussian variance-component
estimation inside a fixed mean structure, but the restricted likelihood depends
on the fixed-effect design that was integrated out.

```{r setup}
library(drmTMB)

criterion_converged <- function(fit) {
  isTRUE(fit$opt$convergence == 0)
}

criterion_table <- function(...) {
  models <- list(...)
  out <- data.frame(
    model = names(models),
    AIC = vapply(models, stats::AIC, numeric(1)),
    BIC = vapply(models, stats::BIC, numeric(1)),
    converged = vapply(models, criterion_converged, logical(1)),
    stringsAsFactors = FALSE
  )
  out$delta_AIC <- out$AIC - min(out$AIC)
  out$delta_BIC <- out$BIC - min(out$BIC)
  out
}
```

## Gaussian REML After Mean-Model Selection

For Gaussian mixed models, first compare different fixed-effect formulas with
ML. This example includes a random intercept for repeated measurements and a
noise predictor `z` that was not used to generate the response.

```{r gaussian-reml-example}
set.seed(2404)
n_id <- 12L
n_each <- 6L
id <- factor(rep(seq_len(n_id), each = n_each))
x <- rnorm(n_id * n_each)
z <- rnorm(n_id * n_each)
u <- rnorm(n_id, sd = 0.65)
mixed_dat <- data.frame(
  id = id,
  x = x,
  z = z,
  y = 0.2 + 0.7 * x + u[id] + rnorm(n_id * n_each, sd = 0.45)
)

fit_mixed_x_ml <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ 1),
  family = gaussian(),
  data = mixed_dat
)

fit_mixed_x_z_ml <- drmTMB(
  bf(y ~ x + z + (1 | id), sigma ~ 1),
  family = gaussian(),
  data = mixed_dat
)

criterion_table(`y ~ x` = fit_mixed_x_ml, `y ~ x + z` = fit_mixed_x_z_ml)
```

Once the fixed-effect mean structure is chosen, refit that structure with
`REML = TRUE` when the target is Gaussian variance-component estimation. The
first REML slice covers ordinary univariate Gaussian random intercepts and
slopes in `mu` with intercept-only `sigma`.

```{r gaussian-reml-fit}
fit_mixed_x_reml <- drmTMB(
  bf(y ~ x + (1 | id), sigma ~ 1),
  family = gaussian(),
  data = mixed_dat,
  REML = TRUE
)

ll_reml <- logLik(fit_mixed_x_reml)
data.frame(
  estimator = fit_mixed_x_reml$estimator,
  restricted_logLik = as.numeric(ll_reml),
  df = attr(ll_reml, "df"),
  residual_sigma = unname(sigma(fit_mixed_x_reml)[1]),
  id_sd = unname(fit_mixed_x_reml$sdpars$mu["(1 | id)"])
)
```

When `lme4` is installed, the restricted log likelihood for this overlap model
matches `lme4::lmer(..., REML = TRUE)`.

```{r gaussian-reml-lme4}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit_lme4_reml <- lme4::lmer(
    y ~ x + (1 | id),
    data = mixed_dat,
    REML = TRUE
  )
  data.frame(
    engine = c("drmTMB", "lme4"),
    restricted_logLik = c(
      as.numeric(logLik(fit_mixed_x_reml)),
      as.numeric(logLik(fit_lme4_reml))
    )
  )
}
```

## Tail Assumption

A robust Student-t model is useful when the biological process or measurement
process can produce unusually large residuals. Here the data are generated
with heavy-tailed residuals, then fitted with a Gaussian model and a Student-t
model that uses the same `mu` and `sigma` formulas.

```{r tail-example}
set.seed(2401)
n <- 220
x <- rnorm(n)
tail_dat <- data.frame(x = x)
tail_dat$y <- 0.2 + 0.7 * x + exp(-0.25) * rt(n, df = 4)

fit_tail_gaussian <- drmTMB(
  bf(y ~ x, sigma ~ 1),
  family = gaussian(),
  data = tail_dat
)

fit_tail_student <- drmTMB(
  bf(y ~ x, sigma ~ 1, nu ~ 1),
  family = student(),
  data = tail_dat
)

criterion_table(
  Gaussian = fit_tail_gaussian,
  `Student-t` = fit_tail_student
)
```

The lower AIC/BIC model is the better criterion fit inside this two-model
candidate set. That result is not a licence to ignore diagnostics: run
`check_drm()` and inspect whether the fitted Student-t `nu` parameter is near
the lower bound or so large that the Student-t model is effectively Gaussian.

```{r tail-check}
check_drm(fit_tail_student)
coef(fit_tail_student, "nu")
```

## Structural Zeros

For count responses, the extra parameter in a zero-inflated model has a
specific interpretation: it is the probability of a separate structural-zero
process. This example generates NB2 counts with structural zeros and compares
ordinary NB2 against ZINB2.

```{r count-example}
set.seed(2402)
n <- 260
x <- rnorm(n)
mu <- exp(log(2.3) + 0.5 * x)
sigma <- 0.65
zi <- plogis(-0.8)

count <- rnbinom(n, size = 1 / sigma^2, mu = mu)
structural_zero <- runif(n) < zi
count[structural_zero] <- 0L
count_dat <- data.frame(count = count, x = x)

fit_nb2 <- drmTMB(
  bf(count ~ x, sigma ~ 1),
  family = nbinom2(),
  data = count_dat
)

fit_zinb2 <- drmTMB(
  bf(count ~ x, sigma ~ 1, zi ~ 1),
  family = nbinom2(),
  data = count_dat
)

criterion_table(NB2 = fit_nb2, ZINB2 = fit_zinb2)
```

If the criterion difference is small, keep both explanations in view. NB2 can
absorb some extra zeros by increasing overdispersion, while ZINB2 separates
structural zeros from count variation. Use `zi` only when a structural-zero
process is plausible for the response and sampling design.

## Scale Formula

Model selection is not only family selection. It can also ask whether a
distributional parameter needs a predictor. In this Gaussian example the mean
changes with `x`, and the residual scale also changes with `x`.

```{r scale-example}
set.seed(2403)
n <- 220
x <- rnorm(n)
sigma <- exp(-0.45 + 0.55 * x)
scale_dat <- data.frame(
  x = x,
  y = 0.3 + 0.55 * x + rnorm(n, sd = sigma)
)

fit_sigma_constant <- drmTMB(
  bf(y ~ x, sigma ~ 1),
  family = gaussian(),
  data = scale_dat
)

fit_sigma_x <- drmTMB(
  bf(y ~ x, sigma ~ x),
  family = gaussian(),
  data = scale_dat
)

criterion_table(`sigma ~ 1` = fit_sigma_constant, `sigma ~ x` = fit_sigma_x)
```

The `sigma ~ x` model estimates a log-scale slope. A positive slope means the
modelled residual scale increases as `x` increases. After selecting this
candidate, interpret the `sigma` coefficient on the ratio scale:

```{r scale-ratio}
exp(coef(fit_sigma_x, "sigma")["x"])
```

## A 200-Replicate Article-Support Simulation

The package includes a seeded article-support simulation. It uses 200
replicates per scenario, which is enough to make the table more stable than a
plumbing smoke run while still being small enough to ship as documentation
evidence. It is not a formal operating-characteristic study over sample sizes,
effect sizes, and candidate-set misspecification.

```{r article-summary}
summary_path <- system.file(
  "sim/reports/model-selection-article-summary.csv",
  package = "drmTMB"
)
if (!nzchar(summary_path)) {
  candidates <- c(
    "../inst/sim/reports/model-selection-article-summary.csv",
    "inst/sim/reports/model-selection-article-summary.csv"
  )
  summary_path <- candidates[file.exists(candidates)][1L]
}

model_selection_article <- read.csv(summary_path)
display_article <- model_selection_article[, c(
  "scenario",
  "selection_target",
  "n_replicate",
  "aic_truth_selection_rate",
  "aic_truth_selection_mcse",
  "bic_truth_selection_rate",
  "bic_truth_selection_mcse",
  "candidate_convergence_rate",
  "candidate_pdHess_rate",
  "candidate_warning_rate"
)]
names(display_article) <- c(
  "scenario",
  "target",
  "replicates",
  "AIC selected target",
  "AIC MCSE",
  "BIC selected target",
  "BIC MCSE",
  "candidate convergence",
  "candidate pdHess",
  "candidate warning"
)
knitr::kable(display_article, digits = 3)
```

The `normal_tail` row is deliberately useful: the Gaussian model is selected,
but the unnecessary Student-t candidate often carries warning or weak-Hessian
status because the extra tail parameter is unnecessary for normal data. The
`heavy_tail` and `extra_zeros` rows show the criterion tradeoff: AIC more often
keeps the extra Student-t or zero-inflation parameter, while BIC often prefers
the simpler candidate under this sample size because its penalty is stronger.
That is the workflow lesson. Selection criteria answer "which candidate has
the smaller penalized likelihood score?", while diagnostics answer "is this
candidate fit stable enough to use?".

## Practical Checklist

Use this sequence when comparing fitted `drmTMB` models:

1. Define the candidate set before looking at AIC/BIC.
2. Fit all candidates to the same response, row set, offsets, and weights.
   For Gaussian mixed models with different fixed-effect formulas, keep
   `REML = FALSE`.
3. Run `check_drm()` and keep convergence, Hessian status, warnings, and
   boundary diagnostics beside the criterion table.
4. Exclude errored fits from selection and treat nonconverged or weak-Hessian
   fits as diagnostic findings, not as ordinary winners.
5. Report AIC and BIC differences, not only the winning model.
6. Interpret the selected model's `mu`, `sigma`, `nu`, `zi`, or `rho12`
   parameters in scientific units or ratios.

When AIC and BIC disagree, describe the tradeoff. AIC is more willing to keep
extra parameters when they improve fit; BIC asks for stronger evidence as
sample size grows. In applied `drmTMB` work, that disagreement is often a sign
to inspect predictions, residuals, and the scientific meaning of the extra
distributional parameter rather than to declare one criterion universally
correct.
