---
title: "Checking and using fitted models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking and using fitted models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)

workflow_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_line(
        colour = "grey90",
        linewidth = 0.3
      ),
      panel.grid.major.y = ggplot2::element_line(
        colour = "grey87",
        linewidth = 0.35
      ),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      strip.background = ggplot2::element_rect(fill = "grey92", colour = NA),
      strip.text = ggplot2::element_text(face = "bold", colour = "grey15"),
      legend.position = "bottom",
      legend.title = ggplot2::element_text(colour = "grey20")
    )
}

workflow_habitat_scales <- function() {
  list(
    ggplot2::scale_colour_manual(
      values = c("reef" = "#0072B2", "kelp" = "#D55E00")
    ),
    ggplot2::scale_fill_manual(
      values = c("reef" = "#0072B2", "kelp" = "#D55E00")
    )
  )
}

workflow_distributional_scales <- function() {
  list(
    ggplot2::scale_colour_manual(
      values = c(
        "kelp" = "#D55E00",
        "reef" = "#0072B2",
        "sigma (habitat not in formula)" = "grey35"
      )
    ),
    ggplot2::scale_fill_manual(
      values = c(
        "kelp" = "#D55E00",
        "reef" = "#0072B2",
        "sigma (habitat not in formula)" = "grey35"
      )
    )
  )
}
```

After fitting a distributional model, the next question is not only "what are
the coefficients?" A useful workflow checks the fit, extracts the fitted
distributional parameters, predicts to meaningful covariate values, examines
residuals, and simulates from the fitted model.

This article uses a small growth example, but the same workflow applies to
other Gaussian location-scale models in `drmTMB`. In the `Start Here` path,
read it after the getting-started fit and the model-status map: this page is
where estimates become checked tables, predictions, intervals, residuals, and
simulation checks.

## Fit the model

Suppose growth is measured along a temperature gradient in two habitat types.
The model asks whether temperature and habitat change expected growth, and
whether temperature also changes residual variability.

Symbolically:

\[
\begin{aligned}
\text{growth}_i \mid \mu_i, \sigma_i
  &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i
  &= \beta_0 + \beta_1 \text{temperature}_i
     + \beta_2 \mathbb{1}(\text{habitat}_i = \text{kelp}),\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{temperature}_i .
\end{aligned}
\]

Matching R syntax:

```{r}
library(drmTMB)

set.seed(12)
n <- 120
fish <- data.frame(
  temperature = runif(n, -1.5, 1.5),
  habitat = factor(sample(c("reef", "kelp"), n, replace = TRUE))
)
eta_mu <- 1 + 0.7 * fish$temperature + 0.35 * (fish$habitat == "kelp")
eta_sigma <- -0.5 + 0.25 * fish$temperature
fish$growth <- rnorm(n, mean = eta_mu, sd = exp(eta_sigma))

fit <- drmTMB(
  bf(growth ~ temperature + habitat, sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Here `growth ~ temperature + habitat` defines the location model, and
`sigma ~ temperature` defines the residual standard deviation model on the
log scale.

## Check before interpreting

Run `check_drm()` before interpreting estimates:

```{r}
check_drm(fit)
```

A clean diagnostic table does not prove that the model is scientifically
correct, but it catches common problems early: optimizer convergence, gradients,
optimizer evaluation counts, finite objective values, Hessian status, dropped
rows, finite fixed-effect standard errors, positive scale values, residual
`rho12` values near the boundary, Student-t `nu` values near the boundary or
close to Gaussian behaviour, known sampling covariance summaries, random-effect
standard deviations near zero, and random-effect design issues when random
effects are present.

## Read the ordinary summary first

The ordinary R habit is still the right first habit in `drmTMB`:

```{r}
summary(fit)
```

`summary()` puts the fixed-effect estimates, response-scale scale, shape,
random-effect SD, correlation summaries, fitted covariance summaries, derived
variance ratios, log likelihood, and convergence code in one place when those
quantities are present. The fixed-effect table includes standard errors when
`TMB::sdreport()` was computed successfully.

Read the printed summary as a map, then use a more specific extractor only
when you need more detail:

| Summary component | What it reports | Use it for |
| --- | --- | --- |
| `coefficients` | Fixed-effect coefficients for each distributional parameter, usually on the model's link scale. | Direction, sign, and approximate uncertainty of terms in `mu`, `sigma`, `rho12`, shape, or zero-inflation formulas. |
| `parameters` | Direct response-scale parameters, fitted parameter ranges, random-effect SDs, and fitted correlations. | Constant `sigma`, residual `rho12`, `sd:mu:(1 | group)`, and quick response-scale uncertainty when `std_error` is available. |
| `covariance` | Fitted random-effect covariance and correlation rows for supported covariance blocks. | Latent correlation layers such as group-level, phylogenetic, or structured-effect covariance summaries. |
| `derived` | Simple variance-ratio summaries whose ingredients are unambiguous. | Repeatability or phylogenetic signal point estimates, including the random-effect and residual variances used in the ratio. |
| `confint` | Confidence-interval rows only when `conf.int = TRUE`. | Checking which requested fixed-effect or profile-likelihood intervals were actually computed. |

The printed summary is meant to answer "what did this fitted model estimate?"
For workflows that need the full object behind a row, use `fixef()`, `sigma()`,
`rho12()`, `ranef()`, `corpairs()`, or `profile_targets()` after reading the
summary.

For models with random effects, `summary()` is also the first place to look for
variance components. This compact example adds a site random intercept to the
location model:

```{r}
set.seed(13)
n_site <- 18
n_per_site <- 6
site <- factor(rep(seq_len(n_site), each = n_per_site))
site_effect <- rnorm(n_site, sd = 0.45)

fish_site <- data.frame(
  site = site,
  temperature = runif(n_site * n_per_site, -1.5, 1.5),
  habitat = factor(
    sample(c("reef", "kelp"), n_site * n_per_site, replace = TRUE)
  )
)

site_mu <- site_effect[as.integer(fish_site$site)]
fish_site$growth <- rnorm(
  nrow(fish_site),
  mean = 1 + 0.7 * fish_site$temperature +
    0.35 * (fish_site$habitat == "kelp") + site_mu,
  sd = 0.35
)

fit_site <- drmTMB(
  bf(growth ~ temperature + habitat + (1 | site), sigma ~ 1),
  family = gaussian(),
  data = fish_site
)

summary(fit_site)
```

In that output, the `sd:mu:(1 | site)` row is the fitted random-intercept
standard deviation on the response scale. Squaring it gives the corresponding
random-effect variance component. When the ingredients are unambiguous, the
`derived` component also reports repeatability and shows both the
random-effect variance and the residual variance used in that ratio.

When `TMB::sdreport()` succeeds, direct response-scale parameter rows such as
constant `sigma` and `sd:mu:(1 | site)` also include delta-method standard
errors. `confint()` uses those same optimized-parameter covariances for its
fast default Wald intervals. SD intervals are formed on the fitted log-SD scale
and exponentiated; correlation intervals are formed on the fitted guarded
atanh scale and transformed back with the model's correlation transform.
Treat these as quick local approximations, then reserve profile or bootstrap
work for the SD or correlation rows that need likelihood-shape or refit-based
uncertainty. `drmTMB` is a frequentist package, so these are confidence
intervals, not Bayesian credible intervals:

```{r}
profile_targets(fit_site)
```

The fastest fitted-model interval check is the default Wald route:

```{r}
confint(fit_site)
```

For a long mixed or phylogenetic model, narrow the table before doing anything
expensive:

```{r}
confint(fit_site, parm = "variance_components")
```

The variance-component shortcut returns constant scale targets and direct
random-effect SD targets. The companion shortcuts `"fixed_effects"`,
`"random_effects"`, and `"correlations"` keep large interval tables readable.
When a report needs a likelihood profile for a specific SD or correlation row,
copy the exact target name from `profile_targets()` and profile only that
target. `profile_precision = "fast"` supplies coarser `TMB::tmbprofile()`
controls (`ystep = 0.5`, `ytol = 2`) unless you override them in `...`; it is
useful as a first pass for long phylogenetic or spatial SD profiles:

```{r, eval = FALSE}
summary(
  fit_site,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "sd:mu:(1 | site)",
  profile_precision = "fast"
)
```

The same target can be requested directly:

```{r, eval = FALSE}
confint(
  fit_site,
  parm = "sd:mu:(1 | site)",
  method = "profile",
  profile_precision = "fast"
)
```

When refit-based uncertainty is the point of the analysis, use the bounded
bootstrap route in `confint()` for selected direct targets. Start with a small
pilot to estimate runtime, then increase `R` for the final analysis:

```{r, eval = FALSE}
confint(
  fit_site,
  parm = "variance_components",
  method = "bootstrap",
  R = 99,
  seed = 20260521
)
```

Bootstrap rows record successful and failed refits in `bootstrap.n` and
`bootstrap.failed`. On Unix-like systems, `parallel = "multicore"` and
`workers = ...` can run refits in parallel; `drmTMB` caps bootstrap workers at
10 to avoid accidental oversubscription.

Each distributional parameter has its own coefficient vector:

```{r}
coef(fit, "mu")
coef(fit, "sigma")
```

The `sigma` coefficients are on the log-standard-deviation scale. A positive
temperature coefficient in `sigma` means that residual variability increases
with temperature. Exponentiating the coefficient gives a multiplicative change
in residual standard deviation for a one-unit temperature increase.

Use the scale of the fitted parameter when deciding what to report:

| Output | Fitted scale | Report when the question is... |
| --- | --- | --- |
| `coef(fit, "mu")` | response units | expected growth, treatment effects, or mean trait differences |
| `coef(fit, "sigma")` | log residual SD | multiplicative residual-SD or residual-variance contrasts |
| `sigma(fit)` | residual SD | fitted predictability or extra heterogeneity on the response scale |
| `rho12(fit)` | residual correlation | residual coupling between two responses after means and residual SDs are modelled |
| `corpairs(fit)` | fitted correlation rows | group-level, phylogenetic, or residual correlation layers, depending on `level` |
| `profile_targets(fit)` | interval target inventory | whether a fitted SD or correlation target can receive a profile interval |

For confidence intervals, keep the fast default summary separate from the
expensive profile or bootstrap step. Wald 95% confidence intervals are
available for fixed effects and direct response-scale parameter rows:

```{r}
summary(fit, conf.int = TRUE)
```

Fixed effects include distributional-parameter coefficients such as `sigma`,
Student-t `nu`, zero-inflation, hurdle, and residual `rho12` formula terms.
Those Wald intervals are on the fitted link scale. For example, a Student-t
`fixef:nu:x` row is an interval for the `log(nu - 2)` coefficient, not a
response-scale degrees-of-freedom interval at a particular covariate value.

Profile-likelihood 95% confidence intervals are available for selected direct
targets such as constant `sigma`, random-effect SDs, random-effect
correlations, and constant residual `rho12`. Fixed effects keep Wald
confidence intervals in the same profile summary unless fixed-effect profile
targets are selected. For example, `fit_site` has `sigma ~ 1`, so its residual
standard deviation is a direct fitted-object target:

```{r, eval = FALSE}
summary(fit_site, conf.int = TRUE, method = "profile", ci_parm = "sigma")
```

For the first model in this article, `sigma` depends on `temperature`. There is
no single response-scale `sigma` target until you name the covariate row you
want to profile:

```{r, eval = FALSE}
sigma_grid <- data.frame(
  temperature = c(-1, 1),
  habitat = factor("reef", levels = levels(fish$habitat))
)
row.names(sigma_grid) <- c("cool_reef", "warm_reef")

confint(fit, parm = "sigma", method = "profile", newdata = sigma_grid)
```

The same row-specific rule applies to bivariate Gaussian scale and residual
correlation models. In a two-response fit with `sigma1 ~ z1`, `sigma2 ~ z2`,
and `rho12 ~ w`, profile the biologically meaningful rows directly:

```{r, eval = FALSE}
biv_grid <- data.frame(
  x = 0,
  z1 = c(-1, 1),
  z2 = c(1, -1),
  w = c(-0.5, 0.5)
)
row.names(biv_grid) <- c("cool_low_coupling", "warm_high_coupling")

confint(fit_biv, parm = "sigma1", method = "profile", newdata = biv_grid)
confint(fit_biv, parm = "sigma2", method = "profile", newdata = biv_grid)
confint(fit_biv, parm = "rho12", method = "profile", newdata = biv_grid)
```

For random-effect SDs and correlations, copy the exact target name from
`profile_targets()`. Random-slope terms keep the coefficient suffix in the
target name, and residual `rho12` remains separate from group-level or
phylogenetic correlation rows:

```{r, eval = FALSE}
profile_targets(fit_slope)

summary(
  fit_slope,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "sd:mu:(1 + temperature | p | site):temperature"
)
summary(
  fit_slope,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "cor:mu:cor((Intercept),temperature | p | site)"
)
summary(
  fit_biv_group,
  conf.int = TRUE,
  method = "profile",
  ci_parm = "cor:mu:cor(mu1:(Intercept),mu2:(Intercept) | p | id)"
)
```

Use the profile call when a report needs fixed-effect confidence intervals and
profile-likelihood intervals for variance components, SDs, or correlations in
one summary. Successful profile rows also include `profile.boundary` and
`profile.message`; values other than `"ok"` tell you to check boundary or
near-correlation-limit behaviour before interpreting the interval.

For correlations, the fast Wald route uses the model's fitted correlation-link
scale, which is a guarded Fisher-z/atanh transform for current correlation
parameters. This differs from a sample-correlation Fisher-z interval with an
effective sample size: `drmTMB` is using the TMB covariance of the fitted
correlation parameter, not a separate `n_eff` heuristic.

Use `profile_targets(fit)` before asking for a profile interval. A row is
profile-ready only when it maps to a direct TMB target and the fitted object
kept its TMB object. Interval-aware output uses `conf.status` to explain rows
without intervals: `newdata_required` means profile a supplied row with
`confint(..., newdata = ...)`, while `derived_interval_unavailable` means the
point estimate combines several quantities and needs a later derived-profile
method.

Read `conf.status` as an action column:

| `conf.status` | What it means | What to do next |
| --- | --- | --- |
| `wald` | A Wald interval was returned for a fixed-effect or direct parameter row. | Use it for routine summaries and as the fastest screen for large models, while remembering it is asymptotic. |
| `profile` | A profile-likelihood interval was returned. | Inspect `profile.boundary` and `profile.message` before treating the interval as stable. |
| `bootstrap` | A simulate/refit percentile interval was returned by `confint()`. | Check `bootstrap.n` and `bootstrap.failed`; increase `R` only after a small runtime pilot works. |
| `profile_ready` | The row is a direct profile target, but the current summary call did not request it. | Use the row's `parm` value in `confint(..., method = "profile")` or in `summary(..., method = "profile", ci_parm = ...)`. |
| `newdata_required` | The fitted surface needs a concrete row before it can be profiled. | Build a biologically meaningful `newdata` row and call `confint(..., newdata = ...)`. |
| `derived_interval_unavailable` | The point estimate combines multiple fitted quantities and has no validated interval method yet. | Report the point estimate as exploratory, or simplify to a direct target if interval support is essential. |
| `wald_unavailable` | Wald uncertainty is not available for this row, or `sdreport()` was skipped or failed. | Use a profile-ready target when one exists; otherwise report the missing interval explicitly. |
| `bootstrap_unavailable` | Fewer than two bootstrap refits produced a finite estimate for the requested target. | Treat the interval as unavailable; inspect model convergence, refit controls, and the bootstrap failure count before increasing `R`. |
| `target_unavailable` or `profile_unavailable` | The row is descriptive or has no current direct interval target. | Treat the interval as unsupported and check `profile_targets(fit)` before trying another route. |
| `not_requested` | The table is reporting a point estimate without an interval request. | Refit is not needed; request intervals only for the rows you plan to interpret. |

Bootstrap intervals are public only through `confint()` for selected direct
targets. `summary(fit, conf.int = TRUE, method = "bootstrap")` and
`corpairs(fit, conf.int = TRUE, method = "bootstrap")` still stop before
interval work begins, and derived q4 covariance rows, repeatability, and
phylogenetic signal still need separate interval designs. For long applied
models, the practical default is therefore: Wald for routine reporting,
targeted `profile_precision = "fast"` profile checks for important SD or
correlation rows, and bootstrap only when refit-based uncertainty is worth the
runtime.

## Predict distributional parameters

`predict()` returns one distributional parameter at a time. For fitted rows,
`fitted(fit)` and `predict(fit, dpar = "mu")` return the same location values
for this fixed-effect model:

```{r}
head(fitted(fit))
head(predict(fit, dpar = "mu"))
head(sigma(fit))
```

For new covariate values, supply `newdata`. These are fixed-effect,
population-level predictions. You can write the grid directly:

```{r}
new_fish <- data.frame(
  temperature = c(-1, 0, 1),
  habitat = factor("reef", levels = levels(fish$habitat))
)

predict(fit, newdata = new_fish, dpar = "mu")
predict(fit, newdata = new_fish, dpar = "sigma")
predict(fit, newdata = new_fish, dpar = "sigma", type = "link")
```

Or build an explicit prediction grid. This keeps the focal term, conditioned
nuisance predictors, and grid rule attached to the data:

```{r}
temperature_grid <- prediction_grid(
  fit,
  focal = "temperature",
  at = list(temperature = c(-1, 0, 1)),
  condition = list(habitat = "reef")
)

attr(temperature_grid, "prediction_grid")
```

Use `predict_parameters()` when the interpretation task needs several
distributional parameters on the same grid:

```{r}
temperature_surface <- predict_parameters(
  fit,
  newdata = temperature_grid,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)

temperature_surface
unique(temperature_surface[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])
```

Here both rows receive Wald 95% confidence intervals because the table uses an
explicit `newdata` grid and both `mu` and `sigma` have ordinary fixed-effect
bases in this fitted model. The interval is a confidence band for the fitted
distributional-parameter surface, not a prediction interval for future raw
growth observations.

If you request intervals for fitted rows without `newdata`, the table keeps the
point estimates and says what is missing:

```{r}
fitted_interval_status <- predict_parameters(
  fit,
  dpar = "mu",
  conf.int = TRUE
)

unique(fitted_interval_status[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])
```

Read `newdata_required` as an instruction to choose a row on purpose. For a
surface, build a compact grid such as `temperature_grid` instead of asking the
plot to decide which fitted rows deserve intervals.

Use an empirical grid when the question is about the fitted response at focal
temperatures while averaging over the fitted-row covariate distribution. Here
each requested temperature is crossed with the fitted model rows, so the
habitat distribution is the one in the fitted data:

```{r}
temperature_empirical <- prediction_grid(
  fit,
  focal = "temperature",
  at = list(temperature = c(-1, 0, 1)),
  margin = "empirical"
)

attr(temperature_empirical, "prediction_grid")
```

Then reduce the empirical grid by the focal term:

```{r}
marginal_parameters(
  fit,
  newdata = temperature_empirical,
  dpar = c("mu", "sigma"),
  by = "temperature"
)
```

These helpers are tables, not plotting functions. They are the current Phase 17
surface for future visualization work: each row names the distributional
parameter, fitted component, prediction scale, estimate, interval status, and
grid values. By default, prediction tables are point-estimate summaries with
`conf.status = "not_requested"` and
`interval_source = "not_available"`. When you request
`conf.int = TRUE` on an explicit `newdata` grid, ordinary fixed-effect
distributional parameters receive Wald fixed-effect intervals with
`conf.status = "wald"` and `interval_source = "wald"`. Rows that cannot receive
that interval, such as modelled random-effect SD surfaces evaluated on a grid,
keep an explicit unavailable status instead of silently drawing a band.

## Predict a random-effect SD surface

Random-effect scale formulas answer a different question from residual
`sigma`. In the next example, `sd(site) ~ reef_cover` asks whether the
among-site spread in expected growth changes along a site-level reef-cover
gradient. The predictor is constant within each site:

```{r}
set.seed(14)
n_site <- 14
n_per_site <- 5
site <- factor(rep(seq_len(n_site), each = n_per_site))
reef_cover_site <- seq(-1, 1, length.out = n_site)
reef_cover <- reef_cover_site[as.integer(site)]
temperature <- runif(n_site * n_per_site, -1.5, 1.5)
site_sd <- exp(-0.65 + 0.45 * reef_cover_site)
site_effect <- rnorm(n_site, sd = site_sd)

fish_site_scale <- data.frame(
  site = site,
  temperature = temperature,
  reef_cover = reef_cover
)
fish_site_scale$growth <- rnorm(
  nrow(fish_site_scale),
  mean = 1 + 0.6 * fish_site_scale$temperature +
    site_effect[as.integer(fish_site_scale$site)],
  sd = 0.35
)

fit_site_scale <- drmTMB(
  bf(
    growth ~ temperature + (1 | site),
    sigma ~ 1,
    sd(site) ~ reef_cover
  ),
  family = gaussian(),
  data = fish_site_scale
)
```

The fitted equation for the site random-effect SD is
`log(sd_site) = alpha_0 + alpha_1 reef_cover`. Use a grid when you want the
site-level SD surface on the response scale:

```{r}
sd_site_grid <- prediction_grid(
  fit_site_scale,
  focal = "reef_cover",
  at = list(reef_cover = c(-1, 0, 1)),
  condition = list(temperature = 0)
)

sd_site_surface <- predict_parameters(
  fit_site_scale,
  newdata = sd_site_grid,
  dpar = "sd(site)"
)

sd_site_surface
```

The table reports the `random-effect-sd-model` component. Its estimates are
random-intercept SDs, not residual SDs and not raw responses. If you request a
Wald interval for this modelled random-effect SD surface, the table keeps the
point estimate and marks the interval status explicitly:

```{r}
sd_site_interval_status <- predict_parameters(
  fit_site_scale,
  newdata = sd_site_grid,
  dpar = "sd(site)",
  conf.int = TRUE
)

unique(sd_site_interval_status[, c(
  "dpar",
  "component",
  "conf.status",
  "interval_source",
  "conf.level"
)])
```

Read that status as a boundary, not as a plotting failure. Direct
random-effect SD surfaces need a profile-likelihood or bootstrap route before a
confidence band is available, so the current plotting helper should show the
line without a ribbon. If the report needs one row per reef-cover value, reduce
the same explicit grid:

```{r}
marginal_parameters(
  fit_site_scale,
  newdata = sd_site_grid,
  dpar = "sd(site)",
  by = "reef_cover"
)
```

## Estimate marginal means for `mu`

When the question is an adjusted mean of the location parameter `mu`, the first
`emmeans` bridge can build an estimated marginal mean table directly from a
fixed-effect univariate fit. Here the reference grid holds `temperature` at
zero and estimates `mu` for each habitat:

```{r}
if (requireNamespace("emmeans", quietly = TRUE)) {
  emmeans::emmeans(
    fit,
    ~habitat,
    at = list(temperature = 0)
  )
}
```

Read this output as estimated marginal means of the native distributional
parameter `mu`, not as summaries of `sigma` or as a random-effect, bivariate,
zero-inflated, hurdle, ordinal, or slope workflow. Generic `emmeans` contrasts
can be formed from this returned `mu` grid, but broader drmTMB-specific contrast
helpers remain a separate future contract. For other distributional targets, use
explicit `prediction_grid()` and `predict_parameters()` tables until the
corresponding reference-grid algebra and tests exist.

The first optional plotting helper consumes that table directly. For a figure,
use a denser grid than the small printed table and request intervals before
plotting:

```{r temperature-surface-plot, fig.cap = "Fitted response-scale `mu` and `sigma` surfaces from `predict_parameters()`; `mu` varies by habitat, while `sigma` is shared because habitat is not in the scale formula.", fig.alt = "Faceted fitted mu and sigma surfaces over temperature. The mu facet has separate kelp and reef lines, while the sigma facet uses one grey line because habitat is not in the sigma formula."}
temperature_plot_grid <- prediction_grid(
  fit,
  focal = c("temperature", "habitat"),
  at = list(temperature = seq(-1.5, 1.5, length.out = 40))
)

pred_temperature <- predict_parameters(
  fit,
  newdata = temperature_plot_grid,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)

unique(pred_temperature[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])

if (requireNamespace("ggplot2", quietly = TRUE)) {
  pred_temperature_display <- pred_temperature[
    pred_temperature$dpar == "mu" | pred_temperature$habitat == "reef",
  ]
  pred_temperature_display$display <- ifelse(
    pred_temperature_display$dpar == "sigma",
    "sigma (habitat not in formula)",
    as.character(pred_temperature_display$habitat)
  )
  pred_temperature_display$display <- factor(
    pred_temperature_display$display,
    levels = c("kelp", "reef", "sigma (habitat not in formula)")
  )

  plot_parameter_surface(
    pred_temperature_display,
    x = "temperature",
    colour = "display"
  ) +
    workflow_distributional_scales() +
    workflow_theme() +
    ggplot2::labs(colour = NULL)
}
```

For a discrete focal predictor, the same interval columns draw interval bars
rather than a ribbon. This example asks for 90% Wald intervals, so the table
records `conf.level = 0.9` on rows where `conf.status = "wald"`:

```{r habitat-contrast-plot, fig.width = 5.8, fig.height = 2.7, fig.cap = "Fitted habitat-level `mu` estimates at temperature zero; vertical intervals are 90% Wald confidence intervals requested from `predict_parameters()`.", fig.alt = "Two habitat-level fitted mu estimates with vertical Wald confidence intervals at temperature zero."}
habitat_plot_grid <- prediction_grid(
  fit,
  focal = "habitat",
  condition = list(temperature = 0)
)

pred_habitat <- predict_parameters(
  fit,
  newdata = habitat_plot_grid,
  dpar = "mu",
  conf.int = TRUE,
  conf.level = 0.90
)

unique(pred_habitat[, c(
  "dpar",
  "conf.status",
  "interval_source",
  "conf.level"
)])

if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_parameter_surface(pred_habitat, x = "habitat", line = FALSE) +
    workflow_theme()
}
```

Treat `conf.level` as the requested level for rows where an interval was
computed or attempted. It is not a success flag by itself; read it together
with `conf.status` and `interval_source`.

The figure is a convenience layer, not a new estimand. Keep reference grids,
marginalization choices, and interval sources visible instead of hiding those
decisions inside a plot.

For reports, pair model surfaces with the raw response pattern. The raw-data
panel uses the observed response scale:

```{r raw-growth-plot, fig.cap = "Raw response-scale growth observations by temperature and habitat, shown before the fitted distributional-parameter surfaces.", fig.alt = "Observed growth values over temperature, coloured by habitat, showing the raw response pattern before model surfaces are drawn."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    fish,
    ggplot2::aes(x = temperature, y = growth, colour = habitat)
  ) +
    ggplot2::geom_point(alpha = 0.68, size = 1.8) +
    workflow_habitat_scales() +
    ggplot2::labs(
      x = "Temperature",
      y = "Observed growth",
      colour = "Habitat"
    ) +
    workflow_theme()
}
```

The fitted panels then use the prediction table. Keep `mu` and `sigma` on their
own y-axes because they are different biological quantities. When a plot keeps
one distributional parameter, `plot_parameter_surface()` labels the y-axis with
the parameter and prediction scale. The bands in these panels come from the
Wald fixed-effect intervals in `pred_temperature`; if those columns were absent
or marked `interval_source = "not_available"`, the helper would draw only
lines and points:

```{r mu-temperature-plot, fig.cap = "Fitted response-scale `mu` surface by habitat; ribbons are Wald confidence bands from the prediction table.", fig.alt = "Fitted mu response-scale surface over temperature with separate kelp and reef Wald confidence bands."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_parameter_surface(
    pred_temperature,
    x = "temperature",
    colour = "habitat",
    dpar = "mu",
    facet = NULL
  ) +
    workflow_habitat_scales() +
    workflow_theme()
}
```

Because `sigma ~ temperature` has no habitat term, the `sigma` plot keeps one
row per temperature. That avoids drawing two identical habitat lines on top of
each other.

```{r sigma-temperature-plot, fig.cap = "Fitted response-scale `sigma` surface over temperature; one band is drawn because the scale formula has no habitat term.", fig.alt = "Fitted sigma response-scale surface over temperature with one Wald confidence band because habitat is not in the sigma formula."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  pred_sigma_temperature <- pred_temperature[
    pred_temperature$dpar == "sigma" & pred_temperature$habitat == "reef",
  ]

  plot_parameter_surface(
    pred_sigma_temperature,
    x = "temperature",
    dpar = "sigma",
    facet = NULL
  ) +
    workflow_theme()
}
```

Do not put raw `growth` points on the `sigma` panel. Raw observations show the
response; fitted `sigma` shows the residual standard deviation implied by the
model. The same rule applies to uncertainty: draw a band only when the table
names a real interval source such as `"wald"` or `"profile"`.

For this Gaussian model, the response-scale `sigma` predictions are residual
standard deviations. The link-scale predictions are the corresponding
log-standard deviations. Report residual variance as `sigma(fit)^2` or
`predict(fit, newdata = new_fish, dpar = "sigma")^2`; do not square the
`sigma` coefficient, because coefficients are fitted on the log-SD scale.

In meta-analytic Gaussian models with `meta_known_V(V = V)`, `sigma(fit)`
still returns the modelled residual heterogeneity scale. Simulation and
Pearson residuals combine the known sampling covariance with that residual
scale internally.

## Inspect residuals

Response residuals are observed minus fitted location:

```{r}
head(residuals(fit))
```

Pearson residuals divide by the fitted observation scale:

```{r}
head(residuals(fit, type = "pearson"))
```

For bivariate Gaussian fits, `residuals(fit, type = "pearson")` returns a
two-column matrix whitened by the fitted `sigma1`, `sigma2`, and residual
correlation `rho12`.

## Simulate from the fitted model

Simulation is a compact way to ask whether the fitted model generates data on
the same scale as the observed response:

```{r}
sims <- simulate(fit, nsim = 3, seed = 1)
head(sims)
```

For this univariate model, the output has one column per simulated data set.
For bivariate Gaussian models, simulations are returned in paired columns such
as `sim_1_y1` and `sim_1_y2`.

## A compact post-fit checklist

For routine fitted models, the post-fit loop is:

1. `check_drm(fit)` before interpreting estimates.
2. `summary(fit)` for fixed effects, response-scale parameters, variance
   components, derived summaries, interval status, and the extractor to use
   next.
3. `profile_targets(fit)` before asking for profile-likelihood intervals on
   fitted SD or correlation targets.
4. `predict(fit, dpar = ...)` for fitted or new-data distributional
   parameters.
5. `residuals(fit, type = "pearson")` to inspect standardized lack of fit.
6. `simulate(fit)` to compare fitted-model behaviour with the observed data.

The same loop will extend as more families arrive. The parameter names change,
but the discipline stays the same: check the fit, then interpret each
distributional parameter on its own scale.
