---
title: "Phase 18 count simulation gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phase 18 count simulation gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article is a developer-facing figure gallery for the first Phase 18 count
pilot. Its reader is a package contributor or method reviewer who wants to see
how a small simulation run becomes inspectable figures before larger grids are
launched.

The gallery is deliberately narrow. It compares the fitted non-zero-inflated
Poisson and NB2 `mu` random-effect smoke surfaces:

```text
count ~ x + (1 | id) + (0 + x | id)
```

For NB2, the pilot also keeps fixed-effect overdispersion, `sigma ~ z`, in the
model. The random-effect terms are ordinary grouped effects on the log mean.
They are not phylogenetic, spatial, animal, `relmat()`, zero-inflation,
hurdle, one-inflation, or shape random effects.

## What the gallery shows

The rendered gallery has five parts:

| Section | What to inspect | Why it matters |
|---|---|---|
| Bias | signed mean estimation error by family and estimand | catches systematic drift |
| RMSE | root-mean-square estimation error by family and estimand | shows typical error size |
| Interval coverage | Wald rows for fixed effects and profile rows for random-effect SDs | separates interval method from estimand |
| Run manifest | cell id, replicate, status, elapsed time, warnings | prevents silent success claims |
| Warning/error ledger | failed fits and warning-bearing fits | keeps numerical failures next to figures |

The current gallery is pilot evidence. It is useful for checking plumbing,
labelling, and obvious failures. It is not a final coverage, power, or runtime
claim.

## Render a local gallery

The helpers live under `inst/sim/` because Phase 18 simulation runs are optional
developer infrastructure, not package examples. A local run can render an HTML
gallery into the ignored `inst/sim/results/` folder:

```{r render-count-gallery, eval = FALSE}
library(drmTMB)

sim_files <- c(
  "sim/R/sim_registry.R",
  "sim/R/sim_utils.R",
  "sim/R/sim_runner.R",
  "sim/R/sim_aggregate.R",
  "sim/R/sim_uncertainty.R",
  "sim/R/sim_plot_data.R",
  "sim/R/sim_gallery.R",
  "sim/dgp/sim_dgp_poisson_mu_random_effect.R",
  "sim/dgp/sim_dgp_nbinom2_mu_random_effect.R",
  "sim/fit/sim_summarise_poisson_mu_random_effect.R",
  "sim/fit/sim_summarise_nbinom2_mu_random_effect.R",
  "sim/run/sim_run_poisson_mu_random_effect_smoke.R",
  "sim/run/sim_run_nbinom2_mu_random_effect_smoke.R",
  "sim/run/sim_summary_poisson_mu_random_effect_smoke.R",
  "sim/run/sim_summary_nbinom2_mu_random_effect_smoke.R",
  "sim/run/sim_summary_count_mu_random_effect_pilot.R",
  "sim/run/sim_render_count_mu_gallery_smoke.R"
)

for (path in sim_files) {
  source(system.file(path, package = "drmTMB", mustWork = TRUE))
}

out <- phase18_render_count_mu_re_gallery_smoke(
  output_dir = "inst/sim/results/count-gallery-smoke",
  overwrite = TRUE,
  notes = "local count-gallery smoke"
)

out$gallery$output_file
```

When working from the source checkout before installation, replace the
`system.file()` call with `source(file.path("inst", path))` after
`devtools::load_all(".")`.

## Read the figures

Start with the manifest and warning/error ledger, then interpret the plots. A
flat zero-bias line is only reassuring when the fitted cells converged and the
warnings are empty or understood.

For fixed effects, the gallery uses Wald intervals because the smoke summaries
already carry standard errors for formula coefficients. For random-effect SDs,
the gallery uses direct profile-likelihood intervals because the SD scale is
bounded below and the Wald approximation is fragile near zero.

The coverage panel shows an approximate Monte Carlo uncertainty range when the
coverage table contains `coverage_mcse`. With one replicate the range is only a
plumbing check. Larger optional grids should increase replicate counts before
turning this gallery into an operating-characteristic report.

## What comes next

The next useful gallery work is not to add more decoration. It is to feed the
same visual grammar with larger count grids and then add sibling panels for the
ready Gaussian location-scale, `meta_V(V = V)`, and spatial one-slope smoke
surfaces. Surfaces still listed as planned should remain in the failure ledger
until their likelihood, extractors, interval targets, and recovery tests exist.
