---
title: "Structured dependence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)

balanced_ultrametric_tree <- function(n_tip = 16L) {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0("sp_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}

simulate_phylo_q4_traits <- function(seed = 20260802) {
  set.seed(seed)
  tree <- balanced_ultrametric_tree(32)
  A_tip <- drmTMB:::drm_phylo_tip_covariance(tree)
  sd_phylo <- c(mu1 = 0.60, mu2 = 0.50, sigma1 = 0.18, sigma2 = 0.16)
  corr_phylo <- matrix(
    c(
      1.00, 0.50, 0.10, 0.05,
      0.50, 1.00, 0.05, 0.10,
      0.10, 0.05, 1.00, 0.25,
      0.05, 0.10, 0.25, 1.00
    ),
    nrow = 4,
    byrow = TRUE,
    dimnames = list(names(sd_phylo), names(sd_phylo))
  )
  z_phylo <- matrix(rnorm(length(tree$tip.label) * 4), ncol = 4)
  phylo_effect <- t(chol(A_tip)) %*%
    z_phylo %*%
    chol(diag(sd_phylo) %*% corr_phylo %*% diag(sd_phylo))
  dimnames(phylo_effect) <- list(tree$tip.label, names(sd_phylo))

  n_each <- 6
  species <- rep(tree$tip.label, each = n_each)
  n <- length(species)
  climate <- rnorm(n)
  habitat_variability <- rnorm(n)
  assay_context <- rep(rep(c(-0.5, 0.5), each = n_each / 2), length(tree$tip.label))

  eta_heat <- 0.35 + 0.30 * climate + phylo_effect[species, "mu1"]
  eta_desiccation <- -0.20 - 0.25 * climate + phylo_effect[species, "mu2"]
  log_sigma_heat <- -1.15 + 0.20 * habitat_variability +
    phylo_effect[species, "sigma1"]
  log_sigma_desiccation <- -1.05 - 0.15 * habitat_variability +
    phylo_effect[species, "sigma2"]
  rho12 <- 0.15
  e_heat <- rnorm(n)
  e_desiccation <- rho12 * e_heat + sqrt(1 - rho12^2) * rnorm(n)

  list(
    tree = tree,
    data = data.frame(
      species = species,
      climate = climate,
      habitat_variability = habitat_variability,
      assay_context = assay_context,
      heat_tolerance = eta_heat + exp(log_sigma_heat) * e_heat,
      desiccation_tolerance = eta_desiccation +
        exp(log_sigma_desiccation) * e_desiccation
    )
  )
}
```

Phylogenetic and spatial dependence are structured random effects. The first
implemented `drmTMB` path is an intercept-only phylogenetic effect in the
univariate Gaussian location formula. A first bivariate path also fits matching
intercept-only `phylo()` terms in `mu1` and `mu2`, estimating the two
phylogenetic location SDs and their mean-mean correlation. Matching labelled
`phylo()` terms across `mu1`, `mu2`, `sigma1`, and `sigma2` now fit the first
constant bivariate phylogenetic location-scale block. Spatial fields,
phylogenetic slopes, and predictor-dependent latent phylogenetic correlations
remain planned extensions.

The scientific question for the implemented path is simple:

> Do predictors explain species trait means after accounting for similarity
> among related species?

In the first fitted path, `species` indexes the rows in the data and `tree`
supplies the relatedness among those species. The model estimates a
phylogenetic deviation in the location parameter `mu` for each species.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

## Current implementation status

This page separates the fitted paths from the roadmap. The fitted phylogenetic
pieces are useful, but they are not yet the full phylogenetic
location-scale-coscale model.

| Model piece | Status | What it means |
| --- | --- | --- |
| `phylo(1 | species, tree = tree)` in univariate `mu` | Implemented | One trait has a phylogenetic random intercept in the location predictor. |
| Matching `phylo(1 | species, tree = tree)` terms in bivariate `mu1` and `mu2` | Implemented first slice | Two traits have correlated phylogenetic location deviations; `corpairs(level = "phylogenetic")` reports the mean-mean row separately from residual `rho12`. |
| Matching labelled `phylo(1 | p | species, tree = tree)` terms in bivariate `mu1` and `mu2` | Implemented | The label is preserved in SD names, the fitted mean-mean correlation, `corpairs(block = "p")`, and profile-target names. |
| Ordinary `(1 | p | id)` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first slice | One ordinary q=4 group block estimates six latent location-location, location-scale, and scale-scale correlations. |
| Matching labelled `phylo()` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first slice | One constant q=4 phylogenetic block estimates four endpoint SDs and six latent location-scale correlations. Partial, unlabelled, mismatched, and slope forms remain rejected. |
| `sd_phylo(species) ~ z` | Planned | This is the separate Box 1 direct-SD family, not a scale formula with a matching random effect. |
| `spatial(1 | site, coords = coords)` or `spatial(1 | site, mesh = mesh)` | Planned | Spatial will mirror the phylogenetic API after the phylogenetic structured-effect path is stable. |

## Mathematical contract

The general structured-effect bridge is:

\[
\eta_d =
  \mathbf{X}_d \boldsymbol{\beta}_d +
  \mathbf{Z}_d \mathbf{z},
\qquad
\mathbf{z} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_z^2 \mathbf{K}).
\]

Here \(d\) is the distributional parameter being modelled. In the first
implemented case, \(d = \mu\): the structured effect is added to the mean
predictor. For phylogenetic models, \(\mathbf{K}\) is a tree-derived
phylogenetic covariance matrix \(\mathbf{A}\). For spatial models,
\(\mathbf{K}\) will be a distance-derived matrix or an SPDE/GMRF precision.

For the implemented univariate Gaussian phylogenetic location model:

\[
y_i \mid \mu_i, \sigma_i
  \sim \operatorname{Normal}(\mu_i, \sigma_i^2),
\]

\[
\mu_i =
  \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu}
  + a_{species[i]},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma},
\]

\[
\mathbf{a} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2\mathbf{A}).
\]

The phylogenetic SD \(\sigma_{phylo}\) measures how strongly related species
remain similar in their fitted mean after the fixed effects have been included.
The residual `sigma` remains the within-observation residual SD. These are
different scale parameters.

The matching implemented R syntax is:

```r
drmTMB(
  drm_formula(
    y ~ x1 + phylo(1 | species, tree = tree),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

## Worked example: thermal tolerance across species

Suppose we measure thermal tolerance for multiple observations per species and
ask whether body size predicts thermal tolerance after accounting for shared
ancestry. In a real analysis, `tree` would usually come from a phylogenetic
data workflow. For this vignette, the setup code creates a small ultrametric
branch-length tree so the example stays self-contained.

```{r}
set.seed(31)
tree <- balanced_ultrametric_tree(16)

species_info <- data.frame(
  species = tree$tip.label,
  body_size = scale(
    rnorm(16, mean = rep(c(-0.4, 0.4), each = 8), sd = 0.35)
  )[, 1]
)

phylo_shift <- rep(c(-0.5, 0.5), each = 8) +
  rep(c(-0.18, 0.18), each = 4, length.out = 16)

n_each <- 5
dat <- species_info[rep(seq_len(nrow(species_info)), each = n_each), ]
dat$body_size <- dat$body_size + rnorm(nrow(dat), sd = 0.08)
dat$thermal_tolerance <- 1.2 + 0.45 * dat$body_size +
  phylo_shift[match(dat$species, species_info$species)] +
  rnorm(nrow(dat), sd = 0.25)

head(dat)
```

The fitted model follows the equation above:

```{r}
fit_phylo <- drmTMB(
  drm_formula(
    thermal_tolerance ~ body_size + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat
)

summary(fit_phylo)
```

How to read this output:

The `mu:body_size` row is the fixed-effect slope for body size in the mean
thermal tolerance model. The `sigma:(Intercept)` row is the log residual SD, so
`exp(coef(fit_phylo, "sigma"))` gives the within-observation residual SD. The
`phylo(1 | species)` random-effect SD is \(\sigma_{phylo}\), the phylogenetic
among-species SD in the mean after body size has been included.

```{r}
phylo_sd <- unname(fit_phylo$sdpars$mu["phylo(1 | species)"])
residual_sd <- exp(coef(fit_phylo, "sigma")["(Intercept)"])
c(phylogenetic_sd = phylo_sd, residual_sd = residual_sd)
```

In this run, related species remain similar in thermal tolerance after
accounting for body size. The fitted phylogenetic SD is larger than the
within-observation residual SD, so shared ancestry explains a visible part of
the remaining mean-level structure.

`check_drm()` reports whether the optimizer, Hessian, scale, dropped-row, and
phylogenetic replication checks passed:

```{r}
check_drm(fit_phylo)
```

## Practical checks for phylogenetic fits

The public `phylo()` term requires an ultrametric tree with branch lengths.
Users pass the tree; `drmTMB` builds the sparse phylogenetic precision
internally. You do not need to construct the phylogenetic covariance matrix or
A-inverse yourself.

If a phylogenetic fit fails before optimization, check these points first:

- the tree object has class `phylo`;
- every observed species in `data$species` appears in `tree$tip.label`;
- species names use the same spelling and underscore/space convention in the
  data and tree;
- the tree has finite positive branch lengths and is ultrametric;
- the model uses the implemented syntax
  `phylo(1 | species, tree = tree)` in a univariate Gaussian `mu` formula, or
  matching terms in bivariate Gaussian `mu1` and `mu2` formulas.

Phylogenetic slopes and structured effects in `rho12` are planned. Partial,
unlabelled, mismatched, and slope q=4 phylogenetic location-scale forms should
error clearly rather than silently fitting a different model.

## Worked example: two traits with phylogenetic location-scale covariance

Now suppose the scientific question is about two tolerance traits measured
repeatedly for the same species. We want climate to explain each trait mean,
habitat variability to explain each residual SD, and assay context to explain
the residual within-observation correlation. We also want the species-level
phylogenetic deviations in both means and both log-SDs to share one latent q=4
covariance block.

The simulated data below are deliberately small enough for the article to run.
The hidden setup creates the tree and species-level phylogenetic effects; the
fitted model uses only public `drmTMB()` syntax.

```{r}
sim_q4 <- simulate_phylo_q4_traits()
tree <- sim_q4$tree
tol_dat <- sim_q4$data

head(tol_dat)
```

The model uses the same labelled `phylo()` term in all four endpoints:

```{r}
fit_phylo_q4 <- suppressWarnings(
  drmTMB(
    drm_formula(
      mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
      mu2 = desiccation_tolerance ~ climate +
        phylo(1 | p | species, tree = tree),
      sigma1 = ~ habitat_variability +
        phylo(1 | p | species, tree = tree),
      sigma2 = ~ habitat_variability +
        phylo(1 | p | species, tree = tree),
      rho12 = ~ assay_context
    ),
    family = c(gaussian(), gaussian()),
    data = tol_dat,
    control = list(eval.max = 1000, iter.max = 1000)
  )
)

summary(fit_phylo_q4)
```

The fixed-effect rows answer the distributional-regression questions. The
`mu1:climate` and `mu2:climate` rows are mean effects. The
`sigma1:habitat_variability` and `sigma2:habitat_variability` rows are effects
on log residual SD. The `rho12:assay_context` row is the residual coscale effect:
it changes the residual correlation between the two traits within the same
observation.

The phylogenetic q=4 block is a different layer. Use `corpairs()` to read the
latent species-level correlations:

```{r}
corpairs(fit_phylo_q4, level = "phylogenetic")
```

Those six rows are not residual correlations. The `mean-mean` row asks whether
species that are phylogenetically high for heat tolerance also tend to be high
for desiccation tolerance. The four `mean-scale` rows ask whether evolutionary
deviations in means are associated with evolutionary deviations in variability.
The `scale-scale` row asks whether the two traits' phylogenetic log-SD
deviations move together.

For a compact scalar view of the residual layer, use `rho12()`:

```{r}
range(rho12(fit_phylo_q4))
```

`check_drm()` keeps these layers separate. In small examples, q=4 models can
warn about optimizer convergence even when gradients, Hessian status, and q=4
covariance diagnostics are acceptable. Treat this article fit as an output
map; for scientific inference, inspect the full diagnostic table and use larger
datasets or simpler covariance structure when needed.

```{r}
q4_check <- check_drm(fit_phylo_q4)
q4_check[q4_check$check %in% c(
  "optimizer_convergence",
  "fixed_gradient",
  "hessian_positive_definite",
  "biv_phylo_q4_covariance"
), ]
```

The fitted q=4 correlations are point summaries. They are listed in
`profile_targets()` as derived `theta_phylo` targets, not as direct
profile-ready intervals:

```{r}
subset(
  profile_targets(fit_phylo_q4),
  startsWith(parm, "cor:phylo:"),
  select = c(parm, target_type, profile_ready, profile_note)
)
```

Predictor-dependent latent phylogenetic correlations are a future layer. The
extractor is `corpairs()`, but the planned formula syntax should be singular:

```r
# Planned, not fitted yet:
corpair(species, block = "p", class = "location-location") ~ w
corpair(species, block = "p", class = "location-scale") ~ w
corpair(species, block = "p", class = "scale-scale") ~ w
```

Use `rho12 = ~ assay_context` for residual correlation predictors today. Use
`corpairs(fit, level = "phylogenetic")` to extract the fitted constant latent
phylogenetic correlations today.

## Planned, not implemented: spatial random intercepts

Spatial fields are deliberately planned to look parallel to phylogenetic
effects, but the spatial likelihood is not implemented yet. The planned syntax
is:

```r
drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 | site, coords = coords),
    sigma ~ temp
  ),
  family = gaussian(),
  data = dat
)
```

The planned `coords` and `mesh` arguments play different roles. `coords`
should name the observation or site coordinates in the modelling data, ideally
in a projected coordinate system such as UTM rather than raw longitude and
latitude. A `mesh` is the computational scaffold for an SPDE/GMRF field: a
triangular network of vertices where the latent spatial field is represented
and then projected back to observations. Users should not interpret mesh
vertices as biological sampling units; they are numerical support points for
the spatial surface.

The spatial term is deliberately parallel to the phylogenetic term:

```r
phylo(1 | species, tree = tree)     # implemented for mu and matching mu1/mu2
phylo(1 | p | species, tree = tree) # implemented for labelled q4 bivariate blocks
spatial(1 | site, coords = coords)  # planned
spatial(1 | site, mesh = mesh)      # planned
```

The mathematical role is the same: add a structured Gaussian random intercept
to a distributional predictor. The computational route differs. `phylo()`
starts from an ultrametric tree with branch lengths and builds a sparse
tree-derived precision. `spatial()` will start from coordinates or a mesh and
build an SPDE/GMRF precision.

The computational plan is to keep using sparse A-inverse ideas for phylogeny
and add SPDE/GMRF structure for spatial fields, with provenance notes for any
code selectively ported from `gllvmTMB`.

Structured phylogenetic and spatial random slopes will be staged more
conservatively than ordinary grouped random slopes. The first structured-effect
target is intercept-only `mu`; the next should be one structured `mu` slope.
Small sets of structured slopes, interaction slopes, and structured effects in
`sigma` or `rho12` should wait until simulation recovery shows that the simpler
models are identifiable.

## Meta-analysis and future bivariate structure

In meta-analysis, the separation among known sampling covariance,
phylogenetic dependence, and spatial dependence is especially transparent:

\[
\begin{aligned}
e
  &\sim \operatorname{MVN}(\mathbf{0}, V)
  &&\text{known sampling covariance},\\
p_{species}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2 A)
  &&\text{phylogenetic structure},\\
\ell_{location}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{space}^2 M)
  &&\text{spatial structure}.
\end{aligned}
\]

The known matrix `V` handles sampling uncertainty. The phylogenetic and spatial
terms add extra biological or environmental dependence after that known
uncertainty has been included. The same design will let `drmTMB` support
structured Gaussian meta-analysis while keeping ecological and evolutionary
meta-analyses readable.

The first phylogenetic target is the location-scale model. In bivariate
Gaussian models, matching `mu1`/`mu2` phylogenetic location terms estimate the
mean-mean phylogenetic correlation. Matching labelled `phylo()` terms across
`mu1`, `mu2`, `sigma1`, and `sigma2` now fit the first constant q=4
phylogenetic location-scale block: one location-location row, four
location-scale rows, and one scale-scale row in `corpairs(level =
"phylogenetic")`. The six q=4 correlations are latent phylogenetic
random-effect correlations, not residual `rho12`.

The first fitted bivariate phylogenetic location-coscale syntax looks like
this:

```r
fit_biv_phylo <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle + phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle + phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

This syntax fits correlated phylogenetic random intercepts in `mu1` and `mu2`.
It estimates two phylogenetic location SDs and one phylogenetic mean-mean
correlation. The `sigma1`, `sigma2`, and residual `rho12` formulas remain
ordinary fixed-effect distributional parameters in this first slice.
Use `phylo(1 | p | species, tree = tree)` in both location formulas when you
want the fitted phylogenetic mean-mean row to use a block label such as `p`.

The full constant q=4 location-scale form uses an explicit covariance-block
label:

```r
fit_biv_phylo_q4 <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle + phylo(1 | p | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle + phylo(1 | p | species, tree = tree),
    sigma1 = ~ lifestyle + phylo(1 | p | species, tree = tree),
    sigma2 = ~ lifestyle + phylo(1 | p | species, tree = tree),
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)

corpairs(fit_biv_phylo_q4, level = "phylogenetic")
```

Partial q=4 syntax, unlabelled all-four syntax, random slopes, and multiple
phylogenetic blocks remain outside this first fitted path. For the first fitted
mean-mean phylogenetic correlation, use `corpars$phylo`,
`corpairs(fit_biv_phylo, level = "phylogenetic")`, or
`summary(fit_biv_phylo)$covariance`. Run `check_drm(fit_biv_phylo)` before
interpreting that row: the current diagnostic flags phylogenetic correlations
near the boundary and phylogenetic location SDs that are tiny relative to the
matching residual scales. If `check_drm(fit_biv_phylo)` does not warn and the
fitted object kept its TMB object, a direct profile interval is available for
the phylogenetic mean-mean correlation target. The q=4 correlations are derived
from `theta_phylo`, so direct profile intervals for those six rows are not yet
claimed:

```r
phylo_target <- "cor:phylo:cor(mu1:(Intercept),mu2:(Intercept) | phylo | species)"

corpairs(fit_biv_phylo, level = "phylogenetic")
summary(fit_biv_phylo)$covariance
confint(fit_biv_phylo, parm = phylo_target, method = "profile")
```

For a staged phylogenetic versus non-phylogenetic species comparison, add an
ordinary labelled species block beside the fitted phylogenetic layer:

```r
fit_biv_phylo_species <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle +
      (1 | species_residual | species) +
      phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle +
      (1 | species_residual | species) +
      phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

That fit has an ordinary group-level species correlation, a phylogenetic
mean-mean correlation, and residual `rho12`. `corpairs()` keeps those layers in
separate rows. Because the ordinary group block and the phylogenetic block use
the same species factor, `check_drm()` records a note rather than letting the
two species-level correlations look automatically identifiable.

The residual `rho12` term remains distinct from the phylogenetic or spatial
correlation matrices. `rho12` is the residual coupling between two responses at
the observation level; `A` and `M` describe dependence among species or places.
In future bivariate structured models, `drmTMB` should also estimate
level-specific correlations such as phylogenetic mean-mean correlations,
non-phylogenetic species correlations, spatial field correlations, and
mean-scale correlations. These are not residual `rho12`; they should be
reported as structured or group-level covariance summaries.
