---
title: "Structural dependence details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural dependence details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)

structural_eye_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.y = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "none"
    )
}

balanced_ultrametric_tree <- function(n_tip = 16L) {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0("sp_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}

simulate_phylo_q4_traits <- function(seed = 20260802) {
  set.seed(seed)
  tree <- balanced_ultrametric_tree(32)
  A_tip <- drmTMB:::drm_phylo_tip_covariance(tree)
  sd_phylo <- c(mu1 = 0.60, mu2 = 0.50, sigma1 = 0.18, sigma2 = 0.16)
  corr_phylo <- matrix(
    c(
      1.00, 0.50, 0.10, 0.05,
      0.50, 1.00, 0.05, 0.10,
      0.10, 0.05, 1.00, 0.25,
      0.05, 0.10, 0.25, 1.00
    ),
    nrow = 4,
    byrow = TRUE,
    dimnames = list(names(sd_phylo), names(sd_phylo))
  )
  z_phylo <- matrix(rnorm(length(tree$tip.label) * 4), ncol = 4)
  phylo_effect <- t(chol(A_tip)) %*%
    z_phylo %*%
    chol(diag(sd_phylo) %*% corr_phylo %*% diag(sd_phylo))
  dimnames(phylo_effect) <- list(tree$tip.label, names(sd_phylo))

  n_each <- 6
  species <- rep(tree$tip.label, each = n_each)
  n <- length(species)
  climate <- rnorm(n)
  habitat_variability <- rnorm(n)
  assay_context <- rep(rep(c(-0.5, 0.5), each = n_each / 2), length(tree$tip.label))

  eta_heat <- 0.35 + 0.30 * climate + phylo_effect[species, "mu1"]
  eta_desiccation <- -0.20 - 0.25 * climate + phylo_effect[species, "mu2"]
  log_sigma_heat <- -1.15 + 0.20 * habitat_variability +
    phylo_effect[species, "sigma1"]
  log_sigma_desiccation <- -1.05 - 0.15 * habitat_variability +
    phylo_effect[species, "sigma2"]
  rho12 <- 0.15
  e_heat <- rnorm(n)
  e_desiccation <- rho12 * e_heat + sqrt(1 - rho12^2) * rnorm(n)

  list(
    tree = tree,
    data = data.frame(
      species = species,
      climate = climate,
      habitat_variability = habitat_variability,
      assay_context = assay_context,
      heat_tolerance = eta_heat + exp(log_sigma_heat) * e_heat,
      desiccation_tolerance = eta_desiccation +
        exp(log_sigma_desiccation) * e_desiccation
    )
  )
}

simulate_spatial_location_data <- function(seed = 20260534) {
  set.seed(seed)
  sites <- paste0("site_", seq_len(12))
  theta <- seq(0, 1.8 * pi, length.out = length(sites))
  coords <- data.frame(
    x = cos(theta) + seq_along(sites) / 30,
    y = sin(theta)
  )
  rownames(coords) <- sites

  distance <- as.matrix(dist(coords))
  range <- stats::median(distance[distance > 0])
  spatial_cov <- exp(-distance / range)
  diag(spatial_cov) <- diag(spatial_cov) + 1e-6
  spatial_intercept <- as.vector(
    t(chol(spatial_cov)) %*% rnorm(length(sites), sd = 0.45)
  )
  spatial_slope <- as.vector(
    t(chol(spatial_cov)) %*% rnorm(length(sites), sd = 0.20)
  )
  names(spatial_intercept) <- sites
  names(spatial_slope) <- sites

  site <- rep(sites, each = 6)
  temp <- rnorm(length(site))
  depth <- rnorm(length(site))
  sigma <- exp(-1.25 + 0.15 * depth)
  y <- 0.4 + 0.35 * temp + 0.20 * depth +
    spatial_intercept[site] + spatial_slope[site] * depth +
    rnorm(length(site), sd = sigma)

  list(
    data = data.frame(y = y, temp = temp, depth = depth, site = site),
    coords = coords
  )
}

simulate_animal_marker_data <- function(seed = 20260519) {
  set.seed(seed)
  n_individual <- 40
  n_each <- 4
  individual_levels <- paste0("id_", seq_len(n_individual))
  individual <- factor(rep(individual_levels, each = n_each))
  sex <- factor(rep(rep(c("female", "male"), each = n_each), n_individual / 2))
  age <- runif(length(individual), 0, 2)
  line_levels <- paste0("line_", seq_len(10))
  breeding_line <- factor(rep(rep(line_levels, each = 4), each = n_each))

  A <- outer(seq_len(n_individual), seq_len(n_individual), function(i, j) {
    0.35^abs(i - j)
  })
  diag(A) <- diag(A) + 0.10
  dimnames(A) <- list(individual_levels, individual_levels)
  Ainv <- solve(A)

  G <- outer(seq_along(line_levels), seq_along(line_levels), function(i, j) {
    0.25^abs(i - j)
  })
  diag(G) <- diag(G) + 0.10
  dimnames(G) <- list(line_levels, line_levels)
  Ginv <- solve(G)

  individual_z_body <- rnorm(n_individual)
  individual_z_beak <- 0.45 * individual_z_body +
    sqrt(1 - 0.45^2) * rnorm(n_individual)
  individual_effect_body <- as.vector(t(chol(A)) %*% individual_z_body) * 0.35
  individual_effect_beak <- as.vector(t(chol(A)) %*% individual_z_beak) * 0.28
  names(individual_effect_body) <- individual_levels
  names(individual_effect_beak) <- individual_levels

  line_z_body <- rnorm(length(line_levels))
  line_z_beak <- 0.35 * line_z_body +
    sqrt(1 - 0.35^2) * rnorm(length(line_levels))
  line_effect_body <- as.vector(t(chol(G)) %*% line_z_body) * 0.25
  line_effect_beak <- as.vector(t(chol(G)) %*% line_z_beak) * 0.20
  names(line_effect_body) <- line_levels
  names(line_effect_beak) <- line_levels

  sigma_body <- exp(-1.0 + 0.15 * (sex == "male"))
  sigma_beak <- exp(-1.1 - 0.10 * (sex == "male"))
  e_body <- rnorm(length(individual))
  e_beak <- 0.20 * e_body + sqrt(1 - 0.20^2) * rnorm(length(individual))
  body_size <- 2.1 + 0.45 * age + 0.20 * (sex == "male") +
    individual_effect_body[individual] + line_effect_body[breeding_line] +
    sigma_body * e_body
  beak_length <- 1.4 + 0.30 * age - 0.12 * (sex == "male") +
    individual_effect_beak[individual] + line_effect_beak[breeding_line] +
    sigma_beak * e_beak

  list(
    data = data.frame(
      body_size = body_size,
      beak_length = beak_length,
      age = age,
      sex = sex,
      individual = individual,
      breeding_line = breeding_line
    ),
    A = A,
    Ainv = Ainv,
    G = G,
    Ginv = Ginv
  )
}
```

Structural dependence means a random effect has a known dependence structure
instead of independent group levels. Read the grammar in biological order:
animal models for pedigree or additive relatedness, phylogenetic models for
tree-structured relatedness, spatial models for coordinate or mesh structure,
combined phylogenetic-plus-spatial models when both sources matter, and
`relmat()` for other known relatedness or precision matrices.

The current fitted `drmTMB` paths are a subset of that ladder. For Gaussian
models, `phylo()`, coordinate `spatial()`, `animal()`, and `relmat()` each fit
one univariate `mu` intercept and one numeric `mu` slope. These one-slope paths
estimate independent structured intercept and slope fields with separate SDs.
Those Gaussian `mu` terms do not themselves establish multiple or labelled
slopes, intercept-slope correlations, or structured residual `rho12`.
Separate exact q1 `sigma` and non-Gaussian gates are fitted for the providers
and families named in the route table below. A first bivariate path also fits matching intercept-only
`phylo()` terms in `mu1` and `mu2`, estimating the two phylogenetic location
SDs and their mean-mean correlation. Matching labelled `phylo()` terms across
`mu1`, `mu2`, `sigma1`, and `sigma2` now fit the first constant bivariate
phylogenetic location-scale block. The first
predictor-dependent phylogenetic correlation regression is also fitted for the
q=2 `mu1`-`mu2` location-location row. The first coordinate-based spatial
paths, `spatial(1 | site, coords = coords)` and
`spatial(1 + x | site, coords = coords)`, are fitted for univariate Gaussian
location models. Matching labelled
`spatial(1 | p | site, coords = coords)` terms in bivariate `mu1` and `mu2`
fit the first coordinate-spatial mean-mean correlation, reported by
`corpairs(level = "spatial")`; matching labelled spatial terms across `mu1`,
`mu2`, `sigma1`, and `sigma2` fit the first constant coordinate-spatial q=4
location-scale block. `animal()` and `relmat()` fit the same first Gaussian
intercept and one-slope `mu` pattern when supplied with a pedigree, precomputed
relationship, or precision matrix, such as `animal(1 | id, pedigree = pedigree)`,
`animal(1 + x | id, Ainv = Ainv)`, or `relmat(1 + x | id, Q = Q)`.
Artifact routing is narrower than fitted syntax. The one-slope routes have
manual Phase 18 artifact tasks named `phylo_mu_slope`, `spatial_mu_slope`,
`animal_mu_slope`, and `relmat_mu_slope`. These tasks are opt-in, excluded from
`task = "all"`, and are not recovery, coverage, or power evidence.
Mesh/SPDE spatial fields, combined phylogenetic-plus-spatial location models,
multiple or labelled spatial slopes, spatial slope correlations,
non-Gaussian spatial effects outside the exact ordinary Poisson/NB2 q1 spatial
`mu` intercept-plus-one-slope, recovery-grade NB2 q1 spatial `sigma`, Student-t
spatial `mu`, Poisson spatial `zi`, fixed-`zi` Poisson spatial `mu`, and
fixed-`zi` NB2 spatial `mu` gates, and
predictor-dependent spatial correlations remain planned extensions. The q1 spatial `sigma` one-slope route has
point-fit/extractor evidence, but its interval gate remains blocked.

The scientific questions for the structural ladder are simple:

> Do predictors explain trait means after accounting for resemblance among
> relatives, species, or nearby sites? Does a response still have smooth
> coordinate-based site deviations, or does a predictor's slope vary smoothly
> across sites?

For animal models, `individual` or `animal` indexes rows in the data. A small
`id`/`dam`/`sire` pedigree or a precomputed additive relatedness or
inverse-relatedness matrix supplies expected genetic similarity. For
phylogenetic models, `species` indexes the rows in the data and
`tree` supplies the relatedness among those species. For spatial models, `site`
indexes the rows in the data and `coords` supplies one coordinate pair per
site, or one row per observation when coordinates are constant within site. In
the fitted animal, phylogenetic, spatial, and `relmat()` paths, the structured
term adds Gaussian coefficients to the location parameter `mu`. The intercept
path adds an individual, species, site, or matrix-indexed deviation in mean
response. The one-slope paths let the effect of one numeric predictor vary
across the same structured levels.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

In this article, a location parameter such as `mu` describes the expected
trait value, a scale parameter such as `sigma` describes residual or
random-effect variability, a shape parameter such as `nu` describes a
family-specific feature such as tail weight or skewness, and a coscale
parameter models residual correlation, such as bivariate `rho12`. Structured
animal, phylogenetic, and spatial effects are not residual coscale terms unless
the fitted model explicitly includes a residual `rho12` formula.

## Reader route: animal, phylo, spatial, phylo+spatial, relmat

Read this page as a five-step structural-dependence ladder. Step 1 asks whether
individuals resemble each other because of pedigree or additive relatedness.
Step 2 asks whether related species remain similar after fixed effects. Step 3
asks whether nearby sites remain similar after fixed effects. Step 4 asks
whether both species relatedness and site coordinates are needed in the same
model. Step 5 is the lower-level `relmat()` route for known dependence matrices
that do not belong to the named animal, phylogenetic, or spatial routes. In the
current package, Steps 1, 2, 3, and 5 have fitted Gaussian paths plus exact
row-specific non-Gaussian gates; Step 4 is documented planned syntax.

For a one-response phylogenetic-plus-spatial endpoint such as
`heat_tolerance`, the conceptual full model is:

\[
\begin{aligned}
y_i &\sim \operatorname{Normal}(\mu_i, \sigma_i^2),\\
\mu_i &=
  \mathbf{x}_i^\top\boldsymbol{\beta}_{\mu}
  + a_{\mathrm{species}[i]} + s_{\mathrm{site}[i]},\\
\mathbf{a} &\sim
  \operatorname{MVN}(\mathbf{0}, \sigma^2_{\mathrm{phylo}}\mathbf{A}),\\
\mathbf{s} &\sim
  \operatorname{MVN}(\mathbf{0}, \sigma^2_{\mathrm{space}}\mathbf{M}),\\
\log(\sigma_i) &=
  \mathbf{z}_i^\top\boldsymbol{\beta}_{\sigma}.
\end{aligned}
\]

Here \(y_i\) is the observed trait, for example heat tolerance for individual
or species-site observation \(i\). The vector \(\mathbf{x}_i\) contains
location predictors such as habitat temperature or body mass, and
\(\boldsymbol{\beta}_{\mu}\) are their average effects on the expected trait
value. The term \(a_{\mathrm{species}[i]}\) is the phylogenetic deviation for the
species attached to observation \(i\), \(\mathbf{A}\) is the phylogenetic
covariance implied by the tree, and \(\sigma_{\mathrm{phylo}}\) is the SD of
tree-structured location deviations. The term \(s_{\mathrm{site}[i]}\) is the
spatial deviation for the sampling site, \(\mathbf{M}\) is the coordinate-based
spatial covariance, and \(\sigma_{\mathrm{space}}\) is the SD of spatial
location deviations. The vector \(\mathbf{z}_i\) contains residual-scale
predictors and \(\boldsymbol{\beta}_{\sigma}\) are their effects on log
residual SD.

The animal-model analogue replaces the species term with an individual
additive effect, for example
\(\mathbf{g} \sim \operatorname{MVN}(\mathbf{0},
\sigma^2_{\mathrm{animal}}\mathbf{A}_{\mathrm{pedigree}})\). That is the
meaning of `animal(1 | individual, pedigree = pedigree)`,
`animal(1 | individual, A = A)`, or `animal(1 | individual, Ainv = Ainv)`.
The pedigree route builds a dense additive relationship matrix for this first
Gaussian location path; sparse large-pedigree precision construction remains
planned.

Read the current route status this way:

| Route | Current status | Use when | Syntax to read |
| --- | --- | --- | --- |
| 1. Animal | Implemented for the documented Gaussian routes plus exact ordinary Poisson/NB2 q1 `mu` intercept-plus-one-slope, NB2 q1 `sigma`, and beta animal gates at their recorded tiers; sparse large-pedigree construction remains planned | Individuals have known additive relatedness from a pedigree or relationship matrix, and the biological question is an animal model or quantitative-genetic random effect. | `animal(1 | individual, pedigree = pedigree)`, `animal(1 + x | individual, Ainv = Ainv)`, or matching labelled Gaussian terms; consult the live ledger before using a non-Gaussian animal gate |
| 2. Phylo | Implemented for the documented Gaussian routes plus exact ordinary Poisson/NB2 q1 `mu` intercept-plus-one-slope, NB2 q1 `sigma`, Student-t q1 `nu`, cumulative-logit q1 `mu`, and recovery-grade Gamma/lognormal q1 `mu` gates | Species share evolutionary history, and related species may have similar trait means, trait variances, or cross-trait latent correlations. | `phylo(1 | species, tree = tree)`, `phylo(1 + x | species, tree = tree)`, labelled Gaussian blocks, or one exact non-Gaussian gate at its recorded tier |
| 3. Spatial | Implemented for the documented Gaussian routes plus exact ordinary Poisson/NB2 q1 `mu` intercept-plus-one-slope, NB2 q1 `sigma`, Student-t q1 `mu`, Poisson q1 `zi`, fixed-`zi` Poisson q1 `mu`, and fixed-`zi` NB2 q1 `mu` gates | Nearby sampling sites may share smooth deviations in the expected response, in one numeric slope such as `depth`, or in paired response means. | `spatial(1 | site, coords = coords)`, `spatial(1 + depth | site, coords = coords)`, matching labelled Gaussian terms, or one exact non-Gaussian gate at its recorded tier |
| 4. Phylo + spatial | Planned, not fitted yet | The same response could be structured by both species relatedness and site coordinates. | Fit separate phylogenetic and spatial sensitivity models for now; simultaneous `phylo()` plus `spatial()` in `mu` is rejected until the joint model has identifiability checks. |
| 5. `relmat()` | Implemented for the documented Gaussian routes plus exact ordinary Poisson/NB2 q1 `mu` intercept-plus-one-slope, NB2 q1 `sigma`, recovery-grade Gamma/lognormal q1 `mu`, and truncated-NB2 q1 `hu` gates | A known covariance or precision matrix does not naturally belong to animal, phylogenetic, or spatial syntax. | `relmat(1 | id, K = K)`, `relmat(1 + x | id, Q = Q)`, matching labelled Gaussian terms, or one exact non-Gaussian gate at its recorded tier |

That status boundary is deliberate. A combined model like this is
scientifically natural:

```r
# Planned, not fitted yet:
heat_tolerance ~ climate +
  phylo(1 | species, tree = tree) +
  spatial(1 | site, coords = coords)
```

It requires careful checks that the tree-structured and coordinate-structured
SDs are both identifiable before `drmTMB` treats the syntax as runnable.

For animal and `relmat()` models, use the known-matrix slices when the matrix
is already available and the target is a Gaussian `mu` or `sigma` intercept,
one numeric Gaussian `mu` slope, a two-response q=2 location covariance, or a
constant q=4 location-scale block. If the question is repeatability without
known relatedness, fit an ordinary `(1 | individual)` or `(1 | line)` Gaussian
model and report that it ignores pedigree or matrix structure. If the question
is species shared ancestry, use the fitted `phylo()` route. If the question is
site structure, use the fitted coordinate-spatial route. If the matrix is
sampling covariance among effect-size estimates, use `meta_V(V = V)`, not
`relmat()`; `relmat()` is for latent random-effect relatedness or precision
matrices.

| Current question | Syntax | Fitted action now |
| --- | --- | --- |
| Animal-model additive individual deviations from a pedigree | `animal(1 | individual, pedigree = pedigree)` | Use the fitted univariate Gaussian `mu` first slice; inspect `check_drm()` and `profile_targets()` before interpreting the SD. |
| Additive genetic variance from a precomputed relationship matrix | `animal(1 | individual, A = A)` or `animal(1 | individual, Ainv = Ainv)` | Use the fitted univariate Gaussian `mu` first slice; inspect `check_drm()` and `profile_targets()` before interpreting the SD. |
| Experimental-line, genomic, or laboratory relatedness that is not a pedigree, tree, or coordinate surface | `relmat(1 | line, K = K)` or `relmat(1 | line, Q = Q)` | Use the fitted univariate Gaussian `mu` first slice, or switch to `phylo()` / `spatial()` when the known structure belongs to those named routes. |
| Known sampling covariance among effect-size estimates | `meta_V(V = V)` | Use the fitted Gaussian meta-analysis route; this is observation covariance, not a latent relatedness random effect. |

## Animal model with a known-matrix first slice

An applied animal-model question might be: after accounting for age and sex,
is there additive individual-level variation in body size that follows a known
pedigree? The fitted first slice uses a precomputed additive relationship or
inverse relationship matrix:

```r
fit_animal <- drmTMB(
  drm_formula(
    body_size ~ age + sex + animal(1 | individual, Ainv = Ainv),
    sigma ~ sex
  ),
  family = gaussian(),
  data = growth
)
```

That animal effect puts the individual-level location deviations in a
precomputed additive relationship matrix. It is not a residual `rho12` coscale
model, and it does not use `meta_V(V = V)`, which is reserved for known
sampling covariance among
observations or effect-size estimates.

```{r}
animal_example <- simulate_animal_marker_data(seed = 1)
growth <- animal_example$data
Ainv <- animal_example$Ainv

fit_animal_known <- drmTMB(
  drm_formula(
    body_size ~ age + sex + animal(1 | individual, Ainv = Ainv),
    sigma ~ sex
  ),
  family = gaussian(),
  data = growth
)

check_drm(fit_animal_known)
summary(fit_animal_known)
```

If two traits were measured on the same individuals, a matching label such as
`p` asks whether the additive deviations in the two location predictors covary.
That covariance is a latent animal-model row; residual `rho12` still describes
within-observation residual correlation after the known relatedness effect is
accounted for.

```{r}
fit_animal_q2_known <- drmTMB(
  drm_formula(
    mu1 = body_size ~ age + sex +
      animal(1 | p | individual, Ainv = Ainv),
    mu2 = beak_length ~ age + sex +
      animal(1 | p | individual, Ainv = Ainv),
    sigma1 = ~ sex,
    sigma2 = ~ sex,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = growth
)

animal_q2_pairs <- corpairs(
  fit_animal_q2_known,
  level = "animal"
)
animal_q2_pairs
animal_q2_checks <- check_drm(fit_animal_q2_known)
animal_q2_checks[grepl("animal|convergence|hessian", animal_q2_checks$check), ]
```

```{r animal-q2-correlation-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "Animal-model q=2 location-location point estimate from `corpairs()`; the dotted vertical line marks zero correlation. Interval calibration for this intercept-only row is not yet validated.", fig.alt = "Single-row point plot for the animal-model mean-mean correlation, with a hollow estimate to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  animal_q2_display <- animal_q2_pairs
  animal_q2_display$display_label <- "Animal\nmu1-mu2"
  plot_corpairs(
    animal_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("animal" = "#0072B2")) +
    ggplot2::scale_fill_manual(values = c("animal" = "#0072B2")) +
    structural_eye_theme() +
    ggplot2::labs(
      title = "Animal-model latent mean correlation",
      subtitle = "Point estimate only; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

Use `relmat()` when the matrix is not naturally a pedigree, tree, or spatial
coordinate surface. Here `Ginv` is a precision matrix for experimental breeding
lines, so the fitted row is a lower-level relatedness covariance rather than an
animal-model covariance.

```{r}
Ginv <- animal_example$Ginv

fit_relmat_q2_known <- drmTMB(
  drm_formula(
    mu1 = body_size ~ age + sex +
      relmat(1 | p | breeding_line, Q = Ginv),
    mu2 = beak_length ~ age + sex +
      relmat(1 | p | breeding_line, Q = Ginv),
    sigma1 = ~ sex,
    sigma2 = ~ sex,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = growth
)

relmat_q2_pairs <- corpairs(
  fit_relmat_q2_known,
  level = "relmat"
)
relmat_q2_pairs
relmat_q2_targets <- profile_targets(fit_relmat_q2_known)
relmat_q2_targets[
  grepl("relmat", relmat_q2_targets$parm),
  c("parm", "target_type", "profile_ready")
]
```

```{r relmat-q2-correlation-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "`relmat()` q=2 location-location point estimate from `corpairs()`; the dotted vertical line marks zero correlation. Interval calibration for this intercept-only row is not yet validated.", fig.alt = "Single-row point plot for the known-matrix relmat mean-mean correlation, with a hollow estimate to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  relmat_q2_display <- relmat_q2_pairs
  relmat_q2_display$display_label <- "relmat\nmu1-mu2"
  plot_corpairs(
    relmat_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("relmat" = "#6A6A6A")) +
    ggplot2::scale_fill_manual(values = c("relmat" = "#6A6A6A")) +
    structural_eye_theme() +
    ggplot2::labs(
      title = "Known-matrix latent mean correlation",
      subtitle = "Point estimate only; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

When the relationship matrix is unavailable, fit an ordinary grouped Gaussian
model as a sensitivity analysis and state the omitted structure:

```{r}
fit_repeatability <- drmTMB(
  drm_formula(
    body_size ~ age + sex + (1 | individual),
    sigma ~ sex
  ),
  family = gaussian(),
  data = growth
)

check_drm(fit_repeatability)
```

That fallback asks whether repeated observations of the same individual share
a common location deviation. It does not estimate additive genetic variance,
does not use pedigree relatedness among individuals, and cannot distinguish
genetic resemblance from permanent environmental or measurement effects. If
the scientific claim requires additive genetic variance, use the fitted
`animal(A/Ainv)` route once the matrix is ready; report the ordinary
random-effect fit only as a sensitivity check.

For matrix inputs, keep the representation explicit. A covariance matrix
such as `A` or `K` is the dense VCV form. It is useful for small examples and
parity checks, but it does not by itself justify large-pedigree or large-matrix
claims. A precision or inverse relationship matrix such as `Ainv` or `Q` is the
preferred scalable route for larger examples. The first slice validates
row-name matching, determinant handling, positive definiteness, diagnostics,
and recovery tests, but speed claims still need larger sparse-matrix evidence.
This is the same distinction as the implemented phylogenetic tree path: users
pass `tree = tree`, and `drmTMB` builds the sparse augmented A-inverse
internally rather than asking users to hand-write a dense tip VCV.

### Parity target for animal, spatial, and relmat

The long-term target is not just to parse `animal()`, `spatial()`, or
`relmat()`. These routes should eventually answer the same classes of
questions as the fitted phylogenetic route, whenever the corresponding
dependence structure is scientifically meaningful and identifiable. The table
below is a roadmap boundary, not a fitted-claim table.

| Layer | Fitted phylogenetic example | Parity target for animal, spatial, and relmat |
| --- | --- | --- |
| Univariate location structure | `phylo(1 | species, tree = tree)` | `animal(1 | individual, pedigree = pedigree)`, `spatial(1 | site, coords = coords)`, and `relmat(1 | line, K = K)` in `mu` |
| Bivariate location-location correlation | Matching labelled `phylo()` terms in `mu1` and `mu2`, reported by `corpairs()` | Matching labelled `spatial()`, `animal()`, and `relmat()` terms now fit the first q=2 slices when their coordinate or known-matrix inputs are available |
| Location-scale q=4 block | Matching labelled `phylo()` terms across `mu1`, `mu2`, `sigma1`, and `sigma2` | Matching labelled all-four spatial, animal, and `relmat()` blocks now fit constant q=4 location-scale covariance |
| Predictor-dependent structured correlation | q=2 phylogenetic `corpair(..., level = "phylogenetic") ~ ecology` | q=2 animal, spatial, and relmat `corpair()` regressions only after constant q=2 fits are stable |
| Direct SD surfaces | `sd_phylo()`, `sd_phylo1()`, and `sd_phylo2()` | Source-specific direct-SD siblings after naming, matrix scale, and interpretation are clear |

Spatial is partway up this ladder: `spatial(1 | site, coords = coords)` can
enter Gaussian `mu`, Gaussian `sigma`, or matching univariate `mu`/`sigma`
blocks; one independent coordinate-spatial `mu` slope, the first bivariate
coordinate-spatial q=2 `mu1`/`mu2` covariance, and the constant q=4
location-scale covariance are fitted. Animal and `relmat()` now have pedigree or
known-matrix Gaussian `mu` and `sigma` intercepts, matching univariate
`mu`/`sigma` correlations, q1 `sigma` one-slope routes, matching labelled q=2
bivariate location covariance, and matching all-four constant q=4
location-scale covariance. The exact A-matrix animal and K/Q relmat sigma
slopes are inference-ready with caveats; the spatial sigma-slope interval gate
remains blocked. Spatial `corpair()`, mesh/SPDE fields, pedigree/Ainv bridge
marshalling, sparse large-pedigree construction, multiple or labelled slopes,
and animal/`relmat()` predictor-dependent `corpair()` routes remain planned.

The focused animal, spatial, and `relmat()` pages cover the current q=2 and
constant q=4 routes. A fitted animal q=4 block looks like:

```r
fit_animal_q4 <- drmTMB(
  drm_formula(
    mu1 = body_size ~ age + sex +
      animal(1 | p | individual, pedigree = pedigree),
    mu2 = beak_length ~ age + sex +
      animal(1 | p | individual, pedigree = pedigree),
    sigma1 = ~ sex + animal(1 | p | individual, pedigree = pedigree),
    sigma2 = ~ sex + animal(1 | p | individual, pedigree = pedigree),
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = growth
)

# Fitted spatial q=2 and q=4 sibling terms:
spatial(1 | p | site, coords = coords)

# Fitted relatedness q=2 and q=4 sibling terms:
relmat(1 | p | line, K = G)
relmat(1 | p | line, Q = Ginv)
```

The next richer animal, spatial, and `relmat()` targets are structured slopes
and predictor-dependent structured correlations, which remain planned.

## Current implementation status

This page separates the fitted paths from the roadmap. The fitted structural
pieces are useful, but they are not yet the full animal, phylo, spatial,
phylo+spatial, and `relmat()` location-scale-coscale ladder.

| Model piece | Status | What it means |
| --- | --- | --- |
| `animal(1 | individual, pedigree = pedigree)` | Implemented first slice | A univariate Gaussian `mu` random intercept using a dense additive relationship matrix built from `id`, `dam`, and `sire` pedigree columns. Sparse large-pedigree precision construction remains planned. |
| `animal(1 | individual, A = A)` or `animal(1 | individual, Ainv = Ainv)` | Implemented first slice | A univariate Gaussian `mu` random intercept using a precomputed additive relatedness or inverse-relatedness matrix. Use `pedigree`, `A`, or `Ainv`; keep `V` for known sampling covariance in meta-analysis. |
| `animal(1 + x | individual, pedigree = pedigree)` or `animal(1 + x | individual, Ainv = Ainv)` | Implemented first one-slope slice | One trait has independent animal-model intercept and numeric slope fields in the location predictor. The two fields have separate SDs and do not estimate an intercept-slope correlation. |
| Matching labelled `animal(1 | p | individual, pedigree = pedigree)`, `animal(1 | p | individual, A = A)`, or `animal(1 | p | individual, Ainv = Ainv)` terms in bivariate `mu1` and `mu2` | Implemented first q=2 slice | Two Gaussian responses have correlated animal-model location deviations from the same pedigree-derived or known matrix. `corpairs(level = "animal")` reports this mean-mean row separately from residual `rho12`. |
| `phylo(1 | species, tree = tree)` in univariate `mu` | Implemented | One trait has a phylogenetic random intercept in the location predictor. |
| `phylo(1 + x | species, tree = tree)` in univariate `mu` | Implemented first one-slope slice | One trait has independent phylogenetic intercept and numeric slope fields in the location predictor. The two fields share the same tree-derived covariance, have separate SDs, and do not estimate an intercept-slope correlation. |
| Matching `phylo(1 | species, tree = tree)` terms in bivariate `mu1` and `mu2` | Implemented first slice | Two traits have correlated phylogenetic location deviations; `corpairs(level = "phylogenetic")` reports the mean-mean row separately from residual `rho12`. |
| Matching labelled `phylo(1 | p | species, tree = tree)` terms in bivariate `mu1` and `mu2` | Implemented | The label is preserved in SD names, the fitted mean-mean correlation, `corpairs(block = "p")`, and profile-target names. |
| Ordinary `(1 | p | id)` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first slice | One ordinary q=4 group block estimates six latent location-location, location-scale, and scale-scale correlations. |
| Matching labelled `phylo()` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first slice | One constant q=4 phylogenetic block estimates four endpoint SDs and six latent correlations: one location-location, four location-scale, and one scale-scale. Partial, unlabelled, mismatched, and slope forms remain rejected. |
| `sd_phylo(species) ~ z` | Implemented for univariate Gaussian `mu` | This is the separate Box 1 direct-SD family: species-level predictors model the SD of a phylogenetic location random effect at observed tips. |
| `sd_phylo1(species) ~ z1` / `sd_phylo2(species) ~ z2` | Implemented for bivariate `mu1`/`mu2` | Response-specific species-level predictors model the SD surfaces of the two phylogenetic location random effects. This remains a Family B direct-SD model, not a residual `sigma1`/`sigma2` model and not a q=4 location-scale block. |
| `corpair(species, level = "phylogenetic", block = "p", from = "mu1", to = "mu2") ~ ecology` | Implemented for q=2 `mu1`/`mu2` | Species-level predictors model the same-species latent phylogenetic location-location correlation through a positive-definite two-field loading contract. `corpairs(level = "phylogenetic")` reports the modelled response-scale mean, range, and number of species values. |
| `spatial(1 | site, coords = coords)` | Implemented first slice | One trait has a coordinate-based spatial random intercept in the location predictor. This is a fixed coordinate covariance foundation, not the future mesh/SPDE path. |
| `spatial(1 + x | site, coords = coords)` | Implemented first one-slope slice | One trait has independent coordinate-spatial intercept and numeric slope fields in the location predictor. The two fields share the same coordinate precision, have separate SDs, and do not estimate an intercept-slope correlation. |
| Matching `spatial(1 | p | site, coords = coords)` terms in bivariate `mu1` and `mu2` | Implemented first q=2 slice | Two Gaussian responses have correlated coordinate-spatial location deviations. `corpairs(level = "spatial")` and `summary()$covariance` report this mean-mean row separately from residual `rho12`. |
| Matching labelled `spatial()` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first q=4 slice | Two Gaussian responses have a constant coordinate-spatial location-scale covariance block with four endpoint SDs and six derived latent spatial correlations. |
| `phylo(1 | species, tree = tree) + spatial(1 | site, coords = coords)` in the same `mu` formula | Planned | This is the intended third structural-dependence route, but the fitter currently stops with a planned-feature message because multiple structural `mu` layers need their own identifiability checks. |
| `relmat(1 | id, K = K)` or `relmat(1 | id, Q = Q)` | Implemented first slice | A lower-level Gaussian `mu` random intercept for user-supplied latent relatedness covariance or precision matrices. |
| `relmat(1 + x | id, K = K)` or `relmat(1 + x | id, Q = Q)` | Implemented first one-slope slice | One trait has independent known-matrix intercept and numeric slope fields in the location predictor. The two fields have separate SDs and do not estimate an intercept-slope correlation. |
| Matching labelled `relmat(1 | p | id, K = K)` or `relmat(1 | p | id, Q = Q)` terms in bivariate `mu1` and `mu2` | Implemented first q=2 slice | Two Gaussian responses have correlated lower-level known-relatedness location deviations from the same matrix. `corpairs(level = "relmat")` reports this mean-mean row separately from residual `rho12`. |
| `spatial(1 | site, mesh = mesh)` | Planned | Mesh inputs will provide expert control for the scalable SPDE/GMRF spatial path. |

The fitted q=2 phylogenetic correlation-regression syntax is:

```r
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "mu2") ~ ecology
```

The predictor-dependent q=4 extensions remain planned:

```r
# Planned q=4 extensions:
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma1") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma2") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma1") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma2") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "sigma1", to = "sigma2") ~ ecology
```

Those formulae are not examples to run today. They mark the intended next layer
after constant phylogenetic `corpairs()` rows: location-location, all four
location-scale pairs, and scale-scale correlation regression. The
ordinary grouped `corpair()` implementation cannot simply be copied here,
because phylogenetic effects share one tree-coupled covariance matrix rather
than independent per-species 2 by 2 blocks.

The first design contract for the location-location row uses two independent
unit phylogenetic fields. For species `l`, the ecological predictor would define
`rho_l = tanh(w_l^T alpha)`, then
`c_l = sqrt((1 + rho_l) / 2)` and `d_l = sqrt((1 - rho_l) / 2)`. The two
location effects would be

```text
a1_l = tau1 (c_l z1_l + d_l z2_l)
a2_l = tau2 (c_l z1_l - d_l z2_l).
```

This q=2 contract is for the location-location row only: `mu1`-`mu2`.
Location-scale rows such as `mu1`-`sigma2` and scale-scale rows such as
`sigma1`-`sigma2` remain deferred until the q=4 version of this
positive-definite construction is designed. This q=2 contract keeps the full
covariance matrix positive definite and reduces to the current constant
bivariate phylogenetic covariance when all `rho_l` are equal. When `rho_l`
varies by species, the model is nonstationary: it changes the tree-coupled
latent covariance surface, not just a residual correlation row.

## A model ladder for two traits

The safest way to read the bivariate syntax is to build the model in layers.
Each layer answers a different biological or statistical question.

### 1. Residual correlation only

Start here when the question is whether two measured traits remain correlated
within the same observation after fixed effects:

```r
fit_resid <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate,
    mu2 = desiccation_tolerance ~ climate,
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

rho12(fit_resid)
```

This model has no phylogenetic random effects. The `rho12()` output is the
residual within-observation correlation between heat tolerance and desiccation
tolerance.

### 2. Add phylogenetic mean-mean correlation

Add matching `phylo()` terms to `mu1` and `mu2` when the question is whether
related species share deviations in the two trait means:

```r
fit_phylo_mean <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | p | species, tree = tree),
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

corpairs(fit_phylo_mean, level = "phylogenetic")
rho12(fit_phylo_mean)
```

This model estimates one latent phylogenetic mean-mean correlation and one
residual `rho12` surface. `corpairs()` and `rho12()` answer different
questions.

The coordinate-spatial sibling uses the same bivariate location layer but
replaces the tree with a coordinate-derived precision:

```r
fit_spatial_mean <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate +
      spatial(1 | p | site, coords = coords),
    mu2 = desiccation_tolerance ~ climate +
      spatial(1 | p | site, coords = coords),
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

corpairs(fit_spatial_mean, level = "spatial")
rho12(fit_spatial_mean)
```

Use this model when nearby sites may share paired deviations in the two response
means. Use the q=4 spatial route only when the spatial question also involves
constant location-scale covariance across `mu1`, `mu2`, `sigma1`, and `sigma2`.
For one-response models, `sigma ~ spatial(1 | site, coords = coords)` fits an
intercept-only residual-scale spatial field, and
`sigma ~ spatial(1 + x | site, coords = coords)` has point-fit/extractor
evidence for the exact q1 one-slope route. Its interval gate remains blocked.
Predictor-dependent spatial `corpair()` regression, multiple or labelled
spatial sigma slopes, and bivariate spatial slope blocks remain planned. Treat
the displayed spatial q2 correlation as a point estimate: interval calibration
for this intercept-only row is not yet validated.

### 3. Let species-level predictors change phylogenetic mean SDs

Use the Box 1 direct-SD family when the question is whether species-level
predictors explain the magnitude of phylogenetic mean variation:

```r
fit_phylo_sd <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate + phylo(1 | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | species, tree = tree),
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context,
    sd_phylo1(species) ~ habitat_variability_species,
    sd_phylo2(species) ~ habitat_variability_species
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

coef(fit_phylo_sd, "sd_phylo1(species)")
coef(fit_phylo_sd, "sd_phylo2(species)")
corpairs(fit_phylo_sd, level = "phylogenetic")
```

This is not a residual scale model. `sd_phylo1()` and `sd_phylo2()` model the
phylogenetic location SD surfaces for the two responses. The cross-response
phylogenetic mean-mean correlation remains constant and is displayed as a point
estimate; this direct-SD route does not have validated interval evidence.

### 4. Move to a q=4 phylogenetic location-scale block

Use the q=4 block when the question is whether phylogenetic deviations in
means and residual log-SDs move together:

```r
fit_phylo_q4 <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | p | species, tree = tree),
    sigma1 = ~ habitat_variability +
      phylo(1 | p | species, tree = tree),
    sigma2 = ~ habitat_variability +
      phylo(1 | p | species, tree = tree),
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

corpairs(fit_phylo_q4, level = "phylogenetic")
rho12(fit_phylo_q4)
```

This model estimates six latent phylogenetic correlations: one
location-location, four location-scale, and one scale-scale. These are constant
correlations in the phylogenetic random-effect covariance block. The four
location-scale rows are `mu1`-`sigma1`, `mu1`-`sigma2`, `mu2`-`sigma1`, and
`mu2`-`sigma2`; the two cross-trait rows can be biologically harder to explain,
but they are statistically part of the same q=4 block. They are not the
residual `rho12` surface.

The six q=4 correlations are point estimates. Direct or fix-and-refit profile
intervals for these derived rows are not yet implemented and validated.

Read the six q=4 rows as latent phylogenetic covariance summaries:

| Pair | Biological reading | Common mistake |
|---|---|---|
| `mu1`-`mu2` | species with positive phylogenetic deviations in trait 1 also tend to have positive or negative deviations in trait 2 | treating it as residual trait coupling |
| `mu1`-`sigma1` | the evolutionary deviation in trait 1's mean is associated with trait 1's residual log-SD | reporting it as a fixed `sigma1` slope |
| `mu1`-`sigma2` | the evolutionary deviation in trait 1's mean is associated with trait 2's residual log-SD | dropping the cross-trait mean-scale row from the interpretation |
| `mu2`-`sigma1` | the evolutionary deviation in trait 2's mean is associated with trait 1's residual log-SD | dropping the cross-trait mean-scale row from the interpretation |
| `mu2`-`sigma2` | the evolutionary deviation in trait 2's mean is associated with trait 2's residual log-SD | treating it as the only mean-scale pair |
| `sigma1`-`sigma2` | species-level residual-scale deviations co-vary across the two traits | treating it as residual `rho12` |

The cross-trait rows are not optional bookkeeping. They are part of the fitted
unstructured q=4 covariance block and should remain visible in examples even
when the biological story is harder to tell.

### 5. Predictor-dependent latent phylogenetic correlations

The q=2 location-location route is fitted:

```r
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "mu2") ~ ecology
```

The q=4 endpoint-specific extensions are still planned:

```r
# Planned q=4 extensions:
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma1") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma2") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma1") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma2") ~ ecology
corpair(species, level = "phylogenetic", block = "p",
        from = "sigma1", to = "sigma2") ~ ecology
```

The fitted q=2 layer makes the phylogenetic location-location random-effect
correlation vary with a species-level ecological predictor, analogous in
spirit to
`rho12 = ~ assay_context` for residual correlation. The implementation is not
identical to residual `rho12` regression: the phylogenetic likelihood must still
build one positive-definite covariance matrix for all species tied together by
the tree. The selected q=2 design uses a two-field loading construction, so the
modelled `rho_l` is the same-species phylogenetic correlation while
between-species covariances remain tied to the tree and become nonstationary
when `rho_l` varies. This first implementation target is only the
location-location row; location-scale and scale-scale correlation regressions
are q=4 extensions and remain deferred. The explicit `level = "phylogenetic"`
keeps this layer separate
from ordinary species effects and future spatial effects. Endpoint-specific
`from` / `to` syntax avoids treating all four location-scale pairs as if they
had the same biological meaning. Use `corpairs()` for fitted constant or
modelled latent phylogenetic correlations and `rho12()` for fitted residual
correlations. The ordinary group-level q=2 analogue is:
`corpair(id, level = "group", block = "p", from = "mu1", to = "mu2") ~ w`
models the correlation between two labelled ordinary location random
intercepts.

## What should get 95% intervals?

There are two different kinds of "random-effect" uncertainty.

The first kind is uncertainty in the covariance parameters: random-effect SDs
and correlations. These are the quantities shown by `sdpars`, `corpars`,
`corpairs()`, and `summary(fit)$covariance`. The admitted interval method is
row-specific rather than a blanket preference for profile likelihood. The q1
`mu` anchors and the exact phylo/relmat slope-only q2 `mu1:x`/`mu2:x` SD rows
use the default location-axis bias-corrected, small-sample-t Wald channel; q1
`sigma` uses raw uncorrected log-SD Wald-z evidence and treats profile as
diagnostic-only at `g = 8`; the Arc 1a REML cells use direct structured-SD
profiles only in their tested discrete domains. Check the
[capability-and-limits guide](capability-and-limits.html#choosing-an-interval-method)
for the evidence channel attached to a particular row.

The intercept-only q2 example on this page is not one of those admitted
slope-only rows. Its default Wald output is diagnostic rather than a validated
reporting interval:

```r
confint(fit_phylo_mean, parm = "variance_components")
```

The direct Wald route uses the fitted log-SD scale for SD targets and the
guarded atanh correlation-link scale for direct correlation targets before
returning intervals on the reported scale. For this intercept-only q2 row,
neither a finite Wald result nor an available target promotes the row beyond
diagnostic-only. A targeted profile can diagnose boundary geometry when
`profile_targets(fit)` lists the target, but it does not replace a validated
channel or promote a row whose interval evidence is diagnostic-only.

For a diagnostic look at likelihood geometry, use `profile_targets()` to find
the target name and `confint(..., method = "profile")` to request a diagnostic
profile. The intercept-only bivariate q2 example below is diagnostic-only under
both Wald and profile channels:

```r
profile_targets(fit_phylo_mean, ready_only = TRUE)

phylo_cor <- "cor:phylo:cor(mu1:(Intercept),mu2:(Intercept) | phylo | species)"
confint(
  fit_phylo_mean,
  parm = phylo_cor,
  method = "profile",
  profile_precision = "fast"
)
```

Diagnostic profile rows include `conf.status = "profile"`. They also include
`profile.boundary` and `profile.message`, so an interval that reaches a lower
SD boundary or approaches a correlation boundary is marked. Do not report
either interval above as validated for this intercept-only q2 fit. The admitted
q2 reporting channel belongs only to the exact phylo/relmat slope-only SD rows
named in the capability guide. `profile_precision = "fast"` is a quicker
diagnostic pass that uses coarser `TMB::tmbprofile()` controls; use default
precision only when a more detailed geometry diagnosis is needed.

The second kind is uncertainty in conditional random-effect modes, such as the
species-specific fitted deviations. Those intervals are useful for caterpillar
plots and diagnostics, but they are not the same as intervals for
phylogenetic SDs or correlations. The current priority is covariance-parameter
intervals first, then conditional-mode uncertainty once the covariance
interfaces are stable.

For q=4 ordinary or phylogenetic blocks, `profile_targets()` lists the six
correlations as derived unstructured-correlation targets. They are reported as
point estimates today, but direct profile intervals for all six rows are not
claimed yet. The first ordinary and phylogenetic `corpair() ~ w` implementations
are narrower: they fit only the q=2 `mu1`-`mu2` location-location pair for
matching labelled ordinary or phylogenetic random intercepts. Those direct
correlation-regression paths expose link-scale fixed-effect coefficients through
`coef()`, `summary()`, `vcov()`, and `profile_targets()`, while `corpairs()`
reports the response-scale mean, minimum, maximum, and number of group-level or
species-level correlation values. The following call is a geometry diagnostic
at a chosen group-level predictor value; profile availability is not validated
interval evidence for this row:

```r
confint(
  fit_ord_corpair,
  parm = 'corpair(id, level = "group", block = "p", from = "mu1", to = "mu2")',
  method = "profile",
  newdata = data.frame(w = 0)
)
```

The animal q=2 path uses the same fitted-versus-planned distinction. Constant
`corpairs(level = "animal")` rows are fitted for matching labelled
`animal(1 | p | individual, pedigree = pedigree)`, `A`, or `Ainv` terms, but
predictor-dependent animal `corpair()` regression and sparse large-pedigree
precision construction remain planned.

The phylogenetic q=2 path exposes the same diagnostic profiling interface, but
the `parm` names the phylogenetic level:

```r
confint(
  fit_phylo_corpair,
  parm = paste0(
    'corpair(species, level = "phylogenetic", block = "p", ',
    'from = "mu1", to = "mu2")'
  ),
  method = "profile",
  newdata = data.frame(ecology = 0)
)
```

`corpairs(conf.int = TRUE)` still reports `newdata_required` for the modelled
summary row because that row averages and ranges over group-level or
species-level correlation values. Neither predictor-dependent q2 example above
has coverage-backed interval validation, so do not use those diagnostic
profiles as reporting intervals. q=4 rows still need a fix-and-refit or
reparameterized profile before 95% intervals are claimed for all six derived
correlations.

## Mathematical contract

The general structured-effect bridge is:

\[
\eta_d =
  \mathbf{X}_d \boldsymbol{\beta}_d +
  \mathbf{Z}_d \mathbf{z},
\qquad
\mathbf{z} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_z^2 \mathbf{K}).
\]

Here \(d\) is the distributional parameter being modelled. For the simplest
fitted examples, \(d = \mu\): the structured effect is added to the mean
predictor. Current Gaussian first slices also include intercept-only structured
`sigma` terms, q1 `sigma` one-slope routes for all four providers, and selected
bivariate covariance blocks. The phylo, A-matrix animal, and K/Q relmat sigma
slopes are inference-ready with caveats; the spatial sigma-slope interval gate,
multiple or labelled structured sigma slopes, and structured `rho12` remain
planned. For phylogenetic models,
\(\mathbf{K}\) is a tree-derived phylogenetic covariance matrix
\(\mathbf{A}\). For spatial models, \(\mathbf{K}\) will be a distance-derived
matrix or an SPDE/GMRF precision.

For the implemented univariate Gaussian phylogenetic location model:

\[
y_i \mid \mu_i, \sigma_i
  \sim \operatorname{Normal}(\mu_i, \sigma_i^2),
\]

\[
\mu_i =
  \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu}
  + a_{species[i]},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma},
\]

\[
\mathbf{a} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2\mathbf{A}).
\]

The phylogenetic SD \(\sigma_{phylo}\) measures how strongly related species
remain similar in their fitted mean after the fixed effects have been included.
The residual `sigma` remains the within-observation residual SD. These are
different scale parameters.

The matching implemented R syntax is:

```r
drmTMB(
  drm_formula(
    y ~ x1 + phylo(1 | species, tree = tree),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

## Worked example: thermal tolerance across species

Suppose we measure thermal tolerance for multiple observations per species and
ask whether body size predicts thermal tolerance after accounting for shared
ancestry. In a real analysis, `tree` would usually come from a phylogenetic
data workflow. For this vignette, the setup code creates a small ultrametric
branch-length tree so the example stays self-contained.

```{r}
set.seed(31)
tree <- balanced_ultrametric_tree(16)

species_info <- data.frame(
  species = tree$tip.label,
  body_size = scale(
    rnorm(16, mean = rep(c(-0.4, 0.4), each = 8), sd = 0.35)
  )[, 1]
)

phylo_shift <- rep(c(-0.5, 0.5), each = 8) +
  rep(c(-0.18, 0.18), each = 4, length.out = 16)

n_each <- 5
dat <- species_info[rep(seq_len(nrow(species_info)), each = n_each), ]
dat$body_size <- dat$body_size + rnorm(nrow(dat), sd = 0.08)
dat$thermal_tolerance <- 1.2 + 0.45 * dat$body_size +
  phylo_shift[match(dat$species, species_info$species)] +
  rnorm(nrow(dat), sd = 0.25)

head(dat)
```

The fitted model follows the equation above:

```{r}
fit_phylo <- drmTMB(
  drm_formula(
    thermal_tolerance ~ body_size + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat
)

summary(fit_phylo)
```

How to read this output:

The `mu:body_size` row is the fixed-effect slope for body size in the mean
thermal tolerance model. The `sigma:(Intercept)` row is the log residual SD, so
`exp(coef(fit_phylo, "sigma"))` gives the within-observation residual SD. The
`phylo(1 | species)` random-effect SD is \(\sigma_{phylo}\), the phylogenetic
among-species SD in the mean after body size has been included.

```{r}
phylo_sd <- unname(fit_phylo$sdpars$mu["phylo(1 | species)"])
residual_sd <- exp(coef(fit_phylo, "sigma")["(Intercept)"])
c(phylogenetic_sd = phylo_sd, residual_sd = residual_sd)
```

In this run, related species remain similar in thermal tolerance after
accounting for body size. The fitted phylogenetic SD is larger than the
within-observation residual SD, so shared ancestry explains a visible part of
the remaining mean-level structure.

`check_drm()` reports whether the optimizer, Hessian, scale, dropped-row, and
phylogenetic replication checks passed:

```{r}
check_drm(fit_phylo)
```

## Practical checks for phylogenetic fits

The public `phylo()` term requires an ultrametric tree with branch lengths.
Users pass the tree; `drmTMB` builds the sparse phylogenetic precision
internally. You do not need to construct the phylogenetic covariance matrix or
A-inverse yourself.

If a phylogenetic fit fails before optimization, check these points first:

- the tree object has class `phylo`;
- every observed species in `data$species` appears in `tree$tip.label`;
- species names use the same spelling and underscore/space convention in the
  data and tree;
- the tree has finite positive branch lengths and is ultrametric;
- the model uses an implemented syntax: Gaussian `phylo()` terms in the
  documented univariate or matching bivariate formulas, ordinary Poisson/NB2
  q1 phylogenetic `mu` intercept-plus-one-slope terms, recovery-grade NB2 q1
  phylogenetic `sigma`, Student-t q1 phylogenetic `nu`, or cumulative-logit q1
  phylogenetic `mu` gates recorded in the live ledger.

Multiple or labelled phylogenetic slopes, phylogenetic slope correlations,
non-Gaussian phylogenetic effects outside those exact gates, and structured
effects in `rho12` remain planned.
Partial, unlabelled, mismatched, and slope q=4 phylogenetic location-scale
forms should error clearly rather than silently fitting a different model.

## Bivariate phylogenetic correlations: model layer by layer

For two Gaussian responses, `drmTMB` separates the residual correlation from
the phylogenetic random-effect correlation. Let
\(\mathbf{y}_i = (y_{1i}, y_{2i})^\top\). A bivariate phylogenetic location
model can be written as:

\[
\mathbf{y}_i
  \mid \boldsymbol{\mu}_i, \boldsymbol{\Omega}_i
  \sim \operatorname{MVN}(\boldsymbol{\mu}_i, \boldsymbol{\Omega}_i),
\]

\[
\begin{aligned}
\mu_{1i} &=
  \mathbf{x}_{1i}^{\top}\boldsymbol{\beta}_1 + a_{1,s[i]},\\
\mu_{2i} &=
  \mathbf{x}_{2i}^{\top}\boldsymbol{\beta}_2 + a_{2,s[i]}.
\end{aligned}
\]

\[
\begin{aligned}
\Omega_{11,i} &= \sigma_{1i}^2,\\
\Omega_{22,i} &= \sigma_{2i}^2,\\
\Omega_{12,i} &= \Omega_{21,i}
  = \rho_{12,i}\sigma_{1i}\sigma_{2i}.
\end{aligned}
\]

The residual coscale formula, `rho12 = ~ assay_context`, models
\(\rho_{12,i}\) in \(\boldsymbol{\Omega}_i\). This is an observation-level
residual correlation.

The phylogenetic location effects are a separate latent layer. Let
\(\mathbf{a}_1\) and \(\mathbf{a}_2\) collect the species-level phylogenetic
location effects for responses 1 and 2. If \(\mathbf{A}\) is the tree-derived
species correlation matrix, then:

\[
\begin{aligned}
\operatorname{Var}(\mathbf{a}_1) &=
  \tau_1^2 \mathbf{A},\\
\operatorname{Var}(\mathbf{a}_2) &=
  \tau_2^2 \mathbf{A},\\
\operatorname{Cov}(\mathbf{a}_1, \mathbf{a}_2) &=
  r_{\mu_1,\mu_2}^{\mathrm{phylo}}\tau_1\tau_2\mathbf{A}.
\end{aligned}
\]

Here \(r_{\mu_1,\mu_2}^{\mathrm{phylo}}\) is the phylogenetic mean-mean
correlation. It is reported by `corpairs(fit, level = "phylogenetic")`, not by
`rho12()`.

The matching R syntax is:

```r
fit_biv_phylo <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | p | species, tree = tree),
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

corpairs(fit_biv_phylo, level = "phylogenetic")
rho12(fit_biv_phylo)
```

In that output, `corpairs()` gives the constant phylogenetic correlation point
estimate among
species-level mean deviations. `rho12()` gives the fitted residual correlation
for each observation, which can vary with `assay_context`. The intercept-only
q2 row does not have validated interval evidence; a callable profile is a
diagnostic, not an inference-ready claim.

This is the key implementation boundary. `drmTMB` can estimate latent
phylogenetic correlations when the matching `phylo()` terms define a shared
covariance block. It can also fit the first predictor-dependent latent
phylogenetic correlation formula for the q=2 location-location row. In other
words, `rho12 = ~ assay_context` is implemented for residual correlation
regression, and
`corpair(species, level = "phylogenetic", block = "p", from = "mu1", to = "mu2") ~ ecology`
is implemented for species-level predictors beside matching labelled `mu1` and
`mu2` `phylo()` terms. The q=4 location-scale and scale-scale correlation
regressions remain planned.

The q=4 phylogenetic location-scale block extends the same idea to four
endpoints. For endpoint \(d \in \{\mu_1, \mu_2, \sigma_1, \sigma_2\}\), let
\(u_{d,s}\) be the phylogenetic effect for species \(s\). The covariance
contract is:

\[
\operatorname{Cov}(u_{d,s}, u_{e,t}) =
\Sigma_{de}^{\mathrm{phylo}} A_{st},
\qquad
d,e \in \{\mu_1, \mu_2, \sigma_1, \sigma_2\}.
\]

The four linear predictors then contain one endpoint-specific phylogenetic
effect:

\[
\mu_{1i} = \mathbf{x}_{1i}^{\top}\boldsymbol{\beta}_1
  + u_{\mu_1,s[i]},
\qquad
\mu_{2i} = \mathbf{x}_{2i}^{\top}\boldsymbol{\beta}_2
  + u_{\mu_2,s[i]},
\]

\[
\log\sigma_{1i} = \mathbf{z}_{1i}^{\top}\boldsymbol{\gamma}_1
  + u_{\sigma_1,s[i]},
\qquad
\log\sigma_{2i} = \mathbf{z}_{2i}^{\top}\boldsymbol{\gamma}_2
  + u_{\sigma_2,s[i]}.
\]

The fitted \(\boldsymbol{\Sigma}^{phylo}\) is written as
\(\mathbf{D}\mathbf{R}\mathbf{D}\), where \(\mathbf{D}\) contains four
phylogenetic SDs and \(\mathbf{R}\) contains six latent correlations. Those six
correlations are the `corpairs(fit, level = "phylogenetic")` rows: one
location-location row, four location-scale rows, and one scale-scale row. The
four location-scale rows are \(u_{\mu_1}\)-\(u_{\sigma_1}\),
\(u_{\mu_1}\)-\(u_{\sigma_2}\), \(u_{\mu_2}\)-\(u_{\sigma_1}\), and
\(u_{\mu_2}\)-\(u_{\sigma_2}\).

## Worked example: two traits with phylogenetic location-scale covariance

Now suppose the scientific question is about two tolerance traits measured
repeatedly for the same species. We want climate to explain each trait mean,
habitat variability to explain each residual SD, and assay context to explain
the residual within-observation correlation. We also want the species-level
phylogenetic deviations in both means and both log-SDs to share one latent q=4
covariance block.

The simulated data below are deliberately small enough for the article to run.
The hidden setup creates the tree and species-level phylogenetic effects; the
fitted model uses only public `drmTMB()` syntax.

```{r}
sim_q4 <- simulate_phylo_q4_traits()
tree <- sim_q4$tree
tol_dat <- sim_q4$data

head(tol_dat)
```

The model uses the same labelled `phylo()` term in all four endpoints:

```{r}
fit_phylo_q4 <- suppressWarnings(
  drmTMB(
    drm_formula(
      mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
      mu2 = desiccation_tolerance ~ climate +
        phylo(1 | p | species, tree = tree),
      sigma1 = ~ habitat_variability +
        phylo(1 | p | species, tree = tree),
      sigma2 = ~ habitat_variability +
        phylo(1 | p | species, tree = tree),
      rho12 = ~ assay_context
    ),
    family = c(gaussian(), gaussian()),
    data = tol_dat,
    control = list(eval.max = 1000, iter.max = 1000)
  )
)

summary(fit_phylo_q4)
```

The fixed-effect rows answer the distributional-regression questions. The
`mu1:climate` and `mu2:climate` rows are mean effects. The
`sigma1:habitat_variability` and `sigma2:habitat_variability` rows are effects
on log residual SD. The `rho12:assay_context` row is the residual coscale effect:
it changes the residual correlation between the two traits within the same
observation.

The phylogenetic q=4 block is a different layer. Use `corpairs()` to read the
latent species-level correlations:

```{r}
corpairs(fit_phylo_q4, level = "phylogenetic")
```

Those six rows are not residual correlations. The `mean-mean` row asks whether
species that are phylogenetically high for heat tolerance also tend to be high
for desiccation tolerance. The four `mean-scale` rows ask whether evolutionary
deviations in means are associated with evolutionary deviations in variability:
same-trait pairs (`mu1`-`sigma1`, `mu2`-`sigma2`) and cross-trait pairs
(`mu1`-`sigma2`, `mu2`-`sigma1`). The cross-trait rows are statistically valid
q=4 covariance rows, but their biological interpretation should be written
with extra care.

The `scale-scale` row asks whether the two traits' phylogenetic log-SD
deviations move together. These q=4 rows are point estimates and remain derived
targets until a direct or fix-and-refit profile is implemented and validated.

For a compact scalar view of the residual layer, use `rho12()`:

```{r}
range(rho12(fit_phylo_q4))
```

`check_drm()` keeps these layers separate. In small examples, q=4 models can
warn about optimizer convergence even when gradients, Hessian status, and q=4
covariance diagnostics are acceptable. Treat this article fit as an output
map; for scientific inference, inspect the full diagnostic table and use larger
datasets or simpler covariance structure when needed.

```{r}
q4_check <- check_drm(fit_phylo_q4)
q4_check[q4_check$check %in% c(
  "optimizer_convergence",
  "fixed_gradient",
  "hessian_positive_definite",
  "biv_phylo_q4_covariance"
), ]
```

The fitted q=4 correlations are point summaries. They are listed in
`profile_targets()` as derived `theta_phylo` targets, not as direct
profile-ready intervals:

```{r}
subset(
  profile_targets(fit_phylo_q4),
  startsWith(parm, "cor:phylo:"),
  select = c(parm, target_type, profile_ready, profile_note)
)
```

The q=2 predictor-dependent latent phylogenetic correlation is fitted for the
location-location row. The extractor is `corpairs()`, and the formula syntax is
singular:

```r
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "mu2") ~ w
```

The predictor-dependent q=4 extensions remain planned:

```r
# Planned q=4 extensions:
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma1") ~ w
corpair(species, level = "phylogenetic", block = "p",
        from = "mu1", to = "sigma2") ~ w
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma1") ~ w
corpair(species, level = "phylogenetic", block = "p",
        from = "mu2", to = "sigma2") ~ w
corpair(species, level = "phylogenetic", block = "p",
        from = "sigma1", to = "sigma2") ~ w
```

This endpoint naming matters because the cross-trait mean-scale pairs can be
biologically harder to interpret than same-trait pairs. Use
`rho12 = ~ assay_context` for residual correlation predictors. Use
`corpairs(fit, level = "phylogenetic")` to extract fitted constant or modelled
latent phylogenetic correlations.

## Implemented third route: coordinate spatial intercepts and one slope

Spatial fields are the third route in the structural-dependence ladder. They
are deliberately shaped to look parallel to phylogenetic effects, but they use
site coordinates rather than a tree. The first fitted paths are
coordinate-based univariate Gaussian `mu` coefficient fields: a random
intercept, and one independent random slope for a numeric predictor. The
example below keeps the data deliberately small so the models can be rebuilt
inside the article:

```{r}
spatial_example <- simulate_spatial_location_data()
spatial_dat <- spatial_example$data
coords <- spatial_example$coords

fit_spatial <- drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 | site, coords = coords),
    sigma ~ depth
  ),
  family = gaussian(),
  data = spatial_dat
)
```

The fitted spatial SD is a structured location-scale component, not the
residual `sigma`. The conditional spatial effects are site-level deviations in
the location predictor, after the fixed effects and residual scale have been
estimated:

```{r}
fit_spatial$sdpars$mu
head(ranef(fit_spatial, "spatial_mu")$terms[[1]])
profile_targets(fit_spatial, ready_only = TRUE)[
  grep("spatial", profile_targets(fit_spatial, ready_only = TRUE)$parm),
  c("parm", "estimate", "profile_ready")
]
```

When the biological question is whether a local environmental gradient has a
different slope in different parts of the sampled landscape, add one numeric
coefficient inside the `spatial()` term:

```{r}
fit_spatial_slope <- drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 + depth | site, coords = coords),
    sigma ~ depth
  ),
  family = gaussian(),
  data = spatial_dat
)

fit_spatial_slope$sdpars$mu
names(ranef(fit_spatial_slope, "spatial_mu")$terms)
```

The two SDs answer different biological questions. `spatial(1 | site)` is the
amount of smooth site-to-site variation in the expected response after the
fixed effects. `spatial(0 + depth | site)` is the amount of smooth site-to-site
variation in the depth effect. Large slope SD means the fitted depth-response
gradient changes across the coordinate surface; it is not a residual
correlation and it is not the residual scale `sigma`.

Before interpreting the spatial SD or mapped site effects, inspect the spatial
diagnostic row. It is deliberately named `spatial_mu_diagnostics`, not
`phylo_mu_replication`, because the fitted layer is a coordinate site field:

```{r}
spatial_check <- check_drm(fit_spatial)
spatial_check[spatial_check$check %in% c(
  "optimizer_convergence",
  "fixed_gradient",
  "hessian_positive_definite",
  "spatial_mu_diagnostics"
), ]
```

The spatial row reports the number of fitted sites, the minimum number of
observations per site, the coordinate range used by the fixed exponential
covariance, the fitted spatial SD, and the ratio between the spatial SD and the
residual scale. A `note` is not a failed model. It tells you to be cautious
when a site is singly observed or when the spatial field is tiny relative to
unstructured residual variation.

The fitted `coords` path accepts either one row per site or one row per
observation, provided coordinates are constant within each site. Internally,
the coordinate path builds a fixed exponential covariance from pairwise site
distances, inverts it to a precision matrix, and estimates one spatial SD for
each fitted spatial coefficient. The intercept model estimates one field. The
one-slope model
`spatial(1 + depth | site, coords = coords)` estimates two independent fields:
`spatial(1 | site)` for site intercept deviations and
`spatial(0 + depth | site)` for site-specific depth slopes. These two fields
share the same coordinate precision but have separate SDs and no
intercept-slope `corpair()` row. This is a small-data foundation and recovery
target, not the scalable mesh/SPDE implementation. The conditional spatial
effects from `ranef()` are shrunken site deviations in the location predictor
after the fixed effects and the spatial covariance have been estimated.

The planned `mesh` argument plays a different role. A `mesh` is the
computational scaffold for an SPDE/GMRF field: a triangular network of vertices
where the latent spatial field is represented and then projected back to
observations. Users should not interpret mesh vertices as biological sampling
units; they are numerical support points for the spatial surface. In other
words, a mesh is not required by the idea of spatial dependence itself. It is
required by the scalable SPDE/GMRF approximation. The friendly API should let
users start with `coords = coords`; `mesh = mesh` is the expert route for
reproducible control over boundaries, barriers, coastlines, or uneven sampling.

The spatial term is deliberately parallel to the phylogenetic term:

```r
# implemented for Gaussian mu/sigma intercepts and matching bivariate blocks
phylo(1 | species, tree = tree)
# implemented for one numeric Gaussian mu slope
phylo(1 + x | species, tree = tree)
# implemented for Gaussian mu/sigma intercepts and matching bivariate blocks
spatial(1 | site, coords = coords)
# implemented for one numeric Gaussian mu slope
spatial(1 + x | site, coords = coords)
# planned
spatial(1 | site, mesh = mesh)
```

The mathematical role is the same: add structured Gaussian coefficient fields
to a distributional predictor. The computational route differs. `phylo()`
starts from an ultrametric tree with branch lengths and builds a sparse
tree-derived precision. The first `spatial()` slice starts from coordinates and
builds a fixed coordinate precision; the future mesh path will build an
SPDE/GMRF precision.

The computational plan is to keep using sparse A-inverse ideas for phylogeny
and add SPDE/GMRF structure for spatial fields. The spatial route should cite
the SPDE/GMRF method literature and the software it builds on. The key
methodological reference is Lindgren, Rue, and Lindstrom (2011), who connected
Gaussian fields to sparse GMRFs through the SPDE approach
(`doi:10.1111/j.1467-9868.2011.00777.x`). The ecological TMB precedent is
`sdmTMB`, described in its
[Journal of Statistical Software paper](https://www.jstatsoft.org/article/view/v115i02)
as a package for spatial and spatiotemporal GLMMs using TMB, `fmesher`, latent
GMRFs, and the SPDE approach. If `drmTMB` imports `fmesher` or asks users to
pass `fmesher` mesh objects, the website and paper should also tell users to
cite `fmesher` as software. Any code selectively ported from `gllvmTMB`,
`sdmTMB`, `fmesher`, or INLA-related sources must be recorded in
`inst/COPYRIGHTS` before the spatial feature is called complete.

Structured random slopes are staged more conservatively than ordinary grouped
random slopes. Coordinate spatial and phylogenetic Gaussian `mu` have first
one-slope paths, and q1 structured `sigma` one-slope paths fit for
phylo/spatial/animal/relmat. The phylo, A-matrix animal, and K/Q relmat sigma
slopes are inference-ready with caveats; spatial is point-fit/extractor only.
These paths are deliberately independent: each returns separate SDs and no
intercept-slope `corpair()` row. Multiple or labelled structured slopes,
mesh/SPDE slopes, and structured slope correlations remain later gates.
The combined phylogeny-plus-spatial route should be a separate additive-block
model with a phylogenetic field and a spatial field, not one large cross-factor
covariance model. A more interesting future two-response target is the
slope1-slope2 correlation for the same covariate across traits, the
plasticity-syndrome question described by O'Dea, Noble, and Nakagawa (2021).

## Meta-analysis and future bivariate structure

In meta-analysis, the separation among known sampling covariance,
phylogenetic dependence, and spatial dependence is especially transparent:

\[
\begin{aligned}
e
  &\sim \operatorname{MVN}(\mathbf{0}, V)
  &&\text{known sampling covariance},\\
p_{species}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2 A)
  &&\text{phylogenetic structure},\\
\ell_{location}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{space}^2 M)
  &&\text{spatial structure}.
\end{aligned}
\]

The known matrix `V` handles sampling uncertainty. The phylogenetic and spatial
terms add extra biological or environmental dependence after that known
uncertainty has been included. The same design will let `drmTMB` support
structured Gaussian meta-analysis while keeping ecological and evolutionary
meta-analyses readable.

The first phylogenetic target is the location-scale model. In bivariate
Gaussian models, matching `mu1`/`mu2` phylogenetic location terms estimate the
mean-mean phylogenetic correlation. Matching labelled `phylo()` terms across
`mu1`, `mu2`, `sigma1`, and `sigma2` now fit the first constant q=4
phylogenetic location-scale block: one location-location row, four
location-scale rows, and one scale-scale row in `corpairs(level =
"phylogenetic")`. The six q=4 correlations are latent phylogenetic
random-effect correlations, not residual `rho12`.

The first fitted bivariate phylogenetic location-coscale syntax looks like
this:

```r
fit_biv_phylo <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle + phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle + phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

This syntax fits correlated phylogenetic random intercepts in `mu1` and `mu2`.
It estimates two phylogenetic location SDs and one phylogenetic mean-mean
correlation. The `sigma1`, `sigma2`, and residual `rho12` formulas remain
ordinary fixed-effect distributional parameters in this first slice.
Use `phylo(1 | p | species, tree = tree)` in both location formulas when you
want the fitted phylogenetic mean-mean row to use a block label such as `p`.

The Box 1 direct-SD alternative for two responses keeps the phylogenetic
effects in the location formulas, but lets species-level predictors explain the
magnitude of each response's phylogenetic location variation:

```r
fit_biv_sd_phylo <- drmTMB(
  formula = drm_formula(
    mu1 = heat_tolerance ~ climate +
      phylo(1 | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | species, tree = tree),
    sigma1 = ~ habitat_variability,
    sigma2 = ~ habitat_variability,
    rho12 = ~ assay_context,
    sd_phylo1(species) ~ habitat_variability_species,
    sd_phylo2(species) ~ habitat_variability_species
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)
```

Here `sd_phylo1(species)` targets the `mu1` phylogenetic location SD surface and
`sd_phylo2(species)` targets the `mu2` surface. The right-hand-side predictors
must be constant within species after model-row filtering. The latent
phylogenetic mean-mean correlation remains a constant fitted correlation, and
`rho12` remains the within-observation residual correlation:

```r
corpairs(fit_biv_sd_phylo, level = "phylogenetic")
rho12(fit_biv_sd_phylo)
```

The `corpairs()` row is the fitted phylogenetic correlation point estimate
between the two location random effects; this direct-SD route does not have
validated interval evidence. The `rho12()` values are residual correlations between
the two responses after the fitted fixed effects and phylogenetic effects have
been included. Run
`check_drm(fit_biv_sd_phylo)` before interpreting the direct-SD slopes; the
diagnostic reports species replication, each fitted species-SD range, and
whether any fitted direct-SD surface is non-positive or non-finite.

Mathematically, the direct-SD model replaces constant phylogenetic location SDs
with species-level SD surfaces. For response \(k \in \{1,2\}\):

\[
\operatorname{Var}(\mathbf{a}_k)
  = \mathbf{D}_k(\mathbf{z})\,\mathbf{A}\,\mathbf{D}_k(\mathbf{z}),
\qquad
\log D_{k,ss}(\mathbf{z}) =
  \mathbf{z}_s^{\top}\boldsymbol{\alpha}_k.
\]

The current bivariate direct-SD path keeps the cross-response phylogenetic
location correlation constant:

\[
\operatorname{Cov}(\mathbf{a}_1, \mathbf{a}_2)
  = r_{\mu_1,\mu_2}^{phylo}
    \mathbf{D}_1(\mathbf{z})\,\mathbf{A}\,\mathbf{D}_2(\mathbf{z}).
\]

This is why `sd_phylo1(species) ~ habitat_variability_species` and
`sd_phylo2(species) ~ habitat_variability_species` answer a different question
from the q=4 location-scale block. They ask whether species-level predictors
change the magnitude of phylogenetic mean variation for each response. They do
not put random effects into `sigma1` or `sigma2`.

The full constant q=4 location-scale form uses an explicit covariance-block
label:

```r
fit_biv_phylo_q4 <- drmTMB(
  formula = drm_formula(
    mu1 = heat_tolerance ~ climate + phylo(1 | p | species, tree = tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | p | species, tree = tree),
    sigma1 = ~ habitat_variability +
      phylo(1 | p | species, tree = tree),
    sigma2 = ~ habitat_variability +
      phylo(1 | p | species, tree = tree),
    rho12 = ~ assay_context
  ),
  family = c(gaussian(), gaussian()),
  data = tolerance
)

corpairs(fit_biv_phylo_q4, level = "phylogenetic")
rho12(fit_biv_phylo_q4)
```

Partial q=4 syntax, unlabelled all-four syntax, random slopes, and multiple
phylogenetic blocks remain outside this first fitted path. The q=4
`corpairs()` output has six phylogenetic rows: one location-location row,
four location-scale rows, and one scale-scale row. The four location-scale rows
include both same-trait and cross-trait mean-scale pairs. These rows are the
latent phylogenetic correlations among species-level effects. They are not the
residual `rho12` curve, which is why the example shows `corpairs()` and
`rho12()` separately.

For the simpler fitted mean-mean phylogenetic correlation point estimate, use
`corpars$phylo`, `corpairs(fit_biv_phylo, level = "phylogenetic")`, or
`summary(fit_biv_phylo)$covariance`. Run `check_drm(fit_biv_phylo)` before
interpreting that row: the current diagnostic flags phylogenetic correlations
near the boundary and phylogenetic location SDs that are tiny relative to the
matching residual scales. `profile_targets()` can expose a callable diagnostic
profile target, but that does not validate an interval or promote this
intercept-only q2 row. The q=4 correlations are derived from `theta_phylo`, so
direct profile intervals for those six rows are not yet claimed:

```r
phylo_target <- "cor:phylo:cor(mu1:(Intercept),mu2:(Intercept) | phylo | species)"

corpairs(fit_biv_phylo, level = "phylogenetic")
summary(fit_biv_phylo)$covariance
profile_targets(fit_biv_phylo)
```

Do not read a phylogenetic species effect as a residual observation-level
correlation. The `phylo()` layer is species-level structure induced by the
tree; residual `rho12` is within-observation coupling between the two measured
responses. A separate non-phylogenetic species-level block can be useful in a
specialized model, but it is an advanced comparison because it tries to
separate two species-level covariance layers that use the same grouping factor.
That comparison should live in a dedicated identifiability example, not in this
introductory ladder.

The residual `rho12` term remains distinct from the phylogenetic or spatial
correlation matrices. `rho12` is the residual coupling between two responses at
the observation level; `A` and `M` describe dependence among species or places.
In future bivariate structural models, `drmTMB` should also estimate
level-specific correlations such as phylogenetic mean-mean correlations,
non-phylogenetic species correlations, spatial field correlations, and
mean-scale correlations. These are not residual `rho12`; they should be
reported as structural or group-level covariance summaries. For the univariate
structural-dependence ladder, keep the implementation status visible: use
`animal(pedigree/A/Ainv = ...)` for the animal-model first slice, use
`phylo()` when the species tree is the fitted dependence structure, use
`spatial()` when the site field is the fitted dependence structure, plan
combined `phylo()` plus `spatial()` only after joint identifiability checks,
and use `relmat(K/Q = ...)`
for lower-level known-matrix latent random effects.
