---
title: "Structured dependence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)

balanced_ultrametric_tree <- function(n_tip = 16L) {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0("sp_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}
```

Phylogenetic and spatial dependence are structured random effects. The first
implemented path in `drmTMB` is an intercept-only phylogenetic effect in the
univariate Gaussian location formula. Spatial fields, phylogenetic slopes,
phylogenetic scale terms, and bivariate structured covariance blocks remain
planned extensions.

The scientific question for the implemented path is simple:

> Do predictors explain species trait means after accounting for similarity
> among related species?

In the first fitted path, `species` indexes the rows in the data and `tree`
supplies the relatedness among those species. The model estimates a
phylogenetic deviation in the location parameter `mu` for each species.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

## Mathematical contract

The general structured-effect bridge is:

\[
\eta_d =
  \mathbf{X}_d \boldsymbol{\beta}_d +
  \mathbf{Z}_d \mathbf{z},
\qquad
\mathbf{z} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_z^2 \mathbf{K}).
\]

Here \(d\) is the distributional parameter being modelled. In the first
implemented case, \(d = \mu\): the structured effect is added to the mean
predictor. For phylogenetic models, \(\mathbf{K}\) is a tree-derived
phylogenetic covariance matrix \(\mathbf{A}\). For spatial models,
\(\mathbf{K}\) will be a distance-derived matrix or an SPDE/GMRF precision.

For the implemented univariate Gaussian phylogenetic location model:

\[
y_i \mid \mu_i, \sigma_i
  \sim \operatorname{Normal}(\mu_i, \sigma_i^2),
\]

\[
\mu_i =
  \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu}
  + a_{species[i]},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma},
\]

\[
\mathbf{a} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2\mathbf{A}).
\]

The phylogenetic SD \(\sigma_{phylo}\) measures how strongly related species
remain similar in their fitted mean after the fixed effects have been included.
The residual `sigma` remains the within-observation residual SD. These are
different scale parameters.

The matching implemented R syntax is:

```r
drmTMB(
  drm_formula(
    y ~ x1 + phylo(1 | species, tree = tree),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

## Worked example: thermal tolerance across species

Suppose we measure thermal tolerance for multiple observations per species and
ask whether body size predicts thermal tolerance after accounting for shared
ancestry. In a real analysis, `tree` would usually come from a phylogenetic
data workflow. For this vignette, the setup code creates a small ultrametric
branch-length tree so the example stays self-contained.

```{r}
set.seed(31)
tree <- balanced_ultrametric_tree(16)

species_info <- data.frame(
  species = tree$tip.label,
  body_size = scale(
    rnorm(16, mean = rep(c(-0.4, 0.4), each = 8), sd = 0.35)
  )[, 1]
)

phylo_shift <- rep(c(-0.5, 0.5), each = 8) +
  rep(c(-0.18, 0.18), each = 4, length.out = 16)

n_each <- 5
dat <- species_info[rep(seq_len(nrow(species_info)), each = n_each), ]
dat$body_size <- dat$body_size + rnorm(nrow(dat), sd = 0.08)
dat$thermal_tolerance <- 1.2 + 0.45 * dat$body_size +
  phylo_shift[match(dat$species, species_info$species)] +
  rnorm(nrow(dat), sd = 0.25)

head(dat)
```

The fitted model follows the equation above:

```{r}
fit_phylo <- drmTMB(
  drm_formula(
    thermal_tolerance ~ body_size + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat
)

summary(fit_phylo)
```

How to read this output:

The `mu:body_size` row is the fixed-effect slope for body size in the mean
thermal tolerance model. The `sigma:(Intercept)` row is the log residual SD, so
`exp(coef(fit_phylo, "sigma"))` gives the within-observation residual SD. The
`phylo(1 | species)` random-effect SD is \(\sigma_{phylo}\), the phylogenetic
among-species SD in the mean after body size has been included.

```{r}
phylo_sd <- unname(fit_phylo$sdpars$mu["phylo(1 | species)"])
residual_sd <- exp(coef(fit_phylo, "sigma")["(Intercept)"])
c(phylogenetic_sd = phylo_sd, residual_sd = residual_sd)
```

In this run, related species remain similar in thermal tolerance after
accounting for body size. The fitted phylogenetic SD is larger than the
within-observation residual SD, so shared ancestry explains a visible part of
the remaining mean-level structure.

`check_drm()` reports whether the optimizer, Hessian, scale, dropped-row, and
phylogenetic replication checks passed:

```{r}
check_drm(fit_phylo)
```

## Practical checks for phylogenetic fits

The public `phylo()` term requires an ultrametric tree with branch lengths.
Users pass the tree; `drmTMB` builds the sparse phylogenetic precision
internally. You do not need to construct the phylogenetic covariance matrix or
A-inverse yourself.

If a phylogenetic fit fails before optimization, check these points first:

- the tree object has class `phylo`;
- every observed species in `data$species` appears in `tree$tip.label`;
- species names use the same spelling and underscore/space convention in the
  data and tree;
- the tree has finite positive branch lengths and is ultrametric;
- the model uses the implemented syntax
  `phylo(1 | species, tree = tree)` in a univariate Gaussian `mu` formula.

Phylogenetic slopes, phylogenetic terms in `sigma`, and bivariate phylogenetic
covariance blocks are planned. They should error clearly rather than silently
fitting a different model.

Developer scaffolds now name and test the future q=4 phylogenetic endpoint
state across `mu1`, `mu2`, `sigma1`, and `sigma2`, but those scaffolds are not
fitted bivariate phylogenetic models. They do not change the accepted formula
syntax, and they do not add `corpairs()` rows for phylogenetic correlations.

## Planned, not implemented: spatial random intercepts

Spatial fields are deliberately planned to look parallel to phylogenetic
effects, but the spatial likelihood is not implemented yet. The planned syntax
is:

```r
drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 | site, coords = coords),
    sigma ~ temp
  ),
  family = gaussian(),
  data = dat
)
```

The planned `coords` and `mesh` arguments play different roles. `coords`
should name the observation or site coordinates in the modelling data, ideally
in a projected coordinate system such as UTM rather than raw longitude and
latitude. A `mesh` is the computational scaffold for an SPDE/GMRF field: a
triangular network of vertices where the latent spatial field is represented
and then projected back to observations. Users should not interpret mesh
vertices as biological sampling units; they are numerical support points for
the spatial surface.

The spatial term is deliberately parallel to the phylogenetic term:

```r
phylo(1 | species, tree = tree)     # implemented for univariate Gaussian mu
spatial(1 | site, coords = coords)  # planned
spatial(1 | site, mesh = mesh)      # planned
```

The mathematical role is the same: add a structured Gaussian random intercept
to a distributional predictor. The computational route differs. `phylo()`
starts from an ultrametric tree with branch lengths and builds a sparse
tree-derived precision. `spatial()` will start from coordinates or a mesh and
build an SPDE/GMRF precision.

The computational plan is to keep using sparse A-inverse ideas for phylogeny
and add SPDE/GMRF structure for spatial fields, with provenance notes for any
code selectively ported from `gllvmTMB`.

Structured phylogenetic and spatial random slopes will be staged more
conservatively than ordinary grouped random slopes. The first structured-effect
target is intercept-only `mu`; the next should be one structured `mu` slope.
Small sets of structured slopes, interaction slopes, and structured effects in
`sigma` or `rho12` should wait until simulation recovery shows that the simpler
models are identifiable.

## Meta-analysis and future bivariate structure

In meta-analysis, the separation among known sampling covariance,
phylogenetic dependence, and spatial dependence is especially transparent:

\[
\begin{aligned}
e
  &\sim \operatorname{MVN}(\mathbf{0}, V)
  &&\text{known sampling covariance},\\
p_{species}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2 A)
  &&\text{phylogenetic structure},\\
\ell_{location}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{space}^2 M)
  &&\text{spatial structure}.
\end{aligned}
\]

The known matrix `V` handles sampling uncertainty. The phylogenetic and spatial
terms add extra biological or environmental dependence after that known
uncertainty has been included. The same design will let `drmTMB` support
structured Gaussian meta-analysis while keeping ecological and evolutionary
meta-analyses readable.

The first phylogenetic target is the location-scale model: trait means are the
first implemented target, and residual SDs are a planned target once the
location path has enough simulation and comparator coverage. In future
bivariate models this could allow several structured correlations, including
mean-mean, scale-scale, and mean-scale phylogenetic correlations.

Planned bivariate phylogenetic location-coscale syntax looks like this:

```r
drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle + phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle + phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

This syntax is planned, not implemented. It captures the intended extension of
phylogenetic location-scale models: after modelling mean and variance, test
whether the residual association between two traits changes with ecological
predictors such as lifestyle.

Current internal q=4 scaffolds only protect the algebra and endpoint names for
future contributors. Applied users should still fit the implemented univariate
phylogenetic Gaussian model, or a non-phylogenetic bivariate Gaussian model,
until a fitted bivariate phylogenetic likelihood, simulation recovery tests, and
reporting rows are present.

The residual `rho12` term remains distinct from the phylogenetic or spatial
correlation matrices. `rho12` is the residual coupling between two responses at
the observation level; `A` and `M` describe dependence among species or places.
In future bivariate structured models, `drmTMB` should also estimate
level-specific correlations such as phylogenetic mean-mean correlations,
non-phylogenetic species correlations, spatial field correlations, and
mean-scale correlations. These are not residual `rho12`; they should be
reported as structured or group-level covariance summaries.
