---
title: "Structured dependence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)

balanced_ultrametric_tree <- function(n_tip = 16L) {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0("sp_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}
```

Phylogenetic and spatial dependence are structured random effects. The first
implemented `drmTMB` path is an intercept-only phylogenetic effect in the
univariate Gaussian location formula. A first bivariate path also fits matching
intercept-only `phylo()` terms in `mu1` and `mu2`, estimating the two
phylogenetic location SDs and their mean-mean correlation. Spatial fields,
phylogenetic slopes, phylogenetic scale terms, and the full bivariate
phylogenetic location-scale block remain planned extensions.

The scientific question for the implemented path is simple:

> Do predictors explain species trait means after accounting for similarity
> among related species?

In the first fitted path, `species` indexes the rows in the data and `tree`
supplies the relatedness among those species. The model estimates a
phylogenetic deviation in the location parameter `mu` for each species.

Throughout this article, `Normal(a, b)` and `MVN(a, B)` use variance or
covariance as the second argument.

## Current implementation status

This page separates the fitted paths from the roadmap. The fitted phylogenetic
pieces are useful, but they are not yet the full phylogenetic
location-scale-coscale model.

| Model piece | Status | What it means |
| --- | --- | --- |
| `phylo(1 | species, tree = tree)` in univariate `mu` | Implemented | One trait has a phylogenetic random intercept in the location predictor. |
| Matching `phylo(1 | species, tree = tree)` terms in bivariate `mu1` and `mu2` | Implemented first slice | Two traits have correlated phylogenetic location deviations; `corpairs(level = "phylogenetic")` reports the mean-mean row separately from residual `rho12`. |
| Ordinary `(1 | p | id)` terms in all four bivariate `mu1`, `mu2`, `sigma1`, and `sigma2` formulas | Implemented first slice | One ordinary q=4 group block estimates six latent location-location, location-scale, and scale-scale correlations. |
| `phylo()` terms in bivariate `sigma1` and `sigma2` | Planned | This is the next phylogenetic PLSM path: a q=4 phylogenetic block spanning `mu1`, `mu2`, `sigma1`, and `sigma2`. |
| `sd_phylo(species) ~ z` | Planned | This is the separate Box 1 direct-SD family, not a scale formula with a matching random effect. |
| `spatial(1 | site, coords = coords)` or `spatial(1 | site, mesh = mesh)` | Planned | Spatial will mirror the phylogenetic API after the phylogenetic structured-effect path is stable. |

## Mathematical contract

The general structured-effect bridge is:

\[
\eta_d =
  \mathbf{X}_d \boldsymbol{\beta}_d +
  \mathbf{Z}_d \mathbf{z},
\qquad
\mathbf{z} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_z^2 \mathbf{K}).
\]

Here \(d\) is the distributional parameter being modelled. In the first
implemented case, \(d = \mu\): the structured effect is added to the mean
predictor. For phylogenetic models, \(\mathbf{K}\) is a tree-derived
phylogenetic covariance matrix \(\mathbf{A}\). For spatial models,
\(\mathbf{K}\) will be a distance-derived matrix or an SPDE/GMRF precision.

For the implemented univariate Gaussian phylogenetic location model:

\[
y_i \mid \mu_i, \sigma_i
  \sim \operatorname{Normal}(\mu_i, \sigma_i^2),
\]

\[
\mu_i =
  \mathbf{x}_{\mu i}^{\top}\boldsymbol{\beta}_{\mu}
  + a_{species[i]},
\qquad
\log(\sigma_i) =
  \mathbf{x}_{\sigma i}^{\top}\boldsymbol{\beta}_{\sigma},
\]

\[
\mathbf{a} \sim
  \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2\mathbf{A}).
\]

The phylogenetic SD \(\sigma_{phylo}\) measures how strongly related species
remain similar in their fitted mean after the fixed effects have been included.
The residual `sigma` remains the within-observation residual SD. These are
different scale parameters.

The matching implemented R syntax is:

```r
drmTMB(
  drm_formula(
    y ~ x1 + phylo(1 | species, tree = tree),
    sigma ~ x1
  ),
  family = gaussian(),
  data = dat
)
```

## Worked example: thermal tolerance across species

Suppose we measure thermal tolerance for multiple observations per species and
ask whether body size predicts thermal tolerance after accounting for shared
ancestry. In a real analysis, `tree` would usually come from a phylogenetic
data workflow. For this vignette, the setup code creates a small ultrametric
branch-length tree so the example stays self-contained.

```{r}
set.seed(31)
tree <- balanced_ultrametric_tree(16)

species_info <- data.frame(
  species = tree$tip.label,
  body_size = scale(
    rnorm(16, mean = rep(c(-0.4, 0.4), each = 8), sd = 0.35)
  )[, 1]
)

phylo_shift <- rep(c(-0.5, 0.5), each = 8) +
  rep(c(-0.18, 0.18), each = 4, length.out = 16)

n_each <- 5
dat <- species_info[rep(seq_len(nrow(species_info)), each = n_each), ]
dat$body_size <- dat$body_size + rnorm(nrow(dat), sd = 0.08)
dat$thermal_tolerance <- 1.2 + 0.45 * dat$body_size +
  phylo_shift[match(dat$species, species_info$species)] +
  rnorm(nrow(dat), sd = 0.25)

head(dat)
```

The fitted model follows the equation above:

```{r}
fit_phylo <- drmTMB(
  drm_formula(
    thermal_tolerance ~ body_size + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat
)

summary(fit_phylo)
```

How to read this output:

The `mu:body_size` row is the fixed-effect slope for body size in the mean
thermal tolerance model. The `sigma:(Intercept)` row is the log residual SD, so
`exp(coef(fit_phylo, "sigma"))` gives the within-observation residual SD. The
`phylo(1 | species)` random-effect SD is \(\sigma_{phylo}\), the phylogenetic
among-species SD in the mean after body size has been included.

```{r}
phylo_sd <- unname(fit_phylo$sdpars$mu["phylo(1 | species)"])
residual_sd <- exp(coef(fit_phylo, "sigma")["(Intercept)"])
c(phylogenetic_sd = phylo_sd, residual_sd = residual_sd)
```

In this run, related species remain similar in thermal tolerance after
accounting for body size. The fitted phylogenetic SD is larger than the
within-observation residual SD, so shared ancestry explains a visible part of
the remaining mean-level structure.

`check_drm()` reports whether the optimizer, Hessian, scale, dropped-row, and
phylogenetic replication checks passed:

```{r}
check_drm(fit_phylo)
```

## Practical checks for phylogenetic fits

The public `phylo()` term requires an ultrametric tree with branch lengths.
Users pass the tree; `drmTMB` builds the sparse phylogenetic precision
internally. You do not need to construct the phylogenetic covariance matrix or
A-inverse yourself.

If a phylogenetic fit fails before optimization, check these points first:

- the tree object has class `phylo`;
- every observed species in `data$species` appears in `tree$tip.label`;
- species names use the same spelling and underscore/space convention in the
  data and tree;
- the tree has finite positive branch lengths and is ultrametric;
- the model uses the implemented syntax
  `phylo(1 | species, tree = tree)` in a univariate Gaussian `mu` formula, or
  matching terms in bivariate Gaussian `mu1` and `mu2` formulas.

Phylogenetic slopes, phylogenetic terms in `sigma`, structured effects in
`rho12`, and the full bivariate phylogenetic location-scale block are planned.
They should error clearly rather than silently fitting a different model.

Developer scaffolds now name and test the future q=4 phylogenetic endpoint
state across `mu1`, `mu2`, `sigma1`, and `sigma2`. The first fitted bivariate
slice uses matching intercept-only `mu1`/`mu2` `phylo()` terms, but the full
q=4 endpoint and reporting rows beyond the fitted mean-mean pair remain planned.

## Planned, not implemented: spatial random intercepts

Spatial fields are deliberately planned to look parallel to phylogenetic
effects, but the spatial likelihood is not implemented yet. The planned syntax
is:

```r
drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 | site, coords = coords),
    sigma ~ temp
  ),
  family = gaussian(),
  data = dat
)
```

The planned `coords` and `mesh` arguments play different roles. `coords`
should name the observation or site coordinates in the modelling data, ideally
in a projected coordinate system such as UTM rather than raw longitude and
latitude. A `mesh` is the computational scaffold for an SPDE/GMRF field: a
triangular network of vertices where the latent spatial field is represented
and then projected back to observations. Users should not interpret mesh
vertices as biological sampling units; they are numerical support points for
the spatial surface.

The spatial term is deliberately parallel to the phylogenetic term:

```r
phylo(1 | species, tree = tree)     # implemented for mu and matching mu1/mu2
spatial(1 | site, coords = coords)  # planned
spatial(1 | site, mesh = mesh)      # planned
```

The mathematical role is the same: add a structured Gaussian random intercept
to a distributional predictor. The computational route differs. `phylo()`
starts from an ultrametric tree with branch lengths and builds a sparse
tree-derived precision. `spatial()` will start from coordinates or a mesh and
build an SPDE/GMRF precision.

The computational plan is to keep using sparse A-inverse ideas for phylogeny
and add SPDE/GMRF structure for spatial fields, with provenance notes for any
code selectively ported from `gllvmTMB`.

Structured phylogenetic and spatial random slopes will be staged more
conservatively than ordinary grouped random slopes. The first structured-effect
target is intercept-only `mu`; the next should be one structured `mu` slope.
Small sets of structured slopes, interaction slopes, and structured effects in
`sigma` or `rho12` should wait until simulation recovery shows that the simpler
models are identifiable.

## Meta-analysis and future bivariate structure

In meta-analysis, the separation among known sampling covariance,
phylogenetic dependence, and spatial dependence is especially transparent:

\[
\begin{aligned}
e
  &\sim \operatorname{MVN}(\mathbf{0}, V)
  &&\text{known sampling covariance},\\
p_{species}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{phylo}^2 A)
  &&\text{phylogenetic structure},\\
\ell_{location}
  &\sim \operatorname{MVN}(\mathbf{0}, \sigma_{space}^2 M)
  &&\text{spatial structure}.
\end{aligned}
\]

The known matrix `V` handles sampling uncertainty. The phylogenetic and spatial
terms add extra biological or environmental dependence after that known
uncertainty has been included. The same design will let `drmTMB` support
structured Gaussian meta-analysis while keeping ecological and evolutionary
meta-analyses readable.

The first phylogenetic target is the location-scale model: trait means are the
first implemented target, and residual SDs are a planned target once the
location path has enough simulation and comparator coverage. In bivariate
Gaussian models, matching `mu1`/`mu2` phylogenetic location terms now estimate
the first mean-mean phylogenetic correlation. Scale-scale and mean-scale
phylogenetic correlations remain planned.

The first fitted bivariate phylogenetic location-coscale syntax looks like
this:

```r
fit_biv_phylo <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle + phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle + phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

This syntax fits correlated phylogenetic random intercepts in `mu1` and `mu2`.
It estimates two phylogenetic location SDs and one phylogenetic mean-mean
correlation. The `sigma1`, `sigma2`, and residual `rho12` formulas remain
ordinary fixed-effect distributional parameters in this first slice.

Current internal q=4 scaffolds only protect the algebra and endpoint names for
future contributors. Applied users who need phylogenetic scale or mean-scale
correlations should still treat those as planned until fitted likelihoods,
simulation recovery tests, and reporting rows are present. For the first fitted
mean-mean phylogenetic correlation, use `corpars$phylo`,
`corpairs(fit_biv_phylo, level = "phylogenetic")`, or
`summary(fit_biv_phylo)$covariance`. Run `check_drm(fit_biv_phylo)` before
interpreting that row: the current diagnostic flags phylogenetic correlations
near the boundary and phylogenetic location SDs that are tiny relative to the
matching residual scales. If `check_drm(fit_biv_phylo)` does not warn and the
fitted object kept its TMB object, a direct profile interval is available for
the phylogenetic correlation target:

```r
phylo_target <- "cor:phylo:cor(mu1:(Intercept),mu2:(Intercept) | phylo | species)"

corpairs(fit_biv_phylo, level = "phylogenetic")
summary(fit_biv_phylo)$covariance
confint(fit_biv_phylo, parm = phylo_target, method = "profile")
```

For a staged phylogenetic versus non-phylogenetic species comparison, add an
ordinary labelled species block beside the fitted phylogenetic layer:

```r
fit_biv_phylo_species <- drmTMB(
  formula = drm_formula(
    mu1 = log_body_mass ~ lifestyle +
      (1 | species_residual | species) +
      phylo(1 | species, tree = tree),
    mu2 = log_litter_size ~ lifestyle +
      (1 | species_residual | species) +
      phylo(1 | species, tree = tree),
    sigma1 = ~ lifestyle,
    sigma2 = ~ lifestyle,
    rho12 = ~ lifestyle
  ),
  family = c(gaussian(), gaussian()),
  data = mammals
)
```

That fit has an ordinary group-level species correlation, a phylogenetic
mean-mean correlation, and residual `rho12`. `corpairs()` keeps those layers in
separate rows. Because the ordinary group block and the phylogenetic block use
the same species factor, `check_drm()` records a note rather than letting the
two species-level correlations look automatically identifiable.

The residual `rho12` term remains distinct from the phylogenetic or spatial
correlation matrices. `rho12` is the residual coupling between two responses at
the observation level; `A` and `M` describe dependence among species or places.
In future bivariate structured models, `drmTMB` should also estimate
level-specific correlations such as phylogenetic mean-mean correlations,
non-phylogenetic species correlations, spatial field correlations, and
mean-scale correlations. These are not residual `rho12`; they should be
reported as structured or group-level covariance summaries.
