---
title: "Phylogenetic stable effects and temporal OU deviations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic stable effects and temporal OU deviations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
has_ape <- requireNamespace("ape", quietly = TRUE)
```

> **Development status.** This paired phylogenetic--OU model has passed its
> parser, dense-likelihood, method, and fixed-mean profile checks. Its retained
> point-recovery gate has not met the predeclared fixed-effect threshold.
> Treat the syntax below as a development workflow for checking a model and its
> diagnostics, not as an inference-qualified routine analysis. In particular,
> do not report confidence intervals from this paired model until its recovery
> and calibration gates are repaired and passed.

A comparative time-series data set can contain three different kinds of
variation. Closely related species can have similar *stable* baselines because
of shared evolutionary history. Each species can also make a short-lived
departure from its baseline that persists across nearby observation times.
Finally, observations have independent residual noise. The first paired
phylogenetic-temporal route keeps these quantities separate.

For observation time \(t\) in species \(i\), it fits

\[
y_{it}=\mathbf{x}_{it}^{T}\boldsymbol{\beta}+b_i+a_{it}+\epsilon_{it},
\]

\[
\mathbf b\sim\mathcal N(\mathbf 0,s_b^2A),\qquad
\operatorname{Cov}(a_{it},a_{js})=\mathbb{1}_{i=j}s_a^2
\exp\{-\lambda|t-s|\},\qquad
\epsilon_{it}\sim\mathcal N(0,\sigma^2).
\]

Here \(A\) is the covariance derived from the phylogenetic tree.
`sd_phylo_stable` describes stable between-species variation that follows this
tree. `sd_temporal` describes the stationary SD of an OU departure *within the
same species*, and `decay_temporal` is its positive decay rate in the units of
the elapsed-time column. `sigma` is independent observation-level noise.

The model is additive. It does **not** say that a temporal departure is shared
between related species. In particular, for two different species, their
covariance is \(s_b^2A_{ij}\), even when their observations occur close
together in time. A separable phylogeny-by-time field is a different future
model with a different scientific question.

## A small comparative time-series example

The example uses genuinely irregular elapsed times. The gap from 0.5 to 2.5
is two time units; `drmTMB` retains that gap and does not replace these values
with ranks.

```{r simulate, eval = has_ape}
library(ape)
set.seed(20260909)

n_species <- 8L
elapsed_values <- c(0, 0.5, 2.5, 5)
tree <- rcoal(n_species, tip.label = paste0("sp", seq_len(n_species)))

# Simulate the stable phylogenetic component once per species.
A <- drmTMB:::drm_phylo_tip_covariance(tree)
stable <- drop(t(chol(A)) %*% rnorm(n_species, sd = 0.45))
names(stable) <- tree$tip.label

# Simulate independent OU departures for each species.
R_ou <- exp(-0.45 * abs(outer(elapsed_values, elapsed_values, "-")))
ou_by_species <- lapply(tree$tip.label, function(species) {
  drop(t(chol(R_ou)) %*% rnorm(length(elapsed_values), sd = 0.55))
})
names(ou_by_species) <- tree$tip.label

dat <- do.call(rbind, lapply(tree$tip.label, function(species) {
  treatment <- rep(c(-0.5, 0.5), length.out = length(elapsed_values))
  data.frame(
    species = species,
    elapsed_days = elapsed_values,
    treatment = treatment,
    y = 1 + 0.4 * treatment + stable[[species]] +
      ou_by_species[[species]] + rnorm(length(elapsed_values), sd = 0.35)
  )
}))
dat$species <- factor(dat$species, levels = tree$tip.label)
head(dat)
```

Fit the phylogenetic stable intercept and OU process with the same `species`
column. The `tree` object is supplied by name, so its tips can be checked
against the observed species before the response is analysed.

```{r fit, eval = has_ape}
fit <- drmTMB(
  bf(
    y ~ treatment +
      phylo(1 | species, tree = tree) +
      temporal(1 | species, time = elapsed_days, structure = "ou"),
    sigma ~ 1
  ),
  data = dat,
  family = gaussian(),
  REML = FALSE
)

summary(fit)
check_drm(fit)
```

The paired route is intentionally narrow: it needs a univariate Gaussian ML
model, `sigma ~ 1`, finite numeric elapsed times, one response for every
species--time key, at least three observed species, at least two distinct times
per species, and at least three positive distinct lags across the data. It
rejects an ordinary `(1 | species)` intercept, different IDs for `phylo()` and
`temporal()`, temporal slopes, a second structured term, non-Gaussian families,
REML, non-unit weights, forecasts, and `newdata` prediction.

## Read the three sources of variation separately

The fit keeps the phylogenetic and temporal components distinct in its
structured metadata. The exact labels are intended to prevent a temporal SD
from being read as evolutionary variation, or vice versa.

```{r components, eval = has_ape}
fit$sdpars$mu
fit$decaypars$temporal
fit$sigma
```

`fitted(fit)` and `residuals(fit)` use the estimated stable and temporal
conditional deviations for these observed rows. `simulate(fit)` draws a new
phylogenetic stable vector and a new independent OU path for each species;
`simulate(fit, re.form = NA)` instead retains the fitted conditional latent
components. These are in-sample tools only.

```{r in-sample-tools, eval = has_ape}
head(fitted(fit))
head(residuals(fit))
simulate(fit, nsim = 2)
simulate(fit, nsim = 2, re.form = NA)
```

## What can be reported now

This development slice has an independently checked fixed-mean likelihood
profile calculation, but its recovery evidence is not yet sufficient to qualify
an interval workflow. For now, use the example to inspect the fitted covariance
components and the diagnostics; do not use `confint()`, Wald covariance,
bootstrap intervals, decay or variance-component intervals, forecasts, or
`newdata` prediction as a reported paired-model result. `check_drm(fit)` is a
necessary diagnostic but cannot turn an unqualified recovery result into
calibrated inference.

The existing [temporal AR1 and OU guide](temporal-random-effects.html) remains
the supported guide for independent-series temporal models. A later
phylogeny-by-OU field will require a separate formula, likelihood, simulation
study, and reader guide because it represents a different covariance model.
