---
title: "Phylogenetic structured effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic structured effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
library(drmTMB)

phylo_guide_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "bottom"
    )
}

phylo_eye_theme <- function() {
  phylo_guide_theme() +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_blank(),
      legend.position = "none"
    )
}

balanced_ultrametric_tree <- function(n_tip = 16L) {
  stopifnot(n_tip >= 2L, log2(n_tip) == floor(log2(n_tip)))
  edges <- matrix(integer(), ncol = 2L)
  edge_lengths <- numeric()
  next_node <- n_tip + 1L

  build <- function(tips) {
    if (length(tips) == 1L) {
      return(tips)
    }
    node <- next_node
    next_node <<- next_node + 1L
    mid <- length(tips) / 2L
    left <- build(tips[seq_len(mid)])
    right <- build(tips[seq.int(mid + 1L, length(tips))])
    edges <<- rbind(edges, c(node, left), c(node, right))
    edge_lengths <<- c(edge_lengths, 1, 1)
    node
  }

  build(seq_len(n_tip))
  structure(
    list(
      edge = edges,
      edge.length = edge_lengths,
      tip.label = paste0("sp_", seq_len(n_tip)),
      Nnode = n_tip - 1L
    ),
    class = "phylo"
  )
}

simulate_phylo_guide_data <- function(seed = 31) {
  set.seed(seed)
  tree <- balanced_ultrametric_tree(16)
  species_info <- data.frame(
    species = tree$tip.label,
    body_size = scale(
      rnorm(16, mean = rep(c(-0.4, 0.4), each = 8), sd = 0.35)
    )[, 1]
  )
  phylo_shift <- rep(c(-0.5, 0.5), each = 8) +
    rep(c(-0.18, 0.18), each = 4, length.out = 16)

  n_each <- 5
  dat <- species_info[rep(seq_len(nrow(species_info)), each = n_each), ]
  dat$body_size <- dat$body_size + rnorm(nrow(dat), sd = 0.08)
  dat$thermal_tolerance <- 1.2 + 0.45 * dat$body_size +
    phylo_shift[match(dat$species, species_info$species)] +
    rnorm(nrow(dat), sd = 0.25)

  list(tree = tree, data = dat)
}

simulate_phylo_q2_guide_data <- function(seed = 42) {
  set.seed(seed)
  tree <- balanced_ultrametric_tree(16)
  A_tip <- drmTMB:::drm_phylo_tip_covariance(tree)
  z1 <- rnorm(length(tree$tip.label))
  z2 <- 0.55 * z1 + sqrt(1 - 0.55^2) * rnorm(length(tree$tip.label))
  u1 <- as.vector(t(chol(A_tip)) %*% z1) * 0.45
  u2 <- as.vector(t(chol(A_tip)) %*% z2) * 0.35
  names(u1) <- tree$tip.label
  names(u2) <- tree$tip.label

  n_each <- 4
  species <- rep(tree$tip.label, each = n_each)
  climate <- rnorm(length(species))
  e1 <- rnorm(length(species))
  e2 <- 0.15 * e1 + sqrt(1 - 0.15^2) * rnorm(length(species))

  list(
    tree = tree,
    data = data.frame(
      species = species,
      climate = climate,
      heat_tolerance = 0.30 + 0.35 * climate + u1[species] + 0.35 * e1,
      desiccation_tolerance = -0.20 - 0.25 * climate + u2[species] + 0.32 * e2
    )
  )
}
```

Use `phylo()` when species share evolutionary history and the scientific
question is whether related species have similar location deviations, structured
location-scale covariance, direct phylogenetic SD surfaces, or a fitted
phylogenetic correlation-pair regression. The fitted route uses a tree supplied
as `tree = tree`.

## What is fitted today

| Question | Syntax | Status |
| --- | --- | --- |
| Does one Gaussian response have phylogenetic location deviations? | `phylo(1 | species, tree = tree)` in `mu` | Fitted. |
| Do two Gaussian response means share a phylogenetic correlation? | matching `phylo(1 | species, tree = tree)` or labelled `phylo(1 | p | species, tree = tree)` terms in `mu1` and `mu2` | Fitted q=2 location-location slice. `corpairs(level = "phylogenetic")` reports the latent phylogenetic row separately from residual `rho12`. |
| Do phylogenetic location and scale deviations form one q=4 block? | matching labelled `phylo(1 | p | species, tree = tree)` terms across `mu1`, `mu2`, `sigma1`, and `sigma2` | Fitted first constant q=4 location-scale slice. Partial, unlabelled, mismatched, and slope forms remain rejected. |
| Do species-level predictors change the phylogenetic location SD? | `sd_phylo(species) ~ z`, `sd_phylo1(species) ~ z1`, or `sd_phylo2(species) ~ z2` | Fitted direct-SD route for phylogenetic location SD surfaces. This is not residual `sigma`. |
| Do species-level predictors change the q=2 phylogenetic mean-mean correlation? | `corpair(species, level = "phylogenetic", block = "p", from = "mu1", to = "mu2") ~ ecology` | Fitted q=2 predictor-dependent correlation route. q=4 endpoint-specific `corpair()` regressions remain planned. |

## Start with the smallest useful model

For one response, start with a phylogenetic location intercept:

```r
fit_phylo <- drmTMB(
  y ~ treatment + phylo(1 | species, tree = tree),
  data = dat,
  family = gaussian()
)
```

For two response means, use matching terms and read the latent correlation with
`corpairs()`:

```r
fit_phylo_q2 <- drmTMB(
  mu1 = trait1 ~ treatment +
    phylo(1 | p | species, tree = tree),
  mu2 = trait2 ~ treatment +
    phylo(1 | p | species, tree = tree),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_phylo_q2, level = "phylogenetic")
rho12(fit_phylo_q2)
```

`corpairs()` reports the fitted latent phylogenetic correlation among
species-level location deviations. `rho12()` reports the residual correlation
between responses after fixed effects and random effects have been included.

## What to inspect

After fitting, inspect the phylogenetic layer before interpreting it:

| Output | Use |
| --- | --- |
| `check_drm(fit)` | Confirm the phylogenetic layer was recognized and review tree/species diagnostics. |
| `fit$sdpars$mu` | Read fitted phylogenetic location SDs. |
| `ranef(fit, "phylo_mu")` | Inspect conditional species deviations on the fitted phylogenetic layer. |
| `summary(fit)$covariance` | Check how phylogenetic SDs and correlations are reported beside other covariance layers. |
| `profile_targets(fit)` | See which phylogenetic SD or constant q=2 correlation targets can be profiled directly. |
| `corpairs(fit, level = "phylogenetic")` | Read fitted constant or modelled phylogenetic correlation rows. |

Direct-SD terms such as `sd_phylo(species) ~ z` model species-level predictors
of a phylogenetic location standard deviation. They do not model residual
`sigma`, and they do not imply a q=4 location-scale block unless the matching
q=4 `phylo()` terms are also fitted.

## Rendered checks

The small examples below separate raw response data, fitted phylogenetic SDs,
and fitted latent correlations. Raw observations can help readers understand
the trait scale, but they are not uncertainty intervals for `phylo()` SDs or
`corpairs()` rows.

```{r phylo-guide-fit}
phylo_example <- simulate_phylo_guide_data()
tree <- phylo_example$tree
dat <- phylo_example$data

fit_phylo <- drmTMB(
  drm_formula(
    thermal_tolerance ~ body_size + phylo(1 | species, tree = tree),
    sigma ~ 1
  ),
  family = gaussian(),
  data = dat
)
```

```{r phylo-sd-figure, fig.width = 6.6, fig.height = 3.4, fig.cap = "Fitted residual `sigma` and phylogenetic location SD from a univariate Gaussian `phylo()` model; points and bars are response-scale estimates with 95% Wald intervals from `confint(fit_phylo, parm = \"variance_components\")`.", fig.alt = "Horizontal interval display comparing residual sigma and the phylogenetic location standard deviation. Both rows have hollow point estimates and finite 95 percent Wald intervals."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  phylo_sd <- confint(fit_phylo, parm = "variance_components")
  phylo_sd$estimate <- c(
    unname(exp(coef(fit_phylo, "sigma")["(Intercept)"])),
    unname(fit_phylo$sdpars$mu["phylo(1 | species)"])
  )
  phylo_sd$label <- c(
    "Residual\nsigma",
    "Phylogenetic\nlocation SD"
  )

  ggplot2::ggplot(phylo_sd, ggplot2::aes(y = label)) +
    ggplot2::geom_vline(
      xintercept = 0,
      linewidth = 0.45,
      linetype = "dashed",
      colour = "grey55"
    ) +
    ggplot2::geom_errorbar(
      ggplot2::aes(xmin = lower, xmax = upper),
      width = 0,
      linewidth = 1.1,
      colour = "#0072B2"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = estimate),
      shape = 21,
      size = 3.5,
      stroke = 1,
      fill = "white",
      colour = "#0072B2"
    ) +
    ggplot2::scale_x_continuous(
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    phylo_eye_theme() +
    ggplot2::labs(
      title = "Phylogenetic SD is separate from residual sigma",
      subtitle = "Bars are 95% Wald intervals on the response scale",
      x = "Fitted standard deviation",
      y = NULL
    )
}
```

For two response means, the phylogenetic correlation row is a fitted latent
correlation among species-level location deviations. It is not the residual
`rho12` correlation and it is not a raw sample correlation among observations.

```{r phylo-q2-correlation-fit}
phylo_q2_example <- simulate_phylo_q2_guide_data()
phylo_q2_tree <- phylo_q2_example$tree
phylo_q2_dat <- phylo_q2_example$data

fit_phylo_q2_example <- drmTMB(
  drm_formula(
    mu1 = heat_tolerance ~ climate +
      phylo(1 | p | species, tree = phylo_q2_tree),
    mu2 = desiccation_tolerance ~ climate +
      phylo(1 | p | species, tree = phylo_q2_tree),
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = phylo_q2_dat
)

phylo_q2_pairs <- corpairs(
  fit_phylo_q2_example,
  level = "phylogenetic",
  conf.int = TRUE,
  ystep = 0.5,
  ytol = 2
)
phylo_q2_pairs
```

```{r phylo-q2-confidence-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "Phylogenetic q=2 location-location correlation from `corpairs()`; the Confidence Eye is a 95% profile interval on the latent phylogenetic-correlation scale and the dotted vertical line marks zero correlation.", fig.alt = "Single-row Confidence Eye for the phylogenetic mean-mean correlation. The eye spans the 95 percent profile interval and the hollow point estimate is to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  phylo_q2_display <- phylo_q2_pairs
  phylo_q2_display$display_label <- "Phylogenetic\nmu1-mu2"

  plot_corpairs(
    phylo_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("phylogenetic" = "#D55E00")) +
    ggplot2::scale_fill_manual(values = c("phylogenetic" = "#D55E00")) +
    phylo_eye_theme() +
    ggplot2::labs(
      title = "Phylogenetic latent mean correlation",
      subtitle = "Eye is a 95% profile interval; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

## Boundaries

The following phylogenetic routes remain planned:

- phylogenetic slopes such as `phylo(1 + x | species, tree = tree)`;
- public phylogenetic matrix inputs such as `phylo(1 | species, A = A)`;
- simultaneous `phylo()` plus `spatial()` layers in the same formula;
- structured random effects in residual `rho12`;
- predictor-dependent q=4 phylogenetic `corpair()` regressions;
- non-Gaussian phylogenetic structured effects beyond the fitted Gaussian
  slices.

Use the [structural-dependence overview](structural-dependence.html) when you
are choosing among `animal()`, `phylo()`, `spatial()`, and `relmat()`. Use the
detailed [structural-dependence tutorial](phylogenetic-spatial.html) when you
need the current worked examples, equations, and broader parity ladder.
