---
title: "Phylogenetic mixed models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic mixed models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
library(drmTMB)

# Suggested packages used only inside guarded chunks. When either is absent the
# corresponding chunk degrades to a short note instead of erroring at build.
has_ape <- requireNamespace("ape", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

Closely related species are not independent observations. They share ancestry,
so a trait value carried by one species is partly inherited from the same
nodes that shaped its relatives. A phylogenetic mixed model encodes that shared
history as a structured random effect: a species-level deviation whose
covariance is read off the tree, so that sister taxa are expected to deviate
from the regression line in the same direction.

`drmTMB` exposes this through the `phylo()` term. This article explains where
the covariance comes from, how to write a `phylo()` model for a Gaussian
response, and how the same syntax carries over to non-Gaussian responses such
as counts. Everything below runs at build time on a small simulated tree, so
you can read the fitted numbers next to the values that generated them.

If you are choosing between `animal()`, `phylo()`, `spatial()`, and `relmat()`,
read the [structural-dependence overview](structural-dependence.html) first.
When observations belong to a *pair* of species drawn from two different trees,
read [two-tree phylogenetic interactions](bipartite-phylogenetic-interactions.html).

## The phylogenetic covariance

A `phylo()` term adds a vector of species-level location deviations $u$ with a
mean of zero and a covariance proportional to the phylogenetic covariance
matrix $A$,

\[
\begin{aligned}
y_{ij} &= \mathbf{x}_{ij}^\top \boldsymbol{\beta}
          + u_{s(i,j)} + \varepsilon_{ij}, \\
\mathbf{u} &\sim \mathcal{N}\!\left(\mathbf{0},\, \sigma_{\text{phylo}}^2\, A\right), \\
\varepsilon_{ij} &\sim \mathcal{N}\!\left(0,\, \sigma^2\right),
\end{aligned}
\]

where $s(i,j)$ is the species of observation $j$ in group $i$. Under a Brownian
motion model of trait evolution, $A_{kl}$ is the height of the most recent
common ancestor of species $k$ and $l$: the longer two species have shared an
evolutionary path, the more strongly their deviations covary. For an
ultrametric tree (all tips equidistant from the root), the diagonal of $A$ is
constant and the off-diagonal entries are the shared root-to-ancestor path
lengths.

Two standard deviations appear, and they answer different questions:

- $\sigma_{\text{phylo}}$ is the **phylogenetic** SD: how much species deviate
  from the fixed-effect prediction in a way that tracks the tree.
- $\sigma$ is the **residual** SD: variation among observations within a
  species, independent of ancestry.

Their ratio is the phylogenetic signal,

\[
\lambda = \frac{\sigma_{\text{phylo}}^2}{\sigma_{\text{phylo}}^2 + \sigma^2},
\]

the proportion of the random-plus-residual variance attributable to phylogeny.
`drmTMB` reports this as a derived quantity (see below). Internally `drmTMB`
does not invert the dense $A$. It uses the Hadfield and Nakagawa (2010) sparse
phylogenetic precision (an $A^{-1}$ built from the tree's branch lengths), which
keeps the augmented-state Laplace approximation fast as the number of species
grows.

## A small tree with `ape::rcoal()`

We simulate a tree small enough to fit in a fraction of a second. `ape::rcoal()`
draws a random **ultrametric** coalescent tree, which is exactly the shape
`phylo()` expects (branch lengths present, all tips contemporaneous).

```{r simulate-tree, eval = has_ape}
library(ape)
set.seed(2026)

n_species <- 16L
tree <- rcoal(n_species, tip.label = paste0("sp", seq_len(n_species)))

c(ultrametric = is.ultrametric(tree), n_tip = length(tree$tip.label))
```

The covariance $A$ implied by this tree is available from the same internal
helper the fitting code uses. We only need it here to *simulate* a trait with
known phylogenetic structure; you never compute it by hand for a real fit.

```{r tip-covariance, eval = has_ape}
A <- drmTMB:::drm_phylo_tip_covariance(tree)
dim(A)
round(A[1:4, 1:4], 3)
```

## Simulating a Gaussian trait with phylogenetic signal

We draw one species-level deviation per tip from $\mathcal{N}(0,
\sigma_{\text{phylo}}^2 A)$ using a Cholesky factor of $A$, add a fixed
covariate effect, and add independent residual noise. Several observations per
species let the model separate the phylogenetic SD from the residual SD.

```{r simulate-gaussian, eval = has_ape}
sd_phylo_true <- 0.8     # phylogenetic SD
sigma_true    <- 0.3     # residual SD
n_per_species <- 6L

# One deviation per species, correlated along the tree.
u <- as.vector(t(chol(A)) %*% rnorm(n_species, sd = sd_phylo_true))
names(u) <- tree$tip.label

species <- rep(tree$tip.label, each = n_per_species)
x <- rnorm(length(species))

# y = intercept + slope * x + phylogenetic deviation + residual noise
trait <- 0.5 - 0.4 * x + u[species] + rnorm(length(species), sd = sigma_true)

dat <- data.frame(
  trait = unname(trait),
  x = x,
  species = species
)
head(dat)
```

## Fitting with `phylo(1 | species)`

The `phylo()` term goes inside the mean (`mu`) formula. It takes a random-effect
specification, `1 | species`, and the tree as `tree = tree`. Wrap the formulas
in `bf()` (the `drmTMB` formula builder) and pass a family, exactly as for any
other `drmTMB` model.

```{r fit-gaussian, eval = has_ape}
fit <- drmTMB(
  bf(trait ~ x + phylo(1 | species, tree = tree),
     sigma ~ 1),
  family = gaussian(),
  data = dat
)

fit$opt$convergence   # 0 indicates the optimizer converged
```

A few accessors return the pieces of the fit:

```{r gaussian-pieces, eval = has_ape}
coef(fit, "mu")              # fixed-effect mean coefficients
fit$sdpars$mu                # phylogenetic location SD
stats::sigma(fit)            # residual SD
```

The recovered slope (`x`) is close to its true value of `-0.4`, and the
phylogenetic and residual SDs are in the neighbourhood of `0.8` and `0.3`. The
fixed intercept need not match the simulated `0.5`: with only `r n_species`
species the species-level deviations carry a non-zero sample mean that the
random effect absorbs, so the intercept and the deviations trade off. This is
expected behaviour, not a fitting error -- the scientifically interpretable
quantities are the slope and the two SDs.

### Reading the phylogenetic signal

`summary()` adds a derived row for the phylogenetic signal $\lambda$, alongside
the random-effect and residual variances it is built from.

```{r gaussian-signal, eval = has_ape}
summary(fit)$derived[, c(
  "quantity", "estimate",
  "random_effect_variance", "residual_variance"
)]
```

`estimate` is $\lambda$. A value well above zero says that, after accounting for
the covariate, related species really do resemble each other more than
unrelated ones. The species-level deviations themselves are available with
`ranef()`:

```{r gaussian-ranef, eval = has_ape}
phylo_dev <- ranef(fit, "phylo_mu")
str(phylo_dev$values)
```

### Confidence interval for the phylogenetic SD

Variance components have their own `confint()` target. The interval is on the
response (SD) scale, obtained by transforming the Wald interval for the
log-SD parameter.

```{r gaussian-confint, eval = has_ape}
confint(fit, parm = "variance_components")[, c(
  "parm", "lower", "upper", "scale"
)]
```

A lower bound comfortably above zero is the interval-based counterpart of a
non-trivial phylogenetic signal. A `check_drm(fit)` call (not shown) reports the
same diagnostics for the recognised phylogenetic layer.

```{r gaussian-figure, eval = has_ape && has_ggplot2, fig.width = 6.6, fig.height = 3.2, fig.cap = "Fitted residual SD and phylogenetic location SD with 95% Wald intervals on the response scale. The two standard deviations answer different questions and are reported separately.", fig.alt = "Two horizontal interval bars. The phylogenetic location standard deviation and the residual standard deviation each have a point estimate and a 95 percent interval, with the phylogenetic interval bounded away from zero."}
vc <- confint(fit, parm = "variance_components")
sd_tab <- data.frame(
  label = c("Phylogenetic\nlocation SD", "Residual\nsigma"),
  estimate = c(
    unname(fit$sdpars$mu[["phylo(1 | species)"]]),
    unname(stats::sigma(fit)[[1L]])
  ),
  lower = c(vc$lower[1L], NA),
  upper = c(vc$upper[1L], NA)
)

ggplot2::ggplot(sd_tab, ggplot2::aes(x = estimate, y = label)) +
  ggplot2::geom_errorbarh(
    ggplot2::aes(xmin = lower, xmax = upper),
    height = 0, linewidth = 1.1, colour = "#0072B2", na.rm = TRUE
  ) +
  ggplot2::geom_point(shape = 21, size = 3.4, stroke = 1,
                      fill = "white", colour = "#0072B2") +
  ggplot2::labs(
    title = "Phylogenetic SD is separate from residual sigma",
    x = "Fitted standard deviation", y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
```

## Non-Gaussian responses

The same `phylo(1 | species, tree = tree)` term works for ordinary Poisson and
negative-binomial (NB2) location models. The phylogenetic deviation now acts on
the linear predictor of the count mean through the log link, so related species
share a baseline abundance. Two differences from the Gaussian case:

- count families have no residual `sigma` formula, so the `phylo()` term is the
  only random piece in `mu`;
- the simulated deviation is built from the **correlation** form of $A$ (unit
  diagonal) scaled by the phylogenetic SD, so the SD is interpretable on the
  log-mean scale.

```{r simulate-count, eval = has_ape}
set.seed(11)

# Standardise A to a correlation matrix, then scale by the phylogenetic SD.
A_cor <- A / outer(sqrt(diag(A)), sqrt(diag(A)))
sd_phylo_count <- 0.5
u_count <- as.vector(t(chol(A_cor)) %*% rnorm(n_species)) * sd_phylo_count
names(u_count) <- tree$tip.label

species_c <- rep(tree$tip.label, each = n_per_species)
x_c <- rep(seq(-1, 1, length.out = n_per_species), times = n_species)

eta <- log(3) - 0.3 * x_c + u_count[species_c]   # log mean
count <- rpois(length(eta), lambda = exp(eta))

dat_count <- data.frame(count = count, x = x_c, species = species_c)
range(dat_count$count)
```

A Poisson fit uses the same call with `family = poisson()`:

```{r fit-poisson, eval = has_ape}
fit_pois <- drmTMB(
  bf(count ~ x + phylo(1 | species, tree = tree)),
  family = poisson(link = "log"),
  data = dat_count
)

fit_pois$opt$convergence
coef(fit_pois, "mu")     # log-mean intercept near log(3) ~ 1.10, slope near -0.3
fit_pois$sdpars$mu       # phylogenetic SD on the log-mean scale
```

If the counts are overdispersed relative to a Poisson, swap in `nbinom2()`. The
NB2 family adds an overdispersion (scale) parameter while keeping the identical
phylogenetic location term.

```{r fit-nbinom2, eval = has_ape}
fit_nb <- drmTMB(
  bf(count ~ x + phylo(1 | species, tree = tree)),
  family = nbinom2(),
  data = dat_count
)

fit_nb$opt$convergence
fit_nb$sdpars$mu         # phylogenetic SD, NB2 mean model
```

The phylogenetic SD is reported on the log-mean (link) scale for count
families, so it is not directly comparable to the Gaussian response-scale SD;
compare it instead to other log-scale effects in the same model.

## A larger fit, for reference

The fits above are tiny by design. For intuition about runtime on a more
realistic tree, the chunk below sketches a 200-species fit. It is marked
`eval = FALSE` so the vignette never blocks on it; the numbers in the comments
are illustrative of the shape of the output, not a benchmarked claim.

```{r large-sketch, eval = FALSE}
set.seed(99)
big_tree <- ape::rcoal(200, tip.label = paste0("t", 1:200))
A_big <- drmTMB:::drm_phylo_tip_covariance(big_tree)
u_big <- as.vector(t(chol(A_big)) %*% rnorm(200, sd = 0.7))
names(u_big) <- big_tree$tip.label

sp <- rep(big_tree$tip.label, each = 4L)
xb <- rnorm(length(sp))
yb <- 0.2 + 0.5 * xb + u_big[sp] + rnorm(length(sp), sd = 0.3)
big <- data.frame(y = yb, x = xb, species = sp)

fit_big <- drmTMB(
  bf(y ~ x + phylo(1 | species, tree = big_tree), sigma ~ 1),
  family = gaussian(),
  data = big
)
fit_big$sdpars$mu
#> phylo(1 | species)
#>              ~0.70
# The sparse-precision path keeps this on the order of a second on a laptop;
# cost grows roughly linearly in the number of species rather than cubically.
```

## What `phylo()` fits today

| Question | Syntax | Status |
| --- | --- | --- |
| Phylogenetic location deviations in one Gaussian response | `phylo(1 | species, tree = tree)` in `mu` | Fitted. |
| Phylogenetic location deviations in a Poisson or NB2 mean | `phylo(1 | species, tree = tree)` in `mu` | Fitted q=1 location intercept. |
| One phylogenetic location SD modelled by a species predictor | `sd_phylo(species) ~ z` | Fitted direct-SD route (not residual `sigma`). |
| Two Gaussian means sharing a phylogenetic correlation | matching `phylo(1 | p | species, tree = tree)` in `mu1` and `mu2` | Fitted; read with `corpairs(level = "phylogenetic")`. |
| Phylogenetic location and scale as one block | matching labelled `phylo(1 | p | species, tree = tree)` across `mu1`, `mu2`, `sigma1`, `sigma2` | Fitted first constant q=4 slice. |

These bivariate and location-scale routes, and the broader parity ladder, are
covered with worked examples in the
[structural-dependence tutorial](phylogenetic-spatial.html).

The following remain planned: multiple phylogenetic slopes and slope
correlations; non-Gaussian phylogenetic *slopes* (only the q=1 intercept is
fitted for counts); public matrix inputs such as `phylo(1 | species, A = A)`;
simultaneous `phylo()` and `spatial()` layers in one formula; and
structured random effects inside residual `rho12`.

## References

Hadfield, J. D. and Nakagawa, S. (2010). General quantitative genetic methods
for comparative biology: phylogenies, taxonomies and multi-trait models for
continuous and categorical characters. *Journal of Evolutionary Biology*,
23(3), 494-508.
