---
title: "Profile-likelihood intervals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile-likelihood intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144
)
```

Profile-likelihood intervals ask a different question from Wald intervals. A
Wald interval uses local curvature near the fitted estimate. A profile interval
fixes one target at a sequence of values, re-optimizes the nuisance parameters,
and keeps the values whose likelihood-ratio distance is still compatible with
the fitted model.

This article shows the full profile object for one Gaussian residual-scale
target. The same workflow applies to other direct targets listed by
`profile_targets()`.

## Fit a small model

The example asks whether a continuous response changes with temperature while
the residual standard deviation is constant:

\[
\begin{aligned}
y_i \mid \mu_i, \sigma
  &\sim \operatorname{Normal}(\mu_i, \sigma^2),\\
\mu_i
  &= \beta_0 + \beta_1 \text{temperature}_i .
\end{aligned}
\]

```{r}
library(drmTMB)

set.seed(20260608)
n <- 70
dat <- data.frame(temperature = runif(n, -1.2, 1.2))
dat$growth <- 0.5 + 0.8 * dat$temperature + rnorm(n, sd = 0.55)

fit <- drmTMB(
  bf(growth ~ temperature, sigma ~ 1),
  family = gaussian(),
  data = dat
)
```

`profile_targets()` names the fitted quantities that can be requested. The
constant residual SD appears twice: once as the fixed-effect coefficient on the
log-`sigma` link scale, and once as the response-scale target `sigma`.

```{r}
profile_targets(fit)
```

## Compute the profile curve

The focused call below profiles the response-scale residual SD. `compare =
TRUE` keeps the first pass and the denser pass in the returned table, so the
plot can show how much extra sampling changed the likelihood-ratio curve.
`profile_precision = "fast"` is a quick demonstration setting; for a final
analysis, rerun important targets with the default controls or a deliberate
`profile_maxit`.

```{r}
prof <- profile(
  fit,
  parm = "sigma",
  level = 0.95,
  compare = TRUE,
  profile_precision = "fast",
  trace = FALSE
)
```

The profile stores the fitted estimate, the interval endpoints, the
likelihood-ratio distance, the pass label, elapsed seconds, and the profile
engine metadata.

```{r}
profile_summary <- unique(prof[c(
  "parm",
  "estimate",
  "conf.low",
  "conf.high",
  "conf.status",
  "profile.message",
  "profile_source"
)])
profile_summary
```

```{r}
profile_timing <- aggregate(
  elapsed ~ profile_pass + profile_controls,
  data = prof,
  FUN = unique
)
profile_timing
```

## Inspect the likelihood-ratio curve

The dashed horizontal line is the 95% likelihood-ratio cutoff. The vertical
line marks the fitted `sigma` estimate, and the interval endpoints show where
the profiled curve crosses the cutoff.

```{r}
plot(prof) +
  ggplot2::labs(
    title = "Residual SD profile",
    subtitle = "Dotted line is the 95% cutoff; dashed lines are profile endpoints.",
    caption = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    legend.position = "bottom"
  )
```

The interval is useful only because the curve crosses the cutoff on both sides
of the estimate. If one side does not cross, or if `conf.status`,
`profile.boundary`, or `profile.message` reports a boundary or failure, treat
the interval as diagnostic evidence rather than a finished uncertainty
summary.

## Request only the endpoints

When a full curve is not needed, `confint()` can return the profile endpoints.
For direct scalar scale, SD, and correlation targets, `profile_engine = "auto"`
uses a faster endpoint solver unless full-profile controls require
`TMB::tmbprofile()`. Use `profile_engine = "tmbprofile"` when you want the
same full-profile route used above.

```{r}
confint(
  fit,
  parm = "sigma",
  method = "profile",
  profile_engine = "tmbprofile",
  profile_precision = "fast"
)
```

The profile table and plot are the safer first read for a new target. The
endpoint interval is the compact result after the curve shape has been checked.
