---
title: "Known-matrix relatedness with relmat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Known-matrix relatedness with relmat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144,
  dev.args = list(bg = "white")
)
library(drmTMB)

relmat_guide_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA),
      legend.position = "bottom"
    )
}

relmat_eye_theme <- function() {
  relmat_guide_theme() +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_blank(),
      legend.position = "none"
    )
}

relatedness_heatmap_data <- function(x) {
  out <- as.data.frame(as.table(x), stringsAsFactors = FALSE)
  names(out) <- c("row", "column", "relatedness")
  out$row <- factor(out$row, levels = rev(rownames(x)))
  out$column <- factor(out$column, levels = colnames(x))
  out
}

simulate_relmat_guide_data <- function(seed = 20260524) {
  set.seed(seed)
  n_line <- 24
  n_each <- 5
  line_levels <- paste0("line_", seq_len(n_line))
  line <- factor(rep(line_levels, each = n_each))
  temperature <- runif(length(line), -1, 1)
  treatment <- factor(rep(rep(c("control", "heated"), each = n_each), n_line / 2))

  K <- outer(seq_len(n_line), seq_len(n_line), function(i, j) {
    0.36^abs(i - j)
  })
  diag(K) <- diag(K) + 0.08
  dimnames(K) <- list(line_levels, line_levels)
  Q <- solve(K)

  line_effect <- as.vector(t(chol(K)) %*% rnorm(n_line)) * 0.42
  names(line_effect) <- line_levels
  sigma <- exp(-1.20 + 0.10 * (treatment == "heated"))
  seed_mass <- 2.2 + 0.50 * temperature + 0.25 * (treatment == "heated") +
    line_effect[line] + rnorm(length(line), sd = sigma)

  list(
    data = data.frame(
      seed_mass = seed_mass,
      temperature = temperature,
      treatment = treatment,
      line = line
    ),
    K = K,
    Q = Q
  )
}

simulate_relmat_q2_guide_data <- function(seed = 1) {
  set.seed(seed)
  n_observation_block <- 40
  n_line <- 10
  n_each <- 4
  observation <- factor(rep(paste0("obs_", seq_len(n_observation_block)), each = n_each))
  sex <- factor(rep(rep(c("female", "male"), each = n_each), n_observation_block / 2))
  age <- runif(length(observation), 0, 2)
  line_levels <- paste0("line_", seq_len(n_line))
  line <- factor(rep(rep(line_levels, each = 4), each = n_each))

  K <- outer(seq_len(n_line), seq_len(n_line), function(i, j) {
    0.25^abs(i - j)
  })
  diag(K) <- diag(K) + 0.10
  dimnames(K) <- list(line_levels, line_levels)
  Q <- solve(K)

  z_mass <- rnorm(n_line)
  z_height <- 0.55 * z_mass + sqrt(1 - 0.55^2) * rnorm(n_line)
  u_mass <- as.vector(t(chol(K)) %*% z_mass) * 0.35
  u_height <- as.vector(t(chol(K)) %*% z_height) * 0.30
  names(u_mass) <- line_levels
  names(u_height) <- line_levels

  e_mass <- rnorm(length(line))
  e_height <- 0.10 * e_mass + sqrt(1 - 0.10^2) * rnorm(length(line))
  sigma_mass <- exp(-1.0 + 0.15 * (sex == "male"))
  sigma_height <- exp(-1.1 - 0.10 * (sex == "male"))

  list(
    data = data.frame(
      line = line,
      age = age,
      sex = sex,
      seed_mass = 2.1 + 0.45 * age + 0.20 * (sex == "male") +
        u_mass[line] + sigma_mass * e_mass,
      plant_height = 1.4 + 0.30 * age - 0.12 * (sex == "male") +
        u_height[line] + sigma_height * e_height
    ),
    Q = Q
  )
}
```

Use `relmat()` when the dependence is among latent group-level deviations and
the matrix is already known. The matrix can be a covariance, a correlation, or
an inverse covariance/precision matrix. The common case is a correlation-like
matrix among lines, strains, plots, populations, or experimental units:

```text
u_id ~ N(0, sd_relmat^2 K)
y_i  = mu_i + u_id[i] + error_i
```

If the natural input is a precision matrix, use `Q` instead:

```text
u_id ~ N(0, sd_relmat^2 Q^-1)
```

This is a higher-level random-effect structure. It is not the same as known
sampling covariance among observed effect-size estimates; that observation-level
matrix belongs to the meta-analysis route with `meta_V(V = V)`.

## When is `relmat()` the right tool?

Use `relmat()` only after the named biological or spatial route is not the
better description.

| Situation | Use | Why |
| --- | --- | --- |
| Genomic or marker-based similarity among inbred lines, strains, or cultivars | `relmat(1 | line, K = G)` | `G` is often a correlation or genomic relationship matrix. The fitted SD turns it into the latent covariance `sd_relmat^2 G`. |
| A lab, assay, or ecological kernel among experimental units | `relmat(1 | unit, K = K_unit)` | The matrix says which unit-level deviations should be similar, but it is not a pedigree, species tree, or coordinate-spatial surface. |
| A precomputed inverse relatedness matrix from another tool | `relmat(1 | id, Q = Q_id)` | The natural object is already precision-like, so `drmTMB` does not need to invert a dense covariance matrix. |
| A graph, network, river, areal, or custom Gaussian Markov random-field precision that was built and checked outside `drmTMB` | `relmat(1 | node, Q = Q_node)` | The scientific structure is a latent node-level field with a known precision, but it is not the current coordinate `spatial()` route. |

Do not use `relmat()` just because you have any covariance matrix. If the
matrix is known sampling covariance for observed estimates, use
`meta_V(V = V)`. If the matrix is additive relatedness from a pedigree or animal
model, use `animal()`. If it comes from an ultrametric species tree, use
`phylo()`. If coordinates define the dependence, use `spatial()`.

## Choosing `K` or `Q`

Use `K` when the matrix is easiest to interpret as relatedness, covariance, or
correlation. A correlation matrix with diagonal 1 is usually the clearest input:
the fitted `sd_relmat` then controls the variance scale of the latent
deviations. A covariance matrix with non-unit diagonal is also allowed, but then
the diagonal scale is part of the known structure.

Use `Q` when the matrix is easiest to interpret as an inverse covariance or
precision. This is common when another model, pedigree engine, graph model, or
Gaussian Markov random-field construction gives a sparse precision matrix. In
that case the latent covariance is proportional to `solve(Q)`, but the model can
work with `Q` directly.

## Two concrete examples

Suppose `G` is a marker-derived genomic relationship matrix among experimental
lines. The scientific question is whether lines that are genomically similar
also have similar unexplained deviations in seed mass:

```r
fit_grm <- drmTMB(
  bf(seed_mass ~ temperature + relmat(1 | line, K = G), sigma ~ temperature),
  data = dat,
  family = gaussian()
)
```

Here `G` can be a correlation matrix. `fit_grm$sdpars$mu` reports the estimated
line-level SD that multiplies that known relatedness matrix. This is not known
sampling error; it is latent among-line structure left after the fixed effects.

Now suppose `Q_river` is a positive-definite precision matrix for sites on a
river network, built outside `drmTMB` from an adjacency or flow model. The
scientific question is whether nearby or connected sites have similar latent
condition deviations:

```r
fit_river <- drmTMB(
  bf(condition ~ treatment + relmat(1 | reach, Q = Q_river), sigma ~ 1),
  data = dat,
  family = gaussian()
)
```

This is a good `Q` example because the natural object is already an inverse
covariance. The model estimates the SD of the latent site field while the known
precision controls the pattern of similarity across reaches.

## What is fitted today

| Question | Syntax | Status |
| --- | --- | --- |
| Does one Gaussian response have deviations structured by a user-supplied latent covariance or precision matrix? | `relmat(1 | id, K = K)` or `relmat(1 | id, Q = Q)` in `mu` | Fitted first slice. `K` is a covariance matrix; `Q` is a precision matrix. |
| Does one numeric fixed-effect slope also have known-matrix deviations? | `relmat(1 + x | id, K = K)` or `relmat(1 + x | id, Q = Q)` in `mu` | Fitted first one-slope slice with independent intercept and slope fields. Multiple structured slopes and intercept-slope correlations remain planned. |
| Does a one-response Gaussian model have known-matrix residual-scale deviations? | `relmat(1 | id, Q = Q)` in `sigma`, alone or matching the `mu` term | Fitted first residual-scale intercept slice. Matching univariate `mu` and `sigma` `relmat()` terms estimate one known-matrix mean-scale correlation; residual-scale structured slopes remain planned. |
| Do two Gaussian response means share a known-matrix latent correlation? | matching labelled `relmat(1 | p | id, K = K)` or `relmat(1 | p | id, Q = Q)` terms in `mu1` and `mu2` | Fitted first q=2 location-location slice. `corpairs(level = "relmat")` reports the latent relatedness correlation separately from residual `rho12`. |
| Does the known-matrix layer link means and residual scales across two responses? | the same labelled `relmat(1 | p | id, Q = Q)` term in `mu1`, `mu2`, `sigma1`, and `sigma2` | Fitted first q=4 location-scale slice. `corpairs(level = "relmat")` reports six constant latent relatedness correlations; profile intervals for those derived correlations remain unavailable. |
| Does the known-matrix layer change multiple slopes or a predictor-dependent correlation? | examples such as `relmat(1 + x + z | id, K = K)` or `relmat()` `corpair()` formulas | Planned. These routes need matrix validation, diagnostics, recovery tests, interval targets, and examples before use. |

Use `animal()` when the matrix is additive relatedness for individual animal
models. Use `phylo()` when the matrix comes from a species tree and the tree
route is available. Use `spatial()` when the structure is induced by sampling
coordinates. Use `relmat()` for validated latent matrices that do not belong to
those named routes.

## Start with the smallest useful model

For one response, start with the fitted location-intercept route:

```r
fit_relmat <- drmTMB(
  y ~ treatment + relmat(1 | line, K = K),
  data = dat,
  family = gaussian()
)
```

For two responses, use the same `relmat()` block label in `mu1` and `mu2`:

```r
fit_relmat_q2 <- drmTMB(
  bf(
    mu1 = trait1 ~ treatment +
      relmat(1 | p | line, Q = Q),
    mu2 = trait2 ~ treatment +
      relmat(1 | p | line, Q = Q)
  ),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_relmat_q2, level = "relmat")
rho12(fit_relmat_q2)
```

`corpairs()` reports the fitted latent known-matrix correlation among line or
unit-level location deviations. `rho12()` reports the residual correlation
between responses after fixed effects and random effects have been included.

For reports, do not leave the q=2 known-matrix correlation as a lone dot when
interval evidence is available. Request interval provenance first, then plot the
explicit table:

```r
relmat_pairs <- corpairs(fit_relmat_q2, level = "relmat", conf.int = TRUE)
plot_corpairs(relmat_pairs)
```

That plot should be read as a latent known-matrix correlation display, not as a
residual `rho12` plot. The q=4 rows below are different: current q=4 `relmat()`
correlations are derived rows, so their tables may expose point estimates and
`conf.status` without profile Confidence Eyes.

When the scientific question is whether known-matrix deviations in means and
residual scales covary, use the same labelled `relmat()` term in all four
bivariate endpoint formulas:

```r
fit_relmat_q4 <- drmTMB(
  bf(
    mu1 = trait1 ~ treatment +
      relmat(1 | p | line, Q = Q),
    mu2 = trait2 ~ treatment +
      relmat(1 | p | line, Q = Q),
    sigma1 = ~ treatment +
      relmat(1 | p | line, Q = Q),
    sigma2 = ~ treatment +
      relmat(1 | p | line, Q = Q)
  ),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_relmat_q4, level = "relmat")
```

The q=4 path is constant across levels and responses: it estimates four
known-matrix endpoint standard deviations and six latent correlations. It is
not a direct-SD model and it does not make the residual correlation `rho12`
known-matrix structured.

## What to inspect

After fitting, inspect the relatedness layer before interpreting it:

| Output | Use |
| --- | --- |
| `check_drm(fit)` | Confirm the `relmat()` layer was recognized and review matrix/replication diagnostics. |
| `fit$sdpars$mu` | Read the fitted known-matrix location SD. |
| `ranef(fit, "relmat_mu")` | Inspect conditional deviations on the fitted relatedness layer. |
| `summary(fit)$covariance` | Check how the `relmat()` SD or q=2 correlation is reported beside other covariance layers. |
| `profile_targets(fit)` | See which relatedness SD or q=2 correlation targets can be profiled directly; q=4 correlations are derived-only in the first slice. |
| `corpairs(fit, level = "relmat")` | Read the fitted q=2 relatedness mean-mean correlation, or the six q=4 relatedness endpoint correlations when the model has the all-four block. |

## Rendered checks

The examples below keep the matrix, standard deviations, and correlation row in
separate visual grammars. The known matrix is raw input structure, not
uncertainty. The SD display uses response-scale estimates with Wald intervals.
The q=2 relatedness correlation has profile interval support, so it uses the
Confidence Eye and a dotted zero line.

```{r relmat-guide-fit}
relmat_example <- simulate_relmat_guide_data()
relmat_dat <- relmat_example$data
K <- relmat_example$K
Q <- relmat_example$Q

fit_relmat <- drmTMB(
  bf(
    seed_mass ~ temperature + treatment + relmat(1 | line, Q = Q),
    sigma ~ 1
  ),
  family = gaussian(),
  data = relmat_dat
)
```

```{r relmat-known-matrix, fig.width = 5.3, fig.height = 4.8, fig.cap = "Known relatedness matrix used by the `relmat()` example; this heatmap shows the supplied latent structure, not model-estimated uncertainty.", fig.alt = "Heatmap of the known relatedness matrix for the relmat example. Values are highest on the diagonal and fade as line identifiers are farther apart in the simulated ordering."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  relmat_matrix <- relatedness_heatmap_data(K)

  ggplot2::ggplot(
    relmat_matrix,
    ggplot2::aes(column, row, fill = relatedness)
  ) +
    ggplot2::geom_tile() +
    ggplot2::coord_equal() +
    ggplot2::scale_fill_gradientn(
      colours = c("#F7FCF5", "#C7E9C0", "#74C476", "#006D2C"),
      name = "Known\nrelatedness"
    ) +
    relmat_guide_theme() +
    ggplot2::theme(
      axis.text = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank()
    ) +
    ggplot2::labs(
      title = "Known-matrix input structure",
      subtitle = "Validated relatedness among experimental lines",
      x = "Line",
      y = "Line"
    )
}
```

```{r relmat-sd-figure, fig.width = 6.6, fig.height = 3.4, fig.cap = "Fitted residual `sigma` and known-matrix location SD from a univariate Gaussian `relmat()` model; points and bars are response-scale estimates with 95% Wald intervals from `confint(fit_relmat, parm = \"variance_components\")`.", fig.alt = "Horizontal interval display comparing residual sigma and the known-matrix relmat location standard deviation. Both rows have hollow point estimates and finite 95 percent Wald intervals."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  relmat_sd <- confint(fit_relmat, parm = "variance_components")
  relmat_sd$estimate <- c(
    unname(exp(coef(fit_relmat, "sigma")["(Intercept)"])),
    unname(fit_relmat$sdpars$mu["relmat(1 | line)"])
  )
  relmat_sd$label <- c("Residual\nsigma", "relmat\nlocation SD")

  ggplot2::ggplot(relmat_sd, ggplot2::aes(y = label)) +
    ggplot2::geom_vline(
      xintercept = 0,
      linewidth = 0.45,
      linetype = "dashed",
      colour = "grey55"
    ) +
    ggplot2::geom_errorbar(
      ggplot2::aes(xmin = lower, xmax = upper),
      width = 0,
      linewidth = 1.1,
      colour = "#009E73"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = estimate),
      shape = 21,
      size = 3.5,
      stroke = 1,
      fill = "white",
      colour = "#009E73"
    ) +
    ggplot2::scale_x_continuous(
      expand = ggplot2::expansion(mult = c(0.02, 0.05))
    ) +
    relmat_eye_theme() +
    ggplot2::labs(
      title = "Known-matrix SD is separate from residual sigma",
      subtitle = "Bars are 95% Wald intervals on the response scale",
      x = "Fitted standard deviation",
      y = NULL
    )
}
```

For two response means, the known-matrix row is a fitted latent correlation
among line-level deviations. It is not residual `rho12` and it is not a raw
correlation among observations.

```{r relmat-q2-correlation-fit}
relmat_q2_example <- simulate_relmat_q2_guide_data()
relmat_q2_dat <- relmat_q2_example$data
Q <- relmat_q2_example$Q

fit_relmat_q2_example <- drmTMB(
  bf(
    mu1 = seed_mass ~ age + sex +
      relmat(1 | p | line, Q = Q),
    mu2 = plant_height ~ age + sex +
      relmat(1 | p | line, Q = Q),
    sigma1 = ~ sex,
    sigma2 = ~ sex,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = relmat_q2_dat
)

relmat_q2_pairs <- corpairs(
  fit_relmat_q2_example,
  level = "relmat",
  conf.int = TRUE,
  ystep = 0.5,
  ytol = 2
)
relmat_q2_pairs
```

```{r relmat-q2-confidence-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "`relmat()` q=2 location-location correlation from `corpairs()`; the Confidence Eye is a 95% profile interval on the latent known-matrix correlation scale and the dotted vertical line marks zero correlation.", fig.alt = "Single-row Confidence Eye for the known-matrix relmat mean-mean correlation. The eye spans the 95 percent profile interval and the hollow point estimate is to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  relmat_q2_display <- relmat_q2_pairs
  relmat_q2_display$display_label <- "relmat\nmu1-mu2"

  plot_corpairs(
    relmat_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("relmat" = "#009E73")) +
    ggplot2::scale_fill_manual(values = c("relmat" = "#009E73")) +
    relmat_eye_theme() +
    ggplot2::labs(
      title = "Known-matrix latent mean correlation",
      subtitle = "Eye is a 95% profile interval; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

## Boundaries

The following `relmat()` routes remain planned:

- multiple structured slopes such as `relmat(1 + x + z | id, K = K)`;
- known-matrix intercept-slope correlations;
- residual-scale structured slopes;
- predictor-dependent `relmat()` `corpair()` regressions;
- generic direct-SD grammar for known-matrix standard deviations;
- non-Gaussian known-matrix structured effects.

Known sampling covariance is a different model layer. If the matrix describes
known observation-level uncertainty for effect-size estimates, use
`meta_V(V = V)` and the [meta-analysis documentation](meta-analysis.html)
instead of `relmat()`. The older `meta_known_V(V = V)` spelling is deprecated
but remains a compatibility alias, not a separate relatedness route.

Use the [structural-dependence overview](structural-dependence.html) when you
are choosing among `animal()`, `phylo()`, `spatial()`, and `relmat()`. Use the
detailed [structural-dependence tutorial](phylogenetic-spatial.html) when you
need the current fitted examples, equations, and broader parity ladder.
