---
title: "Robust continuous responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust continuous responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Continuous biological and environmental measurements, such as seedling height
growth in centimetres, often contain a few large residuals. They may be real
rare events, short-term stress responses, field measurement problems, or
individuals that experienced unmeasured conditions. A Gaussian location-scale
model can still be useful, but a Student-t model asks a slightly different
question: are the location and scale patterns stable when the likelihood allows
heavier tails?

This article introduces the first implemented robust continuous family in
`drmTMB`: fixed-effect Student-t location-scale-shape regression.
Here location means the expected response parameter `mu`, scale means the
positive Student-t core scale `sigma`, and shape means the tail parameter
`nu`. This one-response model has no coscale term; in `drmTMB`, coscale means
modelling residual correlation, such as bivariate `rho12`.

## Model equation and R syntax

The Student-t model has three distributional parameters:

\[
\begin{aligned}
\text{growth}_i \mid \mu_i, \sigma_i, \nu_i
  &\sim \operatorname{Student\text{-}t}(\mu_i, \sigma_i, \nu_i),\\
\mu_i
  &= \beta_0 + \beta_1 \text{dry}_i,\\
\log(\sigma_i)
  &= \gamma_0 + \gamma_1 \text{dry}_i,\\
\nu_i
  &= 2 + \exp(\delta_0).
\end{aligned}
\]

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(growth ~ drought, sigma ~ drought, nu ~ 1),
  family = student(),
  data = seedlings
)
```

Here `mu` is the expected response, `sigma` is the Student-t scale parameter,
and `nu` is the tail-shape parameter. When `nu > 2`, the residual standard
deviation is `sigma * sqrt(nu / (nu - 2))`, so `sigma` should not be read as
the exact residual SD unless `nu` is large. Smaller `nu` means heavier tails.
Large `nu` means the Student-t likelihood is close to Gaussian. In `drmTMB`,

\[
\nu_i = 2 + \exp(\eta_{\nu i}).
\]

so fitted `nu` values stay above 2. This keeps the fitted Student-t
distribution in the finite-variance region.

## A seedling growth example

Suppose seedlings are grown under ambient and dry conditions, and growth is
measured as height increase in centimetres. We expect drought to reduce average
growth, and we also allow drought to change scale among seedlings.

```{r}
library(drmTMB)

set.seed(101)
n <- 180
seedlings <- data.frame(
  drought = factor(rep(c("ambient", "dry"), each = n / 2))
)
dry <- as.numeric(seedlings$drought == "dry")
mu <- 1.2 - 0.45 * dry
sigma <- exp(-1 + 0.35 * dry)
seedlings$growth <- mu + sigma * rt(n, df = 5)
```

In the equation above, `dry_i` is the model-matrix indicator for the dry
treatment level. It corresponds to the `droughtdry` coefficient printed by R.

A Gaussian location-scale model is a useful baseline:

```{r}
fit_gaussian <- drmTMB(
  bf(growth ~ drought, sigma ~ drought),
  family = gaussian(),
  data = seedlings
)
```

The robust version keeps the same location and scale formulas, then adds a
formula for `nu`:

```{r}
fit_student <- drmTMB(
  bf(growth ~ drought, sigma ~ drought, nu ~ 1),
  family = student(),
  data = seedlings
)
```

The extra line `nu ~ 1` says that the tail-shape parameter is estimated as a
constant across observations. Predictor-varying fixed-effect `nu` formulas are
also supported. For example, `nu ~ drought` asks whether the residual tails are
heavier in the dry treatment than in the ambient treatment after `mu` and
`sigma` have been modelled. This first tutorial keeps `nu` constant so the
comparison focuses on location, scale, and residual-tail assumptions.

## Check the fitted robust model

Run `check_drm()` before interpreting the Student-t coefficients:

```{r}
student_checks <- check_drm(fit_student)
student_checks
student_checks[
  student_checks$check == "student_nu",
  c("status", "value", "message")
]
```

The `student_nu` row inspects fitted response-scale `nu` values. A near-boundary
warning means the tail-shape parameter is close to 2, where the finite-variance
constraint matters. A large-`nu` note means the Student-t fit may be close to a
Gaussian fit. If `nu` is large, report that the Student-t fit is close to
Gaussian and check whether the Gaussian and Student-t `mu` and `sigma`
conclusions differ. If `nu` is near the boundary, inspect influential
observations, compare Gaussian and Student-t conclusions, and report that the
fitted tail-shape parameter is near the lower bound.

Keep the `student_nu` status beside AIC, coefficient, and simulation-summary
tables. An `ok` row supports reading the fitted `nu` as an ordinary
finite-variance Student-t shape estimate. A `note` or `warning` should travel
with the result, because it changes how a reader should interpret the tail
assumption even when the optimizer converged.

## Interpret coefficients by parameter

Location and scale coefficients use the same formula grammar as Gaussian
location-scale models:

```{r}
coef(fit_student, "mu")
coef(fit_student, "sigma")
coef(fit_student, "nu")
```

The `mu` coefficient for `droughtdry` estimates the drought difference in
expected growth. The `sigma` coefficient for `droughtdry` is on the log
Student-t scale-parameter scale: a positive value means the fitted scale is
larger under drought. The `nu` coefficient is on the link scale for
`nu = 2 + exp(eta_nu)`. Use prediction to read `nu` on the response scale:

```{r}
head(predict(fit_student, dpar = "nu"))
```

## Compare with the Gaussian model

For this simulated example, the two models share `mu` and `sigma` formulas:
does drought reduce growth, and does it change scale? They differ in the
residual-tail assumption.

```{r}
AIC(fit_gaussian, fit_student)
```

AIC is only one summary. The more important habit is to inspect whether the
scientific conclusions about `mu` and `sigma` are sensitive to the residual
tail assumption.

```{r}
coef(fit_gaussian, "mu")
coef(fit_student, "mu")

coef(fit_gaussian, "sigma")
coef(fit_student, "sigma")
```

```{r robust-student-tail-figure, eval=requireNamespace("ggplot2", quietly = TRUE), fig.width=7.2, fig.height=4.4, fig.cap="Robust-model check for the seedling example. Faint points show observed growth values; overlaid points compare fitted Gaussian and Student-t expected growth by drought treatment. No interval bars are drawn because this fixture is a raw-data and fitted-point comparison, not an interval summary.", fig.alt="Jittered point plot of seedling growth by ambient and dry drought treatments. Faint raw observations show heavy-tailed residual variation, and overlaid points compare Gaussian and Student-t fitted expected growth for each treatment."}
library(ggplot2)

student_plot_grid <- data.frame(
  drought = factor(c("ambient", "dry"), levels = levels(seedlings$drought))
)
student_plot_means <- rbind(
  data.frame(
    student_plot_grid,
    model = "Gaussian",
    fitted_mu = predict(fit_gaussian, newdata = student_plot_grid, dpar = "mu")
  ),
  data.frame(
    student_plot_grid,
    model = "Student-t",
    fitted_mu = predict(fit_student, newdata = student_plot_grid, dpar = "mu")
  )
)

ggplot(seedlings, aes(drought, growth, colour = drought)) +
  geom_jitter(width = 0.12, height = 0, alpha = 0.22, size = 1.1) +
  geom_point(
    data = student_plot_means,
    aes(y = fitted_mu, shape = model),
    position = position_dodge(width = 0.35),
    size = 3.2,
    stroke = 1.1
  ) +
  scale_colour_manual(values = c("ambient" = "#0072B2", "dry" = "#D55E00")) +
  scale_shape_manual(values = c("Gaussian" = 16, "Student-t" = 1)) +
  labs(
    title = "Robust fits should be checked against the raw tails",
    subtitle = "Points are observed growth values; symbols are fitted expected growth",
    x = "Drought treatment",
    y = "Growth",
    colour = "Treatment",
    shape = "Model"
  ) +
  guides(colour = "none") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30")
  )
```

If the `mu` and `sigma` conclusions change strongly, the large residuals are
not a side detail. They are part of the distributional story and should be
reported.

## Current boundary

The implemented Student-t path is intentionally narrow:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2, nu ~ x3),
  family = student(),
  data = dat
)
```

Random effects, known sampling covariance, phylogenetic terms, spatial terms,
and bivariate Student-t models are later phases. Start with this fixed-effect
path when you need a robust continuous comparison to a Gaussian
location-scale model.

Skew-normal residual asymmetry is a separate fitted fixed-effect family:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2, nu ~ x3),
  family = skew_normal(),
  data = dat
)
```

Use this model when the question is residual asymmetry rather than heavy tails.
The fitted first slice has density, Gaussian-limit, positive- and negative-skew
recovery, false-positive, interval-visibility, and Phase 18 smoke/grid artifact
tests. Random effects, known covariance, structured effects, bivariate
skew-normal models, residual `rho12`, aliases such as `skew ~ x`, and
ID-level skewness such as future `skew(id) ~ x` remain later routes. Gaussian,
Student-t, and skew-normal fits answer different residual-distribution
questions, so compare them as sensitivity models rather than aliases.
