---
title: "Robust continuous responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust continuous responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Continuous biological and environmental measurements, such as seedling height
growth in centimetres, often contain a few large residuals. They may be real
rare events, short-term stress responses, field measurement problems, or
individuals that experienced unmeasured conditions. A Gaussian location-scale
model can still be useful, but a Student-t model asks a slightly different
question: are the location and scale patterns stable when the likelihood allows
heavier tails?

This article introduces the first implemented robust continuous family in
`drmTMB`: fixed-effect Student-t location-scale-shape regression.

## Model equation and R syntax

The Student-t model has three distributional parameters:

```text
growth_i | mu_i, sigma_i, nu_i ~ Student-t(mu_i, sigma_i, nu_i)
mu_i = beta_0 + beta_1 dry_i
log(sigma_i) = gamma_0 + gamma_1 dry_i
nu_i = 2 + exp(delta_0)
```

The matching `drmTMB` syntax is:

```r
drmTMB(
  bf(growth ~ drought, sigma ~ drought, nu ~ 1),
  family = student(),
  data = seedlings
)
```

Here `mu` is the expected response, `sigma` is the Student-t scale parameter,
and `nu` is the tail-shape parameter. When `nu > 2`, the residual standard
deviation is `sigma * sqrt(nu / (nu - 2))`, so `sigma` should not be read as
the exact residual SD unless `nu` is large. Smaller `nu` means heavier tails.
Large `nu` means the Student-t likelihood is close to Gaussian. In `drmTMB`,

```text
nu_i = 2 + exp(eta_nu_i)
```

so fitted `nu` values stay above 2. This keeps the fitted Student-t
distribution in the finite-variance region.

## A seedling growth example

Suppose seedlings are grown under ambient and dry conditions, and growth is
measured as height increase in centimetres. We expect drought to reduce average
growth, and we also allow drought to change scale among seedlings.

```{r}
library(drmTMB)

set.seed(101)
n <- 180
seedlings <- data.frame(
  drought = factor(rep(c("ambient", "dry"), each = n / 2))
)
dry <- as.numeric(seedlings$drought == "dry")
mu <- 1.2 - 0.45 * dry
sigma <- exp(-1 + 0.35 * dry)
seedlings$growth <- mu + sigma * rt(n, df = 5)
```

In the equation above, `dry_i` is the model-matrix indicator for the dry
treatment level. It corresponds to the `droughtdry` coefficient printed by R.

A Gaussian location-scale model is a useful baseline:

```{r}
fit_gaussian <- drmTMB(
  bf(growth ~ drought, sigma ~ drought),
  family = gaussian(),
  data = seedlings
)
```

The robust version keeps the same location and scale formulas, then adds a
formula for `nu`:

```{r}
fit_student <- drmTMB(
  bf(growth ~ drought, sigma ~ drought, nu ~ 1),
  family = student(),
  data = seedlings
)
```

The extra line `nu ~ 1` says that tail weight is estimated as a constant across
observations. Predictor-varying fixed-effect `nu` formulas are also supported;
this first tutorial keeps `nu` constant so the comparison focuses on location,
scale, and residual-tail assumptions.

## Check the fitted robust model

Run `check_drm()` before interpreting the Student-t coefficients:

```{r}
check_drm(fit_student)
```

The `student_nu` row inspects fitted response-scale `nu` values. A near-boundary
warning means the heavy-tail parameter is close to 2, where the finite-variance
constraint matters. A large-`nu` note means the Student-t fit may be close to a
Gaussian fit. If `nu` is near the boundary, inspect influential observations,
compare Gaussian and Student-t conclusions, and report that the fitted tail
parameter is near the lower bound.

## Interpret coefficients by parameter

Location and scale coefficients use the same formula grammar as Gaussian
location-scale models:

```{r}
coef(fit_student, "mu")
coef(fit_student, "sigma")
coef(fit_student, "nu")
```

The `mu` coefficient for `droughtdry` estimates the drought difference in
expected growth. The `sigma` coefficient for `droughtdry` is on the log
Student-t scale-parameter scale: a positive value means the fitted scale is
larger under drought. The `nu` coefficient is on the link scale for
`nu = 2 + exp(eta_nu)`. Use prediction to read `nu` on the response scale:

```{r}
head(predict(fit_student, dpar = "nu"))
```

## Compare with the Gaussian model

For this simulated example, the two models share `mu` and `sigma` formulas:
does drought reduce growth, and does it change scale? They differ in the
residual-tail assumption.

```{r}
AIC(fit_gaussian, fit_student)
```

AIC is only one summary. The more important habit is to inspect whether the
scientific conclusions about `mu` and `sigma` are sensitive to the residual
tail assumption.

```{r}
coef(fit_gaussian, "mu")
coef(fit_student, "mu")

coef(fit_gaussian, "sigma")
coef(fit_student, "sigma")
```

If the `mu` and `sigma` conclusions change strongly, the large residuals are
not a side detail. They are part of the distributional story and should be
reported.

## Current boundary

The implemented Student-t path is intentionally narrow:

```r
drmTMB(
  bf(y ~ x1, sigma ~ x2, nu ~ x3),
  family = student(),
  data = dat
)
```

Random effects, known sampling covariance, phylogenetic terms, spatial terms,
and bivariate Student-t models are later phases. Start with this fixed-effect
path when you need a robust continuous comparison to a Gaussian
location-scale model.
