---
title: "Simulation plot grammar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation plot grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This article shows the plot grammar that Phase 18 simulation reports should use
before broad simulation grids are advertised as evidence. It is a display
contract, not a final result report. The small tables below are illustrative
fixtures that mimic the columns a real simulation summary should carry.

Simulation figures should answer the same reader questions across continuous,
proportion, count, and meta-analysis examples:

- Does the estimator recover the truth without important bias?
- How large is the error, measured by RMSE?
- Do interval procedures cover at the advertised rate?
- Does power increase under the intended effect?
- How often do fits converge, hit boundaries, warn, or fail?
- How much runtime does each surface require?

```{r setup}
library(ggplot2)

theme_sim_grammar <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey30", lineheight = 1.05),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      plot.margin = margin(6, 10, 12, 6)
    )
}

surface_palette <- c(
  "Gaussian location-scale" = "#0072B2",
  "Proportion" = "#009E73",
  "Count" = "#D55E00",
  "Meta-analysis" = "#CC79A7"
)

surface_levels <- names(surface_palette)
surface_facet_labels <- c(
  "Gaussian location-scale" = "Gaussian\nlocation-scale",
  "Proportion" = "Proportion",
  "Count" = "Count",
  "Meta-analysis" = "Meta-analysis"
)
estimand_display_labels <- c(
  "beta_x" = "beta_x",
  "sigma" = "sigma",
  "sd_group" = "sd(group)",
  "rho12" = "rho12"
)
```

## Bias and RMSE

Bias and RMSE should be plotted beside the estimand. A single average across
all parameters hides the distinction between fixed effects, residual scale,
random-effect SDs, and residual correlation `rho12`.

```{r bias-rmse-display, fig.width=8.2, fig.height=5.1, fig.alt="Two simulation accuracy plots faceted by simulation surface: an illustrative raincloud-style bias display with pseudo-replicate errors, mean bias points, and MCSE intervals, followed by an aggregate RMSE display with MCSE intervals for four estimands."}
accuracy_summary <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  bias = c(
    0.01, -0.02, 0.03, 0.00,
    0.00, 0.01, 0.02, NA,
    0.02, NA, 0.06, NA,
    -0.01, 0.02, NA, 0.03
  ),
  rmse = c(
    0.09, 0.12, 0.18, 0.16,
    0.07, 0.09, 0.15, NA,
    0.12, NA, 0.22, NA,
    0.08, 0.11, NA, 0.14
  ),
  n_replicate = c(
    300, 300, 260, 280,
    300, 300, 250, NA,
    280, NA, 220, NA,
    300, 300, NA, 260
  )
)
accuracy_summary$surface <- factor(
  accuracy_summary$surface,
  levels = surface_levels
)
accuracy_summary$error_sd <- sqrt(
  pmax(accuracy_summary$rmse^2 - accuracy_summary$bias^2, 0)
)
accuracy_summary$bias_mcse <- with(
  accuracy_summary,
  error_sd / sqrt(n_replicate)
)
accuracy_summary$rmse_mcse <- with(
  accuracy_summary,
  rmse / sqrt(2 * n_replicate)
)
accuracy_summary$bias_low <- accuracy_summary$bias -
  1.96 * accuracy_summary$bias_mcse
accuracy_summary$bias_high <- accuracy_summary$bias +
  1.96 * accuracy_summary$bias_mcse
accuracy_summary$rmse_low <- pmax(
  0,
  accuracy_summary$rmse - 1.96 * accuracy_summary$rmse_mcse
)
accuracy_summary$rmse_high <- accuracy_summary$rmse +
  1.96 * accuracy_summary$rmse_mcse
accuracy_summary$estimand <- factor(
  accuracy_summary$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)

display_replicates <- 49
accuracy_replicates <- do.call(
  rbind,
  lapply(seq_len(nrow(accuracy_summary)), function(i) {
    row <- accuracy_summary[i, ]
    if (is.na(row$bias) || is.na(row$rmse)) {
      return(NULL)
    }
    data.frame(
      surface = row$surface,
      estimand = row$estimand,
      error = row$bias + qnorm(ppoints(display_replicates)) * row$error_sd
    )
  })
)
accuracy_replicates$surface <- factor(
  accuracy_replicates$surface,
  levels = surface_levels
)
accuracy_replicates$estimand <- factor(
  accuracy_replicates$estimand,
  levels = levels(accuracy_summary$estimand)
)
set.seed(20260519)
ggplot(accuracy_replicates, aes(error, estimand)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey55") +
  geom_violin(
    aes(
      fill = surface
    ),
    width = 0.62,
    alpha = 0.16,
    colour = NA,
    scale = "width",
    trim = FALSE
  ) +
  geom_point(
    aes(colour = surface),
    position = position_jitter(height = 0.05, width = 0),
    alpha = 0.22,
    size = 0.65,
    show.legend = FALSE
  ) +
  geom_segment(
    data = accuracy_summary,
    aes(
      x = bias_low,
      xend = bias_high,
      y = estimand,
      yend = estimand,
      colour = surface
    ),
    linewidth = 0.85,
    na.rm = TRUE
  ) +
  geom_point(
    data = accuracy_summary,
    aes(bias, estimand, colour = surface),
    size = 2.4,
    na.rm = TRUE
  ) +
  facet_wrap(
    ~surface,
    nrow = 1,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  scale_fill_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "Bias displays should show replicate-level error",
    subtitle = paste(
      "Clouds and faint dots are illustrative pseudo-replicate errors;",
      "large points and bars are mean bias with 95% MCSE"
    ),
    x = "Estimate minus truth",
    y = NULL,
    colour = "Surface"
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  guides(colour = "none", fill = "none") +
  theme_sim_grammar()

ggplot(accuracy_summary, aes(rmse, estimand, colour = surface)) +
  geom_errorbarh(
    aes(
      xmin = rmse_low,
      xmax = rmse_high
    ),
    height = 0.15,
    linewidth = 0.9,
    alpha = 0.9,
    na.rm = TRUE
  ) +
  geom_point(size = 2.4, na.rm = TRUE) +
  facet_wrap(
    ~surface,
    nrow = 2,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "RMSE needs its own uncertainty display",
    subtitle = "Facets are surfaces; points and bars are RMSE and 95% MCSE intervals for each estimand",
    x = "Root mean-square error",
    y = NULL,
    colour = "Surface"
  ) +
  scale_x_continuous(
    breaks = c(0, 0.10, 0.20),
    limits = c(0, 0.31),
    expand = expansion(mult = c(0.01, 0.04))
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  guides(colour = "none") +
  theme_sim_grammar()
```

The missing rows are part of the message. A count pilot should not quietly fill
`rho12` with zeros, and a proportion example should not invent a random-effect
SD if that surface was not fitted. Real bias reports should replace the fixture
quantiles above with one row per replicate estimate or error, then overplot the
aggregate mean and MCSE interval produced by the Phase 18 summary helpers. RMSE
should remain an aggregate root-mean-square summary with its own MCSE interval,
not a cloud whose center could be mistaken for a mean absolute error.

## Coverage and Power

Coverage and power are proportions, so every point should carry or clearly
reference the number of replicates that contributed to it and a Monte Carlo
uncertainty interval. Real reports should plot stored replicate rows or stored
replicate-block summaries when they are available; the fixture below illustrates
that grammar with block-level proportions plus a binomial MCSE approximation for
the aggregate.

```{r coverage-power-display, fig.width=8.6, fig.height=5.2, fig.alt="Faceted point plot showing illustrative coverage and power with Monte Carlo uncertainty intervals for four simulation surfaces, with one panel per surface so intervals align with their points."}
interval_summary <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  coverage = c(
    0.95, 0.93, 0.91, 0.94,
    0.94, 0.92, 0.90, NA,
    0.92, NA, 0.86, NA,
    0.95, 0.94, NA, 0.92
  ),
  power = c(
    0.82, 0.64, 0.58, 0.70,
    0.78, 0.60, 0.52, NA,
    0.74, NA, 0.46, NA,
    0.80, 0.61, NA, 0.66
  ),
  n_interval = c(
    280, 280, 240, 260,
    280, 280, 230, NA,
    250, NA, 200, NA,
    280, 280, NA, 240
  )
)
interval_summary$surface <- factor(
  interval_summary$surface,
  levels = surface_levels
)
interval_long <- rbind(
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Coverage",
    value = interval_summary$coverage,
    target = 0.95
  ),
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Power",
    value = interval_summary$power,
    target = 0.80
  )
)
interval_long$mcse <- sqrt(
  interval_long$value * (1 - interval_long$value) / interval_long$n_interval
)
interval_long$lower <- pmax(0, interval_long$value - 1.96 * interval_long$mcse)
interval_long$upper <- pmin(1, interval_long$value + 1.96 * interval_long$mcse)
interval_long$measure <- factor(
  interval_long$measure,
  levels = c("Coverage", "Power")
)
interval_long$estimand <- factor(
  interval_long$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)
interval_long$missing_x <- ifelse(
  interval_long$measure == "Coverage",
  0.405,
  0.405
)
interval_target <- unique(interval_long[c("measure", "target")])
interval_present <- interval_long[!is.na(interval_long$value), , drop = FALSE]
interval_missing <- interval_long[is.na(interval_long$value), , drop = FALSE]
make_interval_blocks <- function(row, n_block = 18) {
  block_size <- max(8L, floor(row$n_interval / n_block))
  block_proportion <- qbinom(
    ppoints(n_block),
    size = block_size,
    prob = row$value
  ) / block_size
  data.frame(
    surface = row$surface,
    estimand = row$estimand,
    measure = row$measure,
    block_proportion = block_proportion
  )
}
interval_blocks <- do.call(
  rbind,
  lapply(seq_len(nrow(interval_present)), function(i) {
    make_interval_blocks(interval_present[i, ])
  })
)

ggplot(interval_present, aes(value, estimand, colour = surface)) +
  geom_vline(
    data = interval_target,
    aes(xintercept = target),
    linetype = "dashed",
    colour = "grey55",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = interval_blocks,
    aes(block_proportion, estimand, colour = surface),
    position = position_jitter(height = 0.07, width = 0),
    alpha = 0.18,
    size = 0.70,
    inherit.aes = FALSE
  ) +
  geom_segment(
    aes(x = lower, xend = upper, yend = estimand),
    linewidth = 0.4,
    na.rm = TRUE
  ) +
  geom_point(size = 2.1, na.rm = TRUE) +
  geom_text(
    aes(label = paste0("n=", n_interval)),
    nudge_y = -0.25,
    colour = "grey40",
    size = 1.7,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_text(
    data = interval_missing,
    aes(x = missing_x, y = estimand, label = "not\ntargeted"),
    colour = "grey55",
    size = 2.15,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_grid(
    measure ~ surface,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, guide = "none") +
  scale_x_continuous(
    limits = c(0.38, 1),
    breaks = c(0.4, 0.6, 0.8, 0.95),
    labels = c("0.40", "0.60", "0.80", "0.95")
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  labs(
    title = "Interval coverage and power need Monte Carlo uncertainty",
    subtitle = "Faint dots illustrate replicate-block proportions; large points carry n and 95% binomial MCSE",
    x = "Proportion",
    y = NULL,
    colour = "Surface"
  ) +
  theme_sim_grammar()
```

The plot should make pilot status obvious. With few replicates, a coverage
estimate near 0.95 can still be too noisy for a release claim.

## Convergence and Runtime

Convergence, positive-definite Hessian status, and runtime should sit beside
accuracy. A surface with low bias but frequent warnings or high runtime is not
ready for broad claims.

```{r convergence-runtime-display, fig.alt="Faceted plot showing illustrative convergence rate, Hessian rate, median runtime, and high runtime quantile by simulation surface."}
run_summary <- data.frame(
  surface = surface_levels,
  convergence_rate = c(0.98, 0.96, 0.90, 0.95),
  pdhess_rate = c(0.96, 0.94, 0.86, 0.93),
  elapsed_median = c(0.35, 0.28, 0.52, 0.44),
  elapsed_p90 = c(0.72, 0.58, 1.35, 0.96)
)
run_summary$surface <- factor(run_summary$surface, levels = surface_levels)
run_long <- rbind(
  data.frame(
    surface = run_summary$surface,
    measure = "Converged",
    value = run_summary$convergence_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "pdHess",
    value = run_summary$pdhess_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "Median runtime",
    value = run_summary$elapsed_median
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "90th percentile runtime",
    value = run_summary$elapsed_p90
  )
)
run_long$panel <- ifelse(
  grepl("runtime", run_long$measure),
  "Runtime (seconds)",
  "Fit status proportion"
)
run_long$measure <- factor(
  run_long$measure,
  levels = c(
    "Converged",
    "pdHess",
    "Median runtime",
    "90th percentile runtime"
  )
)

ggplot(run_long, aes(value, surface, fill = measure)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  facet_wrap(~panel, scales = "free_x") +
  scale_fill_manual(values = c(
    "Converged" = "#0072B2",
    "pdHess" = "#56B4E9",
    "Median runtime" = "#009E73",
    "90th percentile runtime" = "#E69F00"
  )) +
  labs(
    title = "Convergence and runtime belong beside accuracy",
    subtitle = "Fast successful fits and stable uncertainty are separate checks",
    x = "Value",
    y = NULL,
    fill = "Measure"
  ) +
  theme_sim_grammar()
```

Runtime summaries should report medians and high quantiles, not only means,
because a few difficult cells can dominate overnight simulation planning.

## Failure Ledger

Failures and warnings are data. They should be summarised by surface and
condition instead of filtered away before plotting.

```{r failure-ledger-display, fig.alt="Stacked bar plot showing illustrative ok, warning, error, boundary, and skipped replicate counts by simulation surface."}
failure_ledger <- data.frame(
  surface = rep(surface_levels, each = 5),
  status = rep(c("ok", "warning", "boundary", "error", "skipped"), 4),
  count = c(
    268, 14, 10, 4, 4,
    262, 18, 12, 5, 3,
    232, 30, 20, 12, 6,
    252, 20, 14, 8, 6
  )
)
failure_ledger$surface <- factor(failure_ledger$surface, levels = surface_levels)
failure_ledger$status <- factor(
  failure_ledger$status,
  levels = c("ok", "warning", "boundary", "error", "skipped")
)

ggplot(failure_ledger, aes(surface, count, fill = status)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    "ok" = "#009E73",
    "warning" = "#E69F00",
    "boundary" = "#56B4E9",
    "error" = "#D55E00",
    "skipped" = "#999999"
  )) +
  labs(
    title = "Warnings and failures stay visible",
    subtitle = "Failure ledgers keep hard cells in the evidence trail",
    x = NULL,
    y = "Replicate count",
    fill = "Status"
  ) +
  theme_sim_grammar()
```

Real reports should pair this figure with a table that preserves warning
classes, error messages, seeds, and cell identifiers. The plot is for scanning;
the ledger is for debugging and reproducibility.

## Minimum Table Contract

A real simulation result can add columns, but it should not drop these fields:

| Field | Purpose |
| --- | --- |
| `surface` | Human-readable model surface, such as Gaussian location-scale, count, proportion, or meta-analysis |
| `cell_id` | The varied condition, such as sample size, effect size, true SD, or known-`V` structure |
| `estimand` | The scientific target, such as `beta_x`, `sigma`, `sd_group`, or `rho12` |
| `truth`, `estimate`, `error` | The value used for bias and RMSE |
| `conf.low`, `conf.high`, `interval_status` | The interval source and whether coverage is meaningful |
| `converged`, `pdHess`, `boundary` | Fit diagnostics that should be summarised and plotted |
| `elapsed`, `warning_count`, `error_message` | Runtime and failure-ledger fields |
| `n_replicate`, `mcse` | Monte Carlo uncertainty for any aggregate claim |

The plotting grammar is deliberately conservative. A feature should not enter a
comprehensive plot until the model surface is fitted, the estimand is named, the
interval source is known, and failed replicates remain visible.
