---
title: "Simulation plot grammar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation plot grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This article shows the plot grammar that Phase 18 simulation reports should use
before broad simulation grids are advertised as evidence. It is a display
contract, not a final result report. The small tables below are illustrative
fixtures that mimic the columns a real simulation summary should carry.

Simulation figures should answer the same reader questions across continuous,
proportion, count, and meta-analysis examples:

- Does the estimator recover the truth without important bias?
- How large is the error, measured by RMSE?
- Do interval procedures cover at the advertised rate?
- Does power increase under the intended effect?
- How often do fits converge, hit boundaries, warn, or fail?
- How much runtime does each surface require?

```{r setup}
library(ggplot2)

theme_sim_grammar <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey30", lineheight = 1.05),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      plot.margin = margin(6, 10, 12, 6)
    )
}

surface_palette <- c(
  "Gaussian location-scale" = "#0072B2",
  "Proportion" = "#009E73",
  "Count" = "#D55E00",
  "Meta-analysis" = "#CC79A7"
)

surface_levels <- names(surface_palette)
surface_facet_labels <- c(
  "Gaussian location-scale" = "Gaussian\nlocation-scale",
  "Proportion" = "Proportion",
  "Count" = "Count",
  "Meta-analysis" = "Meta-analysis"
)
estimand_display_labels <- c(
  "beta_x" = "beta_x",
  "sigma" = "sigma",
  "sd_group" = "sd(group)",
  "rho12" = "rho12"
)
```

## Bias and RMSE

Bias and RMSE should be plotted beside the estimand. A single average across
all parameters hides the distinction between fixed effects, residual scale,
random-effect SDs, and residual correlation `rho12`.

```{r bias-rmse-display, fig.width=8.4, fig.height=5.2, fig.cap="Simulation accuracy fixture for bias and RMSE; faint marks are sampled replicate errors, points are aggregate summaries, and bars are 95% Monte Carlo standard-error intervals.", fig.alt="Two simulation accuracy plots faceted by simulation surface: a raincloud-style bias display with fixture replicate errors, mean bias points, and MCSE intervals, followed by an aggregate RMSE display with MCSE intervals for four estimands."}
accuracy_targets <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  target_bias = c(
    0.01, -0.02, 0.03, 0.00,
    0.00, 0.01, 0.02, NA,
    0.02, NA, 0.06, NA,
    -0.01, 0.02, NA, 0.03
  ),
  target_rmse = c(
    0.09, 0.12, 0.18, 0.16,
    0.07, 0.09, 0.15, NA,
    0.12, NA, 0.22, NA,
    0.08, 0.11, NA, 0.14
  ),
  n_replicate = c(
    300, 300, 260, 280,
    300, 300, 250, NA,
    280, NA, 220, NA,
    300, 300, NA, 260
  )
)
accuracy_targets$error_sd <- sqrt(
  pmax(accuracy_targets$target_rmse^2 - accuracy_targets$target_bias^2, 0)
)
accuracy_replicates <- do.call(
  rbind,
  lapply(seq_len(nrow(accuracy_targets)), function(i) {
    row <- accuracy_targets[i, ]
    if (
      is.na(row$target_bias) ||
        is.na(row$target_rmse) ||
        is.na(row$n_replicate)
    ) {
      return(NULL)
    }
    data.frame(
      surface = row$surface,
      estimand = row$estimand,
      replicate = seq_len(row$n_replicate),
      error = row$target_bias +
        qnorm(ppoints(row$n_replicate)) * row$error_sd
    )
  })
)
summarise_accuracy_rows <- function(rows) {
  data.frame(
    surface = rows$surface[1],
    estimand = rows$estimand[1],
    bias = mean(rows$error),
    rmse = sqrt(mean(rows$error^2)),
    error_sd = stats::sd(rows$error),
    n_replicate = nrow(rows)
  )
}
accuracy_summary <- do.call(
  rbind,
  lapply(
    split(
      accuracy_replicates,
      interaction(accuracy_replicates$surface, accuracy_replicates$estimand, drop = TRUE)
    ),
    summarise_accuracy_rows
  )
)
accuracy_summary <- merge(
  accuracy_targets[c("surface", "estimand", "n_replicate")],
  accuracy_summary,
  by = c("surface", "estimand"),
  all.x = TRUE,
  sort = FALSE
)
accuracy_summary$n_replicate <- ifelse(
  is.na(accuracy_summary$n_replicate.y),
  accuracy_summary$n_replicate.x,
  accuracy_summary$n_replicate.y
)
accuracy_summary$n_replicate.x <- NULL
accuracy_summary$n_replicate.y <- NULL
accuracy_summary$surface <- factor(
  accuracy_summary$surface,
  levels = surface_levels
)
accuracy_summary$bias_mcse <- with(
  accuracy_summary,
  error_sd / sqrt(n_replicate)
)
accuracy_summary$rmse_mcse <- with(
  accuracy_summary,
  rmse / sqrt(2 * n_replicate)
)
accuracy_summary$bias_low <- accuracy_summary$bias -
  1.96 * accuracy_summary$bias_mcse
accuracy_summary$bias_high <- accuracy_summary$bias +
  1.96 * accuracy_summary$bias_mcse
accuracy_summary$rmse_low <- pmax(
  0,
  accuracy_summary$rmse - 1.96 * accuracy_summary$rmse_mcse
)
accuracy_summary$rmse_high <- accuracy_summary$rmse +
  1.96 * accuracy_summary$rmse_mcse
accuracy_summary$estimand <- factor(
  accuracy_summary$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)
accuracy_missing <- accuracy_summary[
  is.na(accuracy_summary$bias) | is.na(accuracy_summary$rmse),
  ,
  drop = FALSE
]
accuracy_missing$missing_label <- "not\ntargeted"
accuracy_missing$bias_missing_x <- 0.50
accuracy_missing$rmse_missing_x <- 0.22

display_replicates <- 49
accuracy_replicate_display <- do.call(
  rbind,
  lapply(split(accuracy_replicates, accuracy_replicates[c("surface", "estimand")]), function(rows) {
    if (nrow(rows) == 0L) {
      return(NULL)
    }
    keep <- round(seq(1, nrow(rows), length.out = display_replicates))
    rows[keep, , drop = FALSE]
  })
)
accuracy_replicate_display$surface <- factor(
  accuracy_replicate_display$surface,
  levels = surface_levels
)
accuracy_replicate_display$estimand <- factor(
  accuracy_replicate_display$estimand,
  levels = levels(accuracy_summary$estimand)
)
set.seed(20260519)
ggplot(accuracy_replicate_display, aes(error, estimand)) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey55") +
  geom_violin(
    aes(
      fill = surface
    ),
    width = 0.62,
    alpha = 0.16,
    colour = NA,
    scale = "width",
    trim = FALSE
  ) +
  geom_point(
    aes(colour = surface),
    position = position_jitter(height = 0.05, width = 0),
    alpha = 0.22,
    size = 0.65,
    show.legend = FALSE
  ) +
  geom_errorbar(
    data = accuracy_summary,
    aes(
      xmin = bias_low,
      xmax = bias_high,
      y = estimand,
      colour = surface
    ),
    inherit.aes = FALSE,
    orientation = "y",
    width = 0.16,
    linewidth = 0.65,
    na.rm = TRUE
  ) +
  geom_point(
    data = accuracy_summary,
    aes(bias, estimand, colour = surface),
    size = 2.4,
    na.rm = TRUE
  ) +
  geom_text(
    data = accuracy_missing,
    aes(
      x = bias_missing_x,
      y = estimand,
      label = missing_label
    ),
    colour = "grey55",
    size = 2.25,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~surface,
    nrow = 1,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  scale_fill_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "Bias displays should show replicate-level error",
    subtitle = "Faint dots sample fixture replicate errors; points/bars show mean bias with 95% MCSE",
    x = "Estimate minus truth",
    y = NULL,
    colour = "Surface"
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  guides(colour = "none", fill = "none") +
  theme_sim_grammar()

ggplot(accuracy_summary, aes(rmse, estimand, colour = surface)) +
  geom_errorbar(
    aes(
      xmin = rmse_low,
      xmax = rmse_high
    ),
    orientation = "y",
    width = 0.15,
    linewidth = 0.9,
    alpha = 0.9,
    na.rm = TRUE
  ) +
  geom_point(size = 2.4, na.rm = TRUE) +
  geom_text(
    data = accuracy_missing,
    aes(
      x = rmse_missing_x,
      y = estimand,
      label = missing_label
    ),
    colour = "grey55",
    size = 2.25,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_wrap(
    ~surface,
    nrow = 2,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, drop = FALSE) +
  labs(
    title = "RMSE needs its own uncertainty display",
    subtitle = "Facets are surfaces; points and bars are RMSE and 95% MCSE intervals for each estimand",
    x = "Root mean-square error",
    y = NULL,
    colour = "Surface"
  ) +
  scale_x_continuous(
    breaks = c(0, 0.10, 0.20),
    limits = c(0, 0.31),
    expand = expansion(mult = c(0.01, 0.04))
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  guides(colour = "none") +
  theme_sim_grammar()
```

The missing rows are part of the message, so unsupported cells stay visible as
`not targeted` lanes. A count pilot should not quietly fill `rho12` with zeros,
and a proportion example should not invent a random-effect SD if that surface
was not fitted. Real bias reports should replace the fixture quantiles above
with one row per replicate estimate or error, then overplot the aggregate mean
and MCSE interval produced by the Phase 18 summary helpers. RMSE should remain
an aggregate root-mean-square summary with its own MCSE interval, not a cloud
whose center could be mistaken for a mean absolute error.

## Coverage and Power

Coverage and power are proportions, so every point should carry or clearly
reference the number of replicates that contributed to it and a Monte Carlo
uncertainty interval. Real reports should plot stored replicate rows or stored
replicate-block summaries when they are available; the fixture below illustrates
that grammar with block-level proportions plus a binomial MCSE approximation for
the aggregate.

The example table still carries `n_interval` for each cell. The rendered plot
keeps the `n` values out of the panel labels so the MCSE bars, points, and
block-level dots stay aligned and readable.

```{r coverage-power-display, fig.width=8.6, fig.height=5.2, fig.cap="Coverage and power fixture; faint dots are replicate-block proportions, points are aggregate proportions, and horizontal bars are 95% binomial MCSE intervals.", fig.alt="Faceted point plot showing illustrative coverage and power with Monte Carlo uncertainty intervals for four simulation surfaces, with one panel per surface so intervals align with their points."}
interval_summary <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  coverage = c(
    0.95, 0.93, 0.91, 0.94,
    0.94, 0.92, 0.90, NA,
    0.92, NA, 0.86, NA,
    0.95, 0.94, NA, 0.92
  ),
  power = c(
    0.82, 0.64, 0.58, 0.70,
    0.78, 0.60, 0.52, NA,
    0.74, NA, 0.46, NA,
    0.80, 0.61, NA, 0.66
  ),
  n_interval = c(
    280, 280, 240, 260,
    280, 280, 230, NA,
    250, NA, 200, NA,
    280, 280, NA, 240
  )
)
interval_summary$surface <- factor(
  interval_summary$surface,
  levels = surface_levels
)
interval_long <- rbind(
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Coverage",
    value = interval_summary$coverage,
    target = 0.95
  ),
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Power",
    value = interval_summary$power,
    target = 0.80
  )
)
interval_long$mcse <- sqrt(
  interval_long$value * (1 - interval_long$value) / interval_long$n_interval
)
interval_long$lower <- pmax(0, interval_long$value - 1.96 * interval_long$mcse)
interval_long$upper <- pmin(1, interval_long$value + 1.96 * interval_long$mcse)
interval_long$measure <- factor(
  interval_long$measure,
  levels = c("Coverage", "Power")
)
interval_long$estimand <- factor(
  interval_long$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)
interval_long$missing_x <- ifelse(
  interval_long$measure == "Coverage",
  0.405,
  0.405
)
interval_target <- unique(interval_long[c("measure", "target")])
interval_present <- interval_long[!is.na(interval_long$value), , drop = FALSE]
interval_missing <- interval_long[is.na(interval_long$value), , drop = FALSE]
make_interval_blocks <- function(row, n_block = 18) {
  block_size <- max(8L, floor(row$n_interval / n_block))
  block_proportion <- qbinom(
    ppoints(n_block),
    size = block_size,
    prob = row$value
  ) / block_size
  data.frame(
    surface = row$surface,
    estimand = row$estimand,
    measure = row$measure,
    block_proportion = block_proportion
  )
}
interval_blocks <- do.call(
  rbind,
  lapply(seq_len(nrow(interval_present)), function(i) {
    make_interval_blocks(interval_present[i, ])
  })
)

ggplot(interval_present, aes(value, estimand, colour = surface)) +
  geom_vline(
    data = interval_target,
    aes(xintercept = target),
    linetype = "dotted",
    colour = "grey55",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = interval_blocks,
    aes(block_proportion, estimand, colour = surface),
    position = position_jitter(height = 0.07, width = 0),
    alpha = 0.18,
    size = 0.70,
    inherit.aes = FALSE
  ) +
  geom_segment(
    aes(x = lower, xend = upper, yend = estimand),
    linewidth = 0.4,
    na.rm = TRUE
  ) +
  geom_point(size = 2.1, na.rm = TRUE) +
  geom_text(
    data = interval_missing,
    aes(x = missing_x, y = estimand, label = "not\ntargeted"),
    colour = "grey55",
    size = 2.15,
    lineheight = 0.9,
    hjust = 0,
    inherit.aes = FALSE
  ) +
  facet_grid(
    measure ~ surface,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = surface_palette, guide = "none") +
  coord_cartesian(xlim = c(0.38, 1)) +
  scale_x_continuous(
    breaks = c(0.4, 0.6, 0.8, 0.95),
    labels = c("0.40", "0.60", "0.80", "0.95")
  ) +
  scale_y_discrete(labels = estimand_display_labels) +
  labs(
    title = "Interval coverage and power need Monte Carlo uncertainty",
    subtitle = "Faint dots: replicate-block proportions; points/bars: aggregate proportions with 95% binomial MCSE",
    x = "Proportion",
    y = NULL,
    colour = "Surface"
  ) +
  theme_sim_grammar()
```

The plot should make pilot status obvious. With few replicates, a coverage
estimate near 0.95 can still be too noisy for a release claim.

## Convergence and Runtime

Convergence, positive-definite Hessian status, and runtime should sit beside
accuracy. A surface with low bias but frequent warnings or high runtime is not
ready for broad claims.

```{r convergence-runtime-display, fig.width=8.6, fig.height=4.6, fig.cap="Simulation readiness fixture for fit status and runtime; proportions, medians, and high runtime quantiles are separate checks shown on their own rows.", fig.alt="Faceted readiness plot with one row each for convergence, positive-definite Hessian rate, median runtime, and high runtime quantile by simulation surface."}
run_summary <- data.frame(
  surface = surface_levels,
  convergence_rate = c(0.98, 0.96, 0.90, 0.95),
  pdhess_rate = c(0.96, 0.94, 0.86, 0.93),
  elapsed_median = c(0.35, 0.28, 0.52, 0.44),
  elapsed_p90 = c(0.72, 0.58, 1.35, 0.96)
)
run_summary$surface <- factor(run_summary$surface, levels = surface_levels)
run_long <- rbind(
  data.frame(
    surface = run_summary$surface,
    measure = "Converged",
    value = run_summary$convergence_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "pdHess",
    value = run_summary$pdhess_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "Median runtime",
    value = run_summary$elapsed_median
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "90th percentile runtime",
    value = run_summary$elapsed_p90
  )
)
run_long$panel <- ifelse(
  grepl("runtime", run_long$measure),
  "Runtime (seconds)",
  "Fit status proportion"
)
run_long$measure <- factor(
  run_long$measure,
  levels = c(
    "Converged",
    "pdHess",
    "Median runtime",
    "90th percentile runtime"
  )
)
run_long$measure_label <- factor(
  run_long$measure,
  levels = rev(levels(run_long$measure))
)

ggplot(run_long, aes(value, measure_label, colour = measure)) +
  geom_segment(
    aes(x = 0, xend = value, yend = measure_label),
    linewidth = 1.2,
    lineend = "round"
  ) +
  geom_point(size = 2.6) +
  facet_grid(
    panel ~ surface,
    scales = "free",
    space = "free_y",
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_colour_manual(values = c(
    "Converged" = "#0072B2",
    "pdHess" = "#56B4E9",
    "Median runtime" = "#009E73",
    "90th percentile runtime" = "#E69F00"
  )) +
  scale_y_discrete(labels = c(
    "Converged" = "converged",
    "pdHess" = "pdHess",
    "Median runtime" = "median",
    "90th percentile runtime" = "90th pct."
  )) +
  scale_x_continuous(
    breaks = c(0, 0.5, 1),
    expand = expansion(mult = c(0, 0.04))
  ) +
  labs(
    title = "Convergence and runtime belong beside accuracy",
    subtitle = "Rows keep fit-status proportions separate from runtime summaries",
    x = "Proportion or seconds",
    y = NULL,
    colour = "Measure"
  ) +
  guides(colour = "none") +
  theme_sim_grammar()
```

Runtime summaries should report medians and high quantiles, not only means,
because a few difficult cells can dominate overnight simulation planning.

## Failure Ledger

Failures and warnings are data. They should be summarised by surface and
condition instead of filtered away before plotting.

```{r failure-ledger-display, fig.width=8.6, fig.height=4.2, fig.cap="Failure-ledger fixture showing replicate counts by status; warnings, boundary cases, errors, and skipped cells remain visible as separate rows instead of stacked slivers.", fig.alt="Faceted horizontal bar plot showing illustrative ok, warning, boundary, error, and skipped replicate counts as separate rows by simulation surface."}
failure_ledger <- data.frame(
  surface = rep(surface_levels, each = 5),
  status = rep(c("ok", "warning", "boundary", "error", "skipped"), 4),
  count = c(
    268, 14, 10, 4, 4,
    262, 18, 12, 5, 3,
    232, 30, 20, 12, 6,
    252, 20, 14, 8, 6
  )
)
failure_ledger$surface <- factor(failure_ledger$surface, levels = surface_levels)
failure_ledger$status <- factor(
  failure_ledger$status,
  levels = c("ok", "warning", "boundary", "error", "skipped")
)
failure_ledger$status_label <- factor(
  failure_ledger$status,
  levels = rev(levels(failure_ledger$status))
)

ggplot(failure_ledger, aes(count, status_label, fill = status)) +
  geom_col(width = 0.64) +
  facet_wrap(
    ~surface,
    nrow = 1,
    labeller = labeller(surface = as_labeller(surface_facet_labels))
  ) +
  scale_fill_manual(values = c(
    "ok" = "#009E73",
    "warning" = "#E69F00",
    "boundary" = "#56B4E9",
    "error" = "#D55E00",
    "skipped" = "#999999"
  )) +
  scale_y_discrete(labels = c(
    "ok" = "ok",
    "warning" = "warning",
    "boundary" = "boundary",
    "error" = "error",
    "skipped" = "skipped"
  )) +
  scale_x_continuous(
    breaks = c(0, 25, 100, 200, 300),
    expand = expansion(mult = c(0, 0.04))
  ) +
  labs(
    title = "Warnings and failures stay visible",
    subtitle = "Separate rows keep rare statuses readable beside the dominant ok count",
    x = "Replicate count",
    y = NULL,
    fill = "Status"
  ) +
  guides(fill = "none") +
  theme_sim_grammar()
```

Real reports should pair this figure with a table that preserves warning
classes, error messages, seeds, and cell identifiers. The plot is for scanning;
the ledger is for debugging and reproducibility.

## Minimum Table Contract

A real simulation result can add columns, but it should not drop these fields:

| Field | Purpose |
| --- | --- |
| `surface` | Human-readable model surface, such as Gaussian location-scale, count, proportion, or meta-analysis |
| `cell_id` | The varied condition, such as sample size, effect size, true SD, or known-`V` structure |
| `estimand` | The scientific target, such as `beta_x`, `sigma`, `sd_group`, or `rho12` |
| `truth`, `estimate`, `error` | The value used for bias and RMSE |
| `conf.low`, `conf.high`, `interval_status` | The interval source and whether coverage is meaningful |
| `converged`, `pdHess`, `boundary` | Fit diagnostics that should be summarised and plotted |
| `elapsed`, `warning_count`, `error_message` | Runtime and failure-ledger fields |
| `n_replicate`, `mcse` | Monte Carlo uncertainty for any aggregate claim |

The plotting grammar is deliberately conservative. A feature should not enter a
comprehensive plot until the model surface is fitted, the estimand is named, the
interval source is known, and failed replicates remain visible.
