---
title: "Simulation plot grammar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation plot grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 144,
  eval = has_ggplot2
)
```

This article shows the plot grammar that Phase 18 simulation reports should use
before broad simulation grids are advertised as evidence. It is a display
contract, not a final result report. The small tables below are illustrative
fixtures that mimic the columns a real simulation summary should carry.

Simulation figures should answer the same reader questions across continuous,
proportion, count, and meta-analysis examples:

- Does the estimator recover the truth without important bias?
- How large is the error, measured by RMSE?
- Do interval procedures cover at the advertised rate?
- Does power increase under the intended effect?
- How often do fits converge, hit boundaries, warn, or fail?
- How much runtime does each surface require?

```{r setup}
library(ggplot2)

theme_sim_grammar <- function() {
  theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(colour = "grey30"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

surface_palette <- c(
  "Continuous" = "#0072B2",
  "Proportion" = "#009E73",
  "Count" = "#D55E00",
  "Meta-analysis" = "#CC79A7"
)

surface_levels <- names(surface_palette)
```

## Bias and RMSE

Bias and RMSE should be plotted beside the estimand. A single average across
all parameters hides the distinction between fixed effects, residual scale,
random-effect SDs, and residual correlation `rho12`.

```{r bias-rmse-display, fig.alt="Faceted dot plot showing illustrative bias and RMSE for four simulation surfaces and four estimands."}
accuracy_summary <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  bias = c(
    0.01, -0.02, 0.03, 0.00,
    0.00, 0.01, 0.02, NA,
    0.02, NA, 0.06, NA,
    -0.01, 0.02, NA, 0.03
  ),
  rmse = c(
    0.09, 0.12, 0.18, 0.16,
    0.07, 0.09, 0.15, NA,
    0.12, NA, 0.22, NA,
    0.08, 0.11, NA, 0.14
  ),
  n_replicate = c(
    300, 300, 260, 280,
    300, 300, 250, NA,
    280, NA, 220, NA,
    300, 300, NA, 260
  )
)
accuracy_summary$surface <- factor(
  accuracy_summary$surface,
  levels = surface_levels
)
accuracy_long <- rbind(
  data.frame(
    accuracy_summary[c("surface", "estimand", "n_replicate")],
    measure = "Bias",
    value = accuracy_summary$bias
  ),
  data.frame(
    accuracy_summary[c("surface", "estimand", "n_replicate")],
    measure = "RMSE",
    value = accuracy_summary$rmse
  )
)
accuracy_long$measure <- factor(accuracy_long$measure, levels = c("Bias", "RMSE"))
accuracy_long$estimand <- factor(
  accuracy_long$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)

ggplot(
  accuracy_long,
  aes(value, estimand, colour = surface, group = surface)
) +
  geom_vline(
    data = data.frame(measure = "Bias"),
    aes(xintercept = 0),
    inherit.aes = FALSE,
    linetype = "dashed",
    colour = "grey55"
  ) +
  geom_point(size = 2.1, na.rm = TRUE) +
  facet_wrap(~measure, scales = "free_x") +
  scale_colour_manual(values = surface_palette) +
  labs(
    title = "Bias and RMSE by estimand",
    subtitle = "Missing points are unsupported or not-yet-targeted estimands",
    x = "Summary value",
    y = NULL,
    colour = "Surface"
  ) +
  theme_sim_grammar()
```

The missing rows are part of the message. A count pilot should not quietly fill
`rho12` with zeros, and a proportion example should not invent a random-effect
SD if that surface was not fitted.

## Coverage and Power

Coverage and power are proportions, so every point should carry the number of
replicates that contributed to it and a Monte Carlo uncertainty interval. The
fixture below uses a binomial MCSE approximation for display.

```{r coverage-power-display, fig.alt="Faceted point plot showing illustrative coverage and power with Monte Carlo uncertainty intervals for four simulation surfaces."}
interval_summary <- data.frame(
  surface = rep(surface_levels, each = 4),
  estimand = rep(c("beta_x", "sigma", "sd_group", "rho12"), 4),
  coverage = c(
    0.95, 0.93, 0.91, 0.94,
    0.94, 0.92, 0.90, NA,
    0.92, NA, 0.86, NA,
    0.95, 0.94, NA, 0.92
  ),
  power = c(
    0.82, 0.64, 0.58, 0.70,
    0.78, 0.60, 0.52, NA,
    0.74, NA, 0.46, NA,
    0.80, 0.61, NA, 0.66
  ),
  n_interval = c(
    280, 280, 240, 260,
    280, 280, 230, NA,
    250, NA, 200, NA,
    280, 280, NA, 240
  )
)
interval_summary$surface <- factor(
  interval_summary$surface,
  levels = surface_levels
)
interval_long <- rbind(
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Coverage",
    value = interval_summary$coverage,
    target = 0.95
  ),
  data.frame(
    interval_summary[c("surface", "estimand", "n_interval")],
    measure = "Power",
    value = interval_summary$power,
    target = 0.80
  )
)
interval_long$mcse <- sqrt(
  interval_long$value * (1 - interval_long$value) / interval_long$n_interval
)
interval_long$lower <- pmax(0, interval_long$value - 1.96 * interval_long$mcse)
interval_long$upper <- pmin(1, interval_long$value + 1.96 * interval_long$mcse)
interval_long$measure <- factor(
  interval_long$measure,
  levels = c("Coverage", "Power")
)
interval_long$estimand <- factor(
  interval_long$estimand,
  levels = rev(c("beta_x", "sigma", "sd_group", "rho12"))
)

ggplot(interval_long, aes(value, estimand, colour = surface)) +
  geom_vline(
    aes(xintercept = target),
    linetype = "dashed",
    colour = "grey55"
  ) +
  geom_segment(
    aes(x = lower, xend = upper, yend = estimand),
    linewidth = 0.55,
    na.rm = TRUE
  ) +
  geom_point(size = 2.1, na.rm = TRUE) +
  facet_wrap(~measure, scales = "free_x") +
  scale_colour_manual(values = surface_palette) +
  labs(
    title = "Interval coverage and power need Monte Carlo uncertainty",
    subtitle = "Dashed lines mark nominal coverage and a planned power benchmark",
    x = "Proportion",
    y = NULL,
    colour = "Surface"
  ) +
  theme_sim_grammar()
```

The plot should make pilot status obvious. With few replicates, a coverage
estimate near 0.95 can still be too noisy for a release claim.

## Convergence and Runtime

Convergence, positive-definite Hessian status, and runtime should sit beside
accuracy. A surface with low bias but frequent warnings or high runtime is not
ready for broad claims.

```{r convergence-runtime-display, fig.alt="Faceted plot showing illustrative convergence rate, Hessian rate, median runtime, and high runtime quantile by simulation surface."}
run_summary <- data.frame(
  surface = surface_levels,
  convergence_rate = c(0.98, 0.96, 0.90, 0.95),
  pdhess_rate = c(0.96, 0.94, 0.86, 0.93),
  elapsed_median = c(0.35, 0.28, 0.52, 0.44),
  elapsed_p90 = c(0.72, 0.58, 1.35, 0.96)
)
run_summary$surface <- factor(run_summary$surface, levels = surface_levels)
run_long <- rbind(
  data.frame(
    surface = run_summary$surface,
    measure = "Converged",
    value = run_summary$convergence_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "pdHess",
    value = run_summary$pdhess_rate
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "Median runtime",
    value = run_summary$elapsed_median
  ),
  data.frame(
    surface = run_summary$surface,
    measure = "90th percentile runtime",
    value = run_summary$elapsed_p90
  )
)
run_long$panel <- ifelse(
  grepl("runtime", run_long$measure),
  "Runtime (seconds)",
  "Fit status proportion"
)
run_long$measure <- factor(
  run_long$measure,
  levels = c(
    "Converged",
    "pdHess",
    "Median runtime",
    "90th percentile runtime"
  )
)

ggplot(run_long, aes(value, surface, fill = measure)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.64) +
  facet_wrap(~panel, scales = "free_x") +
  scale_fill_manual(values = c(
    "Converged" = "#0072B2",
    "pdHess" = "#56B4E9",
    "Median runtime" = "#009E73",
    "90th percentile runtime" = "#E69F00"
  )) +
  labs(
    title = "Convergence and runtime belong beside accuracy",
    subtitle = "Fast successful fits and stable uncertainty are separate checks",
    x = "Value",
    y = NULL,
    fill = "Measure"
  ) +
  theme_sim_grammar()
```

Runtime summaries should report medians and high quantiles, not only means,
because a few difficult cells can dominate overnight simulation planning.

## Failure Ledger

Failures and warnings are data. They should be summarised by surface and
condition instead of filtered away before plotting.

```{r failure-ledger-display, fig.alt="Stacked bar plot showing illustrative ok, warning, error, boundary, and skipped replicate counts by simulation surface."}
failure_ledger <- data.frame(
  surface = rep(surface_levels, each = 5),
  status = rep(c("ok", "warning", "boundary", "error", "skipped"), 4),
  count = c(
    268, 14, 10, 4, 4,
    262, 18, 12, 5, 3,
    232, 30, 20, 12, 6,
    252, 20, 14, 8, 6
  )
)
failure_ledger$surface <- factor(failure_ledger$surface, levels = surface_levels)
failure_ledger$status <- factor(
  failure_ledger$status,
  levels = c("ok", "warning", "boundary", "error", "skipped")
)

ggplot(failure_ledger, aes(surface, count, fill = status)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    "ok" = "#009E73",
    "warning" = "#E69F00",
    "boundary" = "#56B4E9",
    "error" = "#D55E00",
    "skipped" = "#999999"
  )) +
  labs(
    title = "Warnings and failures stay visible",
    subtitle = "Failure ledgers keep hard cells in the evidence trail",
    x = NULL,
    y = "Replicate count",
    fill = "Status"
  ) +
  theme_sim_grammar()
```

Real reports should pair this figure with a table that preserves warning
classes, error messages, seeds, and cell identifiers. The plot is for scanning;
the ledger is for debugging and reproducibility.

## Minimum Table Contract

A real simulation result can add columns, but it should not drop these fields:

| Field | Purpose |
| --- | --- |
| `surface` | Human-readable model surface, such as Gaussian location-scale, count, proportion, or meta-analysis |
| `cell_id` | The varied condition, such as sample size, effect size, true SD, or known-`V` structure |
| `estimand` | The scientific target, such as `beta_x`, `sigma`, `sd_group`, or `rho12` |
| `truth`, `estimate`, `error` | The value used for bias and RMSE |
| `conf.low`, `conf.high`, `interval_status` | The interval source and whether coverage is meaningful |
| `converged`, `pdHess`, `boundary` | Fit diagnostics that should be summarised and plotted |
| `elapsed`, `warning_count`, `error_message` | Runtime and failure-ledger fields |
| `n_replicate`, `mcse` | Monte Carlo uncertainty for any aggregate claim |

The plotting grammar is deliberately conservative. A feature should not enter a
comprehensive plot until the model surface is fitted, the estimand is named, the
interval source is known, and failed replicates remain visible.
