---
title: "Coordinate-spatial structured effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coordinate-spatial structured effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.1,
  dpi = 144
)
library(drmTMB)

spatial_guide_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.title = ggplot2::element_text(colour = "grey15"),
      axis.text = ggplot2::element_text(colour = "grey25"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "grey10",
        margin = ggplot2::margin(b = 4)
      ),
      plot.subtitle = ggplot2::element_text(
        colour = "grey30",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "bottom"
    )
}

spatial_eye_theme <- function() {
  spatial_guide_theme() +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_blank(),
      legend.position = "none"
    )
}

simulate_spatial_guide_data <- function(seed = 20260534) {
  set.seed(seed)
  sites <- paste0("site_", seq_len(12))
  theta <- seq(0, 1.8 * pi, length.out = length(sites))
  coords <- data.frame(
    x = cos(theta) + seq_along(sites) / 30,
    y = sin(theta)
  )
  rownames(coords) <- sites

  distance <- as.matrix(dist(coords))
  range <- stats::median(distance[distance > 0])
  spatial_cov <- exp(-distance / range)
  diag(spatial_cov) <- diag(spatial_cov) + 1e-6
  spatial_intercept <- as.vector(
    t(chol(spatial_cov)) %*% rnorm(length(sites), sd = 0.45)
  )
  spatial_slope <- as.vector(
    t(chol(spatial_cov)) %*% rnorm(length(sites), sd = 0.20)
  )
  names(spatial_intercept) <- sites
  names(spatial_slope) <- sites

  site <- rep(sites, each = 6)
  temp <- rnorm(length(site))
  depth <- rnorm(length(site))
  sigma <- exp(-1.25 + 0.15 * depth)
  y <- 0.4 + 0.35 * temp + 0.20 * depth +
    spatial_intercept[site] + spatial_slope[site] * depth +
    rnorm(length(site), sd = sigma)

  list(
    data = data.frame(y = y, temp = temp, depth = depth, site = site),
    coords = coords
  )
}

simulate_spatial_q2_guide_data <- function(seed = 20260501) {
  set.seed(seed)
  sites <- paste0("site_", seq_len(16))
  theta <- seq(0, 1.9 * pi, length.out = length(sites))
  coords <- data.frame(
    x = cos(theta) + seq_along(sites) / 40,
    y = sin(theta)
  )
  rownames(coords) <- sites

  distance <- as.matrix(dist(coords))
  range <- stats::median(distance[distance > 0])
  spatial_cov <- exp(-distance / range)
  diag(spatial_cov) <- diag(spatial_cov) + 1e-6

  z1 <- rnorm(length(sites))
  z2 <- 0.55 * z1 + sqrt(1 - 0.55^2) * rnorm(length(sites))
  u1 <- as.vector(t(chol(spatial_cov)) %*% z1) * 0.55
  u2 <- as.vector(t(chol(spatial_cov)) %*% z2) * 0.45
  names(u1) <- sites
  names(u2) <- sites

  site <- rep(sites, each = 8)
  x <- rnorm(length(site))
  e1 <- rnorm(length(site))
  e2 <- 0.10 * e1 + sqrt(1 - 0.10^2) * rnorm(length(site))

  list(
    data = data.frame(
      site = site,
      x = x,
      y1 = 0.30 + 0.35 * x + u1[site] + 0.25 * e1,
      y2 = -0.10 - 0.25 * x + u2[site] + 0.24 * e2
    ),
    coords = coords
  )
}
```

Use `spatial()` when named sites have coordinates and nearby sites may have
similar location deviations after fixed effects have been included. The fitted
route uses a coordinate table supplied as `coords = coords`. Mesh and SPDE
inputs remain planned. If you are still choosing among `animal()`, `phylo()`,
`spatial()`, and `relmat()`, read the [structural-dependence
overview](structural-dependence.html) first.

## What is fitted today

| Question | Syntax | Status |
| --- | --- | --- |
| Does one Gaussian response have smooth site-level location deviations? | `spatial(1 | site, coords = coords)` in `mu` | Fitted first coordinate-spatial intercept slice. |
| Does one predictor have a spatially varying slope? | `spatial(1 + depth | site, coords = coords)` in `mu` | Fitted one numeric-slope slice. The intercept and slope fields are independent and have separate SDs. |
| Do two Gaussian response means share a coordinate-spatial correlation? | matching `spatial(1 | p | site, coords = coords)` terms in `mu1` and `mu2` | Fitted q=2 location-location slice. `corpairs(level = "spatial")` reports the latent spatial row separately from residual `rho12`. |
| Do spatial location and scale deviations covary across two Gaussian responses? | matching `spatial(1 | p | site, coords = coords)` terms in `mu1`, `mu2`, `sigma1`, and `sigma2` | Fitted constant q=4 location-scale slice. Six latent spatial rows are reported through `corpairs(level = "spatial")`; q=4 correlations are derived and unavailable for intervals. |

## Start with the smallest useful model

For one response, start with a coordinate-spatial location intercept:

```r
fit_spatial <- drmTMB(
  y ~ treatment + spatial(1 | site, coords = coords),
  data = dat,
  family = gaussian()
)
```

Add one numeric slope only when the scientific question is about spatial
variation in that slope:

```r
fit_spatial_slope <- drmTMB(
  y ~ treatment + depth + spatial(1 + depth | site, coords = coords),
  data = dat,
  family = gaussian()
)
```

For the exact Arc 1a REML route, keep `sigma ~ 1`, use an unlabelled intercept
or independent intercept-plus-one-numeric-slope shape, and set `REML = TRUE`:

```r
fit_spatial_reml <- drmTMB(
  bf(
    y ~ depth + spatial(1 + depth | site, coords = coords),
    sigma ~ 1
  ),
  data = dat,
  family = gaussian(),
  REML = TRUE
)
```

The multi-seed campaign used the coordinate representation shown here, with
`n_each = 20` and exactly `M = {8, 16, 32}` sites. This is not a continuous
minimum-sample-size claim, and it does not admit estimated range, labelled,
slope-only, multiple-slope, scale-side, other bivariate, or non-Gaussian REML
routes. The exact bivariate exception is the matched labelled location-intercept
cell below.

For two response means, use matching labelled terms and read the latent spatial
correlation with `corpairs()`:

```r
fit_spatial_q2 <- drmTMB(
  mu1 = trait1 ~ treatment +
    spatial(1 | p | site, coords = coords),
  mu2 = trait2 ~ treatment +
    spatial(1 | p | site, coords = coords),
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_spatial_q2, level = "spatial")
rho12(fit_spatial_q2)
```

`corpairs()` reports the fitted latent coordinate-spatial correlation among
site-level location deviations. `rho12()` reports the residual correlation
between paired responses after fixed effects and random effects have been
included.

That exact q2 location-intercept model also admits native REML when the three
residual parameters are intercept-only:

```r
fit_spatial_q2_reml <- drmTMB(
  bf(
    mu1 = trait1 ~ treatment +
      spatial(1 | p | site, coords = coords),
    mu2 = trait2 ~ treatment +
      spatial(1 | p | site, coords = coords),
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ 1
  ),
  data = dat,
  family = biv_gaussian(),
  REML = TRUE
)
```

Here the coordinates define a fixed spatial covariance matrix. This cell has
dense-oracle and retained-denominator point-recovery evidence only; it does not
authorize interval, coverage, range-estimation, slope, scale-side, or q4
claims.

For a constant location-scale spatial block, use the same labelled `spatial()`
term in all four bivariate Gaussian endpoints:

```r
fit_spatial_q4 <- drmTMB(
  mu1 = trait1 ~ treatment +
    spatial(1 | p | site, coords = coords),
  mu2 = trait2 ~ treatment +
    spatial(1 | p | site, coords = coords),
  sigma1 = ~ treatment +
    spatial(1 | p | site, coords = coords),
  sigma2 = ~ treatment +
    spatial(1 | p | site, coords = coords),
  rho12 = ~ 1,
  data = dat,
  family = biv_gaussian()
)
corpairs(fit_spatial_q4, level = "spatial")
```

This q=4 route estimates four coordinate-spatial endpoint SDs and six latent
correlations: one location-location, four location-scale, and one scale-scale
row. It is still a constant intercept block, not a spatial slope or
predictor-dependent spatial correlation model.

## What to inspect

After fitting, inspect the spatial layer before interpreting it:

| Output | Use |
| --- | --- |
| `check_drm(fit)` | Confirm the spatial layer was recognized and review coordinate/site diagnostics. |
| `fit$sdpars$mu` | Read fitted spatial location SDs, including separate intercept and slope SDs when a one-slope model is used. |
| `ranef(fit, "spatial_mu")` | Inspect conditional site deviations on the fitted spatial layer. |
| `summary(fit)$covariance` | Check how spatial SDs and q=2 or q=4 spatial correlations are reported beside other covariance layers. |
| `profile_targets(fit)` | See which spatial SD or constant q=2 correlation targets can be profiled directly, and which q=4 correlation rows are derived-unavailable for intervals. |
| `corpairs(fit, level = "spatial")` | Read fitted constant spatial correlation rows. |

The one-slope route is deliberately narrow. The formula term
`spatial(1 + depth | site, coords = coords)` fits an intercept field and one
numeric slope field with the same coordinate precision and separate SDs. It
does not estimate an intercept-slope correlation. Spatial `sigma` is supported
through a separate route -- a standalone `sigma ~ spatial(1 | site)` or
`sigma ~ spatial(1 + depth | site)` field, or the matched location-scale block --
which fits at recovery grade (trust the point estimate, not the interval); this
`mu` one-slope term is not that route.

## Rendered checks

The small example below is only a guide to the output grain. The coordinate
surface is fitted in the location predictor `mu`; raw response values remain on
the response scale and should not be plotted as if they were spatial SDs or
correlations.

```{r spatial-guide-fit}
spatial_example <- simulate_spatial_guide_data()
spatial_dat <- spatial_example$data
coords <- spatial_example$coords

fit_spatial <- drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 | site, coords = coords),
    sigma ~ depth
  ),
  family = gaussian(),
  data = spatial_dat
)

fit_spatial_slope <- drmTMB(
  drm_formula(
    y ~ depth + temp + spatial(1 + depth | site, coords = coords),
    sigma ~ depth
  ),
  family = gaussian(),
  data = spatial_dat
)
```

```{r spatial-site-field-figure, fig.width = 6.6, fig.height = 4.4, fig.cap = "Coordinate-spatial fitted site deviations from `ranef(fit_spatial, \"spatial_mu\")`; points are conditional location-effect estimates, and no interval is drawn because this display is a site-level fitted field rather than a confidence-interval summary.", fig.alt = "Map of twelve sampled sites. Each point is positioned at the site coordinates and coloured by the fitted conditional spatial location deviation; positive deviations are teal and negative deviations are orange."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  spatial_effect <- ranef(fit_spatial, "spatial_mu")$terms[[1]]
  spatial_field <- data.frame(
    site = names(spatial_effect),
    fitted_spatial_deviation = unname(spatial_effect),
    coords[names(spatial_effect), , drop = FALSE],
    row.names = NULL
  )

  ggplot2::ggplot(
    spatial_field,
    ggplot2::aes(
      x = x,
      y = y,
      fill = fitted_spatial_deviation
    )
  ) +
    ggplot2::geom_hline(yintercept = 0, colour = "grey90", linewidth = 0.4) +
    ggplot2::geom_vline(xintercept = 0, colour = "grey90", linewidth = 0.4) +
    ggplot2::geom_point(
      shape = 21,
      size = 7,
      colour = "grey20",
      linewidth = 0.35
    ) +
    ggplot2::scale_fill_gradient2(
      low = "#D55E00",
      mid = "white",
      high = "#009E73",
      midpoint = 0,
      name = "Fitted\nspatial deviation"
    ) +
    ggplot2::coord_equal() +
    spatial_guide_theme() +
    ggplot2::labs(
      title = "Fitted spatial location field",
      subtitle = "Conditional deviations in mu",
      x = "Coordinate x",
      y = "Coordinate y"
    )
}
```

```{r spatial-sd-figure, fig.width = 6.6, fig.height = 3.4, fig.cap = "Fitted coordinate-spatial intercept-field SD (response units) and depth-slope field SD (response units per unit depth) from the one-slope model. Interval calibration for these exact rows is not validated, so no interval bars are drawn.", fig.alt = "Horizontal point display for the spatial intercept and depth-slope standard deviations, without interval bars."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  spatial_sd <- data.frame(
    estimate = unname(fit_spatial_slope$sdpars$mu[
      c("spatial(1 | site)", "spatial(0 + depth | site)")
    ]),
    label = c("Spatial\nintercept SD", "Spatial\ndepth-slope SD"),
    unit = c("Response units", "Response units per depth unit")
  )

  ggplot2::ggplot(spatial_sd, ggplot2::aes(y = label)) +
    ggplot2::geom_vline(
      xintercept = 0,
      linewidth = 0.45,
      linetype = "dashed",
      colour = "grey55"
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = estimate),
      shape = 21,
      size = 3.5,
      stroke = 1,
      fill = "white",
      colour = "#0072B2"
    ) +
    ggplot2::scale_x_continuous(
      expand = ggplot2::expansion(mult = c(0.01, 0.04))
    ) +
    spatial_eye_theme() +
    ggplot2::labs(
      title = "Spatial intercept and slope SDs have different units",
      subtitle = "Point estimates only; interval validation remains planned",
      x = "Fitted spatial standard deviation",
      y = NULL
    )
}
```

For bivariate spatial location models, the current intercept-only q2
`corpairs()` row is point-only: profile mechanics exist, but interval calibration
and coverage remain planned. Native REML for this exact cell requires complete
response pairs, unit weights, intercept-only `sigma1`, `sigma2`, and `rho12`, no
known `meta_V()` covariance, and no additional ordinary random effect, direct-SD
formula, or `corpair()` regression. The raw paired responses do not belong on
this axis.

```{r spatial-q2-correlation-fit}
spatial_q2_example <- simulate_spatial_q2_guide_data()
spatial_q2_dat <- spatial_q2_example$data
spatial_q2_coords <- spatial_q2_example$coords

fit_spatial_q2_example <- drmTMB(
  drm_formula(
    mu1 = y1 ~ x + spatial(1 | p | site, coords = spatial_q2_coords),
    mu2 = y2 ~ x + spatial(1 | p | site, coords = spatial_q2_coords),
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12 = ~ 1
  ),
  family = c(gaussian(), gaussian()),
  data = spatial_q2_dat
)

spatial_q2_pairs <- suppressWarnings(
  corpairs(
    fit_spatial_q2_example,
    level = "spatial"
  )
)
spatial_q2_pairs
```

```{r spatial-q2-confidence-eye, fig.width = 6.4, fig.height = 2.7, fig.cap = "Coordinate-spatial intercept-only q=2 location-location point estimate from `corpairs()`; the dotted vertical line marks zero correlation and no interval is shown because calibration remains planned.", fig.alt = "Single-row point plot for the coordinate-spatial mean-mean correlation, with a hollow point estimate slightly to the right of the dotted zero reference line."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  spatial_q2_display <- spatial_q2_pairs
  spatial_q2_display$display_label <- "Spatial\nmu1-mu2"

  plot_corpairs(
    spatial_q2_display,
    colour = "level",
    label = "display_label",
    facet = NULL
  ) +
    ggplot2::scale_colour_manual(values = c("spatial" = "#009E73")) +
    ggplot2::scale_fill_manual(values = c("spatial" = "#009E73")) +
    spatial_eye_theme() +
    ggplot2::labs(
      title = "Spatial latent mean correlation",
      subtitle = "Point estimate only; dotted line marks zero",
      x = "Correlation estimate"
    )
}
```

## Boundaries

The following spatial routes remain planned:

- mesh or SPDE inputs such as `spatial(1 | site, mesh = mesh)`;
- multiple spatial slopes and spatial slope correlations;
- partial spatial terms in `sigma`, plus spatial terms in `nu`,
  zero-inflation, or `rho12`;
- direct spatial SD surfaces;
- predictor-dependent spatial `corpair()` regressions;
- simultaneous `phylo()` plus `spatial()` layers in the same formula;
- non-Gaussian spatial structured effects outside the exact ordinary Poisson/NB2 q1 spatial `mu` intercept-plus-one-slope, recovery-grade NB2 q1 spatial `sigma`, Student-t spatial `mu`, Poisson spatial `zi`, fixed-`zi` Poisson spatial `mu`, and fixed-`zi` NB2 spatial `mu` gates.

Use the [structural-dependence overview](structural-dependence.html) when you
are choosing among `animal()`, `phylo()`, `spatial()`, and `relmat()`. Use the
detailed [structural-dependence tutorial](phylogenetic-spatial.html) when you
need the current worked examples, equations, and broader parity ladder.
    ggplot2::facet_grid(unit ~ ., scales = "free_x", space = "free_y") +
    ggplot2::theme(strip.text.y = ggplot2::element_text(angle = 0)) +
