---
title: "Temporal AR1 and OU random effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal AR1 and OU random effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Repeated measurements can vary for three different reasons. Individuals or
sites may have stable baseline differences, their deviations may persist through
time, and each observation has independent residual noise. Gaussian temporal
AR1 and OU models separate those components while estimating the effects of
treatments or other predictors.

For observation \\(t\\) in series \\(i\\), the model is

\\[
y_{it}=x_{it}^{T}\\beta+b_i+a_{it}+\\epsilon_{it},
\\]

where \\(b_i\\) is the stable series intercept, \\(a_{it}\\) is a stationary
temporal process with covariance
\\(\\operatorname{Cov}(a_{it},a_{is})=s_a^2\\phi^{|t-s|}\\), and
\\(\\epsilon_{it}\\) is independent residual noise. The process SD \\(s_a\\),
ordinary-intercept SD \\(s_b\\), and residual SD \\(\\sigma\\) measure different
sources of variation. Persistence \\(\\phi\\) is the correlation one occasion
apart and can be negative or positive.

## A repeated-site example

This example creates measurements from sites visited on genuinely irregular
integer occasions. The gap between occasions 1 and 3 remains a two-occasion
gap when the model is fitted.

```{r}
set.seed(20260908)
n_site <- 48
occasion_values <- c(0L, 1L, 3L, 4L, 6L, 7L)
n_time <- length(occasion_values)

site <- factor(rep(paste0("site_", seq_len(n_site)), each = n_time))
occasion <- rep(occasion_values, n_site)
treatment <- rep(rep(c(0, 1), length.out = n_time), n_site)

R_ar1 <- 0.55 ^ abs(outer(occasion_values, occasion_values, "-"))
temporal_effect <- unlist(lapply(seq_len(n_site), function(i) {
  as.vector(t(chol(R_ar1)) %*% rnorm(n_time, sd = 0.70))
}), use.names = FALSE)
site_intercept <- rep(rnorm(n_site, sd = 0.50), each = n_time)

dat <- data.frame(
  y = 1.2 + 0.45 * treatment + 0.08 * occasion +
    site_intercept + temporal_effect + rnorm(n_site * n_time, sd = 0.45),
  treatment = treatment,
  occasion = occasion,
  site = site
)
```

Fit an ordinary random intercept and the temporal process together. Both terms
use `site`; the first captures stable site-to-site differences and the second
captures deviations through time within each site. Here `treatment` compares
treated and untreated visits within sites after accounting for a linear
occasion trend.

```{r}
fit <- drmTMB(
  bf(
    y ~ treatment + occasion +
      (1 | site) +
      temporal(1 | site, time = occasion, structure = "ar1"),
    sigma ~ 1
  ),
  data = dat,
  family = gaussian(),
  REML = FALSE
)

summary(fit, conf.int = TRUE, method = "wald")
confint(fit, method = "wald")
```

The mean-model coefficient rows report the intercept, treatment, and occasion
effects with normal-reference Wald intervals. `fit$sdpars$mu` contains separate
ordinary-intercept and temporal process SDs, while `fit$corpars$temporal` contains persistence. Read a positive
persistence estimate as deviations tending to continue in the same direction
over adjacent occasions; read a negative value as alternating deviations.

```{r}
fit$sdpars$mu
fit$corpars$temporal
head(fitted(fit))
head(residuals(fit))
simulate(fit, nsim = 2)
simulate(fit, nsim = 2, re.form = NA)
```

`fitted()` and `residuals()` are conditional on the estimated site and temporal
deviations. The default `simulate()` call draws fresh stable and temporal
effects, while `re.form = NA` retains the fitted conditional effects. Use the
first for a new realization from the fitted model and the second to simulate
around the observations' fitted latent structure.

## What the intervals cover

For temporal fits, `vcov()` and Wald `confint()` deliberately expose only mean
coefficients, and only for AR1. Their covariance comes from the full observed
marginal-likelihood Hessian, including estimated variance and persistence
parameters. The process SD, ordinary-intercept SD, persistence or decay, and
residual SD are point estimates in this release.

Likelihood profiles are available for temporal mean coefficients in both AR1
and OU models. Use them when reporting a treatment or other regression effect:

```{r, eval = FALSE}
confint(fit, parm = "mu:treatment", method = "profile")
summary(fit, conf.int = TRUE, method = "profile", ci_parm = "mu:treatment")
```

Omit `parm` or `ci_parm` to profile every mean coefficient. Variance-component,
persistence or decay, bootstrap, forecasting, and `newdata` intervals remain
unavailable. `check_drm(fit)` labels a non-positive-definite fitted Hessian as
an irregular fitted-likelihood situation: a mean-coefficient profile can
still be finite there, but it is not coverage-calibrated. The first calibration
pilot found a residual-SD boundary in a primary short-series case, so neither
the Wald nor profile route is yet a general coverage claim.

The fit supports in-sample predictions, fitted values, residuals, and
simulation. It does not support forecasting or `newdata` prediction. Supply a
finite integer time column, retain its real gaps, and ensure that every
site--occasion pair is unique. If a site was measured more than once at an
occasion, aggregate those records before fitting. An AR1-only fit omits `(1 | site)` but keeps the
same `temporal()` term.

## Irregular elapsed time with OU

Use an OU field when elapsed intervals carry meaning, such as visits at 0, 0.5,
2.5, and 5 days. It keeps the same three-way variance separation but estimates
a positive decay rate, so correlation over a gap `d` is `exp(-decay * d)`. It
cannot represent alternating negative temporal correlation.

```{r}
set.seed(20260909)
elapsed_values <- c(0, 0.5, 2.5, 5)
n_ou_site <- 24
site_ou <- factor(rep(paste0("ou_site_", seq_len(n_ou_site)), each = length(elapsed_values)))
elapsed_days <- rep(elapsed_values, n_ou_site)
treatment_ou <- rep(rep(c(0, 1), length.out = length(elapsed_values)), n_ou_site)

R_ou <- exp(-0.45 * abs(outer(elapsed_values, elapsed_values, "-")))
ou_effect <- unlist(lapply(seq_len(n_ou_site), function(i) {
  as.vector(t(chol(R_ou)) %*% rnorm(length(elapsed_values), sd = 0.65))
}), use.names = FALSE)
dat_irregular <- data.frame(
  y = 0.8 + 0.35 * treatment_ou +
    rep(rnorm(n_ou_site, sd = 0.45), each = length(elapsed_values)) +
    ou_effect + rnorm(n_ou_site * length(elapsed_values), sd = 0.40),
  treatment = treatment_ou,
  site = site_ou,
  elapsed_days = elapsed_days
)

ou_fit <- drmTMB(
  bf(y ~ treatment + (1 | site) +
       temporal(1 | site, time = elapsed_days, structure = "ou"), sigma ~ 1),
  data = dat_irregular, family = gaussian(), REML = FALSE
)
summary(ou_fit)
ou_fit$decaypars$temporal
confint(ou_fit, parm = "mu:treatment", method = "profile")
check_drm(ou_fit)
head(fitted(ou_fit))
simulate(ou_fit, nsim = 2)
```

For OU, `elapsed_days` must be finite numeric, genuine gaps are retained, and
duplicate site--time records must be aggregated before fitting. Input order may
be shuffled. In-sample fitted values, residuals, and conditional or fresh
simulation follow the AR1 conventions. Forecasting and `newdata` prediction
remain unavailable. `ou_fit$decaypars$temporal` is the positive decay rate in
the units of `elapsed_days`: if you change days to hours, the numerical rate
changes by the inverse factor while the fitted correlation over a physical gap
does not. OU Wald inference through `vcov()`, `confint(..., method = "wald")`,
and `summary(..., conf.int = TRUE, method = "wald")` remains unavailable.
Use `confint(..., parm = "mu:<coefficient>", method = "profile")` for a mean
effect, read its `conf.status`, and inspect `check_drm(ou_fit)` before
reporting it. Report it only when `check_drm(ou_fit)` has no
`temporal_mean_profile` warning; finite endpoints alone are not calibration
evidence. OU decay and variance-component intervals remain unavailable.
