---
title: "Temporal AR1 random effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal AR1 random effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

Repeated measurements can vary for three different reasons. Individuals or
sites may have stable baseline differences, their deviations may persist from
one occasion to the next, and each observation has independent residual noise.
The Gaussian temporal AR1 model separates those components while estimating
the effects of treatments or other predictors.

For observation \\(t\\) in series \\(i\\), the model is

\\[
y_{it}=x_{it}^{T}\\beta+b_i+a_{it}+\\epsilon_{it},
\\]

where \\(b_i\\) is the stable series intercept, \\(a_{it}\\) is a stationary
temporal process with covariance
\\(\\operatorname{Cov}(a_{it},a_{is})=s_a^2\\phi^{|t-s|}\\), and
\\(\\epsilon_{it}\\) is independent residual noise. The process SD \\(s_a\\),
ordinary-intercept SD \\(s_b\\), and residual SD \\(\\sigma\\) measure different
sources of variation. Persistence \\(\\phi\\) is the correlation one occasion
apart and can be negative or positive.

## A repeated-site example

This example creates measurements from sites visited on genuinely irregular
integer occasions. The gap between occasions 1 and 3 remains a two-occasion
gap when the model is fitted.

```{r}
set.seed(20260908)
n_site <- 48
occasion_values <- c(0L, 1L, 3L, 4L, 6L, 7L)
n_time <- length(occasion_values)

site <- factor(rep(paste0("site_", seq_len(n_site)), each = n_time))
occasion <- rep(occasion_values, n_site)
treatment <- rep(rep(c(0, 1), length.out = n_time), n_site)

R_ar1 <- 0.55 ^ abs(outer(occasion_values, occasion_values, "-"))
temporal_effect <- unlist(lapply(seq_len(n_site), function(i) {
  as.vector(t(chol(R_ar1)) %*% rnorm(n_time, sd = 0.70))
}), use.names = FALSE)
site_intercept <- rep(rnorm(n_site, sd = 0.50), each = n_time)

dat <- data.frame(
  y = 1.2 + 0.45 * treatment + 0.08 * occasion +
    site_intercept + temporal_effect + rnorm(n_site * n_time, sd = 0.45),
  treatment = treatment,
  occasion = occasion,
  site = site
)
```

Fit an ordinary random intercept and the temporal process together. Both terms
use `site`; the first captures stable site-to-site differences and the second
captures deviations through time within each site.

```{r}
fit <- drmTMB(
  bf(
    y ~ treatment + occasion +
      (1 | site) +
      temporal(1 | site, time = occasion, structure = "ar1"),
    sigma ~ 1
  ),
  data = dat,
  family = gaussian(),
  REML = FALSE
)

summary(fit, conf.int = TRUE, method = "wald")
confint(fit, method = "wald")
```

The coefficient rows report treatment and occasion effects with normal-reference
Wald intervals. `fit$sdpars$mu` contains separate ordinary-intercept and temporal
process SDs, while `fit$corpars$temporal` contains persistence. Read a positive
persistence estimate as deviations tending to continue in the same direction
over adjacent occasions; read a negative value as alternating deviations.

```{r}
fit$sdpars$mu
fit$corpars$temporal
head(fitted(fit))
head(residuals(fit))
simulate(fit, nsim = 2)
```

## What the intervals cover

For this first temporal route, `vcov()` and `confint()` deliberately expose
uncertainty for the fixed mean coefficients only. Their covariance comes from
the full observed marginal-likelihood Hessian, including the estimated
variance and persistence parameters. The process SD, ordinary-intercept SD,
persistence, and residual SD are point estimates in this release. Profile and
bootstrap intervals for temporal fits are not available.

The fit supports in-sample predictions, fitted values, residuals, and
simulation. It does not support forecasting or `newdata` prediction. Supply a
finite integer time column, retain its real gaps, and ensure that every
site--occasion pair is unique. An AR1-only fit omits `(1 | site)` but keeps the
same `temporal()` term.
