---
title: "Testing likelihoods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing likelihoods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This article describes how `drmTMB` likelihoods should be checked before they
are treated as implemented. It is written for package contributors: the goal is
to make each model defensible by matching the symbolic model, the R syntax, the
TMB likelihood, and the tests.

The planning reference is `docs/design/05-testing-strategy.md`. This article is
the practical version: what to put in a pull request or phase before asking
users to trust a model.

In this guide, location means a mean-like parameter such as `mu`; scale means a
residual spread or dispersion-like parameter such as `sigma`; shape means a
family-specific parameter such as Student-t `nu`; and coscale means residual
coupling, represented in the bivariate Gaussian model by `rho12`.

## The contract

Every fitted model should have the same four-part contract:

1. a symbolic model that defines the likelihood;
2. the `drmTMB` syntax that represents that model;
3. an independent likelihood, simulation, or comparator check;
4. a clear statement of what is still unsupported.

For example, the fixed-effect Gaussian location-scale model is:

```text
y_i | mu_i, sigma_i ~ Normal(mu_i, sigma_i^2)
mu_i = beta_0 + beta_1 x_i
log(sigma_i) = gamma_0 + gamma_1 x_i
```

The matching user syntax is:

```{r gaussian-ls-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(
    y ~ x,
    sigma ~ x
  ),
  family = gaussian(),
  data = dat
)
```

The test should then simulate from the same equations, fit the same model, and
check recovery of `beta` and `gamma`. If the equation, syntax, or TMB template
changes, the test should change in the same commit.

## Tier 1: comparator checks

Comparator checks compare `drmTMB` with established software in an overlapping
case. They are not enough by themselves, because no other package has exactly
the whole `drmTMB` grammar. They are still powerful: they catch likelihood
constants, scale definitions, covariance assembly, and optimizer plumbing.

The simplest Gaussian mixed-model comparator is:

```text
y_ij | b_j, sigma ~ Normal(beta_0 + beta_1 x_ij + b_j, sigma^2)
b_j ~ Normal(0, sd_id^2)
```

with syntax:

```{r lme4-comparator-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(y ~ x + (1 | id)),
  family = gaussian(),
  data = dat
)

fit_lme4 <- lme4::lmer(
  y ~ x + (1 | id),
  data = dat,
  REML = FALSE
)
```

The comparator test should check fixed effects, random-effect standard
deviations, residual scale, and `logLik()`. Checking the log likelihood matters
because equal coefficients can hide a wrong likelihood constant or covariance
term.

Meta-analysis gives another useful comparator. The implemented dense known-`V`
Gaussian model is:

```text
y ~ Normal(X beta, V + sigma^2 I)
```

where `V` is known sampling covariance and `sigma^2` is residual
heterogeneity. The matching `drmTMB` syntax is:

```{r dense-meta-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(yi ~ x + meta_known_V(V = V)),
  family = gaussian(),
  data = dat
)
```

For the overlapping model, the comparator is:

```{r metafor-comparator-syntax, eval = FALSE}
fit_metafor <- metafor::rma.mv(
  yi = yi,
  V = V,
  mods = ~ x,
  random = ~ 1 | obs,
  data = dat,
  method = "ML"
)
```

Here `obs` has one level per row, so `metafor` estimates one constant
observation-level heterogeneity variance. The corresponding `drmTMB` test
should compare:

```{r metafor-comparator-checks, eval = FALSE}
expect_equal(unname(coef(fit, "mu")),
             unname(stats::coef(fit_metafor)),
             tolerance = 1e-4)
expect_equal(stats::sigma(fit)[[1L]]^2,
             fit_metafor$sigma2[[1L]],
             tolerance = 1e-4)
expect_equal(as.numeric(stats::logLik(fit)),
             as.numeric(stats::logLik(fit_metafor)),
             tolerance = 1e-4)
```

Optional comparator packages should use `testthat::skip_if_not_installed()` so
routine checks stay portable.

Count models also need comparator checks when another package has the same
parameterization. For the overlapping negative-binomial 2 case with constant
overdispersion, `MASS::glm.nb()` estimates `theta = size`, whereas `drmTMB`
uses `sigma = 1 / sqrt(size)`:

```text
y_i ~ NB2(mu_i, size)
log(mu_i) = beta_0 + beta_1 x_i
sigma = 1 / sqrt(size)
```

```{r mass-nb-comparator-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(count ~ x, sigma ~ 1),
  family = nbinom2(),
  data = dat
)

fit_mass <- MASS::glm.nb(count ~ x, data = dat)
```

The comparator should check the `mu` coefficients, the transformed
overdispersion scale, and `logLik()`:

```{r mass-nb-comparator-checks, eval = FALSE}
expect_equal(unname(coef(fit, "mu")),
             unname(stats::coef(fit_mass)),
             tolerance = 1e-4)
expect_equal(sigma(fit)[[1L]],
             1 / sqrt(fit_mass$theta),
             tolerance = 1e-4)
expect_equal(as.numeric(stats::logLik(fit)),
             as.numeric(stats::logLik(fit_mass)),
             tolerance = 1e-4)
```

For beta mean-scale models, the always-available comparator is the base R
`stats::dbeta()` density at the fitted coefficients. The public `sigma`
parameter maps to precision as `phi = 1 / sigma^2`:

```text
logit(mu_i) = X_mu[i, ] beta_mu
log(sigma_i) = X_sigma[i, ] beta_sigma
phi_i = 1 / sigma_i^2
alpha_i = mu_i phi_i
beta_i = (1 - mu_i) phi_i
```

```{r beta-comparator-checks, eval = FALSE}
eta_mu <- as.vector(fit$model$X$mu %*% coef(fit, "mu"))
eta_sigma <- as.vector(fit$model$X$sigma %*% coef(fit, "sigma"))
mu <- plogis(eta_mu)
sigma <- exp(eta_sigma)
phi <- 1 / sigma^2
ll_independent <- sum(stats::dbeta(
  fit$model$y,
  shape1 = mu * phi,
  shape2 = (1 - mu) * phi,
  log = TRUE
))
expect_equal(as.numeric(logLik(fit)), ll_independent, tolerance = 1e-6)
```

## Tier 2: simulation recovery

Simulation recovery is the main internal validation source for new `drmTMB`
features. It should be used whenever a model has no exact comparator, or
whenever the package adds a distributional parameter that other software does
not model in the same way.

A simulation recovery test should:

1. choose known parameters on the link scale;
2. simulate from the symbolic model;
3. fit the matching `drmTMB` model;
4. check optimizer convergence;
5. check estimates on the scale that is most meaningful for the parameter;
6. include at least one edge or malformed-input case nearby.

For the Student-t location-scale-shape model, the implemented equations are:

```text
y_i | mu_i, sigma_i, nu_i ~ Student-t(mu_i, sigma_i, nu_i)
mu_i = beta_0 + beta_1 x_i
log(sigma_i) = gamma_0 + gamma_1 x_i
nu_i = 2 + exp(delta_0 + delta_1 x_i)
```

The matching syntax is:

```{r student-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(
    y ~ x,
    sigma ~ x,
    nu ~ x
  ),
  family = student(),
  data = dat
)
```

The test should check that `nu` stays above 2, because the current
parameterization makes the residual variance finite. For a fitted Student-t
model, `sigma` is a scale parameter; the residual standard deviation is
`sigma * sqrt(nu / (nu - 2))` when `nu > 2`.

## Independent likelihood checks

When the density is simple enough, add a likelihood check that does not call the
TMB objective. This is especially useful for the first implementation of a
family. For the Student-t model above, an independent R calculation has the
form:

```{r independent-student-likelihood, eval = FALSE}
eta_mu <- X_mu %*% beta_mu
eta_sigma <- X_sigma %*% beta_sigma
eta_nu <- X_nu %*% beta_nu

mu <- as.vector(eta_mu)
sigma <- exp(as.vector(eta_sigma))
nu <- 2 + exp(as.vector(eta_nu))

loglik <- sum(
  stats::dt((y - mu) / sigma, df = nu, log = TRUE) - log(sigma)
)
```

The test should compare `logLik(fit)` with `loglik` at the fitted parameter
values. This catches transform mistakes such as forgetting the Jacobian-like
`-log(sigma)` density adjustment for scaled Student-t residuals.

For zero-truncated NB2, the independent likelihood must include the
normalizing probability that the untruncated NB2 count component is positive:

```text
y_i | y_i > 0, mu_i, sigma_i ~ truncated NB2(mu_i, sigma_i)
Pr_trunc(y_i) = Pr_NB2(y_i) / (1 - Pr_NB2(0))
```

```{r independent-truncated-nbinom2-likelihood, eval = FALSE}
eta_mu <- X_mu %*% beta_mu
eta_sigma <- X_sigma %*% beta_sigma

mu <- exp(as.vector(eta_mu))
sigma <- exp(as.vector(eta_sigma))
size <- 1 / sigma^2

log_p0 <- stats::dnbinom(0, mu = mu, size = size, log = TRUE)
log_positive <- log1p(-exp(log_p0))
loglik <- sum(
  stats::dnbinom(y, mu = mu, size = size, log = TRUE) - log_positive
)
```

The matching test should also check that all responses are positive integers,
that zeros are rejected, that `predict(fit, dpar = "mu")` returns the
untruncated NB2 component mean, and that `fitted(fit)` returns
`mu / (1 - Pr_NB2(0))`.

For hurdle NB2, the independent likelihood separates the zero process from the
positive truncated NB2 component:

```text
Pr(y_i = 0) = hu_i
Pr(y_i = k > 0) = (1 - hu_i) Pr_NB2(k) / (1 - Pr_NB2(0))
```

```{r independent-hurdle-nbinom2-likelihood, eval = FALSE}
eta_mu <- X_mu %*% beta_mu
eta_sigma <- X_sigma %*% beta_sigma
eta_hu <- X_hu %*% beta_hu

mu <- exp(as.vector(eta_mu))
sigma <- exp(as.vector(eta_sigma))
hu <- plogis(as.vector(eta_hu))
size <- 1 / sigma^2

log_p0 <- stats::dnbinom(0, mu = mu, size = size, log = TRUE)
log_positive <- log1p(-exp(log_p0))
loglik <- sum(ifelse(
  y == 0,
  log(hu),
  log1p(-hu) + stats::dnbinom(y, mu = mu, size = size, log = TRUE) -
    log_positive
))
```

The matching test should also check that zeros are accepted only when `hu` is
present, that `predict(fit, dpar = "hu")` returns a probability, and that
`fitted(fit)` returns `(1 - hu) * mu / (1 - Pr_NB2(0))`.

## Bivariate and coscale checks

The first bivariate Gaussian location-coscale model uses:

```text
[y1_i, y2_i]' ~ MVN([mu1_i, mu2_i]', Omega_i)
Omega_i =
  [ sigma1_i^2                         rho12_i sigma1_i sigma2_i ]
  [ rho12_i sigma1_i sigma2_i          sigma2_i^2                ]
eta_rho12_i = X_rho12[i, ] beta_rho12
rho12_i = 0.99999999 * tanh(eta_rho12_i)
```

The matching syntax is:

```{r bivariate-rho12-syntax, eval = FALSE}
fit <- drmTMB(
  drm_formula(
    mu1 = y1 ~ x1 + x2,
    mu2 = y2 ~ x1,
    sigma1 = ~ x1 + x2,
    sigma2 = ~ x1,
    rho12 = ~ x1 + x2
  ),
  family = c(gaussian(), gaussian()),
  data = dat
)
```

Tests for this model should include `rho12 = 0`, moderate negative and
positive `rho12`, high negative and positive `rho12` near `+/-0.8`,
predictor-dependent `rho12`, and unequal `sigma1` and `sigma2`.

For bivariate meta-analysis, keep two covariance objects separate:

```text
S_i[1,2] = known within-study sampling covariance
Omega_i[1,2] = rho12_i sigma1_i sigma2_i
```

In the current implementation, bivariate `meta_known_V(V = V)` must evaluate to
a complete-row `2n` by `2n` row-paired covariance matrix, with
`meta_vcov_bivariate()` available as the helper. Conceptually this stacked
matrix contains the known within-study sampling covariance blocks `S_i`.
`rho12` models the unknown residual or between-study coupling. A valid test must
show that fitted `rho12` is not merely reproducing the known sampling
correlation.

## Boundary and rejection tests

Likelihood tests should include unsupported syntax and boundary-prone cases.
For early phases, rejecting a model clearly is a feature, because it protects
users from syntax that looks plausible but has no fitted likelihood path.

Examples of rejection tests:

```{r rejection-test-examples, eval = FALSE}
expect_snapshot(
  {
    drmTMB(
      drm_formula(y ~ x + meta_known_V(V = V), nu ~ x),
      family = student(),
      data = dat
    )
  },
  error = TRUE
)

expect_snapshot(
  {
    drmTMB(
      drm_formula(
        mu1 = y1 ~ x + (1 | id),
        mu2 = y2 ~ x,
        sigma1 = ~ 1,
        sigma2 = ~ 1,
        rho12 = ~ 1
      ),
      family = c(gaussian(), gaussian()),
      data = dat
    )
  },
  error = TRUE
)
```

The exact messages may change, but the rule should not: an unsupported
combination should fail before optimization with an explanation of the missing
feature.

## Fast and long tests

CRAN-facing tests should be small and deterministic. They answer whether the
likelihood is wired correctly for a few representative cases.

Long tests belong in optional scripts, scheduled CI, or a simulation study. They
answer whether the estimator behaves well across sample sizes, effect sizes,
mesh densities, tree sizes, boundary variances, and extreme correlations.

Use this split deliberately:

| Test type | Runs in routine checks | Purpose |
|---|---:|---|
| Unit test | yes | parser, family, extractor, and error-message behaviour |
| Independent likelihood check | yes | density constants, transforms, covariance assembly |
| Small simulation recovery | yes | basic estimator recovery |
| Comparator smoke test | yes, with `skip_if_not_installed()` | parity with established software |
| Large simulation grid | no | operating characteristics across parameter space |
| Bootstrap or profile-likelihood coverage | no, until scheduled | uncertainty calibration |

## Before closing a task

Before a likelihood task is complete, run the after-task audit. The compact
checklist is:

- the symbolic equation, R syntax, and TMB implementation describe the same
  model;
- tests include either an independent likelihood check, a comparator check, or
  simulation recovery;
- at least one unsupported or malformed neighbouring case is checked;
- `docs/design/03-likelihoods.md`, `docs/design/05-testing-strategy.md`, and
  relevant vignettes are synchronized;
- `pkgdown::check_pkgdown()` and `pkgdown::build_site()` pass for user-facing
  documentation changes;
- `docs/dev-log/check-log.md` and an after-task report record the checks, what
  did not go smoothly, and what remains unsupported.

This discipline is deliberately slow at the end of each task. It is much faster
than debugging a package whose equations, examples, tests, and documentation
have drifted apart.
