---
title: "Which scale are you modelling?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which scale are you modelling?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(drmTMB)
```

`drmTMB` deliberately separates residual scale from group-level scale. This
article pairs each piece of R syntax with the symbolic model it represents.
That pairing is the guardrail: before adding a feature to the package, the R
syntax, equations, TMB parameters, tests, and documentation should describe the
same model.

Throughout this article, `Normal(a, b)` uses variance as the second argument.
Thus `Normal(mu_i, sigma_i^2)` means a Gaussian distribution with mean `mu_i`
and standard deviation `sigma_i`.

For non-Gaussian families, `sigma` is still the public variability-facing scale,
but its family-specific meaning changes. For example, beta and NB2 likelihoods
use precision-like internal parameters, while `drmTMB` reports `sigma` so that
larger fitted values mean larger modelled variability. See
`vignette("distribution-families")` for the conversion table before comparing
`sigma` with `phi`, `theta`, or other software-specific dispersion parameters.

## Residual standard deviation

Implemented:

```r
drmTMB(
  bf(growth ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Symbolically:

```text
growth_i | mu_i, sigma_i ~ Normal(mu_i, sigma_i^2)
mu_i = beta_0 + beta_1 temperature_i
log(sigma_i) = gamma_0 + gamma_1 temperature_i
sigma_i = exp(gamma_0 + gamma_1 temperature_i)
```

The `sigma` formula models the residual standard deviation. If `gamma_1` is
positive, warmer observations have a larger residual SD around their fitted
mean. The coefficient is on the log-SD scale, so `exp(gamma_1)` is the
multiplicative change in residual SD for a one-unit change in temperature.

## Mean-model random intercept

Implemented:

```r
drmTMB(
  bf(growth ~ temperature + (1 | population), sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Symbolically:

```text
growth_ij | mu_ij, sigma_ij, b_j ~ Normal(mu_ij, sigma_ij^2)
mu_ij = beta_0 + beta_1 temperature_ij + b_j
log(sigma_ij) = gamma_0 + gamma_1 temperature_ij
b_j = sd_mu_population u_j
u_j ~ Normal(0, 1)
```

Here `sd_mu_population` is the among-population SD in expected growth. It is a
group-level scale parameter for the mean model. It is not the residual
standard deviation `sigma_ij`.

## Residual-scale random intercept

Implemented:

```r
drmTMB(
  bf(
    growth ~ temperature + (1 | population),
    sigma ~ temperature + (1 | population)
  ),
  family = gaussian(),
  data = fish
)
```

Symbolically:

```text
growth_ij | mu_ij, sigma_ij, b_j, a_j ~ Normal(mu_ij, sigma_ij^2)
mu_ij = beta_0 + beta_1 temperature_ij + b_j
log(sigma_ij) = gamma_0 + gamma_1 temperature_ij + a_j

b_j = sd_mu_population u_j
u_j ~ Normal(0, 1)

a_j = sd_sigma_population v_j
v_j ~ Normal(0, 1)
```

This asks whether populations differ in their residual variability after the
fixed effects in the `sigma` formula have been accounted for. It is
group-to-group variation in within-population variability.

It is not the same as asking whether the among-population SD in mean growth
depends on a predictor. That is the `sd(population) ~ x_group` model.

## Random-effect scale formula

Implemented for one or more distinct unlabelled Gaussian `mu` random
intercepts:

```r
drmTMB(
  bf(
    growth ~ temperature + (1 | population),
    sigma ~ temperature,
    sd(population) ~ habitat
  ),
  family = gaussian(),
  data = fish
)
```

Symbolically:

```text
growth_ij | mu_ij, sigma_ij, b_j ~ Normal(mu_ij, sigma_ij^2)
mu_ij = beta_0 + beta_1 temperature_ij + b_j
log(sigma_ij) = gamma_0 + gamma_1 temperature_ij

b_j = sd_mu_population,j u_j
u_j ~ Normal(0, 1)
log(sd_mu_population,j) = alpha_0 + alpha_1 habitat_j
```

The right-hand side of `sd(population) ~ habitat` is group-level. `habitat_j`
must be constant within each population after missing rows are removed.

This model asks whether among-population differences in expected growth are
larger in some habitats. That is a different question from `sigma ~ habitat`,
which asks whether individual observations are more variable around their
fitted mean in some habitats.

A compact copy-run example:

```r
set.seed(1)
n_population <- 24
n_each <- 8
population_info <- data.frame(
  population = factor(seq_len(n_population)),
  habitat = rep(c("closed", "open"), length.out = n_population)
)
fish <- population_info[rep(seq_len(n_population), each = n_each), ]
fish$temperature <- rnorm(nrow(fish))

sd_population <- exp(-0.4 + 0.5 * (population_info$habitat == "open"))
b_population <- rnorm(n_population, sd = sd_population)
sigma <- exp(-0.6 + 0.25 * fish$temperature)
fish$growth <- 1 + 0.7 * fish$temperature +
  b_population[fish$population] +
  rnorm(nrow(fish), sd = sigma)

fit <- drmTMB(
  bf(
    growth ~ temperature + (1 | population),
    sigma ~ temperature,
    sd(population) ~ habitat
  ),
  family = gaussian(),
  data = fish
)

coef(fit, "sd(population)")
head(predict(fit, dpar = "sd(population)"))
```

The `sd(population)` coefficients are on the log-SD scale. A positive
`habitatopen` coefficient means the estimated among-population SD in expected
growth is larger for open-habitat populations than for the baseline habitat.

## One location model, several scale quantities

A model can have one location predictor and several scale quantities. For
example, suppose fish growth is measured across populations and sites:

```r
drmTMB(
  bf(
    growth ~ temperature + (1 | population) + (1 | site),
    sigma ~ temperature,
    sd(population) ~ habitat,
    sd(site) ~ site_area
  ),
  family = gaussian(),
  data = fish
)
```

The matching symbolic model is:

```text
growth_i | mu_i, sigma_i, b_j[i], c_k[i] ~ Normal(mu_i, sigma_i^2)
mu_i = beta_0 + beta_1 temperature_i + b_j[i] + c_k[i]
log(sigma_i) = gamma_0 + gamma_1 temperature_i

b_j = sd_mu_population,j u_j
u_j ~ Normal(0, 1)
log(sd_mu_population,j) = alpha_0 + alpha_1 habitat_j

c_k = sd_mu_site,k r_k
r_k ~ Normal(0, 1)
log(sd_mu_site,k) = kappa_0 + kappa_1 site_area_k
```

This has three scale quantities:

| Quantity | Formula | Interpretation |
|---|---|---|
| `sigma_i` | `sigma ~ temperature` | residual SD for each observation |
| `sd_mu_population,j` | `sd(population) ~ habitat` | among-population SD in expected growth |
| `sd_mu_site,k` | `sd(site) ~ site_area` | among-site SD in expected growth |

There are no random effects in the residual scale unless the `sigma` formula
contains a term such as `(1 | population)`. There is also no correlation
between the population and site random effects in this example; they are
separate grouping factors. Correlated random intercept-slope blocks require a
single grouping factor block such as `(1 + temperature | population)`.

## A copy-run scale audit

This section uses one small ecological example to show what each scale means
in fitted output. Suppose `growth` is a growth-rate measurement from several
populations. Temperature may change the expected growth rate and the residual
SD around that expectation. Populations may also differ in their mean growth,
and that among-population SD may be larger in one habitat.

```{r scale-audit-data}
set.seed(42)
n_population <- 32
n_each <- 6

population_info <- data.frame(
  population = factor(seq_len(n_population)),
  habitat = rep(c("forest", "grassland"), length.out = n_population)
)

fish <- population_info[rep(seq_len(n_population), each = n_each), ]
fish$temperature <- rnorm(nrow(fish))
fish$reliability <- ifelse(seq_len(nrow(fish)) %% 3 == 0, 2, 1)

pop_sd <- exp(-0.8 + 0.7 * (population_info$habitat == "grassland"))
b_population <- rnorm(n_population, sd = pop_sd)
fish$sigma_true <- exp(-0.7 + 0.3 * fish$temperature)
fish$growth <- 1.2 + 0.55 * fish$temperature +
  b_population[fish$population] +
  rnorm(nrow(fish), sd = fish$sigma_true)
```

### Residual scale: `sigma ~ temperature`

The model

\[
\begin{aligned}
growth_i &\sim \mathrm{Normal}(\mu_i, \sigma_i^2),\\
\mu_i &= \beta_0 + \beta_1 temperature_i,\\
\log(\sigma_i) &= \gamma_0 + \gamma_1 temperature_i
\end{aligned}
\]

is fitted by:

```{r residual-scale-fit}
fit_sigma <- drmTMB(
  bf(growth ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data = fish
)

summary(fit_sigma)
round(coef(fit_sigma, "sigma"), 3)
round(range(sigma(fit_sigma)), 3)
```

The `sigma:temperature` coefficient is on the log-SD scale. A positive value
means residual variation in growth is larger at warmer temperatures, after the
mean temperature effect has been modelled. The displayed range of `sigma(fit)`
is on the response scale, so it is the fitted residual SD, not a variance.

### Likelihood weights: `weights = reliability`

Likelihood weights do not define a new biological scale. They multiply the row
log-likelihood contribution:

\[
\ell(\theta) = \sum_i w_i \log f(y_i \mid \theta_i).
\]

The model syntax stays ordinary:

```{r likelihood-weights-fit}
fit_weighted <- drmTMB(
  bf(growth ~ temperature, sigma ~ 1),
  family = gaussian(),
  data = fish,
  weights = reliability
)

summary(fit_weighted)
head(weights(fit_weighted), 8)
```

Here rows with `reliability = 2` count twice in the likelihood. That may be a
reasonable model-fitting choice for replicated or reliability-weighted rows,
but it is not the same as telling the model that an effect size has known
sampling variance. Known sampling variance belongs in `meta_known_V(V = V)`.
At present, dense full known-covariance models reject non-unit likelihood
weights; keep row weighting and known sampling covariance as separate modelling
decisions unless the fitted model explicitly supports their combination.

### Known sampling variance: `meta_known_V(V = vi)`

For meta-analysis, the known sampling variance is part of the observation
model. In a diagonal Gaussian meta-analysis:

\[
\begin{aligned}
y_i &\sim \mathrm{Normal}(\mu_i, v_i + \sigma_i^2),\\
\mu_i &= \beta_0 + \beta_1 treatment_i,\\
\log(\sigma_i) &= \gamma_0 + \gamma_1 treatment_i .
\end{aligned}
\]

The R syntax puts the known variance in the formula, not in `weights`:

```{r meta-known-v-fit}
set.seed(101)
n_effect <- 50
meta <- data.frame(treatment = rep(c(0, 1), each = n_effect / 2))
meta$vi <- runif(n_effect, 0.02, 0.08)

mu_meta <- 0.1 + 0.25 * meta$treatment
sigma_meta <- exp(-1.1 + 0.35 * meta$treatment)
meta$yi <- rnorm(n_effect, mu_meta, sqrt(meta$vi + sigma_meta^2))

fit_meta <- drmTMB(
  bf(yi ~ treatment + meta_known_V(V = vi), sigma ~ treatment),
  family = gaussian(),
  data = meta
)

summary(fit_meta)

meta_report <- data.frame(
  treatment = meta$treatment,
  known_sampling_variance = meta$vi,
  extra_heterogeneity_sd = sigma(fit_meta)
)
meta_report$extra_heterogeneity_variance <-
  meta_report$extra_heterogeneity_sd^2
meta_report$total_observation_variance <-
  meta_report$known_sampling_variance +
  meta_report$extra_heterogeneity_variance

meta_summary <- aggregate(
  meta_report[c(
    "extra_heterogeneity_sd",
    "extra_heterogeneity_variance",
    "total_observation_variance"
  )],
  by = list(treatment = meta_report$treatment),
  FUN = mean
)
round(meta_summary, 3)
```

The `vi` values are known before the model is fitted. The fitted `sigma`
describes extra heterogeneity after those known sampling variances have been
included. In meta-analysis notation this extra heterogeneity is often written
as `tau`, but `drmTMB` keeps the name `sigma` so Gaussian distributional
models use one stable scale vocabulary. If the scientific summary needs a
variance, square the fitted `sigma` to report extra heterogeneity variance.
If it needs the total observation variance, add the known sampling variance
`vi`. This is a reporting conversion, not a separate `tau ~` formula.

### Random-effect scale: `sd(population) ~ habitat`

Now the question changes again. The model

\[
\begin{aligned}
growth_{ij} &\sim \mathrm{Normal}(\mu_{ij}, \sigma_{ij}^2),\\
\mu_{ij} &= \beta_0 + \beta_1 temperature_{ij} + b_j,\\
b_j &= sd_{population,j} u_j,\quad u_j \sim \mathrm{Normal}(0, 1),\\
\log(sd_{population,j}) &= \alpha_0 + \alpha_1 habitat_j
\end{aligned}
\]

asks whether the among-population SD in expected growth differs by habitat:

```{r random-effect-scale-fit}
fit_sd <- drmTMB(
  bf(
    growth ~ temperature + (1 | population),
    sigma ~ temperature,
    sd(population) ~ habitat
  ),
  family = gaussian(),
  data = fish
)

round(coef(fit_sd, "sd(population)"), 3)
round(tapply(
  predict(fit_sd, dpar = "sd(population)"),
  population_info$habitat,
  mean
), 3)
```

The `habitatgrassland` coefficient is on the log-SD scale for the population
random intercept. In this example, grassland populations have a larger fitted
among-population SD in expected growth. That is not residual variation among
individual observations; it is variation among population-level means.

### Residual coscale: `rho12 ~ treatment`

For two responses, coscale means residual coupling after both response means
and residual SDs have been modelled. In the implemented bivariate Gaussian
model:

\[
\begin{aligned}
\begin{bmatrix} activity_i \\ boldness_i \end{bmatrix}
&\sim \mathrm{MVN}
\left(
\begin{bmatrix} \mu_{1i} \\ \mu_{2i} \end{bmatrix},
\begin{bmatrix}
\sigma_{1i}^2 & \rho_{12i}\sigma_{1i}\sigma_{2i}\\
\rho_{12i}\sigma_{1i}\sigma_{2i} & \sigma_{2i}^2
\end{bmatrix}
\right),\\
\rho_{12i} &= \tanh(\eta_{\rho12,i}),\\
\eta_{\rho12,i} &= \delta_0 + \delta_1 treatment_i .
\end{aligned}
\]

The matching syntax is:

```{r residual-coscale-fit}
set.seed(12)
behaviour <- data.frame(treatment = rep(c(0, 1), each = 50))
Sigma0 <- matrix(c(0.6^2, 0.2 * 0.6 * 0.5,
                   0.2 * 0.6 * 0.5, 0.5^2), 2, 2)
Sigma1 <- matrix(c(0.6^2, 0.65 * 0.6 * 0.5,
                   0.65 * 0.6 * 0.5, 0.5^2), 2, 2)
Y <- matrix(NA_real_, nrow(behaviour), 2)

for (i in seq_len(nrow(behaviour))) {
  Sigma_i <- if (behaviour$treatment[i] == 0) Sigma0 else Sigma1
  mu_i <- c(1 + 0.2 * behaviour$treatment[i],
            0.5 + 0.1 * behaviour$treatment[i])
  Y[i, ] <- as.numeric(mu_i + t(chol(Sigma_i)) %*% rnorm(2))
}

behaviour$activity <- Y[, 1]
behaviour$boldness <- Y[, 2]

fit_rho12 <- drmTMB(
  bf(
    mu1 = activity ~ treatment,
    mu2 = boldness ~ treatment,
    sigma1 = ~ treatment,
    sigma2 = ~ treatment,
    rho12 = ~ treatment
  ),
  family = c(gaussian(), gaussian()),
  data = behaviour
)

summary(fit_rho12)
round(coef(fit_rho12, "rho12"), 3)
round(tapply(rho12(fit_rho12), behaviour$treatment, mean), 3)
```

The `rho12:treatment` coefficient is on the correlation-link scale. The final
line shows the same fitted residual coupling on the response scale. In this
simulated example, treatment changes how tightly activity and boldness remain
coupled after their means and residual SDs have been modelled.

## Side-by-side guide

| R syntax | Equation target | Question |
|---|---|---|
| `sigma ~ x` | `log(sigma_i) = X_sigma[i, ] beta_sigma` | Does `x` change residual SD? |
| `sigma ~ x + (1 | id)` | `log(sigma_i) = X_sigma[i, ] beta_sigma + a_id[i]` | Do groups differ in residual SD? |
| `y ~ x + (1 | id)` | `mu_i = X_mu[i, ] beta_mu + b_id[i]` | Do groups differ in mean response? |
| `sd(id) ~ x_group` | `log(sd_mu_id,j) = W_id[j, ] alpha_id` | Does `x_group` change among-group SD in the mean model? |
| `weights = w` | `logLik = sum_i w_i log f(y_i | theta_i)` | Should row `i` contribute more or less to the likelihood? |
| `meta_known_V(V = V)` | `y ~ MVN(mu, V + Omega_est)` | What sampling covariance is known before fitting? |
| `y ~ x + (1 + x | id)` | `[b_0j, b_1j]' ~ MVN(0, Sigma_id)` | Are group intercepts and slopes correlated? |
| `rho12 ~ x` | `eta_rho12_i = X_rho12[i, ] beta_rho12`; `rho12_i = tanh(eta_rho12_i)` | Does `x` change residual coupling between two responses? |

Implementation detail: the C++ likelihood applies a tiny numerical guard to
the `tanh()` transform so fitted covariance matrices stay strictly positive
definite in floating-point arithmetic. The guard is not a biological scaling
factor. Tutorials and model equations should be read as the standard
Fisher-z-style map from an unconstrained predictor to a correlation.

## Correlations: group-level versus residual

Implemented univariate random-slope correlation:

```r
drmTMB(
  bf(growth ~ temperature + (1 + temperature | population), sigma ~ temperature),
  family = gaussian(),
  data = fish
)
```

Symbolically:

```text
mu_ij = beta_0 + beta_1 temperature_ij + b_0j + temperature_ij b_1j
[b_0j, b_1j]' ~ MVN(0, Sigma_population)
Sigma_population =
  [sd0^2,          rho_re sd0 sd1;
   rho_re sd0 sd1, sd1^2]
```

This `rho_re` is a group-level random-effect correlation. It is reported under
`corpars$mu`.

Implemented bivariate residual correlation:

```r
drmTMB(
  bf(
    mu1 = activity ~ treatment,
    mu2 = boldness ~ treatment,
    sigma1 = ~ treatment,
    sigma2 = ~ treatment,
    rho12 = ~ treatment
  ),
  family = c(gaussian(), gaussian()),
  data = behaviour
)
```

Symbolically:

```text
[activity_i, boldness_i]' ~ MVN([mu1_i, mu2_i]', Omega_i)
Omega_i[1, 2] = rho12_i sigma1_i sigma2_i
eta_rho12_i = delta_0 + delta_1 treatment_i
rho12_i = tanh(eta_rho12_i)
```

Here `rho12_i` is residual response-response correlation within an observation
after the response-specific means and residual SDs have been modelled. If the
correlation is among random intercepts or random slopes, it is a group-level
covariance parameter. If the correlation is between the two residual responses
in one row, it is `rho12`. Extract fitted response-scale residual correlations
with `rho12(fit)`. Use `corpairs(fit)` when you want a table that keeps
residual `rho12` separate from ordinary group-level random-effect correlations,
including the implemented bivariate `mu1`/`mu2` random-intercept correlation.

## Current boundary

Current implemented Gaussian syntax covers:

- `sigma ~ x` for residual SD fixed effects;
- `y ~ x + (1 | id)` for `mu` random intercepts;
- `y ~ x + (0 + x | id)` for independent numeric `mu` random slopes;
- `y ~ x + (1 + x | id)` and `(1 + x | p | id)` for one-slope correlated
  `mu` blocks;
- `sigma ~ x + (1 | id)` for residual-scale random intercepts;
- `sd(id) ~ x_group` and multiple distinct `sd(group) ~ x_group` formulas for
  standard deviations of unlabelled Gaussian `mu` random intercepts;
- bivariate fixed-effect `rho12 ~ x` with `family = c(gaussian(), gaussian())`;
- matching labelled `(1 | p | id)` random intercepts in bivariate `mu1` and
  `mu2`.

Current planned but not implemented syntax includes:

- slope-specific and labelled-block random-effect scale models;
- residual-scale random slopes in `sigma`;
- bivariate random slopes and residual-scale random effects;
- cross-formula covariance sharing from repeated labels such as
  `(1 + x | p | id)`;
- phylogenetic slopes, phylogenetic `sigma` terms, bivariate phylogenetic
  effects, and spatial structured effects.
