---
title: "Phase 18 First-Wave Simulation Summary"
output:
  html_document:
    toc: true
params:
  artifact_status_csv: null
  artifact_grain_status_csv: null
  aggregate_csv: null
  manifest_csv: null
  failures_csv: null
  wald_coverage_csv: null
  profile_coverage_csv: null
  bootstrap_coverage_csv: null
  interval_diagnostics_csv: null
  interval_failures_csv: null
  require_complete: false
  max_rows: 20
  notes: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

phase18_read_optional_csv <- function(path) {
  if (
    is.null(path) ||
      length(path) != 1L ||
      is.na(path) ||
      !nzchar(path)
  ) {
    return(NULL)
  }
  if (!file.exists(path)) {
    return(NULL)
  }
  read.csv(path)
}

phase18_read_required_csv <- function(path, label) {
  data <- phase18_read_optional_csv(path)
  if (is.null(data)) {
    stop("`", label, "` must be an existing CSV path.", call. = FALSE)
  }
  data
}

phase18_select_columns <- function(data, priority = character()) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data)
  }
  front <- intersect(priority, names(data))
  data[c(front, setdiff(names(data), front))]
}

phase18_show_table <- function(data, empty_message, priority = character()) {
  if (is.null(data) || nrow(data) == 0L) {
    cat(empty_message, "\n")
  } else {
    out <- phase18_select_columns(data, priority)
    max_rows <- params$max_rows
    if (
      is.numeric(max_rows) &&
        length(max_rows) == 1L &&
        is.finite(max_rows) &&
        max_rows > 0L &&
        nrow(out) > max_rows
    ) {
      cat("Showing first", max_rows, "of", nrow(out), "rows.\n\n")
      out <- utils::head(out, max_rows)
    }
    out
  }
}

phase18_missing_text <- function(data, name, default) {
  if (!name %in% names(data)) {
    return(rep(default, nrow(data)))
  }
  out <- as.character(data[[name]])
  out[is.na(out) | !nzchar(out)] <- default
  out
}

phase18_failure_summary <- function(data) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data.frame())
  }

  surface <- phase18_missing_text(data, "source_surface", "unknown_surface")
  severity <- phase18_missing_text(data, "severity", "unknown_severity")
  message <- phase18_missing_text(data, "message", "missing message")
  key <- paste(surface, severity, sep = "\r")
  groups <- split(seq_len(nrow(data)), key)
  out <- do.call(
    rbind,
    lapply(groups, function(index) {
      data.frame(
        source_surface = surface[index[1L]],
        severity = severity[index[1L]],
        n_event = length(index),
        n_unique_message = length(unique(message[index])),
        example_message = message[index[1L]],
        stringsAsFactors = FALSE
      )
    })
  )
  row.names(out) <- NULL
  out[order(out$source_surface, out$severity), , drop = FALSE]
}

phase18_bias_overview_data <- function(data) {
  if (is.null(data) || nrow(data) == 0L || !"bias" %in% names(data)) {
    return(NULL)
  }

  out <- data
  out$bias <- suppressWarnings(as.numeric(out$bias))
  out <- out[is.finite(out$bias), , drop = FALSE]
  if (nrow(out) == 0L) {
    return(NULL)
  }

  out$source_surface <- phase18_missing_text(
    out,
    "source_surface",
    "unknown_surface"
  )
  if ("parameter" %in% names(out)) {
    out$parameter_label <- phase18_missing_text(out, "parameter", "parameter")
  } else if (all(c("dpar", "term") %in% names(out))) {
    out$parameter_label <- paste(
      phase18_missing_text(out, "dpar", "dpar"),
      phase18_missing_text(out, "term", "term"),
      sep = ":"
    )
  } else {
    out$parameter_label <- paste0("row_", seq_len(nrow(out)))
  }
  if ("cell_id" %in% names(out)) {
    out$parameter_label <- paste(
      phase18_missing_text(out, "cell_id", "cell"),
      out$parameter_label,
      sep = " / "
    )
  }

  total <- nrow(out)
  out <- out[order(abs(out$bias), decreasing = TRUE), , drop = FALSE]
  max_rows <- params$max_rows
  if (
    is.numeric(max_rows) &&
      length(max_rows) == 1L &&
      is.finite(max_rows) &&
      max_rows > 0L &&
      nrow(out) > max_rows
  ) {
    out <- utils::head(out, max_rows)
  }
  out$bias_rank <- seq_len(nrow(out))
  out$plot_label <- factor(
    paste0("#", out$bias_rank),
    levels = rev(paste0("#", out$bias_rank))
  )
  attr(out, "n_total") <- total
  out
}

phase18_replicate_cloud_gate <- function(data) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data.frame())
  }

  source_surface <- phase18_missing_text(
    data,
    "source_surface",
    "unknown_surface"
  )
  grain_status <- phase18_missing_text(
    data,
    "grain_status",
    "unknown_grain_status"
  )
  replicate_allowed <- if ("replicate_cloud_allowed" %in% names(data)) {
    as.logical(data$replicate_cloud_allowed)
  } else {
    grain_status == "replicate_ready"
  }
  replicate_allowed[is.na(replicate_allowed)] <- FALSE

  key <- source_surface
  groups <- split(seq_len(nrow(data)), key)
  out <- do.call(
    rbind,
    lapply(groups, function(index) {
      n_replicate_ready <- sum(
        replicate_allowed[index] | grain_status[index] == "replicate_ready"
      )
      n_aggregate_only <- sum(grain_status[index] == "aggregate_only")
      n_not_cloud_ready <- length(index) - n_replicate_ready
      gate <- if (n_replicate_ready > 0L) {
        "replicate_clouds_allowed"
      } else if (n_aggregate_only > 0L) {
        "aggregate_only_no_clouds"
      } else {
        "no_replicate_artifact_ready"
      }
      display_rule <- switch(
        gate,
        replicate_clouds_allowed = paste(
          "Replicate-error clouds may use replicate-ready rows for this",
          "surface."
        ),
        aggregate_only_no_clouds = paste(
          "Use aggregate points, bars, and MCSE intervals; do not draw",
          "replicate-error clouds."
        ),
        no_replicate_artifact_ready = paste(
          "Do not draw replicate-error clouds until a replicate-ready",
          "artifact exists."
        )
      )
      data.frame(
        source_surface = source_surface[index[1L]],
        replicate_cloud_gate = gate,
        n_replicate_ready = n_replicate_ready,
        n_aggregate_only = n_aggregate_only,
        n_not_cloud_ready = n_not_cloud_ready,
        display_rule = display_rule,
        stringsAsFactors = FALSE
      )
    })
  )
  row.names(out) <- NULL
  out[order(out$source_surface), , drop = FALSE]
}

phase18_add_replicate_cloud_gate <- function(data, gate) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data)
  }
  if (is.null(gate) || nrow(gate) == 0L) {
    data$replicate_cloud_gate <- "no_grain_status_supplied"
    return(data)
  }
  data$source_surface <- phase18_missing_text(
    data,
    "source_surface",
    "unknown_surface"
  )
  match_index <- match(data$source_surface, gate$source_surface)
  out <- data
  out$replicate_cloud_gate <- gate$replicate_cloud_gate[match_index]
  out$replicate_cloud_gate[is.na(out$replicate_cloud_gate)] <-
    "no_grain_status_supplied"
  out
}

phase18_bind_tables <- function(pieces) {
  pieces <- Filter(function(x) is.data.frame(x) && nrow(x) > 0L, pieces)
  if (length(pieces) == 0L) {
    return(data.frame())
  }
  all_names <- unique(unlist(lapply(pieces, names), use.names = FALSE))
  aligned <- lapply(pieces, function(x) {
    missing <- setdiff(all_names, names(x))
    for (name in missing) {
      x[[name]] <- NA
    }
    x[all_names]
  })
  out <- do.call(rbind, aligned)
  row.names(out) <- NULL
  out
}

phase18_bind_interval_coverage <- function(wald, profile, bootstrap) {
  pieces <- list(wald = wald, profile = profile, bootstrap = bootstrap)
  pieces <- lapply(names(pieces), function(method) {
    data <- pieces[[method]]
    if (is.null(data) || nrow(data) == 0L) {
      return(NULL)
    }
    data$interval_method <- method
    data
  })
  phase18_bind_tables(pieces)
}

phase18_interval_coverage_summary <- function(data) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data.frame())
  }
  data$source_surface <- phase18_missing_text(
    data,
    "source_surface",
    "unknown_surface"
  )
  data$interval_method <- phase18_missing_text(
    data,
    "interval_method",
    "unknown_method"
  )
  data$coverage <- if ("coverage" %in% names(data)) {
    suppressWarnings(as.numeric(data$coverage))
  } else {
    rep(NA_real_, nrow(data))
  }
  data$n_interval <- if ("n_interval" %in% names(data)) {
    suppressWarnings(as.numeric(data$n_interval))
  } else {
    rep(NA_real_, nrow(data))
  }
  data$n_covered <- if ("n_covered" %in% names(data)) {
    suppressWarnings(as.numeric(data$n_covered))
  } else {
    rep(NA_real_, nrow(data))
  }
  data$parameter <- phase18_missing_text(data, "parameter", "parameter")

  key <- paste(data$source_surface, data$interval_method, sep = "\r")
  groups <- split(seq_len(nrow(data)), key)
  out <- do.call(
    rbind,
    lapply(groups, function(index) {
      finite_interval <- is.finite(data$n_interval[index])
      finite_covered <- is.finite(data$n_covered[index])
      interval_total <- if (any(finite_interval)) {
        sum(data$n_interval[index][finite_interval])
      } else {
        NA_real_
      }
      covered_total <- if (any(finite_covered)) {
        sum(data$n_covered[index][finite_covered])
      } else {
        NA_real_
      }
      coverage <- if (is.finite(interval_total) && interval_total > 0) {
        covered_total / interval_total
      } else {
        mean(data$coverage[index], na.rm = TRUE)
      }
      if (!is.finite(coverage)) {
        coverage <- NA_real_
      }
      data.frame(
        source_surface = data$source_surface[index[1L]],
        interval_method = data$interval_method[index[1L]],
        n_parameter = length(unique(data$parameter[index])),
        n_interval_total = interval_total,
        n_covered_total = covered_total,
        coverage = coverage,
        stringsAsFactors = FALSE
      )
    })
  )
  row.names(out) <- NULL
  out[order(out$source_surface, out$interval_method), , drop = FALSE]
}

phase18_manifest_summary <- function(data) {
  if (is.null(data) || nrow(data) == 0L) {
    return(data.frame())
  }
  data$source_surface <- phase18_missing_text(
    data,
    "source_surface",
    "unknown_surface"
  )
  data$status <- phase18_missing_text(data, "status", "unknown_status")
  skipped <- if ("skipped" %in% names(data)) {
    as.logical(data$skipped)
  } else {
    rep(FALSE, nrow(data))
  }
  skipped[is.na(skipped)] <- FALSE
  warning_count <- if ("warning_count" %in% names(data)) {
    suppressWarnings(as.numeric(data$warning_count))
  } else {
    rep(0, nrow(data))
  }
  warning_count[!is.finite(warning_count)] <- 0
  error_text <- phase18_missing_text(data, "error", "")
  has_error <- nzchar(error_text) & error_text != "NA"
  elapsed <- if ("elapsed" %in% names(data)) {
    suppressWarnings(as.numeric(data$elapsed))
  } else {
    rep(NA_real_, nrow(data))
  }

  key <- paste(data$source_surface, data$status, sep = "\r")
  groups <- split(seq_len(nrow(data)), key)
  out <- do.call(
    rbind,
    lapply(groups, function(index) {
      elapsed_mean <- mean(elapsed[index], na.rm = TRUE)
      if (!is.finite(elapsed_mean)) {
        elapsed_mean <- NA_real_
      }
      data.frame(
        source_surface = data$source_surface[index[1L]],
        status = data$status[index[1L]],
        n_run = length(index),
        n_skipped = sum(skipped[index]),
        n_warning_run = sum(warning_count[index] > 0),
        warning_total = sum(warning_count[index]),
        n_error = sum(has_error[index]),
        mean_elapsed = elapsed_mean,
        stringsAsFactors = FALSE
      )
    })
  )
  row.names(out) <- NULL
  out[order(out$source_surface, out$status), , drop = FALSE]
}

artifact_status <- phase18_read_required_csv(
  params$artifact_status_csv,
  "artifact_status_csv"
)
artifact_grain_status <- phase18_read_optional_csv(
  params$artifact_grain_status_csv
)
aggregate <- phase18_read_optional_csv(params$aggregate_csv)
manifest <- phase18_read_optional_csv(params$manifest_csv)
failures <- phase18_read_optional_csv(params$failures_csv)
wald_coverage <- phase18_read_optional_csv(params$wald_coverage_csv)
profile_coverage <- phase18_read_optional_csv(params$profile_coverage_csv)
bootstrap_coverage <- phase18_read_optional_csv(params$bootstrap_coverage_csv)
interval_coverage <- phase18_bind_interval_coverage(
  wald_coverage,
  profile_coverage,
  bootstrap_coverage
)
interval_diagnostics <- phase18_read_optional_csv(params$interval_diagnostics_csv)
interval_failures <- phase18_read_optional_csv(params$interval_failures_csv)
replicate_cloud_gate <- phase18_replicate_cloud_gate(artifact_grain_status)

if (
  isTRUE(params$require_complete) &&
    "n_missing" %in% names(artifact_status) &&
    any(artifact_status$n_missing > 0L)
) {
  missing_surface <- artifact_status$surface[artifact_status$n_missing > 0L]
  stop(
    "Phase 18 artifact status has missing artifacts for: ",
    paste(missing_surface, collapse = ", "),
    call. = FALSE
  )
}
```

# Purpose

This first-wave summary report is a staging report for Phase 18 simulation
outputs. It brings artifact status, artifact-grain status, aggregate
operating-characteristic rows, interval diagnostics, and warning/error ledgers
into one reader-facing page before any publication-style figures or claims are
made.

# Notes

`r params$notes`

# Artifact Status

```{r artifact-status}
phase18_show_table(
  artifact_status,
  "No artifact-status CSV supplied.",
  c(
    "surface",
    "n_artifact",
    "n_present",
    "n_missing",
    "n_empty_csv",
    "n_total_csv_rows"
  )
)
```

# Artifact Grain Status

```{r artifact-grain-status}
phase18_show_table(
  artifact_grain_status,
  "No artifact-grain status CSV supplied.",
  c(
    "source_surface",
    "source_artifact",
    "artifact_grain",
    "grain_status",
    "plot_geometry",
    "replicate_cloud_allowed",
    "n_row"
  )
)
```

# Replicate Cloud Gate

```{r replicate-cloud-gate}
phase18_show_table(
  replicate_cloud_gate,
  "No replicate-cloud gate rows supplied.",
  c(
    "source_surface",
    "replicate_cloud_gate",
    "n_replicate_ready",
    "n_aggregate_only",
    "n_not_cloud_ready",
    "display_rule"
  )
)
```

# Aggregate Operating Characteristics

```{r aggregate}
phase18_show_table(
  aggregate,
  "No aggregate table supplied.",
  c(
    "source_surface",
    "source_artifact",
    "surface",
    "family",
    "parameter_class",
    "dpar",
    "term",
    "parameter",
    "n_replicate",
    "bias",
    "rmse",
    "convergence_rate",
    "pd_hessian_rate",
    "warning_rate"
  )
)
```

# Aggregate Bias Overview

```{r aggregate-bias-overview, fig.width=8.5, fig.height=5}
bias_overview <- phase18_bias_overview_data(aggregate)
if (!is.null(bias_overview)) {
  bias_overview <- phase18_add_replicate_cloud_gate(
    bias_overview,
    replicate_cloud_gate
  )
}
if (is.null(bias_overview)) {
  cat("No finite aggregate-bias rows to plot.\n")
} else if (!requireNamespace("ggplot2", quietly = TRUE)) {
  cat("Package `ggplot2` is required for the aggregate-bias overview.\n")
} else {
  n_total <- attr(bias_overview, "n_total")
  if (!is.null(n_total) && n_total > nrow(bias_overview)) {
    cat(
      "Showing first",
      nrow(bias_overview),
      "of",
      n_total,
      "finite aggregate-bias rows by absolute bias.\n\n"
    )
  }
  bias_plot <- ggplot2::ggplot(
    bias_overview,
    ggplot2::aes(
      x = bias,
      y = plot_label,
      colour = source_surface
    )
  ) +
    ggplot2::geom_vline(
      xintercept = 0,
      linetype = "dashed",
      linewidth = 0.5,
      colour = "grey55"
    ) +
    ggplot2::geom_point(size = 2.3, alpha = 0.85) +
    ggplot2::labs(
      x = "Mean estimate minus truth",
      y = "Aggregate row rank by absolute bias",
      colour = "Surface",
      caption = paste(
        "Full parameter names are listed below. Aggregate rows only;",
        "replicate-level clouds require",
        "`replicate_cloud_gate = replicate_clouds_allowed`."
      )
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.major.y = ggplot2::element_line(colour = "grey90"),
      panel.grid.minor = ggplot2::element_blank()
    )
  print(bias_plot)
  bias_columns <- c(
    "bias_rank",
    "source_surface",
    "cell_id",
    "parameter_label",
    "n_replicate",
    "bias",
    "rmse",
    "bias_mcse",
    "replicate_cloud_gate"
  )
  phase18_show_table(
    bias_overview[intersect(bias_columns, names(bias_overview))],
    "No finite aggregate-bias rows to list.",
    bias_columns
  )
}
```

# Interval Coverage Summary

```{r interval-coverage-summary}
phase18_show_table(
  phase18_interval_coverage_summary(interval_coverage),
  "No interval-coverage rows supplied.",
  c(
    "source_surface",
    "interval_method",
    "n_parameter",
    "n_interval_total",
    "n_covered_total",
    "coverage"
  )
)
```

# Interval Diagnostics

```{r interval-diagnostics}
phase18_show_table(
  interval_diagnostics,
  "No interval-diagnostics table supplied.",
  c(
    "source_surface",
    "source_artifact",
    "surface",
    "interval_method",
    "interval_status",
    "coverage",
    "n_interval",
    "n_covered",
    "n_failed"
  )
)
```

# Interval Failure Ledger

```{r interval-failures}
phase18_show_table(
  interval_failures,
  "No interval-failure table supplied.",
  c(
    "source_surface",
    "source_artifact",
    "surface",
    "interval_method",
    "interval_status",
    "parameter",
    "message"
  )
)
```

# Run Manifest Summary

```{r manifest-summary}
phase18_show_table(
  phase18_manifest_summary(manifest),
  "No run-manifest rows to summarise.",
  c(
    "source_surface",
    "status",
    "n_run",
    "n_skipped",
    "n_warning_run",
    "warning_total",
    "n_error",
    "mean_elapsed"
  )
)
```

# Run Manifest

```{r manifest}
phase18_show_table(
  manifest,
  "No run-manifest table supplied.",
  c(
    "source_surface",
    "source_artifact",
    "cell_id",
    "replicate",
    "status",
    "warning_count",
    "error",
    "elapsed"
  )
)
```

# Warning And Error Summary

```{r failure-summary}
phase18_show_table(
  phase18_failure_summary(failures),
  "No warning/error events to summarise.",
  c(
    "source_surface",
    "severity",
    "n_event",
    "n_unique_message",
    "example_message"
  )
)
```

# Warning And Error Ledger

```{r failures}
phase18_show_table(
  failures,
  "No warning/error ledger supplied.",
  c(
    "source_surface",
    "source_artifact",
    "cell_id",
    "replicate",
    "severity",
    "message"
  )
)
```

# Reader Checks

- Which surfaces are represented in the artifact-status and aggregate tables?
- Do any surfaces have missing or empty artifacts that change interpretation?
- Which artifacts are replicate-ready, aggregate-only, or missing grain tags
  before any replicate-cloud figures are drawn?
- Does each surface's replicate-cloud gate allow clouds, or should the report
  keep aggregate points, bars, and MCSE intervals only?
- Are convergence rate, Hessian rate, warning rate, and elapsed-time summaries
  visible for each estimand?
- Does the aggregate-bias overview point to rows that need replicate-level
  simulation displays?
- Which interval methods have coverage rows before deeper diagnostics are read?
- Which interval methods are usable evidence, failed diagnostics, or
  not-requested rows?
- Which warning/error severities recur by surface, and do they need a targeted
  diagnostic before the next larger grid?
- Do manifest summaries show skipped runs, warnings, errors, or elapsed-time
  outliers that should be handled before larger grids?
- Are these rows enough for a visual report, or does the surface need a larger
  grid first?

# Interpretation Boundary

This summary is a staging report. It can identify promising or broken
simulation surfaces, but it does not by itself establish final operating
characteristics. Formal claims still require the planned replicate counts,
Monte Carlo uncertainty, interval status review, and a Florence-reviewed visual
display.
