---
title: "Phase 18 Count Pilot Figure Gallery"
output:
  html_document:
    toc: true
params:
  aggregate_csv: null
  coverage_csv: null
  manifest_csv: null
  failures_csv: null
  notes: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5.4
)

phase18_read_optional_csv <- function(path) {
  if (
    is.null(path) ||
      length(path) != 1L ||
      is.na(path) ||
      !nzchar(path)
  ) {
    return(NULL)
  }
  read.csv(path)
}

aggregate <- phase18_read_optional_csv(params$aggregate_csv)
coverage <- phase18_read_optional_csv(params$coverage_csv)
manifest <- phase18_read_optional_csv(params$manifest_csv)
failures <- phase18_read_optional_csv(params$failures_csv)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

phase18_count_gallery_palette <- c(Poisson = "#0072B2", NB2 = "#D55E00")
phase18_count_gallery_class_palette <- c(
  fixed_effect = "#0072B2",
  random_sd = "#009E73"
)
phase18_count_gallery_class_labels <- c(
  fixed_effect = "Fixed effect",
  random_sd = "Random-effect SD"
)
phase18_count_gallery_term_labels <- c(
  "(Intercept)" = "intercept",
  "x" = "x slope",
  "z" = "z slope",
  "(0 + x | id)" = "x random-slope SD",
  "(1 | id)" = "intercept SD"
)

phase18_count_gallery_theme <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.major.y = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold"),
      plot.title.position = "plot",
      plot.title = ggplot2::element_text(face = "bold", size = 14),
      plot.subtitle = ggplot2::element_text(size = 10, colour = "grey30"),
      plot.caption.position = "plot",
      plot.caption = ggplot2::element_text(size = 8, hjust = 0),
      plot.margin = ggplot2::margin(6, 8, 14, 8)
    )
}

phase18_count_gallery_terms <- function(x) {
  if ("family" %in% names(x)) {
    x$family <- factor(x$family, levels = names(phase18_count_gallery_palette))
  }
  if ("parameter_class" %in% names(x)) {
    x$parameter_class <- factor(
      x$parameter_class,
      levels = names(phase18_count_gallery_class_palette)
    )
  }
  x$term <- ifelse(
    x$term %in% names(phase18_count_gallery_term_labels),
    phase18_count_gallery_term_labels[x$term],
    x$term
  )
  x$term <- factor(x$term, levels = rev(unique(x$term)))
  x
}

phase18_count_gallery_add_mcse_interval <- function(data, metric) {
  mcse <- paste0(metric, "_mcse")
  low <- paste0(metric, "_low")
  high <- paste0(metric, "_high")
  if (!(mcse %in% names(data))) {
    return(data)
  }
  z <- stats::qnorm(0.975)
  data[[low]] <- data[[metric]] - z * data[[mcse]]
  data[[high]] <- data[[metric]] + z * data[[mcse]]
  if (identical(metric, "rmse")) {
    data[[low]] <- pmax(0, data[[low]])
  }
  data
}

phase18_plot_count_bias <- function(data) {
  data <- phase18_count_gallery_terms(data)
  data <- phase18_count_gallery_add_mcse_interval(data, "bias")
  p <- ggplot2::ggplot(
    data,
    ggplot2::aes(
      x = bias,
      y = term,
      colour = parameter_class,
      shape = parameter_class
    )
  ) +
    ggplot2::geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey55")
  if (all(c("bias_low", "bias_high") %in% names(data))) {
    p <- p + ggplot2::geom_segment(
      ggplot2::aes(x = bias_low, xend = bias_high, yend = term),
      linewidth = 0.75,
      alpha = 0.8,
      show.legend = FALSE,
      na.rm = TRUE
    )
  }
  p +
    ggplot2::geom_point(size = 2.7, alpha = 0.95, na.rm = TRUE) +
    ggplot2::facet_grid(
      parameter_class ~ family,
      scales = "free_y",
      space = "free_y",
      labeller = ggplot2::labeller(
        parameter_class = phase18_count_gallery_class_labels
      )
    ) +
    ggplot2::scale_colour_manual(
      values = phase18_count_gallery_class_palette,
      labels = phase18_count_gallery_class_labels
    ) +
    ggplot2::scale_shape_discrete(labels = phase18_count_gallery_class_labels) +
    ggplot2::labs(
      title = "Bias by estimand",
      subtitle = paste(
        "Aggregate mean errors are shown here;",
        "add replicate-error clouds when replicate-level output is supplied."
      ),
      x = "Mean estimate minus truth",
      y = NULL,
      colour = "Parameter class",
      shape = "Parameter class",
      caption = paste(
        "Bars show 95% MCSE when bias_mcse is available;",
        "inspect the manifest and failure ledger before interpretation."
      )
    ) +
    phase18_count_gallery_theme()
}

phase18_plot_count_rmse <- function(data) {
  data <- phase18_count_gallery_terms(data)
  data <- phase18_count_gallery_add_mcse_interval(data, "rmse")
  p <- ggplot2::ggplot(
    data,
    ggplot2::aes(
      x = rmse,
      y = term,
      colour = parameter_class,
      shape = parameter_class
    )
  )
  if (all(c("rmse_low", "rmse_high") %in% names(data))) {
    p <- p + ggplot2::geom_segment(
      ggplot2::aes(x = rmse_low, xend = rmse_high, yend = term),
      linewidth = 0.75,
      alpha = 0.8,
      show.legend = FALSE,
      na.rm = TRUE
    )
  }
  p +
    ggplot2::geom_point(size = 2.7, alpha = 0.95, na.rm = TRUE) +
    ggplot2::facet_grid(
      parameter_class ~ family,
      scales = "free_y",
      space = "free_y",
      labeller = ggplot2::labeller(
        parameter_class = phase18_count_gallery_class_labels
      )
    ) +
    ggplot2::scale_colour_manual(
      values = phase18_count_gallery_class_palette,
      labels = phase18_count_gallery_class_labels
    ) +
    ggplot2::scale_shape_discrete(labels = phase18_count_gallery_class_labels) +
    ggplot2::labs(
      title = "RMSE needs its own uncertainty display",
      subtitle = "Root mean-square error is an aggregate summary, not the center of an absolute-error cloud.",
      x = "Root mean-square error",
      y = NULL,
      colour = "Parameter class",
      shape = "Parameter class",
      caption = paste(
        "Bars show 95% MCSE when rmse_mcse is available;",
        "bias, RMSE, and coverage answer different simulation questions."
      )
    ) +
    phase18_count_gallery_theme()
}

phase18_plot_count_coverage <- function(data) {
  data <- phase18_count_gallery_terms(data)
  if ("coverage_mcse" %in% names(data)) {
    data$coverage_low <- pmax(0, data$coverage - 1.96 * data$coverage_mcse)
    data$coverage_high <- pmin(1, data$coverage + 1.96 * data$coverage_mcse)
  }
  coverage_values <- data$coverage
  if ("coverage_low" %in% names(data)) {
    coverage_values <- c(coverage_values, data$coverage_low)
  }
  coverage_floor <- min(coverage_values, na.rm = TRUE)
  if (!is.finite(coverage_floor)) {
    coverage_floor <- 0.90
  }
  coverage_floor <- max(0, min(0.90, coverage_floor - 0.02))
  data$coverage_panel <- paste(
    phase18_count_gallery_class_labels[as.character(data$parameter_class)],
    data$dpar,
    sep = ": "
  )
  data$coverage_panel <- factor(data$coverage_panel, levels = unique(data$coverage_panel))
  point_position <- ggplot2::position_jitter(width = 0, height = 0.06)
  p <- ggplot2::ggplot(
    data,
    ggplot2::aes(
      x = coverage,
      y = term,
      colour = family,
      shape = interval_method,
      group = interaction(family, interval_method)
    )
  ) +
    ggplot2::geom_vline(xintercept = 0.95, linewidth = 0.3, colour = "grey55")
  if (all(c("coverage_low", "coverage_high") %in% names(data))) {
    p <- p + ggplot2::geom_segment(
      ggplot2::aes(x = coverage_low, xend = coverage_high, yend = term),
      alpha = 0.65,
      linewidth = 0.5,
      show.legend = FALSE
    )
  }
  p +
    ggplot2::geom_point(position = point_position, size = 2.7, alpha = 0.9) +
    ggplot2::facet_wrap(
      ~coverage_panel,
      scales = "free_y",
      ncol = 1
    ) +
    ggplot2::coord_cartesian(xlim = c(coverage_floor, 1.01), clip = "off") +
    ggplot2::scale_x_continuous(
      breaks = c(0.90, 0.95, 1.00),
      labels = c("0.90", "0.95", "1.00")
    ) +
    ggplot2::scale_colour_manual(values = phase18_count_gallery_palette) +
    ggplot2::labs(
      title = "Interval coverage",
      x = "Empirical coverage",
      y = NULL,
      colour = "Family",
      shape = "Interval method",
      caption = paste(
        "The grey line marks nominal 95% coverage;",
        "horizontal ranges show approximate 95% Monte Carlo uncertainty when available."
      )
    ) +
    phase18_count_gallery_theme()
}
```

# Purpose

This gallery is for the first paired count-family Phase 18 pilot. It compares
the ready Poisson and NB2 `mu` random-effect surfaces using plot-ready output
from `phase18_count_mu_re_plot_data()`.

It is not a final simulation report. Use it to inspect bias, RMSE, interval
coverage, convergence status, and warning/error patterns before increasing the
grid size or replicate count.

Notes: `r params$notes`

# Bias

```{r bias-plot}
if (is.null(aggregate)) {
  cat("No aggregate CSV supplied.\n")
} else if (!has_ggplot2) {
  aggregate
} else {
  phase18_plot_count_bias(aggregate)
}
```

# RMSE

```{r rmse-plot}
if (is.null(aggregate)) {
  cat("No aggregate CSV supplied.\n")
} else if (!has_ggplot2) {
  aggregate
} else {
  phase18_plot_count_rmse(aggregate)
}
```

# Interval Coverage

```{r coverage-plot, fig.height=6.1}
if (is.null(coverage)) {
  cat("No coverage CSV supplied.\n")
} else if (!has_ggplot2) {
  coverage
} else {
  phase18_plot_count_coverage(coverage)
}
```

# Run Manifest

```{r manifest-table}
if (is.null(manifest)) {
  cat("No manifest CSV supplied.\n")
} else {
  manifest
}
```

# Warning And Error Ledger

```{r failures-table}
if (is.null(failures)) {
  cat("No warning/error ledger CSV supplied.\n")
} else if (nrow(failures) == 0L) {
  cat("No warnings or errors recorded.\n")
} else {
  failures
}
```

# Florence Checks

- Are the Poisson and NB2 panels visually comparable without hiding family
  differences?
- Are bias, RMSE, and coverage interpreted separately?
- Are fixed-effect and random-effect SD targets visually distinguishable?
- Are failed fits, warnings, and missing intervals visible before interpretation?
- Does the caption or surrounding text say this is a pilot, not a final power
  or coverage claim?
- Does bias use replicate-error clouds only when replicate-level errors are
  available, and otherwise show aggregate means with MCSE?
- Does RMSE stay separate from signed bias and avoid looking like the center of
  an absolute-error cloud?
