---
title: "Phase 18 Count Pilot Figure Gallery"
output:
  html_document:
    toc: true
params:
  aggregate_csv: null
  coverage_csv: null
  manifest_csv: null
  failures_csv: null
  notes: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8
)

phase18_read_optional_csv <- function(path) {
  if (
    is.null(path) ||
      length(path) != 1L ||
      is.na(path) ||
      !nzchar(path)
  ) {
    return(NULL)
  }
  read.csv(path)
}

aggregate <- phase18_read_optional_csv(params$aggregate_csv)
coverage <- phase18_read_optional_csv(params$coverage_csv)
manifest <- phase18_read_optional_csv(params$manifest_csv)
failures <- phase18_read_optional_csv(params$failures_csv)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

# Purpose

This gallery is for the first paired count-family Phase 18 pilot. It compares
the ready Poisson and NB2 `mu` random-effect surfaces using plot-ready output
from `phase18_count_mu_re_plot_data()`.

It is not a final simulation report. Use it to inspect bias, RMSE, interval
coverage, convergence status, and warning/error patterns before increasing the
grid size or replicate count.

Notes: `r params$notes`

# Bias

```{r bias-plot}
if (is.null(aggregate)) {
  cat("No aggregate CSV supplied.\n")
} else if (!has_ggplot2) {
  aggregate
} else {
  ggplot2::ggplot(
    aggregate,
    ggplot2::aes(
      x = term,
      y = bias,
      colour = family,
      shape = parameter_class
    )
  ) +
    ggplot2::geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey55") +
    ggplot2::geom_point(size = 2.5, alpha = 0.9) +
    ggplot2::facet_wrap(~dpar, scales = "free_x") +
    ggplot2::scale_colour_manual(values = c(Poisson = "#0072B2", NB2 = "#D55E00")) +
    ggplot2::labs(
      x = NULL,
      y = "Bias",
      colour = "Family",
      shape = "Parameter class"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
    )
}
```

# RMSE

```{r rmse-plot}
if (is.null(aggregate)) {
  cat("No aggregate CSV supplied.\n")
} else if (!has_ggplot2) {
  aggregate
} else {
  ggplot2::ggplot(
    aggregate,
    ggplot2::aes(
      x = term,
      y = rmse,
      colour = family,
      shape = parameter_class
    )
  ) +
    ggplot2::geom_point(size = 2.5, alpha = 0.9) +
    ggplot2::facet_wrap(~dpar, scales = "free_x") +
    ggplot2::scale_colour_manual(values = c(Poisson = "#0072B2", NB2 = "#D55E00")) +
    ggplot2::labs(
      x = NULL,
      y = "RMSE",
      colour = "Family",
      shape = "Parameter class"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
    )
}
```

# Interval Coverage

```{r coverage-plot}
if (is.null(coverage)) {
  cat("No coverage CSV supplied.\n")
} else if (!has_ggplot2) {
  coverage
} else {
  ggplot2::ggplot(
    coverage,
    ggplot2::aes(
      x = term,
      y = coverage,
      colour = family,
      shape = interval_method
    )
  ) +
    ggplot2::geom_hline(yintercept = 0.95, linewidth = 0.3, colour = "grey55") +
    ggplot2::geom_point(size = 2.5, alpha = 0.9) +
    ggplot2::facet_grid(interval_method ~ dpar, scales = "free_x") +
    ggplot2::coord_cartesian(ylim = c(0, 1)) +
    ggplot2::scale_colour_manual(values = c(Poisson = "#0072B2", NB2 = "#D55E00")) +
    ggplot2::labs(
      x = NULL,
      y = "Coverage",
      colour = "Family",
      shape = "Interval method"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
    )
}
```

# Run Manifest

```{r manifest-table}
if (is.null(manifest)) {
  cat("No manifest CSV supplied.\n")
} else {
  manifest
}
```

# Warning And Error Ledger

```{r failures-table}
if (is.null(failures)) {
  cat("No warning/error ledger CSV supplied.\n")
} else if (nrow(failures) == 0L) {
  cat("No warnings or errors recorded.\n")
} else {
  failures
}
```

# Florence Checks

- Are the Poisson and NB2 panels visually comparable without hiding family
  differences?
- Are bias, RMSE, and coverage interpreted separately?
- Are fixed-effect and random-effect SD targets visually distinguishable?
- Are failed fits, warnings, and missing intervals visible before interpretation?
- Does the caption or surrounding text say this is a pilot, not a final power
  or coverage claim?
