---
title: "Phase 18 Smoke Simulation Report"
output:
  html_document:
    toc: true
params:
  surface: "gaussian_ls"
  aggregate_csv: null
  notes: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Purpose

This smoke report is for checking that a Phase 18 surface can run from data
generation to fitted summaries and grouped operating-characteristic tables. It
is not a comprehensive simulation report and should not be used for power or
coverage claims without the larger grid, Monte Carlo uncertainty, and interval
construction slices.

# Surface

Surface: `r params$surface`

Notes: `r params$notes`

# Aggregate Summary

```{r aggregate-summary, eval=FALSE}
aggregate <- read.csv(params$aggregate_csv)
aggregate
```

# Reader Checks

- What is the biological or methodological question?
- What data-generating model produced the simulated response?
- Which estimand does each row target?
- How many replicates contributed to each aggregate row?
- Are failures, warnings, convergence rates, and Hessian rates visible?
- Are Monte Carlo uncertainty columns present for reported operating
  characteristics?

# Interpretation Boundary

Use this report to check that the simulation machinery is coherent and
readable. Move to a full Phase 18 report only after the surface has enough
replicates, explicit interval targets, MCSEs, and failure-mode summaries.
