---
title: "Phase 18 First-Wave Artifact Status"
output:
  html_document:
    toc: true
params:
  artifact_manifest_csv: null
  artifact_status_csv: null
  require_complete: false
  notes: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

phase18_read_required_csv <- function(path, label) {
  if (
    is.null(path) ||
      length(path) != 1L ||
      is.na(path) ||
      !nzchar(path)
  ) {
    stop("`", label, "` must be supplied.", call. = FALSE)
  }
  if (!file.exists(path)) {
    stop("`", label, "` does not exist: ", path, call. = FALSE)
  }
  read.csv(path)
}

phase18_assert_columns <- function(data, label, required) {
  missing <- setdiff(required, names(data))
  if (length(missing) > 0L) {
    stop(
      "`",
      label,
      "` is missing ",
      paste(missing, collapse = ", "),
      ".",
      call. = FALSE
    )
  }
  invisible(data)
}

artifact_manifest <- phase18_read_required_csv(
  params$artifact_manifest_csv,
  "artifact_manifest_csv"
)
artifact_status <- phase18_read_required_csv(
  params$artifact_status_csv,
  "artifact_status_csv"
)
phase18_assert_columns(
  artifact_manifest,
  "artifact_manifest_csv",
  c("surface", "artifact", "path", "exists", "n_row")
)
phase18_assert_columns(
  artifact_status,
  "artifact_status_csv",
  c("surface", "n_artifact", "n_present", "n_missing", "n_empty_csv")
)

missing_surface <- artifact_status$surface[artifact_status$n_missing > 0L]
if (isTRUE(params$require_complete) && length(missing_surface) > 0L) {
  stop(
    "Phase 18 artifact status has missing artifacts for: ",
    paste(missing_surface, collapse = ", "),
    call. = FALSE
  )
}
```

# Purpose

This first-wave status report checks that Phase 18 grid-output artifacts exist
before a reader-facing simulation report consumes aggregate, replicate,
interval, or warning/error tables. It is an artifact preflight, not a model
accuracy, power, or coverage report.

# Notes

`r params$notes`

# Surface Status

```{r surface-status}
artifact_status
```

# Artifact Manifest

```{r artifact-manifest}
artifact_manifest
```

# Missing Or Empty Artifacts

```{r missing-empty}
subset(
  artifact_manifest,
  !exists | (!is.na(n_row) & n_row == 0L)
)
```

# Reader Checks

- Are all required grid-output CSVs present before the report reads them?
- Are empty CSVs expected optional ledgers, such as no failures, or evidence of
  an incomplete run?
- Which surfaces have enough aggregate, replicate, interval, and failure-ledger
  rows for the next report?
- Which artifacts should be regenerated before a public figure or table is
  interpreted?

# Interpretation Boundary

This status page only verifies artifact availability and row counts. The
statistical report must still inspect model convergence, Hessian status,
interval status, Monte Carlo uncertainty, and warning/error ledgers before
claiming operating characteristics.
