---
title: "Function map and cheat sheet"
description: "Choose the next drmTMB task: specify, fit, check, interpret, predict, or quantify uncertainty."
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Function map and cheat sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE
)
if (!"package:drmTMB" %in% search()) {
  library(drmTMB)
}
inventory_candidates <- c(
  file.path("..", "..", "tools", "function-map-inventory.R"),
  file.path("tools", "function-map-inventory.R")
)
inventory_file <- inventory_candidates[file.exists(inventory_candidates)][1]
if (is.na(inventory_file)) {
  stop("Cannot locate tools/function-map-inventory.R")
}
get("source", envir = baseenv())(normalizePath(inventory_file, mustWork = TRUE))
```

<style>
.drmtmb-map {
  --ink: #173042;
  --muted: #536875;
  margin: 1.5rem 0;
  padding: 1rem;
  background: #f7fafb;
  border: 1px solid #d9e3e8;
  border-radius: 16px;
  box-shadow: 0 8px 24px rgba(23, 48, 66, .08);
  hyphens: none;
  -webkit-hyphens: none;
}
.drmtmb-map *, .drmtmb-downloads * { box-sizing: border-box; }
.drmtmb-map__header { margin-bottom: .9rem; }
.drmtmb-map__header h2 { margin: 0; color: var(--ink); font-size: 1.35rem; }
.drmtmb-map__header p, .drmtmb-map__caption { margin: .2rem 0 0; color: var(--muted); }
.drmtmb-map__grid {
  display: grid;
  grid-template-columns: repeat(3, minmax(0, 1fr));
  gap: .8rem;
}
.drmtmb-map__card {
  min-width: 0;
  padding: .8rem;
  border: 1.5px solid var(--accent);
  border-radius: 12px;
  background: var(--wash);
  hyphens: none;
  -webkit-hyphens: none;
}
.drmtmb-map__card h3 { margin: 0 0 .45rem; color: var(--accent-dark); font-size: .98rem; }
.drmtmb-map__card p { margin: .35rem 0; color: var(--ink); line-height: 1.42; }
.drmtmb-map__card small { display: block; margin-top: .45rem; color: var(--muted); line-height: 1.35; }
.drmtmb-map__card a { color: var(--accent-dark); font-weight: 700; text-decoration-thickness: .08em; }
.drmtmb-map__functions code {
  white-space: normal;
  overflow-wrap: normal;
  word-break: normal;
  hyphens: none;
  color: var(--accent-dark);
}
.drmtmb-map__function { display: inline; font-weight: 600; }
.drmtmb-map__card a:focus-visible, .drmtmb-download:focus-visible {
  outline: 3px solid #052b3f;
  outline-offset: 3px;
}
.map-specify { --accent: #75a843; --accent-dark: #476f1f; --wash: #f4f8ef; }
.map-fit { --accent: #4e86bf; --accent-dark: #245c91; --wash: #eff5fb; }
.map-check { --accent: #8464ad; --accent-dark: #5b3c82; --wash: #f5f1fa; }
.map-interpret { --accent: #df8a45; --accent-dark: #9a501d; --wash: #fdf5ec; }
.map-predict { --accent: #31a0a0; --accent-dark: #126e70; --wash: #eff8f7; }
.map-uncertainty { --accent: #c56b8c; --accent-dark: #8f3e5e; --wash: #fbf1f5; }
.drmtmb-map__route {
  margin: .9rem 0 0;
  padding: .7rem .8rem;
  border: 1px dashed #9eafb8;
  border-radius: 9px;
  background: #fff;
  color: var(--ink);
}
.drmtmb-map__route strong { color: #052b3f; }
.drmtmb-downloads {
  display: grid;
  grid-template-columns: repeat(2, minmax(0, 1fr));
  gap: .8rem;
  margin: 1rem 0 1.25rem;
}
.drmtmb-download {
  display: block;
  padding: .85rem 1rem;
  border: 1px solid #bdd1d8;
  border-left: 5px solid #0f7887;
  border-radius: 10px;
  color: #173042;
  background: #f5f9fa;
  text-decoration: none;
}
.drmtmb-download:hover { background: #ebf4f6; color: #173042; text-decoration: none; }
.drmtmb-download strong, .drmtmb-download small { display: block; }
.drmtmb-download small { margin-top: .2rem; color: #536875; }
.drmtmb-route-table {
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.drmtmb-route-table table {
  min-inline-size: 46rem;
  margin-block-end: 0;
}
@media (max-width: 760px) {
  .drmtmb-map__grid, .drmtmb-downloads { grid-template-columns: 1fr; }
}
</style>

Use this guide when `drmTMB` is installed and you know the scientific task you
want to do next, but not which function starts it. For installation and a
guided first fit, read the fuller [package introduction](drmTMB.html). If you
already know the syntax and need to verify support, go directly to [What can I
fit today?](model-map.html).

This page is a navigation aid, not a claim that every displayed function,
family, structured effect, or interval route is suitable for every model.

`drmTMB` can model location (a family-specific centre or location parameter;
for Gaussian models it is the expected response), scale (residual
variability), shape (distribution features beyond location and scale), and
coscale, meaning residual correlation such as `rho12` between two responses.
Not every family exposes every parameter, and a location parameter need not be
the unconditional response mean.

```{=html}
<div class="drmtmb-downloads" aria-label="Download printable drmTMB function guides">
  <a class="drmtmb-download" href="../cheatsheets/drmTMB-function-map.pdf" download>
    <strong>Download the visual function map (PDF)</strong>
    <small>One landscape page with the six task modules and their key functions.</small>
  </a>
  <a class="drmtmb-download" href="../cheatsheets/drmTMB-function-cheatsheet.pdf" download>
    <strong>Download the one-page function cheat sheet (PDF)</strong>
    <small>A compact landscape quick reference generated from the same task inventory.</small>
  </a>
</div>
<p><small>The linked HTML map below is the accessible version with navigable
links and selectable text. The untagged PDFs are print companions.</small></p>
```

## Find the next step

The ordinary route is **Specify model → Fit → Check fit health**. After checking
the fit, choose the branch that answers the scientific question: interpret
parameters, predict and assess, or quantify uncertainty and simulate. If the
fit is not healthy, revise the specification and fit again. Structured terms
such as `phylo()`, `spatial()`, and `relmat()` belong in the model
specification; they are not a later analysis stage. The `gaussian()` and
`nbinom2()` labels in the first box are examples of family constructors, not an
exhaustive family list.

```{=html}
<section class="drmtmb-map" role="figure" aria-label="drmTMB function map from model specification through fitting and fit checks to interpretation, prediction, uncertainty, and simulation">
  <div class="drmtmb-map__header">
    <h2>drmTMB function map</h2>
    <p>Choose the scientific task first; then follow the relevant function and guide.</p>
  </div>
  <div class="drmtmb-map__grid">
```

```{r function-map-cards, echo = FALSE, results = 'asis'}
cat(function_map_html())
```

```{=html}
  </div>
  <p class="drmtmb-map__route"><strong>Typical route:</strong> Specify → Fit → Check fit health. Then choose the branch that answers your question: interpretation, prediction and assessment, or uncertainty and simulation.</p>
  <p class="drmtmb-map__caption">The map's headings, function names, links, and reading order carry the meaning without relying on colour or arrows.</p>
</section>
```

The **Predict & assess** branch checks fitted-response and fixed-effect
distributional form. It does not by itself establish adequacy of every random
or structured component; follow [Distributional outputs and
adequacy](distributional-outputs-and-adequacy.html) for that boundary.

## Choose a route

<div class="drmtmb-route-table" role="region" aria-label="Scrollable drmTMB task route table" tabindex="0">

| Goal | Start with | Read next |
| --- | --- | --- |
| Learn the model grammar and fit a first model | `bf()`, a family object, `drmTMB()` | [Full package introduction](drmTMB.html) |
| Check numerical fit health before interpreting | `check_drm()`, `summary()` | [Model workflow](model-workflow.html) |
| Resolve an error, warning, or unstable fit | `check_drm()`, the original warning, and the fitted object | [Errors, warnings, and convergence](convergence.html) |
| Extract a parameter or make a prediction table | `coef()`, `fixef()`, `predict_parameters()` | [Distributional outputs and adequacy](distributional-outputs-and-adequacy.html) |
| Study latent-normal association for a paired-outcome class admitted by the capability guide | `biv_associate()`, `association()`, `vcov()`, `confint()` | [Association between outcome pairs](cross-family.html) |
| Work with repeated, related, or spatially structured data | `bf()` with `phylo()`, `spatial()`, or `relmat()`; then `ranef()` | [Structural dependence overview](structural-dependence.html) |
| Request an interval or profile a target | `confint()`, `profile_targets()`, `profile()` | [Checking and using fitted models](model-workflow.html) |
| Fit when some values are missing | `miss_control()`, `mi()`, `imputed()` | [Handling missing data](missing-data.html) |
| Confirm an unfamiliar route is supported | the model and capability guides | [What can I fit today?](model-map.html) |

</div>

## The shortest useful workflow

Suppose an ecologist asks whether an environmental gradient changes both mean
growth and its predictability. For a continuous response, write one formula
for location `mu` and one for residual scale `sigma`, then check the fit before
interpreting it.

```{r shortest-workflow}
set.seed(20260721)
n <- 80
x <- rnorm(n)
dat <- data.frame(
  y = 0.4 + 0.7 * x + rnorm(n, sd = exp(-0.3 + 0.2 * x)),
  x = x
)

fit <- drmTMB(
  bf(y ~ x, sigma ~ x),
  family = gaussian(),
  data = dat
)

fit_checks <- check_drm(fit)
attr(fit_checks, "ok")
table(fit_checks$status)
```

The logical result is the first health verdict. The small status table shows
whether any diagnostic needs attention; inspect every `warning` or `error` row
before continuing.

Here `mu` is expected growth and `sigma` is residual SD. The `sigma`
coefficients use a log-SD link: exponentiating the slope gives the residual-SD
ratio per one-unit increase in `x`. Keep this residual scale separate from the
SD of a group-level random effect.

Make response-scale predictions on an explicit grid:

```{r coefficient-and-prediction}
coef(fit, dpar = "mu")
coef(fit, dpar = "sigma")

grid <- data.frame(x = c(-1, 0, 1))
predict_parameters(
  fit,
  newdata = grid,
  dpar = c("mu", "sigma"),
  conf.int = TRUE
)
```

The prediction table records interval provenance in `conf.status` and
`interval_source`. If `check_drm()` reports a warning or error, or an interval
status is unavailable, follow [Errors, warnings, and
convergence](convergence.html) before reporting the result. A fixed-effect
Wald interval does not establish calibrated coverage for a random-effect SD.

If fitting stops before an object is returned, read the original error and
confirm the exact family-component combination in [What can I fit
today?](model-map.html). If a fit object is returned, run `check_drm(fit)`,
address the flagged optimizer, gradient, Hessian, or design problem, then refit
and rerun the check before interpreting.

## Keep the boundaries visible

`drmTMB()` is the primary R workflow shown here. Julia fitting remains
future/deferred work; the current Julia methods provide compatibility for
legacy Julia-engine objects and are not a route for a new analysis.

This page chooses a route; it does not establish a claim tier. Before a
substantive analysis, confirm the exact family, model component, and inference
boundary in [What can I fit today?](model-map.html) and [What can I
trust?](capability-and-limits.html).
If the exact combination is absent or marked planned or diagnostic-only, do
not treat a successful fit as supported inference; choose an implemented route
or report the limitation.
